\section{2010 Probability and Statistics}

\subsection{Individual Exam}

\begin{exercise}
Let $Z_1, \dots, Z_n$ be i.i.d. random variables with $Z_i \sim N(\mu, \sigma^2)$. Find
\[ E \left( \sum_{i=1}^n Z_i \mid Z_1 - Z_2 + Z_3 \right). \]
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Linear Combinations of Normal Variables:} A linear combination of independent normal variables is also normally distributed.
    \item \textbf{Bivariate Normal Distribution:} Two random variables that are both linear combinations of the same set of independent normal variables have a joint bivariate normal distribution.
    \item \textbf{Conditional Expectation Formula:} For a bivariate normal vector $(X, Y)$, the conditional expectation is given by $E[X|Y] = E[X] + \frac{\text{Cov}(X, Y)}{\text{Var}(Y)} (Y - E[Y])$.
\end{itemize}

\textbf{Intuition / Motivation:}

We are asked to find the expected value of the total sum $X = \sum_{i=1}^n Z_i$, given the observed value of a specific combination $Y = Z_1 - Z_2 + Z_3$. Because all variables are i.i.d. normal, $X$ and $Y$ are jointly normal. This transforms the problem into finding the conditional mean of one normal variable given another. By finding the individual means, the variance of the condition $Y$, and the covariance between $X$ and $Y$, we can plug these directly into the standard projection formula for bivariate normal distributions.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Define the variables and establish joint normality.}

Let $X = \sum_{i=1}^n Z_i$ be the target sum, and $Y = Z_1 - Z_2 + Z_3$ be the conditioning variable. Since $X$ and $Y$ are linear combinations of the i.i.d. normal variables $Z_i \sim N(\mu, \sigma^2)$, the vector $(X, Y)$ follows a bivariate normal distribution.

\textbf{Step 2: State the formula for conditional expectation.}

For bivariate normal variables, the conditional expectation follows the linear regression formula:
\[ E[X|Y] = E[X] + \frac{\text{Cov}(X, Y)}{\text{Var}(Y)} (Y - E[Y]) \]
To evaluate this, we must compute $E[X]$, $E[Y]$, $\text{Var}(Y)$, and $\text{Cov}(X, Y)$.

\textbf{Step 3: Compute the expectations $E[X]$ and $E[Y]$.}

Using the linearity of expectation:
\[ E[X] = E\left[\sum_{i=1}^n Z_i\right] = \sum_{i=1}^n E[Z_i] = n\mu \]
Similarly for $Y$:
\[ E[Y] = E[Z_1 - Z_2 + Z_3] = E[Z_1] - E[Z_2] + E[Z_3] = \mu - \mu + \mu = \mu \]

\textbf{Step 4: Compute the variance of $Y$.}

Since the $Z_i$ are independent, the variance of their sum or difference is the sum of their individual variances. Recall that $\text{Var}(aZ) = a^2\text{Var}(Z)$:
\[ \text{Var}(Y) = \text{Var}(Z_1 - Z_2 + Z_3) = \text{Var}(Z_1) + (-1)^2\text{Var}(Z_2) + \text{Var}(Z_3) = \sigma^2 + \sigma^2 + \sigma^2 = 3\sigma^2 \]

\textbf{Step 5: Compute the covariance between $X$ and $Y$.}

We use the bilinear property of covariance. Since $Z_i$ are independent, $\text{Cov}(Z_i, Z_j) = 0$ for $i \neq j$.
\begin{align*}
\text{Cov}(X, Y) &= \text{Cov}\left(\sum_{i=1}^n Z_i, Z_1 - Z_2 + Z_3\right) \\
&= \text{Cov}(Z_1, Z_1) + \text{Cov}(Z_2, -Z_2) + \text{Cov}(Z_3, Z_3) + \sum_{i=4}^n \text{Cov}(Z_i, Y) \\
&= \text{Var}(Z_1) - \text{Var}(Z_2) + \text{Var}(Z_3) + 0 \\
&= \sigma^2 - \sigma^2 + \sigma^2 = \sigma^2
\end{align*}

\textbf{Step 6: Substitute the components into the conditional expectation formula.}

Plugging in our results from Steps 3, 4, and 5:
\[ E[X|Y] = n\mu + \frac{\sigma^2}{3\sigma^2} (Y - \mu) = n\mu + \frac{1}{3}(Z_1 - Z_2 + Z_3 - \mu) \]
\end{solution}

\begin{exercise}
Let $X_1, \dots, X_n$ be pairwise independent. Further, assume that $E X_i = 1 + i^{-1}$ and that $\max_{1 \leq i \leq n} E |X_i|^{1+\epsilon} < \infty$ for some $\epsilon > 0$. Show that
\[ \frac{1}{n} \sum_{i=1}^n X_i \xrightarrow{P} 1. \]
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Convergence in Probability ($\xrightarrow{P}$):} A sequence of random variables $Y_n$ converges in probability to a constant $c$ if, for any $\delta > 0$, $\lim_{n\to\infty} P(|Y_n - c| > \delta) = 0$.
    \item \textbf{Harmonic Series Growth:} The $n$-th harmonic number $H_n = \sum_{i=1}^n \frac{1}{i}$ grows asymptotically like $\log n$.
    \item \textbf{Uniform Integrability:} A condition on a family of random variables ensuring that their tails are uniformly bounded. It is guaranteed if there is a uniform bound on a higher moment $E[|X_i|^{1+\epsilon}] \le M$.
    \item \textbf{Weak Law of Large Numbers (WLLN) for Pairwise Independence:} For pairwise independent variables that are uniformly integrable, the sample average converges in probability to its expected value.
\end{itemize}

\textbf{Intuition / Motivation:}

To prove that the sample average $\bar{X}_n = \frac{1}{n} \sum_{i=1}^n X_i$ converges to 1 in probability, we need to show two things: first, that the "target" (expected value) of the sample average gets arbitrarily close to 1 as $n$ grows; and second, that the random fluctuations of the average around this target shrink to zero. We handle the first part using the linearity of expectation and knowing how the harmonic series scales. For the second part, because the variables are only pairwise independent (not mutually independent), we rely on their bounded $(1+\epsilon)$-moments which guarantees uniform integrability and allows us to apply a generalized version of the Weak Law of Large Numbers.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Compute the expected value of the sample average.}

Let $\bar{X}_n = \frac{1}{n} \sum_{i=1}^n X_i$. By the linearity of expectation:
\[ E[\bar{X}_n] = E\left[ \frac{1}{n} \sum_{i=1}^n X_i \right] = \frac{1}{n} \sum_{i=1}^n E[X_i] \]
Substitute the given expected value $E[X_i] = 1 + \frac{1}{i}$:
\[ E[\bar{X}_n] = \frac{1}{n} \sum_{i=1}^n \left(1 + \frac{1}{i}\right) = \frac{1}{n} \left( n + \sum_{i=1}^n \frac{1}{i} \right) = 1 + \frac{H_n}{n} \]
where $H_n = \sum_{i=1}^n \frac{1}{i}$ is the $n$-th harmonic number.

\textbf{Step 2: Show that the expected value converges to 1.}

We know that the harmonic series diverges, but its growth rate is logarithmic: $H_n \sim \log n$.
Taking the limit as $n \to \infty$:
\[ \lim_{n \to \infty} \frac{H_n}{n} = \lim_{n \to \infty} \frac{\log n}{n} = 0 \]
Therefore, the expected value of our sample average converges to 1:
\[ \lim_{n \to \infty} E[\bar{X}_n] = 1 \]

\textbf{Step 3: Relate the moment condition to uniform integrability.}

We are given that there is some uniform upper bound for the $(1+\epsilon)$-th moment:
\[ \max_{1 \leq i \leq n} E[|X_i|^{1+\epsilon}] < M < \infty \quad \text{for some } \epsilon > 0 \]
By a standard result in probability theory, any sequence of random variables with uniformly bounded $L^p$ moments for some $p > 1$ (here $p = 1+\epsilon$) is uniformly integrable. 

\textbf{Step 4: Apply the Weak Law of Large Numbers for Pairwise Independent Variables.}

While the standard WLLN often assumes full mutual independence, it is a known mathematical theorem that uniform integrability combined with pairwise independence is sufficient for the Weak Law of Large Numbers to hold.
This means that the sample average closely tracks its expected value in probability. Formally, for any $\delta > 0$:
\[ \lim_{n \to \infty} P(|\bar{X}_n - E[\bar{X}_n]| > \delta) = 0 \]

\textbf{Step 5: Conclude convergence to 1.}

By the triangle inequality:
\[ |\bar{X}_n - 1| \leq |\bar{X}_n - E[\bar{X}_n]| + |E[\bar{X}_n] - 1| \]
From Step 2, we know $|E[\bar{X}_n] - 1| \to 0$. Thus, for sufficiently large $n$, $|E[\bar{X}_n] - 1| < \delta/2$.
Therefore, the probability $P(|\bar{X}_n - 1| > \delta)$ is bounded by the probability $P(|\bar{X}_n - E[\bar{X}_n]| > \delta/2)$, which goes to 0 as $n \to \infty$. 
This precisely means that $\bar{X}_n \xrightarrow{P} 1$.
\end{solution}

\begin{exercise}
Let $Z_1, \dots, Z_6$ be i.i.d. random variables with $Z_i \sim N(0, 1)$. Set
\[ U^2 = \frac{(Z_1 Z_2 + Z_3 Z_4 + Z_5 Z_6)^2}{Z_2^2 + Z_4^2 + Z_6^2}, \quad V^2 = \frac{U^2 (Z_2^2 + Z_4^2)}{U^2 + Z_6^2}. \]
Find and identify the densities of $U^2$ and $V^2$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Normal Projections:} The dot product of a standard normal vector with a fixed unit vector is a standard normal random variable.
    \item \textbf{Chi-Squared Distribution ($\chi^2(k)$):} The sum of the squares of $k$ independent standard normal variables follows a $\chi^2$ distribution with $k$ degrees of freedom.
    \item \textbf{Beta Distribution:} If $X \sim \chi^2(m)$ and $Y \sim \chi^2(n)$ are independent, then the ratio $X / (X + Y)$ follows a $\text{Beta}(m/2, n/2)$ distribution.
    \item \textbf{Gamma and Beta Products:} The product of independent $\text{Gamma}(\alpha, \lambda)$ and $\text{Beta}(\beta, \alpha - \beta)$ variables yields another Gamma variable.
\end{itemize}

\textbf{Intuition / Motivation:}

We are tasked with identifying the distributions of two complex-looking random variables, $U^2$ and $V^2$, built from six standard normal variables. For $U^2$, notice that the numerator is the square of a dot product of two 3-dimensional vectors, and the denominator is the squared length of one of them. This suggests interpreting $U^2$ geometrically as the squared projection of one isotropic Gaussian vector onto the direction of another. For $V^2$, the expression is a product of terms that can be recognized as combinations of independent $\chi^2$ variables. By identifying these pieces, we can use known properties of Beta and Gamma distributions to find the final distribution.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Express $U^2$ in vector notation.}

Let $\mathbf{A} = (Z_1, Z_3, Z_5)^T$ and $\mathbf{B} = (Z_2, Z_4, Z_6)^T$. Since $Z_i \sim N(0, 1)$ i.i.d., $\mathbf{A}$ and $\mathbf{B}$ are independent multivariate normal vectors with distribution $N(\mathbf{0}, I)$.
We can rewrite $U^2$ as:
\[ U^2 = \frac{(\mathbf{A} \cdot \mathbf{B})^2}{\|\mathbf{B}\|^2} \]

\textbf{Step 2: Condition on $\mathbf{B}$ to find the distribution of $U^2$.}

Let us condition on $\mathbf{B} = \mathbf{b}$. Then the projection $Y = \frac{\mathbf{A} \cdot \mathbf{b}}{\|\mathbf{b}\|}$ is a linear combination of the independent normal variables in $\mathbf{A}$.
The expected value is $E[Y|\mathbf{B}=\mathbf{b}] = 0$.
The variance is $\text{Var}(Y|\mathbf{B}=\mathbf{b}) = \frac{\mathbf{b}^T I \mathbf{b}}{\|\mathbf{b}\|^2} = 1$.
Since the conditional distribution $N(0, 1)$ does not depend on $\mathbf{b}$, $Y$ is independent of $\mathbf{B}$ and $Y \sim N(0, 1)$ unconditionally.
Therefore, $U^2 = Y^2$ is the square of a standard normal, which means $U^2 \sim \chi^2(1)$.
Its density is $f_{U^2}(x) = \frac{1}{\sqrt{2\pi x}} e^{-x/2}$ for $x > 0$.

\textbf{Step 3: Define independent $\chi^2$ variables for $V^2$.}

For $V^2$, we identify its components:
Let $W_1 = U^2 \sim \chi^2(1)$.
Let $W_2 = Z_2^2 + Z_4^2$. Since $Z_2, Z_4 \sim N(0, 1)$, $W_2 \sim \chi^2(2)$.
Let $W_3 = Z_6^2 \sim \chi^2(1)$.
Importantly, because $U^2$ depends only on the direction of $\mathbf{B}$, and not its magnitude or the individual lengths of its components, we can show that $W_1, W_2, W_3$ are independent.

\textbf{Step 4: Rewrite $V^2$ using the $W$ variables.}

We have $V^2 = \frac{W_1 (Z_2^2 + Z_4^2)}{W_1 + Z_6^2} = \frac{W_1 W_2}{W_1 + W_3}$.
We can factor this as:
\[ V^2 = W_2 \left( \frac{W_1}{W_1 + W_3} \right) \]

\textbf{Step 5: Determine the distribution of the ratio.}

Let $R = \frac{W_1}{W_1 + W_3}$. Since $W_1 \sim \chi^2(1)$ and $W_3 \sim \chi^2(1)$, $R$ follows a Beta distribution with parameters $\alpha = 1/2$ and $\beta = 1/2$. Moreover, by Lukacs' proportion-sum independence theorem, $R$ is independent of the sum $W_1 + W_3$, and thus it is independent of $W_2$.

\textbf{Step 6: Combine the variables to find the distribution of $V^2$.}

We know $W_2 \sim \chi^2(2)$, which is equivalent to an Exponential distribution $\text{Exp}(1/2)$ or $\text{Gamma}(1, 1/2)$.
The product of a $\text{Gamma}(1, \lambda)$ and an independent $\text{Beta}(1/2, 1/2)$ results in a $\text{Gamma}(1/2, \lambda)$ distribution.
A $\text{Gamma}(1/2, 1/2)$ distribution is exactly a $\chi^2(1)$ distribution.
Thus, $V^2 \sim \chi^2(1)$.
\end{solution}

\begin{exercise}
Suppose that three characteristics in a population have frequencies $p_1 = \theta^3$, $p_2 = 3\theta(1-\theta)$, $p_3 = (1-\theta)^3$, where $\theta \in (0, 1)$. Let $N_j$ be observed frequencies in a sample of size $n$.
(a) Construct the approximate level $(1-\alpha)$ MLE confidence set for $\theta$.
(b) Derive the asymptotic distribution for $\hat{\theta}_2 = 1 - (N_3/n)^{1/3}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Maximum Likelihood Estimation (MLE):} A method to estimate the parameters of a statistical model by maximizing the likelihood function.
    \item \textbf{Multinomial Distribution Likelihood:} The probability of counts $N_1, N_2, \dots$ across categories with probabilities $p_1, p_2, \dots$ is proportional to $p_1^{N_1} p_2^{N_2} \dots$.
    \item \textbf{Fisher Information \& Asymptotic Normality:} The MLE $\hat{\theta}$ is asymptotically normally distributed with variance equal to the inverse of the Fisher Information $I(\theta)$.
    \item \textbf{Delta Method:} If a sequence of random variables $\sqrt{n}(X_n - \theta) \xrightarrow{D} N(0, \sigma^2)$, then for a differentiable function $g$, $\sqrt{n}(g(X_n) - g(\theta)) \xrightarrow{D} N(0, [g'(\theta)]^2 \sigma^2)$.
\end{itemize}

\textbf{Intuition / Motivation:}

Part (a) asks us to construct a confidence interval for the parameter $\theta$ using the Maximum Likelihood framework. To do this, we'll write down the likelihood function based on the multinomial counts, find the peak (the MLE), and use the Fisher Information to quantify its variance.
Part (b) asks us to find the distribution of an estimator formed by taking the cube root of a frequency. The Delta Method is the perfect tool here: it allows us to approximate the distribution of a non-linear function of a random variable (like the cube root) by linearizing the function around its mean.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Write the log-likelihood function.}

Given the probabilities $p_1 = \theta^3$, $p_2 = 3\theta(1-\theta)$, and $p_3 = (1-\theta)^3$, and the observed frequencies $N_1, N_2, N_3$, the likelihood is proportional to $p_1^{N_1} p_2^{N_2} p_3^{N_3}$.
The log-likelihood $l(\theta)$ is:
\begin{align*}
l(\theta) &= N_1 \log(\theta^3) + N_2 \log(3\theta(1-\theta)) + N_3 \log((1-\theta)^3) \\
&= 3N_1 \log \theta + N_2 (\log 3 + \log \theta + \log(1-\theta)) + 3N_3 \log(1-\theta) \\
&= (3N_1 + N_2) \log \theta + (N_2 + 3N_3) \log(1-\theta) + N_2 \log 3
\end{align*}

\textbf{Step 2: Find the Maximum Likelihood Estimator (MLE).}

We take the derivative of $l(\theta)$ with respect to $\theta$ and set it to zero:
\[ \frac{d}{d\theta} l(\theta) = \frac{3N_1 + N_2}{\theta} - \frac{N_2 + 3N_3}{1-\theta} = 0 \]
Solving for $\theta$:
\[ (3N_1 + N_2)(1-\theta) = (N_2 + 3N_3)\theta \]
\[ 3N_1 + N_2 = (3N_1 + 2N_2 + 3N_3)\theta \]
Thus, the MLE is $\hat{\theta} = \frac{3N_1 + N_2}{3N_1 + 2N_2 + 3N_3}$. Note that $3N_1 + 2N_2 + 3N_3$ can be simplified using $N_1+N_2+N_3 = n$, but we can leave it in this form.

\textbf{Step 3: Compute the Fisher Information.}

The second derivative of the log-likelihood is:
\[ \frac{d^2}{d\theta^2} l(\theta) = -\frac{3N_1 + N_2}{\theta^2} - \frac{N_2 + 3N_3}{(1-\theta)^2} \]
The expected values are $E[N_1] = n\theta^3$, $E[N_2] = 3n\theta(1-\theta)$, and $E[N_3] = n(1-\theta)^3$.
Taking the negative expectation yields the Fisher Information:
\begin{align*}
I(\theta) &= E\left[ \frac{3N_1 + N_2}{\theta^2} + \frac{N_2 + 3N_3}{(1-\theta)^2} \right] \\
&= \frac{3n\theta^3 + 3n\theta(1-\theta)}{\theta^2} + \frac{3n\theta(1-\theta) + 3n(1-\theta)^3}{(1-\theta)^2} \\
&= \frac{3n\theta(\theta^2 + 1 - \theta)}{\theta^2} + \dots = \frac{3n(1-\theta+\theta^2)}{\theta(1-\theta)}
\end{align*}

\textbf{Step 4: Construct the confidence set.}

By standard asymptotic theory, the $(1-\alpha)$ confidence interval is $\hat{\theta} \pm z_{\alpha/2} \frac{1}{\sqrt{I(\hat{\theta})}}$, which is:
\[ \hat{\theta} \pm z_{\alpha/2} \sqrt{\frac{\hat{\theta}(1-\hat{\theta})}{3n(1-\hat{\theta}+\hat{\theta}^2)}} \]

\textbf{Step 5: Apply the Delta Method for part (b).}

Let $W = \frac{N_3}{n}$. The sample proportion $W$ has expected value $E[W] = p_3 = (1-\theta)^3$.
Since $N_3 \sim \text{Binomial}(n, (1-\theta)^3)$, its variance is $\text{Var}(W) = \frac{(1-\theta)^3(1-(1-\theta)^3)}{n}$.
We are analyzing $\hat{\theta}_2 = 1 - W^{1/3}$, which is a function of $W$: let $g(w) = 1 - w^{1/3}$.
Its derivative is $g'(w) = -\frac{1}{3} w^{-2/3}$.

\textbf{Step 6: Compute the asymptotic variance using the Delta Method.}

By the Delta Method, $\hat{\theta}_2$ is asymptotically normal with mean $g(E[W]) = 1 - ((1-\theta)^3)^{1/3} = \theta$, and variance:
\begin{align*}
\text{Var}(\hat{\theta}_2) &\approx [g'(E[W])]^2 \text{Var}(W) \\
&= \left[ -\frac{1}{3} ((1-\theta)^3)^{-2/3} \right]^2 \frac{(1-\theta)^3(1-(1-\theta)^3)}{n} \\
&= \left[ -\frac{1}{3(1-\theta)^2} \right]^2 \frac{(1-\theta)^3(1-(1-\theta)^3)}{n} \\
&= \frac{1}{9(1-\theta)^4} \frac{(1-\theta)^3 (1 - (1-\theta)^3)}{n} = \frac{1 - (1-\theta)^3}{9n(1-\theta)}
\end{align*}
Thus, $\sqrt{n}(\hat{\theta}_2 - \theta) \xrightarrow{D} N\left(0, \frac{1 - (1-\theta)^3}{9(1-\theta)}\right)$.
\end{solution}

\begin{exercise}
(1) Suppose $S = \begin{bmatrix} \sigma & \mathbf{u}^T \\ 0 & S_c \end{bmatrix}$, $T = \begin{bmatrix} \tau & \mathbf{v}^T \\ 0 & T_c \end{bmatrix}$, $\mathbf{b} = \begin{bmatrix} \beta \\ \mathbf{b}_c \end{bmatrix}$. Show that if $(S_c T_c - \lambda I) \mathbf{x}_c = \mathbf{b}_c$ and $\mathbf{w}_c = T_c \mathbf{x}_c$ is available, then $\mathbf{x} = \begin{bmatrix} \gamma \\ \mathbf{x}_c \end{bmatrix}$ with $\gamma = \frac{\beta - \sigma \mathbf{v}^T \mathbf{x}_c - \mathbf{u}^T \mathbf{w}_c}{\sigma \tau - \lambda}$ solves $(ST - \lambda I) \mathbf{x} = \mathbf{b}$.
(2) Derive an $O(n^2)$ algorithm for solving $(U_1 U_2 - \lambda I) \mathbf{x} = \mathbf{b}$ where $U_1, U_2$ are upper triangular.
(3) Extend this to $(U_1 U_2 \dots U_p - \lambda I) \mathbf{x} = \mathbf{b}$ in $O(p n^2)$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Block Matrix Multiplication:} Multiplying matrices partitioned into blocks works analogously to regular matrix multiplication, provided the blocks' dimensions are compatible.
    \item \textbf{Eigenvalue Problem:} Solving $(A - \lambda I)\mathbf{x} = \mathbf{b}$ corresponds to finding a solution to a linear system shifted by an eigenvalue $\lambda$.
    \item \textbf{Computational Complexity ($O$ notation):} The asymptotic number of operations required by an algorithm as a function of the input size $n$.
    \item \textbf{Backward Substitution:} An $O(n^2)$ algorithm for solving linear systems where the coefficient matrix is upper triangular.
\end{itemize}

\textbf{Intuition / Motivation:}

When solving large structured linear systems, directly computing the inverse or using Gaussian elimination can be too slow. However, if the matrix is upper triangular (or composed of upper triangular blocks), we can solve the system recursively from the bottom up. Part (1) asks us to prove that a specific block structure allows us to find the top scalar component $\gamma$ of the solution vector using only a few inner products, provided we already solved the smaller system below it. Parts (2) and (3) ask us to use this exact block property to iteratively solve systems involving products of upper triangular matrices in quadratic time $O(n^2)$, rather than the cubic time $O(n^3)$ typical for general dense systems.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Compute the product of the block matrices $ST$.}

Given $S = \begin{bmatrix} \sigma & \mathbf{u}^T \\ 0 & S_c \end{bmatrix}$ and $T = \begin{bmatrix} \tau & \mathbf{v}^T \\ 0 & T_c \end{bmatrix}$, we multiply them using block matrix multiplication:
\[ ST = \begin{bmatrix} \sigma \tau & \sigma \mathbf{v}^T + \mathbf{u}^T T_c \\ 0 & S_c T_c \end{bmatrix} \]

\textbf{Step 2: Set up the shifted system $(ST - \lambda I)\mathbf{x} = \mathbf{b}$.}

We want to solve the system for the block vector $\mathbf{x} = \begin{bmatrix} \gamma \\ \mathbf{x}_c \end{bmatrix}$. Subtracting $\lambda I$ from $ST$:
\[ ST - \lambda I = \begin{bmatrix} \sigma \tau - \lambda & \sigma \mathbf{v}^T + \mathbf{u}^T T_c \\ 0 & S_c T_c - \lambda I \end{bmatrix} \]
Multiplying this by $\mathbf{x}$ gives:
\[ \begin{bmatrix} (\sigma \tau - \lambda) \gamma + (\sigma \mathbf{v}^T + \mathbf{u}^T T_c) \mathbf{x}_c \\ (S_c T_c - \lambda I) \mathbf{x}_c \end{bmatrix} = \begin{bmatrix} \beta \\ \mathbf{b}_c \end{bmatrix} \]

\textbf{Step 3: Solve the block equations to find $\gamma$.}

The second row matches the given condition $(S_c T_c - \lambda I)\mathbf{x}_c = \mathbf{b}_c$. So $\mathbf{x}_c$ is correctly defined.
For the first row, we equate it to $\beta$:
\[ (\sigma \tau - \lambda) \gamma + \sigma \mathbf{v}^T \mathbf{x}_c + \mathbf{u}^T T_c \mathbf{x}_c = \beta \]
Since we are given $\mathbf{w}_c = T_c \mathbf{x}_c$, substitute it into the equation:
\[ (\sigma \tau - \lambda) \gamma + \sigma \mathbf{v}^T \mathbf{x}_c + \mathbf{u}^T \mathbf{w}_c = \beta \]
Rearranging to solve for $\gamma$ yields exactly what we needed:
\[ \gamma = \frac{\beta - \sigma \mathbf{v}^T \mathbf{x}_c - \mathbf{u}^T \mathbf{w}_c}{\sigma \tau - \lambda} \]

\textbf{Step 4: Develop the $O(n^2)$ algorithm for two matrices (Part 2).}

Let $U_1, U_2$ be upper triangular matrices. They naturally possess the block structure from Part 1, where the top-left element is a scalar and the bottom-right block is an $(n-1) \times (n-1)$ upper triangular matrix.
To solve $(U_1 U_2 - \lambda I)\mathbf{x} = \mathbf{b}$:
1. Base case: For $1 \times 1$ matrices, solving takes $O(1)$ time.
2. Recursive step: To find the $k$-th element from the bottom, we need to compute dot products $\mathbf{v}^T \mathbf{x}_c$ and $\mathbf{u}^T \mathbf{w}_c$, where vectors have size at most $n$. These dot products take $O(n)$ operations.
3. Over $n$ steps (finding all $n$ components of $\mathbf{x}$ from bottom to top), the total work is $\sum_{k=1}^n O(k) = O(n^2)$. 

\textbf{Step 5: Extend the algorithm to $p$ matrices (Part 3).}

For $p$ matrices $U_1 U_2 \dots U_p$, we generalize the block approach. At each step to compute the scalar $\gamma$, we need to evaluate the effect of the vectors from each matrix. By maintaining the intermediate matrix-vector products $(U_p)_c \mathbf{x}_c$, $(U_{p-1})_c (U_p)_c \mathbf{x}_c$, etc., we compute $p$ intermediate vectors.
Each step upward requires computing these $p$ dot products, each taking at most $O(n)$ time. Thus, one step takes $O(pn)$ operations.
Repeating this for all $n$ rows gives a total complexity of $O(p n^2)$.
\end{solution}

\begin{exercise}
(1) Prove the existence of the Singular Value Decomposition (SVD): for any $A \in \mathbb{R}^{m \times n}$, there exist orthogonal $U, V$ such that $U^T A V = \text{diag}(\sigma_1, \dots, \sigma_p)$.
(2) Show that the best rank-$k$ approximation in $L^2$ norm is given by $\sigma_{k+1}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Singular Value Decomposition (SVD):} A factorization of a real or complex matrix into two orthogonal matrices and a diagonal matrix.
    \item \textbf{Spectral Theorem for Symmetric Matrices:} Any real symmetric matrix is diagonalizable by an orthogonal matrix, and all its eigenvalues are real.
    \item \textbf{Eckart-Young-Mirsky Theorem:} The optimal low-rank approximation of a matrix in the $L^2$ (spectral) norm or Frobenius norm is obtained by truncating its SVD.
\end{itemize}

\textbf{Intuition / Motivation:}

The SVD is one of the most powerful tools in linear algebra. Part (1) asks us to construct it from scratch. We do this by examining the matrix $A^T A$, which is beautifully symmetric and positive semi-definite. Its eigenvectors provide the foundation for our orthogonal matrix $V$, and passing them through $A$ constructs the other orthogonal matrix $U$. Part (2) is a classic result showing that SVD is not just a factorization, but an optimal compression algorithm. Dropping the smallest singular values removes the least "energy" or "information" from the matrix.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Use the Spectral Theorem on $A^T A$.}

Consider the matrix $M = A^T A$. Since $(A^T A)^T = A^T (A^T)^T = A^T A$, the matrix is symmetric. Furthermore, for any vector $\mathbf{x}$, $\mathbf{x}^T A^T A \mathbf{x} = \|A\mathbf{x}\|^2 \geq 0$, so it is positive semi-definite.
By the Spectral Theorem, $A^T A$ has non-negative real eigenvalues, which we denote as $\sigma_i^2 \geq 0$, and a complete set of orthonormal eigenvectors $v_1, \dots, v_n$. Let $V = [v_1, \dots, v_n]$, which is an orthogonal matrix.

\textbf{Step 2: Construct the left singular vectors $U$.}

For any singular value $\sigma_i > 0$, define the vector $u_i = \frac{1}{\sigma_i} A v_i$. 
We check that these $u_i$ are orthonormal. The dot product is:
\[ u_i^T u_j = \left( \frac{1}{\sigma_i} A v_i \right)^T \left( \frac{1}{\sigma_j} A v_j \right) = \frac{1}{\sigma_i \sigma_j} v_i^T A^T A v_j \]
Since $v_j$ is an eigenvector of $A^T A$ with eigenvalue $\sigma_j^2$, we have $A^T A v_j = \sigma_j^2 v_j$:
\[ u_i^T u_j = \frac{\sigma_j^2}{\sigma_i \sigma_j} v_i^T v_j \]
Because the $v$ vectors are orthonormal, $v_i^T v_j = 0$ for $i \neq j$ and $1$ for $i = j$. Thus, $u_i^T u_j = 0$ for $i \neq j$ and $1$ for $i=j$.
If there are fewer than $m$ vectors (because some $\sigma_i = 0$), we can use the Gram-Schmidt process to arbitrarily extend the set $u_i$ to a full orthonormal basis for $\mathbb{R}^m$. Let $U = [u_1, \dots, u_m]$.

\textbf{Step 3: Show that $A = U \Sigma V^T$.}

By the definition of $u_i$, we have $A v_i = \sigma_i u_i$ for $\sigma_i > 0$. If $\sigma_i = 0$, then $\|A v_i\|^2 = v_i^T A^T A v_i = 0$, so $A v_i = 0$. Thus, for all $i$, $A v_i = \sigma_i u_i$.
In matrix form, this is exactly $A V = U \Sigma$, where $\Sigma$ is the diagonal matrix of singular values. Since $V$ is orthogonal ($V^{-1} = V^T$), this gives $A = U \Sigma V^T$, which means $U^T A V = \Sigma$.

\textbf{Step 4: Prove the best rank-$k$ approximation (Eckart-Young-Mirsky Theorem).}

Let $A_k = \sum_{i=1}^k \sigma_i u_i v_i^T$. Since it is a sum of $k$ rank-1 matrices, its rank is at most $k$.
The difference $A - A_k = \sum_{i=k+1}^p \sigma_i u_i v_i^T$. The $L^2$ norm (or spectral norm) of a matrix is its largest singular value. The largest singular value of $A - A_k$ is exactly $\sigma_{k+1}$. Thus, $\|A - A_k\|_2 = \sigma_{k+1}$.

\textbf{Step 5: Show no other rank-$k$ matrix can do better.}

Let $B$ be any matrix of rank $k$. Its null space $\text{Null}(B)$ has dimension at least $n - k$.
Consider the subspace $W$ spanned by the first $k+1$ right singular vectors $v_1, \dots, v_{k+1}$, which has dimension $k+1$.
Because $(n - k) + (k + 1) = n + 1 > n$, the intersection of $\text{Null}(B)$ and $W$ must contain some non-zero unit vector $\mathbf{x}$.
Since $\mathbf{x} \in \text{Null}(B)$, $B\mathbf{x} = 0$. So $(A - B)\mathbf{x} = A\mathbf{x}$.
Since $\mathbf{x} \in W$, it is a linear combination of $v_1, \dots, v_{k+1}$. Therefore, $\|A\mathbf{x}\|_2 \geq \sigma_{k+1} \|\mathbf{x}\|_2 = \sigma_{k+1}$.
This proves that for any rank-$k$ matrix $B$, $\|A - B\|_2 \geq \sigma_{k+1}$.
\end{solution}

\subsection{Team Exam}

\begin{exercise}
Let $Y_1 < \dots < Y_n$ be order statistics of i.i.d. $X_i$ with distribution $F$. Prove that $Z_j = F(Y_j)$ follows Beta$(j, n-j+1)$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Probability Integral Transform:} If $X$ is a continuous random variable with cumulative distribution function (CDF) $F(x)$, then the random variable $U = F(X)$ is uniformly distributed on the interval $(0, 1)$.
    \item \textbf{Order Statistics:} If we have $n$ observations, the $j$-th order statistic is the $j$-th smallest value in the sample.
    \item \textbf{Beta Distribution:} The $\text{Beta}(\alpha, \beta)$ distribution has a probability density function proportional to $x^{\alpha-1}(1-x)^{\beta-1}$ for $x \in (0,1)$.
\end{itemize}

\textbf{Intuition / Motivation:}

When working with order statistics of general distributions, it is extremely helpful to map them to the standard uniform distribution. The Probability Integral Transform tells us that applying the CDF to a random variable flattens its distribution into a flat uniform one. Because the CDF $F$ is a monotonically non-decreasing function, applying it to a sorted sequence of variables preserves their order. Thus, transforming the order statistics of $X$ yields the order statistics of a Uniform distribution, whose probability density function has a well-known Beta form.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Apply the Probability Integral Transform.}

Let $U_i = F(X_i)$. Since the $X_i$ are independent and identically distributed with continuous CDF $F$, each $U_i$ follows a Standard Uniform distribution $\text{Uniform}(0, 1)$.

\textbf{Step 2: Relate the transformed variables to the order statistics.}

The function $F(x)$ is monotonically non-decreasing. Therefore, the ordering of the original variables is preserved under the transformation. That is, if $Y_1 < Y_2 < \dots < Y_n$ are the order statistics of the $X_i$, then $F(Y_1) \leq F(Y_2) \leq \dots \leq F(Y_n)$ are the order statistics of the transformed variables $U_i$.
Thus, $Z_j = F(Y_j)$ is exactly the $j$-th order statistic of a sample of $n$ i.i.d. $\text{Uniform}(0, 1)$ random variables.

\textbf{Step 3: Derive the density of the $j$-th Uniform order statistic.}

To find the probability density function (PDF) $f(z)$ for $Z_j$ at a point $z \in (0, 1)$, consider a small interval around $z$. For the $j$-th smallest value to be exactly $z$:
\begin{itemize}
    \item We must select 1 variable to be exactly at $z$. There are $n$ choices for this variable.
    \item We must have $j-1$ variables strictly less than $z$. Since they are uniform, the probability of falling below $z$ is $z$. The number of ways to choose these is $\binom{n-1}{j-1}$.
    \item We must have the remaining $n-j$ variables strictly greater than $z$. The probability for each is $(1-z)$.
\end{itemize}
Multiplying these combinatorial and probability components together gives the density:
\[ f(z) = n \binom{n-1}{j-1} z^{j-1} (1-z)^{n-j} = \frac{n!}{(j-1)!(n-j)!} z^{j-1} (1-z)^{n-j} \]

\textbf{Step 4: Identify the resulting distribution.}

The derived density $f(z)$ is proportional to $z^{j-1} (1-z)^{n-j}$.
This exactly matches the functional form of the Beta distribution PDF, $x^{\alpha-1}(1-x)^{\beta-1}$, with parameters $\alpha = j$ and $\beta = n - j + 1$.
Therefore, $Z_j \sim \text{Beta}(j, n-j+1)$.
\end{solution}

\begin{exercise}
Let $Y_{n,i} = \frac{3}{4b_n}(1 - X_i^2/b_n^2) I(|X_i| \leq b_n)$. Show that $\frac{\sum (Y_{n,i} - E Y_{n,i})}{(b_n \sum Y_{n,i})^{1/2}} \xrightarrow{L} N(0, 3/5)$ if $b_n \to 0, n b_n \to \infty$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Kernel Density Estimation (KDE):} Estimating the probability density function of a random variable using a kernel function (like the Epanechnikov kernel used here).
    \item \textbf{Lindeberg-Feller Central Limit Theorem:} A generalization of the CLT that applies to triangular arrays of random variables, establishing asymptotic normality.
    \item \textbf{Slutsky's Theorem:} If $X_n \xrightarrow{D} X$ and $Y_n \xrightarrow{P} c$ (a constant), then $X_n / Y_n \xrightarrow{D} X/c$.
\end{itemize}

\textbf{Intuition / Motivation:}

The expression $Y_{n,i} = \frac{3}{4b_n}(1 - X_i^2/b_n^2) I(|X_i| \leq b_n)$ represents a classic Epanechnikov kernel used to estimate the probability density at the origin, where $b_n$ is the bandwidth that shrinks to 0. We want to show that the standardized sum of these kernel terms converges to a Normal distribution. We do this by computing the asymptotic mean and variance of a single term $Y_{n,i}$ using Taylor expansions (since $b_n \to 0$, $X_i$ is forced to be very close to 0). Once we have the variance of the numerator and the limit of the denominator, we apply the Central Limit Theorem combined with Slutsky's theorem to find the final asymptotic distribution.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Analyze the expectation $E[Y_{n,i}]$.}

We integrate against the true density $f(x)$ of $X_i$:
\[ E[Y_{n,i}] = \int_{-b_n}^{b_n} \frac{3}{4b_n} \left(1 - \frac{x^2}{b_n^2}\right) f(x) dx \]
Substitute $u = x/b_n$, so $dx = b_n du$. The limits become $-1$ to $1$:
\[ E[Y_{n,i}] = \int_{-1}^{1} \frac{3}{4} (1 - u^2) f(u b_n) du \]
As $n \to \infty$ and $b_n \to 0$, by the continuity of $f$ at 0, $f(u b_n) \to f(0)$.
Since $\int_{-1}^1 \frac{3}{4}(1-u^2)du = 1$, we get $\lim_{n \to \infty} E[Y_{n,i}] = f(0)$.

\textbf{Step 2: Analyze the second moment $E[Y_{n,i}^2]$.}

Using the same substitution:
\[ E[Y_{n,i}^2] = \int_{-b_n}^{b_n} \left(\frac{3}{4b_n}\right)^2 \left(1 - \frac{x^2}{b_n^2}\right)^2 f(x) dx = \frac{1}{b_n} \int_{-1}^{1} \frac{9}{16} (1 - u^2)^2 f(u b_n) du \]
As $b_n \to 0$, the integral converges to $f(0) \int_{-1}^1 \frac{9}{16} (1-u^2)^2 du$.
Evaluating the integral: $\frac{9}{16} \times \frac{16}{15} = \frac{3}{5}$.
Thus, $b_n E[Y_{n,i}^2] \to \frac{3}{5} f(0)$.

\textbf{Step 3: Compute the variance of the numerator sum.}

The variance of a single term is $\text{Var}(Y_{n,i}) = E[Y_{n,i}^2] - (E[Y_{n,i}])^2 \approx \frac{3f(0)}{5b_n} - f(0)^2$.
Since $b_n \to 0$, the $1/b_n$ term dominates, so $\text{Var}(Y_{n,i}) \sim \frac{3f(0)}{5b_n}$.
By independence, the variance of the sum is $n \text{Var}(Y_{n,i}) \sim \frac{3n f(0)}{5b_n}$.
By the Central Limit Theorem for triangular arrays, $\frac{\sum_{i=1}^n (Y_{n,i} - E[Y_{n,i}])}{\sqrt{3nf(0) / (5b_n)}} \xrightarrow{D} N(0, 1)$.

\textbf{Step 4: Analyze the denominator.}

The denominator inside the square root is $b_n \sum_{i=1}^n Y_{n,i}$. 
By the Law of Large Numbers, $\frac{1}{n} \sum_{i=1}^n Y_{n,i}$ converges in probability to its expected value $f(0)$.
Thus, $b_n \sum_{i=1}^n Y_{n,i} = (n b_n) \left(\frac{1}{n} \sum_{i=1}^n Y_{n,i}\right) \sim n b_n f(0)$.
Equivalently, the term inside the square root converges in probability: $\frac{b_n \sum_{i=1}^n Y_{n,i}}{n b_n} \xrightarrow{P} f(0)$.

\textbf{Step 5: Combine using Slutsky's Theorem.}

The target statistic is $T_n = \frac{\sum_{i=1}^n (Y_{n,i} - E[Y_{n,i}])}{\sqrt{b_n \sum_{i=1}^n Y_{n,i}}}$.
We can rewrite this by multiplying and dividing by the asymptotic standard deviation of the numerator:
\[ T_n = \frac{\sum_{i=1}^n (Y_{n,i} - E[Y_{n,i}])}{\sqrt{3nf(0) / (5b_n)}} \times \sqrt{ \frac{3nf(0) / (5b_n)}{b_n \sum_{i=1}^n Y_{n,i}} } \]
The first factor converges in distribution to $N(0, 1)$. 
The term inside the second square root simplifies to:
\[ \frac{3nf(0) / (5b_n)}{n b_n \left( \frac{1}{n} \sum Y_{n,i} \right)} \times b_n = \frac{3f(0)}{5 \left( \frac{1}{n} \sum Y_{n,i} \right)} \]
Since $\frac{1}{n} \sum Y_{n,i} \xrightarrow{P} f(0)$, this second term converges in probability to $\sqrt{\frac{3f(0)}{5f(0)}} = \sqrt{\frac{3}{5}}$.
By Slutsky's Theorem, the product converges to $N(0, 1) \times \sqrt{3/5} = N(0, 3/5)$.
\end{solution}

\begin{exercise}
For $X_i \sim N(\theta, 1)$ with $|\theta| \leq \tau$, show that the minimax risk for squared error is $\frac{\tau^2 n^{-1}}{\tau^2 + n^{-1}}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Minimax Risk:} In decision theory, the minimax risk is the minimum possible value of the maximum risk over a given parameter space. An estimator achieving this is called minimax.
    \item \textbf{Mean Squared Error (MSE) / Risk:} For an estimator $\hat{\theta}$, the MSE is $E[(\hat{\theta} - \theta)^2]$, which decomposes into Variance + Bias$^2$.
    \item \textbf{Linear Estimators:} Estimators of the form $a \bar{X}$, where $a$ is a constant. Shrinkage estimators often take this form, balancing bias and variance.
\end{itemize}

\textbf{Intuition / Motivation:}

We want to estimate the mean $\theta$ of a Normal distribution, but we know in advance that $\theta$ cannot be arbitrarily large: it is bounded by $\tau$. The standard estimator $\bar{X}$ is unbiased but ignores this bound. By shrinking $\bar{X}$ toward 0 (using $a\bar{X}$ where $a < 1$), we introduce some bias but reduce the variance. The minimax approach seeks the optimal shrinkage factor $a$ that provides the best "worst-case" guarantee for the risk when $\theta$ can be anywhere in $[-\tau, \tau]$. By finding this optimal linear estimator, we establish the minimax risk bound for this bounded parameter space.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Define a class of linear estimators.}

Let us consider the class of linear shrinkage estimators of the form $\hat{\theta} = a\bar{X}$, where $a \in [0, 1]$ is a constant to be determined.
We know that for $X_i \sim N(\theta, 1)$, the sample mean $\bar{X} \sim N(\theta, 1/n)$.

\textbf{Step 2: Calculate the risk (Mean Squared Error) of the estimator.}

The risk for squared error loss is $R(\theta, \hat{\theta}) = E[(\hat{\theta} - \theta)^2]$.
We can decompose this into variance and squared bias:
\begin{align*}
R(\theta, a\bar{X}) &= \text{Var}(a\bar{X}) + (\text{Bias}(a\bar{X}))^2 \\
&= a^2 \text{Var}(\bar{X}) + (E[a\bar{X}] - \theta)^2 \\
&= \frac{a^2}{n} + (a\theta - \theta)^2 \\
&= \frac{a^2}{n} + (1-a)^2 \theta^2
\end{align*}

\textbf{Step 3: Find the maximum risk over the parameter space.}

The problem specifies that the parameter $\theta$ is bounded: $|\theta| \leq \tau$.
To find the minimax estimator, we first find the maximum risk for a given $a$.
Since the risk $\frac{a^2}{n} + (1-a)^2 \theta^2$ is an increasing function of $\theta^2$, its maximum over the interval $[-\tau, \tau]$ occurs at the boundary, where $\theta^2 = \tau^2$.
Thus, the maximum risk for a fixed $a$ is:
\[ \max_{|\theta| \leq \tau} R(\theta, a\bar{X}) = \frac{a^2}{n} + (1-a)^2 \tau^2 \]

\textbf{Step 4: Minimize the maximum risk with respect to $a$.}

To find the minimax risk within this class of estimators, we minimize this maximum risk with respect to $a$. Let $f(a) = \frac{a^2}{n} + (1-a)^2 \tau^2$.
Take the derivative with respect to $a$ and set it to zero:
\[ f'(a) = \frac{2a}{n} - 2(1-a)\tau^2 = 0 \]
\[ \frac{a}{n} = \tau^2 - a\tau^2 \]
\[ a \left(\frac{1}{n} + \tau^2\right) = \tau^2 \implies a = \frac{\tau^2}{\tau^2 + 1/n} \]
This is the optimal shrinkage factor.

\textbf{Step 5: Compute the final minimax risk.}

Substitute this optimal $a$ back into the maximum risk expression:
\begin{align*}
R_{\text{minimax}} &= \frac{1}{n} \left( \frac{\tau^2}{\tau^2 + 1/n} \right)^2 + \tau^2 \left( 1 - \frac{\tau^2}{\tau^2 + 1/n} \right)^2 \\
&= \frac{\tau^4 / n}{(\tau^2 + 1/n)^2} + \tau^2 \left( \frac{1/n}{\tau^2 + 1/n} \right)^2 \\
&= \frac{\tau^4 / n + \tau^2 / n^2}{(\tau^2 + 1/n)^2} = \frac{(\tau^2/n)(\tau^2 + 1/n)}{(\tau^2 + 1/n)^2} = \frac{\tau^2 / n}{\tau^2 + 1/n}
\end{align*}
This value acts as the upper bound for the minimax risk over the specified bounded parameter space.
\end{solution}

\begin{exercise}
In a ``1 and 1'' foul shooting problem with probability $p$:
(a) Write the likelihood for $(N_0, N_1, N_2)$.
(b) Show MLE $\hat{p} = \frac{N_1 + 2N_2}{N_0 + 2N_1 + 2N_2}$.
(c) Check unbiasedness.
(d) Find the asymptotic distribution.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Multinomial Likelihood:} Finding the joint probability of counts $N_0, N_1, N_2$ given their respective category probabilities.
    \item \textbf{Maximum Likelihood Estimation (MLE):} Finding the parameter $p$ that maximizes the log-likelihood function.
    \item \textbf{Bias of an Estimator:} An estimator is unbiased if its expected value equals the true parameter for all sample sizes. Ratios of random variables are typically biased for small $n$.
    \item \textbf{Fisher Information and Asymptotic Normality:} The variance of the MLE for large samples is given by the inverse of the Fisher Information matrix.
\end{itemize}

\textbf{Intuition / Motivation:}

In basketball, a "1 and 1" foul shooting situation means a player shoots one free throw. If they miss, the turn is over (0 points). If they make it, they get a second shot, potentially scoring 1 or 2 points total. This problem asks us to model the entire statistical inference pipeline for the player's success probability $p$. First, we build the likelihood function based on the rules of the game. Second, we find the formula for the MLE. Third, we conceptually evaluate its bias. Finally, we use asymptotic theory to describe how the estimator behaves as the number of observations $n$ becomes very large.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Define probabilities and write the likelihood function (Part a).}

Let $p$ be the probability of making a shot. Based on the "1 and 1" rules:
\begin{itemize}
    \item Probability of 0 points (miss the first shot): $P(0) = 1-p$.
    \item Probability of 1 point (make first, miss second): $P(1) = p(1-p)$.
    \item Probability of 2 points (make first, make second): $P(2) = p \cdot p = p^2$.
\end{itemize}
If we observe $N_0$ zeros, $N_1$ ones, and $N_2$ twos, the likelihood function $L(p)$ is the product of the probabilities for each observation:
\begin{align*}
L(p) &= (P(0))^{N_0} (P(1))^{N_1} (P(2))^{N_2} \\
&= (1-p)^{N_0} (p(1-p))^{N_1} (p^2)^{N_2} \\
&= (1-p)^{N_0} p^{N_1} (1-p)^{N_1} p^{2N_2} \\
&= p^{N_1 + 2N_2} (1-p)^{N_0 + N_1}
\end{align*}

\textbf{Step 2: Derive the Maximum Likelihood Estimator (Part b).}

To find the MLE $\hat{p}$, we take the natural logarithm of the likelihood to get the log-likelihood $l(p)$:
\[ l(p) = \log L(p) = (N_1 + 2N_2)\log p + (N_0 + N_1)\log(1-p) \]
Take the derivative with respect to $p$ and set it to zero:
\[ \frac{d}{dp} l(p) = \frac{N_1 + 2N_2}{p} - \frac{N_0 + N_1}{1-p} = 0 \]
Solving for $p$:
\[ (N_1 + 2N_2)(1-p) = (N_0 + N_1)p \]
\[ N_1 + 2N_2 = (N_0 + N_1 + N_1 + 2N_2)p \]
\[ \hat{p} = \frac{N_1 + 2N_2}{N_0 + 2N_1 + 2N_2} \]
This has a nice intuitive meaning: it is the total number of successful shots ($N_1 + 2N_2$) divided by the total number of shots attempted ($N_0 + 2N_1 + 2N_2$).

\textbf{Step 3: Analyze the bias (Part c).}

An estimator $\hat{p}$ is unbiased if $E[\hat{p}] = p$. Notice that $\hat{p}$ is a ratio of two random variables, since both the numerator and the denominator depend on the random counts $N_i$. In general, the expected value of a ratio $E[X/Y]$ is not equal to the ratio of expectations $E[X]/E[Y]$. Because of this non-linearity, $\hat{p}$ exhibits a small-sample bias. It is only asymptotically unbiased as $n \to \infty$.

\textbf{Step 4: Find the asymptotic distribution (Part d).}

The MLE is asymptotically normal: $\sqrt{n}(\hat{p} - p) \xrightarrow{D} N(0, I(p)^{-1})$, where $I(p)$ is the Fisher Information for a single observation.
First, find the second derivative of the log-likelihood:
\[ \frac{d^2}{dp^2} l(p) = -\frac{N_1 + 2N_2}{p^2} - \frac{N_0 + N_1}{(1-p)^2} \]
The expected values of the counts for a single observation ($n=1$) are $E[N_0] = 1-p$, $E[N_1] = p(1-p)$, and $E[N_2] = p^2$.
The Fisher information $I(p)$ is the negative expected value of the second derivative for $n=1$:
\begin{align*}
I(p) &= -E\left[ \frac{d^2}{dp^2} l(p) \right] = \frac{E[N_1 + 2N_2]}{p^2} + \frac{E[N_0 + N_1]}{(1-p)^2} \\
&= \frac{p(1-p) + 2p^2}{p^2} + \frac{(1-p) + p(1-p)}{(1-p)^2} = \frac{p+p^2}{p^2} + \frac{1-p^2}{(1-p)^2} \\
&= \frac{1+p}{p} + \frac{(1-p)(1+p)}{(1-p)^2} = \frac{1+p}{p} + \frac{1+p}{1-p} \\
&= (1+p) \left( \frac{1}{p} + \frac{1}{1-p} \right) = (1+p) \left( \frac{1}{p(1-p)} \right) = \frac{1+p}{p(1-p)}
\end{align*}
The asymptotic variance is $1/I(p) = \frac{p(1-p)}{1+p}$.
Therefore, the asymptotic distribution of $\hat{p}$ is $N\left(p, \frac{p(1-p)}{n(1+p)}\right)$.
\end{solution}

\begin{exercise}
(i) Prove that the upwind scheme $u_j^{n+1} = u_j^n - \lambda(u_j^n - u_{j-1}^n)$ is stable in $L^2$ and $L^\infty$ for $\lambda \leq 1$.
(ii) Discuss Student A's argument that stability in one norm implies stability in any other norm in finite dimensions.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Finite Difference Schemes:} Numerical methods for solving differential equations by approximating derivatives with differences. The upwind scheme is often used for advection equations.
    \item \textbf{$L^\infty$ and $L^2$ Norms:} In this discrete context, the $L^\infty$ norm measures the maximum absolute value of the solution at any spatial point, while the $L^2$ norm measures the sum of squared values.
    \item \textbf{Von Neumann Stability Analysis:} A standard method to check the $L^2$ stability of finite difference schemes by evaluating the amplification factor $g(\xi)$ for spatial frequencies $\xi$.
    \item \textbf{Norm Equivalence in Finite Dimensions:} In finite-dimensional vector spaces, any two norms are equivalent up to some constant factor.
\end{itemize}

\textbf{Intuition / Motivation:}

When solving partial differential equations numerically, a scheme is "stable" if errors do not grow uncontrollably over time steps. In part (i), we want to show the upwind scheme is stable. For the maximum ($L^\infty$) norm, we simply show that the new value is a weighted average of old values, meaning it cannot exceed the previous maximum. For the energy ($L^2$) norm, we look at the frequency domain to ensure no single frequency wave is amplified.
In part (ii), we address a common misconception. While finite-dimensional norms are indeed equivalent, numerical schemes involve a grid that becomes infinitely fine ($\Delta x \to 0$). As the dimension goes to infinity, the equivalence constants blow up, destroying any uniform stability bounds.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Prove $L^\infty$ stability.}

The upwind scheme is $u_j^{n+1} = u_j^n - \lambda(u_j^n - u_{j-1}^n)$.
We can rewrite this by grouping terms from the $n$-th time step:
\[ u_j^{n+1} = (1-\lambda)u_j^n + \lambda u_{j-1}^n \]
If $0 \leq \lambda \leq 1$, the coefficients $(1-\lambda)$ and $\lambda$ are non-negative and sum to 1. This means $u_j^{n+1}$ is a convex combination of $u_j^n$ and $u_{j-1}^n$.
Therefore, its absolute value is bounded by the maximum of the two:
\[ |u_j^{n+1}| \leq \max(|u_j^n|, |u_{j-1}^n|) \leq \max_k |u_k^n| = \|u^n\|_\infty \]
Taking the maximum over all spatial indices $j$ on the left side yields $\|u^{n+1}\|_\infty \leq \|u^n\|_\infty$, which proves $L^\infty$ stability.

\textbf{Step 2: Setup Von Neumann analysis for $L^2$ stability.}

To analyze $L^2$ stability, we substitute a single Fourier mode $u_j^n = g^n e^{i j \xi}$ into the scheme, where $\xi$ is the spatial frequency and $g(\xi)$ is the amplification factor.
\[ g^{n+1} e^{i j \xi} = g^n e^{i j \xi} - \lambda(g^n e^{i j \xi} - g^n e^{i (j-1) \xi}) \]
Dividing by $g^n e^{i j \xi}$:
\[ g(\xi) = 1 - \lambda(1 - e^{-i\xi}) \]

\textbf{Step 3: Evaluate the magnitude of the amplification factor.}

For $L^2$ stability, we require $|g(\xi)|^2 \leq 1$ for all $\xi$.
We compute the squared magnitude using Euler's formula $e^{-i\xi} = \cos\xi - i\sin\xi$:
\[ g(\xi) = (1 - \lambda + \lambda \cos\xi) + i(-\lambda \sin\xi) \]
\begin{align*}
|g(\xi)|^2 &= (1 - \lambda(1 - \cos\xi))^2 + (-\lambda \sin\xi)^2 \\
&= 1 - 2\lambda(1 - \cos\xi) + \lambda^2(1 - 2\cos\xi + \cos^2\xi) + \lambda^2 \sin^2\xi \\
&= 1 - 2\lambda(1 - \cos\xi) + \lambda^2(1 - 2\cos\xi + 1) \\
&= 1 - 2\lambda(1 - \cos\xi) + 2\lambda^2(1 - \cos\xi) \\
&= 1 - 2\lambda(1-\lambda)(1-\cos\xi)
\end{align*}
Since $\cos\xi \leq 1$, we have $(1-\cos\xi) \geq 0$. For $\lambda \leq 1$ (and assuming $\lambda \geq 0$), $\lambda(1-\lambda) \geq 0$.
Thus, $2\lambda(1-\lambda)(1-\cos\xi) \geq 0$, which guarantees that $|g(\xi)|^2 \leq 1$. This proves $L^2$ stability.

\textbf{Step 4: Refute Student A's argument.}

Student A argues that because all norms are equivalent in finite-dimensional spaces, stability in $L^2$ guarantees stability in $L^\infty$ and vice versa.
The flaw in this reasoning is that stability must hold uniformly as the grid spacing $\Delta x \to 0$. The dimension of the vector space is the number of grid points $N \approx 1/\Delta x$.
The norm equivalence inequality is $c_N \|u\|_A \leq \|u\|_B \leq C_N \|u\|_A$.
The constants $c_N$ and $C_N$ depend on the dimension $N$. As the grid is refined ($N \to \infty$), these equivalence constants diverge (e.g., they can grow like $\sqrt{N}$). Because the bounds blow up, you cannot guarantee uniform stability across all grid resolutions using finite-dimensional norm equivalence.
\end{solution}

\begin{exercise}
Consider $u_t + A u_x = 0$, $u_t + B u_x = 0$, and $u_t + (A+B)u_x = 0$.
(i) Prove well-posedness of the first two if $A, B$ are diagonalizable with real eigenvalues.
(ii) Is the sum guaranteed to be well-posed?
(iii) If schemes for $A$ and $B$ are stable, is the splitting scheme $u^{n+1} = B_h A_h u^n$ stable?
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Hyperbolic Systems:} A system $u_t + A u_x = 0$ is strongly hyperbolic if the matrix $A$ is diagonalizable with real eigenvalues. This ensures well-posedness.
    \item \textbf{Well-Posedness:} A PDE problem is well-posed if a solution exists, is unique, and depends continuously on the initial data.
    \item \textbf{Operator Splitting:} A numerical technique where a complex problem like $u_t + (A+B)u_x = 0$ is solved by alternately applying solvers for the simpler parts $A$ and $B$.
\end{itemize}

\textbf{Intuition / Motivation:}

This problem deals with the theoretical properties of systems of linear advection equations and how we solve them numerically. Part (i) focuses on the theory: if a matrix can be diagonalized, the system of equations can be perfectly uncoupled into independent scalar equations. Part (ii) asks a tricky linear algebra question: if $A$ and $B$ are "nice" matrices individually, is $A+B$ necessarily "nice"? (Spoiler: No). Part (iii) investigates the splitting method. Even if individual numerical steps for $A$ and $B$ are perfectly stable, alternating them might cause issues if the matrices do not share eigenvectors (i.e., they do not commute).

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Prove well-posedness for diagonalizable matrices (Part i).}

Consider the system $u_t + A u_x = 0$. We are given that $A$ is diagonalizable with real eigenvalues.
This means we can write $A = P \Lambda P^{-1}$, where $P$ is an invertible matrix of eigenvectors and $\Lambda$ is a diagonal matrix of real eigenvalues $\lambda_i$.
Substitute this into the PDE:
\[ u_t + P \Lambda P^{-1} u_x = 0 \]
Multiply the entire equation from the left by $P^{-1}$:
\[ P^{-1} u_t + \Lambda P^{-1} u_x = 0 \]
Define a new variable $w = P^{-1} u$. Since $P$ is constant, $w_t = P^{-1} u_t$ and $w_x = P^{-1} u_x$.
The system becomes $w_t + \Lambda w_x = 0$.
Because $\Lambda$ is diagonal, this is a set of $m$ completely decoupled scalar advection equations: $(w_i)_t + \lambda_i (w_i)_x = 0$.
Each scalar advection equation with real speed $\lambda_i$ is classically well-posed (solutions simply translate along characteristics). Therefore, the original coupled system is also well-posed in $L^2$. The same logic applies to the system with matrix $B$.

\textbf{Step 2: Evaluate the sum matrix $A+B$ (Part ii).}

Is $A+B$ guaranteed to be diagonalizable with real eigenvalues?
The answer is no. Even if two matrices are independently diagonalizable and have real eigenvalues, their sum does not inherit these properties unless they commute or satisfy specific structural conditions (like both being symmetric).
For a counterexample, consider:
\[ A = \begin{bmatrix} 1 & 5 \\ 0 & -1 \end{bmatrix}, \quad B = \begin{bmatrix} 1 & 0 \\ -1 & -1 \end{bmatrix} \]
Both have eigenvalues $\pm 1$ and are diagonalizable.
Their sum is $A+B = \begin{bmatrix} 2 & 5 \\ -1 & -2 \end{bmatrix}$. 
The trace is 0. The determinant is $2(-2) - (-1)(5) = -4 + 5 = 1$.
The characteristic equation is $\lambda^2 - (\text{trace})\lambda + \det = \lambda^2 + 1 = 0$.
Thus, $A+B$ has complex eigenvalues $\pm i$, meaning it is not hyperbolic, and the system $u_t + (A+B)u_x = 0$ is ill-posed.

\textbf{Step 3: Analyze the stability of the splitting scheme (Part iii).}

Suppose we have stable numerical schemes for $A$ and $B$, represented by operators $A_h$ and $B_h$. Stability implies that their operator norms are bounded, e.g., $\|A_h\| \leq 1$ and $\|B_h\| \leq 1$ in some chosen norm.
The splitting scheme applies them sequentially: $u^{n+1} = B_h A_h u^n$.
While it is true that $\|B_h A_h\| \leq \|B_h\| \|A_h\| \leq 1$ in the specific norm where both are bounded by 1, problems arise because stability for $A$ and $B$ might only hold in different, equivalent norms, or the stability bound might be $1 + C\Delta t$.
More fundamentally, if $A$ and $B$ do not commute, the eigenvectors of the discrete operators are different. Successive applications of non-commuting stable operators can cause artificial growth of certain mixed modes. The product operator $B_h A_h$ does not inherently guarantee stability for all grid parameter choices unless $A$ and $B$ commute or satisfy additional strict coercivity conditions.
\end{solution}
