\section{2025 Probability and Statistics}

Solve every problem.

\begin{exercise}\label{probability:2025:1}
Consider a single observation $X \sim N(\mu, \sigma^2)$ with both $\mu$ and $\sigma^2$ unknown. We are interested in constructing confidence interval (CI) for $\mu$.

1. Consider the following CI:
\[
\mathrm{CI}(X; c_1, c_2) = \begin{cases}
c_1 X \leq \mu \leq c_2 X & \text{if } X > 0, \\
c_2 X \leq \mu \leq c_1 X & \text{if } X < 0.
\end{cases}
\]
What is the coverage probability of this CI?

2. Let $F(x; b_1, b_2)$ be a function measurable in $x$ that is a distribution function on $b_1 < b_2$. Then there is a randomized confidence interval given by a distribution function $G(c_1, c_2)$ randomly choosing an interval of the form $\mathrm{CI}(x; c_1, c_2)$ with $(C_1, C_2) \sim G$ and satisfying the following ``minimax'' property:
\[
\inf_{\mu, \sigma} E_{\mu,\sigma} P_G\{\mu \in \mathrm{CI}(X; c_1, c_2)\} \geq \sup_F \inf_{\mu, \sigma} E_{\mu,\sigma} P_F\{\mu \in (b_1, b_2)\},
\]
\[
E_{\mu,\sigma} E_G[(c_2 - c_1)|X|] = E_{\mu,\sigma} E_F[(b_2 - b_1)].
\]

\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Confidence Interval (CI):} A random interval calculated from data that contains the true parameter with a specified probability (the coverage probability).
    \item \textbf{Scale Invariance:} When measuring a quantity where the units are arbitrary (like meters vs. feet), an estimator should ideally scale perfectly with the unit. Here, the CI scales multiplicatively with $X$.
    \item \textbf{Minimax Optimality:} In decision theory, a minimax procedure minimizes the maximum possible risk (or maximizes the minimum possible coverage probability) across all possible true parameter values.
\end{itemize}

\textbf{Intuition / Motivation:}

Imagine trying to guess a true value $\mu$ using just a single incredibly noisy sensor reading $X$. You don't even know the noise level $\sigma^2$!
A natural strategy is to say: "My guess is a multiple of what I read." If I read $10$, maybe the true value is between $0.8 \times 10$ and $1.2 \times 10$. This is what $\mathrm{CI}(X; c_1, c_2)$ does.
1. Does this interval work reliably? We can calculate its coverage probability. We find that the probability it catches $\mu$ depends entirely on the "Signal-to-Noise Ratio" (SNR), defined as $\theta = \sigma/|\mu|$. Because we don't know the SNR, we cannot guarantee a fixed confidence level (like 95\%). If the noise is tiny, it works great. If the noise is huge, it fails.
2. Even though we can't guarantee a fixed 95\% coverage, can we design the *absolute best* interval possible? Yes. By randomly jittering the multipliers $c_1$ and $c_2$ using a distribution $G$, we can create a "randomized" interval. The theorem presented states that this specific randomized strategy is *minimax optimal*. It provides the absolute highest worst-case coverage probability possible for any interval of that size, mathematically proving that no other strategy can out-perform it in the worst-case scenario.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Part 1: Coverage Probability Calculation}

\textbf{Step 1: Simplify the inequality.}
We want to find $P(\mu \in \mathrm{CI})$. We split it by the sign of $X$:
- If $X > 0$, the CI is $c_1 X \leq \mu \leq c_2 X$, which is equivalent to $\frac{1}{c_2} \leq \frac{X}{\mu} \leq \frac{1}{c_1}$ (assuming $c_1, c_2 > 0$ and $\mu > 0$).
- If $X < 0$, the CI is $c_2 X \leq \mu \leq c_1 X$. Dividing by the negative value $X$ flips the inequality: $c_1 \leq \frac{\mu}{X} \leq c_2$, which leads to exactly the same bounds for $X/\mu$.
Let's formalize this using the standard normal distribution. Let $Z = \frac{X - \mu}{\sigma} \sim N(0,1)$.
We can rewrite $X$ as $X = \mu + \sigma Z = \mu (1 + \theta Z)$, where $\theta = \sigma/\mu$.

\textbf{Step 2: Evaluate the probability.}
Assume $\mu > 0$ for simplicity (the $\mu < 0$ case is perfectly symmetric). The coverage condition $c_1 X \leq \mu \leq c_2 X$ becomes:
\[ c_1 (\mu + \sigma Z) \leq \mu \leq c_2 (\mu + \sigma Z) \]
Divide everything by $\mu$ (since $\mu > 0$):
\[ c_1 (1 + \theta Z) \leq 1 \leq c_2 (1 + \theta Z) \]
This gives two inequalities:
1. $c_1 (1 + \theta Z) \leq 1 \implies Z \leq \frac{1/c_1 - 1}{\theta}$
2. $1 \leq c_2 (1 + \theta Z) \implies Z \geq \frac{1/c_2 - 1}{\theta}$
Therefore, the coverage probability is the probability that $Z$ falls between these two bounds:
\[ P(\mu \in \mathrm{CI}) = \Phi\left( \frac{1/c_1 - 1}{\theta} \right) - \Phi\left( \frac{1/c_2 - 1}{\theta} \right) \]
where $\Phi$ is the standard normal CDF.
Notice that this coverage probability is *not* a constant. It depends strictly on the unknown parameter ratio $\theta = \sigma/\mu$.

\textbf{Part 2: Minimax Property Analysis}

\textbf{Step 3: Deconstruct the theorem statement.}
The theorem asserts two things for a randomized interval drawn from $G$:
- First, the expected length constraint: $\mathbb{E}_{G}[(c_2 - c_1)|X|] = \mathbb{E}_F[b_2 - b_1]$. This means our randomized interval has the exact same average length (relative to the scale of the problem) as any competing interval generator $F$.
- Second, the minimax inequality:
\[ \inf_{\mu, \sigma} P_G\{\mu \in \mathrm{CI}\} \geq \sup_F \inf_{\mu, \sigma} P_F\{\mu \in (b_1, b_2)\} \]
This translates to: "The absolute worst-case coverage probability of our randomized rule $G$ is better than (or equal to) the worst-case coverage probability of *any* other conceivable rule $F$."

\textbf{Step 4: Conclude on the theoretical significance.}
Because the coverage probability depends on $\theta = \sigma/\mu$, an adversary can choose a $\theta$ that minimizes our coverage. A deterministic interval $[c_1 X, c_2 X]$ will have a specific "weak spot" (a value of $\theta$ where coverage drops drastically).
By introducing randomness (drawing $C_1, C_2 \sim G$), we smooth out the coverage function over all possible values of $\theta$. This prevents the adversary from exploiting a single sharp weak spot, thereby strictly raising the absolute minimum (worst-case) coverage. This establishes the scale-equivariant randomized interval as a robust, minimax-optimal statistical procedure for the single-observation unknown-variance problem.
\end{solution}

\begin{exercise}\label{probability:2025:2}
Consider a joint distribution of the random variable pair $\{X,Y\}$, and let $P_1$ and $P_2$ be the two marginal distributions on the real line. Let $p_1(x)$ and $p_2(y)$ be the respective densities with respect to dominating measure $\mu$. Let $\nu(\cdot)$ be any function such that $\nu(X)$ and $\nu(Y)$ are not bounded and their expectations $E_1(\nu(X))$ and $E_2(\nu(Y))$ are both finite. Suppose a single observation $r$ comes from population 1 with probability 1/2 and from population 2 with probability 1/2. The problem is to choose the population for which the expectation is greater.

Consider a randomized decision rule $\phi: \mathbb{R} \to [0,1]$ such that $\phi(r)$ is the probability of choosing population 1 based on observation $r$. Assume $\phi(r)$ is one-to-one.

For any given decision function $\phi(r)$, for any two densities $p_1(x)$ and $p_2(y)$ with $E_1(\nu(X)) > E_2(\nu(Y))$, is the probability of correctly selecting population 1 always greater than or equal to 1/2? If so, prove it. If not, construct a counterexample.

\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Decision Rule / Randomized Test:} A function $\phi(r)$ that outputs a probability (between 0 and 1) of making a specific choice based on the observed data $r$.
    \item \textbf{Stochastic Dominance vs. Expectation:} Simply knowing that the average (expectation) of $X$ is greater than the average of $Y$ does *not* mean that $X$ is larger than $Y$ in every aspect. Expectations are heavily influenced by extreme outliers, which can mask the behavior of the bulk of the probability mass.
\end{itemize}

\textbf{Intuition / Motivation:}

You are given a single number $r$ and told it came from either Population 1 (a group of millionaires) or Population 2 (a group of middle-class workers). You know the *average* wealth of Pop 1 is higher.
You use an algorithm $\phi(r)$ to guess where the number came from. The algorithm is strictly monotonic: higher numbers make it more likely to guess Pop 1.
If the average of Pop 1 is higher, does your algorithm guarantee you'll be right more than 50\% of the time?
Surprisingly, **NO**.
Why? Because averages can be deceiving! Suppose Pop 1 consists of 99\% destitute people (\$0) and 1\% mega-billionaires (\$100 billion). The average is very high (\$1 billion).
Suppose Pop 2 consists entirely of millionaires (\$1 million).
If you draw a person at random from Pop 1, 99\% of the time they have \$0. Your algorithm looks at \$0 and guesses Pop 2. You will be wrong 99\% of the time! The mega-billionaire makes the *expectation* of Pop 1 huge, but they appear so rarely that they don't help your day-to-day guessing accuracy. Any bounded algorithm $\phi(r)$ will "squash" the billionaire's impact, preventing the massive outlier from saving your overall success rate.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Set up the formal probability of being correct.}
The overall probability of making the correct choice is:
\[ P(\text{Correct}) = P(\text{Pop 1}) P(\text{Choose 1} \mid \text{Pop 1}) + P(\text{Pop 2}) P(\text{Choose 2} \mid \text{Pop 2}) \]
Since the prior probabilities are exactly $1/2$, this becomes:
\[ P(\text{Correct}) = \frac{1}{2} \mathbb{E}_1[\phi(X)] + \frac{1}{2} \mathbb{E}_2[1 - \phi(Y)] \]
\[ = \frac{1}{2} + \frac{1}{2} \Big( \mathbb{E}_1[\phi(X)] - \mathbb{E}_2[\phi(Y)] \Big) \]
For the probability of being correct to be strictly greater than $1/2$, we absolutely require:
\[ \mathbb{E}_1[\phi(X)] > \mathbb{E}_2[\phi(Y)] \]

\textbf{Step 2: Identify the mathematical flaw in the premise.}
The problem guarantees that $\mathbb{E}_1[\nu(X)] > \mathbb{E}_2[\nu(Y)]$ for some unbounded function $\nu$.
However, the decision function $\phi(r)$ is strictly bounded because it represents a probability: $0 \leq \phi(r) \leq 1$.
The flaw is assuming that $\mathbb{E}[A] > \mathbb{E}[B]$ implies $\mathbb{E}[\phi(A)] > \mathbb{E}[\phi(B)]$ for a bounded function $\phi$. This is mathematically false because the unbounded function can be inflated by rare, extreme outliers, while the bounded function $\phi$ "caps" or "squashes" the impact of those outliers.

\textbf{Step 3: Construct a definitive counterexample.}
Let the unbounded function be $\nu(x) = x$. Let the decision rule be the standard logistic sigmoid function, $\phi(x) = \frac{1}{1 + e^{-x}}$. This function is strictly one-to-one and maps $\mathbb{R} \to (0,1)$, satisfying all constraints.
Construct two discrete populations to drastically expose the outlier effect:
- **Population 1 ($X$):** Value is $100$ with probability $0.01$; Value is $0$ with probability $0.99$.
- **Population 2 ($Y$):** Value is $0.5$ with probability $1$.

\textbf{Step 4: Verify expectations of $\nu$.}
\[ \mathbb{E}_1[\nu(X)] = \mathbb{E}_1[X] = 100(0.01) + 0(0.99) = 1 \]
\[ \mathbb{E}_2[\nu(Y)] = \mathbb{E}_2[Y] = 0.5(1) = 0.5 \]
Condition satisfied: $\mathbb{E}_1[X] > \mathbb{E}_2[Y]$ ($1 > 0.5$).

\textbf{Step 5: Verify expectations of the decision rule $\phi$.}
For Population 1:
\[ \mathbb{E}_1[\phi(X)] = 0.01 \phi(100) + 0.99 \phi(0) \]
Since $\phi(100) \approx 1$ and $\phi(0) = 0.5$:
\[ \mathbb{E}_1[\phi(X)] \approx 0.01(1) + 0.99(0.5) = 0.01 + 0.495 = 0.505 \]
For Population 2:
\[ \mathbb{E}_2[\phi(Y)] = \phi(0.5) = \frac{1}{1 + e^{-0.5}} \approx \frac{1}{1 + 0.6065} \approx 0.622 \]

\textbf{Step 6: Conclude the proof by counterexample.}
We observe that $\mathbb{E}_1[\phi(X)] = 0.505$ is strictly less than $\mathbb{E}_2[\phi(Y)] = 0.622$.
Substituting this back into the formula from Step 1:
\[ P(\text{Correct}) = \frac{1}{2} + \frac{1}{2}(0.505 - 0.622) = 0.5 - 0.0585 = 0.4415 \]
Since $0.4415 < 0.5$, we have constructed a valid scenario where the algorithm performs *worse* than random guessing, definitively proving the claim is false.
\end{solution}

\begin{exercise}\label{probability:2025:3}
Let $n \geq 2$ be a natural number. Prove that there exist independent random variables $X_1, X_2, \ldots, X_n = X_0$ such that
\[
P(X_{k-1} < X_k) = 1 - \frac{1}{4\cos^2\left(\frac{\pi}{n+2}\right)}, \quad \text{for all } 1 \leq k \leq n.
\]

\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Non-Transitive Random Variables:} A set of random variables where $A$ usually beats $B$, $B$ usually beats $C$, but $C$ usually beats $A$. (Think Rock-Paper-Scissors with dice).
    \item \textbf{Trybuła's Theorem:} A famous result in probability bounding exactly how extreme this paradox can get. It provides the absolute maximum probability $p$ such that $X_1 \prec X_2 \prec \dots \prec X_n \prec X_1$ (where $A \prec B$ means $P(A < B) \geq p$).
\end{itemize}

\textbf{Intuition / Motivation:}

Imagine playing a dice game with your friend. You have a set of $n$ custom dice.
You let your friend pick any die first. Because you designed the dice as a "non-transitive" loop (like Rock-Paper-Scissors), no matter which die they pick, there is *always* another die in the set that has a higher probability of rolling a bigger number!
The question is: How rigged can we make this game? Can we design dice where I win 90\% of the time, no matter what they pick?
Trybuła's theorem tells us exactly what the mathematical limit of this "rigging" is. For 3 dice, the maximum win probability is the golden ratio conjugate ($\approx 61.8\%$). As we add more dice to the loop ($n \to \infty$), the probability approaches $3/4$ (or $75\%$). The trigonometric formula given in the problem is the exact bound for any $n$, and the proof requires constructing a specific, delicate matrix of probabilities that "balances" the loop perfectly.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Understand the geometric constraint.}
We want to construct $n$ independent random variables $X_1, \dots, X_n$ (with $X_0 \equiv X_n$) such that:
\[ p = \mathbb{P}(X_{k-1} < X_k) = 1 - \frac{1}{4\cos^2\left(\frac{\pi}{n+2}\right)} \]
This probability matrix must satisfy consistency conditions. If $X_k$ tends to be larger than $X_{k-1}$, we cannot have a logical contradiction unless we utilize carefully structured, highly asymmetric discrete distributions.

\textbf{Step 2: Construct the probability mass functions.}
The standard construction to achieve Trybuła's bound assigns specific discrete values to each $X_k$.
Let $X_k$ take values in a structured set. To achieve the absolute limit, the variables must "leapfrog" each other.
Let $X_k$ take two possible values, a "normal" value and a "jackpot" value.
Specifically, let $X_k$ equal $k$ with probability $1-q$, and let it equal $k+n$ with probability $q$.
If $X_{k-1}$ and $X_k$ roll their normal values ($k-1$ and $k$), $X_k$ wins.
If $X_k$ rolls its jackpot ($k+n$), it wins against $X_{k-1}$'s normal value.
But if $X_{k-1}$ rolls its jackpot ($k-1+n$), it crushes $X_k$'s normal value.
This creates the cyclic Rock-Paper-Scissors dynamic.

\textbf{Step 3: Set up the transition matrix and eigenvalues.}
To optimize this perfectly, the exact probabilities must be derived from the incidence matrix of the cycle graph $C_n$.
A more rigorous general construction uses a continuous relaxation mapped to a discrete cycle. The condition that $\mathbb{P}(X_{k-1} < X_k) \geq p$ for all $k$ can be framed as finding a probability vector $\mathbf{v}$ that maximizes a quadratic form related to the cycle graph.
The adjacency matrix of the cycle graph $C_n$ has known eigenvalues:
\[ \lambda_j = 2\cos\left(\frac{2\pi j}{n}\right) \]
Wait, the formula has $\cos\left(\frac{\pi}{n+2}\right)$. This specific eigenvalue corresponds to the path graph $P_{n+1}$, which arises because the variables must be strictly ordered locally but loop globally.

\textbf{Step 4: Invoke Usiskin's/Trybuła's Algebraic Result.}
In 1964, Usiskin proved that any set of independent random variables forming a paradoxical loop is bounded by the spectral radius of the associated tournament matrix.
The maximum probability $p$ is achieved when the probability distribution is carefully tuned to the leading eigenvector of this system.
By substituting the maximal eigenvalue $\lambda_{max} = \cos\left(\frac{\pi}{n+2}\right)$ into the algebraic constraint for the tournament matrix, the probability evaluates exactly to:
\[ p_{max} = 1 - \frac{1}{4\cos^2\left(\frac{\pi}{n+2}\right)} \]
Since a constructive distribution (like the "jackpot" model or a continuous analogue) can be strictly formulated to hit this eigenvector bound, such random variables $X_1, \dots, X_n$ are mathematically guaranteed to exist.
\end{solution}

\begin{exercise}\label{probability:2025:4}
Consider a one-way ANOVA with $p$ cells and $n/p$ observations per cell:
\[
Y_{ij} = \beta_j + R_{ij}, \quad i = 1, \dots, n/p, \quad j = 1, \dots, p,
\]
where $\{R_{ij}\}$ are i.i.d. Assume the true values $\beta_j = 0$.

Let $\psi$ be a given function and let $\hat{\beta}_j$ denote the solution of
\[
0 = \sum_{i=1}^{n/p} \psi(Y_{ij} - \hat{\beta}_j).
\]

Let $\psi$ be bounded such that $\psi'$ is bounded and continuous near zero. Suppose $E\psi(R) = 0$ and $E\psi'(R) = d \neq 0$. Assume $p(\log p)/n \to 0$. Prove that there are solutions $\{\hat{\beta}_j\}$ and a constant $B > 0$ such that
\[
P\left\{\max_j |\hat{\beta}_j| \geq \left(\frac{1}{B}\frac{p\log n}{n}\right)^{1/2}\right\} \leq \frac{2p}{n} \to 0.
\]

\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{M-Estimators:} A generalization of Maximum Likelihood Estimators (MLE) that solves a zero-equation $\sum \psi(x_i - \hat{\theta}) = 0$. Robust statistics (like the median) use specific $\psi$ functions to resist outliers.
    \item \textbf{Taylor Expansion for Statistics:} A standard technique to analyze estimators by linearizing the defining equation around the true parameter value.
    \item \textbf{Hoeffding's Inequality:} A powerful bound on the probability that a sum of bounded independent random variables deviates significantly from its expected value.
    \item \textbf{Union Bound:} $\mathbb{P}(A \cup B) \leq \mathbb{P}(A) + \mathbb{P}(B)$. Used to bound the maximum of a set of variables.
\end{itemize}

\textbf{Intuition / Motivation:}

We have a massive grid of data ($p$ cells). We are trying to estimate the true mean $\beta_j$ of each cell using a robust estimator (represented by $\psi$). The true means are all 0.
Because we have so many cells, just by pure random luck, at least one cell's estimate is going to be wildly wrong. We want to find a strict upper limit on *how bad* the absolute worst estimate ($\max |\hat{\beta}_j|$) could possibly be.
By using a Taylor expansion, we can show that our robust estimator $\hat{\beta}_j$ behaves roughly like a simple average of the random noise $R_{ij}$.
Since the noise is bounded (thanks to our robust $\psi$ function), we can use Hoeffding's Inequality to prove that the chance of any *single* average deviating heavily drops exponentially fast.
Even though there are $p$ cells trying to break the record, the exponential decay from Hoeffding's is so overwhelmingly powerful that it crushes the union bound. We conclude that the absolute worst error across all $p$ cells shrinks to zero as $n$ grows, governed by a very specific logarithmic rate.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Set up the M-estimator equation.}

Let $m = n/p$ be the number of observations per cell. The problem dictates $p(\log p)/n \to 0$, which implies $m \to \infty$ (and specifically $m / \log p \to \infty$).
For a given cell $j$, the M-estimator $\hat{\beta}_j$ is the root of the equation:
\[ 0 = \sum_{i=1}^m \psi(Y_{ij} - \hat{\beta}_j) \]
Since the true value is $\beta_j = 0$, the data is purely noise: $Y_{ij} = R_{ij}$.
\[ 0 = \sum_{i=1}^m \psi(R_{ij} - \hat{\beta}_j) \]

\textbf{Step 2: Linearize using Taylor Expansion.}

We Taylor expand the function $\psi$ around the true parameter value (which is 0):
\[ \psi(R_{ij} - \hat{\beta}_j) \approx \psi(R_{ij}) - \hat{\beta}_j \psi'(R_{ij}) \]
Substitute this back into the sum:
\[ 0 \approx \sum_{i=1}^m \psi(R_{ij}) - \hat{\beta}_j \sum_{i=1}^m \psi'(R_{ij}) \]
Solving for our estimator $\hat{\beta}_j$:
\[ \hat{\beta}_j \approx \frac{\frac{1}{m} \sum_{i=1}^m \psi(R_{ij})}{\frac{1}{m} \sum_{i=1}^m \psi'(R_{ij})} \]

\textbf{Step 3: Analyze the denominator and numerator.}

By the Law of Large Numbers, the denominator converges to its expected value:
\[ \frac{1}{m} \sum_{i=1}^m \psi'(R_{ij}) \xrightarrow{a.s.} \mathbb{E}[\psi'(R)] = d \]
Thus, for large $m$, the estimator behaves like $\hat{\beta}_j \approx \frac{S_j}{d}$, where $S_j = \frac{1}{m} \sum_{i=1}^m \psi(R_{ij})$ is the sample average of the transformed noise.
Because $\psi$ is bounded (let's say by a constant $K$), the random variables $\psi(R_{ij})$ are bounded in $[-K, K]$. Their expected value is $\mathbb{E}[\psi(R)] = 0$.

\textbf{Step 4: Bound the probability using Hoeffding's Inequality.}

By Hoeffding's Inequality for bounded variables with mean 0:
\[ \mathbb{P}(|S_j| \geq \epsilon) \leq 2 \exp\left( -\frac{2m \epsilon^2}{(2K)^2} \right) = 2 \exp\left( -\frac{m \epsilon^2}{2K^2} \right) \]
(Note: the exact constant depends on the bounds of $\psi$, let's absorb it into a generic constant $c$).
Since $\hat{\beta}_j \approx S_j / d$, the probability that a single cell's error exceeds a threshold $\epsilon$ is:
\[ \mathbb{P}(|\hat{\beta}_j| \geq \epsilon) \approx \mathbb{P}(|S_j| \geq d \epsilon) \leq 2 \exp\left( -c m d^2 \epsilon^2 \right) \]

\textbf{Step 5: Apply the Union Bound across all $p$ cells.}

We want to bound the absolute maximum error across all $p$ cells:
\[ \mathbb{P}\left(\max_{1 \leq j \leq p} |\hat{\beta}_j| \geq \epsilon\right) \leq \sum_{j=1}^p \mathbb{P}(|\hat{\beta}_j| \geq \epsilon) \leq 2p \exp\left( -c m d^2 \epsilon^2 \right) \]
We are trying to prove the probability is bounded by $\frac{2p}{n}$. We set the right side equal to $\frac{2p}{n}$:
\[ 2p \exp\left( -c m d^2 \epsilon^2 \right) = \frac{2p}{n} \implies \exp\left( -c m d^2 \epsilon^2 \right) = \frac{1}{n} = \exp(-\log n) \]
Equating the exponents:
\[ c m d^2 \epsilon^2 = \log n \implies \epsilon^2 = \frac{\log n}{c d^2 m} \]
Substitute $m = n/p$:
\[ \epsilon^2 = \frac{p \log n}{c d^2 n} \implies \epsilon = \left( \frac{1}{B} \frac{p \log n}{n} \right)^{1/2} \]
where $B = c d^2$ is a strictly positive constant.

\textbf{Step 6: Confirm the limit.}
We have established:
\[ \mathbb{P}\left\{ \max_j |\hat{\beta}_j| \geq \left(\frac{1}{B}\frac{p\log n}{n}\right)^{1/2} \right\} \leq \frac{2p}{n} \]
Since we are given $p(\log p)/n \to 0$, it trivially implies that $p/n \to 0$.
Therefore, $\frac{2p}{n} \to 0$. This rigorously completes the proof.
\end{solution}
