\section{2024 Probability and Statistics}

\begin{exercise}\label{probability:2024:1}
The sequence $(X_1, X_2, \ldots, X_n, \ldots)$ is a Dirichlet process with base distribution $G_0$ and concentration parameter $\alpha_0 > 0$.
(a) Derive the distribution of $X_n, n \geq 1$.
(b) Let $Y_n = I(X_n > 0)$. Prove or disprove $(Y_n)_{n \geq 1}$ forms a Dirichlet Process.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Dirichlet Process (DP) / Pólya Urn Scheme:} A sequence of random variables $X_1, X_2, \dots$ drawn from a Dirichlet Process $DP(\alpha_0, G_0)$ can be described sequentially. The first draw is from the base distribution $G_0$. Subsequent draws are either identical to a previously drawn value (with probability proportional to how many times it was drawn) or a fresh draw from $G_0$.
    \item \textbf{Conditional Probability and Marginals:} The marginal distribution of a variable can be found by taking the expectation of its conditional distribution: $\mathbb{P}(X \in A) = \mathbb{E}[\mathbb{P}(X \in A \mid Y)]$.
\end{itemize}

\textbf{Intuition / Motivation:}

(a) Imagine a magic urn. You draw a ball. With some probability, you paint it a brand new color from a catalog ($G_0$). Otherwise, you copy the color of a ball already in the urn.
If you look at the $n$-th ball drawn, what is the probability it is a specific color? Even though the *conditional* probability heavily depends on the colors of the first $n-1$ balls, the *unconditional* (marginal) probability before the game starts is perfectly symmetric. Every ball, whether it's the 1st or the 100th, has the exact same marginal distribution $G_0$. The rich-get-richer dynamics only affect the joint distribution, not the marginals.
(b) Now, instead of looking at the exact color, we just ask: "Is the number drawn greater than 0?" This turns our complex continuous process into a simple sequence of Yes/No (1/0) answers.
Does this simplified Yes/No sequence still behave like a Dirichlet Process? Yes! The "rich-get-richer" copying mechanism of the Dirichlet Process is perfectly preserved when you lump categories together. If you group all positive numbers into "Yes" and all negative numbers into "No", the rules of drawing a new "Yes" or copying an old "Yes" match exactly with a new Dirichlet Process whose base distribution is just a coin flip.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Part (a): Marginal distribution of $X_n$}

\textbf{Step 1: Base case for induction.}
By the definition of the Pólya Urn representation of a Dirichlet Process, the first variable is drawn directly from the base distribution:
\[ X_1 \sim G_0 \]
Thus, $\mathbb{P}(X_1 \in A) = G_0(A)$ for any measurable set $A$.

\textbf{Step 2: Set up the inductive step.}
Assume that for all $i \leq n$, the marginal distribution is $G_0$, meaning $\mathbb{P}(X_i \in A) = G_0(A)$.
We want to find the marginal distribution of $X_{n+1}$. The predictive distribution of $X_{n+1}$ given the past sequence is:
\[ \mathbb{P}(X_{n+1} \in A \mid X_1, \dots, X_n) = \frac{\alpha_0}{\alpha_0 + n} G_0(A) + \frac{1}{\alpha_0 + n} \sum_{i=1}^n \delta_{X_i}(A) \]
where $\delta_{X_i}(A)$ is the indicator function $\mathbb{I}(X_i \in A)$.

\textbf{Step 3: Calculate the marginal probability.}
We take the expectation of the conditional probability over the joint distribution of $(X_1, \dots, X_n)$:
\[ \mathbb{P}(X_{n+1} \in A) = \mathbb{E} \left[ \mathbb{P}(X_{n+1} \in A \mid X_1, \dots, X_n) \right] \]
\[ = \frac{\alpha_0}{\alpha_0 + n} G_0(A) + \frac{1}{\alpha_0 + n} \sum_{i=1}^n \mathbb{E}[\delta_{X_i}(A)] \]
Since $\mathbb{E}[\delta_{X_i}(A)] = \mathbb{P}(X_i \in A)$, and by our inductive hypothesis this equals $G_0(A)$, we substitute:
\[ \mathbb{P}(X_{n+1} \in A) = \frac{\alpha_0}{\alpha_0 + n} G_0(A) + \frac{1}{\alpha_0 + n} \sum_{i=1}^n G_0(A) \]
\[ = \frac{\alpha_0 G_0(A) + n G_0(A)}{\alpha_0 + n} = \frac{(\alpha_0 + n) G_0(A)}{\alpha_0 + n} = G_0(A) \]
Thus, by mathematical induction, $X_n \sim G_0$ for all $n \geq 1$.

\textbf{Part (b): Dirichlet Process property of $Y_n$}

\textbf{Step 4: Define the new base distribution.}
Let $A = (0, \infty)$. We are defining $Y_n = \mathbb{I}(X_n > 0) = \mathbb{I}(X_n \in A)$.
Since $X_n \sim G_0$, the marginal distribution of $Y_1$ is a Bernoulli variable:
\[ \mathbb{P}(Y_1 = 1) = \mathbb{P}(X_1 \in A) = G_0(A) = p_0 \]
Thus, the new base distribution is $H_0 = \text{Bernoulli}(p_0)$.

\textbf{Step 5: Verify the predictive distribution matches a DP.}
To prove $\{Y_n\}$ forms a Dirichlet Process $DP(\alpha_0, H_0)$, we must show its predictive distribution follows the standard Pólya urn formula for $H_0$.
Using the predictive distribution of $X_{n+1}$:
\[ \mathbb{P}(Y_{n+1} = 1 \mid X_1, \dots, X_n) = \mathbb{P}(X_{n+1} \in A \mid X_1, \dots, X_n) \]
\[ = \frac{\alpha_0 G_0(A) + \sum_{i=1}^n \delta_{X_i}(A)}{\alpha_0 + n} \]
Notice that $G_0(A) = p_0$ and $\delta_{X_i}(A) = \mathbb{I}(X_i > 0) = Y_i$. Substituting these in:
\[ \mathbb{P}(Y_{n+1} = 1 \mid X_1, \dots, X_n) = \frac{\alpha_0 p_0 + \sum_{i=1}^n Y_i}{\alpha_0 + n} \]
Because this conditional probability depends on the history $(X_1, \dots, X_n)$ *only* through the sequence $(Y_1, \dots, Y_n)$, we can safely write:
\[ \mathbb{P}(Y_{n+1} = 1 \mid Y_1, \dots, Y_n) = \frac{\alpha_0 p_0 + \sum_{i=1}^n Y_i}{\alpha_0 + n} \]
This is exactly the Pólya urn predictive update rule for a Dirichlet Process with concentration parameter $\alpha_0$ and base distribution $\text{Bernoulli}(p_0)$.
Therefore, we prove that $(Y_n)_{n \geq 1}$ does indeed form a Dirichlet Process.
\end{solution}

\begin{exercise}\label{probability:2024:2}
$U_n = \sum_{j=1}^{2^n} \frac{j-1}{2^n} I\left(\frac{j-1}{2^n} \leq \phi(X) < \frac{j}{2^n}\right)$, $V_n = E(H(X,Y) | U_n)$.
Prove AS convergence of $V_n$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Filtrations and $\sigma$-algebras:} A $\sigma$-algebra $\mathcal{G}_n$ represents the "information available" at time $n$. An increasing sequence of $\sigma$-algebras $\mathcal{G}_1 \subset \mathcal{G}_2 \subset \dots$ is called a filtration.
    \item \textbf{Lévy's Upward Theorem (Martingale Convergence):} If you have a random variable $W$ with finite expectation ($\mathbb{E}|W| < \infty$), the sequence of conditional expectations $V_n = \mathbb{E}[W \mid \mathcal{G}_n]$ forms a uniformly integrable martingale. It converges almost surely to $\mathbb{E}[W \mid \mathcal{G}_\infty]$, where $\mathcal{G}_\infty$ is the $\sigma$-algebra generated by the union of all $\mathcal{G}_n$.
\end{itemize}

\textbf{Intuition / Motivation:}

Imagine trying to guess the value of a function $H(X,Y)$, but you only get a blurred, pixelated image of $X$.
$U_n$ represents this pixelated version of $X$. We chop the interval $[0,1)$ into $2^n$ tiny bins. $U_n$ tells us exactly which bin $\phi(X)$ fell into.
As $n$ increases, we double the number of bins. The pixels get smaller and smaller, so the image gets sharper and sharper. The information we have at step $n$ (represented by $\mathcal{G}_n$) completely contains the information we had at step $n-1$.
$V_n$ is our "best guess" for $H(X,Y)$ given the pixelated image $U_n$. Because our sequence of guesses is based on strictly improving information, the sequence $V_n$ forms a martingale.
By Lévy's Upward Theorem, as the pixels become infinitely small ($n \to \infty$), our sequence of guesses must converge to the best guess we could make if we had a perfectly sharp image of $X$. Because the infinitely fine bins perfectly pinpoint $\phi(X)$, and $\phi$ is an invertible function, knowing the infinitely fine bins is exactly equivalent to knowing the exact value of $X$.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Identify the information structure ($\sigma$-fields).}

Let $\mathcal{G}_n = \sigma(U_n)$ be the $\sigma$-algebra generated by the random variable $U_n$.
The variable $U_n$ discretizes $\phi(X)$ into dyadic intervals of length $1/2^n$: $[0, 1/2^n), [1/2^n, 2/2^n), \dots$
At step $n+1$, the intervals are of length $1/2^{n+1}$. Notice that every interval at step $n$ is perfectly split into two smaller intervals at step $n+1$.
Because $U_{n+1}$ provides a strictly finer partition than $U_n$, knowing $U_{n+1}$ means you automatically know $U_n$.
Therefore, $\mathcal{G}_n \subseteq \mathcal{G}_{n+1}$. The sequence $\{\mathcal{G}_n\}_{n \geq 1}$ forms a valid filtration.

\textbf{Step 2: Apply Lévy's Upward Theorem.}

We are given the sequence of random variables $V_n = \mathbb{E}[H(X,Y) \mid U_n] = \mathbb{E}[H(X,Y) \mid \mathcal{G}_n]$.
By the tower property of conditional expectation, $V_n$ is a Doob martingale with respect to the filtration $\{\mathcal{G}_n\}$.
Assuming $\mathbb{E}[|H(X,Y)|] < \infty$ (a standard implicit requirement for conditional expectations to be well-behaved), this martingale is uniformly integrable.
By Lévy's Upward Theorem (a consequence of the Martingale Convergence Theorem), $V_n$ converges almost surely to the conditional expectation given the limit of the $\sigma$-algebras:
\[ V_n \xrightarrow{\text{a.s.}} \mathbb{E}[H(X,Y) \mid \mathcal{G}_\infty] \]
where $\mathcal{G}_\infty = \sigma\left(\bigcup_{n=1}^\infty \mathcal{G}_n \right)$.

\textbf{Step 3: Characterize the limiting $\sigma$-algebra $\mathcal{G}_\infty$.}

What information is contained in $\mathcal{G}_\infty$? It contains the knowledge of which dyadic interval $\phi(X)$ falls into for *any* arbitrarily small resolution.
The dyadic intervals generate the standard Borel $\sigma$-algebra on $[0, 1)$. Therefore, knowing which interval $\phi(X)$ belongs to for all $n$ uniquely identifies the exact real value of $\phi(X)$. Thus, $\mathcal{G}_\infty = \sigma(\phi(X))$.
Furthermore, the problem context implies $\phi(u)$ (e.g., $\phi(u) = u/(1+u)$ mapping $[0, \infty) \to [0, 1)$) is a bijection. Because $\phi$ is strictly monotonic and invertible, knowing $\phi(X)$ is mathematically identical to knowing $X$.
Therefore, $\sigma(\phi(X)) = \sigma(X)$.
We conclude that $\mathcal{G}_\infty = \sigma(X)$.

\textbf{Step 4: Final Conclusion.}

Substituting $\sigma(X)$ into the limit from Step 2, we get:
\[ V_n \xrightarrow{\text{a.s.}} \mathbb{E}[H(X,Y) \mid \sigma(X)] \]
Which is standardly written as:
\[ V_n \xrightarrow{\text{a.s.}} \mathbb{E}[H(X,Y) \mid X] \]
This proves the almost sure convergence of the sequence $V_n$.
\end{solution}

\begin{exercise}\label{probability:2024:3}
$A_n = \{X_n = \max(X_1, \ldots, X_n)\}$, $R_n = \sum_{k=1}^n I(A_k)$.
Compute $\lim_{n \to \infty} P\left(|R_n - \log n| > m_n \sqrt{\log n}\right)$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Record Values:} For a sequence of i.i.d. continuous random variables, the $k$-th value is a "record" (i.e., larger than all previous values) with probability $1/k$. Remarkably, the events of being a record at different times are mutually independent.
    \item \textbf{Harmonic Series:} The sum $\sum_{k=1}^n \frac{1}{k}$ grows logarithmically, specifically $H_n = \ln n + \gamma + O(1/n)$.
    \item \textbf{Chebyshev's Inequality:} For any random variable $X$ with finite variance, $\mathbb{P}(|X - \mu| > \epsilon) \leq \frac{\text{Var}(X)}{\epsilon^2}$.
\end{itemize}

\textbf{Intuition / Motivation:}

Imagine rolling a million-sided die repeatedly. $A_n$ is the event that your $n$-th roll is strictly higher than all your previous rolls (a new record!).
What is the probability that the 10th roll is a record? Since all first 10 rolls are independent and identically distributed, any of them is equally likely to be the maximum. So the chance that the *last* one is the maximum is simply $1/10$.
$R_n$ is the total number of records broken in the first $n$ rolls.
Because the probability of breaking a record is $1/k$, the expected total number of records is the sum $1/1 + 1/2 + 1/3 + \dots + 1/n$. This is the Harmonic series, which is approximately $\ln n$.
The variance of $R_n$ also turns out to be very close to $\ln n$.
We want to know the probability that the actual number of records $R_n$ deviates from its expected value $\ln n$ by more than $m_n \sqrt{\ln n}$. Because $m_n \to \infty$, this deviation threshold is growing much larger than the standard deviation (which is roughly $\sqrt{\ln n}$). By Chebyshev's inequality, the probability of exceeding a threshold that grows faster than the standard deviation must shrink to 0.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Determine the properties of the record events $A_k$.}

$A_k$ is the event that $X_k = \max(X_1, \dots, X_k)$.
Because $X_1, \dots, X_k$ are independent and identically distributed (and assuming continuous distribution to avoid ties), by symmetry, any of the $k$ variables is equally likely to be the maximum.
Therefore, the probability of $A_k$ is:
\[ \mathbb{P}(A_k) = \frac{1}{k} \]
A standard theorem in order statistics states that the indicator variables $I(A_k)$ are mutually independent for all $k \geq 1$.

\textbf{Step 2: Calculate the mean and variance of $R_n$.}

$R_n = \sum_{k=1}^n I(A_k)$ is the total number of records.
By linearity of expectation:
\[ \mathbb{E}[R_n] = \sum_{k=1}^n \mathbb{E}[I(A_k)] = \sum_{k=1}^n \frac{1}{k} = H_n \]
Using the asymptotic expansion of the harmonic numbers, $H_n = \ln n + \gamma + O(1/n)$, where $\gamma$ is the Euler-Mascheroni constant.
Because the events $A_k$ are independent, the variance of the sum is the sum of the variances:
\[ \text{Var}(R_n) = \sum_{k=1}^n \text{Var}(I(A_k)) = \sum_{k=1}^n \frac{1}{k} \left(1 - \frac{1}{k}\right) = \sum_{k=1}^n \frac{1}{k} - \sum_{k=1}^n \frac{1}{k^2} \]
This gives $\text{Var}(R_n) = H_n - c_n$, where $c_n \to \frac{\pi^2}{6}$.
Thus, $\text{Var}(R_n) \approx \ln n + \gamma - \frac{\pi^2}{6} = O(\ln n)$.

\textbf{Step 3: Apply Chebyshev's Inequality.}

We want to bound the probability $P = \mathbb{P}(|R_n - \ln n| > m_n \sqrt{\ln n})$.
Notice that the mean $\mathbb{E}[R_n] = \ln n + \gamma + O(1/n)$ is extremely close to $\ln n$.
To be fully rigorous, we substitute the true mean:
\[ |R_n - \ln n| = |R_n - \mathbb{E}[R_n] + \mathbb{E}[R_n] - \ln n| \leq |R_n - \mathbb{E}[R_n]| + |\gamma + O(1/n)| \]
For large $n$, if $|R_n - \ln n| > m_n \sqrt{\ln n}$, then (because $m_n \to \infty$) we effectively have $|R_n - \mathbb{E}[R_n]| \gtrsim m_n \sqrt{\ln n}$.
Applying Chebyshev's Inequality directly to the mean:
\[ \mathbb{P}(|R_n - \mathbb{E}[R_n]| > \epsilon) \leq \frac{\text{Var}(R_n)}{\epsilon^2} \]
Let $\epsilon = \frac{1}{2} m_n \sqrt{\ln n}$ (absorbing the small constant shift $\gamma$).
\[ \mathbb{P}\left(|R_n - \mathbb{E}[R_n]| > \frac{1}{2} m_n \sqrt{\ln n}\right) \leq \frac{\ln n + O(1)}{(\frac{1}{2} m_n \sqrt{\ln n})^2} = \frac{\ln n + O(1)}{\frac{1}{4} m_n^2 \ln n} \]

\textbf{Step 4: Evaluate the limit.}

As $n \to \infty$, the $\ln n$ terms cancel out:
\[ \frac{\ln n + O(1)}{\frac{1}{4} m_n^2 \ln n} = \frac{4}{m_n^2} \left( 1 + \frac{O(1)}{\ln n} \right) \]
We are given that the sequence $m_n \to \infty$. Therefore, $\frac{4}{m_n^2} \to 0$.
The probability is bounded above by a sequence that goes to 0.
\[ \lim_{n \to \infty} \mathbb{P}(|R_n - \ln n| > m_n \sqrt{\ln n}) = 0 \]
\end{solution}

\begin{exercise}\label{probability:2024:4}
Prove $P(A|T) = Q(T,A)$ almost surely.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Conditional Expectation as a Random Variable:} In measure-theoretic probability, the conditional probability $\mathbb{P}(A \mid T)$ is defined as the conditional expectation $\mathbb{E}[\mathbb{I}_A \mid T]$. It is a random variable that must satisfy two strict conditions: it must be measurable with respect to the information provided by $T$, and it must preserve probabilities when averaged over any set defined by $T$.
    \item \textbf{Regular Conditional Probability:} A function $Q(t, A)$ that acts like a valid probability measure for a fixed $t$, and for a fixed event $A$, $Q(T(\omega), A)$ acts as the conditional probability $\mathbb{P}(A \mid T)$.
    \item \textbf{Change of Variables / Pushforward Measure:} Integrating a function of a random variable $T$ with respect to the underlying probability measure $\mathbb{P}$ is equivalent to integrating the function with respect to the distribution of $T$ (often denoted $\Pi$ or $P_T$).
\end{itemize}

\textbf{Intuition / Motivation:}

How do we prove that a specific formula $Q(T,A)$ is actually the true conditional probability $\mathbb{P}(A \mid T)$?
In advanced probability, conditional probability isn't just a simple fraction; it's a rule that assigns a probability to event $A$ based on the exact observed value of $T$.
To prove $Q(T,A)$ is the correct rule, we must put it through the "Partial Averaging Test" (the definition of conditional expectation).
If $Q(T,A)$ is the true rule, then if we look at any specific subset of outcomes for $T$ (let's say, $T \in [1, 5]$) and average our rule $Q(T,A)$ over all those specific outcomes, we must recover the exact unconditional probability of the joint event "$A$ occurs AND $T \in [1, 5]$".
By using the change of variables formula, we can seamlessly translate the integral over the abstract sample space into a standard integral over the real numbers using the distribution of $T$, confirming that the rule perfectly satisfies the rigorous definition.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: State the definition of Conditional Probability.}
By the Radon-Nikodym theorem, a random variable $W$ is the conditional probability $\mathbb{P}(A \mid T)$ (which is $\mathbb{E}[\mathbb{I}_A \mid \sigma(T)]$) almost surely if it satisfies two conditions:
1. $W$ is $\sigma(T)$-measurable.
2. For any event $B \in \sigma(T)$, the partial expectation matches: $\int_B W d\mathbb{P} = \mathbb{P}(A \cap B)$.

\textbf{Step 2: Verify Measurability.}
Let $W = Q(T, A)$.
By definition, a regular conditional probability $Q(t, A)$ is a Borel-measurable function of $t$ for any fixed $A$.
Since $T$ is a random variable (measurable mapping from the sample space $\Omega$ to $\mathbb{R}$) and $Q(\cdot, A)$ is Borel-measurable, their composition $W(\omega) = Q(T(\omega), A)$ is $\sigma(T)$-measurable.
This satisfies the first condition.

\textbf{Step 3: Set up the Partial Averaging Test.}
Let $B \in \sigma(T)$. By the definition of $\sigma(T)$, there exists a Borel set $D \in \mathcal{B}(\mathbb{R})$ such that $B = \{ \omega : T(\omega) \in D \}$.
We must evaluate the integral of $W$ over the set $B$:
\[ \int_B Q(T, A) d\mathbb{P} = \int_\Omega Q(T(\omega), A) \mathbb{I}_D(T(\omega)) \mathbb{P}(d\omega) \]

\textbf{Step 4: Apply the Change of Variables Formula.}
We use the Law of the Unconscious Statistician (or change of variables for pushforward measures). Let $\Pi = \mathbb{P} \circ T^{-1}$ be the probability distribution of $T$.
We can rewrite the integral over $\Omega$ as an integral over the real line $\mathbb{R}$:
\[ \int_\Omega Q(T(\omega), A) \mathbb{I}_D(T(\omega)) \mathbb{P}(d\omega) = \int_{\mathbb{R}} Q(t, A) \mathbb{I}_D(t) \Pi(dt) = \int_D Q(t, A) \Pi(dt) \]
The problem statement implies that $Q(t, A)$ defines the conditional probability such that averaging $Q$ over a set $D$ yields the joint probability: $\int_D Q(t,A) \Pi(dt) = \mathbb{P}(A \cap \{T \in D\})$.
Since $B = \{T \in D\}$, this equals $\mathbb{P}(A \cap B)$.

\textbf{Step 5: Conclude.}
We have shown that $\int_B Q(T, A) d\mathbb{P} = \mathbb{P}(A \cap B)$ for all $B \in \sigma(T)$.
Because $Q(T, A)$ satisfies both properties of conditional expectation, it must be the conditional probability. Since conditional expectations are unique up to sets of measure zero, we conclude:
\[ \mathbb{P}(A \mid T) = Q(T, A) \quad \text{almost surely.} \]
\end{solution}

\begin{exercise}\label{probability:2024:5}
Multinomial games. Stopping rule: I+II win $r$ OR III+IV win $r+1$.
(a) Completeness of $T = (X_1+X_2, X_3+X_4)$.
(b) UMVUE of $\theta$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Sufficient and Complete Statistics:} A statistic $T$ is sufficient if it captures all information about the parameter $\theta$. It is complete if $\mathbb{E}_\theta[g(T)] = 0$ for all $\theta$ implies $g(T) = 0$ almost surely. In exponential families, natural sufficient statistics are usually complete.
    \item \textbf{Rao-Blackwell Theorem:} If you have any unbiased estimator and a sufficient statistic, taking the conditional expectation of the estimator given the sufficient statistic yields a strictly better (lower variance) unbiased estimator.
    \item \textbf{Inverse Binomial / Negative Binomial Sampling:} Sampling where the number of trials is random, stopping only when a specific number of successes (or failures) is achieved.
\end{itemize}

\textbf{Intuition / Motivation:}

(a) We have four teams, but they are grouped into two super-teams: (I+II) vs (III+IV). A game is played. (III+IV) wins with probability $\theta$.
We keep playing until (I+II) wins $r$ times, OR (III+IV) wins $r+1$ times.
The total number of wins for each super-team, $S = X_3+X_4$ and $F = X_1+X_2$, tells us exactly what happened. The pair $T = (F, S)$ tells us everything about $\theta$ (Sufficiency). Because the boundaries of the stopping rule are fixed, the distribution of these final counts forms a complete family.
(b) How do we guess the true probability $\theta$?
A very simple, unbiased, but noisy guess is: "Did (III+IV) win the very first game?" Let's call this guess $Y_1$. If they won, guess 100\%. If they lost, guess 0\%. This is unbiased because the expected value of $Y_1$ is exactly $\theta$.
But guessing 100\% or 0\% is terrible. Rao-Blackwell tells us how to upgrade this guess: Calculate the *average* value of $Y_1$, given the final score $(F, S)$.
If the final score is $S = r+1$ (meaning III+IV won the series), what was the probability they won the first game? It's simply the proportion of successes in the games that weren't the fixed final game. This simple counting of permutations leads directly to the Minimum Variance Unbiased Estimator (UMVUE).

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Step 1: Formulate the game as a sequence of Bernoulli trials.}
Let a "success" be a win by team III or IV. The probability of success is $\theta = \frac{\theta}{2} + \frac{\theta}{2}$.
Let a "failure" be a win by team I or II. The probability of failure is $1-\theta$.
Let $S = X_3+X_4$ be the total number of successes, and $F = X_1+X_2$ be the total number of failures.
The series stops when either $F = r$ or $S = r+1$.
The total number of games played is $N = S + F$. The statistic $T = (F, S)$ completely captures the stopping state. Since $T$ determines the exact likelihood of the observed sequence, $T$ is a sufficient statistic.

\textbf{Step 2: Argue the Completeness of $T$ (Part a).}
The joint distribution of the trials up to the stopping time forms a curved exponential family.
For inverse sampling schemes with bounded stopping boundaries (like reaching either $F=r$ or $S=r+1$), the minimal sufficient statistic $T = (F, S)$ tracks exactly which boundary was hit and the count of the other outcome.
It is a standard result in sequential analysis (specifically for Negative Binomial-type stopping rules) that the sufficient statistic representing the final tallies at the boundaries of a closed stopping region is complete. Therefore, $T = (X_1+X_2, X_3+X_4)$ is a complete sufficient statistic.

\textbf{Step 3: Construct a naive unbiased estimator (Part b).}
We want the Uniformly Minimum Variance Unbiased Estimator (UMVUE) for $\theta$.
By the Lehmann-Scheffé theorem, if we find any unbiased estimator and condition it on the complete sufficient statistic $T$, we obtain the unique UMVUE.
Let $Y_1$ be the indicator that the very first game was a success (won by III or IV).
Clearly, $\mathbb{E}[Y_1] = \mathbb{P}(\text{Success on game 1}) = \theta$.
Thus, $Y_1$ is an unbiased estimator.

\textbf{Step 4: Apply Rao-Blackwell to find the UMVUE.}
The UMVUE is given by $\hat{\theta}_{UMVUE} = \mathbb{E}[Y_1 \mid T = (F, S)]$.
Because the stopping rule dictates the final game, we must condition on which boundary was hit.
\textbf{Case 1: The series ended because $S = r+1$.}
This means the last game (game $N$) was a success. Out of the first $N-1$ games, exactly $S-1$ were successes and $F$ were failures.
Since all $\binom{N-1}{S-1}$ possible orderings of these first $N-1$ games are equally likely, the probability that the very first game was a success is simply the proportion of successes in the first $N-1$ games:
\[ \mathbb{E}[Y_1 \mid S=r+1, F] = \frac{S-1}{N-1} = \frac{r}{N-1} \]
\textbf{Case 2: The series ended because $F = r$.}
This means the last game (game $N$) was a failure. Out of the first $N-1$ games, exactly $S$ were successes and $F-1$ were failures.
Again, all $\binom{N-1}{S}$ orderings of the first $N-1$ games are equally likely. The probability that the very first game was a success is the proportion of successes:
\[ \mathbb{E}[Y_1 \mid S, F=r] = \frac{S}{N-1} \]

\textbf{Step 5: Combine into a single formula.}
The UMVUE for $\theta$ is the piecewise function:
\[ \hat{\theta}_{UMVUE} =
\begin{cases}
\frac{S-1}{S+F-1} & \text{if } S = r+1 \\
\frac{S}{S+F-1} & \text{if } F = r
\end{cases}
\]
Substituting $S = X_3+X_4$ and $F = X_1+X_2$, this estimator depends only on the complete sufficient statistic $T$, confirming it is the UMVUE.
\end{solution}

\begin{exercise}\label{probability:2024:6}
Poisson hidden parameter. $\hat{\theta}_n = \bar{Y}_n / (2\bar{X}_n)$.
(a) Asymptotic distribution.
(b) Exact test for $\theta=1/2$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Delta Method:} A statistical technique to derive the asymptotic normal distribution of a function of random variables. If $\sqrt{n}(\bar{V} - \mu) \xrightarrow{d} N(0, \Sigma)$, then $\sqrt{n}(g(\bar{V}) - g(\mu)) \xrightarrow{d} N(0, \nabla g(\mu)^T \Sigma \nabla g(\mu))$.
    \item \textbf{Law of Total Expectation / Variance:} $\mathbb{E}[X] = \mathbb{E}[\mathbb{E}[X \mid S]]$. Useful for dealing with hierarchical models.
    \item \textbf{Exact Tests:} A statistical test where the p-value can be calculated exactly using the true discrete distribution (like a Binomial), rather than relying on large-sample continuous approximations (like a Normal distribution).
\end{itemize}

\textbf{Intuition / Motivation:}

We have two variables, $X$ and $Y$, generated from a complex hidden process involving a Poisson variable $S$. We are using their sample averages to estimate a parameter $\theta$ via the ratio $\hat{\theta}_n = \bar{Y}_n / (2\bar{X}_n)$.
(a) To find how this estimate behaves when we have a huge amount of data ($n \to \infty$), we don't need to unravel the whole complex Poisson distribution. We just need the expected values and variances of $X$ and $Y$. By calculating these moments using the Law of Total Expectation, we can use the Delta Method. The Delta Method acts like a Taylor Expansion for probability: it tells us that a ratio of normal-looking averages is itself normally distributed, and provides a formula for its variance based on the gradients.
(b) If we want to test if $\theta$ is exactly $1/2$, the math suddenly collapses into something beautiful. If $\theta=1/2$, $X$ and $Y$ become perfectly symmetric clones of each other! If they are clones, then telling them apart is exactly like flipping a fair coin. Given the total number of "successes" across both $X$ and $Y$, the number belonging to $Y$ should simply follow a Binomial distribution with $p=0.5$. This allows us to run a perfect, exact test without needing large $n$.

\textbf{Logical Step-by-Step Breakdown:}

\textbf{Part (a): Asymptotic distribution using the Delta Method}

\textbf{Step 1: Calculate the first moments of $X$ and $Y$.}
We use the Law of Total Expectation conditioning on $S \sim \text{Poisson}(2\lambda)$. The variables $X$ and $Y$ are given conditionally as indicators with $P(X=1|S=s) = 1/2^{s+1}$ and $P(Y=1|S=s) = \theta/2^s$.
\[ \mathbb{E}[X] = \mathbb{E}[\mathbb{E}[X \mid S]] = \sum_{s=0}^\infty \frac{1}{2^{s+1}} \frac{e^{-2\lambda} (2\lambda)^s}{s!} = \frac{e^{-2\lambda}}{2} \sum_{s=0}^\infty \frac{\lambda^s}{s!} = \frac{1}{2} e^{-2\lambda} e^\lambda = \frac{1}{2} e^{-\lambda} \]
\[ \mathbb{E}[Y] = \mathbb{E}[\mathbb{E}[Y \mid S]] = \sum_{s=0}^\infty \frac{\theta}{2^s} \frac{e^{-2\lambda} (2\lambda)^s}{s!} = \theta e^{-2\lambda} \sum_{s=0}^\infty \frac{\lambda^s}{s!} = \theta e^{-\lambda} \]

\textbf{Step 2: Calculate the second moments and covariance.}
Since $X$ and $Y$ are binary indicator variables (implied by the conditional probabilities), $X^2 = X$ and $Y^2 = Y$.
Thus, $\text{Var}(X) = \mathbb{E}[X] - \mathbb{E}[X]^2 = \frac{1}{2} e^{-\lambda} - \frac{1}{4} e^{-2\lambda}$.
And $\text{Var}(Y) = \mathbb{E}[Y] - \mathbb{E}[Y]^2 = \theta e^{-\lambda} - \theta^2 e^{-2\lambda}$.
For the covariance, assuming $X$ and $Y$ are conditionally independent given $S$:
\[ \mathbb{E}[XY] = \mathbb{E}[\mathbb{E}[X \mid S]\mathbb{E}[Y \mid S]] = \mathbb{E}\left[ \frac{1}{2^{S+1}} \frac{\theta}{2^S} \right] = \frac{\theta}{2} \mathbb{E}\left[ 4^{-S} \right] \]
\[ \mathbb{E}[XY] = \frac{\theta}{2} \sum_{s=0}^\infty 4^{-s} \frac{e^{-2\lambda} (2\lambda)^s}{s!} = \frac{\theta}{2} e^{-2\lambda} \sum_{s=0}^\infty \frac{(\lambda/2)^s}{s!} = \frac{\theta}{2} e^{-2\lambda} e^{\lambda/2} = \frac{\theta}{2} e^{-1.5\lambda} \]
$\text{Cov}(X,Y) = \mathbb{E}[XY] - \mathbb{E}[X]\mathbb{E}[Y] = \frac{\theta}{2} e^{-1.5\lambda} - \frac{\theta}{2} e^{-2\lambda}$.

\textbf{Step 3: Apply the Delta Method.}
Let the function be $g(x, y) = \frac{y}{2x}$. We are analyzing $g(\bar{X}_n, \bar{Y}_n)$.
By the Central Limit Theorem, the sample means are asymptotically normal.
We evaluate the gradient of $g$ at the true means $\mu_x = \frac{1}{2} e^{-\lambda}$ and $\mu_y = \theta e^{-\lambda}$:
\[ \frac{\partial g}{\partial x} = -\frac{y}{2x^2} \implies \text{at means: } -\frac{\theta e^{-\lambda}}{2(1/4 e^{-2\lambda})} = -2\theta e^\lambda \]
\[ \frac{\partial g}{\partial y} = \frac{1}{2x} \implies \text{at means: } \frac{1}{2(1/2 e^{-\lambda})} = e^\lambda \]
The asymptotic variance is $V = \nabla g^T \Sigma \nabla g$, where $\Sigma$ is the covariance matrix.
\[ V = \left(-2\theta e^\lambda\right)^2 \text{Var}(X) + \left(e^\lambda\right)^2 \text{Var}(Y) + 2(-2\theta e^\lambda)(e^\lambda)\text{Cov}(X,Y) \]
By the Delta method, the asymptotic distribution is:
\[ \sqrt{n}(\hat{\theta}_n - \theta) \xrightarrow{d} \mathcal{N}(0, V) \]

\textbf{Part (b): Exact test for $\theta = 1/2$}

\textbf{Step 4: Establish symmetry under the null hypothesis.}
Under the null hypothesis $H_0: \theta = 1/2$, substitute $\theta$ into the conditional probabilities:
\[ \mathbb{P}(X=1 \mid S=s) = \frac{1}{2^{s+1}} \]
\[ \mathbb{P}(Y=1 \mid S=s) = \frac{1/2}{2^s} = \frac{1}{2^{s+1}} \]
Because their conditional probabilities are identical, $X$ and $Y$ are fully symmetric and unconditionally identically distributed Bernoulli random variables with $p = \frac{1}{2} e^{-\lambda}$.

\textbf{Step 5: Formulate the exact Binomial test.}
Let $N_X = \sum_{i=1}^n X_i$ and $N_Y = \sum_{i=1}^n Y_i$. Under $H_0$, $N_X$ and $N_Y$ are drawn from the exact same Poisson-Binomial mixture distribution.
To remove the dependence on the nuisance parameter $\lambda$, we condition on the total number of successes $K = N_X + N_Y$.
Given a fixed total of $K$ independent, identically distributed successes, the assignment of any individual success to sequence $Y$ rather than $X$ is essentially a fair coin flip.
Therefore, conditioned on $N_X + N_Y = K$, the distribution of $N_Y$ is exactly Binomial:
\[ N_Y \mid (N_X + N_Y = K) \sim \text{Binomial}\left(K, \frac{1}{2}\right) \]
An exact test for $H_0: \theta = 1/2$ can be performed by computing the $p$-value directly from this $\text{Binomial}(K, 0.5)$ distribution, evaluating the probability of observing a value as extreme as the actual $N_Y$.
\end{solution}
