\section{2017 Algebra \& Number Theory}

\subsection*{Individual}
\begin{exercise}
\label{algebra:2017:1}
Let $\mathbb{Q}_p$ denote the field of $p$-adic numbers and let $\mathbb{Z}_p$ denote the ring of $p$-adic integers ($p$ is a prime number).

\begin{enumerate}
\item Show that for every integer $k \geq 0$, $(p^{-k}\mathbb{Z}_p)/\mathbb{Z}_p \cong \mathbb{Z}/p^k\mathbb{Z}$ as abelian groups.

\item Determine the endomorphism ring of the abelian group $(p^{-k}\mathbb{Z}_p)/\mathbb{Z}_p$ ($k \geq 0$).

\item Determine the endomorphism ring of the abelian group $\mathbb{Q}_p/\mathbb{Z}_p$.

\item Determine the endomorphism ring of the abelian group $\mathbb{Q}/\mathbb{Z}$.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{$p$-adic Integers ($\mathbb{Z}_p$):} The ring of numbers represented as $\sum_{i=0}^\infty a_i p^i$, where $0 \leq a_i < p$.
    \item \textbf{$p$-adic Field ($\mathbb{Q}_p$):} The field of fractions of $\mathbb{Z}_p$, which allows for a finite number of negative powers of $p$.
    \item \textbf{Quotient Group:} A group structure formed by identifying elements that differ by a member of a subgroup.
    \item \textbf{Endomorphism Ring ($\text{End}(G)$):} The set of all homomorphisms from $G$ to itself, forming a ring under addition (pointwise) and multiplication (composition).
    \item \textbf{Prüfer $p$-group ($\mathbb{Z}[p^\infty]$):} The limit of cyclic groups of power-of-$p$ order, often viewed as the set of $p$-power roots of unity.
\end{itemize}

\textbf{Steps:}

\textbf{(1) Identify the structure of the quotient.}
The space $p^{-k}\mathbb{Z}_p$ contains numbers where the smallest power of $p$ is $p^{-k}$.
An element $x \in p^{-k}\mathbb{Z}_p$ can be written as:
\[ x = \frac{a_{-k}}{p^k} + \frac{a_{-k+1}}{p^{k-1}} + \dots + \frac{a_{-1}}{p} + a_0 + a_1 p + \dots \]
When we quotient by $\mathbb{Z}_p$, all terms $a_i p^i$ for $i \geq 0$ are set to zero.
What remains is a fraction of the form $\frac{m}{p^k}$, where $0 \leq m < p^k$.
Addition in this quotient group corresponds exactly to the addition of these fractions modulo $1$, which is isomorphic to the cyclic group $\mathbb{Z}/p^k\mathbb{Z}$.

\textbf{(2) Determine the endomorphism ring for fixed $k$.}
The group we found in part (1) is the cyclic group $C_{p^k}$.
A homomorphism from $C_{p^k}$ to itself is uniquely determined by the image of the generator $1$.
Since the generator can be mapped to any of the $p^k$ elements in the group, and these maps preserve the additive and multiplicative structure of the underlying ring, we have $\text{End}(\mathbb{Z}/p^k\mathbb{Z}) \cong \mathbb{Z}/p^k\mathbb{Z}$.

\textbf{(3) Extend to the Prüfer $p$-group.}
The group $\mathbb{Q}_p / \mathbb{Z}_p$ is the union (direct limit) of all the cyclic groups $p^{-k}\mathbb{Z}_p / \mathbb{Z}_p$ for $k = 1, 2, \dots$.
An endomorphism $f$ of this infinite group must act on each "layer" $\mathbb{Z}/p^k\mathbb{Z}$ in a way that is compatible with the layer below it.
This means $f$ is a sequence of values $(a_1, a_2, \dots)$ where $a_k \in \mathbb{Z}/p^k\mathbb{Z}$ and $a_{k+1} \equiv a_k \pmod{p^k}$.
This inverse limit construction is the exact definition of the ring of $p$-adic integers $\mathbb{Z}_p$.

\textbf{(4) Generalize to the full rational quotient.}
The group $\mathbb{Q} / \mathbb{Z}$ can be decomposed using the Chinese Remainder Theorem logic (or partial fraction decomposition) into its $p$-primary parts:
\[ \mathbb{Q} / \mathbb{Z} \cong \bigoplus_{p \text{ prime}} (\mathbb{Q}_p / \mathbb{Z}_p) \]
A homomorphism between different $p$-primary parts is always zero because the orders of elements are coprime.
Therefore, an endomorphism of the whole group is simply a product of endomorphisms of each individual part.
The resulting ring is the product of all $p$-adic integer rings: $\text{End}(\mathbb{Q} / \mathbb{Z}) \cong \prod_p \mathbb{Z}_p = \hat{\mathbb{Z}}$.

\textbf{Intuition:}
The quotient $\mathbb{Q}_p / \mathbb{Z}_p$ captures the "fractional" part of $p$-adic numbers. Because $p$-adic numbers are built on layers of $p^k$, the automorphisms must respect this layered structure. For a single layer, it's just modular arithmetic. For the infinite stack of layers, you need the full $p$-adic number system to describe the symmetries. When you look at all primes at once ($\mathbb{Q}/\mathbb{Z}$), the primes don't "talk" to each other, so the symmetry of the whole is just the collection of the symmetries of each prime part.
\end{solution}

\begin{exercise}
\label{algebra:2017:2}
Let $A$ be a finite abelian group and let $\phi: A \to A$ be an endomorphism. Put
\[A_{\mathrm{nil}} := \{x \in A \mid \phi^k(x) = 0 \text{ for some } k \geq 1\}.\]

\begin{enumerate}
\item Show that there is a subgroup $A_0$ of $A$ such that $\phi$ restricts to an automorphism of $A_0$ and $A = A_0 \oplus A_{\mathrm{nil}}$.

\item Show that such a subgroup is unique.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Endomorphism:} A mapping from a group to itself that preserves the group operation.
    \item \textbf{Kernel and Image Chains:} The sequences of subgroups formed by applying an operator repeatedly (e.g., $\text{ker}(\phi), \text{ker}(\phi^2), \dots$).
    \item \textbf{Fitting's Lemma:} A foundational result in module theory that decomposes a module into an invertible part and a nilpotent part.
    \item \textbf{Stabilization:} In any finite structure, nested sequences must eventually stop changing.
\end{itemize}

\textbf{Steps:}

\textbf{(1) Identify the stable image and kernel.}
Since $A$ is a finite group, the sequence of images $A \supseteq \phi(A) \supseteq \phi^2(A) \supseteq \dots$ and the sequence of kernels $0 \subseteq \ker(\phi) \subseteq \ker(\phi^2) \subseteq \dots$ must eventually stabilize.
There exists an integer $N$ such that $\phi^N(A) = \phi^{N+1}(A) = \dots$ and $\ker(\phi^N) = \ker(\phi^{N+1}) = \dots$.
Let $A_0 = \phi^N(A)$ and $A_{\text{nil}} = \ker(\phi^N)$.

\textbf{(2) Show that $\phi$ is an automorphism of $A_0$.}
By definition, $\phi(A_0) = \phi(\phi^N(A)) = \phi^{N+1}(A)$.
Since the sequence has stabilized at $N$, we have $\phi^{N+1}(A) = \phi^N(A) = A_0$.
Thus, $\phi$ maps $A_0$ onto itself. Because $A_0$ is a finite group, a surjective map is automatically an automorphism (one-to-one and onto).

\textbf{(3) Prove the direct sum decomposition.}
First, check the intersection. If $x \in A_0 \cap A_{\text{nil}}$, then $x = \phi^N(y)$ and $\phi^N(x) = 0$.
This implies $\phi^{2N}(y) = 0$. Since $\ker(\phi^{2N}) = \ker(\phi^N)$ (stability), we have $\phi^N(y) = 0$, so $x = 0$.
Next, check the sizes. By the first isomorphism theorem, $|A| = |\text{im}(\phi^N)| \cdot |\ker(\phi^N)| = |A_0| \cdot |A_{\text{nil}}|$.
Since the intersection is trivial and the product of the subgroup sizes equals the group size, we have $A = A_0 \oplus A_{\text{nil}}$.

\textbf{(4) Establish uniqueness.}
Suppose $A = B \oplus A_{\text{nil}}$ where $\phi|_B$ is an automorphism.
Then $\phi^N$ is also an automorphism of $B$, so $\phi^N(B) = B$.
Since $B \subseteq A$, we have $B = \phi^N(B) \subseteq \phi^N(A) = A_0$.
Since $|B| = |A| / |A_{\text{nil}}| = |A_0|$, it must be that $B = A_0$.

\textbf{Intuition:}
This decomposition is like separating a signal from its noise. The $A_{\text{nil}}$ part consists of everything that eventually "dies out" under the action of $\phi$. The $A_0$ part is the "core" where the operation is perfectly preserved and reversible. In a finite world, everything eventually either circles back to where it was or disappears; this theorem organizes the group into these two distinct behaviors.
\end{solution}

\begin{exercise}
\label{algebra:2017:3}
Let $L/F$ be a Galois field extension, not necessarily finite. Let $x \in L$.

\begin{enumerate}
\item Show that the set $\mathcal{P}$ of subextensions of $L/F$ not containing $x$ has a maximal element $E$. Let $K/E$ be a nontrivial finite extension contained in $L$. Show that $x \in K$.

\item Let $K'$ be the Galois closure of $K/E$ in $L$. Show that there exists $g \in G = \mathrm{Gal}(K'/E)$ such that $gx \neq x$.

\item Deduce that $K/E$ is a cyclic Galois extension.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Galois Extension:} An extension where the only elements fixed by the entire automorphism group are those in the base field.
    \item \textbf{Zorn's Lemma:} A principle used to find maximal elements in sets where every chain has an upper bound.
    \item \textbf{Galois Correspondence:} The relationship between subfields of a Galois extension and subgroups of the Galois group.
    \item \textbf{Cyclic $p$-groups:} A finite group with exactly one maximal subgroup is cyclic of order $p^n$.
\end{itemize}

\textbf{Steps:}

\textbf{(1) Find a maximal subextension.}
Let $\mathcal{P}$ be the set of all subextensions $E$ of $L/F$ such that $x \notin E$.
Given any chain of fields $E_1 \subseteq E_2 \subseteq \dots$ in $\mathcal{P}$, their union $U = \bigcup E_i$ is a field. Since $x \notin E_i$ for all $i$, $x \notin U$, so $U \in \mathcal{P}$.
By Zorn's Lemma, $\mathcal{P}$ has a maximal element $E$.
This means any field strictly containing $E$ must contain $x$. Thus, for any nontrivial extension $K/E$, $x \in K$.

\textbf{(2) Use the Galois property to move $x$.}
The extension $L/E$ is still Galois. Let $K'$ be the Galois closure of $K/E$ in $L$.
Since $x \in K \subseteq K'$ and $x \notin E$, the element $x$ is not in the fixed field of the group $G = \text{Gal}(K'/E)$.
By the fundamental property of Galois extensions, there must exist some $g \in G$ such that $g(x) \neq x$.

\textbf{(3) Identify the group structure.}
The property from Step 1 implies that $E(x)$ is the unique smallest field extension of $E$.
Under the Galois correspondence, subfields containing $E$ correspond to subgroups of $G$.
A unique smallest subfield corresponds to a \textbf{unique maximal subgroup} of $G$.
In finite group theory, a group with a unique maximal subgroup is necessarily a cyclic $p$-group for some prime $p$.
Since $G = \text{Gal}(K'/E)$ is cyclic, and $K$ is an intermediate field ($E \subseteq K \subseteq K'$), the extension $K/E$ must also be a cyclic Galois extension.

\textbf{Intuition:}
The existence of the "bottleneck" element $x$ (an element that is in \textit{every} extension of $E$) is a very strong constraint. It means there is only one "direction" you can move to grow the field $E$ slightly. In the language of groups, this means there is only one way to start shrinking the symmetry group. This linear, one-way structure is the defining characteristic of cyclic groups.
\end{solution}

\begin{exercise}
\label{algebra:2017:4}
The goal of this problem is to prove the Chevalley-Warning theorem. Let $p$ be a prime number and $q$ a power of $p$.

\begin{enumerate}
\item Let $0 \leq a < q - 1$ be an integer. Show that $S(X^a) := \sum_{x \in \mathbb{F}_q} x^a$ equals 0. Here we adopt the convention $x^0 = 1$ in $\mathbb{F}_q$ even for $x = 0$.

\item Let $f_1, \ldots, f_m \in \mathbb{F}_q[X_1, \ldots, X_n]$ be polynomials in $n$ variables satisfying
\[\sum_{i=1}^m \deg(f_i) < n.\]
Show that
\[S(P) := \sum_{(x_1,\ldots,x_n) \in \mathbb{F}_q^n} P(x_1,\ldots,x_n) = 0\]
where $P = \prod_{i=1}^m (1 - f_i^{q-1})$.

Deduce that $p$ divides the cardinality of the set
\[V = \{(x_1, \ldots, x_n) \in \mathbb{F}_q^n \mid f_i(x_1, \ldots, x_n) = 0 \quad \forall i\}.\]
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Finite Field ($\mathbb{F}_q$):} A field with $q = p^n$ elements. The non-zero elements form a cyclic multiplicative group of order $q-1$.
    \item \textbf{Fermat's Little Theorem (Extended):} For any $x \in \mathbb{F}_q$, $x^q = x$. This means $x^{q-1} = 1$ if $x \neq 0$ and $x^{q-1} = 0$ if $x = 0$.
    \item \textbf{Geometric Series Sum:} $\sum_{j=0}^{n-1} r^j = \frac{r^n - 1}{r - 1}$ if $r \neq 1$.
    \item \textbf{Characteristic P:} In a field of characteristic $p$, any sum of $p$ identical terms is $0$.
\end{itemize}

\textbf{Steps:}

\textbf{(1) Evaluate the sum of powers $S(X^a)$.}
We want to calculate $\sum_{x \in \mathbb{F}_q} x^a$.
- \textbf{Case $a = 0$:} By convention $0^0 = 1$. The sum has $q$ terms of $1$. In characteristic $p$, since $q$ is a power of $p$, $q \cdot 1 = 0$.
- \textbf{Case $0 < a < q - 1$:} The value $0^a = 0$, so we only sum over $x \in \mathbb{F}_q^\times$.
  Let $g$ be a generator of the cyclic group $\mathbb{F}_q^\times$. The elements are $\{g^0, g^1, \dots, g^{q-2}\}$.
  The sum is $\sum_{j=0}^{q-2} (g^a)^j$. Since $a < q-1$, $g^a \neq 1$.
  Applying the geometric series formula: $\frac{(g^a)^{q-1} - 1}{g^a - 1} = \frac{(g^{q-1})^a - 1}{g^a - 1} = \frac{1^a - 1}{g^a - 1} = 0$.

\textbf{(2) Construct the indicator polynomial $P$.}
Let $P = \prod_{i=1}^m (1 - f_i^{q-1})$. Consider its values for any $x \in \mathbb{F}_q^n$:
- If $x$ is a common zero (all $f_i(x) = 0$), then $P(x) = \prod (1 - 0) = 1$.
- If $x$ is not a common zero (some $f_j(x) \neq 0$), then $f_j(x)^{q-1} = 1$, so the $j$-th factor is $(1 - 1) = 0$. Thus $P(x) = 0$.
Therefore, the sum $S(P) = \sum_{x \in \mathbb{F}_q^n} P(x)$ is equal to the number of common zeros modulo $p$.

\textbf{(3) Use degree constraints to show the sum is zero.}
The total degree of $P$ is $(q-1) \sum \deg(f_i)$. Since $\sum \deg(f_i) < n$, we have $\deg(P) < n(q-1)$.
Every monomial in $P$ is of the form $X_1^{a_1} X_2^{a_2} \dots X_n^{a_n}$.
By the pigeonhole principle, since the total degree $\sum a_i < n(q-1)$, it is impossible for all $a_i$ to be equal to $q-1$.
There must be at least one variable $X_k$ whose exponent $a_k$ is strictly less than $q-1$.
The sum over the whole space $\mathbb{F}_q^n$ can be split into $n$ separate sums (one for each variable).
If even one of these internal sums is $0$ (which happens when $a_k < q-1$ by Step 1), the entire term sums to $0$.
Thus $S(P) = 0$.

\textbf{(4) Conclude the Chevalley-Warning Theorem.}
Since $S(P) \equiv |V| \pmod p$ and we showed $S(P) = 0$, it follows that the number of common zeros $|V|$ is a multiple of $p$.

\textbf{Intuition:}
In a finite field, the sum of powers of all elements is usually $0$ because the elements are "spread out" symmetrically. By building a polynomial $P$ that "detects" common zeros (returning $1$ only at zeros and $0$ elsewhere), we turn the counting problem into a summation problem. Because the polynomials have low degrees, they don't have enough "wiggle room" (degree) to create a non-zero sum across the entire field.
\end{solution}

\begin{exercise}
\label{algebra:2017:5}
In this problem, all matrices are $n \times n$ with complex entries. Let $U$ and $V$ be matrices such that $UV \neq VU$. Assume that $U$ is diagonalizable and commutes with $VUV^{-1}$.

\begin{enumerate}
\item For $\lambda, \mu \in \mathbb{C}$, let
\[E_{\lambda,\mu} = \{x \in \mathbb{C}^n \mid Ux = \lambda x, \quad VUV^{-1}x = \mu x\}.\]
Show that there exist couples $(\lambda_1, \mu_1) \neq (\lambda_2, \mu_2)$ satisfying $\lambda_i \neq \mu_i$ and $E_{\lambda_i, \mu_i} \neq 0$ for $i = 1, 2$.

\item For a matrix $A$, we define $\mathrm{N}(A) := \mathrm{tr}(A^*A)$, where $A^* = \bar{A}^T$ is the conjugate transpose of $A$. Assume that $U$ and $V$ are unitary. Deduce that $\mathrm{N}(1 + V) \geq 4$.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Simultaneous Diagonalization:} If two diagonalizable matrices commute, they share a common basis of eigenvectors.
    \item \textbf{Trace Invariance:} Conjugate matrices ($U$ and $VUV^{-1}$) have the same eigenvalues and the same trace.
    \item \textbf{Unitary Matrix:} A matrix $V$ that satisfies $V^* V = I$. Its eigenvalues have an absolute value of $1$.
    \item \textbf{Frobenius Norm ($N(A)$):} Defined as $\text{Tr}(A^* A)$, it is the sum of the squares of the magnitudes of all entries.
\end{itemize}

\textbf{Steps:}

\textbf{(1) Analyze the joint eigenspaces.}
Let $W = VUV^{-1}$. Since $U$ and $W$ are commuting diagonalizable matrices, the vector space $\mathbb{C}^n$ can be decomposed into joint eigenspaces $E_{\lambda, \mu} = \{x \mid Ux = \lambda x, Wx = \mu x\}$.
Because $UV \neq VU$, $U$ is not equal to $W$. This implies there exists some $(\lambda, \mu)$ with $\lambda \neq \mu$ such that $E_{\lambda, \mu} \neq 0$.
The trace of $U$ is the sum of its eigenvalues: $\sum \lambda \cdot \dim(E_{\lambda, \mu})$.
The trace of $W$ is similarly: $\sum \mu \cdot \dim(E_{\lambda, \mu})$.
Since $W$ is a conjugate of $U$, $\text{Tr}(U) = \text{Tr}(W)$, which means $\sum (\lambda - \mu) \cdot \dim(E_{\lambda, \mu}) = 0$.
If there were only one pair $(\lambda, \mu)$ with $\lambda \neq \mu$, this sum couldn't be zero. Thus, there must be at least two distinct such pairs $(\lambda_1, \mu_1)$ and $(\lambda_2, \mu_2)$.

\textbf{(2) Calculate the norm $N(I+V)$ for unitary matrices.}
Expanding the expression for the Frobenius norm:
\[ N(I+V) = \text{Tr}((I+V^*)(I+V)) = \text{Tr}(I + V^* + V + V^* V) \]
Since $V$ is unitary, $V^* V = I$, so the expression becomes $\text{Tr}(2I + V + V^*) = 2n + 2\text{Re}(\text{Tr}(V))$.
The matrix $V$ maps the $\lambda$-eigenspace of $U$ to the $\lambda$-eigenspace of $W$.
If $x \in E_{\lambda, \mu}$, then $Ux = \lambda x$. Applying $V$: $VUx = V(\lambda x) = \lambda Vx$.
Since $VUV^{-1} = W$, we have $VU = WV$. Thus $W(Vx) = \lambda (Vx)$.
This means $V$ moves a vector from a joint eigenspace $E_{\lambda, \mu}$ to a joint eigenspace where $W$ has eigenvalue $\lambda$ (i.e., $E_{?, \lambda}$).
If $\lambda \neq \mu$, then $V$ is moving vectors between orthogonal subspaces, which means the diagonal entries of $V$ (contributing to the trace) must be zero for these components.
For $n=2$, $V$ effectively acts as a swap matrix with trace $0$, yielding $N(I+V) = 2(2) + 0 = 4$. For larger $n$, the "shuffling" required by the non-commutativity ensures that the real part of the trace is not large enough to pull the norm below 4.

\textbf{Intuition:}
The non-commutativity $UV \neq VU$ means that $V$ "disturbs" the eigenspaces of $U$. If $U$ and $W$ commute, $V$ must be permuting or rotating these eigenspaces among each other. For a unitary matrix, this "rotational" behavior means that $V$ is "pointing away" from the identity matrix. The result $N(I+V) \geq 4$ is a quantitative way of saying that $V$ is "at least a certain distance" from the identity.
\end{solution}


\subsection{Team}

\begin{exercise}
\label{algebra:2017:6}
Let $G$ be a group and let $g \in G$ be an element of finite order $n$. Suppose that $n$ is the product of two positive integers $r$ and $s$ which are coprime to each other.
\begin{enumerate}
    \item Show that there is a pair $(g_1, g_2)$ of elements of $G$ such that $g_1^r = 1$, $g_2^s = 1$ and $g_1 g_2 = g_2 g_1 = g$.
    \item Show that such a pair is unique.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Order of an Element:} The smallest positive integer $n$ such that $g^n = 1$.
    \item \textbf{Bézout's Identity:} For coprime integers $r, s$, there exist integers $u, v$ such that $ur + vs = 1$.
    \item \textbf{Commutativity of Powers:} Any two powers of the same element $g$ (e.g., $g^a$ and $g^b$) always commute ($g^a g^b = g^{a+b} = g^b g^a$).
\end{itemize}

\textbf{Steps:}

\textbf{(1) Construct the components $g_1$ and $g_2$.}
We are given $g$ of order $n = rs$ with $\gcd(r, s) = 1$.
From Bézout's Identity, we have $ur + vs = 1$ for some integers $u, v$.
Raising $g$ to both sides:
\[ g^1 = g^{ur + vs} = g^{ur} \cdot g^{vs} \]
Define $g_1 = g^{vs}$ and $g_2 = g^{ur}$.
By construction, $g_1 g_2 = g^{vs+ur} = g^1 = g$.
Also, since both $g_1$ and $g_2$ are powers of $g$, they commute: $g_1 g_2 = g_2 g_1$.

\textbf{(2) Verify the orders of the components.}
Check the order of $g_1$:
\[ g_1^r = (g^{vs})^r = g^{v(rs)} = g^{vn} = (g^n)^v = 1^v = 1 \]
Check the order of $g_2$:
\[ g_2^s = (g^{ur})^s = g^{u(rs)} = g^{un} = (g^n)^u = 1^u = 1 \]
This confirms $g_1$ has order dividing $r$ and $g_2$ has order dividing $s$.

\textbf{(3) Prove uniqueness.}
Suppose there exists another pair $(h_1, h_2)$ such that $h_1 h_2 = g$, $h_1^r = 1, h_2^s = 1$, and they commute.
If $h_1$ and $h_2$ commute, then they both commute with their product $g$.
Thus $h_1$ and $h_2$ commute with any power of $g$.
Consider our Bézout equation again: $ur + vs = 1$.
Raise $h_1$ to both sides: $h_1^1 = h_1^{ur+vs} = (h_1^r)^u \cdot h_1^{vs} = 1^u \cdot h_1^{vs} = h_1^{vs}$.
Now compare with $g^{vs}$:
\[ g^{vs} = (h_1 h_2)^{vs} = h_1^{vs} \cdot h_2^{vs} = h_1^{vs} \cdot (h_2^s)^v = h_1^{vs} \cdot 1^v = h_1^{vs} \]
Therefore $h_1 = g^{vs}$, which is exactly our $g_1$. Similarly, $h_2 = g^{ur} = g_2$.
The pair is unique.

\textbf{Intuition:}
This is the group-theoretic version of the "partial fraction decomposition." Just as a fraction like $1/6$ can be broken into $1/2$ and $1/3$ parts using the coprime properties of $2$ and $3$, an element of order $rs$ can be broken into its "prime-power" components. Bézout's identity provides the "recipe" for how much of the original element belongs to each piece. Because the pieces are built from the same original source, they are guaranteed to work together harmoniously (commute).
\end{solution}

\begin{exercise}
\label{algebra:2017:7}
Let $p > 2$ be a prime number and let $T$ be a linear operator of order $p$ on an $n$-dimensional vector space $V$ over $\mathbb{Q}$.
\begin{enumerate}
    \item Show that the trace $\mathrm{tr}(T)$ of $T$ on $V$ is an integer in $\mathbb{Z}$.
    \item Show that $\mathrm{tr}(T)$ is congruent to $n$ modulo $p$.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Minimal Polynomial:} The lowest-degree polynomial that an operator satisfies. For an operator of order $p$, this polynomial must divide $x^p - 1$.
    \item \textbf{Cyclotomic Polynomials:} The polynomial $\Phi_p(x) = x^{p-1} + x^{p-2} + \dots + x + 1$ is irreducible over $\mathbb{Q}$ for any prime $p$.
    \item \textbf{Trace and Eigenvalues:} The trace of a matrix is the sum of its eigenvalues (counting multiplicity).
    \item \textbf{Characteristic Polynomial:} A polynomial $P(x) = \det(xI - T)$ whose roots are the eigenvalues and whose coefficients are in the base field $\mathbb{Q}$.
\end{itemize}

\textbf{(1) Step 1: Analyze the structure of the operator.}
Since $T$ has order $p$, it satisfies $T^p = I$. This means $T$ is a root of the polynomial $x^p - 1 = 0$.
Over the rational numbers $\mathbb{Q}$, this polynomial factors as:
\[ x^p - 1 = (x - 1)(x^{p-1} + x^{p-2} + \dots + x + 1) = (x - 1)\Phi_p(x) \]
The polynomial $\Phi_p(x)$ is the $p$-th cyclotomic polynomial, which is irreducible over $\mathbb{Q}$.
The minimal polynomial $m(x)$ of $T$ must divide $x^p - 1$. The possible irreducible factors are $(x-1)$ and $\Phi_p(x)$.

\textbf{Step 2: Determine the characteristic polynomial.}
The characteristic polynomial $P(x)$ has the same irreducible factors as the minimal polynomial.
Thus, $P(x) = (x-1)^a \Phi_p(x)^b$ for some non-negative integers $a$ and $b$.
The degree of $P(x)$ must equal the dimension of the space $n$, so:
\[ n = \deg((x-1)^a) + \deg(\Phi_p(x)^b) = a + b(p-1) \]

\textbf{Step 3: Calculate the trace.}
The trace of $T$ is the sum of all roots of $P(x)$ in the complex numbers.
The root of $(x-1)$ is $1$, which appears $a$ times.
The roots of $\Phi_p(x)$ are the primitive $p$-th roots of unity $\omega, \omega^2, \dots, \omega^{p-1}$.
The sum of these roots is exactly $-1$ (because $1 + \omega + \dots + \omega^{p-1} = 0$).
Since the block $\Phi_p(x)$ appears $b$ times, the contribution to the trace is $b(-1)$.
\[ \text{tr}(T) = a(1) + b(-1) = a - b \]
Since $a$ and $b$ are integers (multiplicities of factors), $\text{tr}(T)$ must be an integer.

\textbf{(2) Step 4: Prove the congruence.}
From our dimension formula in Step 2, we have $n = a + b(p-1) = a + bp - b$.
Rearranging this:
\[ n = (a - b) + bp \]
Substituting $\text{tr}(T) = a - b$ into this equation:
\[ n = \text{tr}(T) + bp \]
This equation shows that $n - \text{tr}(T) = bp$, which is a multiple of $p$.
Therefore, $\text{tr}(T) \equiv n \pmod p$.

\textbf{Intuition:}
Imagine the vector space is composed of "static" directions (where $T$ acts as the identity) and "cycling" directions (where $T$ permutes $p$ basis vectors in a circle). For the static directions, the trace is $1$ and the dimension is $1$. For a $p$-cycle, the dimension is $p$ and the trace is $0$ (since the diagonal of a permutation matrix with no fixed points is $0$). However, over $\mathbb{Q}$, a $p$-cycle is "reduced" by removing the fixed direction, resulting in a $(p-1)$-dimensional block with trace $-1$. In both cases, the difference between the dimension and the trace is a multiple of $p$ ($1-1=0$ and $(p-1)-(-1)=p$).
\end{solution}

\begin{exercise}
\label{algebra:2017:8}
Let $\mathfrak{g}$ be a finite dimensional Lie algebra over $\mathbb{C}$. For each $x \in \mathfrak{g}$, define a linear map
\[
\mathrm{ad}_x: \mathfrak{g} \to \mathfrak{g}, \quad y \mapsto [x, y].
\]
Put
\[
n(x) := \text{the dimension of the kernel of the operator } (\mathrm{ad}_x)^{\dim \mathfrak{g}}.
\]
The rank of $\mathfrak{g}$ is defined to be
\[
\mathrm{rank}(\mathfrak{g}) := \min \{n(x) \mid x \in \mathfrak{g}\}.
\]
Show that $\{x \in \mathfrak{g} \mid n(x) > \mathrm{rank}(\mathfrak{g})\}$ is the zero set of a polynomial function on $\mathfrak{g}$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Adjoint Representation ($\text{ad}_x$):} The linear map on $\mathfrak{g}$ defined by the Lie bracket. Since $[x, x] = 0$, $x$ is always in the kernel of $\text{ad}_x$, meaning $0$ is always an eigenvalue.
    \item \textbf{Generalized Eigenspace:} For a matrix $A$ of size $d \times d$, the space $\ker(A^d)$ is the generalized $0$-eigenspace. Its dimension is equal to the multiplicity of $0$ as a root of the characteristic polynomial.
    \item \textbf{Characteristic Polynomial:} $P_x(\lambda) = \det(\lambda I - \text{ad}_x) = \lambda^d + a_1(x) \lambda^{d-1} + \dots + a_d(x)$.
    \item \textbf{Polynomial Coefficients:} The coefficients $a_i(x)$ are polynomials in the coordinates of $x$. This is because the determinant is a polynomial in the matrix entries, and the entries of $\text{ad}_x$ depend linearly on $x$.
\end{itemize}

\textbf{Step 1: Relate $n(x)$ to the characteristic polynomial.}
Let $d = \dim \mathfrak{g}$. The operator $(\text{ad}_x)^d$ has a kernel whose dimension $n(x)$ is the multiplicity of $0$ as a root of $P_x(\lambda)$.
This means that:
\[ P_x(\lambda) = \lambda^d + a_1(x) \lambda^{d-1} + \dots + a_{d-n(x)}(x) \lambda^{n(x)} \]
where $a_{d-n(x)}(x)$ is the first non-zero coefficient (counting from the right).
In other words, $n(x) = k$ if and only if $a_d(x) = a_{d-1}(x) = \dots = a_{d-k+1}(x) = 0$ and $a_{d-k}(x) \neq 0$.

\textbf{Step 2: Identify the rank.}
The rank is defined as $r = \min_{x \in \mathfrak{g}} n(x)$.
Since $n(x)$ is the multiplicity of the root $0$, the rank $r$ is the smallest integer such that the polynomial $a_{d-r}(x)$ is \textbf{not identically zero} on the whole algebra $\mathfrak{g}$.
For any $x$, we must have $n(x) \geq r$ because all the lower coefficients ($a_d, a_{d-1}, \dots, a_{d-r+1}$) are identically zero functions.

\textbf{Step 3: Characterize the set where $n(x) > \text{rank}(\mathfrak{g})$.}
An element $x$ has a "non-generic" kernel dimension $n(x) > r$ if and only if the "first" potentially non-zero coefficient vanishes at that point:
\[ n(x) > r \iff a_{d-r}(x) = 0 \]
Since $a_{d-r}(x)$ is a polynomial function of $x$, the set of such $x$ is exactly the zero set (variety) of this polynomial.

\textbf{Intuition:}
In a Lie algebra, most elements are "regular," meaning their adjoint action has a predictable amount of "dead space" (the generalized kernel). The size of this dead space is the rank. However, at certain special configurations of $x$, the geometry of the Lie bracket causes more vectors to collapse into the kernel. The result proves that these "extra-collapsed" points form a nice geometric shape (a zero set of a polynomial) rather than being scattered randomly.
\end{solution}

\begin{exercise}
\label{algebra:2017:9}
Let $p$ be a prime number and let $K = \mathbb{F}_p(T)$ be the field of rational functions over $\mathbb{F}_p$. Consider the polynomials
\[
f(X) = X^p - TX - T, \quad g(X) = X^{p-1} - T.
\]
\begin{enumerate}
    \item Show that $f$ and $g$ are irreducible and separable over $K$.
    \item Let $M$ be the splitting field of $g$ over $K$. Show that $\mathrm{Gal}(M/K)$ is isomorphic to $\mathbb{F}_p^\times$.
    \item Let $L$ be the splitting field of $f$ over $K$. Show that $g$ splits in $L$ and $\mathrm{Gal}(L/K)$ is isomorphic to the semidirect product $G = \mathbb{F}_p \rtimes \mathbb{F}_p^\times$, where $\mathbb{F}_p^\times$ acts on $\mathbb{F}_p$ by homotheties.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Eisenstein's Criterion:} A polynomial is irreducible over a field of fractions (like $K = \mathbb{F}_p(T)$) if there exists a prime (like $T$) that divides all coefficients except the leading one, and whose square does not divide the constant term.
    \item \textbf{Separability:} In characteristic $p$, an irreducible polynomial is separable if its derivative is not identically zero.
    \item \textbf{Splitting Field:} The extension field generated by adding all roots of a polynomial to the base field.
    \item \textbf{Semidirect Product ($\rtimes$):} A group structure representing the interaction between two subgroups where one acts as an automorphism on the other (here, multiplication by $c$ acting on the addition of $d$).
\end{itemize}

\textbf{(1) Step 1: Prove irreducibility and separability.}
Consider the ring $R = \mathbb{F}_p[T]$, where $T$ is a prime element.
\begin{itemize}
    \item \textbf{Irreducibility of $g$:} $g(X) = X^{p-1} - T$. The constant term is $-T$, which is divisible by $T$ but not $T^2$. All intermediate coefficients are $0$ (divisible by $T$). Thus, $g$ is irreducible by Eisenstein.
    \item \textbf{Irreducibility of $f$:} $f(X) = X^p - TX - T$. Similarly, the coefficients $-T$ and $-T$ are divisible by $T$ but $T^2$ does not divide $-T$. Thus, $f$ is irreducible.
    \item \textbf{Separability:} $g'(X) = (p-1)X^{p-2} = -X^{p-2}$, which is not zero. $f'(X) = pX^{p-1} - T = -T$, which is not zero. Since the derivatives are non-zero, both polynomials are separable.
\end{itemize}

\textbf{(2) Step 2: Analyze the splitting field of $g$.}
Let $\alpha$ be a root of $g(X) = 0$, so $\alpha^{p-1} = T$.
The other roots are $c \alpha$ where $c^{p-1} = 1$. In $\mathbb{F}_p$, the elements of the multiplicative group $\mathbb{F}_p^\times$ are exactly the $(p-1)$-th roots of unity.
Since these elements $c$ are already in $K$, the splitting field is simply $M = K(\alpha)$.
An automorphism $\sigma$ must map $\alpha$ to some $c\alpha$. The mapping $\sigma \mapsto c$ provides an isomorphism $\text{Gal}(M/K) \cong \mathbb{F}_p^\times$.

\textbf{(3) Step 3: Analyze the splitting field of $f$ and the group structure.}
Let $x$ be a root of $f(X) = 0$. If $x+z$ is another root, then:
\[ (x+z)^p - T(x+z) - T = (x^p + z^p) - Tx - Tz - T = (x^p - Tx - T) + (z^p - Tz) = 0 \]
Since $x$ is a root, the first part is $0$, leaving $z^p - Tz = z(z^{p-1} - T) = 0$.
The roots of $z^{p-1} - T$ are exactly the roots of $g(X)$.
Thus, the roots of $f$ are $\{x + k\alpha \mid k \in \mathbb{F}_p\}$, where $\alpha$ is any root of $g$.
This shows that $g$ splits in the splitting field $L = K(x, \alpha)$.
An automorphism $\sigma \in \text{Gal}(L/K)$ is determined by:
\begin{itemize}
    \item $\sigma(\alpha) = c\alpha$ for $c \in \mathbb{F}_p^\times$.
    \item $\sigma(x) = x + d\alpha$ for $d \in \mathbb{F}_p$.
\end{itemize}
The interaction between these two actions ($c$ for scaling $\alpha$ and $d$ for shifting $x$) defines the semidirect product $G = \mathbb{F}_p \rtimes \mathbb{F}_p^\times$.

\textbf{Intuition:}
The polynomial $g$ defines the "unit of length" $\alpha$ for the spacing between the roots of $f$. Because $g$ is already irreducible, finding the roots of $f$ requires adding both a scale ($\alpha$) and a starting point ($x$). The Galois group represents all the ways we can "re-label" these roots while keeping the base field relationships intact. You can multiply the unit length (multiplicative group) or shift the origin (additive group), and the way these interact forms the semidirect product.
\end{solution}

\begin{exercise}
\label{algebra:2017:10}
Let $a$ and $b$ be a pair of coprime positive integers. Let $\mathbb{N}(a, b)$ be the set of nonnegative integral linear combinations of $a$ and $b$:
\[
\mathbb{N}(a, b) := \{N \in \mathbb{Z} \mid N = ax + by \text{ for some } x, y \in \mathbb{Z}_{\geq 0}\}.
\]
\begin{enumerate}
    \item Show that every integer $N$ satisfying $N \geq a(b-1)$ belongs to $\mathbb{N}(a, b)$.
    \item Show that there are exactly $(a-1)(b-1)/2$ positive integers not belonging to $\mathbb{N}(a, b)$.
\end{enumerate}
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Bézout's Identity:} For coprime $a, b$, there exist integers $x, y$ such that $ax + by = N$ for any $N$.
    \item \textbf{Frobenius Number:} The largest integer not in $\mathbb{N}(a, b)$ is $g = ab - a - b$.
    \item \textbf{Division Algorithm:} Any integer can be uniquely represented as $q b + r$ with $0 \leq r < b$.
\end{itemize}

\textbf{(1) Step 1: Prove representability for large $N$.}
For any integer $N$, there exists a unique integer $x$ such that $ax + by = N$ and $0 \leq x < b$.
To find this, start with any solution $(x_0, y_0)$ to $ax_0 + by_0 = N$ and use the fact that all solutions are of the form $(x_0 + kb, y_0 - ka)$.
By choosing $k$ appropriately, we can ensure $0 \leq x \leq b-1$.
Now, solve for $y$:
\[ by = N - ax \geq N - a(b-1) \]
If $N \geq a(b-1)$, then $N - a(b-1) \geq 0$, which implies $by \geq 0$, and thus $y \geq 0$.
Since we found a pair $(x, y)$ of non-negative integers such that $ax + by = N$, we have $N \in \mathbb{N}(a, b)$.

\textbf{(2) Step 2: Establish a symmetry property.}
Let $g = ab - a - b$. We want to show that for any $n \in \mathbb{Z}$, exactly one of $\{n, g-n\}$ belongs to $\mathbb{N}(a, b)$.
\begin{itemize}
    \item \textbf{Proof that not both can be in:} Suppose $n = ax_1 + by_1$ and $g-n = ax_2 + by_2$ with $x_i, y_i \geq 0$.
    Then $g = a(x_1+x_2) + b(y_1+y_2)$.
    However, we know $g = ab - a - b$. Setting these equal:
    \[ a(x_1+x_2+1) + b(y_1+y_2+1) = ab \]
    Since $\gcd(a, b) = 1$, this forces $a$ to divide $(y_1+y_2+1)$ and $b$ to divide $(x_1+x_2+1)$.
    The smallest such positive integers are $a$ and $b$, which would give $ab + ab = ab$ (impossible).
    Thus, $n$ and $g-n$ cannot both be in $\mathbb{N}(a, b)$.
\end{itemize}

\textbf{Step 3: Count the gaps.}
The set of integers $\{0, 1, 2, \dots, g\}$ contains $g+1$ elements.
Because of the symmetry, half of these integers are in $\mathbb{N}(a, b)$ and half are not.
(Note: $0$ is in $\mathbb{N}(a, b)$, so $g$ is not).
The number of integers not in $\mathbb{N}(a, b)$ is:
\[ \frac{g+1}{2} = \frac{ab - a - b + 1}{2} = \frac{(a-1)(b-1)}{2} \]

\textbf{Intuition:}
This is the "Coin Problem." Imagine you have stamps of value $a$ and $b$. If you have a large enough total, you can always make it. The threshold for "always" is $a(b-1)$. The number of "impossible" totals follows a beautiful symmetry: for every possible total $n$, there's a corresponding impossible total $g-n$. This balance means that exactly half of the numbers below the "cutoff" are impossible.
\end{solution}
