\section{2014 Algebra \& Number Theory}

\subsection*{Individual}
\begin{exercise}
\l\label{2014:1}

\label{algebra:2014:1}Let $G$ be a finite subgroup of $GL(V)$ where $V \cong \mathbb{C}^n$.
(a) $H = \{ h \in G : h|_V = \eta(h) I \}$ is normal and $H \cong \text{im}(\eta)$.
(b) $|\chi_V(g)| \leq n$, with equality iff $g \in H$.
(c) Irrep $W$ is a direct summand of $V^{\otimes m}$ for some $m$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Finite Group Representation:} A way to view group elements as matrices acting on a vector space.
    \item \textbf{Character ($\chi_V$):} The trace of the matrices in the representation. $\chi_V(g) = \text{Tr}(g)$.
    \item \textbf{Root of Unity:} Any complex number $\lambda$ such that $\lambda^k = 1$. Eigenvalues of finite group elements are always roots of unity.
    \item \textbf{Schur's Lemma / Burnside's Theorem:} Fundamental results relating irreducible representations to the algebra of endomorphisms.
    \item \textbf{Tensor Product ($V^{\otimes m}$):} A way to build larger representations from smaller ones.
\end{itemize}

\textbf{(a) Step 1: Prove $H$ is a normal subgroup.}
The set $H$ consists of elements $h \in G$ that act as "scalar matrices" on $V$. That is, for each $h \in H$, there is a complex number $\eta(h)$ such that the matrix representing $h$ is $\eta(h) I$, where $I$ is the identity matrix.

To show $H$ is normal ($H \triangleleft G$), we take any $h \in H$ and any $g \in G$ and check the conjugation $ghg^{-1}$:
\[ g h g^{-1} = g (\eta(h) I) g^{-1} = \eta(h) g I g^{-1} = \eta(h) g g^{-1} = \eta(h) I. \]
Since the result is still a scalar matrix, $ghg^{-1} \in H$. This proves $H$ is normal.

\textbf{Step 2: Show $H \cong \text{im}(\eta)$.}
The map $\eta: H \to \mathbb{C}^\times$ takes an element $h$ to its scaling factor. This is a group homomorphism because $(h_1 h_2)$ acts as $(\eta(h_1)I)(\eta(h_2)I) = \eta(h_1)\eta(h_2)I$, so $\eta(h_1 h_2) = \eta(h_1)\eta(h_2)$.
If $\eta(h) = 1$, then $h = 1 \cdot I = I$, which is the identity element of $G$. Thus, the kernel of $\eta$ is trivial, making $\eta$ an isomorphism between $H$ and its image in $\mathbb{C}^\times$.

\textbf{(b) Step 3: Bound the character using eigenvalues.}
The character $\chi_V(g)$ is the trace of the matrix representing $g$. Let $\lambda_1, \lambda_2, \dots, \lambda_n$ be the eigenvalues of $g$.
Since $G$ is a finite group, $g^k = I$ for some integer $k$. This means every eigenvalue $\lambda_j$ must satisfy $\lambda_j^k = 1$, so each $\lambda_j$ is a root of unity. 
The absolute value of any root of unity is $| \lambda_j | = 1$.

Using the triangle inequality for complex numbers:
\[ |\chi_V(g)| = | \lambda_1 + \lambda_2 + \dots + \lambda_n | \leq | \lambda_1 | + | \lambda_2 | + \dots + | \lambda_n | = 1 + 1 + \dots + 1 = n. \]
This proves $|\chi_V(g)| \leq n$.

\textbf{Step 4: Determine when equality holds.}
Equality in the triangle inequality $|z_1 + \dots + z_n| = \sum |z_j|$ occurs if and only if all the complex numbers $z_j$ point in the same direction (lie on the same ray from the origin).
Since all our $\lambda_j$ have length 1, they must all be the \textit{exact same} complex number: $\lambda_1 = \lambda_2 = \dots = \lambda_n = \lambda$.
Because finite group matrices are always diagonalizable, a matrix with all eigenvalues equal to $\lambda$ must be the scalar matrix $\lambda I$.
This means $g$ acts as a scalar, so $g \in H$ by definition.

\textbf{(c) Step 5: Decompose the tensor algebra.}
This part is a beautiful application of Burnside's Theorem. We are given that $G \subset GL(V)$, which means $V$ is a \textit{faithful} representation (the only element acting as identity is the identity element).
A theorem by Burnside states that for a faithful representation $V$, every possible irreducible representation (irrep) $W$ of the group $G$ will eventually show up if we keep taking tensor products of $V$ with itself.
Intuitively, as we take $V \otimes V \otimes \dots \otimes V$, we are building up all possible polynomial combinations of the coordinates. Since the space of all functions on a finite group is finite-dimensional, and polynomials are dense, we can generate any "shape" (irrep) we want.
Thus, any irrep $W$ is a direct summand of $V^{\otimes m}$ for some sufficiently large $m$.
\end{solution}

\begin{exercise}
\l\label{2014:2}

\label{algebra:2014:2}(a) $A = (t^{a_i+a_j})$ is PSD. Rank. (b) $B = (1/(1+a_i+a_j))$ is PSD. (c) $B$ is PD iff $a_i$ distinct.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Positive Semi-Definite (PSD):} A symmetric matrix $M$ is PSD if for every vector $v$, $v^T M v \geq 0$. This means all its eigenvalues are non-negative.
    \item \textbf{Gram Matrix:} A matrix whose entries are inner products of vectors. Gram matrices are always PSD.
    \item \textbf{Rank:} The number of linearly independent rows or columns.
    \item \textbf{Analytic Functions:} Functions that can be represented by a power series. If an analytic function is zero on an interval, it is zero everywhere.
\end{itemize}

\textbf{(a) Step 1: Prove $A$ is PSD and find its rank.}
Let's look at the entries of the matrix $A$: $A_{ij} = t^{a_i + a_j} = t^{a_i} \cdot t^{a_j}$.
We can define a vector $\mathbf{x} = (t^{a_1}, t^{a_2}, \dots, t^{a_n})^T$. Then the matrix $A$ can be written as the outer product of this vector with itself:
\[ A = \mathbf{x} \mathbf{x}^T \]
For any vector $v$, we calculate the quadratic form:
\[ v^T A v = v^T (\mathbf{x} \mathbf{x}^T) v = (v^T \mathbf{x})(\mathbf{x}^T v) = (v \cdot \mathbf{x})^2 \]
Since the square of any real number is non-negative, $v^T A v \geq 0$. This proves $A$ is \textbf{Positive Semi-Definite}.

The rank of an outer product $uv^T$ is always 1 (as long as the vectors are non-zero). Since $t > 0$, the vector $\mathbf{x}$ is non-zero, so the \textbf{rank of $A$ is 1}.

\textbf{(b) Step 2: Use an integral trick for matrix $B$.}
We are given $B_{ij} = \frac{1}{1 + a_i + a_j}$. To show this is PSD, we use a classic trick from calculus: the identity $\frac{1}{S} = \int_0^\infty e^{-Sx} dx$.
Plugging in $S = 1 + a_i + a_j$, we get:
\[ B_{ij} = \int_0^\infty e^{-(1 + a_i + a_j)x} dx = \int_0^\infty e^{-x} \cdot e^{-a_ix} \cdot e^{-a_jx} dx \]
Now, let's check the quadratic form $v^T B v$:
\[ v^T B v = \sum_{i,j} v_i B_{ij} v_j = \sum_{i,j} v_i \left( \int_0^\infty e^{-x} e^{-a_ix} e^{-a_jx} dx \right) v_j \]
By pulling the sum inside the integral:
\[ = \int_0^\infty e^{-x} \left( \sum_{i} v_i e^{-a_ix} \right) \left( \sum_{j} v_j e^{-a_jx} \right) dx = \int_0^\infty e^{-x} \left( \sum_{i} v_i e^{-a_ix} \right)^2 dx \]
Since the integrand is a square times a positive exponential, the integral is $\geq 0$. This proves $B$ is \textbf{Positive Semi-Definite}.

\textbf{(c) Step 3: Determine when $B$ is strictly Positive Definite.}
$B$ is Positive Definite (PD) if $v^T B v = 0$ implies $v = 0$.
From our integral, $v^T B v = 0$ happens if and only if the function $f(x) = \sum_{i=1}^n v_i e^{-a_ix}$ is zero for almost all $x \in (0, \infty)$.
Because $f(x)$ is a sum of exponentials, it is an \textit{analytic} function. If an analytic function is zero on an interval, it must be the zero function ($f(x) \equiv 0$).

A set of functions $\{ e^{-a_1x}, e^{-a_2x}, \dots, e^{-a_nx} \}$ is \textbf{linearly independent} if and only if all the exponents $a_i$ are distinct.
\begin{itemize}
    \item If the $a_i$ are distinct, then the only way to get $\sum v_i e^{-a_ix} = 0$ is if all $v_i = 0$. In this case, $B$ is \textbf{Positive Definite}.
    \item If two $a_i$ are the same, the rows of $B$ will be identical, making the matrix singular (rank $< n$). In this case, it is only semi-definite.
\end{itemize}
Thus, $B$ is PD if and only if the $a_i$ are distinct.
\end{solution}

\begin{exercise}
\l\label{2014:3}

\label{algebra:2014:3}$X^2 - 82Y^2 = \pm 2$.
(a) $(x, y) \to (9x-82y, x-9y)$. (b) Roots mod $p^n$. (c) No integer roots.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Norm in Quadratic Fields:} For $d > 0$, the norm of an element $\alpha = x + y\sqrt{d}$ in $\mathbb{Q}(\sqrt{d})$ is $N(\alpha) = x^2 - dy^2$.
    \item \textbf{Units:} Elements with norm $\pm 1$. Multiplying a solution by a unit gives another solution.
    \item \textbf{Hilbert Symbol:} A tool used in local-global principles (Hasse Principle) to check if an equation has roots modulo $p^n$ for all $p$.
    \item \textbf{Continued Fractions:} A method to find the "best" rational approximations to a square root. The denominators in this process are the only possible small values for the norm $x^2 - dy^2$.
\end{itemize}

\textbf{(a) Step 1: Explain the map using field norms.}
The equation is $x^2 - 82y^2 = \pm 2$. This is exactly the condition that the norm of the element $\alpha = x + y\sqrt{82}$ is $N(\alpha) = \pm 2$.
The map given in the problem is:
\[ (x, y) \mapsto (9x - 82y, x - 9y) \]
This map corresponds to multiplying our element $\alpha = x + y\sqrt{82}$ by the constant $\epsilon = 9 - \sqrt{82}$.
Let's check the norm of $\epsilon$: $N(\epsilon) = 9^2 - 82(1)^2 = 81 - 82 = -1$.
A fundamental property of the norm is that it is multiplicative: $N(\alpha \beta) = N(\alpha) N(\beta)$.
If we take a solution $(x, y)$ with norm $\pm 2$ and multiply it by $\epsilon$, the new norm will be:
\[ N(\text{new}) = N(\alpha) \cdot N(\epsilon) = (\pm 2) \cdot (-1) = \mp 2. \]
So, the map flips the sign of the constant on the right-hand side. This is why it maps solutions of one equation to the other.

\textbf{(b) Step 2: Show local solubility (mod $p^n$).}
The Hasse Principle says that if an equation has solutions modulo every prime power $p^n$, it \textit{might} have a rational solution.
For $p$ odd, we need to check if $x^2 - 82y^2 \equiv 2 \pmod p$ has a solution.
Since $82 = 2 \cdot 41$, the equation involves two primes that are both "ramified" or "related" to 2.
Using the properties of the Legendre symbol and Hilbert symbols, it turns out that for any prime $p$ (including $p=2$ and $p=\infty$), the value $2$ is always represented by the quadratic form $x^2 - 82y^2$ in the $p$-adic numbers $\mathbb{Q}_p$.
This means there are no "local" obstructions—the equation looks solvable if you only look at it through the lens of modular arithmetic.

\textbf{(c) Step 3: Prove there are no global integer roots.}
Even if solutions exist modulo every $p^n$, there might not be any integers $x, y$ that work. This is the case here.
We use the theory of \textbf{continued fractions}. For a Pell-like equation $x^2 - dy^2 = k$, if $|k| < \sqrt{d}$, then the fraction $x/y$ must be one of the "convergents" produced by the continued fraction algorithm for $\sqrt{d}$.
Let's calculate the continued fraction for $\sqrt{82}$:
Since $9^2 = 81 < 82 < 10^2$, the first term is $9$.
$\sqrt{82} = 9 + (\sqrt{82}-9) = 9 + \frac{1}{\sqrt{82}+9} = 9 + \frac{1}{18 + (\sqrt{82}-9)}$.
The pattern repeats: $\sqrt{82} = [9, \overline{18}]$.
The "denominators" $Q_i$ produced by this algorithm are the only integers $k$ with $|k| < \sqrt{82} \approx 9.05$ for which $x^2 - 82y^2 = k$ can have a solution.
For the sequence $[9, 18, 18, \dots]$, the $Q_i$ values are:
$Q_0 = 1$
$Q_1 = 1$ (because $82 - 9^2 = 1$)
The values just cycle through $\{1\}$. This means the only possible values for the norm with magnitude less than $9$ are $\pm 1$.
Since $2$ is not $1$, the equation $x^2 - 82y^2 = 2$ has \textbf{no integer solutions}.
\end{solution}

\begin{exercise}
\l\label{2014:4}

\label{algebra:2014:4}$G = S \times T$ non-abelian simple.
(a) 4 normal subgroups. (b) $S \cong T \implies \exists$ maximal not containing factors. (c) Converse.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Simple Group:} A group that has no normal subgroups other than $\{1\}$ and itself.
    \item \textbf{Non-abelian:} A group where $gh \neq hg$ in general. Simple non-abelian groups have a trivial center: $Z(G) = \{1\}$.
    \item \textbf{Goursat's Lemma:} A theorem that describes the subgroups of a product $S \times T$. It relates them to isomorphisms between sub-quotients.
    \item \textbf{Diagonal Subgroup:} For $S \cong T$, the set of pairs $\{(s, \phi(s)) : s \in S\}$ where $\phi$ is an isomorphism.
\end{itemize}

\textbf{(a) Step 1: Identify all normal subgroups of $S \times T$.}
Let $N$ be a normal subgroup of $G = S \times T$. Consider the projection maps $p_S: G \to S$ and $p_T: G \to T$.
The image $p_S(N)$ is a normal subgroup of $S$. Since $S$ is simple, $p_S(N)$ must be either $\{1\}$ or $S$.
The intersection $N \cap (S \times \{1\})$ is also normal in $S \times \{1\}$. Again, by simplicity, this intersection must be either $\{1\} \times \{1\}$ or $S \times \{1\}$.

There are four possible "factor-wise" combinations:
\begin{enumerate}
    \item $N$ contains both factors: $N = S \times T$.
    \item $N$ contains only the first factor: $N = S \times \{1\}$.
    \item $N$ contains only the second factor: $N = \{1\} \times T$.
    \item $N$ contains neither factor but has full projections: This would mean $N$ is a "diagonal" subgroup where $N \cong S \cong T$. However, for a diagonal subgroup to be \textit{normal}, the isomorphism must commute with every inner automorphism. This only happens if the group is abelian (which it isn't) or if the isomorphism is very special. In our non-abelian simple case, such a subgroup is never normal.
\end{enumerate}
Thus, the only normal subgroups are the 4 obvious ones: $\{1\}, S \times \{1\}, \{1\} \times T, S \times T$.

\textbf{(b) Step 2: Show $S \cong T$ implies the existence of a special maximal subgroup.}
Suppose $S \cong T$ and let $\phi: S \to T$ be an isomorphism.
Define the \textbf{diagonal subgroup}:
\[ \Delta = \{ (s, \phi(s)) : s \in S \} \]
Let's check if $\Delta$ contains $S \times \{1\}$. If $(s, 1) \in \Delta$, then $\phi(s) = 1$. Since $\phi$ is an isomorphism, this means $s=1$. So $\Delta$ only shares the identity with the factors.
$\Delta$ is a maximal subgroup because its index is $|S|$, and any larger subgroup would have to contain at least some elements from the factors, which would quickly force it to be the whole group $G$ (due to simplicity).

\textbf{(c) Step 3: Prove the converse.}
Suppose there exists a maximal subgroup $M < S \times T$ that does not contain either $S \times \{1\}$ or $\{1\} \times T$.
By the same logic as in Step 1, the intersection of $M$ with a factor must be trivial: $M \cap (S \times \{1\}) = \{1\}$.
Also, the projection $p_S(M)$ must be $S$ (if it were $\{1\}$, then $M \subseteq \{1\} \times T$, which would make it one of the factors).
According to Goursat's Lemma, a subgroup $M$ with full projections and trivial intersections is precisely the \textit{graph of an isomorphism} between the two groups.
Since $M$ defines a one-to-one mapping between $S$ and $T$ that preserves the group operation, it follows that $S \cong T$.
\end{solution}

\begin{exercise}
\l\label{2014:5}

\label{algebra:2014:5}Chevalley-Warning Theorem.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Finite Field ($\mathbb{F}_q$):} A field with $q = p^k$ elements. The characteristic is $p$.
    \item \textbf{Fermat's Little Theorem (for fields):} For any non-zero element $x \in \mathbb{F}_q$, we have $x^{q-1} = 1$. This means the expression $(1 - f(X)^{q-1})$ is $1$ if $f(X)=0$ and $0$ otherwise.
    \item \textbf{Monomial Summation:} Over a finite field, the sum $\sum_{x \in \mathbb{F}_q} x^a$ is non-zero only if $(q-1)$ divides $a$ (and $a > 0$). If $0 \leq a < q-1$, the sum is $0$.
    \item \textbf{Degree Argument:} If the total degree of a polynomial is less than the number of variables times $(q-1)$, its sum over all points must be zero.
\end{itemize}

\textbf{Step 1: Construct an indicator polynomial.}
We are looking for common zeros of several polynomials $f_1, f_2, \dots, f_r$ in $n$ variables.
Let's define a special polynomial $P(X)$ that "detects" these common zeros:
\[ P(X) = \prod_{i=1}^r \left(1 - f_i(X)^{q-1}\right) \]
Wait, how does this work?
\begin{itemize}
    \item If $X$ is a common zero (meaning $f_i(X) = 0$ for all $i$), then each term $(1-0) = 1$. The product is $1 \cdot 1 \dots 1 = 1$.
    \item If $X$ is \textit{not} a common zero (meaning $f_i(X) \neq 0$ for at least one $i$), then for that specific $i$, $f_i(X)^{q-1} = 1$. So $(1-1) = 0$. The whole product becomes $0$.
\end{itemize}
So $P(X)$ is exactly $1$ on common zeros and $0$ everywhere else.

\textbf{Step 2: Relate the number of roots to the sum of $P(X)$.}
Let $N$ be the number of common roots. If we add up the value of $P(X)$ for every possible point $X$ in the space $\mathbb{F}_q^n$:
\[ \sum_{X \in \mathbb{F}_q^n} P(X) = \underbrace{1 + 1 + \dots + 1}_{N \text{ times}} + \underbrace{0 + 0 + \dots + 0}_{\text{rest}} = N \]
Note that this sum is calculated inside the field $\mathbb{F}_q$, so we really mean $N \pmod p$.

\textbf{Step 3: Analyze the degree of $P(X)$.}
The degree of each $f_i$ is $d_i$. The degree of $(1 - f_i^{q-1})$ is $d_i(q-1)$.
The total degree of the product $P(X)$ is:
\[ \deg(P) = \sum_{i=1}^r d_i(q-1) = (q-1) \sum_{i=1}^r d_i \]
We are given the condition that $n > \sum d_i$. Multiplying by $(q-1)$ gives:
\[ (q-1)n > (q-1) \sum d_i = \deg(P) \]
This means the degree of $P$ is strictly \textit{less} than the number of variables times $(q-1)$.

\textbf{Step 4: Prove the sum is zero.}
Any polynomial is a sum of monomials like $X_1^{a_1} X_2^{a_2} \dots X_n^{a_n}$.
When we sum a monomial over all points $X \in \mathbb{F}_q^n$, we can split the sum variable by variable:
\[ \sum_{X \in \mathbb{F}_q^n} X_1^{a_1} \dots X_n^{a_n} = \left( \sum_{x_1 \in \mathbb{F}_q} x_1^{a_1} \right) \left( \sum_{x_2 \in \mathbb{F}_q} x_2^{a_2} \right) \dots \left( \sum_{x_n \in \mathbb{F}_q} x_n^{a_n} \right) \]
For the total sum to be non-zero, \textbf{every} single individual sum must be non-zero.
As mentioned in the Prerequisites, $\sum x^a$ is non-zero only if $a$ is a multiple of $(q-1)$ (and $a \geq q-1$).
So for the whole product to be non-zero, we need $a_1 \geq q-1, a_2 \geq q-1, \dots, a_n \geq q-1$.
This would mean the total degree $\sum a_i$ must be at least $n(q-1)$.
But we just proved in Step 3 that the degree of $P$ is \textit{less} than $n(q-1)$!
Therefore, every monomial in $P$ must have at least one variable whose exponent is too small to give a non-zero sum.
Consequently, the sum of $P(X)$ over all points must be $0$ in the field.

\textbf{Step 5: Conclude.}
Since $\sum P(X) = 0$ and $\sum P(X) \equiv N \pmod p$, we have $N \equiv 0 \pmod p$.
This means the number of common zeros is a multiple of $p$.
If the polynomials have no constant terms (or if $(0, \dots, 0)$ is a root), then $N \geq 1$. Since $N$ is a multiple of $p$, $N$ must be at least $p$. This proves that a system of low-degree equations always has non-trivial solutions over a finite field.
\end{solution}

\begin{exercise}
\l\label{2014:6}

\label{algebra:2014:6}(a) $AB=BA \implies \det(A^2+B^2) \geq 0$. (b) $k$ commuting matrices.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Matrix Factorization:} $x^2 + y^2 = (x+iy)(x-iy)$. This works for matrices if they commute ($AB=BA$).
    \item \textbf{Determinant Properties:} $\det(MN) = \det(M)\det(N)$ and $\det(\overline{M}) = \overline{\det(M)}$.
    \item \textbf{Simultaneous Triangularization:} If a set of matrices commute, they can be brought into upper triangular form by the same basis change over $\mathbb{C}$. The diagonal elements will be their eigenvalues.
    \item \textbf{Complex Conjugate Pairs:} For real matrices, non-real eigenvalues always come in pairs $(\lambda, \bar{\lambda})$.
\end{itemize}

\textbf{(a) Step 1: Factor the expression.}
We want to show $\det(A^2 + B^2) \geq 0$ for real matrices $A, B$ that commute.
Because $AB = BA$, we can factor the sum of squares just like we do with numbers:
\[ A^2 + B^2 = (A + iB)(A - iB) \]
Let's check: $(A+iB)(A-iB) = A^2 - iAB + iBA + B^2$. Since $AB=BA$, the middle terms cancel out.

\textbf{Step 2: Use properties of the determinant.}
Now, take the determinant of both sides:
\[ \det(A^2 + B^2) = \det(A + iB) \cdot \det(A - iB) \]
Since $A$ and $B$ are real matrices, the matrix $(A - iB)$ is the complex conjugate of $(A + iB)$.
The determinant of a conjugate matrix is the conjugate of the determinant:
\[ \det(A - iB) = \overline{\det(A + iB)} \]
Substituting this back into our product:
\[ \det(A^2 + B^2) = \det(A + iB) \cdot \overline{\det(A + iB)} = | \det(A + iB) |^2 \]
The absolute value squared of any complex number is always a non-negative real number.
Thus, $\det(A^2 + B^2) \geq 0$.

\textbf{(b) Step 3: Generalize to $k$ commuting matrices.}
Now we have $k$ matrices $A_1, A_2, \dots, A_k$ that all commute with each other. We want to show $\det(A_1^2 + \dots + A_k^2) \geq 0$.
Since they all commute, they share a common basis under which they are all upper triangular. Let the eigenvalues of $A_m$ be $\lambda_{m,1}, \lambda_{m,2}, \dots, \lambda_{m,n}$.
The determinant of a sum of matrices in triangular form is the product of the sums of their diagonal entries:
\[ \det\left(\sum_{m=1}^k A_m^2\right) = \prod_{j=1}^n \left( \sum_{m=1}^k \lambda_{m,j}^2 \right) \]

\textbf{Step 4: Analyze the eigenvalue combinations.}
Since the matrices are real, their eigenvalues come in conjugate pairs.
\begin{itemize}
    \item \textbf{Real case:} If for a specific index $j$, all $\lambda_{m,j}$ are real, then $\sum_m \lambda_{m,j}^2$ is a sum of squares of real numbers, which is $\geq 0$.
    \item \textbf{Complex case:} If the eigenvalues are complex, they must come in a pair with another index $j'$ such that $\lambda_{m,j'} = \bar{\lambda}_{m,j}$.
    The contribution of this pair to the product is:
    \[ \left( \sum_m \lambda_{m,j}^2 \right) \cdot \left( \sum_m \bar{\lambda}_{m,j}^2 \right) = \left( \sum_m \lambda_{m,j}^2 \right) \cdot \overline{\left( \sum_m \lambda_{m,j}^2 \right)} = \left| \sum_m \lambda_{m,j}^2 \right|^2 \]
    This product is also $\geq 0$.
\end{itemize}
Since every term or pair of terms in the product is non-negative, the entire determinant must be $\geq 0$.
\end{solution}

\subsection*{Team}
\begin{exercise}
\l\label{2014:7}

\label{algebra:2014:7}Let the special linear group $SL_2(\mathbb{R}) = \{ g = \begin{pmatrix} a & b \\ c & d \end{pmatrix} \in M_2(\mathbb{R}) : \det g = 1 \}$ act on the upper half plane $\mathbb{H} = \{ z = x + iy \in \mathbb{C} : y > 0 \}$ linear fractionally:
\[ \begin{pmatrix} a & b \\ c & d \end{pmatrix} z = \frac{az + b}{cz + d}. \]
\begin{enumerate}
    \item Prove that the action is transitive, i.e., for any two $z_1, z_2 \in \mathbb{H}$, there is $g \in SL_2(\mathbb{R})$ such that $gz_1 = z_2$.
    \item For a fixed $z \in \mathbb{H}$, prove that its stabilizer $G_z = \{ g \in SL_2(\mathbb{R}) : gz = z \}$ is isomorphic to $SO_2(\mathbb{R}) = \{ g \in M_2(\mathbb{R}) : g g^t = 1 \}$.
    \item Let $\Gamma(2) = \{ \begin{pmatrix} a & b \\ c & d \end{pmatrix} \in SL_2(\mathbb{Z}) : a \equiv d \equiv 1 \pmod 2, b \equiv c \equiv 0 \pmod 2 \}$ be a discrete subgroup. How many orbits does its action have on $\mathbb{Q} \cup \{ \infty \}$? Give a representative for each orbit.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Fractional Linear Transformation:} A mapping $z \mapsto \frac{az+b}{cz+d}$. These preserve the upper half-plane if $\det > 0$ and entries are real.
    \item \textbf{Transitive Action:} An action where you can get from any point in the set to any other point using a group element.
    \item \textbf{Stabilizer Subgroup ($G_z$):} The set of elements that leave a specific point $z$ fixed.
    \item \textbf{Isomorphism of Groups:} A one-to-one mapping between groups that preserves the group operation.
    \item \textbf{Principal Congruence Subgroup ($\Gamma(N)$):} Subgroups of $SL_2(\mathbb{Z})$ where matrices are the identity modulo $N$.
\end{itemize}

\textbf{(1) Step 1: Prove transitivity by connecting to $i$.}
To show the action is transitive, we want to prove that for any $z_1, z_2 \in \mathbb{H}$, there exists $g$ such that $gz_1 = z_2$.
It's enough to show that \textit{any} $z = x+iy$ can be reached starting from the point $i = 0 + 1i$.
We want to find $g = \begin{pmatrix} a & b \\ c & d \end{pmatrix}$ such that $\frac{ai+b}{ci+d} = x+iy$.
Consider the matrix $g = \begin{pmatrix} \sqrt{y} & x/\sqrt{y} \\ 0 & 1/\sqrt{y} \end{pmatrix}$.
First, check the determinant: $\det g = \sqrt{y}(1/\sqrt{y}) - 0 = 1$. So $g \in SL_2(\mathbb{R})$.
Now apply it to $i$:
\[ g(i) = \frac{\sqrt{y} \cdot i + x/\sqrt{y}}{0 \cdot i + 1/\sqrt{y}} = \frac{yi/\sqrt{y} + x/\sqrt{y}}{1/\sqrt{y}} = yi + x = z. \]
Since we can reach any $z$ from $i$, we can get from $z_1$ to $z_2$ by going through $i$: $z_1 \xrightarrow{g_1^{-1}} i \xrightarrow{g_2} z_2$. The element $g_2 g_1^{-1}$ does the job.

\textbf{(2) Step 2: Find the stabilizer of $i$.}
Let $z=i$ and find all matrices such that $\frac{ai+b}{ci+d} = i$.
Multiplying by the denominator: $ai + b = i(ci + d) = -c + di$.
For these complex numbers to be equal, their real and imaginary parts must match:
\begin{itemize}
    \item Real parts: $b = -c \implies c = -b$.
    \item Imaginary parts: $a = d$.
\end{itemize}
Substituting these into the determinant condition $\det g = 1$:
\[ ad - bc = a^2 - b(-b) = a^2 + b^2 = 1. \]
The stabilizer consists of matrices of the form $\begin{pmatrix} a & -b \\ b & a \end{pmatrix}$ with $a^2 + b^2 = 1$.
If we write $a = \cos \theta$ and $b = \sin \theta$, these are exactly the rotation matrices in $SO_2(\mathbb{R})$.
Since all points in $\mathbb{H}$ are equivalent (transitivity), their stabilizers are all isomorphic.

\textbf{(3) Step 3: Count the orbits on the "boundary" points.}
The set $\mathbb{Q} \cup \{ \infty \}$ can be thought of as the "cusps" or boundary of the upper half-plane.
The group $SL_2(\mathbb{Z})$ acts transitively on these points, meaning they all belong to one single orbit (usually represented by $\infty$, which is $1/0$).
The subgroup $\Gamma(2)$ is a smaller group (index 6 in $SL_2(\mathbb{Z})$). Its action on the cusps is not transitive.
The number of orbits is equal to the number of ways to pick a point in the projective line $\mathbb{P}^1(\mathbb{Z}/2\mathbb{Z})$.
Over the field with 2 elements ($\mathbb{Z}/2\mathbb{Z}$), there are exactly 3 points:
\begin{itemize}
    \item $(0:1)$ representing $0/1 = 0$.
    \item $(1:1)$ representing $1/1 = 1$.
    \item $(1:0)$ representing $1/0 = \infty$.
\end{itemize}
Since $\Gamma(2)$ matrices are the identity modulo 2, they cannot swap these three points. Thus, there are exactly \textbf{3 orbits}, with representatives $0, 1, \text{ and } \infty$.
\end{solution}

\begin{exercise}
\l\label{2014:8}

\label{algebra:2014:8}Let $p \geq 7$ be an odd prime number.
\begin{enumerate}
    \item Evaluate the rational number $\cos(\pi/7) \cdot \cos(2\pi/7) \cdot \cos(3\pi/7)$.
    \item Show that $\prod_{n=1}^{(p-1)/2} \cos(n\pi/p)$ is a rational number and determine its value.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Double Angle Formula:} $\sin(2\theta) = 2 \sin \theta \cos \theta$.
    \item \textbf{Symmetry of Cosine:} $\cos(\pi - \theta) = -\cos \theta$ and $\sin(\pi - \theta) = \sin \theta$.
    \item \textbf{Roots of Unity:} The equation $x^{2p} = 1$ has roots $e^{i2\pi n / 2p}$.
    \item \textbf{Cyclotomic Polynomials:} The product of differences with roots of unity often results in integers or simple fractions.
\end{itemize}

\textbf{(1) Step 1: Calculate the specific product for $p=7$.}
We want to evaluate $C = \cos(\frac{\pi}{7}) \cos(\frac{2\pi}{7}) \cos(\frac{3\pi}{7})$.
Let's multiply both sides by $\sin(\frac{\pi}{7})$ and apply the double angle formula repeatedly:
\begin{align*}
C \cdot \sin(\frac{\pi}{7}) &= \left[ \sin(\frac{\pi}{7}) \cos(\frac{\pi}{7}) \right] \cos(\frac{2\pi}{7}) \cos(\frac{3\pi}{7}) \\
&= \frac{1}{2} \sin(\frac{2\pi}{7}) \cos(\frac{2\pi}{7}) \cos(\frac{3\pi}{7}) \quad \dots \text{ (apply formula again)} \\
&= \frac{1}{4} \sin(\frac{4\pi}{7}) \cos(\frac{3\pi}{7})
\end{align*}
Note that $\sin(\frac{4\pi}{7}) = \sin(\pi - \frac{3\pi}{7}) = \sin(\frac{3\pi}{7})$.
So, $C \sin(\frac{\pi}{7}) = \frac{1}{4} \sin(\frac{3\pi}{7}) \cos(\frac{3\pi}{7}) = \frac{1}{8} \sin(\frac{6\pi}{7})$.
Again, using symmetry: $\sin(\frac{6\pi}{7}) = \sin(\pi - \frac{\pi}{7}) = \sin(\frac{\pi}{7})$.
Dividing both sides by the non-zero value $\sin(\frac{\pi}{7})$, we get:
\[ C = \frac{1}{8} \]

\textbf{(2) Step 2: Generalize for any prime $p$.}
Consider the product $P = \prod_{n=1}^{(p-1)/2} \cos(\frac{n\pi}{p})$.
Using Euler's formula $2 \cos \theta = e^{i\theta} + e^{-i\theta}$, we can relate the product to the roots of the equation $x^{2p} - 1 = 0$.
The set of values $\{\cos(n\pi/p)\}$ are the real parts of the $2p$-th roots of unity.
From the theory of cyclotomic polynomials, the product of these specific cosine terms can be shown to involve only the number 2 and signs.
The exact value is:
\[ P = \pm 2^{-(p-1)/2} \]
The sign depends on the quadratic residue of 2 modulo $p$:
\begin{itemize}
    \item The sign is positive (+) if $p \equiv 1$ or $7 \pmod 8$ (meaning 2 is a quadratic residue).
    \item The sign is negative (-) if $p \equiv 3$ or $5 \pmod 8$ (meaning 2 is not a quadratic residue).
\end{itemize}
In all cases, since $p$ is a fixed prime, this is a rational number.
\end{solution}

\begin{exercise}
\l\label{2014:9}

\label{algebra:2014:9}For any $3 \times 3$ matrix $A \in M_3(\mathbb{Q})$, let $A^{db}$ be the $6 \times 6$ matrix $A^{db} := \begin{pmatrix} 0 & I_3 \\ A & 0 \end{pmatrix}$.
\begin{enumerate}
    \item Express the characteristic and minimal polynomials of $A^{db}$ over $\mathbb{Q}$ in terms of those of $A$.
    \item Suppose $A, B \in M_3(\mathbb{Q})$ are such that $A^{db}$ and $B^{db}$ are conjugate. Are $A$ and $B$ conjugate?
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Block Matrices:} Matrices composed of smaller matrices.
    \item \textbf{Characteristic Polynomial ($P_A(t)$):} Defined by $\det(tI - A)$.
    \item \textbf{Minimal Polynomial ($m_A(t)$):} The lowest degree polynomial such that $m_A(A) = 0$.
    \item \textbf{Conjugate Matrices:} Two matrices $X, Y$ are conjugate if $Y = PXP^{-1}$. This happens if and only if they have the same invariant factors.
    \item \textbf{Direct Sum:} A matrix with blocks on the diagonal, like $A \oplus A$.
\end{itemize}

\textbf{(1) Step 1: Calculate the characteristic polynomial of $A^{db}$.}
The matrix is $A^{db} = \begin{pmatrix} 0 & I \\ A & 0 \end{pmatrix}$. To find its characteristic polynomial, we calculate:
\[ P_{A^{db}}(t) = \det(tI_{6} - A^{db}) = \det \begin{pmatrix} tI_3 & -I_3 \\ -A & tI_3 \end{pmatrix} \]
For a $2 \times 2$ block matrix $\begin{pmatrix} X & Y \\ Z & W \end{pmatrix}$, if the blocks commute, the determinant is $\det(XW - YZ)$. Here $tI_3$ commutes with everything.
\[ = \det( (tI_3)(tI_3) - (-I_3)(-A) ) = \det( t^2 I_3 - A ) \]
By the definition of the characteristic polynomial for $A$:
\[ P_{A^{db}}(t) = P_A(t^2) \]

\textbf{Step 2: Relate the minimal polynomials.}
Notice that $(A^{db})^2 = \begin{pmatrix} 0 & I \\ A & 0 \end{pmatrix} \begin{pmatrix} 0 & I \\ A & 0 \end{pmatrix} = \begin{pmatrix} A & 0 \\ 0 & A \end{pmatrix} = A \oplus A$.
If $Q(t)$ is any polynomial, then $Q(A^{db})^2 = Q(A) \oplus Q(A)$.
If $t^2$ satisfies the minimal polynomial of $A$ (i.e., $m_A(A) = 0$), then the square of $A^{db}$ vanishes under $m_A$.
Thus, $m_{A^{db}}(t)$ is $m_A(t^2)$ (as long as $A$ is non-singular; otherwise it might be a factor of it).

\textbf{(2) Step 3: Compare conjugacy classes.}
Suppose $A^{db}$ and $B^{db}$ are conjugate. Then their squares must also be conjugate:
\[ (A^{db})^2 \sim (B^{db})^2 \implies A \oplus A \sim B \oplus B \]
A property of direct sums is that $X \oplus X \sim Y \oplus Y$ if and only if $X \sim Y$.
Why? Because the invariant factors of $X \oplus X$ are just the invariant factors of $X$, each listed twice. If two matrices have the same invariant factors (counting duplicates), the original matrices must have had the same invariant factors.
Since $A$ and $B$ have the same invariant factors, they are conjugate. Thus, the answer is \textbf{Yes}.
\end{solution}

\begin{exercise}
\l\label{2014:10}

\label{algebra:2014:10}Classify all groups of order 8.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Lagrange's Theorem:} The order of a subgroup must divide the order of the group.
    \item \textbf{Fundamental Theorem of Finite Abelian Groups:} Every abelian group is a product of cyclic groups.
    \item \textbf{Center of a Group ($Z(G)$):} The set of elements that commute with everything. For groups of prime power order ($p^k$), the center is always non-trivial.
    \item \textbf{Non-abelian Structure:} If $G/Z(G)$ is cyclic, then $G$ must be abelian.
\end{itemize}

\textbf{Step 1: Classify Abelian groups of order 8.}
The size is $8 = 2^3$. According to the Fundamental Theorem, the possible abelian groups correspond to the partitions of 3:
\begin{itemize}
    \item \textbf{Partition $3$:} $\mathbb{Z}/8\mathbb{Z}$ (a single cyclic group $C_8$).
    \item \textbf{Partition $2+1$:} $\mathbb{Z}/4\mathbb{Z} \times \mathbb{Z}/2\mathbb{Z}$ (one large and one small cyclic group).
    \item \textbf{Partition $1+1+1$:} $\mathbb{Z}/2\mathbb{Z} \times \mathbb{Z}/2\mathbb{Z} \times \mathbb{Z}/2\mathbb{Z}$ (the elementary abelian group).
\end{itemize}
These are all distinct because their elements have different maximum orders (8, 4, and 2 respectively).

\textbf{Step 2: Find Non-Abelian groups.}
Let $G$ be non-abelian of order 8. We know the center $Z(G)$ must have size at least 2.
If the center had size 4, then $G/Z(G)$ would have size 2, making it cyclic. But if $G/Z(G)$ is cyclic, then $G$ is abelian.
So the size of $Z(G)$ must be exactly 2.
Also, $G$ cannot have an element of order 8 (that would make it cyclic/abelian).

\textbf{Step 3: Analyze the element orders.}
\begin{enumerate}
    \item \textbf{Case 1: $G$ has an element $r$ of order 4.}
    The subgroup $\langle r \rangle$ is normal (since it has index 2). Let $s$ be an element outside this subgroup.
    Since $srs^{-1}$ must be in $\langle r \rangle$ and $G$ is non-abelian, the only possibility is $srs^{-1} = r^{-1}$.
    \begin{itemize}
        \item If $s^2 = 1$, we get the \textbf{Dihedral Group $D_8$} (symmetries of a square).
        \item If $s^2 = r^2$ (the only other choice for $s^2$ in the center), we get the \textbf{Quaternion Group $Q_8$}.
    \end{itemize}
    \item \textbf{Case 2: All elements have order 2.}
    If $x^2 = 1$ for all $x$, then $(xy)^2 = 1 \implies xyxy = 1 \implies xy = y^{-1}x^{-1} = yx$.
    This would make the group abelian, which contradicts our assumption.
\end{enumerate}

\textbf{Conclusion:}
There are exactly 5 groups of order 8 up to isomorphism:
1. $C_8$ (Abelian)
2. $C_4 \times C_2$ (Abelian)
3. $C_2 \times C_2 \times C_2$ (Abelian)
4. $D_8$ (Non-Abelian)
5. $Q_8$ (Non-Abelian)
\end{solution}

\begin{exercise}
\l\label{2014:11}

\label{algebra:2014:11}Let $V$ be a finite dimensional complex vector space with a non-degenerate symmetric bilinear form $(\cdot, \cdot)$.
\begin{enumerate}
    \item Prove that $(V \otimes_\mathbb{C} V)^{O(V)} \cong \text{End}_{O(V)}(V)$.
    \item Prove that $(V \otimes_\mathbb{C} V)^{O(V)}$ is 1-dimensional.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Dual Space ($V^*$):} The space of linear functions from $V$ to $\mathbb{C}$.
    \item \textbf{Canonical Isomorphism:} If $V$ has a non-degenerate form, there is a natural way to identify vectors in $V$ with functions in $V^*$.
    \item \textbf{Fixed Points ($X^G$):} The subset of $X$ that is not moved by the action of $G$.
    \item \textbf{Schur's Lemma:} A linear map that commutes with an irreducible representation must be a scalar multiple of the identity.
\end{itemize}

\textbf{(1) Step 1: Identify the spaces.}
There is a general rule in linear algebra: $V \otimes V^*$ is naturally isomorphic to the space of all linear operators $\text{End}(V)$.
Because our vector space $V$ has a \textit{non-degenerate symmetric bilinear form}, we can identify $V$ with $V^*$ (the map is $v \mapsto (v, \cdot)$).
Substituting this into our tensor product formula:
\[ V \otimes V \cong V \otimes V^* \cong \text{End}(V) \]

\textbf{Step 2: Compare the group actions.}
When an element $g$ of the orthogonal group $O(V)$ acts on a tensor $v \otimes w$, it maps it to $gv \otimes gw$.
Under our identification with $\text{End}(V)$, this action corresponds to \textbf{conjugation}: an operator $A$ is mapped to $gAg^{-1}$.
The "fixed point subspace" consists of elements that remain unchanged by this action:
\[ gAg^{-1} = A \implies gA = Ag \]
This condition means the operator $A$ must commute with every transformation in the group $O(V)$.
The set of such operators is called the \textbf{centralizer} of the group action, denoted $\text{End}_{O(V)}(V)$.

\textbf{(2) Step 3: Apply Schur's Lemma.}
The standard representation of the orthogonal group $O(V)$ on the space $V$ is \textbf{irreducible}.
Schur's Lemma tells us that any linear operator that commutes with an irreducible group representation must be a scalar multiple of the identity matrix ($A = cI$ for some $c \in \mathbb{C}$).
Since the space of scalar matrices $\{ cI : c \in \mathbb{C} \}$ is 1-dimensional, the fixed point subspace must also be \textbf{1-dimensional}.
The unique invariant element corresponds to the bilinear form itself.
\end{solution}

\begin{exercise}
\l\label{2014:12}

\label{algebra:2014:12}Let $c$ be a non-zero rational integer.
\begin{enumerate}
    \item Factorize $f(x, y, z) = x^3 + cy^3 + c^2z^3 - 3cxyz$ over $\mathbb{C}$.
    \item When $c = \theta^3$ is a cube, show there are only finitely many integer solutions to $f(x, y, z) = 1$.
    \item When $c$ is not a cube, show there are infinitely many integer solutions.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Cubic Identity:} $x^3+y^3+z^3-3xyz = (x+y+z)(x^2+y^2+z^2-xy-yz-zx)$.
    \item \textbf{Complex Roots of Unity:} $\omega = e^{2\pi i/3}$. Note $\omega^2 + \omega + 1 = 0$.
    \item \textbf{Pure Cubic Field ($K = \mathbb{Q}(\sqrt[3]{c})$):} A field extension where we add the cube root of a number.
    \item \textbf{Dirichlet's Unit Theorem:} A theorem that counts how many independent "units" (integers with norm 1) exist in a field.
\end{itemize}

\textbf{(1) Step 1: Factor the polynomial over the complex numbers.}
Let $\theta = \sqrt[3]{c}$. Notice that our polynomial is related to the sum of cubes formula.
Over $\mathbb{C}$, the factors are:
\[ f(x, y, z) = (x + y\theta + z\theta^2)(x + y\omega\theta + z\omega^2\theta^2)(x + y\omega^2\theta + z\omega\theta^2) \]
This is because if we multiply these three linear terms, we recover the original cubic form. The terms in the parentheses are the three "embeddings" of the element $\alpha = x + y\theta + z\theta^2$ into $\mathbb{C}$.

\textbf{(2) Step 2: Analyze the "Cube Case" ($c = \theta^3$).}
If $c$ is already a perfect cube (like $c=8, \theta=2$), then $\sqrt[3]{c}$ is an integer.
The expression $f(x, y, z) = 1$ factors over the integers:
\[ (x + \theta y + \theta^2 z) \cdot (\text{quadratic part}) = 1 \]
Since we are looking for integer solutions, the first factor $(x + \theta y + \theta^2 z)$ must be an integer that divides 1, so it must be $\pm 1$.
This means our variables are restricted to a 2D plane. Within this plane, the second factor (the quadratic part) must also be 1.
A quadratic equation in 2 variables defines a conic section. In this specific case, it defines an ellipse.
Since an ellipse is a bounded curve, it can only contain a finite number of points with integer coordinates (lattice points). Thus, there are only \textbf{finitely many solutions}.

\textbf{(3) Step 3: Analyze the "Non-Cube Case".}
If $c$ is not a perfect cube, then $\theta = \sqrt[3]{c}$ is an irrational number. The set of numbers $\{ x + y\theta + z\theta^2 \}$ forms a ring of integers in the field $K = \mathbb{Q}(\theta)$.
The expression $f(x, y, z)$ is exactly the \textbf{field norm} $N_{K/\mathbb{Q}}(\alpha)$.
According to Dirichlet's Unit Theorem, the group of units (elements with norm $\pm 1$) in this field has a "rank".
The number of independent units is $r = r_1 + r_2 - 1$.
\begin{itemize}
    \item $r_1$ is the number of real embeddings. Here $\sqrt[3]{c}$ has 1 real embedding.
    \item $r_2$ is the number of pairs of complex embeddings. The other two cube roots are complex conjugates, so $r_2 = 1$.
\end{itemize}
Rank $r = 1 + 1 - 1 = 1$.
Since the rank is 1, there exists a "fundamental unit" $\epsilon$. Every power $\epsilon^n$ (for $n=1, 2, 3, \dots$) is a distinct unit with norm 1.
Since there are infinitely many powers of $\epsilon$, there are \textbf{infinitely many integer solutions} $(x, y, z)$.
\end{solution}
