\section{2019 Algebra \& Number Theory}

\subsection*{Individual}
\begin{exercise}
Let $G$ be a finite group. Assume that for any representation $V$ of $G$ over a field of characteristic zero, the character $\chi_V$ takes value in $\mathbb{Q}$. Assume $g$ is an element in $G$ such that $g^{2019} = 1$.

Prove that $g$ and $g^{19}$ are conjugate in $G$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Representation theory of finite groups and characters.
    \item Theorem: Characters of a finite group are rational-valued if and only if for every element $g \in G$, $g$ is conjugate to $g^k$ for all integers $k$ such that $\gcd(k, |g|) = 1$.
    \item Basic number theory: prime factorization and the concept of order of a group element.
\end{itemize}

\textbf{Intuition:}
The problem asks us to show that two powers of a group element are conjugate. The key hint is the condition that all characters are rational. In representation theory, this condition is equivalent to saying that the group's "symmetry" allows us to swap an element with its "coprime powers" via conjugation. Thus, the task reduces to verifying a simple number-theoretic condition about the exponents.

\par

\textbf{Steps:}

1. \textbf{Identify the character property:} We are given that all characters $\chi_V$ of $G$ over characteristic zero take values in $\mathbb{Q}$. A standard result in group theory states that this happens if and only if for any $g \in G$, $g$ is conjugate to $g^k$ for every integer $k$ coprime to the order of $g$, denoted $|g|$.

\par

2. \textbf{Determine the possible orders of $g$:} The problem states that $g^{2019} = 1$. By the definition of the order of an element, $|g|$ must be a divisor of $2019$.

\par

3. \textbf{Factorize the exponent:} To find the divisors of $2019$, we perform its prime factorization. Summing the digits $2+0+1+9 = 12$ shows it is divisible by $3$. We find $2019 = 3 \times 673$. Testing small primes for $673$, we see it is not divisible by $2, 3, 5, 7, 11, 13, 17, 19, 23$. In fact, $673$ is prime. Thus, the possible values for $|g|$ are $\{1, 3, 673, 2019\}$.

\par

4. \textbf{Check the coprimality condition:} We need to see if $k=19$ is coprime to $|g|$. Since $19$ is a prime number, it is coprime to $|g|$ as long as $|g|$ is not a multiple of $19$. Looking at our list $\{1, 3, 673, 2019\}$, none of these are divisible by $19$ (e.g., $673 = 19 \times 35 + 8$).

\par

5. \textbf{Conclusion:} Since $\gcd(19, |g|) = 1$ for all possible orders of $g$, the property of rational characters guarantees that $g$ is conjugate to $g^{19}$ in $G$.
\end{solution}

\begin{exercise}
Let $p$ be a prime number, and let $\mathbb{F}_p$ be the finite field with $p$ elements. Let $F = \mathbb{F}_p(t)$ be the field of rational functions over $\mathbb{F}_p$. Consider all subfields $C$ of $F$ such that $F/C$ is a finite Galois extension.

\begin{enumerate}
    \item Show that among such subfields, there is a smallest one $C_0$, i.e., $C_0$ is contained in any other $C$.
    \item What is the degree of $F/C_0$?
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Galois theory for function fields.
    \item Automorphisms of the rational function field $\mathbb{F}_p(t)$, which are fractional linear transformations.
    \item The isomorphism $\mathrm{Aut}(\mathbb{F}_p(t)) \cong \mathrm{PGL}_2(\mathbb{F}_p)$.
    \item The Fundamental Theorem of Galois Theory: subfields correspond to subgroups.
\end{itemize}

\textbf{Intuition:}
The field of rational functions $F$ has a group of "symmetries" (automorphisms). Any finite Galois extension $F/C$ corresponds to a finite group of these symmetries that keep $C$ fixed. To find the "smallest" subfield $C$, we need to find the "largest" group of symmetries possible, because the more symmetries we require a subfield to satisfy, the smaller that subfield must be.

\par

\textbf{Steps:}

1. \textbf{Identify the automorphism group:} The automorphisms of $F = \mathbb{F}_p(t)$ that fix the constant field $\mathbb{F}_p$ are given by transformations of the form $t \mapsto \frac{at+b}{ct+d}$ where $ad-bc \neq 0$. This group is the Projective General Linear group $\mathrm{PGL}_2(\mathbb{F}_p)$.

\par

2. \textbf{Apply Galois Correspondence:} For any subfield $C$ such that $F/C$ is a finite Galois extension, there exists a finite subgroup $G \subset \mathrm{PGL}_2(\mathbb{F}_p)$ such that $C$ is the fixed field $F^G$. The Fundamental Theorem of Galois Theory tells us that $C_1 \subset C_2$ if and only if $G_2 \subset G_1$.

\par

3. \textbf{Identify the smallest subfield:} A subfield $C_0$ is contained in every other such $C$ if its corresponding group $G_0$ contains every other finite subgroup $G$. Since $\mathrm{PGL}_2(\mathbb{F}_p)$ is itself a finite group (because $\mathbb{F}_p$ is finite), the largest possible subgroup is the entire group $G_0 = \mathrm{PGL}_2(\mathbb{F}_p)$. Thus, $C_0 = F^{\mathrm{PGL}_2(\mathbb{F}_p)}$ is the unique smallest subfield.

\par

4. \textbf{Calculate the degree:} The degree of the extension $[F : C_0]$ is equal to the order of the Galois group $G_0 = \mathrm{PGL}_2(\mathbb{F}_p)$.

\par

5. \textbf{Compute the group order:} The order of $\mathrm{GL}_2(\mathbb{F}_p)$ is the number of bases of $\mathbb{F}_p^2$, which is $(p^2-1)(p^2-p)$. The group $\mathrm{PGL}_2(\mathbb{F}_p)$ is the quotient of $\mathrm{GL}_2(\mathbb{F}_p)$ by the scalar matrices (of which there are $p-1$).
\[ |\mathrm{PGL}_2(\mathbb{F}_p)| = \frac{(p^2-1)(p^2-p)}{p-1} = (p+1)p(p-1) = p^3 - p. \]
Thus, the degree is $p^3 - p$.
\end{solution}

\begin{exercise}
Let $R \subset R'$ be an integral extension of commutative rings. Let $\mathfrak{p}'$ be a prime ideal of $R'$. Prove that $\mathfrak{p}'$ is a maximal ideal of $R'$ if and only if $\mathfrak{p}' \cap R$ is a maximal ideal of $R$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Definitions of prime and maximal ideals.
    \item Concept of an integral extension (elements satisfying monic polynomials).
    \item Result: A commutative ring $D$ is a field if and only if its only ideals are $0$ and $D$.
    \item Result: In an integral extension of integral domains $D \subset D'$, $D$ is a field iff $D'$ is a field.
\end{itemize}

\textbf{Intuition:}
Integral extensions "preserve" the size of the ring in a specific sense: they don't change the "dimension." Being a maximal ideal is equivalent to saying the quotient ring is a field (dimension 0). This proof shows that the property of being a field "lifts" and "descends" through integral extensions.

\par

\textbf{Steps:}

1. \textbf{Reduce to integral domains:} Let $\mathfrak{p} = \mathfrak{p}' \cap R$. This is a prime ideal in $R$. Form the quotient rings $D = R/\mathfrak{p}$ and $D' = R'/\mathfrak{p}'$. Both are integral domains. The inclusion $R \subset R'$ induces an inclusion $D \subset D'$. Since $R'$ is integral over $R$, the quotient $D'$ is integral over $D$.

\par

2. \textbf{Forward direction ($\implies$):} Assume $\mathfrak{p}'$ is maximal. Then $D'$ is a field. We want to show $D$ is a field. Let $x \in D$ be non-zero. Since $x \in D \subset D'$ and $D'$ is a field, $x^{-1}$ exists in $D'$. Because $D'$ is integral over $D$, $x^{-1}$ satisfies an equation:
\[ (x^{-1})^n + a_{n-1}(x^{-1})^{n-1} + \dots + a_0 = 0, \quad a_i \in D. \]
Multiply by $x^{n-1}$ to get $x^{-1} = -(a_{n-1} + a_{n-2}x + \dots + a_0 x^{n-1})$. The right side is in $D$, so $x^{-1} \in D$. Thus $D$ is a field, and $\mathfrak{p}$ is maximal.

\par

3. \textbf{Backward direction ($\impliedby$):} Assume $\mathfrak{p}$ is maximal. Then $D$ is a field. We want to show $D'$ is a field. Let $\alpha \in D'$ be non-zero. Since $D'$ is integral over $D$, $\alpha$ satisfies:
\[ \alpha^m + c_{m-1}\alpha^{m-1} + \dots + c_1 \alpha + c_0 = 0, \quad c_i \in D. \]
Since $D'$ is a domain, we can assume $c_0 \neq 0$ (otherwise divide by $\alpha$). Then $\alpha(\alpha^{m-1} + \dots + c_1) = -c_0$. Since $c_0 \in D$ is a non-zero element of a field, $c_0^{-1}$ exists. Thus $\alpha \cdot [-c_0^{-1}(\alpha^{m-1} + \dots + c_1)] = 1$. The term in brackets is in $D'$, so $\alpha$ is invertible. $D'$ is a field, and $\mathfrak{p}'$ is maximal.
\end{solution}

\begin{exercise}
\begin{enumerate}
    \item Prove that $\mathrm{GL}_n(\mathbb{C})$ is path-connected.
    \item Let $X = \{A \in \mathrm{GL}_n(\mathbb{C}) \mid A^m = \mathrm{Id}\}$. Describe the path-connected component of $X$ and prove your answer.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Topology of matrix groups.
    \item Jordan Normal Form and diagonalizability.
    \item Properties of $\mathbb{C}^*$ (the punctured plane is connected).
    \item Roots of unity and characteristic polynomials.
\end{itemize}

\textbf{Intuition:}
For Part 1, the complex numbers are "wide enough" that removing a small set (like the roots of a polynomial) doesn't disconnect the space. For Part 2, the condition $A^m=I$ forces the matrix to be diagonalizable with very specific eigenvalues. Since you can't continuously change a discrete set of eigenvalues without breaking the $A^m=I$ rule, the components must be determined by the "counts" of each eigenvalue.

\par

\textbf{Steps:}

1. \textbf{Path-connectedness of $\mathrm{GL}_n(\mathbb{C})$:} Consider $\mathrm{GL}_n(\mathbb{C})$ as an open subset of $\mathbb{C}^{n^2}$. It is the complement of the set $Z = \{A \mid \det(A) = 0\}$. In $\mathbb{C}^N$, the zero set of a non-constant polynomial has real codimension 2. Removing a set of codimension 2 from a connected space keeps it path-connected. Alternatively, any matrix can be connected to its Jordan form, then to a diagonal matrix, and each diagonal entry $\lambda \in \mathbb{C}^*$ can be connected to $1$ in the punctured plane $\mathbb{C} \setminus \{0\}$.

\par

2. \textbf{Structure of elements in $X$:} If $A^m = I$, then $A$ satisfies the polynomial $z^m - 1 = 0$. This polynomial has $m$ distinct roots (the $m$-th roots of unity $\omega^k = e^{2\pi i k / m}$). Since the minimal polynomial of $A$ has distinct roots, $A$ is diagonalizable.

\par

3. \textbf{Eigenvalue multiplicities:} Every $A \in X$ is determined up to conjugacy by the multiplicities $n_k$ of each root $\omega^k$ as an eigenvalue. These integers $n_0, n_1, \dots, n_{m-1}$ satisfy $\sum n_k = n$.

\par

4. \textbf{Connectedness of orbits:} Each set of matrices with a fixed set of multiplicities forms a conjugacy orbit $\mathcal{O}_{\mathbf{n}}$. This orbit is the image of the continuous map $\mathrm{GL}_n(\mathbb{C}) \to \mathrm{GL}_n(\mathbb{C})$ given by $P \mapsto P D_{\mathbf{n}} P^{-1}$. Since $\mathrm{GL}_n(\mathbb{C})$ is connected, each orbit is a path-connected set.

\par

5. \textbf{Discreteness of components:} Because the eigenvalues are fixed at discrete points on the unit circle, any continuous path $A(t)$ in $X$ must preserve the characteristic polynomial (and thus the multiplicities). A jump in multiplicities would be a discontinuous change. Therefore, the path-connected components are exactly these conjugacy orbits. The number of components is the number of ways to partition $n$ into $m$ non-negative integer parts.
\end{solution}

\begin{exercise}
The Fibonacci sequence is defined by
\[
F_0 = 0, \quad F_1 = 1, \quad F_{n+2} = F_{n+1} + F_n.
\]
Let $p$ be a prime number.

\begin{enumerate}
    \item Show that if $p \equiv 1, 4 \pmod{5}$, then $p$ divides $F_{p-1}$.
    \item Let $\mathbb{F}_{p^2}$ be the finite field of $p^2$ elements. Show that the norm map $N: \mathbb{F}_{p^2}^\times \to \mathbb{F}_p^\times$ is surjective, and deduce the cardinality of the kernel of $N$.
    \item Show that if $p \equiv 2, 3 \pmod{5}$, then $p$ divides $F_{p+1}$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Binet's Formula for Fibonacci numbers.
    \item Quadratic Reciprocity (Legendre symbol $\left(\frac{5}{p}\right)$).
    \item Finite field arithmetic and the Frobenius automorphism $x \mapsto x^p$.
    \item Norm map $N(x) = x \cdot x^p = x^{p+1}$ in $\mathbb{F}_{p^2}$.
\end{itemize}

\textbf{Intuition:}
Fibonacci numbers are built from the roots of $x^2-x-1=0$. Modulo $p$, these roots either exist in $\mathbb{F}_p$ (if $5$ is a square) or in a quadratic extension $\mathbb{F}_{p^2}$ (if $5$ is not a square). Part 1 and 3 use Fermat's Little Theorem and its extension to handle these two cases.

\par

\textbf{Steps:}

1. \textbf{Binet's Formula:} $F_n = \frac{\alpha^n - \beta^n}{\alpha - \beta}$ where $\alpha, \beta = \frac{1 \pm \sqrt{5}}{2}$.

\par

2. \textbf{Case $p \equiv 1, 4 \pmod 5$:} By Quadratic Reciprocity, $\left(\frac{5}{p}\right) = \left(\frac{p}{5}\right)$. If $p \equiv 1, 4$, then $\left(\frac{p}{5}\right) = 1$. Thus $\sqrt{5}$ exists in $\mathbb{F}_p$. By Fermat's Little Theorem, $\alpha^{p-1} \equiv 1$ and $\beta^{p-1} \equiv 1 \pmod p$. Then $F_{p-1} = \frac{1-1}{\sqrt{5}} = 0$, so $p \mid F_{p-1}$.

\par

3. \textbf{Norm Map Surjectivity:} The norm map $N: \mathbb{F}_{p^2}^\times \to \mathbb{F}_p^\times$ is $x \mapsto x^{p+1}$. Since $\mathbb{F}_{p^2}^\times$ is a cyclic group of order $p^2-1$, the image of the $(p+1)$-th power map has size $\frac{p^2-1}{\gcd(p+1, p^2-1)} = \frac{p^2-1}{p+1} = p-1$. This matches the size of $\mathbb{F}_p^\times$, so it is surjective. The kernel size is $\frac{p^2-1}{p-1} = p+1$.

\par

4. \textbf{Case $p \equiv 2, 3 \pmod 5$:} Here $\left(\frac{5}{p}\right) = -1$, so $\sqrt{5} \notin \mathbb{F}_p$. The roots $\alpha, \beta$ are in $\mathbb{F}_{p^2}$. The Frobenius map $\sigma(x)=x^p$ swaps the roots of the quadratic, so $\alpha^p = \beta$. The norm is $N(\alpha) = \alpha \alpha^p = \alpha \beta = -1$. Thus $\alpha^{p+1} = -1$ and $\beta^{p+1} = -1$. Finally, $F_{p+1} = \frac{-1 - (-1)}{\sqrt{5}} = 0 \pmod p$.
\end{solution}

\subsection*{Team}
\begin{exercise}
Let $S_n$ be the group of permutations of $\{1, 2, \ldots, n\}$. Let $\sigma \in S_n$ be the permutation
\[(1, n)(2, n-1)\cdots(k, n-k+1)\cdots.\]
Prove that the centralizer $Z_{S_n}(\sigma)$ is isomorphic to $S_{[n/2]} \ltimes (\mathbb{Z}/2\mathbb{Z})^{[n/2]}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Group actions and centralizers.
    \item Cycle structure of permutations.
    \item Semidirect products and wreath products.
\end{itemize}

\textbf{Intuition:}
The permutation $\sigma$ is a collection of disjoint "swaps" (transpositions). To commute with $\sigma$, a permutation must either keep a swap as it is, flip the two elements within a swap, or move one entire swap to the position of another swap. This "shuffling of pairs" combined with "flipping within pairs" perfectly describes the semidirect product structure.

\par

\textbf{Steps:}

1. \textbf{Analyze $\sigma$:} Let $k = [n/2]$. $\sigma$ consists of $k$ disjoint transpositions $B_j = \{j, n-j+1\}$. If $n$ is odd, there is one fixed point $k+1$.

\par

2. \textbf{Condition for Centralizer:} $g \in S_n$ commutes with $\sigma$ if it preserves the set of its cycles. This means $g$ must permute the $k$ sets $B_j$.

\par

3. \textbf{Defining the Homomorphism:} Define a map $\phi: Z_{S_n}(\sigma) \to S_k$ that tracks how $g$ permutes the indices of the blocks $B_j$.

\par

4. \textbf{Kernel of the map:} The kernel consists of elements that map each $B_j$ to itself. For each $B_j = \{x, y\}$, $g$ can either be identity or the swap $(x, y)$. This gives $k$ independent choices of $\mathbb{Z}/2\mathbb{Z}$, so the kernel is $(\mathbb{Z}/2\mathbb{Z})^k$.

\par

5. \textbf{Splitting the sequence:} Any permutation of the blocks can be realized by a permutation that preserves the relative order within blocks. This provides a section (a "lift") from $S_k$ back to the centralizer, showing the group is a semidirect product $S_k \ltimes (\mathbb{Z}/2\mathbb{Z})^k$.
\end{solution}

\begin{exercise}
Recall that the algebra of regular functions on a vector space $W$ is the symmetric algebra of linear forms on $W$.

Let $V = \mathbb{C}^2$. Let $\mathbb{C}[\mathrm{End}_{\mathbb{C}}(V)]$ be the algebra of regular functions on $\mathrm{End}_{\mathbb{C}}(V)$. The natural action of the group $G = SL_2(\mathbb{C})$ on $V$ induces an action of $G \times G$ on $\mathrm{End}_{\mathbb{C}}(V)$ by left and right multiplication. Thus we get an action of $G \times G$ on $\mathbb{C}[\mathrm{End}_{\mathbb{C}}(V)]$.

Compute the algebra of fixed points $\mathbb{C}[\mathrm{End}_{\mathbb{C}}(V)]^{G \times G}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Representation theory of $SL_2(\mathbb{C})$.
    \item Symmetric algebras and invariant theory.
    \item Cauchy Identity for Schur modules.
    \item Definition of the determinant.
\end{itemize}

\textbf{Intuition:}
We are looking for functions on $2 \times 2$ matrices that don't change when we multiply the matrix by $SL_2$ matrices on either side. Since $SL_2$ matrices have determinant 1, we know the determinant of the original matrix is a likely candidate. Invariant theory confirms that in this case, the determinant is the only fundamental building block.

\par

\textbf{Steps:}

1. \textbf{Decompose the space:} $\mathrm{End}(V)$ as a $G \times G$ module is $V \otimes V^*$. For $SL_2$, $V \cong V^*$, so this is $V \otimes V$.

\par

2. \textbf{Use Cauchy Identity:} The symmetric algebra $S^*(V \otimes V)$ decomposes into a sum of $S^\lambda V \otimes S^\lambda V$ over partitions $\lambda$.

\par

3. \textbf{Find Invariants:} An invariant exists if the representation $S^\lambda V$ contains the trivial representation. For $SL_2$, the irreducible $S^\lambda V$ is trivial only if $\lambda$ consists of equal-sized rows (e.g., $\lambda = (k, k)$).

\par

4. \textbf{Identify the generator:} For $k=1$, $\lambda=(1,1)$, which corresponds to the determinant function. Since this is the only invariant in its degree and others are powers, the algebra is generated by $\det$.

\par

5. \textbf{Result:} The algebra of fixed points is $\mathbb{C}[\det]$.
\end{solution}

\begin{exercise}
Let $R$ be a Noetherian ring and $I \subset R$ be an ideal. Define the Rees algebra as
\[\mathrm{Rees}(I, R) := \bigoplus_{n \geq 0} I^n t^n \subset R[t].\]
Prove that $\mathrm{Rees}(I, R)$ is Noetherian.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Definition of Noetherian rings and ideals.
    \item Hilbert Basis Theorem: If $R$ is Noetherian, then $R[x]$ is Noetherian.
    \item Property: Any quotient of a Noetherian ring is Noetherian.
\end{itemize}

\textbf{Intuition:}
The Rees algebra is a way to "track" the powers of an ideal using a formal variable $t$. Since the original ring is Noetherian, the ideal $I$ is finitely generated. We show that the entire Rees algebra is also finitely generated as an algebra, which allows us to use the Hilbert Basis Theorem.

\par

\textbf{Steps:}

1. \textbf{Identify Generators:} Since $R$ is Noetherian, $I$ is finitely generated, say $I = (a_1, \dots, a_k)$.

\par

2. \textbf{Construct the Algebra Map:} The Rees algebra $S = \bigoplus I^n t^n$ is generated over $R$ by the elements $\{a_1 t, a_2 t, \dots, a_k t\}$. This is because any element in $I^n$ is a sum of products of the $a_i$'s.

\par

3. \textbf{Use Hilbert Basis Theorem:} Since $R$ is Noetherian, the polynomial ring $R[x_1, \dots, x_k]$ is Noetherian.

\par

4. \textbf{Conclusion:} There is a surjective homomorphism from $R[x_1, \dots, x_k]$ to $S$ sending $x_i \mapsto a_i t$. Since $S$ is a quotient of a Noetherian ring, $S$ is Noetherian.
\end{solution}

\begin{exercise}
1. Show that $\Phi_p/\mathbb{Q}$ is a Galois extension with Galois group $(\mathbb{Z}/p\mathbb{Z})^\times$.
2. Deduce that $\Phi_p$ contains a unique quadratic extension of $\mathbb{Q}$.
3. Write $g_p$ for the Gauss sum $\sum_{a=1}^{p-1} \left(\frac{a}{p}\right) \zeta_p^a$. Show $\overline{g_p} = \left(\frac{-1}{p}\right) g_p$.
4. Determine the unique quadratic extension of $\mathbb{Q}$ contained in $\Phi_p$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Cyclotomic fields and polynomials.
    \item Fundamental Theorem of Galois Theory.
    \item Properties of Gauss sums and Legendre symbols.
\end{itemize}

\textbf{Intuition:}
The Galois group of a cyclotomic field is cyclic. In cyclic groups, subgroups are unique for each divisor of the order. Thus, there's only one subgroup of index 2, which corresponds to exactly one quadratic subfield. The Gauss sum $g_p$ is a "bridge" that lives in the cyclotomic field but whose square is an integer, making it the generator of this subfield.

\par

\textbf{Steps:}

1. \textbf{Galois Group:} The $p$-th cyclotomic polynomial is irreducible of degree $p-1$. Its roots are the primitive $p$-th roots of unity. The automorphisms are $\zeta \mapsto \zeta^a$, giving the group $(\mathbb{Z}/p\mathbb{Z})^\times$.

\par

2. \textbf{Uniqueness of Quadratic Extension:} The group $(\mathbb{Z}/p\mathbb{Z})^\times$ is cyclic of order $p-1$. A cyclic group has exactly one subgroup of any given index. Index 2 corresponds to the quadratic subfield.

\par

3. \textbf{Gauss Sum Conjugation:} Conjugation $\zeta \mapsto \zeta^{-1}$ sends $g_p = \sum \left(\frac{a}{p}\right) \zeta^a$ to $\sum \left(\frac{a}{p}\right) \zeta^{-a}$. Using $\left(\frac{a}{p}\right) = \left(\frac{-1}{p}\right) \left(\frac{-a}{p}\right)$, we see $\overline{g_p} = \left(\frac{-1}{p}\right) g_p$.

\par

4. \textbf{Identify Subfield:} It is known $g_p^2 = \left(\frac{-1}{p}\right) p$. Thus $g_p = \sqrt{(-1)^{(p-1)/2} p}$. This element generates the quadratic extension $\mathbb{Q}(\sqrt{p^*})$, where $p^* = (-1)^{(p-1)/2} p$.
\end{solution}

\begin{exercise}
1. Let $E/F$ be a finite Galois extension with cyclic Galois group generated by $\sigma$. Let $\mathrm{tr}_{E/F}(x) = 0$. Show $x = \sigma(y) - y$ for some $y$.
2. Let $F$ have characteristic $p$, and $E/F$ be Galois of degree $p$. Show $E \cong F[T]/(T^p - T - x)$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Additive Hilbert's Theorem 90.
    \item Trace map in field extensions.
    \item Artin-Schreier theory.
\end{itemize}

\textbf{Intuition:}
Part 1 is a classic result: an element of trace zero in a cyclic extension can always be written as a "difference" under the generator. Part 2 uses this to construct a special element $\alpha$ that shifts by 1 under the Galois action, which forces its minimal polynomial to have the $T^p-T$ form.

\par

\textbf{Steps:}

1. \textbf{Additive Hilbert 90:} For trace zero $x$, we construct $y$ explicitly using the linear independence of characters. A specific sum $y = \sum_{j=0}^{n-1} (\sum_{i=0}^j \sigma^i(x)) \sigma^j(z)$ satisfies $\sigma(y)-y=x$.

\par

2. \textbf{Constructing the Artin-Schreier element:} In characteristic $p$, $\mathrm{tr}(1) = p \cdot 1 = 0$. Thus there exists $\alpha$ such that $\sigma(\alpha) - \alpha = 1$.

\par

3. \textbf{Minimal Polynomial:} The values $\alpha, \alpha+1, \dots, \alpha+p-1$ are the $p$ distinct conjugates of $\alpha$. The product $(T-\alpha)(T-(\alpha+1))\dots(T-(\alpha+p-1))$ simplifies to $T^p-T - (\alpha^p-\alpha)$.

\par

4. \textbf{Conclusion:} Let $x = \alpha^p-\alpha$. Since $\sigma(x)=x$, $x \in F$. Thus $E = F(\alpha)$ is the splitting field of $T^p-T-x$.
\end{solution}
