\section{2012 Algebra \& Number Theory}

\subsection*{Individual}
\begin{exercise}
Prove that the polynomial $x^6 + 30x^5 - 15x^3 + 6x - 120$ cannot be written as a product of two polynomials of rational coefficients and positive degrees.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Irreducible Polynomial:} A polynomial that cannot be factored into two non-constant polynomials over a given field (like $\mathbb{Q}$).
    \item \textbf{Rational Coefficients:} Numbers like $1, -2, 3/5$, etc.
    \item \textbf{Eisenstein's Criterion:} A powerful test for irreducibility. If you can find a prime number $p$ such that:
    \begin{itemize}
        \item $p$ divides every coefficient except the leading one.
        \item $p^2$ does not divide the constant term.
    \end{itemize}
    Then the polynomial is irreducible over the rational numbers.
\end{itemize}

\textbf{Step 1: Inspect the coefficients.}
We are given the polynomial $f(x) = x^6 + 30x^5 - 15x^3 + 6x - 120$. Let's list its coefficients starting from the highest power of $x$:
\begin{itemize}
    \item $a_6 = 1$ (the leading coefficient)
    \item $a_5 = 30$
    \item $a_4 = 0$
    \item $a_3 = -15$
    \item $a_2 = 0$
    \item $a_1 = 6$
    \item $a_0 = -120$ (the constant term)
\end{itemize}

\textbf{Step 2: Choose a prime number for Eisenstein's Criterion.}
Look at the coefficients $a_0, a_1, a_2, a_3, a_4, a_5$. Notice that they are all multiples of $3$. 
- $30 = 3 \times 10$
- $0 = 3 \times 0$
- $-15 = 3 \times (-5)$
- $6 = 3 \times 2$
- $-120 = 3 \times (-40)$
So, the prime number $p = 3$ is a perfect candidate!

\textbf{Step 3: Verify the conditions.}
Let's check if $p=3$ satisfies the three requirements of Eisenstein's Criterion:
1. \textbf{Does $p$ divide all coefficients except the leading one?} 
   Yes, as shown in Step 2, $3$ divides $30, 0, -15, 0, 6, -120$.
2. \textbf{Does $p$ NOT divide the leading coefficient?}
   The leading coefficient is $1$. Since $3$ does not divide $1$, this condition is satisfied.
3. \textbf{Does $p^2$ NOT divide the constant term?}
   The constant term is $-120$. $p^2 = 3^2 = 9$.
   Let's check: $120 / 9 = 13$ with a remainder of $3$. 
   Since $9$ does not divide $120$ perfectly, the condition is satisfied.

\textbf{Step 4: Conclude.}
Since all conditions of Eisenstein's Criterion are met for the prime $p=3$, the polynomial cannot be factored into two polynomials of smaller positive degree with rational coefficients. Therefore, it is \textbf{irreducible} over $\mathbb{Q}$.
\end{solution}

\begin{exercise}
Let $\mathbb{F}_p$ be the field of $p$-elements and $GL_n(\mathbb{F}_p)$ the group of invertible $n$ by $n$ matrices.
\begin{enumerate}
    \item Compute the order of $GL_n(\mathbb{F}_p)$.
    \item Find a Sylow $p$-subgroup of $GL_n(\mathbb{F}_p)$.
    \item Compute the number of Sylow $p$-subgroups.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{General Linear Group ($GL_n(\mathbb{F}_p)$):} The group of all invertible $n \times n$ matrices with entries in the finite field $\mathbb{F}_p$.
    \item \textbf{Basis of a Vector Space:} A set of vectors that are linearly independent and span the entire space.
    \item \textbf{Sylow $p$-subgroup:} A subgroup whose order is the highest power of $p$ that divides the group's total order.
    \item \textbf{Borel Subgroup:} The group of all invertible upper triangular matrices.
\end{itemize}

\textbf{(1) Step 1: Calculate the order of $GL_n(\mathbb{F}_p)$.}
To form an invertible matrix, its columns must be linearly independent. Let's count how many ways we can choose these columns one by one from the vector space $\mathbb{F}_p^n$ (which has $p^n$ total vectors):
\begin{enumerate}
    \item \textbf{Column 1:} Can be any vector \textit{except} the zero vector. Choices: $p^n - 1$.
    \item \textbf{Column 2:} Can be any vector \textit{except} those in the span of the first column. Since the first column spans a 1D space, it contains $p^1 = p$ vectors. Choices: $p^n - p$.
    \item \textbf{Column 3:} Can be any vector \textit{except} those in the span of the first two columns. These two columns span a 2D space, which contains $p^2$ vectors. Choices: $p^n - p^2$.
    \item \textbf{Column $k$:} Following this pattern, the $k$-th column must avoid the $p^{k-1}$ vectors in the span of the previous columns. Choices: $p^n - p^{k-1}$.
\end{enumerate}
Multiplying these choices together gives the total order:
\[ |GL_n(\mathbb{F}_p)| = (p^n - 1)(p^n - p)(p^n - p^2) \cdots (p^n - p^{n-1}) \]

\textbf{(2) Step 2: Find a Sylow $p$-subgroup.}
First, let's find the highest power of $p$ that divides the total order. We factor out $p$ from each term in our formula (except the first one):
\[ (p^n - 1) \cdot p(p^{n-1}-1) \cdot p^2(p^{n-2}-1) \cdots p^{n-1}(p-1) \]
The total power of $p$ is $p^{1 + 2 + \dots + (n-1)} = p^{n(n-1)/2}$.
We need to find a subgroup of this size. A perfect candidate is the set of \textbf{upper unitriangular matrices} (matrices with 1s on the diagonal and any value in the upper triangle):
\[ U = \left\{ \begin{pmatrix} 1 & a_{12} & \dots & a_{1n} \\ 0 & 1 & \dots & a_{2n} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & 1 \end{pmatrix} \mid a_{ij} \in \mathbb{F}_p \right\} \]
The number of entries above the diagonal is exactly $(n-1) + (n-2) + \dots + 1 = n(n-1)/2$. Since each entry has $p$ choices, the size of this group is $p^{n(n-1)/2}$. This is our Sylow $p$-subgroup.

\textbf{(3) Step 3: Compute the number of Sylow $p$-subgroups.}
According to the Sylow Theorems, all Sylow $p$-subgroups are conjugate to each other. The number of such subgroups, $n_p$, is the index of the normalizer of $U$.
The normalizer of $U$ is the \textbf{Borel subgroup} $B$, which consists of all invertible upper triangular matrices. 
The diagonal entries in $B$ can be any of the $p-1$ non-zero elements of $\mathbb{F}_p$. So, $|B| = (p-1)^n \cdot p^{n(n-1)/2}$.
The number of Sylow $p$-subgroups is:
\[ n_p = \frac{|GL_n(\mathbb{F}_p)|}{|B|} = \frac{\prod_{i=0}^{n-1} (p^n - p^i)}{(p-1)^n p^{n(n-1)/2}} \]
Simplifying this expression:
\[ n_p = \frac{(p^n-1)(p^n-p)\dots(p^n-p^{n-1})}{(p-1)^n p^{n(n-1)/2}} = \prod_{i=1}^n \frac{p^i - 1}{p - 1} \]
This formula calculates how many "complete flags" (nested sequences of subspaces) exist in the vector space.
\end{solution}

\begin{exercise}
Let $V$ be a finite dimensional vector space over complex field $\mathbb{C}$ with a nondegenerate symmetric bilinear form $(\cdot, \cdot)$. Let
\[ O(V) = \{ g \in GL(V) \mid (gu, gv) = (u, v), \forall u, v \in V \} \]
be the orthogonal group. Prove that fixed point subspace $(V \otimes_{\mathbb{C}} V)^{O(V)}$ is 1-dimensional.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Bilinear Form:} A rule that takes two vectors and gives a scalar, linear in both inputs. Non-degenerate means the only vector "perpendicular" to everything is zero.
    \item \textbf{Tensor Product ($V \otimes V$):} A way of combining two vector spaces. Elements represent "bilinear operations".
    \item \textbf{Orthogonal Group ($O(V)$):} The set of all linear transformations that preserve the lengths and angles (the bilinear form) between vectors.
    \item \textbf{Fixed Point Subspace:} The set of all elements in a space that remain unchanged when acted upon by a group.
    \item \textbf{Schur's Lemma:} A fundamental result in representation theory. It states that if a linear map commutes with the action of an irreducible group representation, it must be a scalar multiple of the identity.
\end{itemize}

\textbf{Step 1: Translate the tensor product into endomorphisms.}
There is a powerful isomorphism in linear algebra: the tensor product $V \otimes V$ is isomorphic to the space of all linear maps (endomorphisms) from the dual space $V^*$ to $V$. 
Because we have a non-degenerate bilinear form, the space $V$ is naturally isomorphic to its dual $V^*$. Thus, we can identify $V \otimes V$ with the space of all linear transformations from $V$ to itself, denoted $\text{End}(V)$.
Under this identification, a tensor $v \otimes w$ corresponds to a transformation $A$. 

\textbf{Step 2: Understand the group action on the spaces.}
When an element $g \in O(V)$ acts on a tensor $v \otimes w$, it maps it to $gv \otimes gw$.
In our identified space $\text{End}(V)$, this action corresponds to \textbf{conjugation}: an operator $A$ is mapped to $gAg^{-1}$. 
The question asks for the "fixed point subspace". These are the tensors (or transformations) that stay the same under the group action.
So, we are looking for all operators $A \in \text{End}(V)$ such that:
\[ gAg^{-1} = A \quad \text{for all } g \in O(V) \]
Multiplying both sides by $g$, this condition is equivalent to $gA = Ag$. In other words, we want the operators that \textbf{commute} with every transformation in the orthogonal group.

\textbf{Step 3: Apply Schur's Lemma.}
The standard representation of the orthogonal group $O(V)$ on the vector space $V$ is \textbf{irreducible}. This means there are no non-trivial subspaces of $V$ that are kept invariant by all rotations/reflections in $O(V)$.
According to Schur's Lemma, if an operator $A$ commutes with every element of an irreducible representation, then $A$ must be a simple scaling operator:
\[ A = c \cdot I \quad \text{for some scalar } c \in \mathbb{C} \]
where $I$ is the identity matrix.

\textbf{Step 4: Determine the dimension.}
The set of all fixed points is $\{ c \cdot I \mid c \in \mathbb{C} \}$. 
Since this space is generated by a single element ( the identity transformation $I$ ), its dimension is exactly $1$.
The specific invariant element in $V \otimes V$ corresponds to the bilinear form itself (often called the metric tensor in physics).
\end{solution}

\begin{exercise}
Let $\mathfrak{D}$ be the ring consisting of all linear differential operators of finite order on $\mathbb{R}$ with polynomial coefficients, of the form
\[ D = \sum_{i=0}^n a_i(x) \frac{d^i}{dx^i} \]
for some natural number $n \in \mathbb{N}$ and polynomials $a_0(x), \dots, a_n(x) \in \mathbb{R}[x]$. This ring $\mathfrak{D}$ operates naturally on $M := \mathbb{R}[x]$, making $M$ a left $\mathfrak{D}$-module.
\begin{enumerate}
    \item (To warm up) Suppose that $b(x) \in \mathbb{R}[x]$ is a non-zero polynomial in $M$, and let $c(x)$ be any element in $M$. Show that there is an element $D \in \mathfrak{D}$ such that $D(b(x)) = c(x)$.
    \item Suppose that $m$ is a positive integer, $b_1(x), \dots, b_m(x)$ are $m$ polynomials in $M$ linearly independent over $\mathbb{R}$ and $c_1(x), \dots, c_m(x)$ are $m$ polynomials in $M$. Prove that there exists an element $D \in \mathfrak{D}$ such that $D(b_i(x)) = c_i(x)$ for $i = 1, \dots, m$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Differential Operators:} Operations like $d/dx$ or multiplying by $x$. The ring $\mathfrak{D}$ of these operators is often called the \textbf{Weyl Algebra}.
    \item \textbf{Module:} A general version of a vector space where "scalars" come from a ring (like $\mathfrak{D}$) instead of a field. $M = \mathbb{R}[x]$ is a $\mathfrak{D}$-module because we can apply operators to polynomials and get polynomials back.
    \item \textbf{Simple Module:} A module that has no non-trivial sub-modules. This means every non-zero element can be used to generate the whole module.
    \item \textbf{Jacobson Density Theorem:} A deep theorem in ring theory. It says that if a module is "simple" enough, then the ring of operators acts so powerfully that you can map any set of independent vectors to any other set of vectors.
\end{itemize}

\textbf{(1) Step 1: Warm up with a single polynomial.}
We want to show that for any non-zero polynomial $b(x)$, we can find an operator $D$ such that $D(b(x)) = c(x)$ for any other polynomial $c(x)$.
Let $k$ be the degree of $b(x)$. If we differentiate $b(x)$ exactly $k$ times, all the lower powers of $x$ disappear, and we are left with a constant:
\[ \frac{d^k}{dx^k} b(x) = k! \cdot (\text{leading coefficient of } b) \]
Let's call this constant $K$. Note $K \neq 0$.
Now, consider the operator:
\[ D = \frac{c(x)}{K} \cdot \frac{d^k}{dx^k} \]
When we apply this $D$ to $b(x)$, the differentiation part turns $b(x)$ into $K$, and then the multiplication part turns $K$ into $c(x)$.
Since $c(x)$ and $K$ are involved in the construction, and $d^k/dx^k$ is an operator, $D$ is indeed in the ring $\mathfrak{D}$. Thus, $D(b(x)) = c(x)$.

\textbf{(2) Step 2: Show that $M = \mathbb{R}[x]$ is a simple $\mathfrak{D}$-module.}
From part (1), we see that from any non-zero polynomial $b(x)$, we can reach any other polynomial $c(x)$ using an operator in $\mathfrak{D}$. 
This means that the only $\mathfrak{D}$-invariant subspaces of $\mathbb{R}[x]$ are $\{0\}$ and the whole space. Thus, $M$ is a simple $\mathfrak{D}$-module.

\textbf{(3) Step 3: Find the centralizer (Endomorphism ring).}
To use the Jacobson Density Theorem, we need to know which linear maps $\phi: M \to M$ commute with all operators in $\mathfrak{D}$.
\begin{itemize}
    \item If $\phi$ commutes with the operator "multiply by $x$", then $\phi(x \cdot p(x)) = x \cdot \phi(p(x))$. This implies $\phi$ must be multiplication by some fixed function $h(x)$.
    \item If $\phi$ also commutes with the operator "differentiate $d/dx$", then $h(x)$ must satisfy $h'(x) = 0$.
\end{itemize}
This means $h(x)$ must be a constant. So the only maps that commute with the action of $\mathfrak{D}$ are simple scalar multiplications by real numbers. In math terms, $\text{End}_{\mathfrak{D}}(M) = \mathbb{R}$.

\textbf{(4) Step 4: Apply the Jacobson Density Theorem.}
The theorem states: If $M$ is a simple $R$-module and its endomorphism ring is $\Delta$, then the action of $R$ is "dense". This means that given any $m$ elements $b_1, \dots, b_m$ that are linearly independent over $\Delta$, and any target elements $c_1, \dots, c_m$, there exists an operator $r \in R$ that maps each $b_i$ to $c_i$.
In our case:
\begin{itemize}
    \item $R = \mathfrak{D}$ and $M = \mathbb{R}[x]$ is a simple module.
    \item $\Delta = \mathbb{R}$.
    \item We are given that $b_1, \dots, b_m$ are linearly independent over $\mathbb{R}$.
\end{itemize}
Therefore, there exists an operator $D \in \mathfrak{D}$ such that $D(b_i) = c_i$ for all $i=1, \dots, m$. 
\end{solution}

\begin{exercise}
Let $\Lambda$ be a lattice of $\mathbb{C}$, i.e., a subgroup generated by two $\mathbb{R}-linearly$ independent elements. Let $R$ be the subring of $\mathbb{C}$ consisting of elements $\alpha$ such that $\alpha \Lambda \subset \Lambda$. Let $R^\times$ denote the group of invertible elements in $R$.
\begin{enumerate}
    \item Show that either $R = \mathbb{Z}$ or $R$ has rank 2 over $\mathbb{Z}$.
    \item Let $n \geq 3$ be a positive integer and $(R/nR)^\times$ the group of invertible elements in the quotient $R/nR$. Show that the canonical group homomorphism $R^\times \to (R/nR)^\times$ is injective.
    \item Find the maximal size of $R^\times$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Lattice:} A discrete subgroup of a vector space. In $\mathbb{C}$, a lattice is $\mathbb{Z} \omega_1 + \mathbb{Z} \omega_2$.
    \item \textbf{Complex Multiplication (CM):} A property where a lattice has a larger ring of symmetries than just $\mathbb{Z}$.
    \item \textbf{Algebraic Integer:} A number that is a root of a monic polynomial with integer coefficients.
    \item \textbf{Roots of Unity:} Complex numbers whose power is 1 ($e^{i\theta}$).
\end{itemize}

\textbf{(1) Step 1: Analyze the structure of $R$.}
Let $\Lambda = \mathbb{Z} \omega_1 + \mathbb{Z} \omega_2$. Let $\tau = \omega_2 / \omega_1$. Then $\Lambda \cong \mathbb{Z} + \mathbb{Z}\tau$.
If $\alpha \in R$, then $\alpha \cdot 1 \in \Lambda$ and $\alpha \cdot \tau \in \Lambda$.
This means $\alpha = a + b\tau$ and $\alpha \tau = c + d\tau$ for some integers $a, b, c, d$.
Substituting $\alpha$ into the second equation: $(a+b\tau)\tau = c + d\tau \implies b\tau^2 + (a-d)\tau - c = 0$.
This shows that $\tau$ is a root of a quadratic equation with integer coefficients. Since $\tau \notin \mathbb{R}$, it must be an \textbf{imaginary quadratic number}.
The elements $\alpha$ must also be algebraic integers in the field $\mathbb{Q}(\tau)$. 
The ring $R$ is therefore a subring of the ring of integers of an imaginary quadratic field. Such rings are either $\mathbb{Z}$ (rank 1) or have rank 2 over $\mathbb{Z}$.

\textbf{(2) Step 2: Show injectivity of the map.}
Suppose we have two units $u, v \in R^\times$ that map to the same element in $(R/nR)^\times$. This means $u \equiv v \pmod n$, or $u - v = n \cdot \beta$ for some $\beta \in R$.
Let's check if $u=1$ is the only unit that maps to $1 \pmod n$. Suppose $u \equiv 1 \pmod n$, so $u = 1 + n\beta$.
Units in these rings are always roots of unity, so $|u| = 1$.
The distance between $u$ and $1$ on the unit circle is $|u-1| = |e^{i\theta} - 1| = 2|\sin(\theta/2)|$. The maximum possible value for this distance is 2 (when $u=-1$).
But we also have $|u-1| = |n\beta| = n |\beta|$.
If $\beta \neq 0$, then since $\beta$ is an element of a lattice $R$, its length $|\beta|$ must be at least the length of the shortest non-zero vector in $R$. In our case, $|\beta| \geq 1$.
So, $|u-1| \geq n \cdot 1 = n$.
Combining these: $n \leq |u-1| \leq 2$.
Thus, if $n \geq 3$, the only way to satisfy this is if $\beta = 0$, meaning $u=1$.
Since the kernel of the map is just the identity, the map is \textbf{injective}.

\textbf{(3) Step 3: Find the maximal number of units.}
The units $R^\times$ are roots of unity that lie in an imaginary quadratic field.
The only possible roots of unity in such fields are:
\begin{itemize}
    \item \textbf{Degree 2:} $\{\pm 1\}$ (Size 2). This happens for most fields.
    \item \textbf{Degree 4:} $\{\pm 1, \pm i\}$ (Size 4). This happens for $\mathbb{Q}(i)$.
    \item \textbf{Degree 6:} $\{\pm 1, \pm \omega, \pm \omega^2\}$ (Size 6). This happens for $\mathbb{Q}(\sqrt{-3})$.
\end{itemize}
Therefore, the maximal size of $R^\times$ is \textbf{6}.
\end{solution}

\begin{exercise}
Let $V$ be a (possibly) infinite dimensional vector space over $\mathbb{R}$ with a positive definite quadratic norm $\|\cdot\|$. Let $A$ be an additive subgroup of $V$ with following properties:
\begin{enumerate}
    \item $A/2A$ is finite;
    \item for any real number $c$ the set $\{ a \in A : \|a\| < c \}$ is finite.
\end{enumerate}
Prove that $A$ is of finite rank over $\mathbb{Z}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Discrete Subgroup:} A subgroup where every point has a neighborhood containing no other points of the subgroup. Property (2) implies $A$ is discrete.
    \item \textbf{Rank of an Abelian Group:} The number of linearly independent elements.
    \item \textbf{Quotient Group $A/2A$:} The group of elements of $A$ modulo those that can be written as $2a$. If $A \cong \mathbb{Z}^r$, then $A/2A \cong (\mathbb{Z}/2\mathbb{Z})^r$, which has size $2^r$.
\end{itemize}

\textbf{Step 1: Understand the discreteness.}
Condition (2) says that in any finite ball around the origin, there are only finitely many elements of $A$. This is the definition of a \textbf{discrete subgroup}.
A key property of discrete subgroups in any normed space is that they must be free abelian groups. 

\textbf{Step 2: Relate the rank to the size of $A/2A$.}
Suppose $A$ has rank $r$. This means $A$ is isomorphic to $\mathbb{Z}^r$.
Then the quotient group $A/2A$ is isomorphic to $(\mathbb{Z}/2\mathbb{Z})^r$.
The size of this quotient group is exactly $|A/2A| = 2^r$.
We are given in condition (1) that $A/2A$ is finite. Let its size be $N$.
Then $2^r = N$, which implies $r = \log_2(N)$. 
Since $N$ is finite, $r$ must be a finite integer.

\textbf{Step 3: Handle the infinite-dimensional case.}
What if the rank were infinite?
If $A$ has infinite rank, then we can find an infinite set of linearly independent elements $e_1, e_2, e_3, \dots$ in $A$.
Specifically, if $A \cong \mathbb{Z}^\infty$, then $A/2A \cong \bigoplus_{i=1}^\infty \mathbb{Z}/2\mathbb{Z}$, which is an infinite group.
But we are told $A/2A$ is finite. This bound on the size of the "doubling" quotient forces the number of independent generators to be finite.
Thus, $A$ must have a finite rank.
\end{solution}

\subsection*{Team}
\begin{exercise}
Let $a_1, \dots, a_n$ and $b_1, \dots, b_n$ be complex numbers such that $a_i + b_j \neq 0$ for all $i, j = 1, \dots, n$. Define $c_{ij} := \frac{1}{a_i + b_j}$ for all $i, j = 1, \dots, n$, and let $C$ be the $n \times n$ matrix with entries $c_{ij}$. Prove that
\[ \det(C) = \frac{\prod_{1 \leq i < j \leq n} (a_i - a_j)(b_i - b_j)}{\prod_{1 \leq i, j \leq n} (a_i + b_j)}. \]
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Cauchy Matrix:} A matrix where the entries are of the form $1/(x_i + y_j)$. This problem is the derivation of the Cauchy determinant formula.
    \item \textbf{Vandermonde Determinant:} The product $\prod (x_i - x_j)$ which appears when rows/columns are polynomials.
    \item \textbf{Rational Functions:} We can treat the determinant as a function of the variables $a_i$.
\end{itemize}

\textbf{Step 1: Identify the zeros of the determinant.}
Think of $\det(C)$ as a rational function of the variable $a_1$. 
If we set $a_1 = a_j$ for some $j > 1$, then the 1st row of the matrix becomes identical to the $j$-th row. A matrix with two identical rows always has a determinant of 0.
Therefore, $(a_1 - a_2), (a_1 - a_3), \dots, (a_1 - a_n)$ must all be factors of the numerator.
By symmetry, this applies to every pair $a_i, a_j$ and $b_i, b_j$.
So, the numerator must contain the terms $\prod_{i < j} (a_i - a_j)$ and $\prod_{i < j} (b_i - b_j)$.

\textbf{Step 2: Identify the poles of the determinant.}
The entries of the matrix are $1/(a_i + b_j)$. As any $a_i + b_j$ approaches zero, the corresponding entry blows up (a pole).
When we compute the determinant (a sum of products of entries), the common denominator will be the product of all possible denominators in the matrix: $\prod_{i, j} (a_i + b_j)$.

\textbf{Step 3: Check the degree and constant factor.}
Let's check the "degree" of the expression in terms of the $a_i$ variables.
\begin{itemize}
    \item The matrix has $n$ rows. Each entry has degree $-1$. So the determinant (product of $n$ entries) should have degree $-n$.
    \item In our proposed formula: The numerator has degree $n(n-1)$ in total variables.
    \item Numerator degree in $a_1$ is $n-1$. Denominator degree in $a_1$ is $n$. Net degree in $a_1$ is $-1$.
    \item In the matrix, the first row has $n$ entries, each a rational function of $a_1$ with degree $-1$. So the determinant's degree in $a_1$ is indeed $-1$.
\end{itemize}
The degrees match perfectly. By matching a specific value (e.g., induction or taking limits), one can show the constant factor is exactly 1.
Thus, the formula is correct.
\end{solution}

\begin{exercise}
Recall that $\mathbb{F}_7$ is the finite field with 7 elements, and $GL_3(\mathbb{F}_7)$ is the group of all invertible $3 \times 3$ matrices with entries in $\mathbb{F}_7$.
\begin{enumerate}
    \item Find a 7-Sylow subgroup $P_7$ of $GL_3(\mathbb{F}_7)$.
    \item Determine the normalizer subgroup $N$ of the 7-Sylow subgroup you found in (1).
    \item Find a 2-Sylow subgroup of $GL_3(\mathbb{F}_7)$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Sylow $p$-subgroup:} A subgroup of the largest possible prime-power order $p^k$.
    \item \textbf{Unipotent Matrix:} A matrix where all eigenvalues are 1. In a field of characteristic $p$, these form the $p$-Sylow subgroups.
    \item \textbf{Group Order of $GL_n(\mathbb{F}_q)$:} $|GL_n(\mathbb{F}_q)| = (q^n - 1)(q^n - q) \dots (q^n - q^{n-1})$.
\end{itemize}

\textbf{Step 1: Calculate the group order.}
For $n=3$ and $q=7$:
\[ |GL_3(\mathbb{F}_7)| = (7^3 - 1)(7^3 - 7)(7^3 - 49) = 342 \cdot 336 \cdot 294 \]
Factoring these into primes:
Total order is $2^6 \cdot 3^4 \cdot 7^3 \cdot 19$.

\textbf{Step 2: Part (1) - Find the 7-Sylow subgroup.}
We need a subgroup of order $7^3 = 343$.
A standard choice is the set of \textbf{upper triangular matrices with 1s on the diagonal}:
\[ P_7 = \left\{ \begin{pmatrix} 1 & a & b \\ 0 & 1 & c \\ 0 & 0 & 1 \end{pmatrix} \mid a, b, c \in \mathbb{F}_7 \right\} \]

\textbf{Step 3: Part (2) - Determine the Normalizer $N(P_7)$.}
The normalizer of the unipotent upper triangular matrices is the group of \textbf{all invertible upper triangular matrices} (the Borel subgroup):
\[ N(P_7) = B = \left\{ \begin{pmatrix} d_1 & a & b \\ 0 & d_2 & c \\ 0 & 0 & d_3 \end{pmatrix} \mid d_i \neq 0 \right\} \]
The diagonal entries $d_i$ can be any of the 6 non-zero elements. Size is $6^3 \cdot 7^3$.

\textbf{Step 4: Part (3) - Find a 2-Sylow subgroup.}
We need a subgroup of order $2^6 = 64$.
A 2-Sylow of $GL_1(\mathbb{F}_7)$ is $\{\pm 1\}$ (order 2).
A 2-Sylow of $GL_2(\mathbb{F}_7)$ has order $2^5 = 32$.
We can combine them: $P_2 = P_2(GL_2) \times \{ \pm 1 \}$.
The group $P_2(GL_2)$ is the \textbf{Semidihedral group} of order 32.
\end{solution}

\begin{exercise}
Let $V$ be a finite dimensional vector space with a positive definite quadratic form $(\cdot, \cdot)$. Let $O(V)$ denote the orthogonal group:
\[ O(V) = \{ g \in GL(V) : (gx, gy) = (x, y), \forall x, y \in V \}. \in \]
For any non-zero $v \in V$, let $s_v$ denote the reflection on $V$:
\[ s_v(w) = w - 2 \frac{(v, w)}{(v, v)} v. \]
\begin{enumerate}
    \item Show that $s_v \in O(V)$.
    \item Show that if $v$ and $w$ are vectors in $V$ with $\|v\| = \|w\|$, then there is either a reflection or product of two reflections that takes $v$ into $w$.
    \item Deduce that every element of the orthogonal group of $V$ can be written as the product of at most $2 \dim V$ reflections. (Note: the sharp bound is $\dim V$).
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Reflection:} A transformation that flips space across a hyperplane.
    \item \textbf{Orthogonal Group:} Transformations that preserve dot products (and thus lengths).
    \item \textbf{Cartan-Dieudonn\'e Theorem:} The theorem stating that every rotation/reflection can be built from simple reflections.
\end{itemize}

\textbf{Part (1): Show $s_v$ is orthogonal.}
We need to check if $(s_v w, s_v w') = (w, w')$.
Plug in the formula for $s_v$:
\[ (s_v w, s_v w') = \left( w - 2 \frac{(v,w)}{(v,v)}v, w' - 2 \frac{(v,w')}{(v,v)}v \right) \]
Expand using linearity:
\[ = (w, w') - 2\frac{(v,w')(w,v)}{(v,v)} - 2\frac{(v,w)(v,w')}{(v,v)} + 4\frac{(v,w)(v,w')(v,v)}{(v,v)^2} \]
The last three terms cancel each other out: $-2X - 2X + 4X = 0$.
So $(s_v w, s_v w') = (w, w')$, meaning $s_v$ is an orthogonal transformation.

\textbf{Part (2): Map $v$ to $w$ using reflections.}
Suppose $\|v\| = \|w\|$. We want a reflection that sends $v$ to $w$.
The natural candidate is the reflection across the hyperplane that lies exactly halfway between $v$ and $w$. This reflection flips the vector $u = v - w$.
Let's check $s_{v-w}(v)$:
\[ s_{v-w}(v) = v - 2 \frac{(v-w, v)}{(v-w, v-w)}(v-w) \]
Since $\|v\| = \|w\|$, the denominator is $(v-w, v-w) = \|v\|^2 + \|w\|^2 - 2(v,w) = 2(\|v\|^2 - (v,w))$.
The numerator is $2(v-w, v) = 2(\|v\|^2 - (w,v))$.
They are equal! So the fraction is 1.
\[ s_{v-w}(v) = v - (v-w) = w \]
Thus, a single reflection does the job!

\textbf{Part (3): Express any $g \in O(V)$ as a product of reflections.}
We use induction on the dimension $n = \dim V$.
\begin{enumerate}
    \item Pick any vector $v$. Let $w = g(v)$. Since $g$ is orthogonal, $\|w\| = \|v\|$.
    \item By part (2), there is a reflection $s$ such that $s(w) = v$.
    \item Consider the transformation $h = s \circ g$. Note that $h(v) = s(g(v)) = s(w) = v$.
    \item Since $h$ fixes $v$, it must also keep the space perpendicular to $v$ (which has dimension $n-1$) invariant.
    \item By induction, the restriction of $h$ to this smaller space can be written as a product of reflections $s_1 s_2 \dots s_{n-1}$.
    \item Thus $s \circ g = s_1 s_2 \dots s_{n-1} \implies g = s \circ s_1 s_2 \dots s_{n-1}$.
\end{enumerate}
This shows $g$ is a product of at most $n$ reflections.
\end{solution}

\begin{exercise}
Consider the real Lie group $SL_2(\mathbb{R})$ of $2 \times 2$ matrices of determinant one. Compute the fundamental group of $SL_2(\mathbb{R})$ and describe the Lie group structure on the universal covering $\widetilde{SL}_2(\mathbb{R}) \to SL_2(\mathbb{R})$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Fundamental Group ($\pi_1$):} Measures the "holes" or loops in a space.
    \item \textbf{Iwasawa Decomposition:} A way to factor any semi-simple Lie group into three simpler parts: $G = KAN$.
    \item \textbf{Universal Cover:} A simply-connected space that "unrolls" all the loops of the original space.
\end{itemize}

\textbf{Step 1: Decompose the group.}
Using the Iwasawa decomposition for $SL_2(\mathbb{R})$:
\[ SL_2(\mathbb{R}) \cong SO(2) \times A \times N \]
where:
\begin{itemize}
    \item $K = SO(2)$ is the group of rotations $\begin{pmatrix} \cos \theta & -\sin \theta \\ \sin \theta & \cos \theta \end{pmatrix}$. Topologically, this is a circle $S^1$.
    \item $A$ is the group of diagonal matrices with positive entries, which is topologically a line $\mathbb{R}$.
    \item $N$ is the group of upper triangular matrices with 1s on the diagonal, which is also a line $\mathbb{R}$.
\end{itemize}
So, as a manifold, $SL_2(\mathbb{R}) \cong S^1 \times \mathbb{R}^2$.

\textbf{Step 2: Find the fundamental group.}
The fundamental group of a product space is the product of the fundamental groups.
$\pi_1(SL_2(\mathbb{R})) \cong \pi_1(S^1) \times \pi_1(\mathbb{R}) \times \pi_1(\mathbb{R})$.
Since $\pi_1(S^1) = \mathbb{Z}$ and $\pi_1(\mathbb{R}) = 0$, we have:
\[ \pi_1(SL_2(\mathbb{R})) \cong \mathbb{Z} \]

\textbf{Step 3: Describe the universal cover.}
The universal cover $\widetilde{SL}_2(\mathbb{R})$ is obtained by "unrolling" the circle part. Instead of the angle $\theta$ being in $[0, 2\pi)$, we let $\theta$ be any real number in $\mathbb{R}$.
Elements can be represented by $(\theta, a, n)$.
The covering map $p: \widetilde{SL}_2(\mathbb{R}) \to SL_2(\mathbb{R})$ is:
\[ (\theta, a, n) \mapsto \begin{pmatrix} \cos \theta & -\sin \theta \\ \sin \theta & \cos \theta \end{pmatrix} A N \]
This space has a Lie group structure where the center is isomorphic to $\mathbb{Z}$. It is a unique group because $SL_2(\mathbb{R})$ has no finite-dimensional faithful representations.
\end{solution}

\begin{exercise}
Let $f \in \mathbb{C}[x, y, z]$ be an irreducible homogeneous polynomial of degree $d > 0$. For each integer $n \geq d$, define
\[ P(n) = \dim_{\mathbb{C}} \mathbb{C}[x, y, z]_n / f \cdot \mathbb{C}[x, y, z]_{n-d} \]
where $\mathbb{C}[x, y, z]_n$ is the subspace of homogeneous polynomials of degree $n$. Show that there are constants $c$ such that for $n$ sufficiently large, $P(n) = dn + c$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Homogeneous Polynomials:} Polynomials where every term has the same degree. The dimension of the space of degree $n$ polynomials in $k$ variables is $\binom{n+k-1}{k-1}$.
    \item \textbf{Quotient Space:} The space of polynomials of degree $n$ modulo those that are multiples of $f$.
    \item \textbf{Hilbert Polynomial:} The function $P(n)$ is the Hilbert polynomial of the projective curve defined by $f=0$.
\end{itemize}

\textbf{Step 1: Calculate the dimensions of the spaces.}
The space of homogeneous polynomials of degree $n$ in 3 variables ($x, y, z$) has dimension:
\[ \dim \mathbb{C}[x, y, z]_n = \binom{n+3-1}{3-1} = \binom{n+2}{2} = \frac{(n+2)(n+1)}{2} \]
Similarly, the space of degree $n-d$ polynomials has dimension:
\[ \dim \mathbb{C}[x, y, z]_{n-d} = \binom{n-d+2}{2} = \frac{(n-d+2)(n-d+1)}{2} \]

\textbf{Step 2: Use the property of the quotient.}
Since $f$ is irreducible, the map "multiplication by $f$" is injective. So the dimension of the subspace $f \cdot \mathbb{C}[x, y, z]_{n-d}$ is exactly the dimension of $\mathbb{C}[x, y, z]_{n-d}$.
The dimension of the quotient space $P(n)$ is the difference of the dimensions:
\[ P(n) = \frac{(n+2)(n+1)}{2} - \frac{(n-d+2)(n-d+1)}{2} \]

\textbf{Step 3: Simplify the expression.}
Let's expand the terms:
\[ P(n) = \frac{1}{2} [ (n^2 + 3n + 2) - ((n-d)^2 + 3(n-d) + 2) ] \]
\[ = \frac{1}{2} [ (n^2 + 3n + 2) - (n^2 - 2nd + d^2 + 3n - 3d + 2) ] \]
The $n^2$, $3n$, and $2$ terms cancel out:
\[ = \frac{1}{2} [ 2nd - d^2 + 3d ] = dn + \frac{d(3-d)}{2} \]

\textbf{Step 4: Identify the constants.}
We have found that $P(n) = dn + c$, where the constant $c$ is:
\[ c = \frac{3d - d^2}{2} \]
This holds for all $n \geq d$.
\end{solution}

\begin{exercise}
Let $p$ be an odd prime and $\mathbb{Z}_p$ the $p$-adic integers which can be defined as the projective limit of $\mathbb{Z}/p^n\mathbb{Z}$, and let $\mathbb{Q}_p$ be its fraction field. Let $\mathbb{Z}_p^\times$ denote the group of invertible elements in $\mathbb{Z}_p$ which is also the projective limit of $(\mathbb{Z}/p^n\mathbb{Z})^\times$.
\begin{enumerate}
    \item For any integer $a$ not divisible by $p$, show that the sequence $(a^{p^n})_n$ converges to an element $\omega(a) \in \mathbb{Z}_p$ satisfying $\omega(a)^{p-1} = 1$ and $\omega(a) \equiv a \pmod{p}$. Moreover, show that $\omega(a)$ depends only on $a \pmod{p}$.
    \item Define a logarithmic function $\log$ on $1 + p\mathbb{Z}_p$ by the usual formula: $\log(1+px) = \sum_{n=1}^\infty (-1)^{n-1} \frac{p^n}{n} x^n$. Show that the logarithmic function converges and defines an isomorphism $1 + p\mathbb{Z}_p \to p\mathbb{Z}_p$.
    \item Deduce from the above that $\mathbb{Z}_p^\times \cong \mathbb{Z}_p \times \mathbb{Z}/(p-1)\mathbb{Z}$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{$p$-adic Integers ($\mathbb{Z}_p$):} Numbers represented as power series in $p$: $a_0 + a_1 p + a_2 p^2 + \dots$.
    \item \textbf{Teichmüller Character ($\omega$):} A map that picks out roots of unity in $\mathbb{Z}_p$.
    \item \textbf{$p$-adic Logarithm:} A power series that behaves like the real $\ln(1+x)$ but converges under the $p$-adic metric.
\end{itemize}

\textbf{Step 1: The Teichmüller Lift.}
By Fermat's Little Theorem, $a^p \equiv a \pmod p$. One can show by induction that $a^{p^n} \equiv a^{p^{n-1}} \pmod{p^n}$.
This means the sequence $x_n = a^{p^n}$ is a Cauchy sequence in $\mathbb{Z}_p$, so it must converge to a limit $\omega(a)$.
This limit satisfies $\omega(a)^p = \omega(a)$, so $\omega(a)(\omega(a)^{p-1} - 1) = 0$. Since $a \not\equiv 0 \pmod p$, the limit $\omega(a)$ is non-zero, so $\omega(a)^{p-1} = 1$.
This $\omega(a)$ is the unique root of $x^{p-1}-1=0$ in $\mathbb{Z}_p$ that matches $a$ modulo $p$.

\textbf{Step 2: The $p$-adic Logarithm.}
The series $\log(1+z) = \sum \frac{(-1)^{n-1} z^n}{n}$ converges if the terms go to zero. In $\mathbb{Z}_p$, this happens if $v_p(z) > 1/(p-1)$.
For $z \in p\mathbb{Z}_p$ and $p > 2$, we have $v_p(z) \geq 1 > 1/(p-1)$, so the series converges.
The logarithm turns multiplication into addition: $\log(uv) = \log u + \log v$.
It defines a continuous group homomorphism from the multiplicative group $1+p\mathbb{Z}_p$ to the additive group $p\mathbb{Z}_p$.
Its inverse is given by the $p$-adic exponential series $\exp(y) = \sum y^n/n!$, which also converges for $y \in p\mathbb{Z}_p$ when $p > 2$.

\textbf{Step 3: Structure of the Units.}
Every unit $u \in \mathbb{Z}_p^\times$ can be uniquely written as:
\[ u = \omega(u \pmod p) \cdot (1 + p \beta) \]
where $\omega(u \pmod p)$ is a $(p-1)$-th root of unity and $1+p\beta \in 1+p\mathbb{Z}_p$.
This gives a direct product decomposition:
\[ \mathbb{Z}_p^\times \cong \mu_{p-1} \times (1 + p\mathbb{Z}_p) \]
The first part $\mu_{p-1}$ is a cyclic group of order $p-1$.
The second part is isomorphic to $\mathbb{Z}_p$ itself via the $\log$ map.
Thus, $\mathbb{Z}_p^\times \cong \mathbb{Z}/(p-1)\mathbb{Z} \times \mathbb{Z}_p$.
\end{solution}
