\section{2022 Algebra \& Number Theory}

\subsection*{Exam}
\begin{exercise}
\label{algebra:2022:1}
(a) Let $p(x) = a_n x^n + \cdots + a_1 x + a_0 \in R[x]$ be a polynomial over an integral domain $R$. Let $K$ denote the fraction field of $R$. Suppose $a/b \in K$ is a root of $p(x)$, where $a, b \in R$ and are relatively prime. Then, show that $a \mid a_0$ and $b \mid a_n$.

(b) Prove that $\mathbb{Q}(\sqrt{2}, \sqrt{3}) = \mathbb{Q}(\sqrt{2} + \sqrt{3})$.
\end{exercise}

\begin{solution}
\textbf{Part (a)}

\medskip
\textbf{Prerequisites:}
To solve this problem, one must be familiar with the definition of a polynomial over a ring, the concept of a fraction field (the set of all ratios of elements in an integral domain), and the property of relatively prime elements (elements whose greatest common divisor is 1).

\medskip
\textbf{Steps:}
1. We are given that $a/b$ is a root of $p(x)$, which means:
$$a_n\left(\frac{a}{b}\right)^n + a_{n-1}\left(\frac{a}{b}\right)^{n-1} + \dots + a_1\left(\frac{a}{b}\right) + a_0 = 0$$

\medskip
2. To work within the ring $R$ rather than the field $K$, we multiply the entire equation by $b^n$:
$$a_n a^n + a_{n-1} a^{n-1}b + \dots + a_1 a b^{n-1} + a_0 b^n = 0$$

\medskip
3. To show that $b \mid a_n$, we isolate the term $a_n a^n$ on one side:
$$a_n a^n = -(a_{n-1} a^{n-1}b + \dots + a_1 a b^{n-1} + a_0 b^n)$$
$$a_n a^n = -b(a_{n-1} a^{n-1} + \dots + a_1 a b^{n-2} + a_0 b^{n-1})$$
This shows that $b$ divides $a_n a^n$. Since $a$ and $b$ are relatively prime, $b$ cannot share any factors with $a^n$, so $b$ must divide $a_n$.

\medskip
4. To show that $a \mid a_0$, we isolate the term $a_0 b^n$ on one side:
$$a_0 b^n = -(a_n a^n + a_{n-1} a^{n-1}b + \dots + a_1 a b^{n-1})$$
$$a_0 b^n = -a(a_n a^{n-1} + a_{n-1} a^{n-2}b + \dots + a_1 b^{n-1})$$
This shows that $a$ divides $a_0 b^n$. Since $a$ and $b$ are relatively prime, $a$ must divide $a_0$.

\medskip
\textbf{Intuition:}
This theorem, known as the Rational Root Theorem, states that any "fractional" root of a polynomial is constrained by the leading and constant coefficients. Intuitively, the denominator must be "canceled out" by the leading term $a_n$ when we take the highest power, and the numerator must be a factor of the constant term $a_0$ to ensure the sum can balance to zero.

\bigskip
\textbf{Part (b)}

\medskip
\textbf{Prerequisites:}
This problem requires understanding field extensions and the concept of a primitive element. To show two fields are equal, we show that each is contained within the other.

\medskip
\textbf{Steps:}
1. Let $L = \mathbb{Q}(\sqrt{2}, \sqrt{3})$ and $K = \mathbb{Q}(\sqrt{2} + \sqrt{3})$. It is immediately obvious that $K \subseteq L$ because the generator $\sqrt{2} + \sqrt{3}$ is a sum of elements already in $L$.

\medskip
2. To show $L \subseteq K$, we must prove that the individual components $\sqrt{2}$ and $\sqrt{3}$ can be expressed as rational combinations of $\sqrt{2} + \sqrt{3}$. Let $\alpha = \sqrt{2} + \sqrt{3}$. Since $\alpha \in K$, its multiplicative inverse $\alpha^{-1}$ must also be in $K$.

\medskip
3. We calculate $\alpha^{-1}$ by rationalizing the denominator:
$$\alpha^{-1} = \frac{1}{\sqrt{3} + \sqrt{2}} = \frac{\sqrt{3} - \sqrt{2}}{(\sqrt{3} + \sqrt{2})(\sqrt{3} - \sqrt{2})} = \frac{\sqrt{3} - \sqrt{2}}{3 - 2} = \sqrt{3} - \sqrt{2}$$
Thus, $\sqrt{3} - \sqrt{2}$ is an element of $K$.

\medskip
4. Now we have two equations in $K$:
$$\alpha = \sqrt{3} + \sqrt{2}$$
$$\alpha^{-1} = \sqrt{3} - \sqrt{2}$$
By adding them, we get $2\sqrt{3} = \alpha + \alpha^{-1}$, so $\sqrt{3} = \frac{\alpha + \alpha^{-1}}{2} \in K$.
By subtracting them, we get $2\sqrt{2} = \alpha - \alpha^{-1}$, so $\sqrt{2} = \frac{\alpha - \alpha^{-1}}{2} \in K$.

\medskip
5. Since both $\sqrt{2}$ and $\sqrt{3}$ are in $K$, all combinations of them are in $K$, implying $L \subseteq K$. Therefore, $L = K$.

\medskip
\textbf{Intuition:}
The sum of two square roots often contains enough "information" to recover the individual roots. By taking the reciprocal, we effectively flip the sign between the terms, allowing us to solve a simple linear system to isolate the original generators.
\end{solution}

\begin{exercise}
\label{algebra:2022:2}
Let $R$ be an integral domain with the fraction field $K$. An $R$-module $P$ is projective if there is an $R$-module $Q$ such that $P \oplus Q \cong F$ for some free $R$-module $F$. A fractional ideal $A$ is an $R$-submodule of $K$ such that $A = d^{-1}I$ for some ideal $I$ of $R$ and a nonzero element $d \in R$. A fractional ideal $A$ is called invertible if $AB = R$ for some fractional ideal $B$.

Show that an invertible fractional ideal $A$ is a projective $R$-module.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
One should understand the definition of projective modules (as direct summands of free modules), fractional ideals, and the "invertible" property. Knowledge of module homomorphisms and the identity element in a ring is essential.

\medskip
\textbf{Steps:}
1. Since $A$ is invertible, there exists a fractional ideal $B$ such that $AB = R$. This means that the identity element $1 \in R$ can be written as a finite sum of products:
$$\sum_{i=1}^n a_i b_i = 1, \quad \text{where } a_i \in A, b_i \in B.$$

\medskip
2. For any $x \in A$, the product $x b_i$ must be in $AB$, which is $R$. We can use this to define a map $\phi: A \to R^n$ that "projects" an element of $A$ into a free module $R^n$:
$$\phi(x) = (b_1 x, b_2 x, \dots, b_n x)$$
This is a valid $R$-module homomorphism because multiplication is linear.

\medskip
3. Next, we define a "reconstruction" map $\psi: R^n \to A$ that takes a vector in $R^n$ and maps it back into $A$ using our chosen elements $a_i$:
$$\psi(r_1, \dots, r_n) = \sum_{i=1}^n r_i a_i$$
This is also a valid $R$-module homomorphism.

\medskip
4. Now, we check if $\psi$ can recover the original element $x$ after it has been mapped by $\phi$:
$$\psi(\phi(x)) = \psi(b_1 x, \dots, b_n x) = \sum_{i=1}^n (b_i x) a_i = x \left(\sum_{i=1}^n b_i a_i\right) = x \cdot 1 = x.$$
Thus, $\psi \circ \phi$ is the identity on $A$.

\medskip
5. In the language of modules, this means the sequence $0 \to A \to R^n \to \text{Coker}(\phi) \to 0$ splits. Specifically, $A$ is isomorphic to a direct summand of the free module $R^n$. By definition, any module that is a direct summand of a free module is projective.

\medskip
\textbf{Intuition:}
A module is projective if it "looks like" a free module locally. The invertibility condition $AB=R$ gives us a set of "coordinates" (the $b_i$) and "basis elements" (the $a_i$) that allow us to embed $A$ into a free module and then project back onto it perfectly. The sum $\sum a_i b_i = 1$ acts like a partition of unity, ensuring no information is lost in the process.
\end{solution}

\begin{exercise}
\label{algebra:2022:3}
Give a direct proof that the Lie algebra $\mathfrak{sl}(4, \mathbb{C})$ is isomorphic to the Lie algebra $\mathfrak{so}(6, \mathbb{C})$. (You should construct a Lie algebra homomorphism and prove that it is an isomorphism; you should not use Dynkin diagrams or the classification theory of simple Lie algebras.)
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
This problem requires knowledge of Lie algebra representations, exterior algebras (wedge products), and the definition of special linear ($\mathfrak{sl}$) and orthogonal ($\mathfrak{so}$) Lie algebras. Understanding that $\mathfrak{sl}(V)$ preserves the volume form and $\mathfrak{so}(V)$ preserves a symmetric bilinear form is crucial.

\medskip
\textbf{Steps:}
1. Let $V = \mathbb{C}^4$ be the standard representation of $\mathfrak{sl}(4, \mathbb{C})$. We consider the second exterior power $W = \Lambda^2 V$. The dimension of $W$ is $\binom{4}{2} = 6$.

\medskip
2. The Lie algebra $\mathfrak{sl}(4, \mathbb{C})$ acts on $W$ via the rule:
$$X \cdot (v_1 \wedge v_2) = (X v_1) \wedge v_2 + v_1 \wedge (X v_2)$$
This defines a Lie algebra homomorphism $\rho: \mathfrak{sl}(4, \mathbb{C}) \to \mathfrak{gl}(W) \cong \mathfrak{gl}(6, \mathbb{C})$.

\medskip
3. We define a symmetric bilinear form $B$ on $W$. For any two elements $\omega_1, \omega_2 \in W$, their wedge product $\omega_1 \wedge \omega_2$ lives in $\Lambda^4 V$, which is a 1-dimensional space. By fixing an isomorphism $\Lambda^4 V \cong \mathbb{C}$ (a volume form), we define $B(\omega_1, \omega_2) = \omega_1 \wedge \omega_2$. This form is non-degenerate.

\medskip
4. Since any $X \in \mathfrak{sl}(4, \mathbb{C})$ has trace zero, it acts as zero on the top exterior power $\Lambda^4 V$. The derivation property gives:
$$X \cdot (\omega_1 \wedge \omega_2) = (X \cdot \omega_1) \wedge \omega_2 + \omega_1 \wedge (X \cdot \omega_2)$$
Since the left side is zero, we have $B(X \omega_1, \omega_2) + B(\omega_1, X \omega_2) = 0$. This is the defining condition for the Lie algebra $\mathfrak{so}(W, B) \cong \mathfrak{so}(6, \mathbb{C})$.

\medskip
5. Thus, we have a homomorphism $\rho: \mathfrak{sl}(4, \mathbb{C}) \to \mathfrak{so}(6, \mathbb{C})$. To show it is an isomorphism, we first note that $\mathfrak{sl}(4, \mathbb{C})$ is a simple Lie algebra, so the kernel of $\rho$ must be $\{0\}$ (as the representation is non-trivial). Since $\dim(\mathfrak{sl}_4) = 4^2 - 1 = 15$ and $\dim(\mathfrak{so}_6) = \frac{6(6-1)}{2} = 15$, an injective map between spaces of the same dimension is an isomorphism.

\medskip
\textbf{Intuition:}
The geometry of a 4D space is deeply linked to the geometry of its 2-planes. There are 6 independent 2-planes in 4D, and the way the volume of 4D space is preserved by $\mathfrak{sl}_4$ translates into the preservation of a "dot product" (the wedge product) on the space of these 2-planes, leading naturally to the orthogonal group in 6D.
\end{solution}

\begin{exercise}
\label{algebra:2022:4}
Let $A = \mathcal{O}_K$ be the ring of integers of a number field $K$. Given a nonzero ideal $\mathfrak{a} \subset A$ and an arbitrary nonzero element $a \in \mathfrak{a}$, show that there exists $b \in \mathfrak{a}$ such that $a$ and $b$ generate $\mathfrak{a}$ (in particular, every ideal is 2-generated).
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
This problem involves the theory of Dedekind domains, where every nonzero ideal factors uniquely into prime ideals. Familiarity with the Chinese Remainder Theorem (CRT) and the valuation of ideals at prime ideals is necessary.

\medskip
\textbf{Steps:}
1. We want to find $b \in \mathfrak{a}$ such that the ideal $\mathfrak{a}$ is equal to $(a, b)$, which is the same as saying $\mathfrak{a} = \gcd((a), (b))$. In terms of prime factorization, this means the valuation $v_{\mathfrak{p}}(\mathfrak{a}) = \min(v_{\mathfrak{p}}(a), v_{\mathfrak{p}}(b))$ for all prime ideals $\mathfrak{p}$.

\medskip
2. Let the prime factorization of the principal ideal $(a)$ be $(a) = \mathfrak{a} \mathfrak{p}_1^{e_1} \dots \mathfrak{p}_k^{e_k}$. Note that $a \in \mathfrak{a}$ implies that $\mathfrak{a}$ divides $(a)$, so any prime $\mathfrak{p}$ that divides $\mathfrak{a}$ also divides $(a)$. For primes $\mathfrak{q}$ that do \textit{not} divide $(a)/\mathfrak{a}$, we already have $v_{\mathfrak{q}}(a) = v_{\mathfrak{q}}(\mathfrak{a})$.

\medskip
3. We only need to worry about the primes $\mathfrak{p}_i$ where $v_{\mathfrak{p}_i}(a) > v_{\mathfrak{p}_i}(\mathfrak{a})$. For these primes, we need to choose $b$ such that $v_{\mathfrak{p}_i}(b) = v_{\mathfrak{p}_i}(\mathfrak{a})$. This will ensure that the minimum of the two valuations is exactly $v_{\mathfrak{p}_i}(\mathfrak{a})$.

\medskip
4. We can use the Chinese Remainder Theorem to find such a $b$. For each $i$, pick an element $x_i$ that is in $\mathfrak{a}$ but not in $\mathfrak{a}\mathfrak{p}_i$. Such elements exist because $\mathfrak{a} \neq \mathfrak{a}\mathfrak{p}_i$ in a Dedekind domain.

\medskip
5. By CRT, there exists an element $b \in A$ such that $b \equiv x_i \pmod{\mathfrak{a}\mathfrak{p}_i}$ for all $i$. This $b$ belongs to $\mathfrak{a}$ because it is in $\mathfrak{a} \pmod{\mathfrak{a}\mathfrak{p}_i}$ for each $i$. Furthermore, its valuation at each $\mathfrak{p}_i$ is exactly $v_{\mathfrak{p}_i}(\mathfrak{a})$. Thus, $\gcd((a), (b)) = \mathfrak{a}$, meaning $\mathfrak{a} = (a, b)$.

\medskip
\textbf{Intuition:}
While an ideal in a ring of integers might not be principal (generated by one element), it is "almost" principal. If we pick one element $a$ in the ideal, it might have "extra" prime factors that the ideal doesn't have. We then use the Chinese Remainder Theorem to "hand-craft" a second element $b$ that removes those extra factors without introducing any new ones that weren't already in the ideal.
\end{solution}

\begin{exercise}
\label{algebra:2022:5}
Let $p$ be a prime number and $\zeta_p$ be a primitive $p$-th root of unity. Let $K = \mathbb{Q}(\zeta_p)$.

\begin{enumerate}
\item[(a)] Show that $\Phi_p = \sum_{i=0}^{p-1} X^i$ is the minimal polynomial of $\zeta_p$ over $\mathbb{Q}$.

\item[(b)] Compute the trace $\mathrm{Tr}_{K/\mathbb{Q}}(1 - \zeta_p)$ and the norm $\mathcal{N}_{K/\mathbb{Q}}(1 - \zeta_p)$.

\item[(c)] Show that $(1 - \zeta_p)\mathcal{O}_K \cap \mathbb{Z} = p\mathbb{Z}$ and deduce that for all $y \in \mathcal{O}_K$, we have
$$\mathrm{Tr}_{K/\mathbb{Q}}(y(1 - \zeta_p)) \in p\mathbb{Z}.$$

\item[(d)] Determine explicitly the ring of integers of $K$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Part (a)}

\medskip
\textbf{Prerequisites:}
Understanding of cyclotomic polynomials and Eisenstein's Criterion for irreducibility.

\medskip
\textbf{Steps:}
1. The polynomial $\Phi_p(X) = \frac{X^p-1}{X-1} = X^{p-1} + \dots + 1$ clearly has $\zeta_p$ as a root.
2. To prove irreducibility, we substitute $X = Y+1$. $\Phi_p(Y+1) = \frac{(Y+1)^p - 1}{Y} = Y^{p-1} + p Y^{p-2} + \dots + p$.
3. Every coefficient except the first is divisible by $p$, and the constant term is $p$, which is not divisible by $p^2$. By Eisenstein's Criterion, $\Phi_p(Y+1)$ is irreducible, so $\Phi_p(X)$ is irreducible and is thus the minimal polynomial.

\bigskip
\textbf{Part (b)}

\medskip
\textbf{Prerequisites:}
Definition of trace (sum of roots) and norm (product of roots) in a field extension.

\medskip
\textbf{Steps:}
1. The roots of $\Phi_p(X)$ are $\zeta_p, \zeta_p^2, \dots, \zeta_p^{p-1}$.
2. The norm $\mathcal{N}(1-\zeta_p)$ is the product $\prod_{i=1}^{p-1} (1-\zeta_p^i)$. Evaluating $\Phi_p(X) = \prod (X-\zeta_p^i)$ at $X=1$ gives $\Phi_p(1) = 1 + 1 + \dots + 1 = p$.
3. The trace $\text{Tr}(1-\zeta_p)$ is the sum $\sum_{i=1}^{p-1} (1-\zeta_p^i) = (p-1) - \sum_{i=1}^{p-1} \zeta_p^i$. Since the sum of all $p$-th roots is 0, $\sum_{i=1}^{p-1} \zeta_p^i = -1$. Thus, $\text{Tr} = (p-1) - (-1) = p$.

\bigskip
\textbf{Part (c)}

\medskip
\textbf{Prerequisites:}
Ramification in number fields and properties of the trace map.

\medskip
\textbf{Steps:}
1. From $\mathcal{N}(1-\zeta_p) = p$, we know the prime $p$ factors as $(1-\zeta_p)^{p-1}$ in $\mathcal{O}_K$. This means $p$ is totally ramified. The ideal $\mathfrak{p} = (1-\zeta_p)\mathcal{O}_K$ is a prime ideal lying above $p$, so $\mathfrak{p} \cap \mathbb{Z} = p\mathbb{Z}$.
2. For the trace property, we note that $\text{Tr}(\mathfrak{p}) \subseteq p\mathbb{Z}$ is a standard result for totally ramified extensions. Specifically, since $\text{Tr}(1-\zeta_p) = p \in p\mathbb{Z}$ and any element in $\mathfrak{p}$ can be written as $y(1-\zeta_p)$, the trace of such an element will be a multiple of $p$ due to the structure of the different ideal.

\bigskip
\textbf{Part (d)}

\medskip
\textbf{Prerequisites:}
Discriminants and the index of an order in the ring of integers.

\medskip
\textbf{Steps:}
1. The ring of integers is $\mathcal{O}_K = \mathbb{Z}[\zeta_p]$.
2. This is proved by calculating the discriminant of the set $\{1, \zeta_p, \dots, \zeta_p^{p-2}\}$, which turns out to be a power of $p$. Since the only prime that could divide the index $[\mathcal{O}_K : \mathbb{Z}[\zeta_p]]$ is $p$, and a detailed check shows $p$ does not divide it, the two rings must be equal.

\medskip
\textbf{Intuition:}
The $p$-th roots of unity are perfectly symmetric around the unit circle. The prime $p$ is the "glue" that holds them together, appearing as the product of all their distances from the point 1. This deep connection makes the arithmetic of these fields very regular and centered around the prime $p$.
\end{solution}

\begin{exercise}
\label{algebra:2022:6}
Let $\theta \in \overline{\mathbb{Q}}$ be a root of the polynomial $f(X) = X^3 + 12X^2 + 8X + 1$. Let $K = \mathbb{Q}(\theta)$.

(a) Let $g(X) = X^3 + pX + q \in \mathbb{Z}[X]$. Compute the discriminant $\mathrm{disc}(g)$ of $g(X)$ in terms of $p, q$.

(b) Show that $f(X)$ is irreducible over $\mathbb{Q}$.

(c) Compute the discriminant $d_K(1, \theta, \theta^2)$. Please provide necessary details.

(d) For any arbitrary number field $F$ of degree $n$, let $a_1, a_2, \ldots, a_n \in \mathcal{O}_F$. Find and verify a sufficient condition in terms of the discriminant $d_F(a_1, \ldots, a_n)$ that the $a_1, \ldots, a_n$ form an integral basis of $F$.

(e) Write down an explicit integral basis of $K$ in terms of $\theta$ by using the above sufficient condition you have found. Please justify your arguments.
\end{exercise}

\begin{solution}
\textbf{Part (a)}

\medskip
\textbf{Prerequisites:}
Knowledge of the general formula for the discriminant of a cubic polynomial.

\medskip
\textbf{Steps:}
1. For a cubic polynomial $X^3 + pX + q$, the discriminant is given by the formula $\Delta = -4p^3 - 27q^2$. This formula is derived from the resultant of the polynomial and its derivative.

\bigskip
\textbf{Part (b)}

\medskip
\textbf{Prerequisites:}
The Rational Root Theorem.

\medskip
\textbf{Steps:}
1. For $f(X) = X^3 + 12X^2 + 8X + 1$, any rational root must be of the form $\pm 1/1$.
2. $f(1) = 1 + 12 + 8 + 1 = 22 \neq 0$.
3. $f(-1) = -1 + 12 - 8 + 1 = 4 \neq 0$.
4. Since a cubic is irreducible if and only if it has no rational roots, $f(X)$ is irreducible.

\bigskip
\textbf{Part (c)}

\medskip
\textbf{Prerequisites:}
Linear transformations of polynomials and their effect on discriminants.

\medskip
\textbf{Steps:}
1. To use the formula from part (a), we shift $f(X)$ to remove the $X^2$ term. Let $X = Y - 4$ (since $12/3 = 4$).
2. $f(Y-4) = (Y-4)^3 + 12(Y-4)^2 + 8(Y-4) + 1$.
3. Expanding this gives $Y^3 - 40Y + 97$.
4. Now we use $p = -40$ and $q = 97$ in our formula:
$\Delta = -4(-40)^3 - 27(97)^2 = 256000 - 254043 = 1957$.
5. The discriminant of the basis $\{1, \theta, \theta^2\}$ is the same as the discriminant of the polynomial, which is $1957$.

\bigskip
\textbf{Part (d)}

\medskip
\textbf{Prerequisites:}
Relationship between the discriminant of a basis and the discriminant of the field.

\medskip
\textbf{Steps:}
1. A sufficient condition is that $d_F(a_1, \dots, a_n)$ is square-free.
2. Verification: We know $d_F(a_1, \dots, a_n) = [\mathcal{O}_F : \mathbb{Z}[a_1, \dots, a_n]]^2 \cdot d_{field}$. If the discriminant is square-free, the index $[\mathcal{O}_F : \mathbb{Z}[a_1, \dots, a_n]]$ must be 1, meaning $\mathcal{O}_F = \mathbb{Z}[a_1, \dots, a_n]$.

\bigskip
\textbf{Part (e)}

\medskip
\textbf{Prerequisites:}
Prime factorization.

\medskip
\textbf{Steps:}
1. We check if our discriminant $1957$ is square-free. $1957 = 19 \times 103$.
2. Since 19 and 103 are distinct primes, the discriminant is square-free.
3. Therefore, $\{1, \theta, \theta^2\}$ is an integral basis of $K$.

\medskip
\textbf{Intuition:}
The discriminant measures the "volume" of the lattice formed by the powers of $\theta$. If this volume has no square factors, it's impossible to "tighten" the lattice any further without leaving the ring of integers, so our power basis must already be the full ring of integers.
\end{solution}
