\section{2011 Algebra \& Number Theory}

\subsection*{Individual}

\begin{exercise}
Let $K = \mathbb{Q}(\sqrt{-3})$, an imaginary quadratic field.

\begin{enumerate}
\item Does there exist a finite Galois extension $L/\mathbb{Q}$ which contains $K$ such that $\mathrm{Gal}(L/\mathbb{Q}) \cong S_3$?

\item Does there exist a finite Galois extension $L/\mathbb{Q}$ which contains $K$ such that $\mathrm{Gal}(L/\mathbb{Q}) \cong \mathbb{Z}/4\mathbb{Z}$?

\item Does there exist a finite Galois extension $L/\mathbb{Q}$ which contains $K$ such that $\mathrm{Gal}(L/\mathbb{Q}) \cong Q_8$?
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Galois Extension:} A field extension that is both normal (all roots of polynomials stay in the field) and separable. The "symmetries" of this extension form the Galois group.
    \item \textbf{Fundamental Theorem of Galois Theory:} Establishes a 1-to-1 correspondence between subfields of $L$ and subgroups of its Galois group.
    \item \textbf{Splitting Field:} The smallest field containing all roots of a specific polynomial.
    \item \textbf{Cyclic and Dihedral Groups:} $S_3$ (order 6), $\mathbb{Z}/4\mathbb{Z}$ (cyclic order 4), and $Q_8$ (quaternion group order 8).
    \item \textbf{Sum of Two Squares:} A theorem by Fermat states that a positive integer can be written as $x^2 + y^2$ if and only if its prime factors of the form $4k+3$ appear with an even exponent.
\end{itemize}

\textbf{(1) Step 1: Construct an $S_3$ extension.}
The group $S_3$ has 6 elements. A classic example of an $S_3$ extension over $\mathbb{Q}$ is the splitting field of $x^3 - 2$. 
The roots of $x^3 - 2 = 0$ are:
\[ \sqrt[3]{2}, \quad \sqrt[3]{2}\omega, \quad \sqrt[3]{2}\omega^2 \quad \text{where } \omega = e^{2\pi i/3} = \frac{-1 + \sqrt{-3}}{2}. \]
The splitting field is $L = \mathbb{Q}(\sqrt[3]{2}, \omega)$. 
Notice that $\omega$ involves $\sqrt{-3}$, so $\mathbb{Q}(\sqrt{-3}) \subset L$. 
The degree of $L/\mathbb{Q}$ is 6, and its Galois group is $S_3$ because it permutes the three roots of $x^3-2$ in all possible ways.
Since $K = \mathbb{Q}(\sqrt{-3})$ is a subfield of $L$ and the Galois group is $S_3$, the answer is \textbf{Yes}.

\textbf{(2) Step 2: Analyze cyclic extensions of degree 4.}
We want to know if $K = \mathbb{Q}(\sqrt{-3})$ can be inside a field $L$ where $\text{Gal}(L/\mathbb{Q}) \cong \mathbb{Z}/4\mathbb{Z}$.
A cyclic group of order 4 has a unique subgroup of order 2. By Galois theory, $L$ must have exactly one quadratic subfield. In our case, that subfield would have to be $K$.
There is a known theorem: a quadratic field $\mathbb{Q}(\sqrt{d})$ (where $d$ is square-free) is contained in a $\mathbb{Z}/4\mathbb{Z}$ extension if and only if $d$ is a sum of two squares in $\mathbb{Q}$. 
This is because for a cyclic extension $\mathbb{Q} \subset \mathbb{Q}(\sqrt{d}) \subset L$ to exist, the element $d$ must be "norm-compatible," which boils down to $d = a^2 + b^2$ for some $a, b \in \mathbb{Q}$.
For $K = \mathbb{Q}(\sqrt{-3})$, we have $d = -3$. 
Since $-3$ is a negative number, it cannot be written as $a^2 + b^2$ (as squares of real numbers are always non-negative).
Thus, such an extension \textbf{cannot exist}. The answer is \textbf{No}.

\textbf{(3) Step 3: Embed a quadratic field into a $Q_8$ extension.}
The question is whether $\mathbb{Q}(\sqrt{-3})$ can be part of a $Q_8$ extension.
The group $Q_8$ has 8 elements and is non-abelian. Its only 1D subfields are quadratic.
A deep result by Witt provides the condition: a biquadratic field $\mathbb{Q}(\sqrt{a}, \sqrt{b})$ can be embedded in a $Q_8$ extension if a certain quadratic form is equivalent to the identity form.
Practically, $Q_8$ extensions over $\mathbb{Q}$ are well-studied. For example, $L = \mathbb{Q}(\sqrt{(2+\sqrt{2})(3+\sqrt{6})})$ is a $Q_8$ extension.
More simply, for any quadratic field $\mathbb{Q}(\sqrt{d})$, one can check if it fits inside a $Q_8$ group structure. It turns out that for $d = -3$, there are many such extensions. For instance, $\mathbb{Q}(\sqrt{2}, \sqrt{3})$ can be embedded into $Q_8$.
Since there are no fundamental obstructions (like the negative sum of squares in the cyclic case), such extensions do exist. The answer is \textbf{Yes}.
\end{solution}

\begin{exercise}
Let $f: G \to GL_2(\mathbb{C})$ be a 2D rep of a finite group $G$ such that 1 is an eigenvalue of $f(\sigma)$ for every $\sigma \in G$. Prove that $f$ is a direct sum of two 1D reps.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Representation of a Group:} A mapping from group elements to matrices that preserves the group operation.
    \item \textbf{Character ($\chi$):} The trace of the matrices in a representation. $\chi(g) = \text{tr}(f(g))$.
    \item \textbf{Inner Product of Characters:} A way to count how many times one representation is "inside" another. $\langle \chi, \psi \rangle = \frac{1}{|G|} \sum_{g \in G} \chi(g) \overline{\psi(g)}$.
    \item \textbf{Orthogonality Relations:} Irreducible characters form an orthonormal basis for the space of class functions.
\end{itemize}

\textbf{Step 1: Determine the eigenvalues and the character.}
Let's look at a group element $\sigma \in G$. The representation $f(\sigma)$ is a $2 \times 2$ matrix. We are told that $1$ is always an eigenvalue of $f(\sigma)$.
Since it's a $2 \times 2$ matrix, there is one other eigenvalue. Let's call it $\lambda(\sigma)$.
The trace of a matrix is the sum of its eigenvalues. So, the character of this representation is:
\[ \chi(\sigma) = 1 + \lambda(\sigma) \]
Also, the determinant of a matrix is the product of its eigenvalues. So, $\det(f(\sigma)) = 1 \cdot \lambda(\sigma) = \lambda(\sigma)$.
Because the determinant of a representation is itself a 1-dimensional representation (a homomorphism from $G$ to $\mathbb{C}^\times$), we know that $\lambda$ is a 1-dimensional character of $G$.

\textbf{Step 2: Check for the trivial representation.}
We want to see if our 2D representation $f$ contains the trivial 1D representation. To do this, we compute the inner product of our character $\chi$ with the trivial character $\chi_{triv}$ (where $\chi_{triv}(\sigma) = 1$ for all $\sigma$).
\begin{align*}
\langle \chi, \chi_{triv} \rangle &= \frac{1}{|G|} \sum_{\sigma \in G} \chi(\sigma) \cdot 1 \\
&= \frac{1}{|G|} \sum_{\sigma \in G} (1 + \lambda(\sigma)) \\
&= \frac{1}{|G|} \sum_{\sigma \in G} 1 + \frac{1}{|G|} \sum_{\sigma \in G} \lambda(\sigma)
\end{align*}
The first part is just $\frac{|G|}{|G|} = 1$. The second part is the inner product $\langle \lambda, \chi_{triv} \rangle$.

\textbf{Step 3: Analyze the decomposition.}
Since $\lambda$ is a 1-dimensional representation, it is irreducible.
\begin{itemize}
    \item If $\lambda$ is the trivial representation, then $\langle \lambda, \chi_{triv} \rangle = 1$. In this case, $\langle \chi, \chi_{triv} \rangle = 1 + 1 = 2$. This means $f$ is just the sum of two trivial representations: $f \cong \mathbb{C} \oplus \mathbb{C}$.
    \item If $\lambda$ is \textit{not} the trivial representation, then by the orthogonality of irreducible characters, $\langle \lambda, \chi_{triv} \rangle = 0$. In this case, $\langle \chi, \chi_{triv} \rangle = 1 + 0 = 1$. This means the trivial representation appears exactly once in $f$.
\end{itemize}
Since $f$ is 2-dimensional and contains a 1D trivial representation, the "leftover" part must also be 1-dimensional. 
Specifically, if we choose a basis where the first vector is the trivial invariant vector, the matrices will look like $\begin{pmatrix} 1 & 0 \\ 0 & \lambda(\sigma) \end{pmatrix}$.
Thus, $f$ is always a direct sum of two 1D representations!
\end{solution}

\begin{exercise}
Let $F \subset \mathbb{R}$ be the subset of all real numbers that are roots of monic polynomials $f(X) \in \mathbb{Q}[X]$.
\begin{enumerate}
    \item Show that $F$ is a field.
    \item Show that the only field automorphisms of $F$ are the identity automorphism $\sigma(x) = x$ for all $x \in F$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Algebraic Number:} A number that is a root of a non-zero polynomial with rational coefficients.
    \item \textbf{Monic Polynomial:} A polynomial where the leading coefficient (the one for the highest power of $x$) is 1.
    \item \textbf{Field:} A set of numbers where you can add, subtract, multiply, and divide (by non-zero numbers) and stay within the set.
    \item \textbf{Field Automorphism:} A one-to-one mapping from a field to itself that preserves all algebraic operations.
    \item \textbf{Density of Rationals:} Between any two distinct real numbers, there is always a rational number.
\end{itemize}

\textbf{(1) Step 1: Show that $F$ is a field.}
The set $F$ consists of all real numbers that are roots of monic polynomials with rational coefficients. These are precisely the \textbf{real algebraic numbers}.
A fundamental result in algebra is that if two numbers $\alpha$ and $\beta$ are algebraic, then their sum $\alpha + \beta$, product $\alpha \cdot \beta$, and inverse $1/\alpha$ (if $\alpha \neq 0$) are also algebraic.
Since $\mathbb{R}$ is already a field, and the set of all algebraic numbers $\overline{\mathbb{Q}}$ is a field, their intersection $F = \overline{\mathbb{Q}} \cap \mathbb{R}$ must also be a field. This is because the intersection of any two fields is still a field (it inherits the closure under addition, multiplication, etc.).

\textbf{(2) Step 2: Analyze automorphisms and fixed points.}
Let $\sigma$ be any automorphism of the field $F$. By definition, an automorphism must fix the "base" of the field, which is $\mathbb{Q}$. So, $\sigma(q) = q$ for every rational number $q$.
Now, we need to see what $\sigma$ does to irrational numbers in $F$. A key property of the real numbers (unlike complex numbers) is that they have a built-in order ($<$).
We claim that $\sigma$ must preserve this order. That is, if $x < y$, then $\sigma(x) < \sigma(y)$.
To see this, note that $x < y$ is the same as saying $y - x > 0$. In the real numbers, a number is positive if and only if it is a square of another non-zero real number.
If $y - x = z^2$ for some $z \in \mathbb{R}$, then since $y - x$ is algebraic, its square root $z$ is also algebraic. Since $z$ is real, $z \in F$.
Applying the automorphism: $\sigma(y - x) = \sigma(z^2) = (\sigma(z))^2$.
Since the square of any non-zero number in $F$ is positive, we have $\sigma(y - x) > 0$, which means $\sigma(y) - \sigma(x) > 0$, or $\sigma(x) < \sigma(y)$.

\textbf{(3) Step 3: Use the density of rationals to force the identity.}
Suppose there was some number $x \in F$ such that $\sigma(x) \neq x$. 
\begin{itemize}
    \item If $\sigma(x) < x$, then by the density of rational numbers, there must be some rational number $q$ such that $\sigma(x) < q < x$.
    \item However, since $\sigma$ preserves order, $q < x \implies \sigma(q) < \sigma(x)$.
    \item But we know $\sigma(q) = q$ for all rationals. So, $q < \sigma(x)$.
\end{itemize}
This is a direct contradiction! We started with $\sigma(x) < q$ and ended with $q < \sigma(x)$.
The same contradiction occurs if we assume $\sigma(x) > x$.
Therefore, the only possibility is that $\sigma(x) = x$ for every single $x$ in $F$. The only automorphism of $F$ is the identity map.
\end{solution}

\begin{exercise}
Let $V$ be a finite-dimensional vector space over $\mathbb{R}$ and $T: V \to V$ be a linear transformation such that:
\begin{enumerate}
    \item the minimal polynomial of $T$ is irreducible;
    \item there exists a vector $v \in V$ such that $\{ T^i v \mid i \geq 0 \}$ spans $V$.
\end{enumerate}
Show that $V$ contains no non-trivial proper $T$-invariant subspace.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Minimal Polynomial ($m_T(x)$):} The unique monic polynomial of lowest degree such that $m_T(T) = 0$.
    \item \textbf{Characteristic Polynomial ($f(x)$):} Defined by $\det(xI - T)$. By Cayley-Hamilton, $f(T) = 0$.
    \item \textbf{Cyclic Vector:} A vector $v$ such that its orbits under $T$ (i.e., $v, Tv, T^2v, \dots$) span the entire space $V$. A space with a cyclic vector is called a cyclic space.
    \item \textbf{Invariant Subspace:} A subspace $W \subseteq V$ such that $T(w) \in W$ for all $w \in W$.
    \item \textbf{Irreducible Polynomial:} A polynomial that cannot be factored into two non-constant polynomials over the base field (here, $\mathbb{R}$).
\end{itemize}

\textbf{Step 1: Relate the minimal and characteristic polynomials.}
The problem tells us two things:
1. The minimal polynomial $m_T(x)$ is irreducible.
2. The space $V$ has a cyclic vector $v$.
For any linear transformation, the minimal polynomial $m_T(x)$ and the characteristic polynomial $f(x)$ share the same roots. For a cyclic space, there is a stronger relationship: the minimal polynomial is exactly equal to the characteristic polynomial (up to a possible sign). 
Since $m_T(x)$ is irreducible and $f(x)$ is a power of $m_T(x)$ (if $V$ weren't cyclic), but here they must coincide. So, $f(x) = m_T(x)$.

\textbf{Step 2: View the vector space as a Module.}
A powerful way to think about linear transformations is to view the vector space $V$ as a module over the ring of polynomials $\mathbb{R}[x]$. In this view, multiplying a vector $u$ by the variable $x$ is defined as applying the transformation $T$:
\[ x \cdot u = T(u) \]
Since $V$ has a cyclic vector $v$, every vector in $V$ can be written as a polynomial in $T$ applied to $v$: $u = P(T)v$. 
This means $V$ is a cyclic module, isomorphic to the quotient ring:
\[ V \cong \mathbb{R}[x] / \langle f(x) \rangle \]
where $\langle f(x) \rangle$ is the ideal generated by the characteristic polynomial.

\textbf{Step 3: Analyze the structure of the quotient ring.}
We are given that $f(x) = m_T(x)$ is irreducible over $\mathbb{R}$. 
In the ring of polynomials $\mathbb{R}[x]$, an ideal generated by an irreducible polynomial is a \textbf{maximal ideal}.
When you quotient a ring by a maximal ideal, the result is a \textbf{field}.
So, $V$ (viewed as an $\mathbb{R}[x]$-module) is actually a field!

\textbf{Step 4: Conclude about invariant subspaces.}
A $T$-invariant subspace $W \subseteq V$ is precisely a set of vectors that is closed under addition and "multiplication by $x$" (applying $T$). In other words, $W$ is an $\mathbb{R}[x]$-submodule of $V$.
In our isomorphism $V \cong \mathbb{R}[x] / \langle f(x) \rangle$, submodules correspond to \textbf{ideals} of the quotient ring.
Since we've established that this quotient ring is a field, it only has two ideals:
1. The zero ideal $\{0\}$, which corresponds to the trivial subspace $\{0\}$.
2. The entire ring itself, which corresponds to the whole space $V$.
There are no other ideals, which means there are no non-trivial proper $T$-invariant subspaces. $V$ is what we call an \textbf{irreducible representation} or a \textbf{simple module}.
\end{solution}

\begin{exercise}
Given a commutative diagram of Abelian groups:
\[ \begin{tikzcd}
A \arrow[r] \arrow[d, "\alpha"] & B \arrow[r] \arrow[d, "\beta"] & C \arrow[r] \arrow[d, "\gamma"] & D \arrow[r] \arrow[d, "\delta"] & E \arrow[d, "\epsilon"] \\
A' \arrow[r] & B' \arrow[r] & C' \arrow[r] & D' \arrow[r] & E'
\end{tikzcd} \]
such that (i) both rows are exact sequences and (ii) every vertical map, except the middle one $\gamma$, is an isomorphism. Show that the middle map $\gamma: C \to C'$ is also an isomorphism.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Exact Sequence:} A sequence of groups and maps $\dots \to A \xrightarrow{f} B \xrightarrow{g} C \to \dots$ where the image of one map is exactly the kernel of the next ($\text{im } f = \ker g$).
    \item \textbf{Commutative Diagram:} A drawing where if you start at point A and want to get to point B, any path you follow will give the same result (e.g., $\beta \circ f_1 = g_1 \circ \alpha$).
    \item \textbf{Diagram Chase:} A technique where you "pick up" an element and follow it through the arrows to deduce its properties.
    \item \textbf{Five Lemma:} This problem is exactly the statement of the "Five Lemma" from homological algebra.
\end{itemize}

\textbf{Step 1: Setup the labels.}
Let's name the horizontal maps in the top row $f_1, f_2, f_3, f_4$ and in the bottom row $g_1, g_2, g_3, g_4$. We are given that $\alpha, \beta, \delta, \epsilon$ are isomorphisms. We want to show $\gamma$ is an isomorphism (both injective and surjective).

\textbf{Step 2: Show that $\gamma$ is injective.}
Suppose we have an element $c \in C$ such that $\gamma(c) = 0$. We want to prove $c = 0$.
\begin{enumerate}
    \item Look at where $c$ goes in $D$: $f_3(c)$. By commutativity, $\delta(f_3(c)) = g_3(\gamma(c)) = g_3(0) = 0$.
    \item Since $\delta$ is an isomorphism (and thus injective), $\delta(f_3(c)) = 0$ implies $f_3(c) = 0$.
    \item By exactness at $C$, since $c$ is in the kernel of $f_3$, it must be in the image of $f_2$. So, $c = f_2(b)$ for some $b \in B$.
    \item Now, look at $\beta(b)$. By commutativity, $g_2(\beta(b)) = \gamma(f_2(b)) = \gamma(c) = 0$.
    \item By exactness at $B'$, since $\beta(b)$ is in the kernel of $g_2$, it must be in the image of $g_1$. So, $\beta(b) = g_1(a')$ for some $a' \in A'$.
    \item Since $\alpha$ is an isomorphism, $a' = \alpha(a)$ for some $a \in A$.
    \item Substituting this in, $\beta(b) = g_1(\alpha(a)) = \beta(f_1(a))$ (by commutativity).
    \item Since $\beta$ is an isomorphism, we can "cancel" it: $b = f_1(a)$.
    \item Finally, $c = f_2(b) = f_2(f_1(a))$. By exactness at $B$, the composition $f_2 \circ f_1 = 0$. So $c = 0$. Thus, $\gamma$ is injective!
\end{enumerate}

\textbf{Step 3: Show that $\gamma$ is surjective.}
Now, take any element $c' \in C'$. we want to find a $c \in C$ such that $\gamma(c) = c'$.
\begin{enumerate}
    \item Look at $d' = g_3(c')$. Since $\delta$ is an isomorphism, there exists $d \in D$ such that $\delta(d) = d'$.
    \item Note that $\epsilon(f_4(d)) = g_4(\delta(d)) = g_4(d') = g_4(g_3(c')) = 0$ (by exactness at $D'$).
    \item Since $\epsilon$ is injective, $f_4(d) = 0$. By exactness at $D$, $d = f_3(c_0)$ for some $c_0 \in C$.
    \item Compare $\gamma(c_0)$ with $c'$: $g_3(\gamma(c_0)) = \delta(f_3(c_0)) = \delta(d) = d' = g_3(c')$.
    \item This means $g_3(\gamma(c_0) - c') = 0$. By exactness at $C'$, $\gamma(c_0) - c' = g_2(b')$ for some $b' \in B'$.
    \item Since $\beta$ is surjective, $b' = \beta(b)$ for some $b \in B$.
    \item Let $c = c_0 - f_2(b)$. Then $\gamma(c) = \gamma(c_0) - \gamma(f_2(b)) = \gamma(c_0) - g_2(\beta(b)) = \gamma(c_0) - g_2(b') = c'$.
    \item We found our $c$! Thus, $\gamma$ is surjective.
\end{enumerate}
Since $\gamma$ is both injective and surjective, it is an isomorphism.
\end{solution}

\begin{exercise}
Group of order 150 is not simple.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Simple Group:} Has no normal subgroups except $\{e\}$ and itself.
    \item \textbf{Sylow $p$-subgroup:} A subgroup of order $p^k$, where $p^k$ is the highest power of $p$ dividing the group order.
    \item \textbf{Sylow's Third Theorem:} The number of Sylow $p$-subgroups, $n_p$, must divide the group order and satisfy $n_p \equiv 1 \pmod p$.
    \item \textbf{Lagrange's Theorem:} The order of a subgroup must divide the order of the group.
\end{itemize}

\textbf{Step 1: Analyze the group order.}
The group has size $|G| = 150 = 2 \cdot 3 \cdot 5^2 = 2 \cdot 3 \cdot 25$.
Let's check the Sylow 5-subgroups. These are subgroups of order 25.
The number of such subgroups, $n_5$, must satisfy:
\begin{itemize}
    \item $n_5$ divides $150/25 = 6$.
    \item $n_5 \equiv 1 \pmod 5$.
\end{itemize}
The only possibilities for $n_5$ are 1 and 6.

\textbf{Step 2: Case $n_5 = 1$.}
If $n_5 = 1$, the Sylow 5-subgroup is unique. A unique Sylow subgroup is always normal.
Since $G$ has a normal subgroup of order 25 (which is neither $\{e\}$ nor $G$), the group $G$ is not simple.

\textbf{Step 3: Case $n_5 = 6$.}
What if there are 6 such subgroups? In this case, $G$ acts on these 6 subgroups by conjugation.
This defines a homomorphism $\phi: G \to S_6$.
If $G$ were simple, $\phi$ would have to be injective (otherwise the kernel would be a normal subgroup).
If $\phi$ is injective, $G$ is isomorphic to a subgroup of $S_6$.
This would mean every element in $G$ corresponds to a unique permutation in $S_6$.
However, consider a Sylow 5-subgroup $P$ of $G$, which has order 25. If $G$ is a subgroup of $S_6$, then $P$ must also be a subgroup of $S_6$.
By Lagrange's theorem, the order of any subgroup must divide the order of the group.
The order of $S_6$ is $6! = 720$.
Let's check: Does 25 divide 720?
$720 = 2^4 \cdot 3^2 \cdot 5^1$. 
Notice that 25 ($5^2$) does \textbf{not} divide 720. 
This is a contradiction! A group of order 720 cannot have a subgroup of order 25.
Therefore, the case $n_5 = 6$ is impossible.
The only remaining possibility is $n_5 = 1$, which proves $G$ is not simple.
\end{solution}

\subsection*{Team}
\begin{exercise}
Let $F'$ be a field and $F$ the algebraic closure of $F'$. Let $f(x, y)$ and $g(x, y)$ be polynomials in $F[x, y]$ such that $\gcd(f, g) = 1$ in $F[x, y]$. Show that there are only finitely many $(a, b) \in F^2$ such that $f(a, b) = g(a, b) = 0$. Can you generalize this to the case of more than two variables?
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{GCD of Polynomials:} The greatest common divisor. $\gcd(f, g)=1$ means they share no common non-constant polynomial factor.
    \item \textbf{Bezout's Theorem:} A powerful result in algebraic geometry stating that two plane curves of degrees $m$ and $n$ intersect in exactly $mn$ points (counting multiplicity and points at infinity), provided they share no common component.
    \item \textbf{Regular Sequence / Dimension:} Tools to describe the size of the set of solutions (variety) in higher dimensions.
\end{itemize}

\textbf{Step 1: Understand the geometric meaning.}
The equations $f(x, y) = 0$ and $g(x, y) = 0$ define two curves in the plane $F^2$.
The condition $\gcd(f, g) = 1$ means these curves have no common component (like a shared line or sub-curve).
If they shared a common component, their GCD would be a polynomial defining that component.

\textbf{Step 2: Apply Bezout's Theorem.}
Since the two curves share no common component, they can only intersect at a finite number of points.
According to Bezout's Theorem, if $\deg f = m$ and $\deg g = n$, the number of intersection points is at most $m \cdot n$.
This is a finite number, so there are only finitely many solutions $(a, b)$.

\textbf{Step 3: Consider the generalization.}
If we have three variables $x, y, z$ and two polynomials $f, g$, the solutions to $f=g=0$ will typically form a 1-dimensional curve, not a finite set of points.
Even if $\gcd(f, g) = 1$, we don't get a finite number of points. We need a third polynomial $h$ to "cut" the curve down to points.
The correct generalization is: if we have $n$ polynomials in $n$ variables, they have finitely many common zeros if and only if they form a \textbf{regular sequence} (or more precisely, if the quotient ring $F[x_1, \dots, x_n] / (f_1, \dots, f_n)$ is finite-dimensional as a vector space over $F$). 
In simpler terms, each new equation must "drop the dimension" by exactly one.
\end{solution}

\begin{exercise}
Let $D$ be a PID, and $D^n$ the free module of rank $n$ over $D$. Prove that any submodule of $D^n$ is a free module of rank $m \leq n$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{PID (Principal Ideal Domain):} A ring where every ideal is generated by a single element (e.g., $\mathbb{Z}$).
    \item \textbf{Free Module:} A module that has a basis, meaning every element can be uniquely written as a linear combination of basis elements.
    \item \textbf{Exact Sequence:} $0 \to A \to B \to C \to 0$ means $A$ is a submodule of $B$ and $C \cong B/A$.
    \item \textbf{Splitting:} A short exact sequence of modules splits if the last module is free, meaning $B \cong A \oplus C$.
\end{itemize}

\textbf{Step 1: Use induction on the rank $n$.}
\begin{itemize}
    \item \textbf{Base Case ($n=1$):} $D^1 = D$. A submodule of $D$ is simply an ideal $I \subseteq D$. Since $D$ is a PID, $I = \langle d \rangle$. If $d=0$, the rank is 0. If $d \neq 0$, then $I \cong D$ (the map $r \mapsto rd$ is an isomorphism), so it's a free module of rank 1. In both cases, the rank $m \leq 1$.
\end{itemize}

\textbf{Step 2: The Inductive Step.}
Assume the theorem is true for modules of rank $n-1$. Let $M$ be a submodule of $D^n$.
Let $\pi: D^n \to D$ be the projection onto the first coordinate: $\pi(d_1, \dots, d_n) = d_1$.
The image $\pi(M)$ is an ideal in $D$. Since $D$ is a PID, $\pi(M) = \langle d \rangle$.
There are two possibilities:
\begin{enumerate}
    \item If $d = 0$, then every element in $M$ has 0 as its first coordinate. This means $M$ is actually a submodule of $\{0\} \times D^{n-1} \cong D^{n-1}$. By our induction hypothesis, $M$ is free of rank $m \leq n-1$.
    \item If $d \neq 0$, then $\pi(M) \cong D$ as a module (as shown in the base case). We have a short exact sequence:
    \[ 0 \to \ker(\pi|_M) \to M \xrightarrow{\pi} \pi(M) \to 0 \]
    Because $\pi(M) \cong D$ is a free module, the sequence \textbf{splits}. This is a fundamental property of free modules: maps into them can be "lifted" back.
\end{enumerate}

\textbf{Step 3: Combine the pieces.}
Since the sequence splits, we have an isomorphism:
\[ M \cong \ker(\pi|_M) \oplus \pi(M) \]
The kernel $\ker(\pi|_M)$ consists of elements in $M$ with first coordinate 0, so it is a submodule of $D^{n-1}$. By the induction hypothesis, it is free of rank $m' \leq n-1$.
The image $\pi(M)$ is free of rank 1.
Therefore, $M$ is the direct sum of two free modules, which makes it a free module itself. Its rank is $m = m' + 1 \leq (n-1) + 1 = n$.
This completes the induction.
\end{solution}

\begin{exercise}
Identify pairs of integers $n \neq m \in \mathbb{Z}_+$ such that the quotient rings $\mathbb{Z}[x, y]/(x^2 - y^n) \cong \mathbb{Z}[x, y]/(x^2 - y^m)$; and identify pairs of integers $n \neq m \in \mathbb{Z}_+$ such that $\mathbb{Z}[x, y]/(x^2 - y^n) \not\cong \mathbb{Z}[x, y]/(x^2 - y^m)$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Integral Domain:} A ring with no zero divisors (if $ab=0$, then $a=0$ or $b=0$).
    \item \textbf{Normalization:} The process of finding all elements in the field of fractions that are roots of monic polynomials with coefficients in the ring.
    \item \textbf{Numerical Semigroup:} The set of non-negative integers that can be formed by adding up a fixed set of "generators" (e.g., $\langle 2, 3 \rangle = \{0, 2, 3, 4, \dots\}$).
\end{itemize}

\textbf{Step 1: Check for zero divisors (Even vs. Odd).}
Let $R_n = \mathbb{Z}[x, y] / \langle x^2 - y^n \rangle$.
If $n$ is an even number, say $n=2k$, then we can factor the expression:
\[ x^2 - y^{2k} = (x - y^k)(x + y^k) \]
In the ring $R_n$, the product of $(x-y^k)$ and $(x+y^k)$ is zero, but neither factor is zero. This means $R_n$ has \textbf{zero divisors}.
If $n$ is an odd number, the polynomial $x^2 - y^n$ is irreducible, so $R_n$ is an \textbf{integral domain}.
Since an isomorphism must preserve the property of being an integral domain, a ring with even $n$ can never be isomorphic to a ring with odd $n$.

\textbf{Step 2: Analyze the odd case using curves.}
For odd $n$, the equation $x^2 = y^n$ defines a curve. We can parameterize it using a new variable $t$.
Let $y = t^2$ and $x = t^n$. Then $(t^n)^2 = (t^2)^n = t^{2n}$, so it fits!
The ring $R_n$ is isomorphic to the subring of polynomials $\mathbb{Z}[t^2, t^n] \subset \mathbb{Z}[t]$.
This subring consists of all polynomials where only certain powers of $t$ are allowed. The allowed powers form a \textbf{numerical semigroup} $S_n = \langle 2, n \rangle$.

\textbf{Step 3: Compare the semigroups.}
The semigroup $S_n = \{ 2a + nb \mid a, b \in \mathbb{N} \}$ contains almost all integers. The integers it \textit{misses} are called "gaps".
For $S_n = \langle 2, n \rangle$ (where $n$ is odd), the missed integers are $1, 3, 5, \dots, n-2$.
The number of such gaps is exactly $(n-1)/2$.
If two rings $R_n$ and $R_m$ were isomorphic, their semigroups would have to be the same, and thus they would have to have the same number of gaps.
$(n-1)/2 = (m-1)/2$ implies $n = m$.
So, even within the odd numbers, different $n$ give different rings.

\textbf{Conclusion:}
There are \textbf{no pairs} $n \neq m$ such that the rings are isomorphic. Every distinct $n$ gives a non-isomorphic ring.
\end{solution}

\begin{exercise}
Is it possible to find an integer $n > 1$ such that the sum
\[ 1 + \frac{1}{2} + \frac{1}{3} + \dots + \frac{1}{n} \]
is an integer?
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Harmonic Number ($H_n$):} The sum of the reciprocals of the first $n$ natural numbers.
    \item \textbf{2-adic valuation:} A way of looking at how many factors of 2 are "hidden" in the denominator of a fraction.
    \item \textbf{Least Common Multiple (LCM):} The smallest number that is a multiple of all numbers in a set.
\end{itemize}

\textbf{Step 1: Identify the "dominant" power of 2.}
In the set of numbers $\{1, 2, 3, \dots, n\}$, there is always a unique largest power of 2. Let's call it $2^k$. 
Why is it unique? If there were another number $m$ in the set that was a multiple of $2^k$, it would have to be $2 \cdot 2^k = 2^{k+1}$ or more. But $2^k$ was chosen as the \textit{largest} power of 2 less than or equal to $n$. So $2^{k+1}$ must be greater than $n$.
Thus, $2^k$ is the \textbf{only} number in our set that is divisible by $2^k$. Every other number in the set is divisible by a \textit{smaller} power of 2.

\textbf{Step 2: Clear the denominators.}
Let $L$ be the least common multiple of all the numbers $1, 2, 3, \dots, n$.
We want to check if $H_n = 1 + \frac{1}{2} + \dots + \frac{1}{n}$ is an integer. 
If it were an integer, then multiplying it by $L/2^{k-1}$ would also give an integer.
\[ \frac{L}{2^{k-1}} \cdot H_n = \frac{L}{2^{k-1}} \left( 1 + \frac{1}{2} + \dots + \frac{1}{2^k} + \dots + \frac{1}{n} \right) \]

\textbf{Step 3: Analyze the terms of the sum.}
Consider the whole sum $H_n$ multiplied by $L$:
\[ L \cdot H_n = \frac{L}{1} + \frac{L}{2} + \dots + \frac{L}{2^k} + \dots + \frac{L}{n} \]
Every term $L/j$ in this sum is an integer because $L$ is a multiple of every $j$.
Now, consider which of these integers are \textbf{odd} or \textbf{even}.
The number $L$ contains the factor $2^k$ exactly once (by the definition of LCM and our observation in Step 1). 
\begin{itemize}
    \item For any $j \neq 2^k$, the power of 2 dividing $j$ is \textit{strictly less} than $2^k$. Therefore, $L/j$ will still have at least one factor of 2 remaining. So, $L/j$ is an \textbf{even} integer.
    \item For $j = 2^k$, the factor $2^k$ in the denominator perfectly cancels the $2^k$ in $L$. What's left in $L/2^k$ are only odd factors. So, $L/2^k$ is an \textbf{odd} integer.
\end{itemize}

\textbf{Step 4: Conclude.}
The sum $L \cdot H_n$ is a sum of many even integers plus exactly \textbf{one} odd integer.
\[ \text{Even} + \text{Even} + \dots + \text{Odd} + \dots + \text{Even} = \text{Odd} \]
So $L \cdot H_n$ is an odd integer.
But if $H_n$ were an integer, then $L \cdot H_n$ would have to be a multiple of $L$.
Since $L$ is an even number (for $n > 1$), $L \cdot H_n$ would have to be even.
We have a contradiction: $L \cdot H_n$ is both odd and even.
Therefore, $H_n$ can \textbf{never} be an integer for $n > 1$.
\end{solution}

\begin{exercise}
Recall that $\mathbb{F}_7$ is the finite field with 7 elements, and $GL_3(\mathbb{F}_7)$ is the group of all invertible $3 \times 3$ matrices with entries in $\mathbb{F}_7$.
\begin{enumerate}
    \item Find a 7-Sylow subgroup $P_7$ of $GL_3(\mathbb{F}_7)$.
    \item Determine the normalizer subgroup $N$ of the 7-Sylow subgroup you found in (1).
    \item Find a 2-Sylow subgroup of $GL_3(\mathbb{F}_7)$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Sylow $p$-subgroup:} A subgroup of the largest possible prime-power order $p^k$.
    \item \textbf{Normalizer ($N(P)$):} The set of all group elements $g$ such that $gPg^{-1} = P$.
    \item \textbf{Unipotent Matrix:} A matrix where all eigenvalues are 1. In a field of characteristic $p$, these form the $p$-Sylow subgroups.
    \item \textbf{Group Order of $GL_n(\mathbb{F}_q)$:} $|GL_n(\mathbb{F}_q)| = (q^n - 1)(q^n - q) \dots (q^n - q^{n-1})$.
\end{itemize}

\textbf{Step 1: Calculate the group order.}
For $n=3$ and $q=7$:
\[ |GL_3(\mathbb{F}_7)| = (7^3 - 1)(7^3 - 7)(7^3 - 49) = 342 \cdot 336 \cdot 294 \]
Factoring these numbers into primes:
\begin{itemize}
    \item $342 = 2 \cdot 171 = 2 \cdot 9 \cdot 19 = 2 \cdot 3^2 \cdot 19$
    \item $336 = 2^4 \cdot 3 \cdot 7$
    \item $294 = 2 \cdot 3 \cdot 7^2$
\end{itemize}
Combining them: Total order is $2^6 \cdot 3^4 \cdot 7^3 \cdot 19$.

\textbf{Step 2: Part (1) - Find the 7-Sylow subgroup.}
We need a subgroup of order $7^3 = 343$.
A standard choice for $p$-Sylow subgroups in matrix groups is the set of \textbf{upper triangular matrices with 1s on the diagonal} (called unipotent matrices):
\[ P_7 = \left\{ \begin{pmatrix} 1 & a & b \\ 0 & 1 & c \\ 0 & 0 & 1 \end{pmatrix} \mid a, b, c \in \mathbb{F}_7 \right\} \]
There are 3 variable entries ($a, b, c$), and each has 7 choices. So the size is $7 \times 7 \times 7 = 7^3$. This is our $P_7$.

\textbf{Step 3: Part (2) - Determine the Normalizer $N(P_7)$.}
The normalizer consists of matrices $g$ that preserve this upper-triangular structure when they conjugate $P_7$.
It is a well-known result that the normalizer of the unipotent upper triangular matrices is the group of \textbf{all invertible upper triangular matrices} (called the Borel subgroup $B$):
\[ N(P_7) = B = \left\{ \begin{pmatrix} d_1 & a & b \\ 0 & d_2 & c \\ 0 & 0 & d_3 \end{pmatrix} \mid d_i \neq 0, \text{ and } a, b, c \in \mathbb{F}_7 \right\} \]
The diagonal entries $d_1, d_2, d_3$ can be any of the 6 non-zero elements in $\mathbb{F}_7$. The entries $a, b, c$ can be anything.
The size of $N(P_7)$ is $6^3 \cdot 7^3$.

\textbf{Step 4: Part (3) - Find a 2-Sylow subgroup.}
We need a subgroup of order $2^6 = 64$.
A 2-Sylow of $GL_3$ can be built from a 2-Sylow of $GL_2$ embedded in the top-left, and a 2-Sylow of $GL_1$ in the bottom right.
$|GL_1(\mathbb{F}_7)| = 6$, 2-part is 2.
$|GL_2(\mathbb{F}_7)| = 2^5 \cdot 3^2 \cdot 7$.
So $2^5 \times 2 = 2^6$. Perfect.
A 2-Sylow subgroup of $GL_2(\mathbb{F}_7)$ is the \textbf{Semidihedral group} of order 32. 
So, a 2-Sylow of $GL_3$ is $SD_{32} \times \{ \pm 1 \}$.
\end{solution}

\begin{exercise}
Show that $SL_2(\mathbb{Z})$ is generated by $T = \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}$ and $S = \begin{pmatrix} 0 & 1 \\ -1 & 0 \end{pmatrix}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{$SL_2(\mathbb{Z})$:} The group of $2 \times 2$ matrices with integer entries and determinant 1.
    \item \textbf{Generators:} A set of elements that can form every other element in the group through multiplication and inverses.
    \item \textbf{Euclidean Algorithm:} A method to reduce numbers by repeatedly taking remainders.
\end{itemize}

\textbf{Step 1: Understand the effect of $T$ and $S$.}
Multiplying a matrix $M = \begin{pmatrix} a & b \\ c & d \end{pmatrix}$ by $T^n$ on the left performs a row operation:
\[ T^n \begin{pmatrix} a & b \\ c & d \end{pmatrix} = \begin{pmatrix} 1 & n \\ 0 & 1 \end{pmatrix} \begin{pmatrix} a & b \\ c & d \end{pmatrix} = \begin{pmatrix} a+nc & b+nd \\ c & d \end{pmatrix} \]
This allows us to change the top-left entry $a$ while keeping $c$ the same.
Multiplying by $S$ swaps the rows (with a sign flip):
\[ S \begin{pmatrix} a & b \\ c & d \end{pmatrix} = \begin{pmatrix} 0 & 1 \\ -1 & 0 \end{pmatrix} \begin{pmatrix} a & b \\ c & d \end{pmatrix} = \begin{pmatrix} c & d \\ -a & -b \end{pmatrix} \]

\textbf{Step 2: Use the Euclidean Algorithm on the first column.}
We want to show that by repeatedly applying $T$ and $S$, we can transform any matrix $M$ into the identity matrix $I$.
Suppose we have a matrix with $c \neq 0$.
1. Use $T^n$ to reduce $a$ modulo $c$. We can choose $n$ such that $|a+nc| \leq |c|/2$.
2. Then apply $S$. This moves the smaller remainder to the bottom-left position.
3. Repeat this process. Each time, the bottom-left entry $c$ gets strictly smaller in absolute value.
Since the entries are integers, this process must eventually terminate when the bottom-left entry becomes $0$.

\textbf{Step 3: Analyze the resulting matrix.}
When $c=0$, the matrix looks like $\begin{pmatrix} a & b \\ 0 & d \end{pmatrix}$.
Since the determinant must be 1, we have $ad - 0b = 1 \implies ad = 1$.
Because $a$ and $d$ are integers, the only possibilities are $a=d=1$ or $a=d=-1$.
\begin{itemize}
    \item If $a=d=1$, the matrix is $\begin{pmatrix} 1 & b \\ 0 & 1 \end{pmatrix}$, which is exactly $T^b$.
    \item If $a=d=-1$, the matrix is $\begin{pmatrix} -1 & b \\ 0 & -1 \end{pmatrix}$. Note that $S^2 = \begin{pmatrix} -1 & 0 \\ 0 & -1 \end{pmatrix}$. So this matrix is $S^2 T^{-b}$.
\end{itemize}
In all cases, we have shown that every matrix in $SL_2(\mathbb{Z})$ can be written as a product of $S$ and $T$. Thus, they generate the group.
\end{solution}
