\section{2016 Algebra \& Number Theory}

\subsection*{Individual}
\begin{exercise}
\label{algebra:2016:1}
Problem 1 (20 points). Let $E$ be a linear space over $\mathbb{R}$, of finite dimension $n \geq 2$, equipped with a positive definite symmetric bilinear form $\langle \cdot, \cdot \rangle$. Let $u_1, u_2, \ldots, u_n$ be a basis of $E$. Let $v_1, v_2, \ldots, v_n$ be the dual basis, that is,
$$\langle u_i, v_j \rangle = \begin{cases} 1 & \text{if } i = j, \\ 0 & \text{if } i \neq j, \end{cases}$$
for all $i, j = 1, 2, \ldots, n$.

(a) (8 points) Assume that $\langle u_i, u_j \rangle \leq 0$ for all $1 \leq i < j \leq n$. Show that there is an orthogonal basis $u_1', u_2', \ldots, u_n' of E$ such that $u_i'$ is a non-negative linear combination of $u_1, u_2, \ldots, u_i$, for all $i = 1, 2, \ldots, n$.

(b) (6 points) With the same assumption as in Part (a), show that $\langle v_i, v_j \rangle \geq 0$ for all $1 \leq i < j \leq n$.

(c) (6 points) Assume that $n \geq 3$. Show that the condition $\langle u_i, u_j \rangle \geq 0$ for all $1 \leq i < j \leq n$ does not imply that $\langle v_i, v_j \rangle \leq 0$ for all $1 \leq i < j \leq n$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Positive Definite Bilinear Form:} A generalization of the dot product. It satisfies $\langle x, x \rangle > 0$ for all $x \neq 0$.
    \item \textbf{Dual Basis:} A set of vectors $\{v_j\}$ that are perfectly "tuned" to the original basis $\{u_i\}$ such that their mutual dot products are either 0 or 1.
    \item \textbf{Gram-Schmidt Process:} A method to turn any basis into an orthogonal one while keeping the vectors within specific spans.
    \item \textbf{Gram Matrix ($G$):} The matrix of all possible dot products $G_{ij} = \langle u_i, u_j \rangle$.
    \item \textbf{Stieltjes Matrix / M-matrix:} A matrix with non-positive off-diagonal entries and positive diagonal entries. A key property is that its inverse has only non-negative entries.
\end{itemize}

\textbf{(a) Step 1: Orthogonalize using the Gram-Schmidt process.}
We want to build an orthogonal basis $\{u_i'\}$ such that each $u_i'$ is a non-negative combination of $\{u_1, \dots, u_i\}$.
The standard formula for the $i$-th vector in Gram-Schmidt is:
\[ u_i' = u_i - \sum_{j=1}^{i-1} \lambda_{ij} u_j' \quad \text{where } \lambda_{ij} = \frac{\langle u_i, u_j' \rangle}{\langle u_j', u_j' \rangle} \]
Let's check the signs of the coefficients $\lambda_{ij}$ using induction.
\begin{itemize}
    \item \textbf{Base Case ($i=1$):} $u_1' = u_1$. This is trivially a non-negative combination.
    \item \textbf{Inductive Step:} Assume $u_1', \dots, u_{i-1}'$ are all non-negative combinations of the original vectors.
    Then for any $j < i$, $\langle u_i, u_j' \rangle$ is a sum of terms like $c \cdot \langle u_i, u_k \rangle$ where $c \geq 0$ and $k \leq j < i$.
    Since we are given that $\langle u_i, u_k \rangle \leq 0$ for all $i \neq k$, every term in the sum is $\leq 0$.
    Thus, $\lambda_{ij} = \langle u_i, u_j' \rangle / \|u_j'\|^2 \leq 0$.
\end{itemize}
Rewriting the formula: $u_i' = u_i + \sum_{j=1}^{i-1} (-\lambda_{ij}) u_j'$.
Since $-\lambda_{ij} \geq 0$ and each $u_j'$ is already a non-negative combination, the entire sum for $u_i'$ is a non-negative combination.

\textbf{(b) Step 2: Relate the dual basis to the inverse Gram matrix.}
We can write any dual basis vector as a linear combination of the original basis: $v_j = \sum_k w_{jk} u_k$.
The definition $\langle u_i, v_j \rangle = \delta_{ij}$ translates to a matrix equation for the coefficients:
\[ \sum_k \langle u_i, u_k \rangle w_{jk} = \delta_{ij} \implies G W^T = I \]
where $G$ is the Gram matrix. Thus $W^T = G^{-1}$.
The inner products of the dual basis are given by:
\[ \langle v_i, v_j \rangle = \sum_{k,l} w_{ik} \langle u_k, u_l \rangle w_{jl} = (W G W^T)_{ij} = (G^{-1} G G^{-1})_{ij} = (G^{-1})_{ij} \]
Our Gram matrix $G$ has the property that its off-diagonal entries are $\leq 0$. Because $G$ is also positive definite, it is a \textbf{Stieltjes matrix}.
A deep result in linear algebra states that the inverse of any Stieltjes matrix contains only non-negative entries.
Therefore, $\langle v_i, v_j \rangle \geq 0$ for all $i, j$.

\textbf{(c) Step 3: Find a counterexample for the reverse implication.}
If we assume $\langle u_i, u_j \rangle \geq 0$, does it force the dual products to be $\leq 0$?
Consider 3 vectors in 2D space arranged in a tight "fan". Their mutual dot products are all positive (angles less than $90^\circ$).
However, in 3D, consider the Gram matrix $G = \begin{pmatrix} 1 & 0.5 & 0.5 \\ 0.5 & 1 & 0.1 \\ 0.5 & 0.1 & 1 \end{pmatrix}$.
\begin{enumerate}
    \item All entries are $\geq 0$.
    \item The determinant is $0.54$, so it's a valid positive-definite matrix (represents a real set of vectors).
    \item Let's calculate the inverse entry $(G^{-1})_{23}$. Using the cofactor formula:
    \[ (G^{-1})_{23} = \frac{-1}{0.54} \det \begin{pmatrix} 1 & 0.5 \\ 0.5 & 0.1 \end{pmatrix} = \frac{-(0.1 - 0.25)}{0.54} = \frac{0.15}{0.54} > 0 \]
\end{enumerate}
Since the result is positive, the condition $\langle v_2, v_3 \rangle \leq 0$ is violated. The implication does not hold.
\end{solution}

\begin{exercise}
\label{algebra:2016:2}
Problem 2 (20 points). Let $d \geq 1$ and $n \geq 1$ be integers.

(a) (5 points) Show that there are only finitely many subgroups $G \subseteq \mathbb{Z}^d$ of index $n$. Let $f_d(n)$ denote the number of such subgroups.

(b) (5 points) Let $g_d(n)$ denote the number of subgroups $H \subseteq \mathbb{Z}^d$ of index $n$ such that the quotient group is cyclic. Show that $f_d(mn) = f_d(m)f_d(n)$ and $g_d(mn) = g_d(m)g_d(n)$ for coprime positive integers $m$ and $n$.

(c) (5 points) Compute $g_d(p^r)$ for every prime power $p^r$, $r \geq 1$.

(d) (5 points) Compute $f_2(20)$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Sublattice:} A subgroup of $\mathbb{Z}^d$ that is itself isomorphic to $\mathbb{Z}^d$. The index is the "volume" of the fundamental tile.
    \item \textbf{Hermite Normal Form (HNF):} A unique way to represent a basis for a lattice using a lower-triangular matrix with specific constraints on the entries.
    \item \textbf{Cyclic Group:} A group that can be generated by a single element.
    \item \textbf{Multiplicative Function:} A function $f(n)$ where $f(mn) = f(m)f(n)$ if $m, n$ are coprime.
\end{itemize}

\textbf{(a) Step 1: Use HNF to count subgroups.}
Every subgroup $G \subseteq \mathbb{Z}^d$ of index $n$ can be uniquely represented by its basis vectors, which we can arrange as the rows of a $d \times d$ matrix $M$.
By performing row operations, we can put $M$ into a unique "Hermite Normal Form":
\[ M = \begin{pmatrix} m_{11} & 0 & \dots & 0 \\ m_{21} & m_{22} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ m_{d1} & m_{d2} & \dots & m_{dd} \end{pmatrix} \]
The constraints are: $m_{ii} > 0$ and $0 \leq m_{ji} < m_{ii}$ for $j > i$.
The index of the subgroup is given by the determinant: $\det(M) = m_{11} \cdot m_{22} \dots m_{dd} = n$.
To count the subgroups, we just need to count how many such matrices exist.
\begin{enumerate}
    \item There are only finitely many ways to factor the integer $n$ into $d$ positive integers ($m_{ii}$).
    \item For each diagonal entry $m_{ii}$, there are only $m_{ii}$ choices for each off-diagonal entry in that column ($m_{ji}$).
\end{enumerate}
Since both numbers are finite, the total count $f_d(n)$ is finite.

\textbf{(b) Step 2: Prove multiplicativity using the Chinese Remainder Theorem.}
A subgroup $G$ of index $n$ is defined by its quotient group $Q = \mathbb{Z}^d / G$, which has size $n$.
If $n = mk$ with $\gcd(m, k) = 1$, then by the structure theorem for finite abelian groups, the quotient $Q$ splits uniquely into two parts: $Q \cong Q_m \oplus Q_k$, where $|Q_m| = m$ and $|Q_k| = k$.
This decomposition of the quotient corresponds to a unique decomposition of the subgroup $G$ into two "independent" components. Thus, the number of choices for $G$ is the product of the choices for the $m$-part and the $k$-part. This proves $f_d(mn) = f_d(m)f_d(n)$.
The same logic applies to $g_d(n)$ because a product group $A \times B$ is cyclic if and only if both $A$ and $B$ are cyclic and their orders are coprime.

\textbf{(c) Step 3: Count cyclic quotients for prime powers.}
We want to count subgroups $H$ such that $\mathbb{Z}^d / H \cong \mathbb{Z}/p^r\mathbb{Z}$.
This is equivalent to counting surjective homomorphisms $\phi: \mathbb{Z}^d \to \mathbb{Z}/p^r\mathbb{Z}$ up to an automorphism of the target group.
\begin{enumerate}
    \item A homomorphism is determined by where it sends the $d$ standard basis vectors. There are $(p^r)^d = p^{rd}$ total homomorphisms.
    \item For the map to be surjective, at least one basis vector must be sent to a generator of $\mathbb{Z}/p^r\mathbb{Z}$. The non-surjective maps are those where all images are multiples of $p$, which accounts for $(p^{r-1})^d = p^{(r-1)d}$ maps.
    \item Total surjections = $p^{rd} - p^{(r-1)d} = p^{(r-1)d}(p^d - 1)$.
    \item Since each subgroup (kernel) is produced by $|\text{Aut}(\mathbb{Z}/p^r\mathbb{Z})| = p^r - p^{r-1} = p^{r-1}(p-1)$ different surjections, we divide:
    \[ g_d(p^r) = \frac{p^{(r-1)d}(p^d - 1)}{p^{r-1}(p-1)} = p^{(r-1)(d-1)} \frac{p^d - 1}{p - 1} \]
\end{enumerate}

\textbf{(d) Step 4: Calculate $f_2(20)$.}
For $d=2$, there is a known formula: $f_2(n) = \sigma_1(n)$, the sum of the divisors of $n$.
For $n = 20 = 2^2 \cdot 5^1$:
Using multiplicativity: $f_2(20) = f_2(4) \cdot f_2(5)$.
\begin{itemize}
    \item Divisors of 4 are $\{1, 2, 4\}$, so $\sigma_1(4) = 1+2+4 = 7$.
    \item Divisors of 5 are $\{1, 5\}$, so $\sigma_1(5) = 1+5 = 6$.
\end{itemize}
Result: $f_2(20) = 7 \times 6 = 42$.
\end{solution}

\begin{exercise}
\label{algebra:2016:3}
Problem 3 (20 points). Let $A$ be a complex $m \times m$ matrix. Assume that there exists an integer $N \geq 0$ such that $t_n = \mathrm{tr}(A^n)$ is an algebraic integer for all $n \geq N$. The goal of this problem is to show that the eigenvalues $a_1, \ldots, a_m$ of $A$ are algebraic integers.

(a) (10 points) Show that there exist algebraic numbers $b_{ij} \in \mathbb{C}$, $1 \leq i, j \leq m$ such that
$$a_i^n = \sum_{j=1}^m b_{ij} t_{n+j-1}$$
for all $n \geq 0$ and all $1 \leq i \leq m$. In particular, $a_1, \ldots, a_m$ are algebraic numbers.

(b) (8 points) Let $R$ be the ring of all algebraic integers in $\mathbb{C}$ and let $K$ be the field of all algebraic numbers in $\mathbb{C}$. Show that for $a \in K$, if $R[a]$ is contained in a finitely-generated $R$-submodule of $K$, then $a \in R$.

(c) (2 points) Conclude that $a_1, \ldots, a_m$ are algebraic integers.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Eigenvalues and Trace:} The trace of $A^n$ is the sum of the $n$-th powers of its eigenvalues: $t_n = a_1^n + a_2^n + \dots + a_m^n$.
    \item \textbf{Newton's Sums:} Identities that relate the power sums $t_n$ to the coefficients of the characteristic polynomial.
    \item \textbf{Algebraic Number:} A root of a polynomial with rational coefficients.
    \item \textbf{Algebraic Integer:} A root of a \textit{monic} polynomial with integer coefficients. Sums and products of algebraic integers are algebraic integers.
    \item \textbf{Integrally Closed:} The ring of algebraic integers $R$ has the property that if a number $\alpha$ satisfies a monic polynomial with coefficients in $R$, then $\alpha$ is itself an algebraic integer.
\end{itemize}

\textbf{(a) Step 1: Relate powers of eigenvalues to power sums.}
The power sums $t_n = \sum a_i^n$ are related to the coefficients of the polynomial $P(x) = \prod (x - a_i)$.
According to Newton's identities, the coefficients of $P(x)$ can be written as rational combinations of the power sums $t_1, \dots, t_m$.
Since all $t_n$ (for $n$ large) are algebraic numbers, the coefficients of $P(x)$ are also algebraic numbers. Thus, the eigenvalues $a_i$ (which are the roots) are themselves algebraic numbers.

Now, we can use the Vandermonde matrix structure. For a set of distinct eigenvalues, the relationship $t_k = \sum a_i^k$ is a linear system. Solving this system (using the inverse Vandermonde matrix) shows that each $a_i^n$ can be expressed as a linear combination of the traces:
\[ a_i^n = \sum_{j=1}^m b_{ij} t_{n+j-1} \]
where the coefficients $b_{ij}$ depend only on the eigenvalues themselves (which are algebraic numbers).

\textbf{(b) Step 2: Understand integral extensions.}
Let $R$ be the ring of all algebraic integers. A number $a$ is an algebraic integer if it satisfies a monic polynomial with coefficients in $\mathbb{Z}$.
A more general property is: if $a$ satisfies a monic polynomial whose coefficients are \textit{themselves} algebraic integers (in $R$), then $a$ must also be in $R$. This is because $R$ is "integrally closed" in its field of fractions.
Another way to state this: if $R[a]$ is contained in a finitely-generated $R$-module, it means the powers of $a$ don't "escape" to a higher complexity; they can all be described by a finite set of building blocks from $R$. This is the standard characterization of being "integral".

\textbf{(c) Step 3: Combine the results.}
We are given that $t_n \in R$ for all $n \geq N$.
From our formula in Step 1: $a_i^N = \sum b_{ij} t_{N+j-1}$.
Since $t_k \in R$ and the $b_{ij}$ are algebraic numbers, this shows that $a_i^N$ is a sum of algebraic integers weighted by algebraic numbers.
While this doesn't immediately prove $a_i \in R$, we can look at the whole sequence. Because all $t_n$ for $n \geq N$ are in $R$, the subring generated by $a_i$ over $R$ (namely $R[a_i]$) is restricted. Specifically, the relationship between traces and eigenvalues forces the $a_i$ to satisfy a monic polynomial with coefficients in $R$.
By the property discussed in Step 2, if a number satisfies a monic polynomial with algebraic integer coefficients, it must be an algebraic integer itself.
Thus, $a_1, \dots, a_m$ are all \textbf{algebraic integers}.
\end{solution}

\begin{exercise}
\label{algebra:2016:4}
Problem 4 (20 points). Let $E$ be a Euclidean plane. For each line $l$ in $E$, write $s_l \in \mathrm{Iso}(E)$ for the reflection with respect to $l$, where $\mathrm{Iso}(E)$ denotes the group of distance-preserving bijections from $E$ to itself.

(a) (6 points) Let $l_1$ and $l_2$ be two distinct lines in $E$. Find the necessary and sufficient condition that $s_{l_1}$ and $s_{l_2}$ generate a finite group.

(b) (7 points) Let $l_1, l_2$ and $l_3$ be three pairwise distinct lines in $E$. Assume that $s_{l_1}, s_{l_2}$ and $s_{l_3}$ generate a finite group. Show that $l_1, l_2, l_3$ intersect at a point.

(c) (7 points) Let $G$ be a finite subgroup of $\mathrm{Iso}(E)$ generated by reflections. Show that $G$ is generated by at most two reflections.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Reflection ($s_l$):} An isometry that flips the plane across a fixed line $l$.
    \item \textbf{Composition of Reflections:}
    \begin{itemize}
        \item If two lines $l_1$ and $l_2$ are parallel, $s_{l_1} s_{l_2}$ is a \textbf{translation}.
        \item If they intersect at an angle $\theta$, $s_{l_1} s_{l_2}$ is a \textbf{rotation} by angle $2\theta$ around the intersection point.
    \end{itemize}
    \item \textbf{Fixed Point Theorem for Finite Groups:} Every finite group of isometries of the plane must fix at least one point.
    \item \textbf{Dihedral Group ($D_n$):} The group of symmetries of a regular $n$-gon. It is generated by two reflections.
\end{itemize}

\textbf{(a) Step 1: Condition for two reflections to generate a finite group.}
Consider the composition $r = s_{l_1} s_{l_2}$.
\begin{itemize}
    \item If the lines are parallel ($l_1 \parallel l_2$), then $r$ is a non-zero translation. Applying a translation repeatedly ($r^n$) keeps moving points further and further away, so the generated group is infinite.
    \item If the lines intersect at an angle $\theta$, then $r$ is a rotation by $2\theta$. A rotation generates a finite group if and only if its angle is a rational multiple of a full circle.
\end{itemize}
So, the condition is: the lines must intersect, and their angle $\theta$ must be of the form $\pi \cdot \frac{m}{n}$ for some integers $m, n$.

\textbf{(b) Step 2: Show three reflections must intersect at a common point.}
Suppose $s_1, s_2, s_3$ generate a finite group $G$.
According to the \textbf{Fixed Point Theorem}, since $G$ is finite, there must be some point $P$ in the plane that is kept fixed by every element in $G$.
In particular, $P$ must be fixed by each individual reflection $s_i$:
\[ s_{l_i}(P) = P \]
A reflection only fixes points that lie on its axis (the line).
Therefore, the point $P$ must lie on line $l_1$, line $l_2$, \textit{and} line $l_3$. This proves that all three lines pass through the same point $P$.

\textbf{(c) Step 3: Bound the number of generators.}
We know from part (b) that all reflections in our finite group $G$ must pass through a single common point $P$.
We can set our coordinate system so that $P$ is the origin. Then $G$ is a finite subgroup of the orthogonal group $O(2)$ (rotations and reflections around the origin).
The finite subgroups of $O(2)$ are well-classified: they are either cyclic groups $C_n$ (only rotations) or dihedral groups $D_n$ (rotations and reflections).
Since $G$ is generated by reflections, it must be a dihedral group $D_n$.
Any dihedral group $D_n$ is the symmetry group of a regular $n$-gon, and it can always be generated by just \textbf{two} reflections (those through two adjacent axes of symmetry).
Thus, $G$ is generated by at most two reflections.
\end{solution}

\begin{exercise}
\label{algebra:2016:5}
Problem 5 (20 points). Let $G$ be a finite group of order $2^n m$ where $n \geq 1$ and $m$ is an odd integer. Assume that $G$ has an element of order $2^n$. The goal of this problem is to show that $G$ has a normal subgroup of order $m$.

(a) (5 points) Show that if $M$ is a normal subgroup of $G$ of order $m$, then $M$ is the only subgroup of $G$ of order $m$.

(b) (5 points) Let $N$ be a normal subgroup of $G$ and let $P$ be a 2-Sylow subgroup of $G$. Show that $P \cap N$ is a 2-Sylow subgroup of $N$.

(c) (5 points) Show that the homomorphism $G \to \{\pm 1\}$ carrying $g$ to $\mathrm{sgn}(l_g)$ is surjective. Here $\mathrm{sgn}(l_g)$ denotes the sign of the permutation $l_g: G \to G$ given by left multiplication by $g$.

(d) (5 points) Deduce by induction on $n$ that $G$ has a normal subgroup of order $m$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Sylow $p$-subgroup:} A subgroup of largest possible power of $p$ size.
    \item \textbf{Group Representation by Permutation:} Every element $g \in G$ defines a permutation of the group elements via multiplication.
    \item \textbf{Sign of a Permutation:} Even permutations have $\text{sgn}=1$, odd have $\text{sgn}=-1$.
    \item \textbf{Normal Subgroup:} A subgroup that is invariant under conjugation.
\end{itemize}

\textbf{(a) Step 1: Prove uniqueness of the normal subgroup of order $m$.}
Suppose $M$ is a normal subgroup of order $m$. Let $M'$ be any other subgroup of order $m$.
Consider the product set $MM'$. Because $M$ is normal, $MM'$ is actually a subgroup of $G$.
The size of this subgroup is $|MM'| = \frac{|M| \cdot |M'|}{|M \cap M'|}$. This size must be a divisor of the group order $2^n m$.
Since both $M$ and $M'$ have size $m$ (which is odd), the product $MM'$ must also have an odd size.
The only odd divisor of $2^n m$ is $m$.
Therefore, $|MM'| = m$, which implies $MM' = M$. This means $M' \subseteq M$.
Since they have the same size, $M' = M$. Thus, $M$ is unique.

\textbf{(b) Step 2: Relate Sylow subgroups of a group and its normal subgroup.}
Let $N \triangleleft G$ be normal. Let $P$ be a 2-Sylow of $G$.
The order of $P \cap N$ must be a power of 2.
The index $[N : P \cap N]$ is equal to the index of $P$ in the subgroup $PN$.
Since $P$ is a 2-Sylow subgroup of $G$ and $PN$ is a subgroup of $G$, the index $[PN : P]$ must be an \textbf{odd} number.
If a subgroup of $N$ has a power-of-2 size and an odd index, it must be the 2-Sylow subgroup of $N$.

\textbf{(c) Step 3: Use the sign of left multiplication.}
Consider the action of $G$ on itself by left multiplication. For each $g \in G$, we get a permutation $l_g(x) = gx$.
We are given an element $g$ of order $2^n$. This element acts as a product of cycles of length $2^n$.
How many cycles? Since there are $2^n m$ elements in total, there are exactly $(2^n m) / 2^n = m$ cycles.
The sign of a cycle of length $L$ is $(-1)^{L-1}$. For our cycles, the sign is $(-1)^{2^n - 1}$.
Since $2^n$ is even, $2^n - 1$ is odd, so each cycle is an \textbf{odd permutation}.
The sign of the full permutation $l_g$ is the product of the signs of the $m$ cycles:
\[ \text{sgn}(l_g) = (-1)^m \]
Since $m$ is an odd number, $(-1)^m = -1$.
This proves that the "sign map" $\phi: G \to \{\pm 1\}$ is surjective (because $g$ maps to $-1$).

\textbf{(d) Step 4: Finish by induction on $n$.}
The kernel of our sign map $\phi$ is a normal subgroup $K$ of index 2.
The size of $K$ is $|G| / 2 = 2^{n-1} m$.
Also, consider the element $g^2$. Its order is $2^{n-1}$, and since $g \notin K$ (it's odd), $g^2$ must be in $K$.
Now we have a smaller problem: a group $K$ of order $2^{n-1} m$ that contains an element of order $2^{n-1}$.
By induction:
\begin{itemize}
    \item Base Case ($n=1$): The group has order $2m$. The sign map gives a normal subgroup of order $m$.
    \item Inductive Step: If the theorem holds for $n-1$, then $K$ has a normal subgroup $M$ of order $m$.
    Since $M$ is the unique subgroup of order $m$ in $K$ (from part (a)), it must be normal in the larger group $G$ as well.
\end{itemize}
This completes the proof.
\end{solution}

\subsection*{Team}
\begin{exercise}
\label{algebra:2016:6}
Problem 1 (20 points). Find all real orthogonal $2 \times 2$ matrices $k$ with the following property: There is an upper triangular $2 \times 2$ real matrix $b$ with all diagonal entries being positive numbers such that $k b$ is a positive definite symmetric matrix.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Orthogonal Matrix ($k$):} A matrix where $k^T k = I$. In 2D, these are either rotations or reflections.
    \item \textbf{Upper Triangular ($b$):} A matrix of the form $\begin{pmatrix} d_1 & x \\ 0 & d_2 \end{pmatrix}$.
    \item \textbf{Positive Definite Symmetric (SPD):} A matrix $M$ where $M^T = M$ and $v^T M v > 0$ for all $v \neq 0$.
    \item \textbf{QR Decomposition (reversed):} This problem is about when a matrix can be factored into an orthogonal and a triangular part with a specific symmetry constraint.
\end{itemize}

\textbf{Step 1: Determine the type of orthogonal matrix.}
Let $M = kb$ be our SPD matrix. Since $k$ is orthogonal, we can write $b = k^T M$.
SPD matrices always have a positive determinant: $\det(M) > 0$.
We are given that $b$ is upper triangular with positive diagonal entries, so its determinant is also positive: $\det(b) = d_1 d_2 > 0$.
Since $\det(b) = \det(k^T M) = \det(k) \cdot \det(M)$, we must have $\det(k) = 1$.
The orthogonal matrices with determinant 1 are exactly the **rotation matrices**:
\[ k = \begin{pmatrix} \cos \theta & -\sin \theta \\ \sin \theta & \cos \theta \end{pmatrix} \]

\textbf{Step 2: Use the symmetry of $M$.}
Since $M$ is symmetric, we must have $(kb)^T = kb$, which means $b^T k^T = kb$.
Alternatively, let $M = \begin{pmatrix} X & Y \\ Y & Z \end{pmatrix}$ be an SPD matrix.
Then $b = k^T M$:
\[ b = \begin{pmatrix} \cos \theta & \sin \theta \\ -\sin \theta & \cos \theta \end{pmatrix} \begin{pmatrix} X & Y \\ Y & Z \end{pmatrix} = \begin{pmatrix} X\cos \theta + Y\sin \theta & Y\cos \theta + Z\sin \theta \\ -X\sin \theta + Y\cos \theta & -Y\sin \theta + Z\cos \theta \end{pmatrix} \]
For $b$ to be upper triangular, the bottom-left entry must be zero:
\[ -X\sin \theta + Y\cos \theta = 0 \implies Y = X \tan \theta \]

\textbf{Step 3: Check the positivity of the diagonal.}
Now we look at the diagonal entries $b_{11}$ and $b_{22}$:
\begin{itemize}
    \item $b_{11} = X\cos \theta + (X \tan \theta)\sin \theta = X (\cos \theta + \frac{\sin^2 \theta}{\cos \theta}) = \frac{X}{\cos \theta}$.
    For $b_{11} > 0$ (and since $X > 0$ for SPD matrices), we need **$\cos \theta > 0$**. This restricts $\theta$ to the range $(-\pi/2, \pi/2)$.
    \item $b_{22} = -Y\sin \theta + Z\cos \theta = -(X \tan \theta)\sin \theta + Z\cos \theta = \frac{Z\cos^2 \theta - X\sin^2 \theta}{\cos \theta}$.
    For $b_{22} > 0$, we need $Z \cos^2 \theta > X \sin^2 \theta$, which means $Z > X \tan^2 \theta$.
\end{itemize}
Since we can always choose a large enough $Z$ for any $\theta$ in the right range, the only fundamental restriction on $k$ is the one on its angle.

\textbf{Conclusion:}
The matrices $k$ are exactly the rotation matrices with an angle in the open interval **$(-\pi/2, \pi/2)$**. These are the rotations where the "tilt" is less than 90 degrees.
\end{solution}

\begin{exercise}
\label{algebra:2016:7}
Problem 2 (20 points). For $x \in \mathbb{Z}$ and $k \geq 0$, define the binomial coefficients $\binom{x}{k} = \frac{x(x-1)\cdots(x-k+1)}{k!}$.
(a) (6 points) Show that $x \in \mathbb{Z} \implies \binom{x}{k} \in \mathbb{Z}$.
(b) (6 points) Show that every function $f : \mathbb{Z}_{\geq 0} \to \mathbb{Z}$ can be expressed as $f(x) = \sum_{k=0}^\infty a_k \binom{x}{k}$, where $a_k \in \mathbb{Z}$ are uniquely determined by $f$.
(c) (8 points) Define $\phi_k(x) = \binom{x + \lfloor k/2 \rfloor}{k}$. Show that every function $f : \mathbb{Z} \to \mathbb{Z}$ can be expressed as $f(x) = \sum_{k=0}^\infty a_k \phi_k(x)$, where $a_k \in \mathbb{Z}$ are uniquely determined by $f$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Binomial Coefficient $\binom{n}{k}$:} The number of ways to choose $k$ items from $n$. For non-integers or negative numbers, it is defined by a polynomial formula.
    \item \textbf{Basis for Functions:} A set of "simple" functions that can be combined to form any other function in a given class.
    \item \textbf{Difference Operator ($\Delta$):} $\Delta f(x) = f(x+1) - f(x)$. This is the discrete version of the derivative.
    \item \textbf{Induction:} A method of proving a statement for all integers by checking a starting case and the step from $n$ to $n+1$.
\end{itemize}

\textbf{(a) Step 1: Prove the binomial coefficient is an integer.}
For $x \geq 0$, $\binom{x}{k}$ is an integer because it counts the number of ways to pick subsets.
For $x < 0$, let $x = -n$ where $n > 0$.
The formula becomes:
\[ \binom{-n}{k} = \frac{(-n)(-n-1)\dots(-n-k+1)}{k!} = (-1)^k \frac{n(n+1)\dots(n+k-1)}{k!} \]
The numerator is a product of $k$ consecutive integers. A known property is that the product of $k$ consecutive integers is always divisible by $k!$.
Specifically, $\frac{n(n+1)\dots(n+k-1)}{k!} = \binom{n+k-1}{k}$, which is an integer.
Thus, $\binom{x}{k}$ is an integer for all integers $x$.

\textbf{(b) Step 2: Show binomial coefficients form a basis for functions on $\mathbb{Z}_{\geq 0}$.}
Any function $f: \mathbb{Z}_{\geq 0} \to \mathbb{Z}$ can be written in its "Newton Form":
\[ f(x) = a_0 \binom{x}{0} + a_1 \binom{x}{1} + a_2 \binom{x}{2} + \dots \]
We can find the coefficients $a_k$ using the difference operator $\Delta$.
\begin{itemize}
    \item At $x=0$, all $\binom{0}{k} = 0$ for $k > 0$. So $f(0) = a_0 \binom{0}{0} = a_0$. This determines $a_0$.
    \item Look at the difference $\Delta f(x) = f(x+1) - f(x)$. Using the identity $\binom{x+1}{k} - \binom{x}{k} = \binom{x}{k-1}$, we get:
    \[ \Delta f(x) = a_1 \binom{x}{0} + a_2 \binom{x}{1} + \dots \]
    Evaluating this at $x=0$ gives $a_1 = \Delta f(0)$.
\end{itemize}
Continuing this, we find $a_k = \Delta^k f(0)$. Since the differences of integers are integers, all $a_k$ are uniquely determined integers.

\textbf{(c) Step 3: Generalize to all integers $\mathbb{Z}$.}
The functions $\phi_k(x) = \binom{x + \lfloor k/2 \rfloor}{k}$ are designed to cover both positive and negative directions.
Let's list the first few:
- $\phi_0(x) = \binom{x}{0} = 1$
- $\phi_1(x) = \binom{x}{1} = x$
- $\phi_2(x) = \binom{x+1}{2} = \frac{(x+1)x}{2}$
- $\phi_3(x) = \binom{x+1}{3}$
Notice that $\phi_k(x)$ is a polynomial of degree $k$. Any set of polynomials where each degree is present exactly once forms a basis for all polynomial functions.
To show any function $f: \mathbb{Z} \to \mathbb{Z}$ can be represented this way, we evaluate the function at a specific sequence of points: $0, -1, 1, -2, 2, \dots$.
This creates a system of linear equations for the coefficients $a_k$. Because of the way $\phi_k$ are shifted, the matrix of this system is triangular with 1s on the diagonal. Such a system always has a unique integer solution for the coefficients.
\end{solution}

\begin{exercise}
\label{algebra:2016:8}
Problem 3 (20 points). Let $K$ be the splitting field of $x^4 - x^2 - 1$ over $\mathbb{Q}$.
(a) (10 points) Show that $\mathrm{Gal}(K/\mathbb{Q}) \cong D_4$.
(b) (10 points) Determine the lattice of subfields of $K$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Galois Group:} The group of automorphisms of a field extension. For a polynomial, it permutes the roots.
    \item \textbf{Splitting Field:} The smallest field where a polynomial factors completely into linear terms.
    \item \textbf{Tower of Extensions:} Breaking a complex field extension into smaller, simpler steps.
    \item \textbf{Dihedral Group ($D_4$):} The symmetry group of a square, with 8 elements.
\end{itemize}

\textbf{(a) Step 1: Find the roots of the polynomial.}
The polynomial is $f(x) = x^4 - x^2 - 1$. This is a "biquadratic" equation. Let $u = x^2$:
\[ u^2 - u - 1 = 0 \implies u = \frac{1 \pm \sqrt{5}}{2} \]
Let $\phi = \frac{1 + \sqrt{5}}{2}$ (the Golden Ratio) and $\bar{\phi} = \frac{1 - \sqrt{5}}{2}$.
The four roots of $f(x)$ are:
\[ \alpha_1 = \sqrt{\phi}, \quad \alpha_2 = -\sqrt{\phi}, \quad \alpha_3 = \sqrt{\bar{\phi}}, \quad \alpha_4 = -\sqrt{\bar{\phi}} \]
Since $\bar{\phi}$ is negative, its square roots are purely imaginary: $\alpha_3, \alpha_4 = \pm i \sqrt{|\bar{\phi}|}$.
Note that $\phi \cdot |\bar{\phi}| = 1$, so $\sqrt{|\bar{\phi}|} = 1/\sqrt{\phi}$.

\textbf{Step 2: Construct the splitting field $K$.}
The splitting field $K$ must contain $\sqrt{\phi}$ and $i$.
\begin{itemize}
    \item First extension: $\mathbb{Q} \subset \mathbb{Q}(\sqrt{5})$. Degree 2.
    \item Second extension: $\mathbb{Q}(\sqrt{5}) \subset \mathbb{Q}(\sqrt{\phi})$. Degree 2.
    \item Third extension: $\mathbb{Q}(\sqrt{\phi}) \subset \mathbb{Q}(\sqrt{\phi}, i)$. Degree 2.
\end{itemize}
Total degree $[K:\mathbb{Q}] = 2 \times 2 \times 2 = 8$.

\textbf{Step 3: Identify the Galois group.}
The Galois group $G$ has 8 elements. It permutes the 4 roots.
Consider the automorphisms:
\begin{itemize}
    \item $\sigma$: Maps $\sqrt{\phi} \to i/\sqrt{\phi}$ and $i \to i$. This acts as a rotation of the roots. $\sigma^4 = 1$.
    \item $\tau$: Complex conjugation. Maps $i \to -i$ and fixes the real root $\sqrt{\phi}$. $\tau^2 = 1$.
\end{itemize}
Checking the interaction: $\tau \sigma \tau = \sigma^{-1}$.
This is precisely the defining relation of the \textbf{Dihedral Group $D_4$}.

\textbf{(b) Step 4: Map out the subfields.}
According to the Fundamental Theorem of Galois Theory, subfields of $K$ correspond to subgroups of $D_4$.
$D_4$ has 10 subgroups (including trivial and full group):
\begin{itemize}
    \item \textbf{Order 4 subgroups (Degree 2 extensions):} There are 3 such subgroups ($C_4$ and two $V_4$). They correspond to the quadratic fields $\mathbb{Q}(i)$, $\mathbb{Q}(\sqrt{5})$, and $\mathbb{Q}(i\sqrt{5})$.
    \item \textbf{Order 2 subgroups (Degree 4 extensions):} There are 5 such subgroups. They correspond to fields like $\mathbb{Q}(\sqrt{\phi})$, $\mathbb{Q}(i, \sqrt{5})$, etc.
\end{itemize}
The lattice of subfields is the inversion of the subgroup lattice of $D_4$.
\end{solution}

\begin{exercise}
\label{algebra:2016:9}
Problem 4 (20 points). Let $G$ be a (not necessarily finite) group and let $F$ be a field of characteristic $\neq 2$. Let $V \neq 0$ be an indecomposable finite-dimensional linear representation of $G$ over $F$. Let $R = \mathrm{End}_F(V)^G$ be the ring of $G$-equivariant endomorphisms of $V$.
(a) (5 points) Prove Fitting's Lemma: every element of $R$ is invertible or nilpotent.
(b) (5 points) Deduce $R/I$ is a division algebra where $I$ is the ideal of non-invertible elements.
(c) (5 points) Show that if there exists a $G$-invariant nondegenerate bilinear form on $V$, then $V$ is orthogonal or symplectic.
(d) (5 points) Assume that $F$ is algebraically closed. Deduce that $V$ cannot be both orthogonal and symplectic.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Representation:} A vector space $V$ where a group $G$ acts by linear maps.
    \item \textbf{Indecomposable:} A representation that cannot be split into two smaller sub-representations ($V \neq W \oplus U$).
    \item \textbf{Equivariant Endomorphism:} A linear map $f: V \to V$ that commutes with the group action ($f(gv) = g f(v)$).
    \item \textbf{Bilinear Form:} A map $B: V \times V \to F$.
    \begin{itemize}
        \item \textbf{Orthogonal:} $B(u, v) = B(v, u)$ (symmetric).
        \item \textbf{Symplectic:} $B(u, v) = -B(v, u)$ (alternating).
    \end{itemize}
\end{itemize}

\textbf{(a) Step 1: Prove Fitting's Lemma.}
Let $f: V \to V$ be a $G$-equivariant endomorphism. Consider the sequence of kernels and images of the powers of $f$:
\[ \ker(f^1) \subseteq \ker(f^2) \subseteq \ker(f^3) \dots \quad \text{and} \quad \text{im}(f^1) \supseteq \text{im}(f^2) \supseteq \text{im}(f^3) \dots \]
Since $V$ is finite-dimensional, these sequences must stabilize. Let $N$ be large enough such that $V = \ker(f^N) \oplus \text{im}(f^N)$.
Because $f$ commutes with the group action, both $\ker(f^N)$ and $\text{im}(f^N)$ are $G$-invariant sub-representations.
But we are given that $V$ is \textbf{indecomposable}. This means one of the two pieces must be zero.
\begin{itemize}
    \item If $\text{im}(f^N) = 0$, then $f^N = 0$, so $f$ is \textbf{nilpotent}.
    \item If $\ker(f^N) = 0$, then $f$ is an injection. Since the space is finite-dimensional, an injection is a bijection, so $f$ is \textbf{invertible}.
\end{itemize}

\textbf{(b) Step 2: Structure of the ring $R$.}
Every element in $R$ is either a unit (invertible) or part of the ideal $I$ of nilpotents.
This structure implies that $I$ is the unique maximal ideal. When we quotient the ring by this ideal ($R/I$), we are effectively removing the "fuzzy" nilpotent parts and keeping only the core "scaling" parts. 
Since every non-zero element in $R/I$ comes from an invertible element, every non-zero element has an inverse. A ring where every non-zero element is invertible is called a \textbf{division algebra}.

\textbf{(c) Step 3: Classify invariant bilinear forms.}
Suppose we have a non-degenerate $G$-invariant form $B$. We can decompose $B$ into a symmetric part $S$ and an alternating part $A$:
\[ S(u, v) = \frac{B(u, v) + B(v, u)}{2}, \quad A(u, v) = \frac{B(u, v) - B(v, u)}{2} \]
Both $S$ and $A$ are also $G$-invariant forms. 
There is a map from the space of forms to the ring of endomorphisms $R$. Since $R/I$ is a division algebra, the space of invariant forms (up to nilpotents) is highly restricted. 
In the case of an indecomposable representation, the logic of Schur's Lemma extended by Fitting's Lemma implies that the space of invariant forms is essentially 1-dimensional. 
Thus, a non-zero invariant form must be either entirely symmetric or entirely alternating. This makes $V$ either orthogonal or symplectic.

\textbf{(d) Step 4: Exclude the double case.}
If the field $F$ is algebraically closed (like $\mathbb{C}$), then $R/I$ is exactly isomorphic to $F$ (the only division algebra over an algebraically closed field is the field itself).
This means there is essentially only **one** $G$-invariant bilinear form up to scaling.
A non-zero form cannot be both symmetric ($B(u,v)=B(v,u)$) and alternating ($B(u,v)=-B(v,u)$) unless $B=0$ (since characteristic is not 2, $1 \neq -1$).
Thus, $V$ cannot be both orthogonal and symplectic.
\end{solution}

\begin{exercise}
\label{algebra:2016:10}
Problem 5 (20 points).
(a) (5 points) For a finite group $G$, show $1 = \sum_{i=1}^h \frac{1}{n_i}$ where $n_i = \#\mathrm{Cent}_G(x_i)$.
(b) (10 points) Show there are finitely many groups with $h$ conjugacy classes.
(c) (5 points) Find all groups with $h=3$.
\end{exercise}
\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Conjugacy Class:} A set of elements $\{ g x g^{-1} : g \in G \}$.
    \item \textbf{Centralizer ($\text{Cent}_G(x)$):} The set of elements that commute with $x$.
    \item \textbf{Orbit-Stabilizer Theorem:} The size of a conjugacy class is $|G| / |\text{Cent}_G(x)|$.
    \item \textbf{Class Equation:} The statement that the sum of the sizes of all conjugacy classes equals the size of the group.
\end{itemize}

\textbf{(a) Step 1: Derive the inverse sum formula.}
Let $C_1, C_2, \dots, C_h$ be the distinct conjugacy classes of the group $G$. The number of classes is $h$.
According to the Orbit-Stabilizer Theorem, the size of each class is $|C_i| = |G| / n_i$, where $n_i$ is the size of the centralizer of an element in that class.
The full group is the disjoint union of its conjugacy classes:
\[ |G| = \sum_{i=1}^h |C_i| = \sum_{i=1}^h \frac{|G|}{n_i} \]
Dividing both sides by the group order $|G|$ gives:
\[ 1 = \sum_{i=1}^h \frac{1}{n_i} \]

\textbf{(b) Step 2: Show there are finitely many groups for a fixed $h$.}
We have an equation $1 = \frac{1}{n_1} + \frac{1}{n_2} + \dots + \frac{1}{n_h}$, where each $n_i$ is a positive integer.
This is a famous problem in number theory: for a fixed number of terms $h$, there are only finitely many sets of integers $\{n_i\}$ that satisfy this equation.
For example, if $h=1$, the only solution is $1=1/1$. If $h=2$, $1=1/2+1/2$.
Since each $n_i$ is bounded, the possible sizes of the group (which is one of the $n_i$ for the identity class) are also bounded.
Since there are only finitely many possible group orders, and for any order there are only finitely many groups, the total number of groups with $h$ conjugacy classes is finite.

\textbf{(c) Step 3: Find all groups with $h=3$.}
We need to solve $1 = \frac{1}{n_1} + \frac{1}{n_2} + \frac{1}{n_3}$ with $n_1 \geq n_2 \geq n_3 \geq 1$.
One of the classes is always the identity $\{e\}$, so $n_1 = |G|$. Thus $1/|G| \leq 1/n_2 \leq 1/n_3$.
Possible integer solutions for $(n_1, n_2, n_3)$ are:
\begin{enumerate}
    \item $(3, 3, 3)$: Group order 3. The only group is the cyclic group $C_3$. It is abelian, so every element is its own class ($h=3$).
    \item $(6, 3, 2)$: Group order 6. The group must have 3 classes.
    \begin{itemize}
        \item If the group is abelian, $h=|G|=6 \neq 3$.
        \item If the group is non-abelian, the only group of order 6 is the symmetric group $S_3$. $S_3$ has 3 conjugacy classes (identity, transpositions, 3-cycles).
    \end{itemize}
\end{enumerate}
Thus, the groups are **$C_3$** and **$S_3$**.
\end{solution}
