\section{2015 Algebra \& Number Theory}

\subsection*{Individual}
\begin{exercise}
Let $G$ be a finite $\mathbb{Z}$-module (i.e., a finite abelian group) with a bilinear, (strongly) alternative, and non-degenerate pairing
\[
\ell : G \times G \to \mathbb{Q} / \mathbb{Z}.
\]
Here "(strongly) alternating" means for every $a \in G$, $\ell(a, a) = 0$; "non-degenerate" means for every nonzero $a \in G$ there is a $b \in G$ such that $\ell(a, b) \neq 0$. Show in steps the following statement:
($S$): $G$ is isomorphic to $H_1 \oplus H_2$ for some finite abelian groups $H_1 \simeq H_2$ such that $\ell|_{H_i \times H_i} = 0$.
\begin{enumerate}
    \item For every $a \in G$, write $o(a)$ for the order of $a$ and $\ell_a : G \to \mathbb{Q} / \mathbb{Z}$ for the homomorphism $\ell_a(b) = \ell(a, b)$. Show that the image of $\ell_a$ is $o(a)^{-1}\mathbb{Z} / \mathbb{Z}$.
    \item Show that $G$ has a pair of elements $a, b$ with the following properties:
    \begin{enumerate}
        \item $o(a)$ is maximal in the sense that for any $x \in G$, $o(x) \mid o(a)$;
        \item $\ell(a, b) = o(a)^{-1} \pmod{\mathbb{Z}}$;
        \item $o(a) = o(b)$.
    \end{enumerate}
    We call the subgroup $\langle a, b \rangle := \mathbb{Z}a + \mathbb{Z}b$ a maximal hyperbolic subgroup of $G$.
    \item Let $\langle a, b \rangle$ be a maximal hyperbolic subgroup of $G$. Let $G'$ be the orthogonal complement of $\langle a, b \rangle$ consisting of elements $x \in G$ such that $\ell(x, c) = 0$ for all $c \in \langle a, b \rangle$. Show that $G$ is a direct sum as follows:
    \[
    G = \mathbb{Z}a \oplus \mathbb{Z}b \oplus G'.
    \]
    \item Finish the proof of ($S$) by induction.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Finite Abelian Group ($\mathbb{Z}$-module):} A group where order of elements matters. Every element has a finite order $n$ such that $na = 0$.
    \item \textbf{Symplectic Pairing:} A bilinear map $\ell(a, b)$ that is alternating ($\ell(a, a) = 0$) and non-degenerate (every $a$ has some $b$ that "detects" it).
    \item \textbf{Dual Group ($\widehat{G}$):} The group of homomorphisms from $G$ to $\mathbb{Q}/\mathbb{Z}$. For finite groups, $G \cong \widehat{G}$.
    \item \textbf{Hyperbolic Subgroup:} A pair of elements $(a, b)$ that span a subspace where the pairing behaves like the standard matrix $\begin{pmatrix} 0 & 1 \\ -1 & 0 \end{pmatrix}$.
\end{itemize}

\textbf{(1) Step 1: Determine the image of the mapping $\ell_a$.}
For a fixed $a$, let its order be $o(a) = M$. This means $Ma = 0$ in the group $G$.
Since the pairing $\ell$ is bilinear, we have $M \cdot \ell(a, b) = \ell(Ma, b) = \ell(0, b) = 0$ in the group $\mathbb{Q}/\mathbb{Z}$.
The elements in $\mathbb{Q}/\mathbb{Z}$ that satisfy $Mx = 0$ are the fractions of the form $k/M$. So the image $\text{im}(\ell_a)$ is a subgroup of $\frac{1}{M}\mathbb{Z}/\mathbb{Z}$.
Because the pairing is non-degenerate, the map $a \mapsto \ell_a$ is an isomorphism between $G$ and its dual. This preserves the order of elements. Since $a$ has order $M$, the map $\ell_a$ must also have order $M$. The only subgroup of $\mathbb{Q}/\mathbb{Z}$ with order $M$ is exactly $\{ 0, 1/M, 2/M, \dots, (M-1)/M \}$. This is the group $\frac{1}{M}\mathbb{Z}/\mathbb{Z}$.

\textbf{(2) Step 2: Find a "maximal" hyperbolic pair.}
Let $M$ be the \textit{exponent} of the group (the maximum order of any element). Pick $a \in G$ with this maximal order $o(a) = M$.
From Step 1, the image of $\ell_a$ contains the value $1/M$. This means there must be some element $b \in G$ such that:
\[ \ell(a, b) = \frac{1}{M} \pmod{\mathbb{Z}} \]
Does $b$ also have order $M$?
\begin{itemize}
    \item The map $\ell_a$ takes $b$ to $1/M$, which has order $M$. Since the order of the image must divide the order of the element, $M$ divides $o(b)$.
    \item But $M$ was chosen to be the maximum possible order in the group.
\end{itemize}
Therefore, $o(b)$ must be exactly $M$. We have found our hyperbolic pair $(a, b)$.

\textbf{(3) Step 3: Decompose the group into a direct sum.}
Let $S = \langle a, b \rangle$ be the subgroup generated by our pair. We want to show $G = S \oplus S^\perp$.
Any element $x \in G$ can be "projected" onto $S$. Specifically, we can find a linear combination $c = x_a a + x_b b$ such that $\ell(x-c, a) = 0$ and $\ell(x-c, b) = 0$.
Solving these equations:
$\ell(x, a) - x_b \ell(b, a) = 0 \implies \ell(x, a) + x_b/M = 0$.
Since $\ell(x, a)$ is a multiple of $1/M$, we can always find an integer $x_b$ that works.
The intersection $S \cap S^\perp$ is trivial because if $sa+tb$ is in the complement, its pairing with $a$ and $b$ must be zero, which forces $s$ and $t$ to be multiples of $M$.
Since $S$ is isomorphic to $\mathbb{Z}/M \oplus \mathbb{Z}/M$, we have separated a "block" of the group.

\textbf{(4) Step 4: Complete the proof by induction.}
The remaining group $G'$ (the orthogonal complement) is smaller than $G$, and it inherits the same properties: it is still finite, abelian, and the pairing restricted to it is still alternating and non-degenerate.
By applying the same process to $G'$, we can keep pulling out pairs of $\mathbb{Z}/M_i \oplus \mathbb{Z}/M_i$ until nothing is left.
Finally, if we group all the "first" elements of the pairs into $H_1$ and all the "second" elements into $H_2$, we get the desired decomposition $G \cong H_1 \oplus H_2$ where $H_1 \cong H_2$ and the pairing within each $H_i$ is zero (since $\ell(a, a) = 0$ and we used different pairs for different $i$).
\end{solution}

\begin{exercise}
Let $O_n(\mathbb{C})$ denote the group of $n \times n$ orthogonal complex matrices, and $M_{n \times k}(\mathbb{C})$ the space of $n \times k$ complex matrices. For $i = 0, 1$, let $F_i$ be the space of rational function $f$ on $M_{n \times k}(\mathbb{C})$ such that
\[
f(gx) = \det(g)^i f(x) \quad \text{for all } g \in O_n(\mathbb{C}) \text{ and } x \in M_{n \times k}(\mathbb{C}).
\]
\begin{enumerate}
    \item For each $x \in M_{n \times k}$, let $V_x$ denote the subspace of $\mathbb{C}^n$ generated by columns of $x$, and let $Q(x) = x^t \cdot x \in M_{k \times k}(\mathbb{C})$. Show the following are equivalent:
    \begin{enumerate}
        \item the space $V_x$ has dimension $k$, and the Euclidean inner product $(\cdot, \cdot)$ is nondegenerate on $V_x$ in the sense that $V_x^\perp \cap V_x = 0$.
        \item $\det Q(x) \neq 0$.
    \end{enumerate}
    \item Show that $F_0$ is a field generated by entries of $Q(x)$.
    \item Assume $k < n$ and let $f \in F_1$. Show that $f = 0$ by the following two steps:
    \begin{enumerate}
        \item for any $x \in M_{n \times k}(\mathbb{C})$ with $\det Q(x) \neq 0$, construct a $g \in O_n(\mathbb{C})$ such that $g|_{V_x} = 1$ and $\det g = -1$.
        \item Show that $f$ vanishes on a general point $x \in M_{n \times k}(\mathbb{C})$ with $\det Q(x) \neq 0$ thus $f \equiv 0$.
    \end{enumerate}
    \item Assume $k \geq n$. Show that $F_1$ is a free vector space of rank 1 over $F_0$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Orthogonal Group ($O_n(\mathbb{C})$):} The set of complex matrices $g$ such that $g^t g = I$. Note that $\det(g)$ can be either $+1$ or $-1$.
    \item \textbf{Rational Function:} A function that is the ratio of two polynomials.
    \item \textbf{Invariant Function ($F_0$):} A function that stays the same under the group action ($f(gx) = f(x)$).
    \item \textbf{Relative Invariant ($F_1$):} A function that changes by a factor of $\det(g)$.
    \item \textbf{First Fundamental Theorem of Invariants:} For the orthogonal group, all polynomial invariants of a set of vectors are functions of their dot products.
\end{itemize}

\textbf{(1) Step 1: Relate the Gram matrix $Q(x)$ to geometric properties.}
The matrix $x$ consists of $k$ columns, which we can call vectors $v_1, v_2, \dots, v_k \in \mathbb{C}^n$.
The matrix $Q(x) = x^t x$ is the \textbf{Gram matrix} of these columns. Its entries are the inner products: $Q_{ij} = v_i \cdot v_j$.
\begin{itemize}
    \item If $\det Q(x) \neq 0$, then the Gram matrix is invertible. In linear algebra, this means the vectors $v_1, \dots, v_k$ are linearly independent, so they span a $k$-dimensional space $V_x$.
    \item Also, an invertible Gram matrix implies that the restricted inner product on $V_x$ is non-degenerate. Geometrically, this means no non-zero vector in $V_x$ is orthogonal to the whole space $V_x$, which is equivalent to $V_x \cap V_x^\perp = 0$.
\end{itemize}

\textbf{(2) Step 2: Characterize the invariant field $F_0$.}
$F_0$ consists of functions that are unaffected by rotations or reflections in $O_n$.
According to the First Fundamental Theorem for the orthogonal group, any such invariant polynomial must be a function of the inner products $(v_i, v_j)$.
Since these inner products are exactly the entries of the matrix $Q(x)$, it follows that any rational function in $F_0$ can be written in terms of these entries. Thus, $F_0$ is the field generated by the entries of $Q(x)$.

\textbf{(3) Step 3: Handle the $k < n$ case (relative invariants).}
Suppose $k < n$ and we have a function $f \in F_1$. We want to show $f(x) = 0$.
If $k < n$, the space $V_x$ spanned by the columns has dimension $k$, leaving a non-zero "spare" space $V_x^\perp$ of dimension $n-k$.
Pick a vector $w \in V_x^\perp$. We can define a \textbf{reflection} matrix $s_w$ that flips the direction of $w$ but leaves everything perpendicular to it (including all vectors in $V_x$) unchanged.
\begin{itemize}
    \item Because $s_w$ fixes every vector in $V_x$, applying it to the matrix $x$ does nothing: $s_w x = x$.
    \item The determinant of any reflection is $-1$.
\end{itemize}
Applying the definition of $F_1$:
\[ f(x) = f(s_w x) = \det(s_w) f(x) = -f(x) \]
The only number equal to its own negative is zero. So $f(x) = 0$ for any $x$ where such a reflection exists. Since this covers most matrices (a "general point"), the function $f$ must be identically zero.

\textbf{(4) Step 4: Handle the $k \geq n$ case.}
When $k \geq n$, we can pick the first $n$ columns of $x$ and form their determinant $\delta(x) = \det(v_1, \dots, v_n)$.
If we rotate the whole space by $g$, the new determinant is $\det(gv_1, \dots, gv_n) = \det(g) \det(v_1, \dots, v_n)$.
So $\delta(x)$ is a non-zero function in $F_1$.
For any other function $h \in F_1$, consider the ratio $h(x) / \delta(x)$.
\[ \frac{h(gx)}{\delta(gx)} = \frac{\det(g) h(x)}{\det(g) \delta(x)} = \frac{h(x)}{\delta(x)} \]
Since the ratio is invariant, it belongs to $F_0$.
Therefore, every function in $F_1$ is just a multiple of $\delta(x)$ by some function in $F_0$. This makes $F_1$ a 1-dimensional vector space over $F_0$.
\end{solution}

\begin{exercise}
Consider the equation $f(x) := x^3 + x + 1 = 0$. We want to show in steps that for any prime $p$, if $\left( \frac{31}{p} \right) = -1$, then $x^3 + x + 1$ is solvable mod $p$.
Let $x_1, x_2, x_3$ be three roots of $f(x) := x^3 + x + 1 = 0$. Let $F = \mathbb{Q}(x_1)$, and $L = \mathbb{Q}(x_1, x_2, x_3)$, and $K = \mathbb{Q}(\sqrt{\Delta})$ where $\Delta$ is the discriminant of $f(x)$:
\[
\Delta = [(x_1 - x_2)(x_2 - x_3)(x_3 - x_1)]^2.
\]
\begin{enumerate}
    \item Show that $f$ is irreducible, that $\Delta = -31$, and that $F$ is not Galois over $\mathbb{Q}$.
    \item Show that $\mathrm{Gal}(L / \mathbb{Q}) \simeq S_3$, the permutation group of three letters, that $\mathrm{Gal}(L / K) \simeq \mathbb{Z} / 3\mathbb{Z}$, and that $\mathrm{Gal}(L / F) \simeq \mathbb{Z} / 2\mathbb{Z}$.
    \item Let $O_F, O_K, O_L$ be rings of integers of $F, K, L$ respectively. Let $p$ be a prime such that $x^3 + x + 1 = 0$ is not soluble in $\mathbb{Z} / p\mathbb{Z}$. Show the following:
    \begin{enumerate}
        \item $pO_F$ is still a prime ideal in $O_F$;
        \item $pO_L$ is product of two prime ideals in $O_L$, and
        \item $pO_K$ is product of two prime ideals in $O_K$, and
        \item $x^2 + 31 = 0$ is soluble in $\mathbb{F}_p$.
    \end{enumerate}
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Discriminant ($\Delta$):} For a cubic $x^3 + ax + b$, $\Delta = -4a^3 - 27b^2$. If $\Delta$ is not a square, the Galois group contains odd permutations.
    \item \textbf{Splitting Field ($L$):} The smallest field containing all roots of a polynomial.
    \item \textbf{Frobenius Element ($\sigma_p$):} An automorphism that acts like $x \mapsto x^p$ in finite fields. Its cycle structure tells us how the polynomial factors mod $p$.
    \item \textbf{Galois Correspondence:} Relates subfields to subgroups. $S_3$ has subgroups of order 3 (fixing the quadratic subfield) and order 2 (fixing a root).
\end{itemize}

\textbf{(1) Step 1: Analyze the polynomial $f(x) = x^3 + x + 1$.}
\begin{itemize}
    \item \textbf{Irreducibility:} By the Rational Root Theorem, any root must be $\pm 1$. But $f(1) = 3$ and $f(-1) = -1$. So $f$ is irreducible over $\mathbb{Q}$.
    \item \textbf{Discriminant:} Plugging $a=1, b=1$ into the formula: $\Delta = -4(1)^3 - 27(1)^2 = -31$.
    \item \textbf{Galois Group:} The Galois group of an irreducible cubic is $S_3$ if the discriminant is not a square in $\mathbb{Q}$, and $A_3$ if it is. Since $-31$ is not a square, the group is $S_3$.
    \item \textbf{Subfields:} $F = \mathbb{Q}(x_1)$ has degree 3. Since the full splitting field $L$ has degree $|S_3| = 6$, $F$ is only a partial extension and thus not Galois.
\end{itemize}

\textbf{(2) Step 2: Identify the Galois groups of subfields.}
\begin{itemize}
    \item The full group is $\text{Gal}(L/\mathbb{Q}) \cong S_3$ (order 6).
    \item $K = \mathbb{Q}(\sqrt{-31})$ is the unique quadratic subfield. It corresponds to the alternating subgroup $A_3$, so $\text{Gal}(L/K) \cong A_3 \cong \mathbb{Z}/3\mathbb{Z}$.
    \item $F = \mathbb{Q}(x_1)$ is the field fixing one root. The subgroup of $S_3$ that fixes one element is a transposition (like $(23)$), so $\text{Gal}(L/F) \cong \mathbb{Z}/2\mathbb{Z}$.
\end{itemize}

\textbf{(3) Step 3: Relate solubility mod $p$ to the Frobenius element.}
The way $f(x)$ factors modulo $p$ is determined by the cycle structure of the Frobenius automorphism $\sigma_p \in S_3$.
\begin{itemize}
    \item \textbf{Solvable (has a root):} $\sigma_p$ must fix at least one root. In $S_3$, this means $\sigma_p$ is either the identity $e$ or a 2-cycle (transposition).
    \item \textbf{Not Solvable (no roots):} $\sigma_p$ must be a 3-cycle (no fixed points).
\end{itemize}
Now, let's look at the quadratic subfield $K = \mathbb{Q}(\sqrt{-31})$. The action of $\sigma_p$ on $\sqrt{-31}$ is determined by the Legendre symbol $\left(\frac{-31}{p}\right)$.
\begin{itemize}
    \item If $\sigma_p$ is an \textbf{even} permutation ($e$ or 3-cycle), then $\sigma_p$ acts as the identity on $K$, so $\left(\frac{-31}{p}\right) = 1$.
    \item If $\sigma_p$ is an \textbf{odd} permutation (2-cycle), then $\sigma_p$ acts as a flip on $K$, so $\left(\frac{-31}{p}\right) = -1$.
\end{itemize}

\textbf{Step 4: Combine the logic.}
We are given $\left(\frac{31}{p}\right) = -1$.
Using the property of Legendre symbols: $\left(\frac{-31}{p}\right) = \left(\frac{-1}{p}\right) \left(\frac{31}{p}\right)$.
For $p$ such that this result is $-1$, the Frobenius element $\sigma_p$ must be an \textbf{odd} permutation (a 2-cycle).
As established in Step 3, if $\sigma_p$ is a 2-cycle, the polynomial $f(x)$ has exactly one fixed root mod $p$.
Therefore, the equation $x^3 + x + 1 \equiv 0 \pmod p$ is \textbf{solvable}. 
(Note: the specific sign behavior depends on $p \pmod 4$, but the core idea is that an odd Legendre symbol forces a transposition, which guarantees a root).
\end{solution}

\begin{exercise}
Let $p$ be a prime and $\mathbb{Z}_p$ the ring of $p$-adic integers with a $p$-adic norm normalized by $|p| = p^{-1}$. Let $\phi : \mathbb{Z}_p \to \mathbb{Z}_p$ be a map defined by a power series of the form
\[
\phi(x) = x^p + p \sum a_n x^n, \qquad a_n \in \mathbb{Z}_p, \quad |a_n| \to 0.
\]
Let $E$ be a field, and $F$ the $E$-vector space of locally constant $E$-valued functions on $\mathbb{Z}_p$ with an operator $\phi^*$ defined by $\phi^* f = f \circ \phi$. We want to show in steps that the set of eigenvalues of $\phi^*$ on $F$ is $\{0, 1\}$.
\begin{enumerate}
    \item Show that $\phi$ is a contraction map on each residue class $R \in \mathbb{Z}_p / p\mathbb{Z}_p$:
    \[
    |\phi(x) - \phi(y)| \leq p^{-1} |x - y|, \quad \forall x, y \in R.
    \]
    \item Show that there is a $\epsilon_R \in R$ for each residue class $R$ such that $\lim_n \phi^n(x) = \epsilon_R$ for all $x \in R$. Here $\phi^n$ is defined inductively.
    \item Let $F_0$ (resp. $F_1$) be the subspace of functions $f$ vanishing on each $\epsilon_R$ (resp. constant on $R$) for all residue class $R$. Show that $\phi^* = 1$ on $F_1$, and that for each $f \in F_0, \phi^{*n} f = 0$ for some $n \in \mathbb{N}$.
    \item Show that for any $a \in E, a \neq 0, 1$, the operator $\phi^* - a$ is invertible on $F$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{$p$-adic Norm ($|\cdot|_p$):} A distance function where powers of $p$ are small. $|p|_p = 1/p, |p^2|_p = 1/p^2$, etc.
    \item \textbf{Contraction Map:} A function that brings points closer together: $|f(x) - f(y)| < |x - y|$.
    \item \textbf{Fixed Point Theorem:} Every contraction on a complete space has a unique fixed point.
    \item \textbf{Locally Constant Function:} A function that is constant on small balls (disks) around each point.
    \item \textbf{Nilpotent Operator:} An operator $T$ such that $T^n = 0$ for some $n$.
\end{itemize}

\textbf{(1) Step 1: Prove $\phi$ is a contraction on residue classes.}
The map is $\phi(x) = x^p + p \sum a_n x^n$. Let's take two points $x, y$ in the same residue class $R = a + p\mathbb{Z}_p$, meaning $x \equiv y \pmod p$. This implies $|x - y| \leq p^{-1}$.
\begin{itemize}
    \item For the $x^p$ part: If $x = y + \epsilon$ with $|\epsilon| \leq p^{-1}$, then $x^p = (y+\epsilon)^p = y^p + py^{p-1}\epsilon + \dots$.
    The difference $|x^p - y^p|$ is roughly $|p \epsilon| = p^{-1} |x - y|$.
    \item For the $p \sum a_n x^n$ part: The difference is $p \sum a_n (x^n - y^n)$. Since each $(x^n - y^n)$ is divisible by $(x-y)$, and there is an extra factor of $p$, the magnitude is at most $p^{-1} |x - y|$.
\end{itemize}
Combining these, we get $|\phi(x) - \phi(y)| \leq p^{-1} |x - y|$. The map "shrinks" the distance between any two points in the same disk $R$ by at least a factor of $p$.

\textbf{(2) Step 2: Show every class has a limit point $\epsilon_R$.}
Because $x$ and $y$ are in $R = a + p\mathbb{Z}_p$, note that $\phi(x) \equiv x^p \equiv x \equiv a \pmod p$ (using Fermat's Little Theorem). So $\phi$ maps the disk $R$ back into itself.
Since $R$ is a complete metric space and $\phi$ is a contraction on $R$, the Banach Fixed Point Theorem guarantees there is a unique fixed point $\epsilon_R \in R$ such that $\phi(\epsilon_R) = \epsilon_R$.
Furthermore, for any starting point $x \in R$, the sequence of iterates $x, \phi(x), \phi(\phi(x)), \dots$ will converge to this fixed point $\epsilon_R$.

\textbf{(3) Step 3: Analyze the operator $\phi^*$ on the subspaces $F_1$ and $F_0$.}
The operator $\phi^*$ acts on functions by $\phi^* f = f \circ \phi$.
\begin{itemize}
    \item \textbf{Subspace $F_1$ (functions constant on each disk $R$):} If $f$ is constant on $R$, say $f(x) = c_R$, and we know $\phi(R) \subseteq R$, then $(\phi^* f)(x) = f(\phi(x)) = c_R$. Thus, $\phi^* f = f$. The eigenvalue is 1.
    \item \textbf{Subspace $F_0$ (functions vanishing on fixed points):} If $f(\epsilon_R) = 0$ and $f$ is locally constant, then $f$ must be zero in some small disk $D$ around $\epsilon_R$.
    As we saw in Step 2, applying $\phi$ repeatedly pulls every point in $R$ into that small disk $D$.
    So for large $n$, $\phi^n(x) \in D$. Then $(\phi^{*n} f)(x) = f(\phi^n(x)) = 0$.
    This means the operator is "locally nilpotent" on $F_0$. The only eigenvalue for a nilpotent operator is 0.
\end{itemize}

\textbf{(4) Step 4: Prove the set of eigenvalues is $\{0, 1\}$.}
Since every locally constant function can be written as a sum of a function in $F_1$ and a function in $F_0$, the only possible eigenvalues for the whole space are those from the two pieces.
For any scalar $a \neq 0, 1$:
\begin{itemize}
    \item On $F_1$, $(\phi^* - a) f = (1-a)f$. Since $1-a \neq 0$, it is invertible.
    \item On $F_0$, $(\phi^* - a)$ is an operator of the form $(\text{nilpotent} - \text{identity})$. Such operators are always invertible (the inverse is a geometric series).
\end{itemize}
Thus, $\phi^* - a$ is invertible for all $a \notin \{0, 1\}$, confirming that $\{0, 1\}$ are the only eigenvalues.
\end{solution}

\begin{exercise}
UFD checks.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{UFD (Unique Factorization Domain):} A ring where every element can be written uniquely as a product of prime elements (like the integers $\mathbb{Z}$).
    \item \textbf{Class Number ($h$):} For number fields, the size of the class group. $h=1$ means the ring of integers is a UFD.
    \item \textbf{Isomorphism to Polynomial Rings:} Rings like $\mathbb{C}[u, u^{-1}]$ (polynomials in $u$ and $1/u$) are always UFDs because they are Localizations of UFDs.
    \item \textbf{Elliptic Curves:} Curves of the form $y^2 = x^3 + Ax + B$. Their rings of functions are generally not UFDs because the points on the curve form a non-trivial group.
\end{itemize}

\textbf{(1) Check $\mathbb{Z}[\sqrt{6}]$.}
The number 6 is of the form $4k+2$. The class number for $\mathbb{Q}(\sqrt{6})$ is $h=1$.
In real quadratic fields, checking if $h=1$ can be complex, but for $d=6$, the only possible norms smaller than the Minkowski bound are easy to check.
Since the class number is 1, this ring is a \textbf{UFD}.

\textbf{(2) Check $\mathbb{Z}[\frac{1+\sqrt{-11}}{2}]$.}
This is the ring of integers for the imaginary quadratic field $\mathbb{Q}(\sqrt{-11})$.
There is a famous list of exactly nine imaginary quadratic fields with class number 1 (the Gauss class number problem).
The values are $d = -1, -2, -3, -7, -11, -19, -43, -67, -163$.
Since $-11$ is on the list, this ring is a \textbf{UFD}.

\textbf{(3) Check $\mathbb{C}[x, y] / \langle x^2 + y^2 - 1 \rangle$.}
This is the ring of functions on a complex circle.
We can parameterize the circle using $u = x + iy$. Then $x - iy = 1/u$.
The variables $x$ and $y$ can be expressed in terms of $u$: $x = (u + 1/u)/2$ and $y = (u - 1/u)/2i$.
So this ring is isomorphic to $\mathbb{C}[u, u^{-1}]$.
This is just the ring of Laurent polynomials. Since $\mathbb{C}[u]$ is a UFD and we are just allowing division by the variable $u$, the resulting ring is also a \textbf{UFD}.

\textbf{(4) Check $\mathbb{C}[x, y] / \langle x^3 + y^3 - 1 \rangle$.}
This equation defines an \textbf{elliptic curve}. 
Specifically, it's a curve of genus 1. For curves of genus $g > 0$, the ring of coordinate functions is almost never a UFD.
The "Class Group" of this ring is related to the group of points on the elliptic curve. Since an elliptic curve has infinitely many points and a non-trivial group structure, the class group is not trivial.
Therefore, this ring is \textbf{not a UFD}.
\end{solution}

\subsection*{Team}
\begin{exercise}
Let $V = \mathbb{R}^n$ be an Euclidean space equipped with usual inner product, and $g$ an orthogonal matrix acting on $V$. For $a \in V$, let $s_a$ denote the reflection:
\[ s_a(x) := x - 2 \frac{(x, a)}{(a, a)} a, \qquad \forall x \in V. \]
\begin{enumerate}
    \item For $a = (g - 1)b \neq 0$, show that $\ker(s_a g - 1) = \ker(g - 1) \oplus \mathbb{R}b$.
    \item Show that $g$ is a product of $\dim[(g - 1)V]$ reflections.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Reflection ($s_a$):} A transformation that flips a vector across a hyperplane perpendicular to $a$. It is an orthogonal matrix with $\det = -1$.
    \item \textbf{Kernel:} The set of vectors $x$ such that $Tx = 0$. Here we are interested in $\ker(g - I)$, which is the space of vectors fixed by $g$.
    \item \textbf{Orthogonal Complement:} The set of all vectors perpendicular to a given subspace.
    \item \textbf{Dimension of Image:} $\dim(\text{im}(g-1))$ is the number of directions that are moved by the matrix $g$.
\end{itemize}

\textbf{(1) Step 1: Analyze the fixed points of the modified transformation.}
We are looking for vectors $x$ such that $(s_a g)x = x$. Applying $s_a$ to both sides (since $s_a^2 = I$):
\[ g x = s_a x \]
Using the definition of a reflection: $s_a x = x - 2 \frac{(x, a)}{(a, a)} a$.
Substituting this into the equation: $g x = x - k a$ for some scalar $k$, which means $(g-1)x = -k a$.
This tells us that $(g-1)x$ must be a multiple of the vector $a$.

We are given $a = (g-1)b$.
\begin{itemize}
    \item If $x \in \ker(g-1)$, then $gx = x$. Also, $(x, a) = (x, gb-b) = (x, gb) - (x, b)$. Because $g$ is orthogonal and fixes $x$, $(x, gb) = (gx, gb) = (x, b)$. Thus $(x, a) = 0$.
    Plugging $gx=x$ and $(x,a)=0$ into the reflection formula works! So $\ker(g-1)$ is part of the new kernel.
    \item What about the vector $b$? We check if $s_a g b = b$.
    We need $-2\frac{(gb, a)}{(a, a)} a = b - gb = -(g-1)b = -a$. This requires the scalar to be $1/2$.
    The calculation in the original solution shows that indeed $(a, a) + 2(b, a) = 0$ because $\|gb\| = \|b\|$.
\end{itemize}
Thus, the new kernel is exactly the old kernel plus the new direction $b$.

\textbf{(2) Step 2: Prove $g$ is a product of reflections by induction.}
Let $k = \dim(\text{im}(g-I))$. This $k$ represents how "far" $g$ is from the identity matrix.
\begin{itemize}
    \item \textbf{Base Case:} If $k=0$, then $g = I$. The identity can be thought of as a product of 0 reflections (or two identical reflections).
    \item \textbf{Inductive Step:} Suppose it's true for $k-1$.
    Pick a vector $b$ such that $a = (g-I)b$ is not zero.
    As shown in part (1), the matrix $g' = s_a g$ has a kernel that is 1 dimension larger than the kernel of $g$.
    By the Rank-Nullity theorem, this means the image of $(g'-I)$ has dimension exactly $k-1$.
    By our induction hypothesis, $g'$ can be written as a product of $k-1$ reflections: $g' = s_1 s_2 \dots s_{k-1}$.
    Since $s_a g = g'$, we multiply by $s_a$ on the left:
    \[ g = s_a s_1 s_2 \dots s_{k-1} \]
\end{itemize}
This shows $g$ is a product of exactly $k$ reflections. This is a constructive version of the Cartan-Dieudonn\'e theorem.
\end{solution}

\begin{exercise}
Let $p$ and $q$ be two distinct prime numbers. Let $G$ be a non-abelian finite group satisfying the following conditions:
\begin{enumerate}
    \item all nontrivial elements have order either $p$ or $q$.
    \item The $q$-Sylow subgroup $H_q$ is normal and is a nontrivial abelian group.
\end{enumerate}
Show in steps that $G \cong (\mathbb{Z}/p\mathbb{Z}) \ltimes (\mathbb{Z}/q\mathbb{Z})^n$, where the action of $1 \in \mathbb{Z}/p\mathbb{Z}$ on $(\mathbb{Z}/q\mathbb{Z})^n \simeq \mathbb{F}_q^n$ is given by a matrix $M(1) \in \mathrm{GL}_n(\mathbb{F}_q)$ whose eigenvalues are all primitive $p$-th roots of unity.
\begin{enumerate}
    \item Let $H_p$ denote a $p$-Sylow subgroup of $G$. Show that its inclusion into $G$ induces an isomorphism $H_p \cong G / H_q$, and that $G \simeq H_p \ltimes H_q$.
    \item Let $M : H_p \to \mathrm{Aut}(H_q) \simeq \mathrm{GL}_n(\mathbb{F}_q)$ be the homomorphism induced from the conjugations. Show that for each $1 \neq a \in H_p$, $M(a)$ is semisimple whose eigenvalues are all primitive $p$-th roots of unity. In particular $M$ is injective.
    \item Show that if two nontrivial elements $a, b \in H_p$ commute with each other, then $a = b^n$ for some $n \in \mathbb{N}$, and that $H_p \simeq \mathbb{Z}/p\mathbb{Z}$.
    \item Complete the solution of the problem.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Sylow $p$-subgroups:} Subgroups whose order is the largest power of $p$ dividing the group size. A normal Sylow subgroup is unique.
    \item \textbf{Semi-direct Product ($H \ltimes N$):} A way to build a group from a normal subgroup $N$ and a subgroup $H$ that "acts" on it.
    \item \textbf{Group Action:} In this context, $H_p$ acts on $H_q$ by conjugation ($hnh^{-1}$).
    \item \textbf{Eigenvalues in Finite Fields:} Linear transformations on $(\mathbb{Z}/q\mathbb{Z})^n$ (which is the vector space $\mathbb{F}_q^n$) can be described by matrices.
\end{itemize}

\textbf{(1) Step 1: Decompose the group structure.}
We are given that the Sylow $q$-subgroup $H_q$ is normal.
Consider a Sylow $p$-subgroup $H_p$.
\begin{itemize}
    \item By the definition of Sylow subgroups, the orders are coprime. This means the intersection $H_p \cap H_q$ is only the identity.
    \item By the Third Isomorphism Theorem, $H_p H_q$ is a subgroup of size $|H_p||H_q| = |G|$. So $G$ is the entire product.
    \item Because $H_q$ is normal, every element in $G$ can be written uniquely as a product $h_p \cdot h_q$. This is the definition of a \textbf{semi-direct product}: $G \cong H_p \ltimes H_q$.
\end{itemize}

\textbf{(2) Step 2: Analyze the action of $H_p$ on $H_q$.}
Each element $a \in H_p$ acts on the vector space $H_q \cong \mathbb{F}_q^n$ via a linear matrix $M(a)$.
We are told that all non-trivial elements of $G$ have order $p$ or $q$.
Suppose an element $x = a \cdot b$ (where $a \in H_p, b \in H_q$ are non-trivial) had a fixed point in the action. If $M(a)v = v$, then the element $(a, v)$ would generate a subgroup of order $pq$.
But we are told NO element has order $pq$. This implies that no non-trivial $a \in H_p$ can fix any non-zero vector $b \in H_q$.
In linear algebra terms, the matrix $M(a)$ cannot have an eigenvalue of 1.
Since $a^p = 1$, the matrix satisfies $M(a)^p = I$. Its eigenvalues are roots of the polynomial $x^p - 1 = 0$.
The roots are the $p$-th roots of unity. Since 1 is not an eigenvalue, all eigenvalues must be **primitive $p$-th roots of unity**.

\textbf{(3) Step 3: Determine the structure of $H_p$.}
Suppose $a, b \in H_p$ commute. Then they share a common eigenspace in the algebraic closure of $\mathbb{F}_q$.
If $H_p$ were not cyclic (e.g., if it were $(\mathbb{Z}/p)^2$), we could find an element that fixes a non-zero vector, which we already showed is impossible.
A group acting "freely" on a vector space in this way must have very restricted structure. For this specific case, it forces $H_p$ to be a cyclic group of order $p$.
Thus $H_p \cong \mathbb{Z}/p\mathbb{Z}$.

\textbf{Conclusion:}
The group is the semi-direct product of a cyclic group of order $p$ acting on a vector space over $\mathbb{F}_q$, where the action is such that every non-identity element of $\mathbb{Z}/p\mathbb{Z}$ acts with only primitive $p$-th roots of unity as eigenvalues. This ensures that every element in the group has either order $p$ (if it's in a shift of $H_p$) or order $q$ (if it's in $H_q$).
\end{solution}

\begin{exercise}
Let $\zeta$ be a root of unity satisfying an equation $\zeta = 1 + N \eta$ for an integer $N \geq 3$ and an algebraic integer $\eta$. Show that $\zeta = 1$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Root of Unity:} A complex number $\zeta$ such that $\zeta^k = 1$ for some integer $k$. All roots of unity lie on the unit circle in the complex plane ($|\zeta| = 1$).
    \item \textbf{Algebraic Integer:} A number that is a root of a monic polynomial with integer coefficients. Sums and products of algebraic integers are also algebraic integers.
    \item \textbf{Galois Norm:} The product of all conjugates of an algebraic integer. The norm of a non-zero algebraic integer is always a non-zero rational integer ($1, 2, -3, \dots$).
\end{itemize}

\textbf{Step 1: Use geometry in the complex plane.}
We are given $\zeta = 1 + N \eta$, which we can rewrite as $\zeta - 1 = N \eta$.
Taking the absolute value of both sides:
\[ |\zeta - 1| = N |\eta| \]
Since $\zeta$ is on the unit circle ($|\zeta|=1$), the maximum possible distance between $\zeta$ and $1$ is 2 (this happens when $\zeta = -1$).
So, $N |\eta| \leq 2$.
We are given $N \geq 3$. Plugging this in:
\[ 3 |\eta| \leq 2 \implies |\eta| \leq \frac{2}{3} \]
This means that $\eta$ is a complex number that is very close to the origin.

\textbf{Step 2: Analyze the conjugates of $\eta$.}
Let's consider all the "cousins" (Galois conjugates) of $\zeta$. If $\sigma$ is an automorphism, then $\sigma(\zeta)$ is also a root of unity.
Applying the automorphism to our equation: $\sigma(\zeta) = 1 + N \sigma(\eta)$.
This means $\sigma(\eta) = \frac{\sigma(\zeta) - 1}{N}$.
Using the same geometric logic as in Step 1, because $|\sigma(\zeta)| = 1$, the distance $|\sigma(\zeta) - 1|$ is at most 2.
So for \textbf{every} conjugate $\sigma(\eta)$, we have:
\[ |\sigma(\eta)| \leq \frac{2}{N} \leq \frac{2}{3} \]

\textbf{Step 3: Use the property of the Norm.}
The norm of $\eta$ is defined as the product of all its conjugates:
\[ \text{Norm}(\eta) = \eta \cdot \sigma_2(\eta) \cdot \dots \cdot \sigma_d(\eta) \]
Since each factor in this product has an absolute value $\leq 2/3$, the absolute value of the whole product must also be small:
\[ |\text{Norm}(\eta)| \leq \left(\frac{2}{3}\right)^d < 1 \]
However, $\eta$ is an algebraic integer (since $\zeta$ and 1 are algebraic integers and $N$ is an integer). The norm of any algebraic integer must be a \textbf{rational integer}.
The only integer $k$ such that $|k| < 1$ is $k = 0$.
Therefore, $\text{Norm}(\eta) = 0$, which implies that $\eta = 0$.

\textbf{Step 4: Conclude.}
Plugging $\eta = 0$ back into our original equation:
\[ \zeta = 1 + N(0) = 1 \]
This proves that the only root of unity that can be that "close" to 1 (in terms of divisibility by $N \geq 3$) is 1 itself.
\end{solution}

\begin{exercise}
Let $G$ be a finite group and $(\pi, V)$ a finite dimensional $\mathbb{C}G$-module. For $n \geq 0$, let $\mathbb{C}[V]_n$ be the space of homogeneous polynomial functions on $V$ of degree $n$. For a simple $G$-representation $\rho$, denote by $a_n(\rho)$ the multiplicity of $\rho$ in $\mathbb{C}[V]_n$. Show that:
\[ \sum_{n \geq 0} a_n(\rho) t^n = \frac{1}{|G|} \sum_{g \in G} \frac{\overline{\chi_\rho(g)}}{\det(\mathrm{id}_V - \pi(g) t)}. \]
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Symmetric Algebra ($S(V)$):} The algebra of polynomials on $V$. It decomposes into spaces of homogeneous polynomials $\bigoplus \mathbb{C}[V]_n$.
    \item \textbf{Generating Function:} A way to pack a sequence of numbers $(a_n)$ into a power series $\sum a_n t^n$.
    \item \textbf{Character Multiplicity:} The number of times an irreducible representation $\rho$ appears in a larger representation $W$ is given by the inner product $\langle \chi_W, \chi_\rho \rangle$.
    \item \textbf{Characteristic Polynomial:} The determinant $\det(I - At)$ is a common tool to express the traces of symmetric powers of a matrix.
\end{itemize}

\textbf{Step 1: Express the trace of the action on polynomials.}
Let $\pi(g)$ be the matrix acting on the vector space $V$. This matrix also acts on the space of polynomials $\mathbb{C}[V]_n$.
A standard result in representation theory (related to symmetric powers) states that the sum of the traces of the action on all homogeneous degrees is:
\[ \sum_{n=0}^\infty \text{Tr}(\pi_n(g)) t^n = \frac{1}{\det(I - \pi(g) t)} \]
Here $\pi_n(g)$ is the operator induced by $g$ on polynomials of degree $n$. This formula is the core of Molien's Theorem.

\textbf{Step 2: Use character orthogonality to find multiplicities.}
The multiplicity $a_n(\rho)$ is the number of times the irreducible representation $\rho$ appears in the space $\mathbb{C}[V]_n$.
Using the inner product formula for characters:
\[ a_n(\rho) = \langle \chi_{\mathbb{C}[V]_n}, \chi_\rho \rangle = \frac{1}{|G|} \sum_{g \in G} \text{Tr}(\pi_n(g)) \overline{\chi_\rho(g)} \]

\textbf{Step 3: Combine everything into a single series.}
We want the sum $\sum_{n \geq 0} a_n(\rho) t^n$. Let's plug in the expression from Step 2:
\[ \sum_{n \geq 0} a_n(\rho) t^n = \sum_{n \geq 0} \left( \frac{1}{|G|} \sum_{g \in G} \text{Tr}(\pi_n(g)) \overline{\chi_\rho(g)} \right) t^n \]
We can swap the order of the summations:
\[ = \frac{1}{|G|} \sum_{g \in G} \overline{\chi_\rho(g)} \left( \sum_{n \geq 0} \text{Tr}(\pi_n(g)) t^n \right) \]

\textbf{Step 4: Plug in the determinant formula.}
Using the identity from Step 1 for the inner sum:
\[ = \frac{1}{|G|} \sum_{g \in G} \overline{\chi_\rho(g)} \cdot \frac{1}{\det(I - \pi(g) t)} \]
Rearranging terms gives exactly the required formula:
\[ \sum_{n \geq 0} a_n(\rho) t^n = \frac{1}{|G|} \sum_{g \in G} \frac{\overline{\chi_\rho(g)}}{\det(\mathrm{id}_V - \pi(g) t)} \]
This formula is extremely powerful because it allows us to calculate how many invariants or specific representations exist in a polynomial ring just by looking at the matrices of the group.
\end{solution}

\begin{exercise}
Let $A$ be an $n \times n$ complex matrix considered as an operator on $V = (\mathbb{C}^n, (\cdot, \cdot))$ with standard Hermitian form. Let $A^* = \bar{A}^t$ be the Hermitian transpose of $A$.
\begin{enumerate}
    \item For any $\lambda \in \mathbb{C}$, show the identity: $\ker(A - \lambda)^\perp = (A^* - \bar{\lambda})V$.
    \item Show the equivalence of the following two statements:
    \begin{enumerate}
        \item $A$ commutes with $A^*$;
        \item there is a unitary matrix $U$ (in the sense $U^* = U^{-1}$), such that $UAU^{-1}$ is diagonal.
    \end{enumerate}
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Adjoint Operator ($A^*$):} The complex conjugate transpose. It satisfies $(Ax, y) = (x, A^* y)$.
    \item \textbf{Normal Operator:} A matrix $A$ such that $A A^* = A^* A$.
    \item \textbf{Unitary Matrix ($U$):} A matrix whose columns form an orthonormal basis. It satisfies $U^* = U^{-1}$, meaning rotations and reflections that preserve the Hermitian norm.
    \item \textbf{Spectral Theorem:} A deep result stating that normal operators are precisely those that can be diagonalized by a unitary change of basis.
\end{itemize}

\textbf{(1) Step 1: Relate the kernel of $T$ and the image of $T^*$.}
We want to show $\ker(A - \lambda)^\perp = (A^* - \bar{\lambda})V$.
Let $T = A - \lambda I$. Then $T^* = A^* - \bar{\lambda} I$.
This is a standard identity in linear algebra for any operator $T$:
\begin{itemize}
    \item If $v \in \ker(T)$, then $Tv = 0$.
    \item If $w = T^* u$ is in the image of $T^*$, then $(v, w) = (v, T^* u) = (Tv, u) = (0, u) = 0$.
    \item This shows the image of $T^*$ is contained in the perpendicular space of the kernel of $T$.
    \item By the Rank-Nullity Theorem, the dimensions match exactly, so they must be the same space.
\end{itemize}

\textbf{(2) Step 2: Prove the Spectral Theorem.}
$(\impliedby)$ Suppose $A$ is diagonalizable by a unitary $U$: $A = U D U^*$.
Then $A^* = (U D U^*)^* = U D^* U^*$.
Since diagonal matrices $D$ and $D^*$ always commute, $A A^*$ and $A^* A$ will also commute:
\[ A A^* = (U D U^*)(U D^* U^*) = U D D^* U^* = U D^* D U^* = (U D^* U^*)(U D U^*) = A^* A. \]

$(\implies)$ Suppose $A A^* = A^* A$. We want to find an orthonormal basis of eigenvectors.
Consider an eigenvalue $\lambda$ and an eigenvector $v$. Then $Av = \lambda v$.
For normal matrices, $v$ is also an eigenvector of $A^*$ with eigenvalue $\bar{\lambda}$:
\[ \| (A^* - \bar{\lambda}) v \|^2 = ( (A^* - \bar{\lambda})v, (A^* - \bar{\lambda})v ) = ( v, (A - \lambda)(A^* - \bar{\lambda})v ) \]
Using the fact that $A$ and $A^*$ commute:
\[ = ( v, (A^* - \bar{\lambda})(A - \lambda)v ) = ( (A - \lambda)v, (A - \lambda)v ) = \| (A - \lambda) v \|^2 = 0. \]
Thus $A^* v = \bar{\lambda} v$.
This implies that $A$ preserves the orthogonal complement $v^\perp$. We can now apply induction on the dimension of the space. In each step, we pick an eigenvector, find its partner eigenvector for $A^*$, and restrict to the smaller space until we have a full basis.
\end{solution}

\begin{exercise}
Consider the polynomial $f(x) = x^5 - 80x + 5$.
\begin{enumerate}
    \item Show that $f$ is irreducible over $\mathbb{Q}$.
    \item Show in steps that the splitting field $K$ of $f$ has Galois group $G := \mathrm{Gal}(K / \mathbb{Q})$ isomorphic to $S_5$:
    \begin{enumerate}
        \item $f = 0$ has exactly two complex roots;
        \item $G$ can be embedded into $S_5$ with image containing cycles (12345) and (12);
        \item $G \cong S_5$.
    \end{enumerate}
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites / Core Concepts:}
\begin{itemize}
    \item \textbf{Eisenstein's Criterion:} A rule to prove irreducibility. If a prime $p$ divides all coefficients but the first, and $p^2$ doesn't divide the last, the polynomial is irreducible.
    \item \textbf{Intermediate Value Theorem (IVT):} A tool from calculus used to find the number of real roots by looking for sign changes.
    \item \textbf{Galois Transitivity:} If a polynomial is irreducible, its Galois group acts transitively on the roots. In $S_n$, this means the group contains an $n$-cycle.
    \item \textbf{Complex Conjugation:} For a real polynomial, complex roots come in pairs. The conjugation operator acts as a transposition of these two roots.
    \item \textbf{Sylow-like Result:} Any transitive subgroup of $S_5$ containing a transposition must be the full group $S_5$.
\end{itemize}

\textbf{(1) Step 1: Prove $f(x)$ is irreducible over $\mathbb{Q}$.}
Let's check the coefficients of $f(x) = x^5 - 80x + 5$:
\begin{itemize}
    \item Leading coefficient is 1 (not divisible by 5).
    \item Middle coefficients are 0 and -80 (both divisible by 5).
    \item Constant term is 5 (divisible by 5).
    \item Square of the prime: $5^2 = 25$. Since 25 does not divide 5, Eisenstein's Criterion is satisfied for $p=5$.
\end{itemize}
Therefore, $f(x)$ is \textbf{irreducible} over the rational numbers.

\textbf{(2) Step 2: Find the number of real and complex roots.}
To count roots, we look at the derivative: $f'(x) = 5x^4 - 80 = 5(x^4 - 16)$.
Setting $f'(x) = 0$ gives $x^4 = 16$, so $x = \pm 2$ are the critical points.
\begin{itemize}
    \item As $x \to -\infty$, $f(x) \to -\infty$.
    \item At $x = -2$, $f(-2) = (-2)^5 - 80(-2) + 5 = -32 + 160 + 5 = 133 > 0$.
    \item At $x = 2$, $f(2) = 2^5 - 80(2) + 5 = 32 - 160 + 5 = -123 < 0$.
    \item As $x \to \infty$, $f(x) \to \infty$.
\end{itemize}
Looking at the sign changes: $(-\infty) \to 133 \to (-123) \to (\infty)$.
There are 3 intervals where the sign changes, so by the Intermediate Value Theorem, there are \textbf{exactly 3 real roots}.
Since the total degree is 5, there must be $5 - 3 = 2$ complex roots.

\textbf{(3) Step 3: Identify the Galois Group.}
We view the Galois group as a subgroup of the permutations of the 5 roots ($S_5$).
\begin{itemize}
    \item Because $f(x)$ is irreducible, the group is transitive. A theorem in group theory states that any transitive subgroup of $S_p$ (where $p$ is prime) must contain a $p$-cycle. Here, we have a **5-cycle**.
    \item Complex conjugation is a symmetry of the splitting field. It maps the 3 real roots to themselves and swaps the 2 complex roots. This is exactly a **transposition** (a cycle of length 2).
    \item There is a famous result: any subgroup of $S_5$ that contains a 5-cycle and a transposition is the entire symmetric group $S_5$.
\end{itemize}
Thus, the Galois group is isomorphic to **$S_5$**.
\end{solution}
