\section{2010 Applied \& Computational Mathematics}

\subsection{Individual}

\begin{exercise}
\label{applied:2010:1}
Let $Z_{1}, \cdots, Z_{n}$ be i.i.d. random variables with $Z_{i} \sim N(\mu, \sigma^{2})$. Find $E(\sum_{i=1}^{n} Z_{i} | Z_{1} - Z_{2} + Z_{3})$.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item Let $S = \sum_{i=1}^n Z_i$ and $D = Z_1 - Z_2 + Z_3$. Note that $(S, D)$ follows a bivariate normal distribution since they are linear combinations of independent normal variables.
    \item Compute the parameters:
    \begin{itemize}
        \item $E(S) = n\mu$.
        \item $E(D) = \mu - \mu + \mu = \mu$.
        \item $Var(D) = Var(Z_1) + Var(-Z_2) + Var(Z_3) = 3\sigma^2$.
        \item $Cov(S, D) = Cov(\sum Z_i, Z_1 - Z_2 + Z_3) = Cov(Z_1, Z_1) - Cov(Z_2, Z_2) + Cov(Z_3, Z_3) = \sigma^2 - \sigma^2 + \sigma^2 = \sigma^2$.
    \end{itemize}
    \item Use the conditional expectation formula for bivariate normal variables:
    \[ E(S|D) = E(S) + \frac{Cov(S, D)}{Var(D)}(D - E(D)) \]
    \item Substitute the values:
    \[ E(S|D) = n\mu + \frac{\sigma^2}{3\sigma^2}(D - \mu) = n\mu + \frac{1}{3}(Z_1 - Z_2 + Z_3 - \mu). \]
\end{enumerate}

\textbf{Intuition:}
The sum of all variables and the specific linear combination $Z_1 - Z_2 + Z_3$ are correlated because they share common variables. In a normal setting, the conditional expectation is a linear function of the conditioning variable, with the slope determined by the ratio of the covariance to the variance of the condition.
\end{solution}

\begin{exercise}
\label{applied:2010:2}
Let $X_{1}, \cdots, X_{n}$ be pairwise independent. Further, assume that $E X_{i} = 1 + i^{-1}$ and that $\max_{1 \leq i \leq n} E | X_{i} |^{1 + \epsilon} < \infty$ for some $\epsilon > 0$. Show that $\frac{1}{n} \sum_{i=1}^{n} X_{i} \xrightarrow{P} 1$.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item Compute the expected value of the average:
    \[ E\left(\frac{1}{n} \sum_{i=1}^n X_i\right) = \frac{1}{n} \sum_{i=1}^n (1 + i^{-1}) = 1 + \frac{1}{n} \sum_{i=1}^n \frac{1}{i} = 1 + \frac{H_n}{n}. \]
    Since $H_n \sim \ln n$, $\frac{H_n}{n} \to 0$ as $n \to \infty$, so the expectation converges to 1.
    \item To show convergence in probability, we need to show that the variance (or a related dispersion measure) goes to zero. Since the variables are only pairwise independent, the standard WLLN requires slightly more care.
    \item The condition $\max E |X_i|^{1+\epsilon} < \infty$ implies uniform integrability and allows the use of the weak law for pairwise independent variables. Specifically, we can use a truncation argument: let $Y_i = X_i I(|X_i| \le n)$.
    \item For pairwise independent variables with bounded $(1+\epsilon)$ moments, the average $\frac{1}{n} \sum (X_i - E X_i)$ converges to 0 in probability (and even almost surely if the moments are $L^2$).
\end{enumerate}

\textbf{Intuition:}
The "pairwise independent" condition is weaker than full independence but still allows for the law of large numbers to hold. The bounded $(1+\epsilon)$ moment ensures that no single variable's tail can dominate the sum, providing enough regularity for the average to concentrate around its mean.
\end{solution}

\begin{exercise}
\label{applied:2010:3}
Let $Z_{1}, \cdots, Z_{6}$ be i.i.d. random variables with $Z_{i} \sim N(0, 1)$. Set $U^{2} = \frac{(Z_{1} Z_{2} + Z_{3} Z_{4} + Z_{5} Z_{6})^{2}}{Z_{2}^{2} + Z_{4}^{2} + Z_{6}^{2}}$, $V^{2} = \frac{U^{2} (Z_{2}^{2} + Z_{4}^{2})}{U^{2} + Z_{6}^{2}}$. Find and identify the densities of $U^{2}$ and $V^{2}$.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item Let $\mathbf{X} = (Z_1, Z_3, Z_5)^T$ and $\mathbf{Y} = (Z_2, Z_4, Z_6)^T$. Then $U^2 = \frac{(\mathbf{X} \cdot \mathbf{Y})^2}{\|\mathbf{Y}\|^2}$.
    \item Conditional on $\mathbf{Y}$, the term $\frac{\mathbf{X} \cdot \mathbf{Y}}{\|\mathbf{Y}\|}$ is a linear combination of independent standard normals, which is itself $N(0, 1)$. Since this distribution does not depend on $\mathbf{Y}$, $U = \frac{\mathbf{X} \cdot \mathbf{Y}}{\|\mathbf{Y}\|}$ is independent of $\mathbf{Y}$ and $U \sim N(0, 1)$.
    \item Thus $U^2 \sim \chi^2(1)$.
    \item For $V^2 = \frac{U^2 (Z_2^2 + Z_4^2)}{U^2 + Z_6^2}$, note that $U^2, Z_2^2+Z_4^2, Z_6^2$ are independent. Let $A = U^2 \sim \chi^2(1)$, $B = Z_6^2 \sim \chi^2(1)$, and $C = Z_2^2 + Z_4^2 \sim \chi^2(2)$.
    \item $V^2 = C \cdot \frac{A}{A+B}$. Since $A, B$ are i.i.d. $\chi^2(1)$, $W = \frac{A}{A+B} \sim Beta(\frac{1}{2}, \frac{1}{2})$.
    \item $C \sim \chi^2(2)$ is an Exponential distribution with mean 2, i.e., $f_C(c) = \frac{1}{2} e^{-c/2}$.
    \item The density of $V^2$ can be found by the product of Beta and Gamma distributions.
\end{enumerate}

\textbf{Intuition:}
The variable $U$ represents the projection of one random vector onto another. Due to the rotational invariance of the normal distribution, this projection is always standard normal. $V^2$ then combines this with other $\chi^2$ components, resulting in a distribution related to the ratio of variances.
\end{solution}

\begin{exercise}
\label{applied:2010:4}
Consider a population with three characteristics with frequencies: $p_1 = \theta^3$, $p_2 = 3\theta(1-\theta)$, $p_3 = (1-\theta)^3$. Note that these sum to 1.
(a) Construct approx level $(1-\alpha)$ MLE confidence set.
(b) Derive asymptotic distribution for freq substitution estimator $\hat{\theta}_2 = 1 - (N_3/n)^{1/3}$.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item \textbf{MLE:} The log-likelihood is $\ell(\theta) = N_1 \ln \theta^3 + N_2 \ln(3\theta(1-\theta)) + N_3 \ln(1-\theta)^3$.
    \item $\ell'(\theta) = \frac{3N_1+N_2}{\theta} - \frac{N_2+3N_3}{1-\theta} = 0 \implies \hat{\theta} = \frac{3N_1+N_2}{3N_1+2N_2+3N_3}$.
    \item Fisher Information $I(\theta) = -E[\ell''(\theta)]$ gives the asymptotic variance $Var(\hat{\theta}) \approx \frac{1}{n I(\theta)}$. The confidence set is $\hat{\theta} \pm z_{\alpha/2} \sqrt{\widehat{Var}(\hat{\theta})}$.
    \item \textbf{Substitution Estimator:} For $\hat{\theta}_2 = 1 - (\frac{N_3}{n})^{1/3}$, we use the Delta Method.
    \item $\frac{N_3}{n} \xrightarrow{d} AN(p_3, \frac{p_3(1-p_3)}{n})$. Let $g(x) = 1 - x^{1/3}$. Then $g'(x) = -\frac{1}{3} x^{-2/3}$.
    \item Asymptotic distribution: $\sqrt{n}(\hat{\theta}_2 - \theta) \xrightarrow{d} N(0, [g'(p_3)]^2 p_3(1-p_3))$.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2010:5}
\begin{enumerate}
    \item Suppose
    \[
    S = \begin{bmatrix} \sigma & \mathbf{u}^T \\ 0 & S_c \end{bmatrix}, \quad T = \begin{bmatrix} \tau & \mathbf{v}^T \\ 0 & T_c \end{bmatrix}, \quad \mathbf{b} = \begin{bmatrix} \beta \\ \mathbf{b}_c \end{bmatrix},
    \]
    where $\sigma, \tau$ and $\beta$ are scalars, $S_c$ and $T_c$ are $n$-by-$n$ matrices, and $\mathbf{b}_c$ is an $n$-vector. Show that if there exists a vector $\mathbf{x}_c$ such that $(S_c T_c - \lambda I) \mathbf{x}_c = \mathbf{b}_c$ and $\mathbf{w}_c = T_c \mathbf{x}_c$ is available, then
    \[
    \mathbf{x} = \begin{bmatrix} \gamma \\ \mathbf{x}_c \end{bmatrix}, \quad \gamma = \frac{\beta - \sigma \mathbf{v}^T \mathbf{x}_c - \mathbf{u}^T \mathbf{w}_c}{\sigma \tau - \lambda}
    \]
    solves $(ST - \lambda I) \mathbf{x} = \mathbf{b}$.

    \item Hence or otherwise, derive an $O(n^2)$ algorithm for solving the linear system $(U_1 U_2 - \lambda I) \mathbf{x} = \mathbf{b}$ where $U_1$ and $U_2$ are $n$-by-$n$ upper triangular matrices, and $(U_1 U_2 - \lambda I)$ is nonsingular. Please write down your algorithm and prove that it is indeed of $O(n^2)$ complexity.

    \item Hence or otherwise, derive an $O(pn^2)$ algorithm for solving the linear system $(U_1 U_2 \cdots U_p - \lambda I) \mathbf{x} = \mathbf{b}$ where $\{U_i\}_{i=1}^p$ are all $n$-by-$n$ upper triangular matrices, and $(U_1 U_2 \cdots U_p - \lambda I)$ is nonsingular. Please write down your algorithm and prove that it is indeed of $O(pn^2)$ complexity.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Block matrix multiplication and inversion.
    \item Fundamental properties of upper triangular matrices (product is triangular, solving via back-substitution).
    \item Order of complexity for matrix-vector products.
\end{itemize}

\textbf{Steps:}
\begin{enumerate}
    \item \textbf{Identity Verification:} Compute the product $ST = \begin{bmatrix} \sigma \tau & \sigma \mathbf{v}^T + \mathbf{u}^T T_c \\ 0 & S_c T_c \end{bmatrix}$. The system $(ST - \lambda I)\mathbf{x} = \mathbf{b}$ yields:
    \[ \begin{bmatrix} \sigma \tau - \lambda & \sigma \mathbf{v}^T + \mathbf{u}^T T_c \\ 0 & S_c T_c - \lambda I \end{bmatrix} \begin{bmatrix} \gamma \\ \mathbf{x}_c \end{bmatrix} = \begin{bmatrix} \beta \\ \mathbf{b}_c \end{bmatrix}. \]
    The lower block gives $(S_c T_c - \lambda I)\mathbf{x}_c = \mathbf{b}_c$. The upper block gives $(\sigma \tau - \lambda)\gamma + \sigma \mathbf{v}^T \mathbf{x}_c + \mathbf{u}^T (T_c \mathbf{x}_c) = \beta$. Solving for $\gamma$ with $\mathbf{w}_c = T_c \mathbf{x}_c$ yields the desired formula.

    \item \textbf{$O(n^2)$ Algorithm for $U_1 U_2$:} We use back-substitution. Let $(U_1)_{ii} = \sigma_i$ and $(U_2)_{ii} = \tau_i$. Starting from $i=n$ down to $1$:
    \begin{itemize}
        \item Compute $x_i = (\beta_i - \sigma_i \sum_{j=i+1}^n (U_2)_{ij} x_j - \sum_{j=i+1}^n (U_1)_{ij} w_j) / (\sigma_i \tau_i - \lambda)$.
        \item Update $w_i = \tau_i x_i + \sum_{j=i+1}^n (U_2)_{ij} x_j$.
    \end{itemize}
    Each step $i$ involves $O(n-i)$ additions and multiplications. Total complexity $\sum_{i=1}^n O(n-i) = O(n^2)$.

    \item \textbf{$O(pn^2)$ Algorithm for $p$ matrices:} Let $M^{(k)} = U_k \cdots U_p$. To solve $(M^{(1)} - \lambda I) \mathbf{x} = \mathbf{b}$, we define intermediate vectors $\mathbf{w}^{(k)} = M^{(k+1)} \mathbf{x}$. By solving the components backward $i=n, \dots, 1$ and for each $i$ computing the sequence of $p$ intermediate values, the logic from Part (2) repeats $p$ times per row entry. Total complexity is $O(p n^2)$.
\end{enumerate}

\textbf{Intuition:}
The problem leverages the fact that the product of upper triangular matrices is itself upper triangular. Instead of explicitly computing the matrix product $U_1 U_2$ (which is $O(n^3)$), we can solve the system by maintaining partial products of the triangular matrices with the solution vector, essentially extending the standard back-substitution method to handle the nested structure.
\end{solution}

\begin{exercise}
\label{applied:2010:6}
\begin{enumerate}
    \item Let $A \in \mathbb{R}^{m \times n}$. Show that there exists an $m$-by-$m$ orthogonal matrix $U$ and an $n$-by-$n$ orthogonal matrix $V$ such that $U^T A V = \text{diag}(\sigma_1, \dots, \sigma_p)$, where $p = \min\{m, n\}$ and $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_p \geq 0$.

    \item Let $\text{rank}(A) = r$. Show that for any positive integer $k < r$
    \[
    \min_{\text{rank}(B) = k} \| A - B \|_2 = \sigma_{k+1}.
    \]
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Spectral theorem for real symmetric matrices.
    \item Definition and properties of orthogonal matrices.
    \item Vector norm and induced matrix 2-norm properties.
\end{itemize}

\textbf{Steps:}
\begin{enumerate}
    \item \textbf{Existence of SVD:} Consider $A^T A \in \mathbb{R}^{n \times n}$. Since it is symmetric and positive semi-definite, it has orthonormal eigenvectors $v_1, \dots, v_n$ with eigenvalues $\lambda_1 \geq \dots \geq \lambda_n \geq 0$. Define $\sigma_i = \sqrt{\lambda_i}$. For $\sigma_i > 0$, let $u_i = A v_i / \sigma_i$. Orthonormality follows: $u_i^T u_j = \frac{v_i^T A^T A v_j}{\sigma_i \sigma_j} = \frac{\lambda_j \delta_{ij}}{\sigma_i \sigma_j} = \delta_{ij}$. Complete $\{u_i\}$ to an orthonormal basis of $\mathbb{R}^m$. Then $A V = U \Sigma \implies U^T A V = \Sigma$.

    \item \textbf{Low-Rank Approximation (Eckart-Young):}
    \begin{itemize}
        \item For $B_k = \sum_{i=1}^k \sigma_i u_i v_i^T$, $\| A - B_k \|_2 = \sigma_{k+1}$.
        \item For any $B$ with $\text{rank}(B) = k$, the null space $\mathcal{N}(B)$ has dimension $n-k$. The subspace $V_{k+1} = \text{span}\{v_1, \dots, v_{k+1}\}$ has dimension $k+1$.
        \item There exists a unit vector $z \in \mathcal{N}(B) \cap V_{k+1}$. Then $\|(A-B)z\|_2 = \|Az\|_2$.
        \item Writing $z = \sum_{i=1}^{k+1} \alpha_i v_i$, $\|Az\|_2^2 = \sum_{i=1}^{k+1} \sigma_i^2 \alpha_i^2 \geq \sigma_{k+1}^2$. Thus $\|A-B\|_2 \geq \sigma_{k+1}$.
    \end{itemize}
\end{enumerate}

\textbf{Intuition:}
The Singular Value Decomposition (SVD) identifies the principal axes of the transformation $A$. The 2-norm of a matrix is determined by its largest singular value. When approximating with a lower rank, the "best" strategy is to remove the components corresponding to the smallest singular values, as they contribute the least to the "energy" or "variance" of the mapping.
\end{solution}

\subsection{Team}

\begin{exercise}
\label{applied:2010:7}
Let $X_1, \dots, X_n$ be i.i.d. with distribution function $F$. Let $Y_1 < \dots < Y_n$ be the order statistics. Prove that $Z_j = F(Y_j)$ has a Beta distribution $Beta(j, n-j+1)$.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item Note that if $X \sim F$, then $F(X) \sim U(0, 1)$ (Probability Integral Transform).
    \item Let $U_1, \dots, U_n$ be i.i.d. $U(0, 1)$. The order statistics $U_{(j)} = F(Y_j)$ are the variables $Z_j$ we are interested in.
    \item The probability density function of the $j$-th order statistic from a $U(0, 1)$ distribution is given by:
    \[ f_{U_{(j)}}(z) = \frac{n!}{(j-1)!(n-j)!} z^{j-1} (1-z)^{n-j}, \quad 0 < z < 1. \]
    \item This is precisely the PDF of the $Beta(j, n-j+1)$ distribution.
\end{enumerate}

\textbf{Intuition:}
The transformation $F(X)$ maps any continuous distribution to the uniform distribution on $[0,1]$. The order statistics of uniform variables naturally follow Beta distributions, as they partition the interval $[0,1]$ into segments whose lengths are related to the counts of variables falling in each segment.
\end{solution}

\begin{exercise}
\label{applied:2010:8}
Investigate the convergence to a Normal distribution for the sequence of random variables $Y_{n,i} = \frac{3}{4b_n}(1 - X_i^2/b_n^2)I(|X_i| \le b_n)$, where $X_i$ are i.i.d. and $b_n$ is a sequence of constants.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item This problem involves studying the asymptotic behavior of a sum of kernel-like terms. 
    \item Check the conditions for the Central Limit Theorem (CLT) for the sum $\sum Y_{n,i}$. 
    \item Calculate the mean and variance of $Y_{n,i}$. The variance will depend on the scale $b_n$ and the underlying distribution of $X_i$.
    \item Usually, if $b_n \to 0$ or $b_n \to \infty$ at specific rates, the normalized sum will converge to a normal distribution by the Lindeberg-Feller CLT.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2010:9}
Prove the minimax error bound: $\min \sup_{|\theta| \le \tau} E (\sum a_i X_i + a_{n+1} - \theta)^2 = \frac{\tau^2 n^{-1}}{\tau^2 + n^{-1}}$ for a suitable model (e.g. $X_i \sim N(\theta, 1)$).
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item This is a classic result in minimax estimation. Consider the linear estimator $\delta = w \bar{X}$.
    \item The risk is $R(\theta, w) = E(w \bar{X} - \theta)^2 = w^2 \frac{1}{n} + (w-1)^2 \theta^2$.
    \item To find the minimax estimator, we look for the $w$ that minimizes the maximum risk over $\theta \in [-\tau, \tau]$.
    \item The maximum risk occurs at $\theta = \pm \tau$. Thus we minimize $f(w) = \frac{w^2}{n} + (w-1)^2 \tau^2$.
    \item Taking the derivative: $\frac{2w}{n} + 2(w-1)\tau^2 = 0 \implies w = \frac{\tau^2}{\tau^2 + n^{-1}}$.
    \item Substituting $w$ back into the risk gives the bound $\frac{\tau^2 n^{-1}}{\tau^2 + n^{-1}}$.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2010:10}
In a foul shooting model, let $N_0, N_1, N_2$ be the counts of players missing both, making one, and making both shots respectively in $n$ pairs of shots. Derive the MLE $\hat{p} = \frac{N_1 + 2N_2}{N_0 + 2N_1 + 2N_2}$ (assuming independent shots) and find its asymptotic distribution.
\end{exercise}

\begin{solution}
\textbf{Steps:}
\begin{enumerate}
    \item If shots are independent with probability $p$, the counts $N_0, N_1, N_2$ follow a multinomial distribution with probabilities $(1-p)^2, 2p(1-p), p^2$.
    \item The log-likelihood is $\ell(p) = N_0 \ln(1-p)^2 + N_1 \ln(2p(1-p)) + N_2 \ln p^2$.
    \item $\ell'(p) = \frac{-2N_0}{1-p} + \frac{N_1}{p} - \frac{N_1}{1-p} + \frac{2N_2}{p} = 0$.
    \item Solving for $p$ gives $\hat{p} = \frac{N_1 + 2N_2}{2(N_0 + N_1 + N_2)}$. Note the denominator in the prompt might assume a different total count convention.
    \item By the property of MLE, $\sqrt{n}(\hat{p} - p) \xrightarrow{d} N(0, I(p)^{-1})$.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2010:11}
When considering finite difference schemes approximating partial differential equations (PDEs), for example, the scheme
\[
u_j^{n+1} = u_j^n - \lambda(u_j^n - u_{j-1}^n) \tag{1}
\]
where $\lambda = \frac{\Delta t}{\Delta x}$, approximating the PDE
\[
u_t + u_x = 0, \tag{2}
\]
we are often interested in stability, namely
\[
\|u^n\| \leq C\|u^0\|, \quad n \Delta t \leq T \tag{3}
\]
for a constant $C = C(T)$ independent of the time step $\Delta t$ and the spatial mesh size $\Delta x$. Here $\|\cdot\|$ is a given norm, for example the $L^2$ norm or the $L^\infty$ norm, of the numerical solution vector $u^n = (u_1^n, u_2^n, \dots, u_N^n)$. The mesh points are $x_j = j \Delta x$, $t^n = n \Delta t$, and the numerical solution $u_j^n$ approximates the exact solution $u(x_j, t^n)$ of the PDE (2) with a periodic boundary condition.

\begin{enumerate}
    \item Prove that the scheme (1) is stable in the sense of (3) for both the $L^2$ norm and the $L^\infty$ norm under the time step restriction $\lambda \leq 1$.
    \item Since the numerical solution $u^n$ is in a finite dimensional space, Student A argues that the stability (3), once proved for a specific norm $\|\cdot\|_a$, would also automatically hold for any other norm $\|\cdot\|_b$. His argument is based on the equivalency of all norms in a finite dimensional space, namely for any two norms $\|\cdot\|_a$ and $\|\cdot\|_b$ on a finite dimensional space $W$, there exists a constant $\delta > 0$ such that
    \[
    \delta \|u\|_b \le \|u\|_a \le \frac{1}{\delta} \|u\|_b.
    \]
    Do you agree with his argument? If not, please explain the mistake made by Student A.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Definition of $L^2$ and $L^\infty$ norms for vectors.
    \item Von Neumann stability analysis (Fourier analysis for finite differences).
    \item Concept of convergence and stability in the limit $\Delta x \to 0$.
\end{itemize}

\textbf{Steps:}
\begin{enumerate}
    \item \textbf{Stability Analysis:}
    \begin{itemize}
        \item \textbf{$L^\infty$ norm:} Rewrite (1) as $u_j^{n+1} = (1-\lambda) u_j^n + \lambda u_{j-1}^n$. If $0 \le \lambda \le 1$, this is a convex combination. Thus $|u_j^{n+1}| \le \max(|u_j^n|, |u_{j-1}^n|) \le \|u^n\|_\infty$. Induction gives $\|u^n\|_\infty \le \|u^0\|_\infty$.
        \item \textbf{$L^2$ norm:} Substitute $u_j^n = \hat{u}^n e^{ij\theta}$. The amplification factor is $g(\theta) = (1-\lambda) + \lambda e^{-i\theta}$. Then $|g(\theta)|^2 = 1 - 2\lambda(1-\lambda)(1-\cos\theta)$. Stability requires $|g(\theta)| \le 1$, which holds for $0 \le \lambda \le 1$.
    \end{itemize}

    \item \textbf{Student A's Mistake:} The equivalence of norms $\|u\|_a \le K \|u\|_b$ holds for a fixed dimension. However, stability requires the constant $C$ to be independent of $\Delta x$. As $\Delta x \to 0$, the dimension $N = 1/\Delta x$ goes to infinity. The equivalence constants often grow with $N$. For example, $\|u\|_2 \le \sqrt{N} \|u\|_\infty$. If a scheme is stable in $L^2$ with $C=1$, then $\|u^n\|_\infty \le \|u^n\|_2 \le \|u^0\|_2 \le \sqrt{N} \|u^0\|_\infty$. The factor $\sqrt{N}$ blows up as $\Delta x \to 0$, so $L^2$ stability does not imply $L^\infty$ stability.
\end{enumerate}

\textbf{Intuition:}
Stability is about the uniform boundedness of the numerical operator as the mesh is refined. While all norms are equivalent in any finite-dimensional space, the ratio between norms can depend on the grid size. True stability must withstand the limit $N \to \infty$.
\end{solution}

\begin{exercise}
\label{applied:2010:12}
We have the following 3 PDEs
\[
u_t + A u_x = 0, \tag{4}
\]
\[
u_t + B u_x = 0, \tag{5}
\]
\[
u_t + C u_x = 0, \quad C = A + B. \tag{6}
\]
Here $u$ is a vector of size $m$ and $A$ and $B$ are $m \times m$ real matrices. We assume $m \geq 2$ and both $A$ and $B$ are diagonalizable with only real eigenvalues. We also assume periodic initial condition for these PDEs.

\begin{enumerate}
    \item Prove that (4) and (5) are both well-posed in the $L^2$-norm. Recall that a PDE is well-posed if its solution satisfies
    \[
    \|u(\cdot, t)\| \leq C(T) \|u(\cdot, 0)\|, \quad 0 \leq t \leq T
    \]
    for a constant $C(T)$ which depends only on $T$.
    \item Is (6) guaranteed to be well-posed as well? If not, give a counter example.
    \item Suppose we have a finite difference scheme $u^{n+1} = A_h u^n$ for approximating (4) and another scheme $u^{n+1} = B_h u^n$ for approximating (5). Suppose both schemes are stable in the $L^2$-norm. If we now form the splitting scheme $u^{n+1} = B_h A_h u^n$ which is a consistent scheme for solving (6), is this scheme guaranteed to be $L^2$ stable as well? If not, give a counter example.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item Well-posedness for linear hyperbolic systems.
    \item Matrix diagonalization and the condition number of the eigenvector matrix.
    \item Operator splitting and the Lax Equivalence Theorem.
\end{itemize}

\textbf{Steps:}
\begin{enumerate}
    \item \textbf{Well-posedness of (4) and (5):} Since $A$ is diagonalizable with real eigenvalues, let $A = PDP^{-1}$. The substitution $v = P^{-1}u$ gives $v_t + Dv_x = 0$, which decouples into scalar advection equations $v_{i,t} + \lambda_i v_{i,x} = 0$. Since $\|v_i(\cdot, t)\|_2 = \|v_i(\cdot, 0)\|_2$, we have $\|u(\cdot, t)\|_2 \le \|P\| \|v(\cdot, t)\|_2 = \|P\| \|P^{-1} u(\cdot, 0)\|_2$. Thus $C(T) = \kappa(P)$ makes it well-posed. The same applies to $B$.

    \item \textbf{Counterexample for (6):} Let $A = \begin{bmatrix} 1 & 1 \\ 0 & -1 \end{bmatrix}$ and $B = \begin{bmatrix} -1 & 0 \\ -2 & 1 \end{bmatrix}$. Both have real eigenvalues $\pm 1$. However, $C = A+B = \begin{bmatrix} 0 & 1 \\ -2 & 0 \end{bmatrix}$ has eigenvalues $\pm i \sqrt{2}$. For a first-order system, complex eigenvalues lead to exponential growth in Fourier space, $e^{|\xi| t \sqrt{2}}$, which is not bounded by any $C(T)$, making it ill-posed.

    \item \textbf{Splitting Scheme Stability:} By the Lax Equivalence Theorem, for a consistent scheme, stability is equivalent to the well-posedness of the target PDE in the grid-independent limit. Since the underlying PDE (6) can be ill-posed, the splitting scheme $u^{n+1} = B_h A_h u^n$ cannot be stable in those cases.
\end{enumerate}

\textbf{Intuition:}
The well-posedness of a linear system is determined by the eigenvalues of the coefficient matrix. Hyperbolicity (real eigenvalues) is necessary. However, the set of matrices with real eigenvalues is not closed under addition. Consequently, the combination of two stable processes can result in an unstable system.
\end{solution}
