\section{2022 Applied & Computational Mathematics}

\begin{exercise}
Consider $\{p_{i}(x)\}_{i=0}^{\infty}$, orthogonal polynomials associated with inner product:
\[
\langle f,g\rangle=\int_{-1}^{1}f(x)g(x)w(x)\,dx,\quad w(x)>0\text{ for }x\in(-1,1),
\]
where $p_{i}(x)$ has degree $i$. Let $x_{0},x_{1},\dots,x_{n}$ be roots of $p_{n+1}(x)$. Construct orthonormal basis in subspace of polynomials degree $\leq n$ such that coefficients in expansion equal scaled values at nodes $x_{0},\dots,x_{n}$.
\end{exercise}

\begin{solution}
Let $V = \mathbb{P}_n$ be the space of polynomials of degree at most $n$. The roots $x_0, x_1, \dots, x_n$ of the orthogonal polynomial $p_{n+1}(x)$ are the nodes for the $(n+1)$-point Gaussian quadrature rule associated with the weight function $w(x)$. Let $w_0, w_1, \dots, w_n$ be the corresponding Gaussian weights, given by
\[ w_k = \int_{-1}^{1} L_k(x) w(x) \, dx = \int_{-1}^{1} L_k^2(x) w(x) \, dx, \]
where $L_k(x) = \prod_{j \neq k} \frac{x-x_j}{x_k-x_j}$ are the Lagrange basis polynomials.

Gaussian quadrature is exact for polynomials of degree up to $2n+1$. Since for any $i, j \in \{0, \dots, n\}$, the product $L_i(x)L_j(x)$ has degree at most $2n$, we have:
\[ \langle L_i, L_j \rangle = \int_{-1}^{1} L_i(x)L_j(x)w(x) \, dx = \sum_{k=0}^{n} w_k L_i(x_k) L_j(x_k). \]
By the property of Lagrange polynomials, $L_i(x_k) = \delta_{ik}$. Thus,
\[ \langle L_i, L_j \rangle = \sum_{k=0}^{n} w_k \delta_{ik} \delta_{jk} = w_i \delta_{ij}. \]
This shows that the Lagrange polynomials are orthogonal with respect to the given inner product. To obtain an orthonormal basis $\{\phi_i\}_{i=0}^n$, we normalize them:
\[ \phi_i(x) = \frac{L_i(x)}{\sqrt{w_i}}. \]
For any polynomial $p(x) \in \mathbb{P}_n$, we can write its expansion in this basis as $p(x) = \sum_{i=0}^n a_i \phi_i(x)$. The coefficients $a_i$ are given by:
\[ a_i = \langle p, \phi_i \rangle = \int_{-1}^{1} p(x) \frac{L_i(x)}{\sqrt{w_i}} w(x) \, dx. \]
Since $p(x)L_i(x)$ has degree at most $2n$, Gaussian quadrature is again exact:
\[ a_i = \sum_{k=0}^n w_k p(x_k) \frac{L_i(x_k)}{\sqrt{w_i}} = \sum_{k=0}^n w_k p(x_k) \frac{\delta_{ik}}{\sqrt{w_i}} = \sqrt{w_i} p(x_i). \]
Thus, the coefficients $a_i$ are precisely the values of the polynomial at the nodes $x_i$ scaled by $\sqrt{w_i}$.
\end{solution}

\begin{exercise}
Consider 2D fixed point iteration:
\[
x_{k+1}=f(x_{k},y_{k}),\quad y_{k+1}=g(x_{k},y_{k}).
\]

Assume $\vec{H}(x,y)=(f(x,y),g(x,y))^{T}$ is $C^{1}$ and Jacobian infinity norm $<1$ at unique fixed point $(x_{\infty},y_{\infty})$.

Consider new iteration:
\[
x_{k+1}=f(x_{k},y_{k}),\quad y_{k+1}=g(x_{k+1},y_{k}).
\]

Prove iteration (2) converges to same fixed point for initial conditions sufficiently close to fixed point.
\end{exercise}

\begin{solution}
Let $(x^*, y^*) = (x_\infty, y_\infty)$ be the unique fixed point. The first iteration can be written as $\mathbf{z}_{k+1} = \vec{H}(\mathbf{z}_k)$, where $\mathbf{z} = (x, y)^T$. Convergence is guaranteed if the spectral radius of the Jacobian $J = \nabla \vec{H}$ at the fixed point is less than 1. The problem states $\|J\|_\infty < 1$, which implies $\rho(J) \leq \|J\|_\infty < 1$.

Now consider the second iteration. Let $e_{x,k} = x_k - x^*$ and $e_{y,k} = y_k - y^*$. Linearizing around the fixed point, we have:
\begin{align*}
e_{x,k+1} &\approx \frac{\partial f}{\partial x} e_{x,k} + \frac{\partial f}{\partial y} e_{y,k} \\
e_{y,k+1} &\approx \frac{\partial g}{\partial x} e_{x,k+1} + \frac{\partial g}{\partial y} e_{y,k} = \frac{\partial g}{\partial x} \left( \frac{\partial f}{\partial x} e_{x,k} + \frac{\partial f}{\partial y} e_{y,k} \right) + \frac{\partial g}{\partial y} e_{y,k}
\end{align*}
Let the Jacobian be $J = \begin{pmatrix} f_x & f_y \\ g_x & g_y \end{pmatrix}$. The error evolution is $e_{k+1} \approx M e_k$ with
\[ M = \begin{pmatrix} f_x & f_y \\ g_x f_x & g_x f_y + g_y \end{pmatrix}. \]
To prove local convergence, we need to show that the spectral radius $\rho(M) < 1$. We compute the infinity norm of $M$:
\[ \|M\|_\infty = \max \{ |f_x| + |f_y|, |g_x f_x| + |g_x f_y + g_y| \}. \]
The first row sum is $|f_x| + |f_y|$, which is the first row sum of $J$. By assumption, $|f_x| + |f_y| \leq \|J\|_\infty < 1$.
For the second row, we have:
\[ |g_x f_x| + |g_x f_y + g_y| \leq |g_x| (|f_x| + |f_y|) + |g_y| \leq |g_x| \|J\|_\infty + |g_y|. \]
Since $\|J\|_\infty < 1$ and $|g_x|, |g_y| \geq 0$, we have:
\[ |g_x| \|J\|_\infty + |g_y| \leq |g_x| + |g_y| \leq \|J\|_\infty < 1. \]
Thus, $\|M\|_\infty \leq \|J\|_\infty < 1$, which implies $\rho(M) < 1$. By the Ostrowski Theorem for fixed point iterations, the iteration converges locally to $(x^*, y^*)$.
\end{solution}

\begin{exercise}
Let $A\in\mathbb{R}^{m\times m}$ with entries $a_{ij}$ satisfying:
\[
a_{ii}\geq\sum_{j\neq i}|a_{ij}|+2,\quad a_{ii}\leq 7.
\]

\begin{enumerate}
\item[(a)] Prove $A^{-1}$ exists.
\item[(b)] Prove $\|A\|_{\infty}$ is max row sum of absolute values.
\item[(c)] Find lower and upper bounds for $\|A\|_{\infty}$.
\item[(d)] If $A=A^{T}$, find bounds for $\|A\|_{2}$ and $\|A^{-1}\|_{2}$.
\end{enumerate}
\end{exercise}

\begin{solution}
(a) The condition $a_{ii} \geq \sum_{j\neq i}|a_{ij}| + 2$ implies that $a_{ii} > \sum_{j\neq i}|a_{ij}|$. This means the matrix $A$ is strictly diagonally dominant. By the Levy-Desplanques Theorem, any strictly diagonally dominant matrix is nonsingular, so $A^{-1}$ exists.

(b) The matrix infinity norm is defined as $\|A\|_\infty = \sup_{x \neq 0} \frac{\|Ax\|_\infty}{\|x\|_\infty}$. For any vector $x$, the $i$-th component of $Ax$ is $(Ax)_i = \sum_{j=1}^m a_{ij} x_j$. Thus, $|(Ax)_i| \leq \sum_j |a_{ij}| |x_j| \leq (\sum_j |a_{ij}|) \|x\|_\infty$. Taking the maximum over $i$ gives $\|Ax\|_\infty \leq (\max_i \sum_j |a_{ij}|) \|x\|_\infty$. By choosing $x_j = \text{sign}(a_{kj})$ for a row $k$ that maximizes the sum, we achieve equality. Thus, $\|A\|_\infty = \max_i \sum_j |a_{ij}|$.

(c) Let $R_i = \sum_{j \neq i} |a_{ij}|$. The $i$-th row sum of absolute values is $|a_{ii}| + R_i = a_{ii} + R_i$ (since $a_{ii} \geq 2$). 
From the given condition, $R_i \leq a_{ii} - 2$. Thus, $a_{ii} + R_i \leq a_{ii} + a_{ii} - 2 = 2a_{ii} - 2$.
Using $a_{ii} \leq 7$, we get $\|A\|_\infty \leq 2(7) - 2 = 12$.
For the lower bound, since $R_i \geq 0$, $a_{ii} + R_i \geq a_{ii} \geq 2$. Thus $2 \leq \|A\|_\infty \leq 12$.

(d) If $A = A^T$, then $\|A\|_2 = \rho(A)$, the spectral radius. By Gershgorin Disk Theorem, all eigenvalues $\lambda$ of $A$ lie in the union of disks $D_i = [a_{ii} - R_i, a_{ii} + R_i]$. Since $a_{ii} - R_i \geq 2$ and $a_{ii} + R_i \leq 12$, all eigenvalues satisfy $2 \leq \lambda \leq 12$. Thus, $2 \leq \|A\|_2 \leq 12$.
For $A^{-1}$, the eigenvalues are $1/\lambda$. Thus, $1/12 \leq \|A^{-1}\|_2 \leq 1/2$.
\end{solution}

\begin{exercise}
Consider ODE system: $\frac{d}{dt}u=f(u)$, $u(0)=u_{0}$.

Assume $f(u)$ has property that forward Euler (FE): $U^{n+1}=U^{n}+kf(U^{n})$ satisfies $\|U^{n+1}\|\leq\|U^{n}\|$ for all $0<k\leq k_{\mathrm{FE}}$.

Consider 2-stage Runge-Kutta:
\[
U^{(1)}=U^{n}+k\beta_{10}f(U^{n}),
\]
\[
U^{n+1}=\{\alpha_{20}U^{n}+k\beta_{20}f(U^{n})\}+\{\alpha_{21}U^{(1)}+k\beta_{21}f(U^{(1)})\}
\]
where $\beta_{10},\beta_{20},\beta_{21},\alpha_{20},\alpha_{21}\geq 0$ and $\alpha_{20}+\alpha_{21}=1$.

\begin{enumerate}
\item[(a)] Prove the RK method also satisfies $\|U^{n+1}\|\leq\|U^{n}\|$ under appropriate step restriction $0\leq k\leq k^{*}$, determining $k^{*}$ in terms of $k_{\mathrm{FE}}$.
\item[(b)] Determine coefficients $\beta_{10},\beta_{20},\beta_{21},\alpha_{20},\alpha_{21}$ so that: (i) method is second-order accurate; (ii) maximum allowed $k^{*}$ is as large as possible.
\end{enumerate}
\end{exercise}

\begin{solution}
(a) The RK method can be viewed as a convex combination of Forward Euler steps.
$U^{(1)} = U^n + k\beta_{10}f(U^n)$ is a FE step with step size $k\beta_{10}$. It is stable if $k\beta_{10} \leq k_{\mathrm{FE}}$, i.e., $k \leq k_{\mathrm{FE}}/\beta_{10}$.
The formula for $U^{n+1}$ is:
$U^{n+1} = \alpha_{20} (U^n + \frac{k\beta_{20}}{\alpha_{20}} f(U^n)) + \alpha_{21} (U^{(1)} + \frac{k\beta_{21}}{\alpha_{21}} f(U^{(1)}))$.
Each term in parentheses is a FE step from $U^n$ and $U^{(1)}$ respectively. They are stable if:
$\frac{k\beta_{20}}{\alpha_{20}} \leq k_{\mathrm{FE}}$ and $\frac{k\beta_{21}}{\alpha_{21}} \leq k_{\mathrm{FE}}$.
Under these conditions, $\|U^{(1)}\| \leq \|U^n\|$ and the terms in parentheses have norms $\leq \|U^n\|$ and $\leq \|U^{(1)}\|$ respectively. Thus,
$\|U^{n+1}\| \leq \alpha_{20} \|U^n\| + \alpha_{21} \|U^{(1)}\| \leq (\alpha_{20} + \alpha_{21}) \|U^n\| = \|U^n\|$.
The total restriction is $k \leq k^* = k_{\mathrm{FE}} \cdot \min \left( \frac{1}{\beta_{10}}, \frac{\alpha_{20}}{\beta_{20}}, \frac{\alpha_{21}}{\beta_{21}} \right)$.

(b) For second-order accuracy, we expand $U^{n+1}$ in Taylor series:
$U^{n+1} = \alpha_{20} U^n + k\beta_{20} f(U^n) + \alpha_{21}(U^n + k\beta_{10}f(U^n)) + k\beta_{21}f(U^n + k\beta_{10}f(U^n))$
$= (\alpha_{20}+\alpha_{21})U^n + k(\beta_{20} + \alpha_{21}\beta_{10} + \beta_{21})f(U^n) + k^2(\beta_{21}\beta_{10})f'(U^n)f(U^n) + O(k^3)$.
Matching with $u(t+k) = u + kf + \frac{k^2}{2}f'f$:
1) $\alpha_{20} + \alpha_{21} = 1$
2) $\beta_{20} + \alpha_{21}\beta_{10} + \beta_{21} = 1$
3) $\beta_{10}\beta_{21} = 1/2$.
To maximize $k^*$, we choose the coefficients of Strong Stability Preserving (SSP) RK2 (Heun's method):
$\alpha_{20} = 1/2, \alpha_{21} = 1/2, \beta_{10} = 1, \beta_{20} = 0, \beta_{21} = 1/2$.
Verification: $1/2+1/2=1$; $0 + 1/2(1) + 1/2 = 1$; $(1)(1/2) = 1/2$.
The stability limit is $k^* = k_{\mathrm{FE}} \cdot \min(1, \infty, \frac{1/2}{1/2}) = k_{\mathrm{FE}}$. This is the maximum possible for a 2nd order 2-stage method.
\end{solution}

\begin{exercise}
Construct third-order accurate Lax-Wendroff-type method for $u_{t}+au_{x}=0$ ($a>0$ constant):

\begin{enumerate}
\item[(a)] Expand $u(t+k,x)$ in Taylor series keeping first four terms. Replace time derivatives by spatial derivatives using equation. Construct cubic polynomial through $U_{j-2}^{n},U_{j-1}^{n},U_{j}^{n},U_{j+1}^{n}$. Approximate spatial derivatives by exact derivatives of this cubic.
\item[(b)] Verify truncation error is $O(k^{3})$ if $h=O(k)$.
\end{enumerate}
\end{exercise}

\begin{solution}
(a) The Taylor expansion in time is:
$u(t+k, x) = u + k u_t + \frac{k^2}{2} u_{tt} + \frac{k^3}{6} u_{ttt} + O(k^4)$.
From $u_t = -au_x$, we have $u_{tt} = a^2 u_{xx}$ and $u_{ttt} = -a^3 u_{xxx}$. Substituting these:
$u(t+k, x) = u - ak u_x + \frac{a^2k^2}{2} u_{xx} - \frac{a^3k^3}{6} u_{xxx} + O(k^4)$.
Let $P(x)$ be the cubic interpolating $\{U_{j-2}, U_{j-1}, U_j, U_{j+1}\}$ at $\{x_{j-2}, x_{j-1}, x_j, x_{j+1}\}$. Using the points $x_i = ih$, the derivatives at $x_j$ are:
$u_x \approx \frac{U_{j-2} - 6U_{j-1} + 3U_j + 2U_{j+1}}{6h}$
$u_{xx} \approx \frac{U_{j-1} - 2U_j + U_{j+1}}{h^2}$ (standard central difference for 2nd derivative)
$u_{xxx} \approx \frac{-U_{j-2} + 3U_{j-1} - 3U_j + U_{j+1}}{h^3}$
The resulting scheme is:
$U_j^{n+1} = U_j^n - \frac{ak}{6h}(U_{j-2} - 6U_{j-1} + 3U_j + 2U_{j+1}) + \frac{a^2k^2}{2h^2}(U_{j-1} - 2U_j + U_{j+1}) - \frac{a^3k^3}{6h^3}(-U_{j-2} + 3U_{j-1} - 3U_j + U_{j+1})$.

(b) The truncation error is $T = \frac{u(t+k) - u(t)}{k} - (\text{finite difference terms})$.
The Taylor series in time was matched up to $O(k^3)$. The spatial derivative approximations derived from a cubic polynomial are:
$P'(x_j) = u_x + O(h^3)$, $P''(x_j) = u_{xx} + O(h^2)$, $P'''(x_j) = u_{xxx} + O(h)$.
Substituting these into the scheme:
$U_j^{n+1} = u - ak(u_x + O(h^3)) + \frac{a^2k^2}{2}(u_{xx} + O(h^2)) - \frac{a^3k^3}{6}(u_{xxx} + O(h))$.
Comparing with the exact Taylor series $u(t+k) = u - aku_x + \frac{a^2k^2}{2}u_{xx} - \frac{a^3k^3}{6}u_{xxx} + O(k^4)$, the error is:
$E = O(k^4) + ak O(h^3) + k^2 O(h^2) + k^3 O(h)$.
If $h = O(k)$, then $E = O(k^4)$. The local truncation error per step is $E/k = O(k^3)$.
\end{solution}

\begin{exercise}
Suppose you have $\$60K$ to invest with 3 options, must invest in multiples of $\$10K$. If $d_{i}$ invested in option $i$, net value is $r_{i}(d_{i})$:
\[
r_{1}(d_{1})=(7d_{1}+2)\times 10,\quad r_{2}(d_{2})=(3d_{2}+7)\times 10,\quad r_{3}(d_{3})=(4d_{3}+5)\times 10,
\]
and $r_{1}(0)=r_{2}(0)=r_{3}(0)$. Maximize net value:
\[
\max_{d_{1},d_{2},d_{3}} r_{1}(d_{1})+r_{2}(d_{2})+r_{3}(d_{3}),
\]
subject to $d_{1}+d_{2}+d_{3}\leq 6$, $d_{i}\geq 0$ integers.
\end{exercise}

\begin{solution}
This is a resource allocation problem that can be solved using Dynamic Programming. Let $r_i(d_i)$ be the return from option $i$ with $d_i$ units of $\$10K$. Note $r_i(0)=0$.
The returns for $d \in \{1, 2, 3, 4, 5, 6\}$ are:
$r_1: [90, 160, 230, 300, 370, 440]$ (formula $70d+20$)
$r_2: [100, 130, 160, 190, 220, 250]$ (formula $30d+70$)
$r_3: [90, 130, 170, 210, 250, 290]$ (formula $40d+50$)

Let $f_i(s)$ be the maximum return using options $1$ to $i$ with total investment $s$.
$f_1(s) = r_1(s)$.
$f_2(s) = \max_{0 \leq d \leq s} \{ r_2(d) + f_1(s-d) \}$.
Computing $f_2(s)$:
$f_2(0) = 0$
$f_2(1) = \max(r_2(1)+0, 0+r_1(1)) = \max(100, 90) = 100$ ($d_2=1$)
$f_2(2) = \max(r_2(2)+0, r_2(1)+r_1(1), 0+r_1(2)) = \max(130, 190, 160) = 190$ ($d_1=1, d_2=1$)
$f_2(3) = \max(r_2(3)+0, r_2(2)+r_1(1), r_2(1)+r_1(2), 0+r_1(3)) = \max(160, 220, 260, 230) = 260$ ($d_1=2, d_2=1$)
$f_2(4) = \max(r_2(4), r_2(3)+r_1(1), r_2(2)+r_1(2), r_2(1)+r_1(3), r_1(4)) = \max(190, 250, 290, 330, 300) = 330$ ($d_1=3, d_2=1$)
$f_2(5) = \max(220, 280, 320, 360, 400, 370) = 400$ ($d_1=4, d_2=1$)
$f_2(6) = \max(250, 310, 350, 390, 430, 470, 440) = 470$ ($d_1=5, d_2=1$)

Finally, $f_3(6) = \max_{0 \leq d \leq 6} \{ r_3(d) + f_2(6-d) \}$:
$d_3=0 \implies f_2(6) = 470$
$d_3=1 \implies r_3(1) + f_2(5) = 90 + 400 = 490$
$d_3=2 \implies r_3(2) + f_2(4) = 130 + 330 = 460$
$d_3=3 \implies r_3(3) + f_2(3) = 170 + 260 = 430$
$d_3=4 \implies r_3(4) + f_2(2) = 210 + 190 = 400$
$d_3=5 \implies r_3(5) + f_2(1) = 250 + 100 = 350$
$d_3=6 \implies r_3(6) + f_2(0) = 290 + 0 = 290$

The maximum net value is $\$490$, achieved with $d_1=4, d_2=1, d_3=1$.
Total investment: $40K + 10K + 10K = 60K$.
Total return: $r_1(4) + r_2(1) + r_3(1) = 300 + 100 + 90 = 490$.
\end{solution}
