\section{2016 Applied \& Computational Mathematics}

\subsection{Individual}

\begin{exercise}
\label{applied:2016:1}
Consider the implicit leapfrog scheme:
\[
\frac{u_{m}^{n+1}-u_{m}^{n-1}}{2k}+a\left(1+\frac{h^{2}}{6}\delta^{2}\right)^{-1}\delta_{0}u_{m}^{n}=f_{m}^{n}
\]
for the one-way wave equation $u_{t}+au_{x}=f$.

Here $\delta^{2}$ is the central second difference operator and $\delta_{0}$ is the central first difference operator.

\begin{enumerate}
\item Show that the scheme is of order $(2,4)$.
\item Show that the scheme is stable if and only if $|ak/h|<1/\sqrt{3}$.
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item We expand the operators in Taylor series. The time derivative approximation is:
    \[ \frac{u(x, t+k) - u(x, t-k)}{2k} = u_t + \frac{k^2}{6} u_{ttt} + O(k^4). \]
    The spatial operators are:
    \[ \delta_0 u = \frac{u(x+h) - u(x-h)}{2h} = u_x + \frac{h^2}{6} u_{xxx} + \frac{h^4}{120} u_{xxxxx} + O(h^6), \]
    \[ \delta^2 u = \frac{u(x+h) - 2u(x) + u(x-h)}{h^2} = u_{xx} + \frac{h^2}{12} u_{xxxx} + O(h^4). \]
    The implicit operator $(1 + \frac{h^2}{6} \delta^2)^{-1}$ can be expanded as $1 - \frac{h^2}{6} \delta^2 + O(h^4)$. Thus:
    \begin{align*}
        \left(1 + \frac{h^2}{6} \delta^2\right)^{-1} \delta_0 u &= \left(1 - \frac{h^2}{6} \frac{\partial^2}{\partial x^2} + O(h^4)\right) \left(u_x + \frac{h^2}{6} u_{xxx} + O(h^4)\right) \\
        &= u_x + \frac{h^2}{6} u_{xxx} - \frac{h^2}{6} u_{xxx} + O(h^4) = u_x + O(h^4).
    \end{align*}
    Substituting back into the scheme, we get $u_t + O(k^2) + a(u_x + O(h^4)) = f$, which proves the order is $(2, 4)$.

    \item Let $f=0$ and substitute the ansatz $u_m^n = g^n e^{im\xi h}$. Let $\theta = \xi h$. We have:
    \[ \delta_0 e^{im\theta} = \frac{i \sin \theta}{h} e^{im\theta}, \quad \delta^2 e^{im\theta} = -\frac{4 \sin^2(\theta/2)}{h^2} e^{im\theta}. \]
    The amplification factor $g$ satisfies:
    \[ \frac{g - g^{-1}}{2k} + a \frac{1}{1 - \frac{h^2}{6} \frac{4 \sin^2(\theta/2)}{h^2}} \frac{i \sin \theta}{h} = 0. \]
    Using $1 - \frac{2}{3} \sin^2(\theta/2) = \frac{2 + \cos \theta}{3}$, the equation becomes:
    \[ g^2 + 2i \lambda \phi(\theta) g - 1 = 0, \quad \text{where } \lambda = \frac{ak}{h}, \, \phi(\theta) = \frac{3 \sin \theta}{2 + \cos \theta}. \]
    For stability, we require $|g| \leq 1$. The roots are $g = -i \lambda \phi(\theta) \pm \sqrt{1 - \lambda^2 \phi(\theta)^2}$. Stability holds if $\lambda^2 \phi(\theta)^2 \leq 1$ for all $\theta$. 
    Analyzing $\phi(\theta)$: $\phi'(\theta) = \frac{3(6\cos \theta + 3)}{(2+\cos \theta)^2}$. The maximum occurs at $\cos \theta = -1/2$ (i.e., $\theta = 2\pi/3$), where $\phi(2\pi/3) = \sqrt{3}$.
    Thus, stability requires $|\lambda| \sqrt{3} \leq 1$, or $|ak/h| \leq 1/\sqrt{3}$.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2016:2}
A simple version of an enzyme-mediated chemical reaction is:
\[
\mathrm{S}+\mathrm{E}\xleftarrow[k_{2}]{k_{1}}\mathrm{C}\xrightarrow[k_{3}]{k_{3}}\mathrm{P}+\mathrm{E}
\]
where $S$ is substrate reactant, $P$ is product concentration, and $E$ is enzyme (catalyst).

Assume initial conditions: $S(0)=S_{0}$, $E(0)=E_{0}$, $C(0)=0$, $P(0)=0$; $\mathbf{k}_{1},\mathbf{k}_{2},\mathbf{k}_{3}$ are reaction rate constants.

\begin{enumerate}
\item[(a)] Convert the chemical reaction into rate equations (ODEs) for $S(T)$, $E(T)$, $C(T)$, $P(T)$. Nondimensionalize using:
\[
T=t/(k_{1}E_{0}),\quad S(T)=S_{0}s(t),\quad P(T)=S_{0}p(t),\quad E(T)=E_{0}e(t),\quad C(T)=E_{0}c(t),
\]
\[
\epsilon=\frac{E_{0}}{S_{0}}\ll 1,\quad \lambda=\frac{k_{2}}{k_{1}S_{0}},\quad \mu=\frac{k_{2}+k_{3}}{k_{1}S_{0}}.
\]
\item[(b)] Use expansions $s(t)=s_{0}(t)+\epsilon s_{1}(t)+O(\epsilon^{2})$, etc., to determine equations for leading order slow solution. Show $s_{0}(t)$ and $p_{0}(t)$ satisfy Michaelis-Menten equations:
\[
\dot{s}_{0}(t)=-(\mu-\lambda)\frac{s_{0}}{\mu+s_{0}},\quad \dot{p}_{0}(t)=(\mu-\lambda)\frac{s_{0}}{\mu+s_{0}}.
\]
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item[(a)] The law of mass action yields:
    \begin{align*}
        \frac{dS}{dT} &= -k_1 SE + k_2 C, \\
        \frac{dE}{dT} &= -k_1 SE + (k_2 + k_3) C, \\
        \frac{dC}{dT} &= k_1 SE - (k_2 + k_3) C, \\
        \frac{dP}{dT} &= k_3 C.
    \end{align*}
    From $dE/dT + dC/dT = 0$, we have $E+C = E_0$, so $e+c=1$. Using the provided scaling $t = k_1 E_0 T$:
    \begin{align*}
        \frac{ds}{dt} &= -s(1-c) + \lambda c, \\
        \epsilon \frac{dc}{dt} &= s(1-c) - \mu c, \\
        \frac{dp}{dt} &= (\mu - \lambda) c.
    \end{align*}
    
    \item[(b)] In the limit $\epsilon \to 0$, the fast equation for $c$ reaches quasi-steady state:
    \[ s_0(1-c_0) - \mu c_0 = 0 \implies c_0 = \frac{s_0}{s_0 + \mu}. \]
    Substituting $c_0$ into the slow equations for $s$ and $p$:
    \[ \dot{s}_0 = -s_0(1-c_0) + \lambda c_0 = - \left( s_0 \frac{\mu}{s_0 + \mu} - \lambda \frac{s_0}{s_0 + \mu} \right) = -(\mu - \lambda) \frac{s_0}{s_0 + \mu}. \]
    Similarly,
    \[ \dot{p}_0 = (\mu - \lambda) c_0 = (\mu - \lambda) \frac{s_0}{s_0 + \mu}. \]
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2016:3}
A vector $\mathbf{u}=(u_{1},\dots,u_{n})\in\mathbb{N}^{n}$ is multiplicatively dependent if there is a nonzero vector $\mathbf{k}=(k_{1},\dots,k_{n})\in\mathbb{Z}^{n}$ for which:
\[
u_{1}^{k_{1}}\dots u_{n}^{k_{n}}=1. \tag{1}
\]

This is important in factorization and primality testing algorithms. Prove that if $\mathbf{u}=(u_{1},\dots,u_{n})\in\mathbb{N}^{n}$ is multiplicatively dependent with $\|\mathbf{u}\|_{\infty}\leq H$ where $\|\mathbf{u}\|_{\infty}=\max_{1\leq i\leq n}|u_{i}|$, then there exists a nonzero $\mathbf{k}=(k_{1},\dots,k_{n})\in\mathbb{Z}^{n}$ with:
\[
\|\mathbf{k}\|_{\infty}\leq\left(\frac{2n\log H}{\log 2}\right)^{n-1}.
\]

Thus for fixed $n$, it can be found in polynomial time of order $(\log H)^{n(n-1)}$.

Comment: To solve this problem, you can use the following statement (without proof) which informally means that if a system of homogeneous equations with integer coefficients has a nontrivial solution then it has an integer solution with reasonably small components.

Let $A=(a_{ij})_{i,j=1}^{m,n}$ be an $m\times n$ matrix of rank $r\leq n-1$ with integer entries of size at most $H$. Then there is a nonzero integer vector $\mathbf{x}=(x_{1},\dots,x_{n})\in\mathbb{Z}^{n}$ such that $A\mathbf{x}=\mathbf{0}$ and $\|\mathbf{x}\|_{\infty}\leq(2nH)^{n-1}$.
\end{exercise}

\begin{solution}
Let $p_1, \dots, p_m$ be all the distinct prime factors of the integers $u_1, \dots, u_n$. We can write each $u_i$ as a product of these primes:
\[ u_i = \prod_{j=1}^m p_j^{a_{ji}}, \quad a_{ji} \in \mathbb{N} \cup \{0\}. \]
The condition $u_1^{k_1} \dots u_n^{k_n} = 1$ is equivalent to:
\[ \prod_{j=1}^m p_j^{\sum_{i=1}^n a_{ji} k_i} = 1 \iff \sum_{i=1}^n a_{ji} k_i = 0 \text{ for each } j=1, \dots, m. \]
This is a system of linear equations $A\mathbf{k} = \mathbf{0}$, where $A = (a_{ji})$ is an $m \times n$ matrix. Since $\mathbf{u}$ is multiplicatively dependent, there exists a nonzero $\mathbf{k} \in \mathbb{Z}^n$ in the kernel, implying $\text{rank}(A) \leq n-1$.
The entries $a_{ji}$ are bounded as follows: $2^{a_{ji}} \leq p_j^{a_{ji}} \leq u_i \leq H$, so $a_{ji} \leq \frac{\log H}{\log 2}$.
Applying the provided statement with the bound $H' = \frac{\log H}{\log 2}$ on the entries of $A$, there exists a nonzero $\mathbf{k} \in \mathbb{Z}^n$ such that:
\[ \|\mathbf{k}\|_\infty \leq (2nH')^{n-1} = \left( \frac{2n \log H}{\log 2} \right)^{n-1}. \]
The complexity result follows as we search for $\mathbf{k}$ in a box of volume $(2\|\mathbf{k}\|_\infty + 1)^n$, which is $O((\log H)^{n(n-1)})$.
\end{solution}

\begin{exercise}
\label{applied:2016:4}
Consider a symmetric matrix $A_{n\times n}$ with simple eigenvalue $\lambda_{i}$ satisfying $| \lambda_{j}-\lambda_{i}|=O(1)$ for $j\neq i$.

In inverse iteration to compute eigenvalue and eigenvector, one solves:
\[
(A-\mu I)y_{k+1}=x_{k},
\]
where $\mu$ is approximation of $\lambda_{i}$, $\|x_{k}\|=1$, and obtains $x_{k+1}=y_{k+1}/\|y_{k+1}\|$.

However, for $\mu$ close to $\lambda_{i}$, $A-\mu I$ has very small eigenvalue and the linear system is ill-conditioned. There may be large error in numerical solution $\tilde{y}_{k+1}$. Even with large error in $\tilde{y}_{k+1}$, the $\tilde{x}_{k+1}=\tilde{y}_{k+1}/\|\tilde{y}_{k+1}\|$ is accurate.

\begin{enumerate}
\item[(1)] $\tilde{y}_{k+1}$ satisfies $(A-\mu I+\delta A)\tilde{y}_{k+1}=x_{k}$ where $\|\delta A\|=O(\epsilon)$ and $\epsilon$ is machine precision. Show:
\[
\|(A-\lambda_{i})\frac{\tilde{y}_{k+1}}{\|\tilde{y}_{k+1}\|}\|\leq|\mu-\lambda_{i}|+\|\delta A\|+\frac{1}{\|\tilde{y}_{k+1}\|}.
\]
\item[(2)] Let $\alpha_{i}=x_{k}^{t}q_{i}$ where $q_{i}$ is normalized eigenvector for $\lambda_{i}$. Show:
\[
\|\tilde{y}_{k+1}\|\geq\frac{|\alpha_{i}|}{|\mu-\lambda_{i}|+\|\delta A\|}.
\]
\item[(3)] Conclude: $\|x_{k+1}-(\pm)q_{i}\|=O(|\lambda_{i}-\mu|+\epsilon)$.
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item[(1)] From $(A - \mu I + \delta A) \tilde{y}_{k+1} = x_k$, we rewrite $(A - \lambda_i I) \tilde{y}_{k+1}$ as:
    \[ (A - \lambda_i I) \tilde{y}_{k+1} = (A - \mu I + \delta A) \tilde{y}_{k+1} + (\mu - \lambda_i) \tilde{y}_{k+1} - \delta A \tilde{y}_{k+1} = x_k + (\mu - \lambda_i) \tilde{y}_{k+1} - \delta A \tilde{y}_{k+1}. \]
    Dividing by $\|\tilde{y}_{k+1}\|$ and taking the norm:
    \begin{align*}
        \left\| (A - \lambda_i I) \frac{\tilde{y}_{k+1}}{\|\tilde{y}_{k+1}\|} \right\| &\leq \frac{\|x_k\|}{\|\tilde{y}_{k+1}\|} + |\mu - \lambda_i| \frac{\|\tilde{y}_{k+1}\|}{\|\tilde{y}_{k+1}\|} + \|\delta A\| \frac{\|\tilde{y}_{k+1}\|}{\|\tilde{y}_{k+1}\|} \\
        &= \frac{1}{\|\tilde{y}_{k+1}\|} + |\mu - \lambda_i| + \|\delta A\|.
    \end{align*}
    
    \item[(2)] Project the relation $x_k = (A - \mu I + \delta A) \tilde{y}_{k+1}$ onto $q_i$:
    \[ \alpha_i = q_i^T x_k = q_i^T (A - \lambda_i I + (\lambda_i - \mu) I + \delta A) \tilde{y}_{k+1}. \]
    Since $q_i^T (A - \lambda_i I) = 0$, we have $\alpha_i = (\lambda_i - \mu) q_i^T \tilde{y}_{k+1} + q_i^T \delta A \tilde{y}_{k+1}$. Thus:
    \[ |\alpha_i| \leq |\lambda_i - \mu| \|q_i\| \|\tilde{y}_{k+1}\| + \|q_i\| \|\delta A\| \|\tilde{y}_{k+1}\| = (|\lambda_i - \mu| + \|\delta A\|) \|\tilde{y}_{k+1}\|, \]
    which gives $\|\tilde{y}_{k+1}\| \geq \frac{|\alpha_i|}{|\mu - \lambda_i| + \|\delta A\|}$.
    
    \item[(3)] From (1) and (2), we see that the residual $\eta = \|(A - \lambda_i) \tilde{x}_{k+1}\|$ is bounded by $O(|\lambda_i - \mu| + \epsilon)$. For a symmetric matrix with a simple eigenvalue $\lambda_i$, the distance from a unit vector $\tilde{x}_{k+1}$ to the eigenspace spanned by $q_i$ is $O(\eta / \text{gap})$, where $\text{gap} = \min_{j \neq i} |\lambda_j - \lambda_i|$. Since the gap is $O(1)$, we conclude $\|\tilde{x}_{k+1} - (\pm q_i)\| = O(|\lambda_i - \mu| + \epsilon)$.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2016:5}
A function $f:\mathbb{R}^{n}\to\mathbb{R}$ in $C^{2} is called strongly convex if its Hessian satisfies $\nabla^{2}f\succeq mI$ for some $m>0$. Show the following are equivalent:

\begin{enumerate}
\item[(a)] $f$ is strongly convex: $\nabla^{2}f(x)\succeq mI$ for all $x$.
\item[(b)] For any $t\in[0,1]$, any $x,y\in\mathbb{R}^{n}$:
\[
f(tx+(1-t)y)\leq tf(x)+(1-t)f(y)-\frac{m}{2}t(1-t)\|x-y\|^{2}.
\]
\item[(c)] $\langle\nabla f(x)-\nabla f(y),x-y\rangle\geq m\|x-y\|^{2}$ for any $x,y\in\mathbb{R}^{n}$.
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{itemize}
    \item \textbf{(a) $\implies$ (b):} Let $g(x) = f(x) - \frac{m}{2}\|x\|^2$. Then $\nabla^2 g(x) = \nabla^2 f(x) - mI \succeq 0$, so $g$ is convex. The convexity of $g$ implies:
    \[ g(tx + (1-t)y) \leq tg(x) + (1-t)g(y). \]
    Substituting back $f$:
    \[ f(tx+(1-t)y) - \frac{m}{2}\|tx+(1-t)y\|^2 \leq t\left(f(x) - \frac{m}{2}\|x\|^2\right) + (1-t)\left(f(y) - \frac{m}{2}\|y\|^2\right). \]
    Rearranging and using the identity $\|tx+(1-t)y\|^2 = t\|x\|^2 + (1-t)\|y\|^2 - t(1-t)\|x-y\|^2$, we obtain:
    \[ f(tx+(1-t)y) \leq tf(x) + (1-t)f(y) - \frac{m}{2}t(1-t)\|x-y\|^2. \]

    \item \textbf{(b) $\implies$ (c):} Divide (b) by $t$ and rearrange:
    \[ \frac{f(y+t(x-y)) - f(y)}{t} \leq f(x) - f(y) - \frac{m}{2}(1-t)\|x-y\|^2. \]
    Taking the limit $t \to 0$ yields $f(x) \geq f(y) + \langle \nabla f(y), x-y \rangle + \frac{m}{2}\|x-y\|^2$. Summing this with the symmetric inequality for $f(y)$ gives:
    \[ f(x) + f(y) \geq f(y) + f(x) + \langle \nabla f(y), x-y \rangle + \langle \nabla f(x), y-x \rangle + m\|x-y\|^2, \]
    which simplifies to $\langle \nabla f(x) - \nabla f(y), x-y \rangle \geq m\|x-y\|^2$.

    \item \textbf{(c) $\implies$ (a):} For any $v \in \mathbb{R}^n$ and $t > 0$, (c) with $y = x+tv$ gives:
    \[ \langle \nabla f(x+tv) - \nabla f(x), tv \rangle \geq m\|tv\|^2 \implies \frac{\langle \nabla f(x+tv) - \nabla f(x), v \rangle}{t} \geq m\|v\|^2. \]
    Taking $t \to 0$ gives the directional derivative of the gradient, which is $v^T \nabla^2 f(x) v \geq m\|v\|^2$. Since this holds for all $v$, $\nabla^2 f(x) \succeq mI$.
\end{itemize}
\end{solution}

\subsection{Team}

\begin{exercise}
\label{applied:2016:6}
For solving the following partial differential equation
\begin{equation}
u_t + u_x = 0, \quad -\infty \leq x \leq \infty \tag{1}
\end{equation}
with compactly supported initial condition, we consider the following one-step, three-point scheme on a uniform mesh $x_j = j\Delta x$ with spatial mesh size $\Delta x$:
\begin{equation}
u_j^{n+1} = a u_j^n + b u_{j-1}^n + c u_{j-2}^n, \quad j = \dots, -1, 0, 1, \dots \tag{2}
\end{equation}
where $a, b, c$ are constants which may depend on the mesh ratio $\lambda = \frac{\Delta t}{\Delta x}$. Here $\Delta t$ is the time step, and $u_j^n$ approximates the exact solution at $u(x_j, t^n)$ with $t^n = n\Delta t$.
\begin{enumerate}
    \item Find the constants $a, b, c$ such that the scheme is second order accurate.
    \item Find the CFL number $\lambda_0$ such that the scheme, with the constants determined by the step above, is stable in $L^2$ under the time step restriction $\lambda \leq \lambda_0$.
    \item If the PDE is defined on $(0, \infty)$ with an initial condition compactly supported in $(0, \infty)$ and a boundary condition $u(0, t) = g(t)$, how would you modify the scheme so that it can be applied? Can you prove the stability and accuracy of your modified scheme?
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item We determine the coefficients $a, b, c$ by Taylor expanding $u(x_j, t^{n+1})$ and comparing with the right hand side.
    Let $x = x_j, t = t^n$. Expanding around $(x, t)$:
    \[ u(x, t+\Delta t) = u + \Delta t u_t + \frac{\Delta t^2}{2} u_{tt} + O(\Delta t^3) = u - \lambda \Delta x u_x + \frac{\lambda^2 \Delta x^2}{2} u_{xx} + O(\Delta x^3) \]
    since $u_t = -u_x$. The RHS is:
    \[ a u + b(u - \Delta x u_x + \frac{\Delta x^2}{2} u_{xx}) + c(u - 2\Delta x u_x + 2\Delta x^2 u_{xx}) + O(\Delta x^3) \]
    Matching terms in $u, u_x, u_{xx}$:
    \begin{itemize}
        \item $a+b+c = 1$
        \item $b+2c = \lambda$
        \item $b+4c = \lambda^2$
    \end{itemize}
    Solving this system gives $2c = \lambda^2 - \lambda \implies c = \frac{\lambda(\lambda-1)}{2}$; $b = \lambda - 2c = \lambda(2-\lambda)$; and $a = 1 - b - c = \frac{(\lambda-1)(\lambda-2)}{2}$.
    \item This is the Beam-Warming scheme. The amplification factor is $g(\theta) = a + b e^{-i\theta} + c e^{-2i\theta}$. Analysis of $|g(\theta)|$ shows that the scheme is stable for $0 \leq \lambda \leq 2$. Thus the CFL number is $\lambda_0 = 2$.
    \item For the problem on $(0, \infty)$, the scheme requires $u_{j-1}$ and $u_{j-2}$ which are available for $j \ge 2$. At $j=1$, $u_{-1}$ is needed, and at $j=0$, the value is given by $u(0, t) = g(t)$. We can use a different scheme (like Lax-Wendroff) at the point $j=1$ or a first-order upwind to maintain stability. To maintain second-order accuracy, a second-order boundary treatment or ghost-point extrapolation is required.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2016:7}
Inverse problem. Answer the famous Mark Kac's question: ``can you hear the shape of drum?'' for the special case.
Consider the one-dimensional oscillator $\ddot{x} = -u'(x)$ with symmetric potential $u(-x) = u(x)$, $u(0) = u'(0) = 0$, $u'(x) > 0$ for $x > 0$, $\lim_{x \to \infty} u(x) = \infty$. Denote the inverse function of $y = u(x)$, $x \geq 0$ as $x = u^{-1}(y) = \phi(y)$.
\begin{enumerate}
    \item[(a)] For any solution $x(t)$, show there is a conservation of energy
    \[
    \frac{\dot{x}^2(t)}{2} + u(x(t)) \equiv e
    \]
    where $e$ is a constant.
    \item[(b)] For any energy $e > 0$, find a periodic solution with total energy $e$. Show that the period is given by
    \[
    P(e) = 2\sqrt{2} \int_0^{\phi(e)} \frac{dx}{\sqrt{e - u(x)}}.
    \]
    \item[(c)] Show that
    \[
    \phi(z) = \frac{1}{2\pi\sqrt{2}} \int_0^z \frac{P(e) de}{\sqrt{z - e}}.
    \]
    \item[(d)] In the case of iso-chronous $P(e) \equiv 2\pi$, show that $\phi(z) = \sqrt{2z}$. Then you have $u(x) = \frac{1}{2}x^2$, $x(t) = a\cos(t) + b\sin(t)$, the famous harmonic oscillator.
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item[(a)] Total energy $E = \frac{1}{2}\dot{x}^2 + u(x)$. Taking the time derivative: $\dot{E} = \dot{x}\ddot{x} + u'(x)\dot{x} = \dot{x}(\ddot{x} + u'(x)) = 0$ since $\ddot{x} = -u'(x)$. thus $E$ is constant.
    \item[(b)] For energy $e$, $\dot{x} = \pm\sqrt{2(e-u(x))}$. The particle moves between $\pm \phi(e)$. The time to go from $0$ to $\phi(e)$ is $T = \int_0^{\phi(e)} \frac{dx}{\sqrt{2(e-u(x))}}$. The period is $P(e) = 4T = 2\sqrt{2} \int_0^{\phi(e)} \frac{dx}{\sqrt{e - u(x)}}$.
    \item[(c)] Change variable $u(x) = y$, so $x = \phi(y)$ and $dx = \phi'(y) dy$:
    \[ P(e) = 2\sqrt{2} \int_0^e \frac{\phi'(y)}{\sqrt{e-y}} dy. \]
    This is Abel's integral equation for $\phi'$. Multiplying by $(z-e)^{-1/2}$ and integrating:
    \[ \int_0^z \frac{P(e)}{\sqrt{z-e}} de = 2\sqrt{2} \int_0^z \int_0^e \frac{\phi'(y)}{\sqrt{(z-e)(e-y)}} dy de = 2\sqrt{2} \int_0^z \phi'(y) \left( \int_y^z \frac{de}{\sqrt{(z-e)(e-y)}} \right) dy. \]
    The inner integral is $\pi$. Thus $\int_0^z \frac{P(e)}{\sqrt{z-e}} de = 2\sqrt{2} \pi \phi(z)$, yielding $\phi(z) = \frac{1}{2\pi\sqrt{2}} \int_0^z \frac{P(e) de}{\sqrt{z-e}}$.
    \item[(d)] If $P(e) = 2\pi$, then $\phi(z) = \frac{1}{\sqrt{2}} \int_0^z \frac{de}{\sqrt{z-e}} = [-\sqrt{2}\sqrt{z-e}]_0^z = \sqrt{2z}$. Thus $x = \sqrt{2u(x)} \implies u(x) = \frac{1}{2}x^2$.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2016:8}
The following statement informally means that if a system of homogeneous equations with integer coefficients has a nontrivial solution then it has an integer solution with reasonably small components.
Let $A = (a_{ij})$ be an $m \times n$ matrix of rank $r \leq n-1$ with integer entries of size at most $H$, that is, $|a_{ij}| \leq H$.
Prove that there is an integer non-zero vector $\mathbf{x} = (x_1, \ldots, x_n) \in \mathbb{Z}^n$ such that $A\mathbf{x} = \mathbf{0}$ and
\[
\|\mathbf{x}\|_\infty \leq (2nH)^{n-1}
\]
where $\|\mathbf{x}\|_\infty = \max_{1 \leq i \leq n} |x_i|$.
\end{exercise}

\begin{solution}
This is a statement of Siegel's Lemma. Let $X$ be a positive integer. Consider the set of vectors $S = \{ \mathbf{x} \in \mathbb{Z}^n : 0 \le x_i \le X \}$. The size of $S$ is $(X+1)^n$. For each $\mathbf{x} \in S$, let $\mathbf{y} = A\mathbf{x}$. Since $A$ has rank $r$ and entries bounded by $H$, the components $y_j$ are bounded by $nHX$ in absolute value. The number of possible values for $\mathbf{y}$ in the $r$-dimensional image of $A$ is bounded by $(2nHX+1)^r$. By the pigeonhole principle, if $(X+1)^n > (2nHX+1)^r$, there exist $\mathbf{x}_1 \neq \mathbf{x}_2$ such that $A\mathbf{x}_1 = A\mathbf{x}_2$. Then $\mathbf{x} = \mathbf{x}_1 - \mathbf{x}_2$ is a non-zero integer solution with $\|\mathbf{x}\|_\infty \leq X$. Setting $r = n-1$ and choosing $X = (2nH)^{n-1}$ satisfies the required bound.
\end{solution}

\begin{exercise}
\label{applied:2016:9}
This problem considers an iterative scheme $x_{k+1} = x_k + \beta_k p_k$ for the linear system $Ax = b$, where $A \in \mathbb{R}^{n \times n}$ is a given non-singular matrix and $b \in \mathbb{R}^n$. $x_k$ denotes the approximate solution at the $k$-th iteration, $\beta_k$ is a scalar and $p_k \in \mathbb{R}^n$ is a search direction. If $x_k$ is given, $x_{k+1}$ is determined such that the residual $r_{k+1} := b - Ax_{k+1}$ is minimized with respect to the 2-norm.
\begin{enumerate}
    \item Determine $\beta_k$.
    \item Prove that the residual $r_{k+1}$ is orthogonal to $Ap_k$ with respect to the usual inner-product.
    \item Prove that the residuals satisfy $\|r_{k+1}\| \leq \|r_k\| \sin(\alpha)$ where $\alpha$ is the angle between $r_k$ and $Ap_k$.
    \item Assume that the inner product of $r_k$ and $Ap_k$ is non-zero. Will the above scheme always converge?
    \item Assume that $A$ is positive definite. We take the search direction $p_k = r_k$. Show that the above scheme converges for any initial guess $x_0$.
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item $\beta_k$ minimizes $\|r_{k+1}\|^2 = \|r_k - \beta_k A p_k\|^2 = \|r_k\|^2 - 2\beta_k r_k^T A p_k + \beta_k^2 \|A p_k\|^2$. Setting the derivative to zero gives $\beta_k = \frac{r_k^T A p_k}{\|A p_k\|^2}$.
    \item $r_{k+1}^T A p_k = (r_k - \beta_k A p_k)^T A p_k = r_k^T A p_k - \frac{r_k^T A p_k}{\|A p_k\|^2} \|A p_k\|^2 = 0$.
    \item $\|r_{k+1}\|^2 = \|r_k\|^2 - \frac{(r_k^T A p_k)^2}{\|A p_k\|^2} = \|r_k\|^2 (1 - \cos^2 \alpha) = \|r_k\|^2 \sin^2 \alpha$, so $\|r_{k+1}\| = \|r_k\| |\sin \alpha|$.
    \item It depends on the sequence $p_k$. If $p_k$ are chosen such that $\alpha$ is bounded away from $\pi/2$, then $\|r_k\| \to 0$. However, if $A p_k$ becomes orthogonal to $r_k$, the method can stall.
    \item If $A$ is positive definite and $p_k = r_k$, then $r_k^T A r_k \geq \lambda_{min} \|r_k\|^2$ and $\|A r_k\| \leq \lambda_{max} \|r_k\|$. Thus $\cos \alpha \geq \lambda_{min} / \lambda_{max} = 1/\kappa(A) > 0$. Since $\sin^2 \alpha = 1 - \cos^2 \alpha \le 1 - 1/\kappa(A)^2 < 1$, the residual norm strictly decreases and the scheme converges.
\end{enumerate}
\end{solution}

\begin{exercise}
\label{applied:2016:10}
Let $f: \mathbb{R}^n \to \mathbb{R}$ be convex and in $C^1$. Suppose $f$ has a local minimum $x^*$.
\begin{enumerate}
    \item Must this local minimum $x^*$ be a global minimum?
    \item Consider the following backward gradient method: starting from any $x^0 \in \mathbb{R}^n$ define
    \[
    x^k = x^{k-1} - t \nabla f(x^k), \quad k \geq 1,
    \]
    where $t > 0$ is a fixed step size. Do you need any condition on $t$ to guarantee $\{f(x^k)\}$ converge? Prove your convergence argument, if $\{f(x^k)\}$ converges.
    \item Suppose $f$ is strongly convex, that is, $\exists m > 0$ such that $\langle \nabla f(x) - \nabla f(y), x-y \rangle \geq m \|x-y\|^2$. Under this additional condition, show that $\{x^k\}$ converges.
\end{enumerate}
\end{exercise}

\begin{solution}
\begin{enumerate}
    \item Yes. For a convex function, if $x^*$ is not a global minimum, there exists $y$ such that $f(y) < f(x^*)$. By convexity, the function values on the segment $[x^*, y]$ are strictly less than $f(x^*)$ near $x^*$, contradicting local optimality.
    \item The scheme defines $x^k = \text{argmin}_x \{ f(x) + \frac{1}{2t} \|x - x^{k-1}\|^2 \}$. Thus $f(x^k) + \frac{1}{2t} \|x^k - x^{k-1}\|^2 \le f(x^{k-1})$. This implies $f(x^k)$ is a non-increasing sequence. Since $f$ is bounded below, it converges. No condition on $t > 0$ is needed for monotonicity.
    \item Using strong convexity: $\langle \nabla f(x^k) - \nabla f(x^*), x^k - x^* \rangle \geq m \|x^k - x^*\|^2$. Substituting $\nabla f(x^k) = \frac{x^{k-1} - x^k}{t}$, we get $\langle x^{k-1} - x^k, x^k - x^* \rangle \geq mt \|x^k - x^*\|^2$.
    Using $\|x^{k-1} - x^*\|^2 = \|x^{k-1} - x^k + x^k - x^*\|^2 = \|x^{k-1} - x^k\|^2 + \|x^k - x^*\|^2 + 2\langle x^{k-1} - x^k, x^k - x^* \rangle \geq \|x^k - x^*\|^2 (1 + 2mt)$. Thus $\|x^k - x^*\| \leq \frac{1}{\sqrt{1+2mt}} \|x^{k-1} - x^*\|$, which implies convergence.
\end{enumerate}
\end{solution}
