\section{2013 Analysis \& Differential Equations}

\subsection{Individual}

\begin{exercise}
Suppose that $f$ is an integrable function on $\mathbb{R}^d$. For each $\alpha>0$, let $E_{\alpha}=\{x\mid |f(x)|>\alpha\}$. Prove that:
\[
\int_{\mathbb{R}^d}|f(x)|\,dx = \int_0^{\infty} m(E_{\alpha})\,d\alpha.
\]
\end{exercise}

\begin{solution}
We use Fubini's theorem (or specifically Tonelli's theorem for non-negative functions) to express the integral of $|f(x)|$ as a double integral. Note that for any $x \in \mathbb{R}^d$, we have:
\[
|f(x)| = \int_0^{|f(x)|} 1 \, d\alpha = \int_0^\infty \chi_{\{\alpha < |f(x)|\}}(\alpha) \, d\alpha
\]
where $\chi$ denotes the characteristic function. Substituting this into the integral over $\mathbb{R}^d$, we get:
\[
\int_{\mathbb{R}^d} |f(x)| \, dx = \int_{\mathbb{R}^d} \left( \int_0^\infty \chi_{\{\alpha < |f(x)|\}}(\alpha) \, d\alpha \right) dx.
\]
Since the integrand $\chi_{\{\alpha < |f(x)|\}}(\alpha)$ is non-negative and measurable on $\mathbb{R}^d \times (0, \infty)$, we can swap the order of integration by Tonelli's theorem:
\[
\int_{\mathbb{R}^d} |f(x)| \, dx = \int_0^\infty \left( \int_{\mathbb{R}^d} \chi_{\{\alpha < |f(x)|\}}(x) \, dx \right) d\alpha.
\]
The inner integral is exactly the measure of the set where $|f(x)| > \alpha$:
\[
\int_{\mathbb{R}^d} \chi_{\{x \mid |f(x)| > \alpha\}}(x) \, dx = m(\{x \in \mathbb{R}^d \mid |f(x)| > \alpha\}) = m(E_\alpha).
\]
Thus, we obtain:
\[
\int_{\mathbb{R}^d} |f(x)| \, dx = \int_0^\infty m(E_\alpha) \, d\alpha.
\]
\end{solution}

\begin{exercise}
Let $p(z)$ be a polynomial of degree $d\geq 2$, with distinct roots $a_1,a_2,\cdots,a_d$. Show that
\[
\sum_{i=1}^d \frac{1}{p'(a_i)} = 0.
\]
\end{exercise}

\begin{solution}
Consider the rational function $f(z) = \frac{1}{p(z)}$. Since $p(z)$ is a polynomial of degree $d \geq 2$ with distinct roots $a_1, \dots, a_d$, the function $f(z)$ has simple poles at each $a_i$. The residue of $f(z)$ at $z = a_i$ is given by:
\[
\text{Res}(f, a_i) = \lim_{z \to a_i} (z - a_i) \frac{1}{p(z)} = \frac{1}{p'(a_i)}.
\]
By the Residue Theorem, if $C_R$ is a circle of radius $R$ centered at the origin that encloses all roots $a_1, \dots, a_d$, then:
\[
\oint_{C_R} \frac{1}{p(z)} \, dz = 2\pi i \sum_{i=1}^d \text{Res}(f, a_i) = 2\pi i \sum_{i=1}^d \frac{1}{p'(a_i)}.
\]
Now we analyze the behavior of the integral as $R \to \infty$. Since $p(z)$ is a polynomial of degree $d$, for large $|z|$, we have $|p(z)| \geq C |z|^d$ for some constant $C > 0$. Therefore, on the circle $C_R$ ($|z| = R$):
\[
\left| \frac{1}{p(z)} \right| \leq \frac{1}{C R^d}.
\]
The length of the contour $C_R$ is $2\pi R$. Thus, the integral is bounded by:
\[
\left| \oint_{C_R} \frac{1}{p(z)} \, dz \right| \leq 2\pi R \cdot \frac{1}{C R^d} = \frac{2\pi}{C R^{d-1}}.
\]
Since $d \geq 2$, the exponent $d-1 \geq 1$, so as $R \to \infty$, the integral tends to $0$:
\[
\lim_{R \to \infty} \oint_{C_R} \frac{1}{p(z)} \, dz = 0.
\]
It follows that:
\[
2\pi i \sum_{i=1}^d \frac{1}{p'(a_i)} = 0 \implies \sum_{i=1}^d \frac{1}{p'(a_i)} = 0.
\]
\end{solution}

\begin{exercise}
Let $\alpha$ be a number such that $\alpha/\pi$ is not a rational number.
\begin{enumerate}
    \item Show that for any integer $k$:
    \[
    \lim_{N\to\infty} \frac{1}{N} \sum_{n=1}^N e^{ik(x+n\alpha)} = \frac{1}{2\pi}\int_{-\pi}^{\pi} e^{ikt}\,dt.
    \]
    \item For every continuous periodic function $f:\mathbb{R}\to\mathbb{C}$ of period $2\pi$, we have
    \[
    \lim_{N\to\infty} \frac{1}{N}\sum_{n=1}^N f(x+n\alpha) = \frac{1}{2\pi}\int_{-\pi}^{\pi} f(t)\,dt.
    \]
\end{enumerate}
\end{exercise}

\begin{solution}
First, consider the case $f(t) = e^{ikt}$ for an integer $k$.
If $k = 0$, then $f(t) = 1$. The sum is $\frac{1}{N} \sum_{n=1}^N 1 = 1$, and the integral is $\frac{1}{2\pi} \int_{-\pi}^\pi 1 \, dt = 1$. The equality holds.
If $k \neq 0$, the integral is $\frac{1}{2\pi} \int_{-\pi}^\pi e^{ikt} \, dt = 0$.
The sum is:
\[
\frac{1}{N} \sum_{n=1}^N e^{ik(x+n\alpha)} = \frac{e^{ikx}}{N} \sum_{n=1}^N (e^{ik\alpha})^n = \frac{e^{ikx}}{N} e^{ik\alpha} \frac{1 - e^{ikN\alpha}}{1 - e^{ik\alpha}}.
\]
Since $\alpha/\pi \notin \mathbb{Q}$, $k\alpha$ is not a multiple of $2\pi$, so $e^{ik\alpha} \neq 1$ and the denominator is non-zero. The numerator is bounded in absolute value by $|e^{ikx} \cdot e^{ik\alpha} \cdot 2| = 2$. Thus:
\[
\left| \frac{1}{N} \sum_{n=1}^N e^{ik(x+n\alpha)} \right| \leq \frac{1}{N} \cdot \frac{2}{|1 - e^{ik\alpha}|} \to 0 \text{ as } N \to \infty.
\]
This proves the result for any $k \neq 0$.

Now let $f$ be any continuous $2\pi$-periodic function. By the Stone-Weierstrass theorem, $f$ can be uniformly approximated by trigonometric polynomials. For any $\epsilon > 0$, there exists a trigonometric polynomial $P(t) = \sum_{k=-M}^M c_k e^{ikt}$ such that $\sup_t |f(t) - P(t)| < \epsilon$.
Then:
\[
\left| \frac{1}{N} \sum_{n=1}^N f(x+n\alpha) - \int f \right| \leq \left| \frac{1}{N} \sum (f - P) \right| + \left| \frac{1}{N} \sum P - \int P \right| + \left| \int P - \int f \right|
\]
where $\int$ denotes $\frac{1}{2\pi} \int_{-\pi}^\pi$.
The first and third terms are bounded by $\epsilon$. The middle term tends to 0 as $N \to \infty$ because it is a finite linear combination of terms that tend to 0. Since $\epsilon$ is arbitrary, the limit holds.
\end{solution}

\begin{exercise}
Let $u$ be a positive harmonic function over the punctured complex plane $\mathbb{C}\setminus\{0\}$. Show that $u$ must be a constant function.
\end{exercise}

\begin{solution}
Consider the universal covering map $p: \mathbb{C} \to \mathbb{C} \setminus \{0\}$ given by $p(z) = e^z$.
Define $v(z) = u(e^z)$. Since $u$ is harmonic on $\mathbb{C} \setminus \{0\}$, $v$ is harmonic on $\mathbb{C}$ (the composition of a harmonic function with a holomorphic function is harmonic).
Because $u$ is a positive function, $v$ is also a positive harmonic function on the entire complex plane $\mathbb{C} \cong \mathbb{R}^2$.
Liouville's theorem for harmonic functions states that any harmonic function on $\mathbb{R}^n$ that is bounded from below (or above) must be constant.
Since $v(z) > 0$ for all $z \in \mathbb{R}^2$, it is bounded from below, and therefore $v$ is constant.
Since $v(z) = u(e^z)$ is constant and the map $e^z$ is surjective onto $\mathbb{C} \setminus \{0\}$, the original function $u$ must also be constant.
\end{solution}

\begin{exercise}
Suppose $H=L^2(B)$, $B$ is the unit ball in $\mathbb{R}^d$. Let $K(x,y)$ be a measurable function on $B\times B$ that satisfies
\[
|K(x,y)| \leq A|x-y|^{-d+\alpha}
\]
for some $\alpha>0$, whenever $x,y\in B$. Define
\[
Tf(x) = \int_B K(x,y)f(y)\,dy.
\]
\begin{enumerate}
    \item Prove that $T$ is a bounded operator on $H$.
    \item Prove that $T$ is compact.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{1. Boundedness:}
We use Schur's Test. An operator $T$ with kernel $K$ is bounded on $L^2$ if there exist constants $C_1, C_2$ such that:
\[
\int_B |K(x,y)| \, dy \leq C_1 \text{ for a.e. } x \in B, \quad \int_B |K(x,y)| \, dx \leq C_2 \text{ for a.e. } y \in B.
\]
Given $|K(x,y)| \leq A |x-y|^{-d+\alpha}$, we compute:
\[
\int_B |x-y|^{-d+\alpha} \, dy \leq \int_{|z| \leq 2} |z|^{-d+\alpha} \, dz
\]
where we used the fact that for $x, y \in B$, $|x-y| \leq 2$. In polar coordinates in $\mathbb{R}^d$:
\[
\int_{|z| \leq 2} |z|^{-d+\alpha} \, dz = \omega_{d-1} \int_0^2 r^{-d+\alpha} r^{d-1} \, dr = \omega_{d-1} \int_0^2 r^{\alpha-1} \, dr = \omega_{d-1} \frac{2^\alpha}{\alpha} < \infty
\]
where $\omega_{d-1}$ is the surface area of the unit sphere. Since the kernel is symmetric in its bound, the same estimate holds for the integral over $x$. Thus $T$ is bounded on $L^2(B)$.

\textbf{2. Compactness:}
Define a sequence of kernels $K_n(x,y) = K(x,y)$ if $|x-y| \geq 1/n$ and $0$ otherwise.
Each $K_n$ is a bounded function on $B \times B$ (since $|x-y| \geq 1/n$ implies $|K| \leq A n^{d-\alpha}$).
Since $B$ is a set of finite measure, $K_n \in L^2(B \times B)$.
A kernel operator is Hilbert-Schmidt if its kernel is in $L^2(B \times B)$. Hilbert-Schmidt operators are compact.
Now we show that $T_n \to T$ in the operator norm. $T - T_n$ is the operator with kernel $G_n(x,y) = K(x,y)$ if $|x-y| < 1/n$.
By the same Schur's test estimate:
\[
\|T - T_n\| \leq \sup_x \int_{|x-y| < 1/n} A |x-y|^{-d+\alpha} \, dy = A \omega_{d-1} \int_0^{1/n} r^{\alpha-1} \, dr = \frac{A \omega_{d-1}}{\alpha n^\alpha}.
\]
As $n \to \infty$, $\|T - T_n\| \to 0$. Since $T$ is the limit of compact operators in the operator norm, $T$ is compact.
\end{solution}

\begin{exercise}
Let $A$ be an $n\times n$ real non-degenerate symmetric matrix. For $\lambda\in\mathbb{R}^+$, we define:
\[
\int_{-\infty}^{\infty} \exp\left(i\lambda x^2\right)dx = \lim_{\epsilon\to 0^+}\int_{-\infty}^{\infty} \exp\left(i\lambda x^2 - \frac{1}{2}\epsilon x^2\right)dx.
\]
Show that:
\[
\int_{\mathbb{R}^n} \exp\left(i\frac{\lambda}{2}\langle Ax,x\rangle - i\langle x,\xi\rangle\right)dx = \left(\frac{2\pi}{\lambda}\right)^{n/2}|\det A|^{-1/2}\exp\left(-\frac{i}{2\lambda}\langle A^{-1}\xi,\xi\rangle\right)\exp\left(\frac{i\pi}{4}\operatorname{sgn}A\right).
\]
Here $\lambda\in\mathbb{R}^+$, $\xi\in\mathbb{R}^n$, $\operatorname{sgn}(A) = \nu_+(A) - \nu_-(A)$, where $\nu_+(A)$ ($\nu_-(A)$) is the number of positive (negative) eigenvalues of $A$.
\end{exercise}

\begin{solution}
Let $A = Q D Q^T$ be the spectral decomposition of $A$, where $Q$ is orthogonal and $D = \text{diag}(\mu_1, \dots, \mu_n)$ is the diagonal matrix of eigenvalues.
Change variables $x = Q y$. Then $dx = dy$, $\langle Ax, x \rangle = \sum \mu_j y_j^2$, and $\langle x, \xi \rangle = \langle y, Q^T \xi \rangle = \sum y_j \eta_j$ where $\eta = Q^T \xi$.
The integral factorizes:
\[
\int_{\mathbb{R}^n} \exp\left(i \sum \left( \frac{\lambda}{2} \mu_j y_j^2 - \eta_j y_j \right) \right) dy = \prod_{j=1}^n \int_{-\infty}^\infty \exp\left( i \frac{\lambda \mu_j}{2} y_j^2 - i \eta_j y_j \right) dy_j.
\]
Completing the square in the exponent: $i \frac{\lambda \mu_j}{2} y_j^2 - i \eta_j y_j = i \frac{\lambda \mu_j}{2} \left( y_j - \frac{\eta_j}{\lambda \mu_j} \right)^2 - \frac{i \eta_j^2}{2 \lambda \mu_j}$.
The 1D integral is:
\[
I_j = \exp\left( - \frac{i \eta_j^2}{2 \lambda \mu_j} \right) \int_{-\infty}^\infty \exp\left( i \frac{\lambda \mu_j}{2} z^2 \right) dz.
\]
Using the definition $\int \exp(ia z^2) = \sqrt{\pi/|a|} \exp(i \frac{\pi}{4} \text{sgn}(a))$:
\[
\int_{-\infty}^\infty \exp\left( i \frac{\lambda \mu_j}{2} z^2 \right) dz = \sqrt{\frac{2\pi}{\lambda |\mu_j|}} \exp\left( i \frac{\pi}{4} \text{sgn}(\mu_j) \right).
\]
Multiplying all $I_j$ together:
\[
\prod_{j=1}^n \sqrt{\frac{2\pi}{\lambda |\mu_j|}} \exp\left( i \frac{\pi}{4} \text{sgn}(\mu_j) \right) \exp\left( - \frac{i \eta_j^2}{2 \lambda \mu_j} \right) = \left( \frac{2\pi}{\lambda} \right)^{n/2} \left( \prod |\mu_j| \right)^{-1/2} \exp\left( \frac{i \pi}{4} \sum \text{sgn}(\mu_j) \right) \exp\left( - \frac{i}{2\lambda} \sum \frac{\eta_j^2}{\mu_j} \right).
\]
Observe that:
\begin{itemize}
    \item $\prod |\mu_j| = |\det A|$.
    \item $\sum \text{sgn}(\mu_j) = \nu_+(A) - \nu_-(A) = \text{sgn}(A)$.
    \item $\sum \frac{\eta_j^2}{\mu_j} = \langle D^{-1} \eta, \eta \rangle = \langle D^{-1} Q^T \xi, Q^T \xi \rangle = \langle Q D^{-1} Q^T \xi, \xi \rangle = \langle A^{-1} \xi, \xi \rangle$.
\end{itemize}
Substituting these back gives the desired result:
\[
\left( \frac{2\pi}{\lambda} \right)^{n/2} |\det A|^{-1/2} \exp\left( \frac{i \pi}{4} \text{sgn}(A) \right) \exp\left( - \frac{i}{2\lambda} \langle A^{-1} \xi, \xi \rangle \right).
\]
\end{solution}

\subsection{Team}

\begin{exercise}
Suppose $\Delta = \{z \in \mathbb{C} \mid |z| < 1\}$ is the open unit disk in the complex plane. Show that for any holomorphic function $f: \Delta \to \Delta$,
\[
\frac{|f'(z)|}{1 - |f(z)|^2} \leq \frac{1}{1 - |z|^2} \tag{1}
\]
for all $z$ in $\Delta$. If equality holds in (1) for some $z_0 \in \Delta$, show that $f \in \operatorname{Aut}(\Delta)$, and that the equality holds for all $z \in \Delta$.
\end{exercise}

\begin{solution}
This is the Schwarz-Pick Theorem.
Let $\psi_a(z) = \frac{a-z}{1-\bar{a}z}$ be an automorphism of $\Delta$ that maps $a$ to $0$.
Define $g = \psi_{f(z_0)} \circ f \circ \psi_{z_0}^{-1}$.
Then $g(0) = \psi_{f(z_0)}(f(z_0)) = 0$.
By Schwarz Lemma, $|g'(0)| \leq 1$ and $|g(w)| \leq |w|$.
Using the chain rule:
$g'(0) = \psi_{f(z_0)}'(f(z_0)) \cdot f'(z_0) \cdot (\psi_{z_0}^{-1})'(0)$.
Since $\psi_a'(a) = \frac{-1}{1-|a|^2}$ and $(\psi_a^{-1})'(0) = -(1-|a|^2)$:
$|g'(0)| = \left| \frac{-1}{1-|f(z_0)|^2} \right| \cdot |f'(z_0)| \cdot |1-|z_0|^2| = \frac{|f'(z_0)|(1-|z_0|^2)}{1-|f(z_0)|^2} \leq 1$.
This gives the desired inequality.
If equality holds at $z_0$, then $|g'(0)| = 1$, so by Schwarz Lemma $g(w) = e^{i\theta} w$ for some $\theta$, which is an automorphism.
Thus $f = \psi_{f(z_0)}^{-1} \circ g \circ \psi_{z_0}$ is a composition of automorphisms, hence $f \in \operatorname{Aut}(\Delta)$.
For any automorphism, the equality in Schwarz-Pick holds everywhere.
\end{solution}

\begin{exercise}
Let $f$ be a function of bounded variation on $[a, b]$, $f_1$ its generalized derivative as a measure, i.e. $f(x) - f(a) = \int_a^x f_1(y) dy$ for every $x \in [a, b]$ and $f_1(x)$ is an integrable function on $[a, b]$. Let $f'$ be its weak derivative as a generalized function, i.e. $\int_a^b f(x) g'(x) dx = -\int_a^b f'(x) g(x) dx$, for any smooth function $g(x)$ on $[a, b], g(a) = g(b) = 0$. Show that:
\begin{enumerate}
    \item If $f$ is absolutely continuous, then $f' = f_1$.
    \item If the weak derivative $f'$ of $f$ is an integrable function on $[a, b]$, then $f(x)$ is equal to an absolutely continuous function outside a set of measure zero.
\end{enumerate}
\end{exercise}

\begin{solution}
a) If $f$ is absolutely continuous, then $f(x) = f(a) + \int_a^x f'(x) dx$ where $f'$ is the derivative almost everywhere. By the fundamental theorem of calculus for Lebesgue integrals, $f_1 = f'$ a.e.
Plugging into the weak derivative formula and using integration by parts for absolutely continuous functions:
$\int_a^b f(x) g'(x) dx = [f(x)g(x)]_a^b - \int_a^b f'(x) g(x) dx = -\int_a^b f'(x) g(x) dx$.
Thus the weak derivative coincides with the pointwise derivative $f_1$.

b) If the weak derivative $f'$ is integrable, let $F(x) = \int_a^x f'(t) dt$. $F$ is absolutely continuous.
Then for any test function $g$:
$\int f g' = -\int f' g = -\int F' g = \int F g'$.
So $\int (f-F) g' = 0$ for all $g$.
This implies $f-F$ is constant almost everywhere.
Thus $f(x) = F(x) + C$ a.e., where $F(x)+C$ is absolutely continuous.
\end{solution}

\begin{exercise}
Show that the convex hull of the roots of any polynomial contains all its critical points as well as all the zeros of higher derivatives of the polynomial.
\end{exercise}

\begin{solution}
This is the Gauss-Lucas Theorem.
Let $P(z) = a \prod_{i=1}^n (z - z_i)$. Then
$\frac{P'(z)}{P(z)} = \sum_{i=1}^n \frac{1}{z - z_i}$.
If $z$ is a root of $P'(z)$ but not $P(z)$, then $\sum \frac{1}{z - z_i} = 0$.
Taking the conjugate: $\sum \frac{1}{\bar{z} - \bar{z}_i} = \sum \frac{z - z_i}{|z - z_i|^2} = 0$.
This can be written as $z \sum \frac{1}{|z - z_i|^2} = \sum \frac{z_i}{|z - z_i|^2}$.
So $z = \frac{\sum w_i z_i}{\sum w_i}$ where $w_i = \frac{1}{|z - z_i|^2} > 0$.
Thus $z$ is a convex combination of $z_i$, so $z$ is in the convex hull.
If $z$ is also a root of $P(z)$, it is already in the convex hull.
By induction, the roots of any higher derivative $P^{(k)}(z)$ are in the convex hull of the roots of $P^{(k-1)}(z)$, which is contained in the convex hull of the roots of $P(z)$.
\end{solution}

\begin{exercise}
Let $D \subset \mathbb{R}^3$ be an open domain. Show that every smooth vector field $\mathbf{F} = (P, Q, R)$ over $D$ can be written as $\mathbf{F} = \mathbf{F}_1 + \mathbf{F}_2$ such that $\operatorname{rot}(\mathbf{F}_1) = 0$, $\operatorname{div}(\mathbf{F}_2) = 0$.
\end{exercise}

\begin{solution}
This is the Helmholtz Decomposition.
We want $\mathbf{F} = \nabla \phi + \nabla \times \mathbf{A}$.
Then $\operatorname{rot}(\mathbf{F}_1) = \operatorname{rot}(\nabla \phi) = 0$ and $\operatorname{div}(\mathbf{F}_2) = \operatorname{div}(\nabla \times \mathbf{A}) = 0$.
For a domain like $\mathbb{R}^3$ or a bounded domain with boundary conditions, we solve Poisson's equation.
$\operatorname{div} \mathbf{F} = \operatorname{div} \mathbf{F}_1 = \Delta \phi$.
Let $\phi(x) = -\frac{1}{4\pi} \int_D \frac{\operatorname{div} \mathbf{F}(y)}{|x-y|} dy$.
Then $\mathbf{F}_1 = \nabla \phi$ satisfies the condition.
Then $\mathbf{F}_2 = \mathbf{F} - \mathbf{F}_1$ satisfies $\operatorname{div} \mathbf{F}_2 = \operatorname{div} \mathbf{F} - \Delta \phi = 0$.
\end{solution}

\begin{exercise}
Let $\mathbf{H}$ be a Hilbert space and $\mathbf{A}$ a compact self-adjoint linear operator over $\mathbf{H}$. Show that there exists an orthonormal basis of $\mathbf{H}$ consisting of eigenvectors $\varphi_n$ of $\mathbf{A}$ with non-zero eigenvalues $\lambda_n$ such that every vector $\xi \in \mathbf{H}$ can be written as: $\xi = \sum_k c_k \varphi_k + \xi'$, where $\xi' \in \operatorname{Ker} \mathbf{A}$. We also have $\mathbf{A}\xi = \sum_k \lambda_k c_k \varphi_k$. If there are infinitely many eigenvectors, then $\lim_{n \to \infty} \lambda_n = 0$.
\end{exercise}

\begin{solution}
This is the Spectral Theorem for compact self-adjoint operators.
The nonzero eigenvalues are real and have finite multiplicity.
Let $\{\lambda_k\}$ be the distinct nonzero eigenvalues and $E_k$ the corresponding finite-dimensional eigenspaces.
Because $\mathbf{A}$ is self-adjoint, eigenspaces for different eigenvalues are orthogonal.
Let $\{\varphi_{k,j}\}$ be an orthonormal basis for $E_k$.
The span of these eigenvectors is dense in $(\operatorname{Ker} \mathbf{A})^\perp$.
Thus $\mathbf{H} = \overline{\bigoplus E_k} \oplus \operatorname{Ker} \mathbf{A}$.
Any $\xi \in \mathbf{H}$ can be projected onto the kernel and the sum of eigenspaces.
The fact that $\lambda_n \to 0$ follows from the compactness of $\mathbf{A}$. If $\lambda_n$ did not go to zero, there would be an infinite-dimensional subspace where $\mathbf{A}$ is bounded away from zero, contradicting compactness.
\end{solution}

\begin{exercise}
A function $f : \mathbb{R} \to \mathbb{R}$ is called convex if $f(\lambda x + (1 - \lambda) x') \leq \lambda f(x) + (1 - \lambda) f(x')$ for $0 \leq \lambda \leq 1$ and each $x, x' \in \mathbb{R}$. We assume $|f(x)| < \infty$ for all $x$.
\begin{enumerate}
    \item Show that a convex function $f$ is continuous and the function $g(y) = \max_{x \in \mathbb{R}} (xy - f(x))$ is a well-defined convex function over $\mathbb{R}$.
    \item Show that a convex function $f$ is differentiable except at most countably many points.
    \item $f$ is differentiable everywhere if both $f$ and $g$ are strictly convex.
\end{enumerate}
\end{exercise}

\begin{solution}
a) \textbf{Continuity:} For any $x$, consider $x-h, x, x+h$. By convexity, the slope is increasing. This implies the function is locally bounded and Lipschitz, hence continuous.
\textbf{Legendre Transform:} $g(y) = \sup_x (xy - f(x))$. Since $xy-f(x)$ is linear in $y$, $g(y)$ is the supremum of linear functions, hence convex. It is well-defined if $f(x)$ grows faster than linearly at infinity.

b) \textbf{Differentiability:} A convex function has monotonic left and right derivatives $f'_-(x) \leq f'_+(x)$. These are monotonic functions, so they have at most countably many jump discontinuities. $f$ is differentiable where $f'_-(x) = f'_+(x)$.

c) Strict convexity of $g$ implies that the maximizer $x$ in $xy-f(x)$ is unique for each $y$. This uniqueness is related to the differentiability of $f$. Specifically, $f$ is differentiable if and only if $g$ is strictly convex (under certain growth conditions).
\end{solution}


