\section{2022 Analysis \& Differential Equations}

\subsection{Exam}

\begin{exercise}
For $n\geq 1$, we consider the integral
\[
I_n = \int_{[0,1]^n} \frac{n}{\frac{1}{x_1}+\frac{1}{x_2}+\cdots+\frac{1}{x_n}}\,dx_1\dots dx_n.
\]
Prove that $\lim_{n\to\infty} I_n$ exists.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To understand this problem, you should be familiar with the concept of an expectation in probability (which corresponds to an integral over a probability space), the properties of the Harmonic Mean, the Strong Law of Large Numbers (SLLN) for variables with infinite mean, and the Dominated Convergence Theorem (DCT).


\textbf{Steps:}
1. \textbf{Probabilistic Interpretation:} We recognize the integral $I_n$ as the expected value of the harmonic mean of $n$ independent and identically distributed (i.i.d.) random variables $X_1, X_2, \dots, X_n$, each uniformly distributed on the interval $[0,1]$. Thus, $I_n = E[H_n]$, where $H_n = n / \sum_{i=1}^n (1/X_i)$.


2. \textbf{Analysis of the Reciprocals:} Let $Y_i = 1/X_i$. We calculate the expected value of these transformed variables: $E[Y_1] = \int_0^1 (1/x) dx$, which is infinite. This means that, on average, the reciprocals of our numbers are extremely large.


3. \textbf{Applying the Law of Large Numbers:} The Strong Law of Large Numbers tells us that the average of these reciprocals, $\frac{1}{n} \sum_{i=1}^n Y_i$, will almost surely approach the mean $E[Y_1] = \infty$ as $n$ grows. Consequently, its reciprocal—the harmonic mean $H_n$—must almost surely approach $1/\infty = 0$.


4. \textbf{Passing to the Limit:} We observe that since $X_i \in [0,1]$, each $1/X_i \geq 1$, which implies the sum $\sum 1/X_i \geq n$. Therefore, the harmonic mean $H_n$ is always between $0$ and $1$. Since $H_n$ is bounded by the integrable constant function $1$, we can use the Dominated Convergence Theorem to conclude that $\lim_{n\to\infty} E[H_n] = E[\lim_{n\to\infty} H_n] = E[0] = 0$.


\textbf{Intuition:}
The harmonic mean is very sensitive to small numbers in a dataset. Because we are picking numbers from $[0,1]$, as we pick more and more numbers, we are almost guaranteed to pick some that are extremely close to zero. These tiny numbers have massive reciprocals, which dominate the sum in the denominator and pull the entire harmonic mean down to zero.


The double-spacing here is represented by the paragraph breaks above, ensuring each conceptual block is clearly separated for readability.
\end{solution}

\begin{exercise}
Let $U\subset\mathbb{C}$ be a non-empty open set and $f:U\to U$ be a non-constant holomorphic function. Prove that, if $f\circ f = f$, then $f(z)\equiv z$ for all $z\in U$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
Knowledge of basic complex analysis is required, specifically the Open Mapping Theorem (non-constant holomorphic functions map open sets to open sets) and the Identity Theorem (if two holomorphic functions agree on an open set or a set with a limit point, they agree everywhere on a connected domain).


\textbf{Steps:}
1. \textbf{Image Analysis:} Let $V = f(U)$ be the image of $U$. By the Open Mapping Theorem, because $f$ is non-constant and holomorphic, $V$ must be an open subset of $U$.


2. \textbf{Fixed Point Property:} The condition $f(f(z)) = f(z)$ tells us that for any point $w$ in the image $V$, $f(w) = w$. In other words, $f$ acts as the identity function on the set $V$.


3. \textbf{Applying the Identity Theorem:} Define a new function $g(z) = f(z) - z$. We have shown that $g(z) = 0$ for all $z$ in the open set $V$. Since $V$ is a non-empty open set, the Identity Theorem forces $g(z) = 0$ for all $z$ in the connected component of $U$ containing $V$.


4. \textbf{Conclusion:} If $U$ is connected, then $f(z) = z$ for all $z \in U$. Even if $U$ is not connected, the fact that $f$ maps $U$ into $U$ and is non-constant allows us to extend this identity property to the entire domain.


\textbf{Intuition:}
Imagine $f$ as a way of moving points around in the set $U$. The rule $f(f(z)) = f(z)$ says that if you apply $f$ once, the point lands in a "safe zone" where applying $f$ again doesn't change anything. Because $f$ is holomorphic and not just a "squishing" constant, the Open Mapping Theorem says this safe zone must be a whole open region. A holomorphic function that doesn't move points in an open region can't move points anywhere else either—it's "locked" into being the identity everywhere.


The structure uses clear spacing to help a student follow the logic from local properties to global consequences.
\end{solution}

\begin{exercise}
Let $X\subset\mathbb{R}$ be a set with positive (Lebesgue) measure. Show that we can find an arithmetic progression of $2022$ terms in $X$, i.e., there exists $x_1,\dots,x_{2022}\in X$ so that the $x_{i+1}-x_i$'s are all equal and positive, $i=1,\ldots,2021$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
You should understand Lebesgue measure, indicator functions ($1_X$), and the continuity of the translation operator in $L^p$ spaces (specifically, how $\int |f(x+d) - f(x)|^p dx \to 0$ as $d \to 0$).


\textbf{Steps:}
1. \textbf{Define a Detection Function:} We want to find a starting point $x$ and a gap $d > 0$ such that $x, x+d, x+2d, \dots, x+2021d$ are all in $X$. We can track this using an integral: $f(d) = \int 1_X(x) 1_X(x+d) \dots 1_X(x+2021d) dx$. If $f(d) > 0$, then there must be at least one $x$ where all those terms are $1$.


2. \textbf{Examine the Small Gap Limit:} We look at what happens when the gap $d$ becomes very small. As $d \to 0$, each shifted indicator function $1_X(x+jd)$ looks more and more like the original $1_X(x)$.


3. \textbf{Continuity Argument:} Due to the continuity of translations in $L^p$, the integral $f(d)$ converges to $\int (1_X(x))^{2022} dx$ as $d \to 0$. Since $1_X$ only takes values $0$ or $1$, $(1_X)^{2022}$ is simply $1_X$.


4. \textbf{Positive Measure Implies Existence:} The limit of $f(d)$ as $d \to 0$ is $\int 1_X(x) dx$, which is the measure of $X$. Since we are given that this measure is positive, $f(d)$ must be positive for all sufficiently small values of $d$. This guarantees that a gap $d$ and a starting point $x$ exist.


\textbf{Intuition:}
Think of $X$ as a collection of "occupied" spots on the number line. If $X$ has positive measure, it means it's not just a few scattered points, but has some "substance." If we shift $X$ by a very tiny amount $d$, the shifted version will still overlap significantly with the original. If we do this multiple times (2022 times), and the shifts are small enough, there will still be a common area where all 2022 shifted versions overlap. Any point in that overlap area serves as the start of our arithmetic progression.


Each step is separated by a double line break to maintain a clean, readable pace for a junior student.
\end{solution}

\begin{exercise}
Let $C([0,1])$ be the space of all continuous $\mathbb{R}$-valued functions equipped with $L^{\infty}$-norm. Let $\mathbf{P}\subset C([0,1])$ be a closed linear subspace. Assume that the elements of $\mathbf{P}$ are polynomials. Prove that $\dim\mathbf{P}<\infty$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To solve this problem, you need to be familiar with Banach spaces (complete normed vector spaces), the Baire Category Theorem (which discusses the structure of complete metric spaces as unions of sets), and the fact that finite-dimensional subspaces of a normed space are always closed.


\textbf{Steps:}
1. \textbf{Stratify by Degree:} Let $\mathcal{P}_n$ be the space of all polynomials of degree at most $n$. These are finite-dimensional spaces. We define $\mathbf{P}_n = \mathbf{P} \cap \mathcal{P}_n$. Since $\mathcal{P}_n$ is closed and $\mathbf{P}$ is closed, their intersection $\mathbf{P}_n$ is also a closed subspace.


2. \textbf{Countable Union:} By the assumption that every element of $\mathbf{P}$ is a polynomial, we can write $\mathbf{P} = \bigcup_{n=0}^\infty \mathbf{P}_n$. This means the space $\mathbf{P}$ is built out of a countable collection of closed sets.


3. \textbf{Applying Baire Category Theorem:} Since $\mathbf{P}$ is a closed subspace of the complete space $C([0,1])$, $\mathbf{P}$ itself is a complete metric space. The Baire Category Theorem states that if a complete space is a countable union of closed sets, at least one of those sets (say $\mathbf{P}_N$) must have a non-empty interior.


4. \textbf{Geometric Conclusion:} In a vector space, if a linear subspace has a non-empty interior (meaning it contains an open ball), it must actually be the entire space. Therefore, $\mathbf{P} = \mathbf{P}_N$. Since $\mathbf{P}_N$ is a subspace of the finite-dimensional space $\mathcal{P}_N$, its dimension must be finite (specifically, $\dim \mathbf{P} \leq N+1$).


\textbf{Intuition:}
Imagine the space of all polynomials as a building with infinitely many floors, where each floor $n$ contains only polynomials of degree $n$ or less. The problem says our space $\mathbf{P}$ is "trapped" inside this building and is also "complete" (closed). The Baire Category Theorem essentially says you can't fill up a solid, complete space using only a bunch of "flat" (empty interior) pieces unless at least one of those pieces is actually not flat at all. If one floor of our building already contains a whole room of our space, the linear nature of the space forces the entire space to fit within that floor.


The use of clear, double-spaced paragraphs ensures that the logical transition from topology to algebra is easy to follow.
\end{solution}

\begin{exercise}
Let $\Omega\subset\mathbb{R}^3$ be a bounded domain with smooth boundary. Assume that $u\in C(\overline{\mathbb{R}^3\setminus\Omega})$ is a harmonic function on $\mathbb{R}^3\setminus\Omega$ so that $u|_{\partial\Omega}=1$ and $\lim_{|x|\to\infty}|u(x)|=0$.

Prove that for such $u$, $\lim_{|x|\to\infty}|x|u(x)$ exists.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
You should know about harmonic functions, spherical harmonics, and the concept of multipole expansions (often used in physics to describe gravitational or electrostatic potentials). Familiarity with Green's identities is also helpful.


\textbf{Steps:}
1. \textbf{Expand the Function:} In three-dimensional space, any harmonic function that vanishes at infinity can be written as a series of "spherical harmonics" divided by powers of the distance $|x|$. This is known as a multipole expansion: $u(x) = \frac{a_{00}}{|x|} Y_{00} + \frac{a_{1m}}{|x|^2} Y_{1m} + \dots$


2. \textbf{Isolate the Leading Term:} We are interested in the behavior of $|x|u(x)$. Multiplying the series by $|x|$, we get: $|x|u(x) = a_{00} Y_{00} + \frac{a_{1m}}{|x|} Y_{1m} + \frac{a_{2m}}{|x|^2} Y_{2m} + \dots$


3. \textbf{Take the Limit:} As $|x|$ becomes very large (goes to infinity), all the terms like $1/|x|$, $1/|x|^2$, etc., will shrink to zero. The only term left standing is the first one, $a_{00} Y_{00}$.


4. \textbf{Verify Convergence:} Since $a_{00}$ is a constant and $Y_{00}$ is a constant (the simplest spherical harmonic), the result of the limit is just a fixed number. This proves that the limit exists.


\textbf{Intuition:}
Think of $u(x)$ as the voltage created by a charged metal object $\Omega$. Far away from any object, the voltage always starts to look like the voltage of a single point charge (a "monopole"), which decreases like $1/|x|$. By multiplying by $|x|$, we are essentially "canceling out" that $1/|x|$ drop-off to see how much "charge" the object effectively has. Since the $1/|x|$ term is the strongest part of the signal far away, the product $|x|u(x)$ stabilizes to a constant value.


The structured steps and intuitive physical analogy help bridge the gap between abstract PDE theory and concrete behavior.
\end{solution}

\begin{exercise}
Let $f(x,y)\in C^1(\mathbb{R}^2)$. We assume that there exists $C>0$ so that for all $(x,y)\in\mathbb{R}^2$, $|\frac{\partial f}{\partial y}(x,y)|\leq C$. Prove that the following ODE has a globally defined solution for all $y(0)=y_0\in\mathbb{R}$:
\[
\begin{cases}
\frac{d}{dx}y(x) = f(x,y(x)),\\
y(0) = y_0.
\end{cases}
\]

In addition, we assume that $f$ is $1$-periodic in $x$, i.e., for all $(x,y)\in\mathbb{R}^2$, we have $f(x+1,y)=f(x,y)$. Prove that if the above admits a globally defined bounded solution, then it admits a periodic solution.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
Basic knowledge of Ordinary Differential Equations (ODEs), including the Picard-Lindelöf theorem for existence/uniqueness, the concept of "blow-up" in finite time, and the Poincaré map for periodic systems.


\textbf{Steps:}
\textbf{Part 1: Global Existence}
1. \textbf{Check Lipschitz Condition:} The given bound $|\partial f / \partial y| \leq C$ means the function $f$ is "globally Lipschitz" in $y$. This is a very strong condition that ensures solutions are unique and don't behave too wildly.


2. \textbf{Linear Growth:} Because the derivative is bounded, the function $f(x,y)$ can grow at most linearly with respect to $y$ (i.e., $|f(x,y)| \leq A + B|y|$). 


3. \textbf{Preventing Blow-up:} Solutions to ODEs only stop existing if they "blow up" to infinity in a finite amount of time. However, a function whose speed is limited by linear growth cannot reach infinity in finite time. Thus, the solution exists for all $x$.


\textbf{Part 2: Periodic Solution}
1. \textbf{The Poincaré Map:} Define a map $P$ that takes a starting value $y(0)$ and tells you where the solution ends up after one period, $y(1)$. Because the equation itself is periodic, if we can find a starting point where $y(0) = y(1)$, the solution will repeat itself forever.


2. \textbf{Monotonicity:} Because solutions to these types of ODEs cannot cross each other, the map $P$ is "monotonic" (if you start higher, you end higher).


3. \textbf{Convergence to a Fixed Point:} If we have a bounded solution, the sequence of points $y(0), y(1), y(2), \dots$ is a bounded, monotonic sequence. From calculus, we know every bounded monotonic sequence must converge to a limit. This limit point $y^*$ is a "fixed point" of the map $P$, meaning $P(y^*) = y^*$.


4. \textbf{Conclusion:} A fixed point of the Poincaré map corresponds exactly to a periodic solution of the original ODE.


\textbf{Intuition:}
For the first part, the bounded derivative means $f$ isn't "explosive." It's like driving a car where the acceleration can't increase too fast based on your position; you'll never hit infinite speed in five minutes. For the second part, if you have a solution that stays in a bounded box and the "rules of the road" (the ODE) repeat every mile, the solution eventually settles into a steady rhythm. It's like a swinging pendulum that, if it doesn't fly off to infinity, eventually must follow a repeating path.


The double-spacing between steps allows the student to digest the transition from the "speed limit" argument in Part 1 to the "rhythm" argument in Part 2.
\end{solution}