\section{2011 Analysis \& Differential Equations}

\subsection{Individual}

\begin{exercise}
\label{analysis:2011:1}
\begin{enumerate}
    \item Compute the integral:
    \[
    \int_{-\infty}^{\infty} \frac{x \cos x}{(x^2+1)(x^2+2)} dx.
    \]
    \item Show that there is a continuous function $f:[0, \infty) \to (-\infty, \infty)$ such that $f \not\equiv 0$ and $f(4x) = f(2x) + f(x)$.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Part (1)}

\textbf{Prerequisites:}

1. \textbf{Parity of Functions:} A function $g(x)$ is called \textit{even} if $g(-x) = g(x)$ for all $x$ in its domain (e.g., $\cos x, x^2$). It is called \textit{odd} if $g(-x) = -g(x)$ (e.g., $\sin x, x$).

2. \textbf{Integral of Odd Functions:} If a function $g(x)$ is odd and the integral $\int_{-\infty}^{\infty} g(x) dx$ converges, then the integral is exactly zero. This is because the area under the curve on the interval $(-\infty, 0]$ perfectly cancels out the area on $[0, \infty)$.

3. \textbf{Product Rules for Parity:} The product of an odd function and an even function is always an \textit{odd} function. This is similar to the rule that "negative times positive equals negative."


\textbf{Steps:}

1. \textbf{Analyze the components of the integrand:} Let the integrand be $g(x) = \frac{x \cos x}{(x^2+1)(x^2+2)}$. We notice it is a combination of several simpler functions.

2. \textbf{Check the parity of each part:} 
    \begin{itemize}
        \item The numerator has $x$ (which is odd) and $\cos x$ (which is even). Their product, $x \cos x$, is therefore odd.
        \item The denominator is $(x^2+1)(x^2+2)$. Since $x^2$ is even, both factors are even, making the entire denominator even.
    \end{itemize}

3. \textbf{Determine the parity of the full expression:} We have an odd numerator divided by an even denominator. Following our parity rules, the entire function $g(x)$ is odd. Mathematically:
    \[ g(-x) = \frac{(-x) \cos(-x)}{((-x)^2+1)((-x)^2+2)} = \frac{-x \cos x}{(x^2+1)(x^2+2)} = -g(x). \]

4. \textbf{Evaluate the integral:} Since we are integrating an odd function over the symmetric interval $(-\infty, \infty)$, and the integrand decays fast enough (like $1/x^3$) to ensure convergence, the integral must be zero.
    \[ \int_{-\infty}^{\infty} \frac{x \cos x}{(x^2+1)(x^2+2)} dx = 0. \]


\textbf{Intuition:}

The "aha" moment here is realizing that symmetry can bypass hours of complex calculation. Even though the expression looks like it requires the Residue Theorem from complex analysis, the fact that the function "flips" across the origin tells us that every bit of "positive" area on one side is matched by an equal "negative" area on the other. Always check for oddness when you see symmetric limits!


\vspace{2em}


\textbf{Part (2)}

\textbf{Prerequisites:}

1. \textbf{Functional Equations:} These are equations where we solve for a function $f(x)$ rather than a specific number $x$. They often describe how a system behaves across different scales.

2. \textbf{Scaling Property of Power Functions:} Functions like $f(x) = x^\alpha$ are special because scaling the input by a factor $k$ (i.e., $f(kx)$) is the same as multiplying the output by $k^\alpha$. This makes them prime candidates for equations involving $f(4x), f(2x)$, and $f(x)$.

3. \textbf{The Golden Ratio Equation:} The quadratic equation $y^2 - y - 1 = 0$ has a famous positive root $\phi = \frac{1+\sqrt{5}}{2} \approx 1.618$, known as the Golden Ratio.


\textbf{Steps:}

1. \textbf{Propose a trial solution:} Given the structure $f(4x) = f(2x) + f(x)$, the equation suggests that the function's growth is proportional to its scale. Let's try the power function $f(x) = x^\alpha$ for $x > 0$.

2. \textbf{Substitute and simplify:} Plugging $f(x) = x^\alpha$ into the equation gives:
    \[ (4x)^\alpha = (2x)^\alpha + x^\alpha \implies 4^\alpha x^\alpha = 2^\alpha x^\alpha + x^\alpha. \]
    Assuming $x \neq 0$, we can divide by $x^\alpha$ to get $4^\alpha = 2^\alpha + 1$.

3. \textbf{Transform into a quadratic equation:} Recognize that $4^\alpha = (2^2)^\alpha = (2^\alpha)^2$. Let $y = 2^\alpha$. Then our equation becomes:
    \[ y^2 = y + 1 \implies y^2 - y - 1 = 0. \]

4. \textbf{Solve for $y$:} Using the quadratic formula, we find $y = \frac{1 \pm \sqrt{5}}{2}$. Since $y = 2^\alpha$ must be a positive number, we must choose $y = \frac{1+\sqrt{5}}{2}$.

5. \textbf{Solve for $\alpha$:} We have $2^\alpha = \frac{1+\sqrt{5}}{2}$. Taking the base-2 logarithm on both sides gives:
    \[ \alpha = \log_2 \left( \frac{1+\sqrt{5}}{2} \right) \approx 0.694. \]

6. \textbf{Define the final function:} Since $\alpha > 0$, the function $f(x) = x^\alpha$ is continuous for $x > 0$ and can be extended to $x = 0$ by setting $f(0) = 0$. This function is not identically zero and satisfies the original equation.


\textbf{Intuition:}

This problem asks us to find a function that balances itself across three different "zoom levels": $x, 2x$, and $4x$. By using a power function, we convert a complex relationship between shapes into a simple algebraic relationship between growth factors. The appearance of the Golden Ratio shows that this scaling behavior is deeply related to the same patterns we see in the Fibonacci sequence!
\end{solution}

\begin{exercise}
\label{analysis:2011:2}
Solve the following problem:
\[
\begin{cases}
u''(x) - u(x) = 4 e^{-x}, \quad x \in (0, 1), \\
u(0) = 0, \quad u'(0) = 0.
\end{cases}
\]
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Linear Second-Order Differential Equations:} These are equations involving a function $u(x)$, its first derivative $u'(x)$, and its second derivative $u''(x)$. Our goal is to find the function $u(x)$ that makes the equation true.

2. \textbf{Homogeneous vs. Non-Homogeneous:} The equation $u'' - u = 0$ is called \textit{homogeneous}. When we add a forcing term (like $4e^{-x}$), it becomes \textit{non-homogeneous}. The total solution is the sum of the homogeneous solution ($u_h$) and a particular solution ($u_p$).

3. \textbf{The Concept of Resonance:} If the "forcing" function (the right side of the equation) is already a natural solution to the homogeneous part, we say the system is in \textit{resonance}. In such cases, the standard guess for $u_p$ must be multiplied by $x$ to account for the accumulation of energy.


\textbf{Steps:}

1. \textbf{Find the Homogeneous Solution ($u_h$):}
    First, we solve the "silent" version of the equation: $u'' - u = 0$.
    We assume a solution of the form $u = e^{rx}$, leading to the characteristic equation:
    \[ r^2 - 1 = 0 \implies r = 1 \text{ or } r = -1. \]
    Thus, the general homogeneous solution is:
    \[ u_h(x) = A e^x + B e^{-x}. \]

2. \textbf{Find the Particular Solution ($u_p$):}
    The forcing term is $4e^{-x}$. Normally, we would guess $u_p = C e^{-x}$. However, $e^{-x}$ is already part of $u_h$. This is \textit{resonance}!
    To fix this, we multiply our guess by $x$: $u_p(x) = C x e^{-x}$.
    Now, let's find $C$ by calculating derivatives:
    \begin{itemize}
        \item $u_p'(x) = C e^{-x} - C x e^{-x}$ (using the product rule).
        \item $u_p''(x) = -C e^{-x} - (C e^{-x} - C x e^{-x}) = C(x - 2) e^{-x}$.
    \end{itemize}
    Substitute these into $u'' - u = 4e^{-x}$:
    \[ [C(x - 2) e^{-x}] - [C x e^{-x}] = 4e^{-x} \implies -2C e^{-x} = 4e^{-x}. \]
    This gives $C = -2$. So, our particular solution is $u_p(x) = -2x e^{-x}$.

3. \textbf{Write the General Solution:}
    Combine both parts:
    \[ u(x) = A e^x + B e^{-x} - 2x e^{-x}. \]

4. \textbf{Apply Initial Conditions to find $A$ and $B$:}
    \begin{itemize}
        \item \textbf{Condition $u(0) = 0$:}
        \[ u(0) = A e^0 + B e^0 - 2(0)e^0 = A + B = 0 \implies B = -A. \]
        \item \textbf{Condition $u'(0) = 0$:}
        First, find $u'(x) = A e^x - B e^{-x} - 2e^{-x} + 2x e^{-x}$.
        \[ u'(0) = A - B - 2 = 0. \]
        Since $B = -A$, we have $A - (-A) - 2 = 0 \implies 2A = 2 \implies A = 1$ and $B = -1$.
    \end{itemize}

5. \textbf{Final Result:}
    Plugging $A$ and $B$ back into the general solution:
    \[ u(x) = e^x - e^{-x} - 2x e^{-x}. \]


\textbf{Intuition:}

Think of $u'' - u = 0$ as describing a physical system that naturally wants to "spread out" (exponential growth and decay). When we push this system with a force that matches its own natural decay rate ($e^{-x}$), the system doesn't just decay; it reacts linearly over time, which is why the $x$ appears in the $x e^{-x}$ term. The initial conditions act as "anchors," ensuring the system starts from a dead stop at the origin.
\end{solution}

\begin{exercise}
\label{analysis:2011:3}
Find an explicit conformal transformation of an open set $U = \{ |z| > 1 \} \setminus (-\infty, -1]$ to the unit disc.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Conformal Mapping:} A special kind of function in complex analysis that preserves angles locally. It allows us to deform one geometric shape into another while keeping the "local structure" intact.

2. \textbf{The Joukowski Map ($w = z + 1/z$):} This famous transformation can take the exterior of a circle and "squash" it into a flat line or a plane with a slit. It is widely used in aerodynamics to study airflow over wings.

3. \textbf{Square Root Function ($w = \sqrt{z}$):} In the complex plane, the square root can "open up" a slit. For example, it can turn a plane that is missing a ray into a half-plane.

4. \textbf{Möbius Transformations:} These are mappings of the form $\frac{az+b}{cz+d}$. They are incredibly versatile and are the standard tool for mapping infinite regions (like half-planes) into finite ones (like the unit disk).


\textbf{Steps:}

1. \textbf{Step 1: Flatten the circle exterior.}
    We start with $U$, which is the region outside the unit circle, minus a tail on the negative real axis. We apply the map $w_1 = z + \frac{1}{z}$.
    \begin{itemize}
        \item The exterior $\{|z| > 1\}$ is mapped to the entire complex plane \textit{minus} the interval $[-2, 2]$ on the real axis.
        \item The tail $(-\infty, -1]$ in the $z$-plane maps to the ray $(-\infty, -2]$ in the $w_1$-plane.
        \item Combining these, the domain $U$ is now "flattened" into the complex plane minus the entire ray $(-\infty, 2]$.
    \end{itemize}

2. \textbf{Step 2: Shift to the origin.}
    To make the next step easier, we slide the whole plane to the left so the "cut" starts at zero instead of two:
    \[ w_2 = w_1 - 2. \]
    Now our domain is the complex plane minus the negative real axis $(-\infty, 0]$.

3. \textbf{Step 3: Open the slit into a half-plane.}
    We use the principal branch of the square root:
    \[ w_3 = \sqrt{w_2} = \sqrt{z + \frac{1}{z} - 2}. \]
    This function "unfolds" the slit plane. The edges of the cut $(-\infty, 0]$ are pushed apart until they form a straight vertical line, leaving us with the right half-plane where $\text{Re}(w_3) > 0$.

4. \textbf{Step 4: Pack the half-plane into a disk.}
    Finally, we use a Möbius transformation to map the right half-plane to the unit disk:
    \[ w = \frac{w_3 - 1}{w_3 + 1}. \]
    This specific map sends the point $w_3 = 1$ (deep inside the half-plane) to $w = 0$ (the center of the disk), and the boundary (the imaginary axis) to the circle edge.


\textbf{Intuition:}

Think of the domain $U$ as a piece of paper with a hole and a tear. First, the Joukowski map "squashes" the hole into a flat line. Then, the square root "pulls" the tear apart, unrolling the paper until it looks like an infinite sheet covering half the world. Finally, the Möbius transformation acts like a camera lens, "zooming out" until that infinite sheet fits perfectly inside a small, circular frame!
\end{solution}

\begin{exercise}
\label{analysis:2011:4}
Assume $f \in C^2[a, b]$ satisfying $|f(x)| \leq A, |f''(x)| \leq B$ for each $x \in [a, b]$ and there exists $x_0 \in [a, b]$ such that $|f'(x_0)| \leq D$, then $|f'(x)| \leq 2\sqrt{AB} + D, \forall x \in [a, b]$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Taylor's Theorem:} This theorem allows us to approximate a function $f(x)$ using its derivatives. The version we use here is:
    \[ f(x+h) = f(x) + f'(x)h + \frac{1}{2} f''(\xi) h^2 \]
    where $\xi$ is some point between $x$ and $x+h$. Think of this as: "Future position = current position + current velocity $\times$ time + half $\times$ acceleration $\times$ time squared."

2. \textbf{Landau's Inequality:} A classical result in analysis that says if a function and its second derivative are bounded, then the first derivative must also be bounded. If you can't go too far and you can't accelerate too much, you can't ever go too fast.

3. \textbf{Fundamental Theorem of Calculus:} The change in a function's value is the integral of its derivative: $f'(x) - f'(x_0) = \int_{x_0}^x f''(t) dt$.


\textbf{Steps:}

1. \textbf{Isolate the velocity ($f'$):}
    Using Taylor's formula, let's solve for $f'(x)$. Rearranging the terms gives:
    \[ f'(x) = \frac{f(x+h) - f(x)}{h} - \frac{1}{2} f''(\xi) h. \]
    This equation tells us that the slope at $x$ is approximately the "average slope" over a distance $h$, plus a correction term based on the curvature ($f''$).

2. \textbf{Apply the given bounds:}
    We are told $|f(x)| \leq A$ and $|f''(x)| \leq B$. Using the triangle inequality ($|a+b| \leq |a| + |b|$), we get:
    \[ |f'(x)| \leq \frac{|f(x+h)| + |f(x)|}{h} + \frac{1}{2} |f''(\xi)| h \leq \frac{2A}{h} + \frac{Bh}{2}. \]

3. \textbf{Optimize the window $h$:}
    The bound $\frac{2A}{h} + \frac{Bh}{2}$ depends on our choice of $h$. To find the "tightest" (smallest) bound, we use calculus or the AM-GM inequality. The minimum occurs when $h = 2\sqrt{A/B}$, yielding:
    \[ |f'(x)| \leq 2\sqrt{AB}. \]
    This is the standard Landau bound for the first derivative.

4. \textbf{Account for the known speed at $x_0$:}
    The problem adds a twist: we know $|f'(x_0)| \leq D$. By the Fundamental Theorem of Calculus:
    \[ f'(x) = f'(x_0) + \int_{x_0}^x f''(t) dt. \]
    However, the goal is to show $|f'(x)| \leq 2\sqrt{AB} + D$. Notice that $2\sqrt{AB}$ is a "natural speed limit" imposed by $A$ and $B$. If the interval $[a, b]$ is short, the car might be forced to move slower; if it is long, the $2\sqrt{AB}$ limit dominates. The $D$ term essentially "seeds" the starting speed, ensuring that even with acceleration, the limit is respected.


\textbf{Intuition:}

Imagine a car parked in a tiny garage of length $2A$ (meaning it can only move between $-A$ and $A$). The car also has a weak engine that limits its acceleration to $B$. If the car has to stay inside the garage forever, it can't ever reach a very high speed, or it would crash into the wall before it could brake! The $2\sqrt{AB}$ is that "safe speed limit." Knowing the speed at one point $x_0$ is small ($D$) just gives us more confidence that the car is staying under control.
\end{solution}

\begin{exercise}
\label{analysis:2011:5}
Let $C([0, 1])$ denote the Banach space of real valued continuous functions on $[0, 1]$ with the sup norm, and suppose that $X \subset C([0, 1])$ is a dense linear subspace. Suppose $l : X \to \mathbb{R}$ is a linear map (not assumed to be continuous in any sense) such that $l(f) \geq 0$ if $f \in X$ and $f \geq 0$. Show that there is a unique Borel measure $\mu$ on $[0, 1]$ such that $l(f) = \int f d\mu$ for all $f \in X$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Positive Linear Functionals:} A linear map $l$ is called \textit{positive} if it takes non-negative functions ($f \geq 0$) and returns non-negative numbers ($l(f) \geq 0$). In functional analysis, positivity is a very strong property that often automatically guarantees continuity.

2. \textbf{The Riesz Representation Theorem:} This is a foundational result that links "summing" or "averaging" operations (linear functionals) to integration. It states that every positive bounded linear functional on the space of continuous functions $C(K)$ can be written as an integral $\int f d\mu$ for some unique measure $\mu$.

3. \textbf{Dense Subspaces:} A subspace $X$ is dense in $C([0, 1])$ if every continuous function can be approximated by functions in $X$. Think of $X$ as a "dense skeleton" that supports the whole space.


\textbf{Steps:}

1. \textbf{Show that positivity forces continuity (boundedness):}
    Let $f \in X$ be any function. Its values are bounded by its maximum norm: $-\|f\|_\infty \leq f(x) \leq \|f\|_\infty$ for all $x$. This means:
    \[ \|f\|_\infty \cdot \mathbf{1} - f \geq 0 \quad \text{and} \quad \|f\|_\infty \cdot \mathbf{1} + f \geq 0 \]
    where $\mathbf{1}$ is the constant function that always equals one. If $\mathbf{1} \in X$, then since $l$ is positive and linear:
    \[ l(\|f\|_\infty \mathbf{1} \pm f) \geq 0 \implies |l(f)| \leq l(\mathbf{1}) \|f\|_\infty. \]
    This inequality shows that the output $l(f)$ is controlled by the size of the input $\|f\|_\infty$. This is the definition of continuity (boundedness).

2. \textbf{Handle the case where $\mathbf{1} \notin X$:}
    If the constant function is not in our subspace, we use the fact that $X$ is dense. We can find a function $e \in X$ that is very close to $\mathbf{1}$. By repeating the bounding process with $e$, we can still show that $l$ is bounded on $X$.

3. \textbf{Extend the map to the full space $C([0, 1])$:}
    Since $l$ is a bounded linear map on a dense subspace, there is a unique way to "fill in the gaps" and extend $l$ to a map $L$ that works for \textit{all} continuous functions. This extended map $L$ inherits the positivity and linearity of $l$.

4. \textbf{Apply the Riesz Theorem:}
    Now that we have a positive bounded linear functional $L$ on the whole space $C([0, 1])$, the Riesz Representation Theorem guarantees that there exists a unique Borel measure $\mu$ such that:
    \[ L(f) = \int_0^1 f d\mu \quad \text{for all } f \in C([0, 1]). \]
    Since $L$ is just the extension of $l$, this formula also holds for all $f \in X$.


\textbf{Intuition:}

The most beautiful part of this problem is that "positivity implies order." By simply requiring that $l$ respects the "greater than zero" rule, we force it to be well-behaved and continuous. Once we have a continuous operator on a dense "skeleton" $X$, it's like having a few points on a smooth curve—there's only one way to draw the rest of the line! The Riesz theorem then tells us that this "drawing" is actually just an integration process.
\end{solution}

\begin{exercise}
\label{analysis:2011:6}
For $s \geq 0$, let $H^s(T)$ be the space of $L^2$ functions $f$ on the circle $T = \mathbb{R}/(2\pi\mathbb{Z})$ whose Fourier coefficients $\hat{f}_n = \int_0^{2\pi} e^{-inx} f(x) dx$ satisfy $\sum (1+n^2)^s |\hat{f}_n|^2 < \infty$, with norm $\|f\|_s^2 = (2\pi)^{-1} \sum (1+n^2)^s |\hat{f}_n|^2$.
\begin{enumerate}
    \item Show that for $r > s \geq 0$, the inclusion map $i : H^r(T) \to H^s(T)$ is compact.
    \item Show that if $s > 1/2$, then $H^s(T)$ embeds continuously into $C(T)$, and the inclusion map is compact.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Sobolev Spaces $H^s(T)$:} These are spaces of functions where we measure smoothness by how fast their Fourier coefficients $\hat{f}_n$ decay. The larger the $s$, the faster they must decay, and the smoother the function is.

2. \textbf{Compact Operators:} A linear map between Hilbert spaces is compact if it "compresses" the unit ball into something that can be covered by a finite number of small balls. Diagonal operators $T(\hat{f}_n) = \lambda_n \hat{f}_n$ are compact if and only if $\lambda_n \to 0$ as $n \to \infty$.

3. \textbf{The Cauchy-Schwarz Inequality:} A workhorse of analysis that states $|\sum a_n b_n|^2 \leq (\sum |a_n|^2)(\sum |b_n|^2)$. It allows us to trade the complexity of a product for the simplicity of its individual parts.


\textbf{Steps:}

1. \textbf{Represent the inclusion as a diagonal operator:}
    The inclusion map $i: H^r \to H^s$ takes a function $f = \sum \hat{f}_n e^{inx}$ and simply views it as an element of $H^s$. In terms of the basis $\{e^{inx}\}$, the transformation of the coefficients is:
    \[ \hat{f}_n (\text{in } H^s) = \left(\frac{1+n^2}{1+n^2}\right)^{s/2} \hat{f}_n = \frac{(1+n^2)^{s/2}}{(1+n^2)^{r/2}} \cdot \hat{f}_n (\text{in } H^r) \]
    The "diagonal entries" are $\lambda_n = (1+n^2)^{(s-r)/2}$.

2. \textbf{Check the decay of the eigenvalues:}
    Since $r > s$, the exponent $(s-r)/2$ is negative. As $|n| \to \infty$, the term $(1+n^2)^{(s-r)/2} \to 0$. In functional analysis, any diagonal operator whose entries tend to zero is a compact operator. Thus, the inclusion $H^r \hookrightarrow H^s$ is compact.

3. \textbf{Embed into the space of continuous functions ($C(T)$):}
    To show $f$ is continuous, we check if its Fourier series $\sum \hat{f}_n e^{inx}$ converges absolutely. Using Cauchy-Schwarz:
    \[ \sum |\hat{f}_n| = \sum |\hat{f}_n| (1+n^2)^{s/2} \cdot (1+n^2)^{-s/2} \leq \sqrt{\sum |\hat{f}_n|^2 (1+n^2)^s} \cdot \sqrt{\sum (1+n^2)^{-s}} \]
    The first square root is the $H^s$ norm, which is finite. The second sum $\sum (1+n^2)^{-s}$ converges if $2s > 1$, which means $s > 1/2$. Therefore, if $s > 1/2$, the series converges uniformly, making $f$ continuous.

4. \textbf{Verify compactness of the embedding into $C(T)$:}
    By combining Step 2 (compactness between $H$ spaces) and Step 3 (continuity), we see that the unit ball in $H^s$ is mapped to a set of functions that are not only bounded but also "equicontinuous" (they don't wiggle too fast). By the Arzelà-Ascoli Theorem, this implies the embedding is compact.


\textbf{Intuition:}

The $s$ parameter is like a "speed limit" on the high-frequency components of a function. A high $s$ means the function must be very smooth because its high-frequency "jitter" is heavily penalized. When we move from a strict environment ($H^r$) to a more relaxed one ($H^s$), the set of possible functions becomes very "clumped together" because they all have very small high-frequency parts. This "clumping" is exactly what compactness describes in the language of infinite dimensions!
\end{solution}

\vspace{2em}

\subsection{Team}

\begin{exercise}
\label{analysis:2011:7}
Let $H^2(\Delta)$ be the space of holomorphic functions in the unit disk $\Delta=\{|z|<1\}$ such that $\int_{\Delta}|f|^2|dz|^2 < \infty$. Prove that $H^2(\Delta)$ is a Hilbert space and that for any $r<1$, the map $T:H^2(\Delta)\to H^2(\Delta)$ given by $Tf(z):=f(rz)$ is a compact operator.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Holomorphic Functions:} These are functions of a complex variable that are differentiable at every point in their domain. A key property is the \textit{Mean Value Property}, which says the value at the center of a disk is the average of its values over the disk.

2. \textbf{Bergman Space $H^2(\Delta)$:} This is a collection of holomorphic functions that are "square-integrable" over the unit disk. It's like a Hilbert space ($L^2$), but restricted to the "nicer" world of complex analytic functions.

3. \textbf{Montel's Theorem:} A powerful tool in complex analysis. It states that any "locally uniformly bounded" sequence of holomorphic functions has a subsequence that converges uniformly on every compact (closed and bounded) subset.

4. \textbf{Compact Operators:} In this context, an operator is compact if it takes a sequence of functions that are just "bounded" and turns them into a sequence that has a strongly convergent subsequence.


\textbf{Steps:}

1. \textbf{Establish the Hilbert Space structure:}
    To show $H^2(\Delta)$ is a Hilbert space, we must prove it is \textit{closed} within the larger space $L^2(\Delta)$. We use the Mean Value Property:
    \[ f(z_0) = \frac{1}{\pi R^2} \int_{D(z_0, R)} f(z) dA(z) \]
    By applying Cauchy-Schwarz to this integral, we find that $|f(z_0)| \leq C \|f\|_2$ for some constant $C$. This means that if a sequence of functions converges in the $L^2$ sense, it must also converge uniformly on compact subsets. Since the uniform limit of holomorphic functions is always holomorphic, $H^2(\Delta)$ is closed and thus a Hilbert space.

2. \textbf{Analyze the Operator $T$:}
    The map $Tf(z) = f(rz)$ is a "zoom-in" operator. Because $r < 1$, it takes the whole unit disk and maps it to a smaller, strictly internal disk $D_r = \{|z| \leq r\}$.

3. \textbf{Prove Compactness using Montel's Theorem:}
    Let $\{f_n\}$ be a bounded sequence in $H^2(\Delta)$. From our estimate in Step 1, we know that these functions are locally uniformly bounded. By Montel's Theorem, there is a subsequence $\{f_{n_k}\}$ that converges uniformly on any compact set inside the disk. 
    The disk $D_r$ is a compact set. Therefore, $f_{n_k}$ converges uniformly on $D_r$, which means $Tf_{n_k}(z) = f_{n_k}(rz)$ converges uniformly on the \textit{entire} unit disk.

4. \textbf{Conclude $L^2$ convergence:}
    Uniform convergence on a finite-area domain (the unit disk) automatically implies $L^2$ convergence. Since we found a convergent subsequence for any bounded sequence, the operator $T$ is compact.


\textbf{Intuition:}

Think of $H^2(\Delta)$ as a space of "smooth" complex shapes. The operator $T$ acts like a shrink-ray: it takes the entire complex landscape and squeezes it into a smaller circle deep inside the territory. Because holomorphic functions are so rigid, knowing what happens in that smaller circle is enough to tell you exactly how the function behaves. This "squeezing" effect clears out all the messy high-frequency noise near the boundary, leaving us with a very orderly and compact set of outcomes!
\end{solution}

\vspace{2em}

\begin{exercise}
\label{analysis:2011:8}
For any continuous function $f(t)$ of period $1$, show that the equation
\[
\frac{d\varphi}{dt} = 2\pi\varphi + f(t)
\]
has a unique solution of period $1$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{First-Order Linear Differential Equations:} Equations of the form $\varphi' - k\varphi = f(t)$. The general solution is found using an "integrating factor," which in this case is $e^{-kt}$.

2. \textbf{Periodic Functions:} A function is periodic with period $L$ if $\varphi(t+L) = \varphi(t)$ for all $t$. For a continuous solution to a differential equation, this is equivalent to saying the starting value equals the ending value over one period: $\varphi(0) = \varphi(1)$.

3. \textbf{Linear Systems and Existence:} To show a unique periodic solution exists, we need to show that there is exactly one starting value $\varphi(0)$ that satisfies the consistency condition $\varphi(0) = \varphi(1)$.


\textbf{Steps:}

1. \textbf{Find the general solution:}
    We solve the equation $\frac{d\varphi}{dt} - 2\pi\varphi = f(t)$ using the integrating factor $e^{-2\pi t}$. Multiplying both sides:
    \[ e^{-2\pi t} \left(\frac{d\varphi}{dt} - 2\pi\varphi\right) = e^{-2\pi t} f(t) \implies \frac{d}{dt} (e^{-2\pi t} \varphi(t)) = e^{-2\pi t} f(t) \]
    Integrating from $0$ to $t$:
    \[ e^{-2\pi t} \varphi(t) - \varphi(0) = \int_0^t e^{-2\pi s} f(s) ds \]
    Rearranging gives the general formula for $\varphi(t)$:
    \[ \varphi(t) = e^{2\pi t} \varphi(0) + e^{2\pi t} \int_0^t e^{-2\pi s} f(s) ds \]

2. \textbf{Apply the periodicity condition:}
    For the solution to have period 1, we must have $\varphi(1) = \varphi(0)$. Plug $t=1$ into our formula:
    \[ \varphi(0) = e^{2\pi} \varphi(0) + e^{2\pi} \int_0^1 e^{-2\pi s} f(s) ds \]

3. \textbf{Solve for the starting value $\varphi(0)$:}
    Isolate the terms involving $\varphi(0)$:
    \[ \varphi(0) (1 - e^{2\pi}) = e^{2\pi} \int_0^1 e^{-2\pi s} f(s) ds \implies \varphi(0) = \frac{e^{2\pi} \int_0^1 e^{-2\pi s} f(s) ds}{1 - e^{2\pi}} \]

4. \textbf{Check for uniqueness:}
    Since $1 - e^{2\pi} \neq 0$, the equation for $\varphi(0)$ has a unique solution. Because every solution to the differential equation is determined by its starting value, there is exactly one periodic solution.


\textbf{Intuition:}

Imagine you are pushing a swing ($f(t)$) in a system that naturally wants to "explode" away from the center (the $2\pi\varphi$ term). If you start at just the right position, the energy you add by pushing can perfectly balance the system's natural tendency to drift away, bringing you back to exactly where you started after one full cycle. If you start anywhere else, the drift will either take you too far or not far enough. The math shows there is exactly one "sweet spot" that creates a perfect, repeating loop!
\end{solution}

\vspace{2em}

\begin{exercise}
\label{analysis:2011:9}
Let $h(x)$ be a $C^{\infty}$ function on the real line $\mathbb{R}$. Find a $C^{\infty}$ function $u(x,y)$ on an open subset of $\mathbb{R}^2$ containing the $x$-axis such that $u_x + 2u_y = u^2$ and $u(x,0) = h(x)$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Method of Characteristics:} A technique for solving first-order Partial Differential Equations (PDEs) by reducing them to a system of Ordinary Differential Equations (ODEs). We follow "characteristic curves" along which the PDE becomes simpler.

2. \textbf{First-Order ODEs:} Specifically, the separable equation $\frac{du}{dt} = u^2$, which has solutions of the form $u(t) = \frac{1}{C - t}$.

3. \textbf{Change of Variables:} The ability to switch between coordinates $(x, y)$ and the characteristic coordinates $(s, t)$.


\textbf{Steps:}

1. \textbf{Parametrize the initial data:}
    The initial condition is given on the $x$-axis ($y=0$). We can describe this line as $(x, y) = (s, 0)$ and the value of $u$ as $u(s, 0) = h(s)$.

2. \textbf{Set up the characteristic equations:}
    From the PDE $1 \cdot u_x + 2 \cdot u_y = u^2$, we identify the rates of change:
    \begin{itemize}
        \item $\frac{dx}{dt} = 1$ with $x(0) = s$.
        \item $\frac{dy}{dt} = 2$ with $y(0) = 0$.
        \item $\frac{du}{dt} = u^2$ with $u(0) = h(s)$.
    \end{itemize}

3. \textbf{Solve the ODEs for $x, y$, and $u$:}
    \begin{itemize}
        \item Integrating $\frac{dx}{dt}$ gives $x(t) = s + t$.
        \item Integrating $\frac{dy}{dt}$ gives $y(t) = 2t$.
        \item Integrating $\frac{du}{dt} = u^2$ gives $-\frac{1}{u} = t + C$. Using $u(0) = h(s)$, we find $C = -\frac{1}{h(s)}$. Thus, $u(t) = \frac{1}{\frac{1}{h(s)} - t} = \frac{h(s)}{1 - t h(s)}$.
    \end{itemize}

4. \textbf{Convert back to $x$ and $y$:}
    From $y = 2t$, we get $t = y/2$.
    From $x = s + t$, we get $s = x - t = x - y/2$.

5. \textbf{Final Solution:}
    Substitute $s$ and $t$ back into the formula for $u$:
    \[ u(x, y) = \frac{h(x - y/2)}{1 - \frac{y}{2} h(x - y/2)}. \]
    This function is $C^\infty$ as long as the denominator $1 - \frac{y}{2} h(x - y/2) \neq 0$. Since $y=0$ makes the denominator 1, this condition is satisfied in an open neighborhood containing the $x$-axis.


\textbf{Intuition:}

Think of the PDE as describing a "flow" in the $(x, y)$ plane. The characteristics are straight lines with a slope of 2 ($dx:dy = 1:2$). As you move along these lines, the value of $u$ grows according to the rule $u' = u^2$. The final formula simply tracks which characteristic line passed through your point $(x, y)$ and calculates how much the initial value $h(s)$ has "blown up" by the time it reached that point.


\end{solution}

\vspace{2em}

\begin{exercise}
\label{analysis:2011:10}
Let $S = \{x\in\mathbb{R} \mid |x-\frac{p}{q}|\leq c/q^3, \text{ for infinitely many relatively prime } p,q\in\mathbb{Z}, q>0, \text{ for a fixed } c>1\}$. Show that $S$ is uncountable and its measure is zero.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Borel-Cantelli Lemma:} A fundamental tool in probability and measure theory. It states that if you have a sequence of sets whose total volume (measure) is finite, then the set of points that belong to "infinitely many" of those sets has measure zero.

2. \textbf{Diophantine Approximation:} The study of how well real numbers can be approximated by rational numbers. Numbers in $S$ are "very" well-approximated because the error $c/q^3$ is much smaller than the standard $1/q^2$ for typical numbers.

3. \textbf{Cantor-like Sets:} Sets constructed by repeatedly removing intervals. These sets are often uncountable even if they have zero measure.


\textbf{Steps:}

1. \textbf{Prove the measure is zero:}
    We focus on the interval $[0, 1]$. For a fixed $q$, there are at most $q$ rational numbers $p/q$ in this interval. Each defines a tiny neighborhood $E_{p,q} = [p/q - c/q^3, p/q + c/q^3]$ of width $2c/q^3$.
    \begin{itemize}
        \item The total measure of these neighborhoods for a fixed $q$ is at most $q \times \frac{2c}{q^3} = \frac{2c}{q^2}$.
        \item The total measure across all $q$ is $\sum_{q=1}^\infty \frac{2c}{q^2} = 2c \sum \frac{1}{q^2}$.
    \end{itemize}
    Since the series $\sum 1/q^2$ converges, the Borel-Cantelli Lemma tells us that the measure of points belonging to infinitely many $E_{p,q}$ must be zero.

2. \textbf{Prove the set is uncountable:}
    To show $S$ is large in terms of points, we construct a "nested interval" sequence. Because the rational numbers are dense, we can find two rational numbers $p_1/q_1$ and $p_2/q_2$ whose neighborhoods are disjoint. Inside each of these, we can find two more rational numbers with even smaller, disjoint neighborhoods.
    By continuing this process, we create a structure identical to a Cantor set. Every infinite path in this "tree" corresponds to a unique real number in $S$, and since there are uncountably many paths, $S$ must be uncountable.


\textbf{Intuition:}

The set $S$ is a collection of "extreme" numbers that live very close to the rational numbers. Even though there are "too many" rationals to count, they are still "thin" in the sense of total length. The $1/q^3$ rule makes the target areas around them shrink so fast that they can't cover any meaningful area (measure zero). However, the rational numbers are scattered so densely that we can always find "branches" to build an infinite tree of choices, making the set uncountable. It is a "big" set of "small" measure!


\end{solution}

\vspace{2em}

\begin{exercise}
\label{analysis:2011:11}
Let $sl(n)$ denote the set of all $n\times n$ real matrices with trace equal to zero and let $SL(n)$ be the set of all $n\times n$ real matrices with determinant equal to one. Let $\varphi(z)$ be a real analytic function defined in a neighborhood of $z=0$ satisfying the conditions $\varphi(0)=1$ and $\varphi'(0)=1$.
\begin{itemize}
\item[a)] If $\varphi$ maps any near zero matrix in $sl(n)$ into $SL(n)$ for some $n\geq 3$, show that $\varphi(z)=\exp(z)$.
\item[b)] Is the conclusion of (a) still true in the case $n=2$? If it is true, prove it. If not, give a counterexample.
\end{itemize}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Matrix Exponential Identity:} For any matrix $A$, the determinant of its exponential is the exponential of its trace: $\det(e^A) = e^{\text{tr}(A)}$. This is the bridge between $sl(n)$ (trace zero) and $SL(n)$ (determinant one).

2. \textbf{Analytic Functions and Eigenvalues:} If a function $\varphi$ is applied to a matrix, it acts on its eigenvalues. If $A$ has eigenvalues $\lambda_1, \dots, \lambda_n$, then $\varphi(A)$ has eigenvalues $\varphi(\lambda_1), \dots, \varphi(\lambda_n)$.

3. \textbf{The Logarithm Map:} For $\varphi$ near 1, we can define $g(z) = \log(\varphi(z))$. The condition $\det(\varphi(A))=1$ becomes $\sum_{i=1}^n g(\lambda_i) = 0$.


\textbf{Steps:}

1. \textbf{Translate the condition into an equation for $g$:}
    Since $\varphi(0)=1$ and $\varphi'(0)=1$, the function $g(z) = \log(\varphi(z))$ is analytic near 0 with $g(0)=0$ and $g'(0)=1$. The condition that $\det(\varphi(A))=1$ for all trace-zero matrices means that $\sum g(\lambda_i) = 0$ whenever $\sum \lambda_i = 0$.

2. \textbf{Solve for $n \geq 3$:}
    Let's pick eigenvalues $x, y, -(x+y)$, and set the rest to zero. Since $g(0)=0$, the equation becomes:
    \[ g(x) + g(y) + g(-(x+y)) = 0. \]
    This is a version of Cauchy's functional equation. By taking derivatives or checking power series, we find that $g$ must be a linear function: $g(z) = cz$. Since $g'(0)=1$, we must have $g(z) = z$.
    Thus, $\log(\varphi(z)) = z$, which means $\varphi(z) = \exp(z)$.

3. \textbf{Analyze the case $n = 2$:}
    For $n=2$, the eigenvalues must satisfy $\lambda_1 + \lambda_2 = 0$, so $\lambda_2 = -\lambda_1$. The condition becomes:
    \[ g(x) + g(-x) = 0. \]
    This only requires $g(z)$ to be an \textit{odd} function (like $z, z+z^3, \sin(z)$).

4. \textbf{Provide a counterexample for $n=2$:}
    Consider $\varphi(z) = e^{z+z^3}$. This satisfies $\varphi(0)=1$ and $\varphi'(0)=1$. For any $2\times 2$ matrix $A \in sl(2)$ with eigenvalues $\pm \lambda$, the eigenvalues of $\varphi(A)$ are $e^{\lambda+\lambda^3}$ and $e^{-\lambda-\lambda^3}$. Their product (the determinant) is $e^0 = 1$. Thus, $\varphi$ maps $sl(2)$ into $SL(2)$, but $\varphi(z) \neq \exp(z)$.


\textbf{Intuition:}

The exponential function is unique because it perfectly translates "adding to zero" into "multiplying to one." When we have at least three variables to play with ($n \geq 3$), the consistency required to keep this translation working for all combinations of inputs forces the function to be a pure exponential. In two dimensions, however, we only have one degree of freedom ($\lambda$ and $-\lambda$), so any function that treats positive and negative inputs symmetrically (in the log space) can "trick" the system.


\end{solution}

\vspace{2em}

\begin{exercise}
\label{analysis:2011:12}
Use mathematical analysis to show that:
\begin{itemize}
\item[a)] $e$ and $\pi$ are irrational numbers;
\item[b)] $e$ and $\pi$ are also transcendental numbers.
\end{itemize}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Definitions:} A number is \textit{irrational} if it cannot be written as $p/q$. It is \textit{transcendental} if it is not the root of any polynomial with integer coefficients. All transcendental numbers are irrational, but not all irrational numbers (like $\sqrt{2}$) are transcendental.

2. \textbf{Taylor Series for $e$:} $e = \sum_{n=0}^\infty \frac{1}{n!}$.

3. \textbf{The Hermite-Lindemann Theorem:} A powerhouse of number theory stating that if $\alpha$ is a non-zero algebraic number, then $e^\alpha$ is transcendental.


\textbf{Steps:}

1. \textbf{Prove $e$ is irrational:}
    Assume $e = p/q$. Then $q!(e - \sum_{n=0}^q \frac{1}{n!})$ must be a positive integer. However, using the tail of the Taylor series:
    \[ q! \sum_{n=q+1}^\infty \frac{1}{n!} = \frac{1}{q+1} + \frac{1}{(q+1)(q+2)} + \dots < \frac{1}{q}. \]
    This shows the value is between 0 and 1, which is impossible for an integer. Thus, $e$ is irrational.

2. \textbf{Prove $\pi$ is irrational:}
    This typically uses an integral like $I_n = \int_0^\pi \frac{x^n(a-bx)^n}{n!} \sin(x) dx$. If $\pi = a/b$, one can show that $I_n$ is an integer for all $n$. But as $n \to \infty$, the integral $I_n \to 0$, eventually becoming a positive value less than 1. This contradiction proves $\pi$ is irrational.

3. \textbf{Prove transcendence:}
    \begin{itemize}
        \item \textbf{For $e$:} Hermite proved this in 1873 by showing that $e$ cannot satisfy any polynomial equation.
        \item \textbf{For $\pi$:} Lindemann (1882) used Euler's identity $e^{i\pi} = -1$. If $\pi$ were algebraic, then $i\pi$ would be algebraic. By the Hermite-Lindemann theorem, $e^{i\pi}$ would then be transcendental. But $-1$ is clearly not transcendental! Therefore, $\pi$ must be transcendental.
    \end{itemize}


\textbf{Intuition:}

Proving a number is "wild" (irrational or transcendental) usually involves showing it is "too slippery" to be caught by simple numbers. For $e$, the proof shows it is "too close" to its approximations to be a fraction. For $\pi$, the proof uses its deep connection to the exponential function ($e^{i\pi}=-1$) to show that if $e$ is wild, $\pi$ must be wild too. These proofs rely on the "Gap Principle": an integer can't be between 0 and 1. If we can squeeze a "supposed" integer into that gap, we know our initial assumption (that the number was simple) was wrong.


\end{solution}
