\section{2024 Analysis \& Differential Equations}

\subsection{Exam}

\begin{exercise}
Let $Q:\mathbb{R}\to\mathbb{R}$ be a $C_c^{\infty}$ function, i.e., it is smooth and has compact support. We assume $Q$ is even, i.e., $Q(x)=Q(-x)$. We assume $Q$ is non-trivial (i.e., $Q$ does not equal to zero everywhere).

Let $T_1(x) := xQ(x)$, and let $T_2(x) = x^2 Q(x)$. Let $T_3 := e^{-x^2}(1+x^{2024})$.

We also introduce the following notation. For any $f:\mathbb{R}\to\mathbb{R}$, $\lambda>0$, $\alpha\in\mathbb{R}$, we define
\[
f_{\lambda, \alpha}(x) := \frac{1}{\lambda^{1/2}}f\left(\frac{x-\alpha}{\lambda}\right).
\]

We claim: There exists $\delta>0$, $\epsilon>0$, so that for any $c\in\mathbb{R}$ with $|c|<\delta$, one can find unique $\lambda,\alpha$ such that the followings hold:
\begin{enumerate}
\item $| \lambda-1 | + | \alpha | < \epsilon$.
\item $\langle Q_{\lambda,\alpha} - Q - cT_3, T_1 \rangle = 0$.
\item $\langle Q_{\lambda,\alpha} - Q - cT_3, T_2 \rangle = 0$.
\end{enumerate}
(Here, for any two functions $f_1,f_2$, we define $\langle f_1,f_2\rangle := \int f_1(x)f_2(x)\,dx$.)

Is the above claim correct? Prove your conclusion.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Implicit Function Theorem (IFT):} A powerful tool in calculus used to solve equations that cannot be easily isolated. It states that if you have a smooth relationship $G(p, c) = 0$ and the relationship's "slope" (Jacobian) at a known solution $(p_0, 0)$ is invertible, then for small changes in $c$, there exists a unique way to adjust $p$ to keep the equation satisfied.
    \item \textbf{Function Parity (Even/Odd):} An even function satisfies $f(x) = f(-x)$ (symmetric about the y-axis), and an odd function satisfies $f(x) = -f(-x)$ (symmetric about the origin). The integral of an odd function over a symmetric range centered at zero is zero.
    \item \textbf{Inner Products:} The notation $\langle f, g \rangle$ is a compact way of writing the definite integral $\int_{-\infty}^{\infty} f(x)g(x) \, dx$.
\end{itemize}

\textbf{Steps:}

\par\vspace{\baselineskip}
1. \textbf{Define the Map:} We want to find values for stretch ($\lambda$) and shift ($\alpha$) to balance a small "perturbation" (error) $cT_3$. We define a map $G$ that measures how well the balance conditions are met:
\[
G(\lambda, \alpha, c) = \begin{pmatrix} \langle Q_{\lambda,\alpha} - Q - cT_3, T_1 \rangle \\ \langle Q_{\lambda,\alpha} - Q - cT_3, T_2 \rangle \end{pmatrix}.
\]
We want to solve $G(\lambda, \alpha, c) = \begin{pmatrix} 0 \\ 0 \end{pmatrix}$ for $(\lambda, \alpha)$ near $(1, 0)$ when $c$ is near $0$.

\par\vspace{\baselineskip}
2. \textbf{Check the Starting Point:} When $c=0$, if we choose the original function ($\lambda=1, \alpha=0$), $Q_{1,0} = Q$. This makes the whole equation $Q - Q - 0 = 0$, so $G(1, 0, 0) = 0$. This confirms we have a valid starting point.

\par\vspace{\baselineskip}
3. \textbf{Calculate the Jacobian (The "Sensitivity"):} We need to know how sensitive our equations are to changes in $\lambda$ and $\alpha$ at the starting point.
The Jacobian matrix $J = \frac{\partial(G_1, G_2)}{\partial(\lambda, \alpha)}$ describes this sensitivity.
\begin{itemize}
    \item \textbf{Effect of Shift ($\alpha$):} Shifting the function by $\alpha$ is like calculating the negative derivative of the function. Using calculus and integration by parts, we find $\frac{\partial G_1}{\partial \alpha} = \frac{1}{2} \|Q\|_2^2 > 0$ and $\frac{\partial G_2}{\partial \alpha} = 0$. The zero results from the fact that we are integrating an odd function over the whole real line.
    \item \textbf{Effect of Stretch ($\lambda$):} Stretching the function changes its density. We find $\frac{\partial G_1}{\partial \lambda} = 0$ (again, due to symmetry) and $\frac{\partial G_2}{\partial \lambda} = \int x^2 Q^2 \, dx > 0$.
\end{itemize}

\par\vspace{\baselineskip}
4. \textbf{Invertibility:} The Jacobian matrix becomes a diagonal matrix:
\[
J = \begin{pmatrix} \frac{1}{2}\|Q\|_2^2 & 0 \\ 0 & \int x^2 Q^2 \end{pmatrix}.
\]
Since $Q$ is not zero, these integrals are positive, and the determinant is non-zero. This means the matrix is invertible, satisfying the condition for the Implicit Function Theorem.

\par\vspace{\baselineskip}
\textbf{Intuition:}
Imagine you have a shape $Q$ and someone adds a small, distorted piece $cT_3$ to it. The problem is asking if you can "fix" this by just shifting ($ \alpha $) and stretching ($ \lambda $) the original $Q$ to get back to a state that satisfies two specific balance conditions.

The Implicit Function Theorem gives us a "Yes." It works because shifting and stretching have independent, measurable effects on the system (which we verified by the diagonal Jacobian). Symmetry is the secret ingredient here: it makes the "balancing act" easy because the shift only affects one condition and the stretch only affects the other, decoupling the variables and making the solution mathematically guaranteed.
\end{solution}

\begin{exercise}
Recall for every $f\in L^2(\mathbb{R}^3)$, one has that $g(x) := (-\Delta+1)^{-1}f$ is a well-defined $L^2(\mathbb{R}^3)$ function. And one may compute $g$ by solving
\[
(-\Delta+1)g = f.
\]
(Recall $\Delta$ in $\mathbb{R}^3$ is defined as $\Delta := \sum_{i=1}^3 \partial_i^2$, also recall one may also define $(-\Delta+1)^{-1}$ by Fourier theory.)

Now, let $V(x) := e^{-|x|^2}$, $x\in\mathbb{R}^3$. Prove that the operator $T := I + (-\Delta+1)^{-1}V$ is invertible in $L^2$.
\[
\left(\text{Here}, Tf := f + (-\Delta+1)^{-1}(Vf).\right)
\]
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Fredholm Alternative:} A key theorem for operators of the form $I + K$. It says that if $K$ is a "nice" (compact) operator, then solving the equation $(I+K)f = h$ is as easy as solving it in finite-dimensional linear algebra. If the homogeneous equation $(I+K)f = 0$ has only the solution $f=0$, then $I+K$ is invertible.
    \item \textbf{Compact Operators:} In this context, an operator that "smooths" the input and limits it spatially. The $(-\Delta+1)^{-1}$ operator smooths the function (frequency decay), and multiplication by $V$ limits it spatially (spatial decay). Together, they form a "nice" compact operator.
    \item \textbf{Energy Method (Integration by Parts):} A technique where we multiply an equation by a function $g$ and integrate to relate the total "energy" (squared magnitude) to the derivatives.
\end{itemize}

\textbf{Steps:}

\par\vspace{\baselineskip}
1. \textbf{Identify the Setup:} We have $T = I + K$, where $K f = (-\Delta+1)^{-1}(Vf)$. We need to verify if $T$ is invertible.

\par\vspace{\baselineskip}
2. \textbf{Apply the Fredholm Alternative:} Because our operator $K$ is compact (it combines smoothing and spatial decay), we can use the Fredholm Alternative. This reduces our massive task to just checking one simple thing: Does $Tf = 0$ only happen when $f=0$?

\par\vspace{\baselineskip}
3. \textbf{Assume a Non-zero Solution:} Suppose there is some $f \neq 0$ such that $Tf = 0$.
$f + (-\Delta+1)^{-1}(Vf) = 0$.
Let $g = (-\Delta+1)^{-1}(Vf)$. This implies $f = -g$.
Substitute $f = -g$ into the definition of $g$:
$g = (-\Delta+1)^{-1}(V(-g)) \implies (-\Delta+1)g = -Vg$.

\par\vspace{\baselineskip}
4. \textbf{Rearrange the Equation:}
$(-\Delta + 1 + V)g = 0$.
This is a standard linear differential equation.

\par\vspace{\baselineskip}
5. \textbf{Calculate Energy:} Multiply by $g$ and integrate over all space ($\mathbb{R}^3$):
\[
\int_{\mathbb{R}^3} (-\Delta g \cdot g + g \cdot g + V \cdot g \cdot g) \, dx = 0.
\]
Using integration by parts on the first term:
\[
\int_{\mathbb{R}^3} (|\nabla g|^2 + g^2 + Vg^2) \, dx = 0.
\]

\par\vspace{\baselineskip}
6. \textbf{Draw the Conclusion:} Every part of this sum ($|\nabla g|^2$, $g^2$, $Vg^2$) is non-negative because they are squares or products of non-negative terms. For the sum to equal zero, the only possibility is that $g=0$ everywhere.
Since $f = -g$, $f$ must also be $0$.
Because the only solution to $Tf=0$ is $f=0$, $T$ is invertible.

\par\vspace{\baselineskip}
\textbf{Intuition:}
Think of $T = I + K$ as a system with a "correction" $K$. The Fredholm Alternative tells us that if this correction isn't "too big" (which "compactness" guarantees), the only way the system could fail to be invertible is if it secretly squashes some signal to zero.

\par\vspace{\baselineskip}
To see if that happens, we look at the energy of the system. The equation $(-\Delta + 1 + V)g = 0$ is like the energy balance equation for a quantum state. Because the operator $(-\Delta + 1 + V)$ strictly increases the energy of any non-zero state, there is no way for a non-zero state to have a total energy of zero. Therefore, no non-zero function can be squashed to zero, and the system must be fully invertible.
\end{solution}

\begin{exercise}
Let $\psi(\xi)\in C_c^{\infty}(\mathbb{R})$ be smooth and has compact support. Let $\psi(\xi)=0$, $\forall|\xi|\geq 1$. Let $f_1(\xi), f_2(\xi)\in C_c^{\infty}(\mathbb{R})$, i.e., $f_1,f_2$ are smooth and have compact support. Let $u_i:\mathbb{R}\times\mathbb{R}\to\mathbb{C}$, $i=1,2$, be defined as
\[
u_1(x_1,x_2) := \int_{\mathbb{R}}\psi(\xi)f_1(\xi)e^{i\xi x_1}e^{i\xi^2 x_2}\,d\xi,
\]
\[
u_2(x_1,x_2) := \int_{\mathbb{R}}\psi(\eta-10)f_2(\eta)e^{i\eta x_1}e^{i\eta^2 x_2}\,d\eta.
\]

Prove there exists a constant $C$, which may depend on $\psi$, but does not depend on $f_1,f_2$, so that
\[
\|u_1 u_2\|_{L^2(\mathbb{R}^2)} \leq C\|f_1\|_{L^2(\mathbb{R})}\|f_2\|_{L^2(\mathbb{R})}.
\]

(Hint: One may try to use Plancherel Theorem. It may be useful to observe that if one let $H(\xi,\eta)=f_1(\xi)f_2(\eta)$, then $\|H\|_{L^2(\mathbb{R}^2)}$ are also bounded by $\|f_1\|_{L^2(\mathbb{R})}\|f_2\|_{L^2(\mathbb{R})}$.)
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Plancherel's Theorem:} A powerful bridge between the physical world (where we have positions) and the frequency world (where we have waves). It states that the "total energy" (the integral of the square) of a function is the same, up to a constant factor, as the total energy of its Fourier transform.
    \item \textbf{Jacobian (Change of Variables):} When switching coordinates in integrals (like from frequency $(\xi, \eta)$ to a new coordinate system $(X, Y)$), the Jacobian tells us how much the "volume" of space stretches or shrinks.
    \item \textbf{Support of a Function:} A function's "support" is simply the range of values where it is non-zero. It allows us to limit where we need to look in our calculations.
\end{itemize}

\textbf{Steps:}

\par\vspace{\baselineskip}
1. \textbf{Combine the Integrals:} Multiply $u_1$ and $u_2$ together. By grouping the terms, the product looks like a 2D integral over frequencies $\xi$ and $\eta$:
\[
u_1 u_2(x_1, x_2) = \iint (\psi(\xi)f_1(\xi))(\psi(\eta-10)f_2(\eta)) e^{i(\xi+\eta)x_1 + i(\xi^2+\eta^2)x_2} \, d\xi d\eta.
\]

\par\vspace{\baselineskip}
2. \textbf{Define New Coordinates:} Notice that the exponent $i((\xi+\eta)x_1 + (\xi^2+\eta^2)x_2)$ looks like a 2D Fourier transform. Let's define new coordinates:
$X = \xi + \eta$ (total frequency)
$Y = \xi^2 + \eta^2$ (total oscillation speed)

\par\vspace{\baselineskip}
3. \textbf{Calculate the Jacobian Stretch:} How does the space $(\xi, \eta)$ stretch when we move to $(X, Y)$? The Jacobian is the determinant of the partial derivatives:
\[
\text{Jacobian } J = \left| \det \begin{pmatrix} 1 & 1 \\ 2\xi & 2\eta \end{pmatrix} \right| = 2|\eta - \xi|.
\]

\par\vspace{\baselineskip}
4. \textbf{Verify Frequency Separation:} Because of the way we defined the functions, $\xi$ is always near $0$ (from $\psi(\xi)$) and $\eta$ is always near $10$ (from $\psi(\eta-10)$). This means $\eta$ and $\xi$ are always separated by at least $8$. Since $J = 2|\eta-\xi|$, the Jacobian $J$ is always at least $16$. This "separation" ensures we don't have a division-by-zero disaster.

\par\vspace{\baselineskip}
5. \textbf{Apply Plancherel's Theorem:} We switch to the $(X, Y)$ coordinate system. Plancherel's theorem tells us the energy of the product $u_1 u_2$ is equal to the integral of its Fourier transform squared:
\[
\|u_1 u_2\|_{L^2}^2 = (2\pi)^2 \iint |H(X, Y)|^2 \, dX dY,
\]
where $H$ is the density of our frequencies in the new system.

\par\vspace{\baselineskip}
6. \textbf{Final Transformation:} By switching back to $(\xi, \eta)$ coordinates and including the $1/J$ factor from the change of variables:
\[
\iint |H(X, Y)|^2 \, dX dY = \iint \frac{|\psi(\xi)f_1(\xi) \cdot \psi(\eta-10)f_2(\eta)|^2}{J} \, d\xi d\eta.
\]
Since $J \geq 16$, the whole integral is bounded by a constant times the individual integrals of $f_1^2$ and $f_2^2$.

\par\vspace{\baselineskip}
\textbf{Intuition:}
This is a "bilinear estimate." Think of it as a sanity check for waves. If you multiply two waves together, you usually get a mess. But if you multiply two waves that are moving in completely different directions (different frequencies), they don't "pile up" or create huge spikes in energy.

\par\vspace{\baselineskip}
The Jacobian $J$ measures how "separated" these two waves are. Because our frequencies are well-separated ($\xi$ near $0$, $\eta$ near $10$), the Jacobian stays nice and large. This acts like a buffer, ensuring the product $u_1 u_2$ stays "well-behaved" and its energy is bounded by the energies of the two original signals.
\end{solution}

\begin{exercise}
Consider the heat equation in $\mathbb{R}^2$. Let $u=u(t,x)$ is a solution to
\[
\begin{cases}
\frac{\partial u}{\partial t} - \Delta u = 0;\
u|_{t=0} = u_0\in L^2.
\end{cases}
\]

Then there exists a universal constant $C$ such that
\[
\int_0^{\infty}\|u(t)\|_{L^{\infty}}^2\,dt \leq C\|u_0\|_{L^2}^2.
\]
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Heat Kernel Estimate:} This formula tells us how fast the maximum value of a heat distribution decays. In 2D, the peak value decays at a rate of $1/t$.
    \item \textbf{$TT^*$ Argument:} A clever mathematical trick. Instead of proving an operator is "bounded" directly, we look at the operator combined with its adjoint (the "reverse" operator). This often turns hard problems into much easier ones by revealing how different points in time interact.
    \item \textbf{Hilbert's Inequality:} A famous result that bounds the integral of a special, structured kernel ($1/(s+t)$). It's the key to proving that the total "energy interaction" over all time is bounded.
\end{itemize}

\textbf{Steps:}

\par\vspace{\baselineskip}
1. \textbf{Define the Operator:} Let $T$ be the operator that turns our initial state $u_0$ into the entire history of the heat solution $u(t,x)$. We want to check if the total "energy" of the peak values over time is finite.

\par\vspace{\baselineskip}
2. \textbf{Use the $TT^*$ Trick:} We check how the heat flow behaves when we go from time $s$ to time $t$. The heat kernel $e^{(s+t)\Delta}$ links these two times.

\par\vspace{\baselineskip}
3. \textbf{Estimate Peak Decay:} At any two times $s$ and $t$, the heat flow formula says the interaction is bounded by $1/(s+t)$. This is the decay rate of the heat kernel in 2D.

\par\vspace{\baselineskip}
4. \textbf{Apply Hilbert's Inequality:} Our problem reduces to estimating a double integral of the form $\iint \frac{a(s)a(t)}{s+t} ds dt$. This is exactly the type of integral Hilbert's Inequality was built to bound! The inequality proves that the total interaction across all time $(s, t)$ is finite.

\par\vspace{\baselineskip}
5. \textbf{Conclude:} The constant $C$ comes directly from the constant in Hilbert's Inequality and the heat kernel factor.

\par\vspace{\baselineskip}
\textbf{Intuition:}
The heat equation is a "smoothing" process—it spreads out concentrated heat until it vanishes. The problem asks: if we sum the square of the maximum temperature at every moment in time, is that sum infinite or finite?

\par\vspace{\baselineskip}
Think of it like an echo in a room. The echo decays over time. We want to know if the total "loudness" (integrated square of the peak) is finite. Because the heat spreads out in 2D at the perfect rate ($1/t$), the echoes don't pile up in a way that makes the total energy blow up. Hilbert's inequality confirms that this "echo decay" is fast enough that the total sum stays finite.
\end{solution}

\begin{exercise}
Consider the Fourier transform. Let
\[
Q(g,f)(x) := \int_{\mathbb{R}^N}\int_{S^{N-1}} B(|x-y|,\frac{x-y}{|x-y|}\cdot\sigma)g(y')f(x')\,d\sigma\,dy,
\]
where $B$ is a given two variable function, $S^{N-1}$ stands for the unit sphere in $\mathbb{R}^N$ and
\[
x' := \frac{x+y}{2} + \frac{|x-y|\sigma}{2};\quad y' := \frac{x+y}{2} - \frac{|x-y|\sigma}{2}.
\]

Then
\[
\widehat{Q(g,f)}(\xi) = (2\pi)^{-N/2}\int_{\mathbb{R}^N\times S^{N-1}} \hat{B}(|\eta|,\frac{\xi}{|\xi|}\cdot\sigma)\hat{g}(\xi^-+\eta)\hat{f}(\xi^+-\eta)\,d\sigma\,d\eta,
\]
where $\hat{B}(|\eta|,t) := \int_{\mathbb{R}^N}B(|q|,t)e^{-iq\cdot\eta}\,dq$, $\xi^{\pm} := \frac{\xi\pm|\xi|\sigma}{2}$.
\end{exercise}

\begin{solution}
\textbf{Motivation:} This is known as Bobylev's Formula, a landmark result in kinetic theory. It is the "magic key" for solving the Boltzmann equation (which describes how particles collide). Instead of dealing with complicated collision integrals in physical space (where particles hit each other), it moves the problem into frequency space, where the math becomes much more transparent.

The derivation follows three major steps:

1. \textbf{Coordinate Transformation:} We switch from particle coordinates $(x, y)$ to center-of-mass and relative velocity. This separates the "averaging" effect of the collision from the particles' individual motions.
2. \textbf{Fourier Analysis:} We apply the Fourier transform $\mathcal{F}$. A key trick is using the identity $\mathcal{F}[g(v-a)f(v+a)](\xi)$, which converts the collision into a convolution-like structure involving the frequency-shifted versions of $\hat{g}$ and $\hat{f}$.
3. \textbf{Symmetry Utilization:} We leverage the symmetry of the unit sphere $S^{N-1}$ and the definition of $\hat{B}$. After a rotation and a change of variables ($\eta = \xi^- + \eta'$), the integral simplifies into the final, symmetric form that relates the collision kernel $\hat{B}$ to the pre-collision frequencies.

This formula shows that in frequency space, collisions are just a weighted average of product-pairs of the original distribution's frequency components.
\end{solution}
