\section{2016 Analysis \& Differential Equations}

\subsection{Individual}

\begin{exercise}
Suppose that $F$ is continuous on $[a,b]$, $F'(x)$ exists for every $x\in(a,b)$, and $F'(x)$ is integrable. Prove that $F$ is absolutely continuous and
\[
F(b) - F(a) = \int_a^b F'(x)\,dx.
\]
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Fundamental Theorem of Calculus (FTC):} In its simplest form, it relates differentiation and integration. Here we use the Lebesgue version.
    \item \textbf{Absolute Continuity:} A function $F$ is absolutely continuous if it can be written as the integral of its derivative.
    \item \textbf{Luzin N Property:} A function has this property if it maps sets of measure zero to sets of measure zero. Absolutely continuous functions possess this property.
    \item \textbf{Lebesgue Integrability:} The derivative $F'$ is given as integrable, which is a necessary condition for the integral to be well-defined.
\end{itemize}

\textbf{Intuition:}
The goal is to show that a function is completely determined by its derivative if that derivative exists everywhere and is "well-behaved" (integrable). The "Intuition" here is that if two functions have the same derivative almost everywhere and don't "jump" or "wiggle" too much (the Luzin N property), they must differ only by a constant. Since we can construct a function $G(x)$ that is the integral of $F'$, we just need to prove that $F(x) - G(x)$ is a constant.

\textbf{Steps:}

1. \textbf{Construct an auxiliary function:}
Define $G(x) = \int_a^x F'(t) dt$ for $x \in [a, b]$. 
Because $F'$ is integrable (given), this function $G(x)$ is well-defined. According to the properties of the Lebesgue integral, any function defined as an indefinite integral is automatically \textit{absolutely continuous}.

2. \textbf{Examine the derivatives:}
A key property of $G(x)$ is that $G'(x) = F'(x)$ for \textit{almost every} $x \in [a, b]$. 
We are given that $F'(x)$ exists for \textit{every} $x \in (a, b)$.
Now, consider the difference function $H(x) = F(x) - G(x)$.
This function $H$ is continuous because both $F$ and $G$ are continuous.
Furthermore, $H'(x) = F'(x) - G'(x) = 0$ for almost every $x \in [a, b]$.

3. \textbf{Apply the Luzin N property:}
$F(x)$ has a derivative everywhere in $(a, b)$. A theorem by Banach and Zarecki states that if a continuous function has a derivative everywhere (or even except on a countable set), it satisfies the \textit{Luzin N property}.
Since $G$ is absolutely continuous, it also satisfies the Luzin N property.
Therefore, their difference $H = F - G$ also satisfies the Luzin N property.

4. \textbf{Conclude the constant nature of H:}
We have a continuous function $H$ that:
(a) Has a derivative equal to zero almost everywhere.
(b) Satisfies the Luzin N property.
A fundamental result in real analysis states that any such function must be \textit{constant}.
Thus, $H(x) = C$ for some constant $C$.

5. \textbf{Final Result:}
To find $C$, we plug in $x = a$:
$H(a) = F(a) - G(a) = F(a) - 0 = F(a)$.
So, $H(x) = F(a)$ for all $x$.
This means $F(x) - G(x) = F(a)$, or $F(x) = F(a) + \int_a^x F'(t) dt$.
This identity confirms that $F$ is absolutely continuous and that $F(b) - F(a) = \int_a^b F'(x) dx$.
\end{solution}

\begin{exercise}
Suppose that $f$ is integrable on $\mathbb{R}^n$, let $K_{\delta}(x) = \delta^{-\frac{n}{2}}e^{-\frac{\pi|x|^2}{\delta}}$ for each $\delta>0$. Prove that the convolution
\[
(f*K_{\delta})(x) = \int_{\mathbb{R}^n} f(x-y)K_{\delta}(y)\,dy
\]
is integrable and $\| (f*K_{\delta}) - f\|_{L^1(\mathbb{R}^n)} \to 0$ as $\delta\to 0$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Convolution:} The operation $(f * g)(x) = \int_{\mathbb{R}^n} f(x-y)g(y) dy$.
    \item \textbf{Young's Inequality (for $L^1$):} $\|f * g\|_1 \leq \|f\|_1 \|g\|_1$.
    \item \textbf{Gaussian Integral:} $\int_{-\infty}^\infty e^{-ax^2} dx = \sqrt{\pi/a}$. For $\mathbb{R}^n$, $\int_{\mathbb{R}^n} e^{-\pi |x|^2} dx = 1$.
    \item \textbf{Continuity of Translation in $L^1$:} For any $f \in L^1(\mathbb{R}^n)$, the mapping $y \mapsto f(\cdot - y)$ is continuous from $\mathbb{R}^n$ to $L^1(\mathbb{R}^n)$. That is, $\lim_{y \to 0} \|f(\cdot - y) - f(\cdot)\|_1 = 0$.
\end{itemize}

\textbf{Intuition:}
The function $K_\delta(x)$ is a "mollifier" or an "approximation to the identity." As $\delta$ becomes very small, $K_\delta$ becomes a very sharp peak at the origin, while the area under the curve remains exactly 1. Convolving $f$ with $K_\delta$ is like taking a weighted average of $f$ around each point $x$. Since the weight is concentrated near the origin when $\delta$ is small, the result $(f * K_\delta)(x)$ should be very close to $f(x)$.

\textbf{Steps:}

\textbf{Part 1: Proof of Integrability}
1. \textbf{Apply Young's Inequality:}
By Young's inequality for convolutions, the $L^1$ norm of the product is bounded by the product of the $L^1$ norms:
$\|f * K_\delta\|_1 \leq \|f\|_1 \|K_\delta\|_1$.

2. \textbf{Compute the norm of the Kernel:}
We need to calculate $\|K_\delta\|_1 = \int_{\mathbb{R}^n} \delta^{-n/2} e^{-\pi |x|^2 / \delta} dx$.
Use the substitution $u = x / \sqrt{\delta}$, which implies $dx = \delta^{n/2} du$.
The integral becomes:
$\int_{\mathbb{R}^n} \delta^{-n/2} e^{-\pi |u|^2} (\delta^{n/2} du) = \int_{\mathbb{R}^n} e^{-\pi |u|^2} du = 1$.
Since $\|f\|_1$ is finite and $\|K_\delta\|_1 = 1$, the convolution $\|f * K_\delta\|_1$ is finite, meaning it is integrable.

\textbf{Part 2: Proof of $L^1$ Convergence}
1. \textbf{Setup the difference integral:}
Since $\int K_\delta(y) dy = 1$, we can write $f(x) = \int f(x) K_\delta(y) dy$.
Then, $\|f * K_\delta - f\|_1 = \int_{\mathbb{R}^n} | \int_{\mathbb{R}^n} (f(x-y) - f(x)) K_\delta(y) dy | dx$.
By Fubini's Theorem (swapping the order of integration):
$\|f * K_\delta - f\|_1 \leq \int_{\mathbb{R}^n} K_\delta(y) \left( \int_{\mathbb{R}^n} |f(x-y) - f(x)| dx \right) dy$.

2. \textbf{Analyze the inner integral:}
Let $\omega_f(y) = \int_{\mathbb{R}^n} |f(x-y) - f(x)| dx$. This is the $L^1$ "modulus of continuity."
We know two things:
(a) $\lim_{y \to 0} \omega_f(y) = 0$.
(b) $\omega_f(y) \leq \|f(\cdot-y)\|_1 + \|f\|_1 = 2\|f\|_1$ (boundedness).

3. \textbf{Split the domain of $y$:}
For any $\epsilon > 0$, choose $\eta > 0$ such that $|y| < \eta \implies \omega_f(y) < \epsilon$.
Split the outer integral into two parts: $|y| < \eta$ and $|y| \geq \eta$.
$I_1 = \int_{|y|<\eta} K_\delta(y) \omega_f(y) dy < \epsilon \int K_\delta(y) dy = \epsilon$.
$I_2 = \int_{|y|\geq \eta} K_\delta(y) \omega_f(y) dy \leq 2\|f\|_1 \int_{|y|\geq \eta} K_\delta(y) dy$.

4. \textbf{Take the limit as $\delta \to 0$:}
As $\delta \to 0$, the mass of $K_\delta$ concentrates at the origin.
Specifically, $\int_{|y|\geq \eta} K_\delta(y) dy = \int_{|u| \geq \eta/\sqrt{\delta}} e^{-\pi |u|^2} du$.
As $\delta \to 0$, the lower limit $\eta/\sqrt{\delta} \to \infty$, so the integral goes to 0.
Thus, $\limsup_{\delta \to 0} \|f * K_\delta - f\|_1 \leq \epsilon$. Since $\epsilon$ is arbitrary, the limit is 0.
\end{solution}

\begin{exercise}
Prove that a bounded function on interval $I=[a,b]$ is Riemann integrable if and only if its set of discontinuities has measure zero.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Riemann Integrability:} A bounded function $f$ on $[a, b]$ is Riemann integrable if for every $\epsilon > 0$, there exists a partition $P$ such that the difference between the upper and lower Darboux sums $U(P, f) - L(P, f) < \epsilon$.
    \item \textbf{Oscillation:} The oscillation of $f$ at a point $x$ is $\omega(f, x) = \lim_{\delta \to 0} \sup \{|f(y) - f(z)| : y, z \in (x-\delta, x+\delta) \cap [a, b]\}$. A function is continuous at $x$ if and only if $\omega(f, x) = 0$.
    \item \textbf{Lebesgue Measure Zero:} A set $S \subset \mathbb{R}$ has measure zero if for every $\epsilon > 0$, it can be covered by a countable collection of open intervals whose total length is less than $\epsilon$.
\end{itemize}

\textbf{Intuition:}
Riemann integration works by approximating a function with rectangles. If a function has too many discontinuities, the heights of these rectangles (the "gap" between the maximum and minimum values in a subinterval) won't shrink to zero even as the widths do. Lebesgue's criterion tells us that as long as the "bad" points (discontinuities) are few enough to be ignored in terms of length (measure zero), the function remains integrable.

\textbf{Steps:}

\textbf{Step 1: Characterize the set of discontinuities}
Let $D$ be the set of points where $f$ is discontinuous. As noted, $x \in D \iff \omega(f, x) > 0$.
We can write $D$ as a countable union: $D = \bigcup_{k=1}^\infty D_{1/k}$, where $D_\epsilon = \{x \in [a, b] : \omega(f, x) \geq \epsilon\}$.
It is a known property that $D_\epsilon$ is always a \textit{closed} and \textit{bounded} (hence compact) set.

\textbf{Step 2: Proof that $m(D)=0 \implies$ Riemann Integrability}
1. Assume $m(D) = 0$. This implies $m(D_\epsilon) = 0$ for all $\epsilon > 0$.
2. Fix $\epsilon > 0$. Since $m(D_\epsilon) = 0$, we can cover $D_\epsilon$ by a finite number of open intervals $(a_i, b_i)$ with total length $\sum (b_i - a_i) < \epsilon$. (Finite because $D_\epsilon$ is compact).
3. The remaining part of $[a, b] \setminus \bigcup (a_i, b_i)$ is a collection of closed intervals. On these intervals, the oscillation of $f$ is strictly less than $\epsilon$.
4. We can create a partition $P$ using the endpoints of these intervals.
5. In the sum $U(P, f) - L(P, f) = \sum (M_j - m_j) \Delta x_j$:
   - For subintervals covering $D_\epsilon$, the total contribution is $\leq (M - m) \cdot \sum \Delta x_j < (M - m)\epsilon$, where $M, m$ are the global bounds of $f$.
   - For the other subintervals, the contribution is $\leq \epsilon \cdot (b - a)$.
6. Thus, $U - L \leq \epsilon(M - m + b - a)$. Since $\epsilon$ is arbitrary, $f$ is Riemann integrable.

\textbf{Step 3: Proof that Riemann Integrability $\implies m(D)=0$}
1. Assume $f$ is Riemann integrable. Then for any $\eta > 0$, there is a partition $P$ such that $U(P, f) - L(P, f) < \eta$.
2. For a fixed $\epsilon > 0$, consider the set $D_\epsilon$. If a subinterval $I_j = [x_{j-1}, x_j]$ of the partition contains a point of $D_\epsilon$ in its interior, then $M_j - m_j \geq \epsilon$.
3. Let $J$ be the set of indices of subintervals that contain points of $D_\epsilon$.
4. Then $\eta > U - L \geq \sum_{j \in J} (M_j - m_j) \Delta x_j \geq \epsilon \sum_{j \in J} \Delta x_j$.
5. This means the total length of subintervals covering $D_\epsilon$ (excluding partition points, which are finite) is $\sum_{j \in J} \Delta x_j < \eta / \epsilon$.
6. Since $\eta$ can be chosen arbitrarily small, $m(D_\epsilon) = 0$.
7. Since $D = \bigcup D_{1/k}$ and each $D_{1/k}$ has measure zero, their union $D$ also has measure zero.
\end{solution}

\begin{exercise}
\begin{itemize}
    \item[1)] Let $w=\frac{1}{2}(z+\frac{1}{z})$. Show that it maps $\{|z|>1\}$ to $\mathbb{C}\setminus[-1,1]$.
    \item[2)] Compute $\int_0^\infty \frac{\log x}{x^2-1} dx$.
\end{itemize}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
\begin{itemize}
    \item \textbf{Complex Mapping:} Representing $z = x + iy$ as $re^{i\theta}$.
    \item \textbf{Ellipses:} The standard equation $u^2/a^2 + v^2/b^2 = 1$ with semi-axes $a$ and $b$.
    \item \textbf{Improper Integrals:} Handling $\log x$ and singularities at $x=1$.
    \item \textbf{Taylor Series:} $1/(1-r) = \sum r^n$ for $|r| < 1$.
    \item \textbf{Riemann Zeta Function:} $\sum_{k=1}^\infty \frac{1}{k^2} = \frac{\pi^2}{6}$ (Basel Problem).
\end{itemize}

\textbf{Intuition:}
\textbf{Part 1:} The map $w = \frac{1}{2}(z + 1/z)$ is known as the Zhukovsky transform. It "squashes" the complex plane. Think of circles $|z|=r$. For $r=1$, it maps the circle to the segment $[-1, 1]$ twice. For $r > 1$, it maps circles to larger and larger ellipses that eventually cover the entire plane as $r \to \infty$.

\textbf{Part 2:} The integral $\int_0^\infty \frac{\log x}{x^2-1} dx$ looks tricky because of the $x^2-1$ in the denominator. However, the $\log x$ also vanishes at $x=1$, making the singularity removable. The symmetry between $[0, 1]$ and $[1, \infty]$ under the map $x \to 1/x$ allows us to simplify the problem to an interval $[0, 1]$ where we can use series.

\textbf{Steps:}

\textbf{Part 1: The Mapping}
1. Let $z = re^{i\theta}$ where $r > 1$. Then $1/z = \frac{1}{r}e^{-i\theta}$.
2. Substitute into $w = u + iv$:
   $u + iv = \frac{1}{2}(re^{i\theta} + \frac{1}{r}e^{-i\theta}) = \frac{1}{2}(r(\cos\theta + i\sin\theta) + \frac{1}{r}(\cos\theta - i\sin\theta))$.
3. Separate the real and imaginary parts:
   $u = \frac{1}{2}(r + 1/r)\cos\theta$ and $v = \frac{1}{2}(r - 1/r)\sin\theta$.
4. Define $a = \frac{1}{2}(r + 1/r)$ and $b = \frac{1}{2}(r - 1/r)$. Then $\frac{u^2}{a^2} + \frac{v^2}{b^2} = \cos^2\theta + \sin^2\theta = 1$.
5. This is an ellipse. Since $r > 1$, $a > 1$ and $b > 0$. As $r \to 1^+$, $a \to 1$ and $b \to 0$, so the ellipse collapses to the interval $[-1, 1]$.
6. As $r \to \infty$, $a, b \to \infty$. Thus, the image of $\{|z| > 1\}$ is the entire plane $\mathbb{C}$ minus the "collapsed" starting segment $[-1, 1]$.

\textbf{Part 2: The Integral}
1. Split the integral: $I = \int_0^1 \frac{\log x}{x^2-1} dx + \int_1^\infty \frac{\log x}{x^2-1} dx$.
2. In the second part, let $x = 1/t$. Then $dx = -1/t^2 dt$.
   $\int_1^\infty \frac{\log x}{x^2-1} dx = \int_1^0 \frac{\log(1/t)}{(1/t)^2-1} (-1/t^2) dt = \int_0^1 \frac{-\log t}{\frac{1-t^2}{t^2}} \frac{1}{t^2} dt = \int_0^1 \frac{\log t}{t^2-1} dt$.
3. Thus, $I = 2 \int_0^1 \frac{\log x}{x^2-1} dx$.
4. Note that $\frac{1}{x^2-1} = \frac{-1}{1-x^2} = -\sum_{n=0}^\infty x^{2n}$.
   $I = -2 \int_0^1 \log x \left( \sum_{n=0}^\infty x^{2n} \right) dx = 2 \sum_{n=0}^\infty \int_0^1 (-x^{2n} \log x) dx$.
5. Using integration by parts (or the formula $\int_0^1 x^m \log x dx = \frac{-1}{(m+1)^2}$):
   $\int_0^1 x^{2n} \log x dx = \left[ \frac{x^{2n+1}}{2n+1}\log x \right]_0^1 - \int_0^1 \frac{x^{2n+1}}{2n+1} \frac{1}{x} dx = 0 - \frac{1}{(2n+1)^2}$.
6. $I = 2 \sum_{n=0}^\infty \frac{1}{(2n+1)^2}$.
7. We know $\sum_{k=1}^\infty \frac{1}{k^2} = \sum_{k \text{ even}} \frac{1}{k^2} + \sum_{k \text{ odd}} \frac{1}{k^2} = \frac{1}{4}\sum \frac{1}{j^2} + \sum \frac{1}{(2n+1)^2}$.
   So $\sum_{n=0}^\infty \frac{1}{(2n+1)^2} = (1 - 1/4) \zeta(2) = \frac{3}{4} \cdot \frac{\pi^2}{6} = \frac{\pi^2}{8}$.
8. Therefore, $I = 2 \cdot \frac{\pi^2}{8} = \frac{\pi^2}{4}$.
\end{solution}

\begin{exercise}
Let $f$ be a doubly periodic meromorphic function. Prove that the number of zeros and poles are equal.
\end{exercise}

\begin{solution}
By the Argument Principle, the number of zeros minus the number of poles in the fundamental parallelogram $\Omega$ is:
$Z - P = \frac{1}{2\pi i} \int_{\partial \Omega} \frac{f'(z)}{f(z)} dz$.
Since $f$ is doubly periodic, $f'/f$ is also doubly periodic.
The integral over opposite sides of the parallelogram cancel out because the path is in opposite directions and the function values are the same.
$\int_{\text{side } 1} + \int_{\text{side } 3} = \int_{z_0}^{z_0+\omega_1} \frac{f'(z)}{f(z)} dz + \int_{z_0+\omega_1+\omega_2}^{z_0+\omega_2} \frac{f'(z)}{f(z)} dz = 0$.
Similarly for the other pair of sides.
Thus $Z - P = 0 \implies Z = P$.
\end{solution}

\begin{exercise}
Prove that the spectrum of a bounded self-adjoint operator is a bounded closed subset of the real line.
\end{exercise}

\begin{solution}
1) \textbf{Bounded and Closed:} The spectrum $\sigma(A)$ is always closed and contained in $\{|\lambda| \leq \|A\|\}$ for any bounded operator.
2) \textbf{Real Line:} Let $\lambda = a + ib$ with $b \neq 0$.
$\|(A-\lambda I)x\|^2 = \|(A-aI)x - ibx\|^2 = \|(A-aI)x\|^2 + b^2\|x\|^2$ (since $A-aI$ is self-adjoint, cross terms cancel).
Thus $\|(A-\lambda I)x\| \geq |b|\|x\|$.
This implies $A-\lambda I$ is injective and has a closed range.
Since $(A-\lambda I)^* = A - \bar{\lambda} I$ also satisfies the same estimate, its kernel is $\{0\}$.
So the range of $A-\lambda I$ is dense.
A dense closed range is the whole space, so $A-\lambda I$ is invertible.
Thus $\lambda \notin \sigma(A)$, so $\sigma(A) \subset \mathbb{R}$.
\end{solution}

\subsection*{Team}

\begin{exercise}
Let $D \subset \mathbb{R}^d, d \geq 2$ be a compact convex set with smooth boundary $\partial D$ so that the origin belongs to the interior of $D$. For every $x \in \partial D$ let $\alpha(x) \in (0, \infty)$ be the angle between the position vector $x$ and the outer normal vector $n(x)$. Let $\omega_d$ be the surface area of the unit sphere in $\mathbb{R}^d$. Compute:
\[
\frac{1}{\omega_d} \int_{\partial D} \frac{\cos(\alpha(x))}{|x|^{d-1}} d\sigma(x)
\]
where $d\sigma$ denotes the surface measure on $\partial D$.
\end{exercise}

\begin{exercise}
Let $p > 0$ and suppose $f_n, f \in L^p[0,1]$ and $\|f_n - f\|_p = (\int_0^1 |f_n(x) - f(x)|^p dx)^{1/p} \to 0$ as $n \to \infty$.
\begin{itemize}
    \item[a)] Show that for every $\epsilon > 0$
    \[
    \lim_{n \to \infty} m(\{x \in [0, 1] \mid |f_n(x) - f(x)| > \epsilon\}) = 0.
    \]
    Here $m$ is the Lebesgue measure.
    \item[b)] Show that there exists a subsequence $f_{n_j}$ such that $f_{n_j}(x) \to f(x)$ for almost every $x \in [0, 1]$.
\end{itemize}
\end{exercise}

\begin{exercise}
\begin{itemize}
    \item[1)] Let $f$ be a holomorphic function on the unit disk $D = \{ z \in \mathbb{C} \mid |z| < 1 \}$ except 0. Assume $f \in L^2(D)$, i.e., $\int_D |f(z)|^2 dz d\bar{z} < \infty$, then 0 is a removable singularity.
    \item[2)] Let $f_n$ be a sequence of holomorphic functions over a domain $\Omega \subset \mathbb{C}$ converging to $f$ uniformly on any compact subset of $\Omega$, does the sequence of its derivatives $f_n'$ also have this property?
\end{itemize}
\end{exercise}

\begin{exercise}
Consider the torus $T^2 = \mathbb{C} / \Lambda, \Lambda = \{ m + in \mid m, n \in \mathbb{Z} \}$, i.e., $z_1, z_2 \in \mathbb{C}$ are equivalent if and only if there are integers $m, n$ such that $z_2 = z_1 + m + in$ and $T^2$ is the space of equivalence classes. Show that the group of holomorphic automorphisms of $T^2$ is $SL(2, \mathbb{Z})$ of $2 \times 2$ integer matrices of determinant 1.
\end{exercise}

\begin{exercise}
Let $\{e_n\}$ be an orthonormal basis of $l_2$ of square integrable functions over a circle. Let $A: l_2 \to l_2$ be a linear operator defined by $A e_1 = 0, A e_n = \frac{e_{n-1}}{n-1}$ for $n > 1$. Show that $A$ is a compact operator and $A$ has no eigenvectors. What is the spectrum of $A$?
\end{exercise}

\begin{exercise}
If $M = [0, 1]$ is the unit interval, the heat kernel on $M$ can be written
\[
p(x, y, t) = \sum_k \phi_k(x) \phi_k(y) e^{\lambda_k t},
\]
where $\{\lambda_k\}$ is an enumeration of the eigenvalues of $\Delta = \frac{d^2}{dx^2}$ on $M$ and $\{\phi_k\}$ are the corresponding eigenfunctions which vanish on $\partial M$.
\begin{itemize}
    \item[i)] Calculate $\{\lambda_k\}$ and the corresponding eigenfunctions.
    \item[ii)] Prove that $|p(x, y, t)| \leq C t^{-1/2}$, for all $x, y$, and $0 < t < 1$.
    \item[iii)] What is the exponential rate of decay of $p(x, y, t)$ as $t \to \infty$, i.e., compute:
    \[
    \lim_{t \to \infty} \log(p(x, y, t)).
    \]
\end{itemize}
\end{exercise}
