\chapter{Introduction and Overview}

This chapter provides a systematic overview of the six mathematical disciplines in the S.-T. Yau College Student Mathematics Contest (2010--2025), with a topic-by-topic breakdown for every year and every subject.

\section{Contest Structure Overview}
\label{sec:overview}

The contest consists of six tracks, each representing a fundamental area of mathematics. Below is the number of questions per year for each track.

\begin{itemize}
    \item \textbf{Algebra \& Number Theory}: 2010: 12, 2011: 12, 2012: 12, 2013: 11, 2014: 12, 2015: 11, 2016: 5, 2017: 5, 2018: 5, 2019: 10, 2020: 6, 2021: 6, 2022: 6, 2023: 6, 2024: 6, 2025: 5
    \item \textbf{Analysis \& Differential Equations}: 2010: 14, 2011: 12, 2012: 12, 2013: 6, 2014: 12, 2015: 12, 2016: 6, 2017: 6, 2018: 6, 2019: 6, 2020: 6, 2021: 6, 2022: 6, 2023: 6, 2024: 6, 2025: 6
    \item \textbf{Geometry \& Topology}: 2011: 12, 2012: 12, 2013: 6, 2014: 12, 2015: 12, 2016: 12, 2017: 12, 2018: 6, 2019: 10, 2020: 6, 2021: 6, 2022: 6, 2023: 6, 2024: 6, 2025: 6
    \item \textbf{Probability \& Statistics}: 2010: 5, 2011: 5, 2012: 6, 2013: 5, 2014: 10, 2015: 10, 2016: 5, 2017: 5, 2018: 5, 2019: 8, 2020: 7, 2021: 6, 2022: 5, 2023: 5, 2024: 6, 2025: 5
    \item \textbf{Applied \& Computational Mathematics}: 2011: 11, 2012: 11, 2013: 6, 2014: 9, 2015: 10, 2016: 5, 2017: 5, 2018: 4, 2019: 10, 2020: 6, 2021: 6, 2022: 6, 2023: 6, 2024: 6, 2025: 6
    \item \textbf{Mathematical Physics}: 2022: 6, 2023: 6, 2024: 6, 2025: 6
\end{itemize}

\medskip
\textbf{Notable structural changes:}
\begin{itemize}
    \item \textbf{2010--2016}: Algebra and Analysis featured 10--14 questions (Individual + Team rounds).
    \item \textbf{2017--2025}: Most subjects reduced to 5--6 questions per year (single Exam round).
    \item \textbf{Physics}: First introduced in 2022.
\end{itemize}

\section{Algebra \& Number Theory}
\label{sec:algebra}

Algebra and Number Theory problems in the Yau Contest span from 2010 to 2025, covering a rich spectrum of topics including group theory, representation theory, field theory, linear algebra, and number theory. The problems range from concrete computational exercises to abstract theoretical proofs, reflecting the deep interconnectedness of modern algebra.

\subsection{Overview of Topics}

The algebra problems can be categorized into several interconnected areas:

\medskip
\noindent\textbf{Group Theory and Representations.} This is the most heavily-tested area, appearing in nearly every year. Key topics include:
\begin{itemize}
    \item Classification of finite groups (groups of order 26, 99, 150, etc.)
    \item Sylow theorems and subgroup analysis (GL$_n(\mathbb{F}_p)$, $SL_2(\mathbb{F}_p)$)
    \item Representation theory of finite groups (characters, irreducible representations, $S_n$ and $SL_2(\mathbb{C})$)
    \item Galois groups as subgroups of $S_n$ (especially $S_3$, $S_5$, $D_4$, $V_4$)
    \item Semi-direct products and group extensions ($\mathbb{Z}/p\mathbb{Z} \ltimes (\mathbb{Z}/q\mathbb{Z})^n$)
\end{itemize}

\medskip
\noindent\textbf{Field Theory and Galois Theory.} Core topics include:
\begin{itemize}
    \item Splitting fields and Galois groups (cyclotomic fields, $x^4-2$, $x^5-80x+5$)
    \item Kummer theory and biquadratic extensions
    \item Discriminants and ramification in number fields
    \item Artin-Schreier and Artin-Schreier-Witt extensions
    \item Infinite Galois theory (maximal extensions, profinite groups)
\end{itemize}

\medskip
\noindent\textbf{Linear Algebra and Matrix Theory.} Frequently tested concepts:
\begin{itemize}
    \item Orthogonal and symplectic groups ($O(V)$, $SO(V)$, $Sp(V)$)
    \item Skew-symmetric matrices and their exponentials
    \item Positive semidefinite matrices (Hankel type, $A_{ij} = t^{a_i+a_j}$)
    \item Characteristic and minimal polynomials
    \item Simultaneous triangularization of commuting matrices
\end{itemize}

\medskip
\noindent\textbf{Number Theory.} Algebraic number theory dominates:
\begin{itemize}
    \item Algebraic integers and rings of integers
    \item $p$-adic numbers and valuations ($\mathbb{Z}_p$, $\mathbb{Q}_p$)
    \item Cyclotomic polynomials and fields
    \item Diophantine equations (Mordell curves, Pell equations)
    \item Quadratic reciprocity and Fibonacci numbers
    \item Chevalley-Warning theorem on finite fields
\end{itemize}

\medskip
\noindent\textbf{Commutative Algebra and Algebraic Geometry.}
\begin{itemize}
    \item Integral extensions and normalization
    \item Noetherian rings and Rees algebras
    \item Coordinate rings ($k[x,y]/(y^2-x^3)$, $k[x,y]/(x^2-y^3)$)
    \item Hilbert's Theorem 90 (additive and multiplicative forms)
    \item Dedekind domains and ideal factorization
\end{itemize}

\subsection{Problems by Year}

\begin{tabular}{p{1.2cm}p{13.5cm}}
\textbf{2010} & Commutative operators and common eigenvectors; Matrix representation of $\mathbb{C}$; Conjugacy class ratio $\overline{c}(G) \leq 5/8$; Splitting field of $x^8-5$; Mason-Stothers theorem; $SL_2(\mathbb{Z})$ generators; Irreducible $x^5-80x+5$ with Galois group $S_5$; $SL_2(\mathbb{C})$ representations and Clebsch-Gordan decomposition. \\[4pt]
\textbf{2011} & $S_3$, $\mathbb{Z}/4\mathbb{Z}$, $Q_8$ extensions containing $\mathbb{Q}(\sqrt{-3})$; 2D representations with eigenvalue 1; Real algebraic numbers field automorphism; $T$-invariant subspaces; Groups of order 150 (not simple); Bezout's theorem for curves; PID submodules; Ring isomorphism $\mathbb{Z}[x,y]/(x^2-y^n)$; Harmonic sum non-integer; $GL_3(\mathbb{F}_7)$ Sylow subgroups. \\[4pt]
\textbf{2012} & Eisenstein irreducibility criterion; Order and Sylow subgroups of $GL_n(\mathbb{F}_p)$; Orthogonal group invariants $(V \otimes V)^{O(V)}$; Weyl algebra on polynomials; Lattices in $\mathbb{C}$; Cauchy determinant identity; Reflections and orthogonal group (Cartan-Dieudonn\'e); $\pi_1(SL_2(\mathbb{R}))$; $p$-adic integers (Teichm\"uller lift, $\mathbb{Z}_p^\times \cong \mathbb{Z}_p \times \mathbb{Z}/(p-1)\mathbb{Z}$). \\[4pt]
\textbf{2013} & Groups of order 26 (classification); Quiver representations (Orpheus problem); Irreducible cubics over $\mathbb{F}_5$; $SL_2(\mathbb{R})$ representations ($V_k = \mathrm{Sym}^k$, Clebsch-Gordan); Polynomials with roots mod $p$ but not in $\mathbb{Q}$ (Chebotarev); Skew-symmetric matrices $e^A$ orthogonal; Trace in finite extensions; $SL_2(\mathbb{F}_p)$ element orders; $S_4$ character table; Finite field polynomial counting; Galois group of $x^4-2$. \\[4pt]
\textbf{2014} & Scalar action in representations $H = \{h : h|_V = \eta(h)I\}$; PSD matrices $A_{ij} = t^{a_i+a_j}$; Pell-type equation $X^2-82Y^2=\pm2$; Goursat's lemma on $S \times T$ normal subgroups; Chevalley-Warning theorem; Commuting matrices $\det(A^2+B^2) \geq 0$; $SL_2(\mathbb{R})$ action on hyperbolic plane; Trigonometric product $\cos(\pi/7)\cos(2\pi/7)\cos(3\pi/7)=1/8$; Groups of order 8 classification; $SL_2(\mathbb{C})$ tensor product decomposition. \\[4pt]
\textbf{2015} & Bilinear forms on finite abelian groups (symplectic pairings); $O_n$-invariants on $M_{n \times k}$; Galois group of $x^3+x+1$ (cubic resolvent); $p$-adic contraction mapping; UFD verification; Orthogonal group generated by reflections; Semi-direct products $(\mathbb{Z}/p\mathbb{Z}) \ltimes (\mathbb{Z}/q\mathbb{Z})^n$; Algebraic integer $\zeta = 1 + N\eta$ forces $\zeta=1$; Molien's theorem on invariants; Normal operators and spectral theorem; $x^5-80x+5$ irreducible with Galois group $S_5$. \\[4pt]
\textbf{2016} & Stieltjes matrices (Gram matrix with non-positive off-diagonal); Subgroups of $\mathbb{Z}^d$ (Hermite normal form); Algebraic integers from traces ($\mathrm{tr}(A^n)$ integral implies eigenvalues integral); Euclidean isometries and reflections; Groups of order $2^n m$ with element of order $2^n$; $K=AB$ with $A$ orthogonal, $B$ upper triangular; Binomial coefficients over $\mathbb{Z}$; Galois group of $x^4-x^2-1$ is $D_4$; Fitting's lemma for endomorphisms; Finite groups with $h$ conjugacy classes. \\[4pt]
\textbf{2017} & Pr\"ufer groups $(p^{-k}\mathbb{Z}_p)/\mathbb{Z}_p$; Fitting decomposition for finite abelian groups; Infinite Galois theory (maximal proper subextensions); Chevalley-Warning; Matrix commutation $UV \neq VU$; Chinese remainder for cyclic groups; $p$-adic operators and fixed points; Lie algebra rank and nullcone; Function field extensions $\mathbb{F}_p(t)$; Numerical semigroups $\mathbb{N}(a,b)$. \\[4pt]
\textbf{2018} & Mersenne and Fermat numbers (coprimality); $\mathbb{Q}(\sqrt[3]{5})$ Galois closure and decomposition law; Groups of order 99 are abelian; Vector spaces over finite/infinite fields; $p$-adic valuation matrix structure; Local rings and Fitting's lemma; Binomial coefficient bounds; Cyclotomic polynomials mod $p$; Character values $B(\chi)$; Extension of linear maps. \\[4pt]
\textbf{2019} & Rational-valued characters imply $g \sim g^k$ for $(k,|g|)=1$; Function field Galois theory ($\mathbb{F}_p(t)$); Integral extensions ($\mathfrak{p}'$ maximal iff $\mathfrak{p}' \cap R$ maximal); $GL_n(\mathbb{C})$ path components; Fibonacci numbers and quadratic reciprocity; Centralizer in $S_n$ for double transposition; End$(V)^{G \times G}$ invariants; Noetherian Rees algebras; Gauss sums and quadratic subfields of cyclotomic fields; Hilbert's Theorem 90 additive form. \\[4pt]
\textbf{2020} & Newton's identities on symmetric polynomials; Nakayama's lemma and Cayley-Hamilton; Algebraic curves $y^2 = x^3+x$ (elliptic curve normalization); Cyclotomic polynomial irreducibility; $p$-adic power maps ($x \in (1+p\mathbb{Z}_p)^p \iff x \in 1+p^2\mathbb{Z}_p$); Fermat's Last Theorem in $\mathbb{Z}_p$; Spherical completeness of $\mathbb{Q}_p$ and extensions. \\[4pt]
\textbf{2021} & Artin-Schreier polynomial $x^p-x-a$ irreducibility; Galois group of $x^3-3x+1$ is $\mathbb{Z}/3\mathbb{Z}$; Coordinate ring $k[x,y]/(x^2-y^3)$; Sum of square roots $\sqrt{p_1}+\cdots+\sqrt{p_n}$ irrational; Mordell curve $x^2+13=y^3$ (class number 2); Unramified $p$-adic extensions and Teichm\"uller lifts; $\overline{\mathbb{Q}_p}$ not complete. \\[4pt]
\textbf{2022} & Rational root theorem; Invertible fractional ideals are projective; $\mathfrak{sl}(4,\mathbb{C}) \cong \mathfrak{so}(6,\mathbb{C})$ via $\Lambda^2 V$; Ideal generation in rings of integers ($2$-generated ideals); Cyclotomic fields ($\Phi_p$ minimal polynomial, discriminant, ring of integers); Discriminant of cubic fields and integral bases. \\[4pt]
\textbf{2023} & $SL_n(F)$ generated by elementary matrices; Solvability of $GL_n(F)$; Projective = locally free for finitely presented flat modules; Profinite groups (open subgroups of index $n$); Finite presentation and Schanuel's lemma; $p$-adic matrices (roots of unity); Continuous representations of $\mathbb{Z}_p$; Logarithm map on $K^\times$ and $K^\times/K^{\times,n}$ finite. \\[4pt]
\textbf{2024} & Faithful representations (regular rep); Faithful 1D reps iff cyclic; Abelian groups: $n$-dimensional faithful rep iff generated by $n$ elements; Finite subgroups of $GL_2(\mathbb{R})$ are $C_n$ or $D_{2n}$; Cartan and Iwasawa decompositions in $GL_n$; Imperfect fields ($x^p - a$ irreducibility); $k((t))/k(t)$ transcendental; $p$-adic exponential and $a^x$ continuity. \\[4pt]
\textbf{2025} & Kummer theory: $V_4$ extension iff $[F^*:F^{*2}] > 2$; Biquadratic polynomials $X^4+aX^2+b$; Semidirect products $C_e \rtimes C_f$ with $p$-adic Frobenius; Local fields and unramified extensions; Galois descent for unramified $p$-adic extensions; Splitting field of $x^5-x+1$ has Galois group $S_5$.
\end{tabular}

\subsection{Recurring Themes}

Several themes appear repeatedly across multiple years, indicating their fundamental importance:

\medskip
\noindent\textbf{Classification Problems.} Many problems ask for complete classifications: groups of a given order, finite subgroups of $GL_2(\mathbb{R})$, rings of integers in number fields, or irreducible polynomials over finite fields. The Sylow theorems and structure theorems for abelian groups are the primary tools.

\medskip
\noindent\textbf{Character Theory and Representations.} Problems on $S_4$, $S_5$, $SL_2(\mathbb{C})$, and general finite groups test understanding of irreducible characters, direct sum decompositions (Clebsch-Gordan formula), and the relationship between algebraic properties and representation-theoretic data.

\medskip
\noindent\textbf{Galois Theory and Field Extensions.} Computing Galois groups of polynomials ($x^3-3x+1$, $x^4-2$, $x^5-80x+5$), determining splitting behavior of primes in number fields, and understanding the structure of cyclotomic fields appear frequently.

\medskip
\noindent\textbf{Matrix Analysis.} Problems involving $GL_n(\mathbb{F}_p)$ Sylow subgroups, orthogonal and symplectic groups, PSD matrices, and $p$-adic matrix valuations demonstrate the importance of concrete linear algebra computations.

\medskip
\noindent\textbf{$p$-adic Numbers and Valuations.} The $p$-adic absolute value, $\mathbb{Z}_p$, Teichm\"uller lifts, and the logarithm map on $1+p\mathbb{Z}_p$ form a recurring topic in later years.

\medskip
\noindent\textbf{The Chevalley-Warning Theorem.} This result on zero sets of polynomials over finite fields appears in 2014 and 2017, demonstrating its utility in combinatorial number theory.

\subsection{Structural Changes Over the Years}

The problems show some evolution in emphasis over the 16-year span:

\begin{itemize}
    \item \textbf{Early years (2010-2014):} More classical algebra: group classification, classical Galois theory, elementary number theory (Mersenne/Fermat numbers), and concrete matrix calculations.

    \item \textbf{Middle years (2015-2019):} Greater abstraction: integral extensions, Noetherian rings, infinite Galois theory, profinite groups, and representation-theoretic invariants. The problems require more sophisticated algebraic machinery.

    \item \textbf{Recent years (2020-2025):} Increasingly technical: $p$-adic analysis, DVR theory, local fields, descent arguments, and advanced representation theory over finite fields. The problems integrate multiple theoretical frameworks.
\end{itemize}

Despite these shifts, the core themes remain constant: finite group theory, field extensions, linear algebra, and algebraic number theory form the enduring foundation of the Algebra \& Number Theory portion of the contest.

\section{Analysis \& Differential Equations}
\label{sec:analysis}

\subsection{Subject Overview}

The Analysis and Differential Equations section of the Yau College Student Mathematics Contests encompasses a broad spectrum of topics spanning real analysis, complex analysis, functional analysis, and the theory of ordinary and partial differential equations. The problems are designed to test deep understanding of fundamental principles, creative problem-solving abilities, and the capacity to connect ideas across different areas of mathematics. The contest format typically presents problems that require both rigorous theoretical reasoning and computational skill.

\subsection{Topics and Problems by Year}

\begin{description}

\item[2010] The contest featured problems on Jensen's inequality applied to $\sin x / x$, Fresnel integrals, and the characterization of continuity points as $G_\delta$ sets. Complex analysis appeared through Blaschke products with double zeros and the Poisson integral formula for the half-plane. Topics in functional analysis included Hilbert-Schmidt operators, Lyapunov stability theory for perturbed harmonic oscillators, and the Steinhaus theorem on difference sets. The team problems explored Fourier coefficient convergence in $L^1$ and $L^2$ spaces and uniqueness results for elliptic PDEs of the form $\Delta u + f(u) = 0$.

\item[2011] This year emphasized contour integration and residue calculus with problems involving integrals of rational-trigonometric functions. The conformal mapping problem required constructing a map from a slit plane to the unit disk. Theoretical analysis appeared through the Landau-Kolmogorov inequality relating derivatives to function values, the Riesz representation theorem for positive linear functionals, and Sobolev space embeddings with compactness results. The team section covered $H^2(\Delta)$ as a Hilbert space, first-order PDEs via the method of characteristics, Diophantine approximation with measure-zero sets, matrix group properties ($sl(n)$, $SL(n)$), and proofs of the irrationality and transcendence of $e$ and $\pi$.

\item[2012] Problems included contour integrals with branch cuts for the expression $x^p/(1+x^2)$, conformal mapping of a lune to the unit disk, and a measurable selection theorem for fiber counting functions. The functional analysis problems established the $L^p$ duality via H\"older's inequality, derived the Poisson integral formula for the half-space, and studied Gaussian integrals with positive definite matrices. Topics in complex analysis covered biholomorphic automorphisms of simply connected domains and the Cantor-like set of numbers avoiding the digit 4. Team problems addressed tail behavior of $L^p$ functions, orthonormal bases in Hilbert spaces, and the Volterra operator showing compactness without eigenvalues.

\item[2013] This year featured the layer-cake representation for integrals, residue calculations for polynomials, and Weyl's equidistribution theorem for irrational rotation. The Gauss-Lucas theorem on the convex hull of polynomial roots appeared alongside the Helmholtz decomposition for vector fields. Problems on operator theory included the spectral theorem for compact self-adjoint operators, while real analysis covered convex functions and Legendre transforms. The fractional integral operator problem required proving boundedness and compactness of integral operators with kernel $|x-y|^{-d+\alpha}$.

\item[2014] The Hyers-Ulam stability problem for Cauchy's functional equation opened the competition. Conformal mapping appeared through the modulus invariance for annuli, while the Poisson kernel problem required proving boundary convergence in the upper half-plane. The team section presented the Heisenberg uncertainty principle via Fourier analysis, reproducing kernel Hilbert spaces (Bergman spaces), the Poincar\'e recurrence theorem for volume-preserving maps, and the strong maximum principle for subharmonic functions. Problems also included Thomae's function discontinuities and approximation to identity via mollifiers.

\item[2015] Fatou's lemma techniques for $L^2$ convergence, the moment problem showing that vanishing integrals imply zero functions, and Liouville-type theorems for entire functions with growth restrictions featured prominently. Finite Blaschke products were characterized, and functional analysis problems addressed Fredholm operators with Atkinson's theorem. The team section covered Sobolev space embeddings, the fundamental lemma of calculus of variations, and Euler-Lagrange equations. Additional topics included $L^1$ continuity under translation, Cartan's uniqueness theorem for holomorphic automorphisms, and the Dirichlet problem with Poincar\'e inequality.

\item[2016] The year opened with Lebesgue's fundamental theorem on absolute continuity, Gaussian approximation to identity, and Riemann's integrability criterion. Conformal mapping appeared through the transformation $w = \frac{1}{2}(z + 1/z)$ mapping the exterior of the unit circle to $\mathbb{C} \setminus [-1,1]$. Problems covered doubly periodic meromorphic functions, spectral properties of self-adjoint operators, and removable singularities for holomorphic functions. The team section addressed convergence of subsequences in $L^p$, uniform convergence of derivatives of holomorphic functions, automorphisms of the torus ($SL(2,\mathbb{Z})$), Volterra-type operators, and heat kernel expansion on the unit interval.

\item[2017] Abel summability of Fourier series and weak convergence criteria in Hilbert spaces were examined. Holomorphic function growth restrictions near zero led to discussions of removable singularities. The intersection of two disks required careful conformal mapping. Harnack's inequality for positive harmonic functions and the Poincar\'e inequality on rectangles appeared in the team problems, alongside uniqueness theorems for zero Fourier coefficients, Dini derivatives, Schwarz lemma for Blaschke products, branch cuts for $\sqrt{1-z^2}$, Laplace equation invariance under orthogonal transformations, and Hilbert-Schmidt operators.

\item[2018] Weak convergence in $L^2$ spaces via the Banach-Alaoglu theorem and characterization of finite-dimensional subspaces opened the competition. Maximum modulus principles, convergence of harmonic functions, and upper semicontinuity of measures were explored. The Newton potential and B\^ocher's theorem for positive harmonic functions on punctured space appeared. Team problems addressed subsequence convergence in $L^2$, Paley-Wiener theorems for Schwartz functions, Lebesgue differentiation theorems, completeness of $C[0,1]$ with the $L^1$ metric, and Green's function representation via conformal mapping.

\item[2019] Convex function inequalities (Jensen-type), Gagliardo-Nirenberg-Sobolev inequalities, and Liouville's theorem for harmonic functions with logarithmic bounds demonstrated the year's focus on inequalities. The primitive existence problem for $(z^2-1)^{-2019}$ required high-order residue calculations. Subcritical elliptic PDEs ($\Delta u = |u|^{q-1}u + f$) were solved via direct methods. Team problems covered Paley-Wiener shifts, interior Schauder estimates, ergodic theory (Kronecker's theorem), infinite product construction for entire functions with controlled growth, and finite speed of propagation for semilinear wave equations.

\item[2020] Approximate identities and the Lebesgue differentiation theorem opened the year, followed by Fourier coefficient decay analysis tied to jump discontinuities (featuring Tintin, Haddock, Dupont, and Dupond). Caccioppoli inequalities for elliptic equations and Picard's theorem for entire curves on hyperbolic surfaces demonstrated advanced PDE techniques. The team section covered periodic orbits in Hamiltonian systems ($x'' + x + x^3 = 0$) and measure-preserving Blaschke products.

\item[2021] The year emphasized $L^2$ harmonic function triviality, uniform convergence in finite-dimensional subspaces, and Steinhaus-type theorems on difference sets. The Hadamard three-lines theorem appeared in the context of bounded holomorphic functions on a strip. Functional analysis problems addressed perturbation methods for elliptic equations with small nonlinearities and periodic solution existence via the Poincar\'e map.

\item[2022] Probabilistic methods appeared through the harmonic mean of uniform random variables. Idempotent holomorphic mappings ($f \circ f = f$) led to the identity theorem. Szemer\'edi-type results guaranteed arithmetic progressions in sets of positive measure. The Baire category theorem showed infinite-dimensional polynomial subspaces cannot be closed in $C[0,1]$. Topics included multipole expansions for harmonic functions, global ODE existence with bounded derivatives, and periodic solutions via Poincar\'e maps.

\item[2023] The Lie-Trotter product formula and commutator conditions for matrix exponentials demonstrated the year's analytical-algebraic focus. Shooting methods for boundary value problems with monotonicity conditions appeared. Bochner's theorem on positive definite functions and weakly convergent dilated periodic functions showed the breadth of topics. Additional problems covered growth restrictions on entire functions ($\sqrt{|z|}|\cos z|$), the Feynman-Kac representation for parabolic equations, and space-time estimates for heat equations.

\item[2024] Advanced topics included modulation stability via the implicit function theorem, invertibility of perturbed Schr\"odinger operators using Fredholm theory, bilinear estimates for dispersive equations via Fourier analysis, and the $TT^*$ method for space-time estimates. Bobylev's formula for the Boltzmann collision operator represented the most technical problem of the year.

\item[2025] The Euler-Cauchy differential equation with initial conditions demonstrated classical ODE techniques. Hardy's inequality with optimal constants and Hadamard's three-circles theorem for subharmonic functions represented advanced real and complex analysis. Winding numbers and homotopy classification of curves appeared alongside the Urysohn lemma in metric spaces. BMO (Bounded Mean Oscillation) space properties concluded the competition with sophisticated functional analysis.

\end{description}

\subsection{Recurring Themes}

Several core topics appear repeatedly across multiple years, reflecting their fundamental importance in mathematical analysis:

\begin{itemize}
\item \textbf{Complex Analysis:} Conformal mapping (annuli, slits, intersections of disks), Blaschke products and finite products, harmonic and subharmonic functions, maximum principles, residue calculus, and Riemann surface concepts appear throughout.

\item \textbf{Real Analysis and Measure Theory:} $L^p$ spaces and embeddings, Fourier series and transforms, approximation theory, measure-theoretic convergence theorems, and differentiation of integrals recur frequently.

\item \textbf{Functional Analysis:} Operator theory including compact operators, spectral properties of self-adjoint operators, Hilbert and Banach space theory, reproducing kernel spaces, and fixed-point theorems form a consistent thread.

\item \textbf{ODEs and PDEs:} Existence and uniqueness theory, boundary value problems, stability analysis, elliptic equations, parabolic equations (heat equation), and hyperbolic equations (wave equation) appear across all years.

\item \textbf{Inequalities:} Landau-Kolmogorov, H\"older, Jensen, Cauchy-Schwarz, Hardy, uncertainty principles, and logarithmic convexity demonstrate the central role of estimates in analysis.
\end{itemize}

\subsection{Structural Evolution}

The contest format evolved noticeably over the sixteen-year period. Early years (2010--2016) maintained a clear distinction between \texttt{Individual} and \texttt{Team} sections, with six problems in each section. Beginning around 2017, some corruption in section markers appears in the source files. From 2018 onward, most years adopted a single \texttt{Exam} section, suggesting a transition from team-based to individual examination format.

The mathematical content evolved as well. Earlier years featured more classical topics in complex analysis and conformal mapping. Later years incorporated increasingly sophisticated techniques from harmonic analysis (BMO spaces, Schauder estimates), PDE theory (Feynman-Kac formulas, bilinear estimates), and modern functional analysis (Fredholm operators, modulation spaces). Despite these changes, the core emphasis on rigorous analysis, creative problem-solving, and connections across subdisciplines remained constant throughout the competition's history.

\section{Geometry \& Topology}
\label{sec:geometry}

The \textit{Geometry \& Topology} section of the Yau Contest problems encompasses a rich interplay between
differential geometry and algebraic topology, covering topics from classical manifold theory to modern
Riemannian geometry. Over sixteen years of competition problems (2010--2025), several core themes emerge
consistently, reflecting the fundamental importance of these areas in modern mathematics.

\subsection{Overview of Subject}

Geometry \& Topology studies the properties of spaces that are preserved under continuous deformations.
The field divides roughly into two complementary perspectives:
\begin{itemize}
    \item \textbf{Topology} focuses on global properties preserved under continuous transformations---connectedness,
          holes in various dimensions, and the fundamental group structure that classifies how loops can wrap
          around the space.
    \item \textbf{Geometry} adds metric structure to topology, measuring distances, angles, and curvature.
          The interplay between curvature (local property) and topology (global property) is codified in
          classical theorems such as Gauss-Bonnet, Myer's, and the Sphere Theorem.
\end{itemize}

The competition problems typically require mastery of both viewpoints, often applying topological invariants
to solve geometric problems or using geometric intuition to understand topological phenomena.

\subsection-Problems by Year}

\subsubsection*{2010}
\begin{itemize}
    \item \textbf{Individual:} Hyperbolic metric on punctured disc; compact hypersurfaces and the second
          fundamental form; orientability of $\mathbb{RP}^n$; covering space criteria; symplectic forms
          on $\mathbb{R}^4$ and impossibility on $S^4$; isoperimetric inequality via Wirtinger's inequality;
          Minkowski integral formulas for surfaces.
    \item \textbf{Team:} Conformal maps and stereographic projection; maximum principle on surfaces;
          Whitney embedding theorem; critical points of functions on compact manifolds; construction of
          spaces with specified homology; degree theory for maps $S^2 \to S^2$.
\end{itemize}

\subsubsection*{2011}
\begin{itemize}
    \item \textbf{Individual:} Essential maps to spheres versus tori (highlighting $\pi_n(S^n)$ vs $\pi_n(T^n)$);
          isometries on compact metric spaces; linking number and its invariance; Poincar\'e dual on the
          torus; Bertrand curves and their characterization; Willmore energy for canal surfaces.
    \item \textbf{Team:} Universal cover compactness and $\pi_1$ finiteness; homology of simplex skeletons;
          divergence theorem; Mayer-Vietoris sequence in de Rham cohomology; scalar curvature as
          average of Ricci curvature; Schur's lemma on Einstein manifolds.
\end{itemize}

\subsubsection*{2012}
\begin{itemize}
    \item \textbf{Individual:} Hopf fibration and $\pi_3(S^2) \cong \mathbb{Z}$; Frobenius integrability theorem;
          orientation-preserving homeomorphisms of $\mathbb{CP}^2$; isoperimetric inequality; Minkowski
          formulas; constant curvature metrics via structure equations.
    \item \textbf{Team:} $\mathbb{RP}^n$ as differentiable manifold; rank equals $n$ implies covering map;
          uniqueness of Levi-Civita connection; maps $S^n \to S^n$ with odd degree; Stokes theorem;
          Gauss-Bonnet formula for surfaces with boundary.
\end{itemize}

\subsubsection*{2013}
\begin{itemize}
    \item \textbf{Individual:} Quotient spaces $T^2/\{p,q\}$ and their fundamental groups; isometries on
          compact metric spaces; conjugate points and Jacobi fields; determination of Riemann tensor from
          sectional curvatures; Synge's theorem (positive curvature implies simply connected for even
          dimensions); minimal surfaces and the transformation $K \to -K$ preserving curvature.
    \item \textbf{Team:} Complement of linked circles in $\mathbb{R}^3$; idempotent self-maps and image
          submanifolds; torus diffeomorphisms; Einstein manifolds in dimension 3; Myers' diameter bound;
          total torsion and characterization of spheres and planes.
\end{itemize}

\subsubsection*{2014}
\begin{itemize}
    \item \textbf{Individual:} Quotient spaces from antipodal identifications on spheres; Bernstein's theorem
          on entire minimal graphs; Poincar\'e dual for sloped curves on tori; Riemannian submersions
          and their properties; Willmore inequality for surfaces; $T^1(S^2) \cong \mathbb{RP}^3$.
    \item \textbf{Team:} Wedge sums and fundamental group; orientability of quotients by properly
          discontinuous actions; connected sums $\mathbb{RP}^3 \# \mathbb{RP}^3$; bi-invariant metrics
          on Lie groups and non-negative curvature; completeness of upper half-plane models.
\end{itemize}

\subsubsection*{2015}
\begin{itemize}
    \item \textbf{Individual:} Tangent bundle triviality for $S^n \times S^m$ (Euler characteristic criterion);
          $\mathbb{RP}^2$ not a manifold boundary; flow box theorem; Hopf's theorem (CMC spheres are round);
          Myers' rigidity theorem; normal coordinates expansion of the metric tensor.
    \item \textbf{Team:} $SO(3)$ as $S^3/\{\pm 1\}$; reduced suspension homology; Lie derivative formula
          $[X,Y] = L_X Y$; convex surface rigidity via Minkowski problem; second Bianchi identity;
          Hessian comparison for Riemannian manifolds.
\end{itemize}

\subsubsection*{2016}
\begin{itemize}
    \item \textbf{Individual:} Euler characteristic relations $\chi(M) = \frac{1}{2}\chi(\partial M)$ for odd
          dimensions; de Rham cohomology of punctured torus; Alexandrov (umbilic) theorem for hypersurfaces;
          existence of cutoff functions; Bochner vanishing ($Ric > 0 \implies H^1 = 0$); first eigenvalue
          of minimal hypersurfaces in $S^{n+1}$.
    \item \textbf{Team:} $\mathbb{CP}^{2n}$ covering only itself; reduced suspension; Frobenius theorem;
          great circles as only geodesics on spheres; Riemann curvature bounds from sectional curvature
          range; Simons' identity and gap theorem for minimal hypersurfaces.
\end{itemize}

\subsubsection*{2017}
\begin{itemize}
    \item \textbf{Individual:} Poincar\'e-Hopf theorem (zero Euler characteristic implies non-vanishing
          vector field); cellular homology of spaces $(S^2 \vee S^2) \cup_f D^3$; Lie derivative equals
          Lie bracket; Liouville formula for geodesic curvature; distance as supremum of 1-Lipschitz
          functions; minimal submanifold curvature pinching.
    \item \textbf{Team:} Fundamental group of glued M\"obius bands (Klein bottle); mapping torus homology
          (Wang sequence); geodesic torsion formula; gradient of distance function gives geodesics;
          Minkowski problem rigidity.
\end{itemize}

\subsubsection*{2018}
\begin{itemize}
    \item \textbf{Individual:} Connected sum formulas for $\pi_1$ and homology; maps $S^2 \to S^1 \times S^1$;
          classification of vector bundles over $S^1$; Crofton formula on $S^2$; integral inequality for
          mean curvature of submanifolds; quarter-pinching sphere theorem.
    \item \textbf{Team:} Homology of $(S^2 \times S^2) \cup_{S^2} D^3$; complements of knots and links
          ($\pi_1 \cong \mathbb{Z}$ and free groups); contact forms and non-existence of closed
          multiples; hypersurfaces $x_1^4 + \cdots + x_{n+1}^4 = 1$ as manifolds; total absolute
          Gaussian curvature inequality.
\end{itemize}

\subsubsection*{2019}
\begin{itemize}
    \item \textbf{Individual:} Configuration spaces and Arnold relations in braid group cohomology; signature
          of 4-manifolds ($\mathbb{CP}^2$, $S^2 \times S^2$); fundamental group of $\mathbb{R}^4$ quotients;
          Chern classes and the transgression formula; Bochner's formula for eigenfunctions.
    \item \textbf{Team:} Non-triviality of $TS^2$ (Euler class obstruction); cohomology of $\mathbb{R}^3
          \setminus X$ for points, lines, planes; degree-1 involutions on $T^2$ without fixed points;
          dimensions and $\pi_1$ of $SU(n)$ and $SO(n)$; Killing fields on non-positive curvature manifolds.
\end{itemize}

\subsubsection*{2020}
\begin{itemize}
    \item \textbf{Individual/Team:} Cobordism of $\mathbb{RP}^{2n}$ (non-boundary) versus $\mathbb{RP}^3$
          (boundary); rigidity of $\mathbb{R}^n$ under $Ric \geq 0$ with model ends; homotopy groups of
          $n$-torus ($\pi_1 = \mathbb{Z}^n$, higher groups trivial); mean curvature of planes in
          hyperbolic 3-space; Frankel's theorem (positive curvature forces geodesics to intersect);
          Jordan-Brouwer separation and orientability of embedded hypersurfaces.
\end{itemize}

\subsubsection*{2021}
\begin{itemize}
    \item \textbf{Individual:} Hopf invariant construction for maps $S^{11} \to S^6$; non-existence of
          complete metric on $\mathbb{R}^2$ with $Ric = 1$; identification of Jacobi fields in normal
          coordinates; compactness of $SO(n)$ and isomorphism $SO(3) \cong \mathbb{RP}^3$; $\pi_1(SO(n))$.
    \item \textbf{Team:} Covering map homeomorphism criteria; free group subgroups and Nielsen-Schreier
          formula; H-space fundamental group (Abelian); Hessian bounds for distance functions; area
          bounds for minimal surfaces in negative curvature.
\end{itemize}

\subsubsection*{2022}
\begin{itemize}
    \item \textbf{Individual:} Tangent bundle triviality (Euler characteristic criterion for $S^{2n} \times
          S^{2n}$ vs $S^{2n} \times S^{2n-1}$); Berger metrics on $S^3$ and scalar curvature; homotopy
          groups $\pi_2(U(n))$, $\pi_3(U(n))$ via Bott periodicity; non-immersion of $\mathbb{CP}^2$ into
          $\mathbb{R}^6$ via Pontryagin classes.
    \item \textbf{Team:} Conformal change of Gaussian curvature equation; fixed point theory for
          $\mathbb{CP}^n$ via Lefschetz numbers.
\end{itemize}

\subsubsection*{2023}
\begin{itemize}
    \item \textbf{Individual:} Degree maps $S^{2n} \to \mathbb{CP}^n$ (impossible) versus $\mathbb{CP}^n
          \to S^{2n}$ (possible via quotient); impossibility of embedded closed minimal surfaces in
          $\mathbb{R}^3$; Euler characteristic parity for simply connected 6-manifolds with $H_2 = \mathbb{Z}_2$;
          average Ricci curvature formula; free actions on spheres and group structure ($\mathbb{Z}_2$ for
          even dimensions, no $\mathbb{Z}_p^2$ for odd).
    \item \textbf{Team:} Eigenvalue evolution under metric deformations.
\end{itemize}

\subsubsection*{2024}
\begin{itemize}
    \item \textbf{Individual:} Euler characteristic obstruction ($\chi = 1$) for CW complexes to be
          manifolds; area bounds for minimal surfaces under $sec \leq -1$; symmetric products
          $\mathsf{SP}^n(\mathbb{CP}^1) \cong \mathbb{CP}^n$; geodesic loops to infinity and lines in
          universal covers; H-spaces ($\pi_1$ Abelian) and Adams' theorem on $S^{2022}$; shrinkers
          and the sphere $S^n(\sqrt{2n})$.
    \item \textbf{Team:} Mean curvature flow and不老.
\end{itemize}

\subsubsection*{2025}
\begin{itemize}
    \item \textbf{Individual:} Tangent bundle triviality criteria (same as 2022); Berger metrics
          computation; $\pi_2(U(n)) = 0$, $\pi_3(U(n)) = \mathbb{Z}$ for $n \geq 2$; Pontryagin
          class obstruction to immersing $\mathbb{CP}^2$ in $\mathbb{R}^6$; conformal change equation
          $\Delta u - K + \widetilde{K}e^{2u} = 0$; fixed point theory for $\mathbb{CP}^n$ and
          $\mathbb{CP}^2 \times \mathbb{CP}^2$.
\end{itemize}

\subsection*{Recurring Themes}

Across all sixteen years, certain topics appear with remarkable frequency, indicating their fundamental
importance in the field:

\medskip\noindent\textbf{Algebraic Topology.}
\begin{itemize}
    \item \textit{Fundamental group and homology}: Computing $\pi_1$ and $H_*$ for constructed spaces
          (connected sums, wedges, quotients, mapping tori) appears in nearly every year.
    \item \textit{Covering spaces}: The theory of covering maps, deck transformations, and universal covers
          is tested repeatedly, including the fundamental theorem that $\pi_1$ finiteness implies compact
          universal cover.
    \item \textit{Cohomology operations}: de Rham cohomology, Poincar\'e duality, Mayer-Vietoris sequences,
          and characteristic classes (Stiefel-Whitney, Pontryagin, Chern) appear frequently.
\end{itemize}

\medskip\noindent\textbf{Riemannian Geometry.}
\begin{itemize}
    \item \textit{Curvature comparison theorems}: Myer's diameter bound, Bishop-Gromov volume comparison,
          and Hessian/Ricci comparison appear as standard tools.
    \item \textit{Geodesics}: Geodesic completeness, conjugate points, Jacobi fields, and the geodesic
          equation are perennial favorites.
    \item \textit{Sphere theorems}: Both the classical quarter-pinching theorem and its variations
          (positive curvature $\implies$ homeomorphic to sphere) appear multiple times.
\end{itemize}

\medskip\noindent\textbf{Classical Manifolds.}
The same spaces recur constantly as test subjects:
\begin{itemize}
    \item Spheres $S^n$: Tangent bundles, geodesics, isometries, Hopf fibrations
    \item Real projective space $\mathbb{RP}^n$: Orientability, fundamental group, cobordism
    \item Complex projective space $\mathbb{CP}^n$: cohomology ring, fixed points, covering properties
    \item Tori $T^n$: Fundamental group, homotopy groups, maps of various degrees
    \item Lie groups $SO(n), SU(n), U(n)$: Manifold structure, curvatures, homotopy groups
\end{itemize}

\subsection*{Structural Changes Over the Years}

The competition has maintained remarkable structural consistency while evolving in subtle ways:

\begin{enumerate}
    \item \textbf{Format stability}: Both Individual and Team exams consistently contain 6 problems each,
          allowing comprehensive coverage of the syllabus.

    \item \textbf{Topic balance}: While algebraic topology and Riemannian geometry maintain equal
          prominence, recent years (2022--2025) show increased emphasis on:
          \begin{itemize}
              \item Characteristic class calculations (Pontryagin classes for immersion obstructions)
              \item Modern analytical methods (eigenvalue evolution, mean curvature flow)
              \item Sophisticated topological constructions (Hopf invariant, Arnold relations)
          \end{itemize}

    \item \textbf{Pedagogical depth}: Earlier years (2010--2014) tended toward classical theorems with
          established proofs, while later years incorporate more specialized results (Adams' Hopf invariant
          one theorem, Sormani's geodesic loops to infinity, Simons' gap theorem).

    \item \textbf{Interdisciplinary connections}: Problems increasingly bridge geometry/topology with
          other areas: the Willmore energy connects to physics (Plateau problem), shrinkers connect
          to geometric analysis (mean curvature flow), and H-spaces connect to homotopy theory.
\end{enumerate}

\section{Probability \& Statistics}
\label{sec:probability}

The Probability and Statistics section of the S.-T. Yau College Student Mathematics Contests spans 16 years of competition problems (2010--2025), presenting a comprehensive survey of fundamental concepts, classical theorems, and modern applications in probability theory and statistical inference.

\subsection{Overview of Subject}

The problems in this collection cover the two main branches of the subject:
\begin{itemize}
    \item \textbf{Probability Theory}: The study of random phenomena, including discrete and continuous distributions, limit theorems, stochastic processes, and probabilistic inequalities.
    \item \textbf{Statistical Inference}: Methods for drawing conclusions from data, including point estimation, confidence intervals, hypothesis testing, and asymptotic theory.
\end{itemize}

The contest problems emphasize rigorous mathematical reasoning, often requiring clever applications of classical results rather than routine calculations.

\subsection-Problems by Year}

\textbf{2010}: Conditional expectation for normal variables; Weak Law of Large Numbers for pairwise independent sequences; chi-squared distributions and quadratic forms; MLE confidence sets; SVD and rank-k approximation; order statistics and Beta distribution; foul shooting problem (MLE); numerical stability analysis; hyperbolic PDE well-posedness.

\textbf{2011}: Orthogonal polynomials and Chebyshev polynomials; Lagrange interpolation; hypergeometric and binomial distributions (urn models); Jensen's inequality for i.i.d. variables; Bernoulli distribution, UMVU estimation, Cram\'er-Rao bound; normal error model, Fisher information, ancillary statistics; Box-Muller transform; median minimization property; location-scale Bayesian inference; Poisson distribution and UMVUE.

\textbf{2012}: P\'olya's urn model; conditional expectation properties; Radon-Nikodym derivative; independence and sigma-algebras; joint distribution factorization; Wilcoxon and Mann-Whitney statistics; symmetric distributions; exponential and F-distributions; Borel-Cantelli lemma; geometric-exponential mixture.

\textbf{2013}: Borel-Cantelli and convergence in probability; truncated bivariate normal; record values and order statistics; p-value properties; uniform distribution and confidence intervals; enemy tank problem (capture-recapture); characteristic functions; Diophantine approximation; double exponential prior and posterior; Lasso regularization.

\textbf{2014}: Stein's lemma (normal characterization); Khinchin's theorem for uncorrelated sequences; Radon-Nikodym derivative; maximum of uniform samples; multinomial MLE; Slutsky's theorem; ratio of normals; independence and conditioning; shifted exponential MLE.

\textbf{2015}: Correlation bounds for bivariate normals; expected waiting time for consecutive heads; Gaussian sequences convergence in $L^p$; exponential distribution, consistent estimation, efficiency; normal heteroscedastic model; airplane seating problem; random indexing convergence; Dirichlet and exponential relationships; covariance and joint normality; Lasso oracle properties.

\textbf{2016}: Non-backtracking random walk on $\mathbb{Z}^2$; random matrix operator norm (moment method); variance of binomial count; Jensen Kantorovich inequality; SLLN for random walks; random walk on binary tree; median-of-means concentration; modular arithmetic and fair games; Borel-Cantelli applications; random points on circle.

\textbf{2017}: Bayesian updating (submarine search); random walk last vertex on cycle; gift exchange party (Poisson limiting distribution); multivariate normal conditional independence (precision matrix); $t$-distribution derivation; hypercube concentration measure; convergence with bounded expectations; dollar game random walk; simple random walk intersection times; Poisson zero proportion estimation, MLE, asymptotic relative efficiency.

\textbf{2018}: Hypergeometric remaining balls problem; binary tree signal-to-noise threshold; SLLN for coin flips; Erd\H{o}s--R\'enyi clique size concentration; SIS epidemic model PDE; order statistics spacing asymptotics; light bulb toggling process; convex optimization and exponential families; histogram density estimation; empirical Bayes normal means.

\textbf{2019}: Random variable growth rate ($\limsup X_n^{1/n}$); Markov chain irreducibility and recurrence; random power series zeros (Salem-Zygmund); Fisher's exact test, Neyman-Rubin causal model, ATE estimation; SLLN convergence implications; harmonic number approximation; Erd\H{o}s--R\'enyi isolated vertices phase transition.

\textbf{2020}: Hoeffding inequality for bounded variables; Stein's lemma for Poisson characterization; gambler's ruin hitting probability; rotational invariance implies joint normality; re-randomization for causal inference; odd sample size limitations.

\textbf{2021}: Gaussian convergence preservation; total variation distance and coupling (maximal coupling); pattern waiting times in coin sequences; Markov chain contraction mapping; linear model UMVUE; heteroscedastic normal minimax estimation.

\textbf{2022}: Optional stopping for gambler's ruin; Bernoulli supremum distribution; uniform digit distributions; GLM MLE convergence rates; median estimation in two-way layout.

\textbf{2023}: Moment convergence under uniform integrability; Markov chain transition inequalities; Bernoulli SLLN rates; random matrix singularity probability; clinical trial randomization; propensity score and IPW estimation.

\textbf{2024}: Non-transitive dice constructions; martingale convergence; record time asymptotics; regular conditional probability; multinomial stopping boundaries; Poisson hidden parameter estimation.

\textbf{2025}: Single observation confidence intervals; minimax decision theory; non-transitive random variables (Trybu\l a paradox); one-way ANOVA with M-estimators.

\subsection{Recurring Themes}

Several topics appear repeatedly across multiple years, reflecting their fundamental importance in probability and statistics:

\begin{enumerate}
    \item \textbf{Limit Theorems}: Borel-Cantelli lemma, Strong Law of Large Numbers, Central Limit Theorem, and related results appear in nearly every year (2010, 2012, 2013, 2015, 2017, 2018, 2019, 2020, 2021, 2023).

    \item \textbf{Conditional Expectation and Martingales}: Properties of conditional expectation, martingale convergence, and applications appear in 2010, 2012, 2013, 2014, 2019, 2020, 2022, 2024.

    \item \textbf{Normal Distribution}: Problems involving the normal distribution (characterization, conditional distributions, $t$-distribution derivations) appear in 2010, 2011, 2013, 2014, 2015, 2017, 2020, 2021.

    \item \textbf{Estimation Theory}: UMVU estimation, Cram\'er-Rao bound, MLE, and asymptotic distributions appear in 2010, 2011, 2012, 2013, 2014, 2015, 2018, 2021, 2022, 2025.

    \item \textbf{Stochastic Processes}: Random walks, Markov chains, and branching processes appear in 2010, 2016, 2017, 2018, 2019, 2020, 2022.

    \item \textbf{Inequalities}: Probabilistic inequalities including Jensen, Cauchy-Schwarz, Hoeffding, and concentration inequalities appear in 2010, 2011, 2015, 2016, 2020, 2023.

    \item \textbf{Bayesian Statistics}: Prior selection, posterior distributions, and non-informative priors appear in 2011, 2013, 2017, 2018, 2024.

    \item \textbf{Causal Inference}: The Neyman-Rubin potential outcomes framework, randomization inference, and IPW estimation appear in 2017, 2019, 2020, 2023.
\end{enumerate}

\subsection{Structural Changes Over the Years}

\begin{enumerate}
    \item \textbf{Problem Structure}: Earlier years (2010--2016) typically had 5 individual problems plus 5 team problems per year. Starting around 2017, the structure became more variable, with some years having Part I (Probability) and Part II (Statistics) divisions.

    \item \textbf{Modern Topics}: Later years show increased interest in:
    \begin{itemize}
        \item Machine learning-related statistics (Lasso, empirical Bayes, M-estimators) in 2013, 2015, 2017, 2022, 2025
        \item Causal inference and experimental design in 2017, 2019, 2020, 2023
        \item Network models and random graphs in 2018, 2019, 2023
        \item High-dimensional statistics in 2013, 2015, 2022
    \end{itemize}

    \item \textbf{Computational Topics}: Some problems incorporated numerical analysis and algorithms (2010: stability of schemes; 2018: histogram estimation; 2024: EM algorithm).

    \item \textbf{Applied Problems}: Real-world applications became more prominent, including clinical trial design (2017, 2023), epidemic models (2018), search problems (2017 submarine), and game theory (2017, 2018, 2022).
\end{enumerate}

\subsection{Key Mathematical Techniques}

Students are expected to master a variety of techniques:
\begin{itemize}
    \item \textbf{Characteristic functions and generating functions} for distribution analysis
    \item \textbf{Moment generating functions} for limit theorems
    \item \textbf{Change of variables} and Jacobian calculations
    \item \textbf{Exponential family} theory and completeness
    \item \textbf{Bayesian decision theory} and risk evaluation
    \item \textbf{Stochastic ordering} and coupling arguments
    \item \textbf{Orthogonal polynomial} constructions
    \item \textbf{Concentration inequalities} (Hoeffding, Markov, Chebyshev)
\end{itemize}

This collection provides a thorough survey of classical and modern probability and statistics, requiring mastery of both theoretical foundations and creative problem-solving skills.

\section{Applied \& Computational Mathematics}
\label{sec:applied}

The Applied \& Computational Mathematics category of the S.-T. Yau College Student Mathematics Contests encompasses a rich and diverse collection of problems spanning numerical analysis, optimization, linear algebra, differential equations, and their applications. From 2011 to 2025, the contest has consistently tested contestants on both theoretical understanding and practical computational skills, reflecting the essential role of computational mathematics in modern science and engineering.

\subsection{Overview of Subject Matter}

The problems in this category can be broadly classified into several interconnected areas:

\begin{itemize}
\item \textbf{Numerical Analysis of PDEs:} A substantial portion of each year's problems concerns the discretization and numerical solution of partial differential equations, including finite difference, finite element, and discontinuous Galerkin methods. Stability analysis, consistency, and convergence (via the Lax Equivalence Theorem) are recurring themes.

\item \textbf{Iterative Methods for Linear Systems:} The conjugate gradient (CG) method, GMRES, Arnoldi iteration, and various stationary iterative methods appear frequently. These problems often involve properties of Krylov subspaces, spectral analysis, and convergence rates.

\item \textbf{Optimization and Convex Analysis:} Problems on gradient descent, strong convexity, descent lemmas, co-coercivity, and metric projections reflect the importance of optimization in computational mathematics.

\item \textbf{Orthogonal Polynomials and Approximation Theory:} Chebyshev, Legendre, Hermite, and other orthogonal polynomials appear in problems involving Gaussian quadrature, polynomial interpolation, and best approximation.

\item \textbf{Linear Algebra:} Eigenvalue problems, singular value decomposition (SVD), condition numbers, and matrix perturbation theory are perennial favorites.

\item \textbf{Differential Equations:} Both ODEs (including stiff systems) and PDEs (wave, heat, Schr\"odinger, Fokker-Planck equations) are well represented.

\item \textbf{Applications:} Problems from physics and biology---enzyme kinetics (Michaelis-Menten), projectile motion scaling, neuron dynamics (FitzHugh-Nagumo), and Stokes drag---demonstrate the practical relevance of the methods studied.
\end{itemize}

\subsection{Topics and Problems by Year}

\medskip
\noindent\textbf{2011.} The inaugural year established several recurring themes. Individual problems covered: the half-angle formula with Catalan numbers (generating functions); efficient gcd computation for polynomials via modular exponentiation; $L^2$ stability of the semi-discrete upwind scheme for conservation laws; Richardson's iteration with optimal step size; and energy estimates for backward Euler discretization of the heat equation. The team problems focused on the conjugate gradient method (optimal step choice, $A$-orthogonality of search directions, Krylov subspaces) and the Beam-Warming scheme via Taylor expansion.

\medskip
\noindent\textbf{2012.} This year emphasized numerical integration and approximation. Individual problems included: determining constants for a two-point integration formula (a precursor to Gaussian quadrature with one fixed node); the moving least squares approximation and its error analysis; and analysis of the FTCS scheme for $u_t + u_x = 0$ (notably found to be unconditionally unstable). The team problems covered: efficient RSA factorization for twin primes; superconvergence of FEM for 1D Poisson (exactness at nodes); and similarity solutions for the porous medium equation $h_t = \Delta(h^2)$.

\medskip
\noindent\textbf{2013.} Problems explored weak solutions and analytical techniques. Individual work included: weak solutions of the wave equation (distributional sense); finite time extinction and hyper-contractivity for ODEs ($y' \leq \alpha - \beta y^a$); nondimensionalization of Stokes drag; and error estimates for $L^2$ projection into piecewise polynomial spaces. Team problems covered: scaling laws and slowly varying functions; Fokker-Planck, Witten Laplacian, and Schr\"odinger operators (showing eigenvalue equivalence); and energy conservation for a wave guide problem $u_t + u_x = 0$, $v_t - v_x = 0$.

\medskip
\noindent\textbf{2014.} A particularly rich year with diverse topics. Individual problems included: complete monotonicity of moment sequences (Bernstein's theorem); Eisenstein polynomials and multivariate generalizations; $L^2$ stability of the discontinuous Galerkin method for conservation laws; the GMRES algorithm and Arnoldi process; solving $x_k = 2x_{k-1} + b_k$ and condition number analysis; Gauss-Chebyshev quadrature; the uncertainty principle $\|x\|_0 \|\hat{x}\|_0 \geq N$; eigenvalues of an augmented system matrix $\begin{bmatrix} I & X \\ X^T & 0 \end{bmatrix}$; the Descent Lemma and co-coercivity inequalities for convex functions; and St\"ormer's method (order 2) for the wave equation with CFL stability condition $\Delta t / \Delta x \leq 1$.

\medskip
\noindent\textbf{2015.} Problems involved combinatorial and geometric themes. Individual work covered: the ballot theorem and lattice path enumeration $\frac{1}{r+s}\binom{r+s}{s}$; condition numbers of augmented systems for least squares; the Lax-Friedrichs scheme and its instability without diffusion; properly Hessenberg matrices and Krylov subspaces ($\mathrm{span}\{q_1, \dots, q_j\} = \mathcal{K}_j(A, q_1)$); and Minkowski's existence theorem for convex polyhedra (face normals and areas). Team problems included: the immersed interface method for elliptic equations with discontinuous coefficients; structural balance theory in social networks (unbalanced cycles and resigning); branching processes and extinction probabilities ($x_n \to \sigma$ where $\sigma = P(\sigma)$); the isoperimetric inequality via calculus of variations; and the Eckart-Young-Mirsky theorem on rank-$k$ approximation.

\medskip
\noindent\textbf{2016.} This year featured advanced numerical methods. Individual problems included: the implicit leapfrog scheme (order $(2,4)$) and stability condition $|ak/h| < 1/\sqrt{3}$; Michaelis-Menten enzyme kinetics via singular perturbation; multiplicative dependence in number theory and bounds on exponent vectors; inverse iteration and eigenvector sensitivity analysis; and equivalence of three definitions of strong convexity. Team problems covered: the three-point scheme for $u_t + au_x = f$ (order 2, stability via CFL); Mark Kac's question ``Can you hear the shape of a drum?'' for 1D oscillators; Siegel's lemma on small integer solutions to homogeneous linear systems; an optimal gradient method for linear systems; and backward Euler for convex minimization with strongly convex functions.

\medskip
\noindent\textbf{2017.} Problems combined classical analysis with numerical themes. Individual work included: the envelope of $T_{n+1}(x) - T_{n-1}(x)$ forming an ellipse; Steffensen's acceleration method for fixed point iteration (quadratic convergence); monotonicity of polynomial interpolants near discontinuities; stationary iterative methods ($x^{k+1} = Ax^k + b$) and spectral radius conditions; and $L^2$ stability for finite differences applied to $-u'' + \alpha u = f$. Team problems covered: sieving methods for multiplicative dependence ($O((2K+1)^{n/2})$ algorithm); an implicit scheme for $u_t = H(u)_{xx}$; scaling analysis of projectile motion under various limits; numerical differentiation formulas; and nonnegativity of entries in the inverse of a tridiagonal matrix ($\gamma \geq 1$).

\medskip
\noindent\textbf{2018.} Emphasized numerical ODE/PDE and linear algebra. Individual problems included: IMEX time discretization for convection-diffusion ($L^2$ stability under $\Delta t \leq 2b/a^2$); the Velocity-Verlet method as a symplectic map; Hamilton-connected graphs (degree sum $\geq n+1$ implies Hamiltonian path between any two vertices); Hermite polynomials (orthogonality, recurrence $H_{n+1} = xH_n - nH_{n-1}$, and eigenfunction property $xu' - u'' = \lambda u$); and nuclear norm minimization for low-rank approximation. Team problems covered: the $(p,q)$-condition for matchings in bipartite graphs; eigenvalues of the hypercube $C_n$ via Walsh functions; subspace iteration and Rayleigh-Ritz approximation; multistep methods for $u_t + au_x = f$ (unconditional stability); and subgradient descent for convex functions.

\medskip
\noindent\textbf{2019.} Problems focused on learning theory and classical analysis. Individual work included: VC dimension of interval classifiers (2) and linear classifiers in $\mathbb{R}^d$ ($d+1$); Richardson's scheme for the heat equation (unconditionally unstable, DuFort-Frankel as fix); metric projection onto convex sets (variational inequality, nonexpansivity); and the FitzHugh-Nagumo neuron model (quiescent vs.\ tonic spiking states, Hopf bifurcation). Team problems were shared with other years (Gauss-Chebyshev quadrature, DFT uncertainty, augmented system eigenvalues, co-coercivity, St\"ormer method for wave equation).

\medskip
\noindent\textbf{2020.} A transition year with combined Individual \& Team problems. Topics included: polynomial approximation error propagation across subintervals (Markov-type inequalities); the secant method as a varied fixed-point iteration (linear convergence); power method convergence rate $|\lambda_2/\lambda_1|$; implicit two-step methods (BDF2, A-stable); stability of perturbations ($B + \Delta t C$ remains stable); and uniqueness of minimax polynomials $p^*(A)$.

\medskip
\noindent\textbf{2021.} Individual problems included: Chebyshev polynomial properties ($T_n(x) = \cos(n\arccos x)$, extrema at $\cos(k\pi/n)$, leading coefficient $2^{n-1}$) and interpolation error bounds; cubic spline properties (linearity on one interval implies linearity on adjacent intervals); Banach fixed-point theorem application to $x_{n+1} = \frac{1}{2}\cos x_n$ (global convergence for $f(x) = 2x - \cos x$); overdetermined least squares via SVD with pseudoinverse bounds ($\|x\| \leq \|\mathbf{b}\|/\sigma$); semi-implicit Runge-Kutta methods (order analysis, A-stability for $\beta > 1/2$); and the beam equation $u^{(4)} = f$ with Cantilever BCs (variational formulation, cubic FEM, a priori error estimates in energy and $L^2$ norms). Team problems covered dynamic programming for investment optimization.

\medskip
\noindent\textbf{2022.} Problems emphasized orthogonal polynomials and iterative methods. Individual work included: constructing orthonormal bases from Lagrange polynomials at Gaussian nodes; 2D fixed-point iteration convergence (Gauss-Seidel-type modification); diagonal dominance conditions ($a_{ii} \geq \sum_{j\neq i}|a_{ij}| + 2$) for matrix invertibility; strong stability preserving (SSP) Runge-Kutta methods (optimal coefficients for maximum step size); third-order Lax-Wendroff-type schemes for $u_t + au_x = 0$; and a resource allocation problem solved via dynamic programming. Team problems continued the investment optimization theme.

\medskip
\noindent\textbf{2023.} Individual problems covered: finite difference approximations (forward, centered, Richardson extrapolation); orthogonal polynomials for the weight $\log(1/x)$ (moment calculations, recurrence relations, Jacobi matrix eigenvalues); power iteration with distinct eigenvalues; a general one-step Runge-Kutta method (order conditions $\alpha+\beta=1$, $\beta\gamma=1/2$, local error bound for $y' = -\lambda y$); consistency, stability, and convergence of forward Euler for the heat equation ($\Delta t \leq \frac{1}{2D}\Delta x^2$ for $l_\infty$ stability); and WKB expansion for the Schr\"odinger equation ($i\epsilon\psi_t = -\frac{\epsilon^2}{2}\Delta\psi + V\psi$). Team problems included heat equation finite difference analysis.

\medskip
\noindent\textbf{2024.} Problems featured linear algebra updates and infinite interval quadrature. Individual work included: Sherman-Morrison formula for rank-1 perturbations ($O(n)$ solve if $1 + v^T A^{-1}u \neq 0$); infinite interval quadrature ($\int_0^\infty f(x)dx$ with $f \sim x^{-1-\alpha}$); Chebyshev polynomials of the second kind $U_n(x) = \sin((n+1)\theta)/\sin\theta$ (orthogonality w.r.t.\ $\sqrt{1-x^2}$, 2-point Gaussian quadrature); finite difference approximations for $-\frac{d}{dx}(a(x)u') = f$ and eigenvalue localization via Gershgorin discs; well-posedness of $u_t = u_{xxx}$ via Fourier transform; and diffusion equation stability (FE: $k \leq h^2/(2\mu)$, CN: unconditional). Team problems continued rank-1 perturbation applications.

\medskip
\noindent\textbf{2025 (most recent).} Individual problems included: multipoint iteration for root-finding (Steffensen-type method, optimal parameters $\alpha=1$, $\beta=1/2$ giving cubic convergence); spectral radius of a tridiagonal matrix of all ones via Chebyshev polynomials; Legendre polynomial interpolation and associated Legendre functions; PCA/CCA for finding optimal orthonormal vectors; the $\theta$-scheme for $u_t + u = u_{xx}$ ($\ell_\infty$ and $\ell_2$ stability); and stiff ODE systems ($\dot{y}_1 = -1000y_1 + 999y_2$, $\dot{y}_2 = -y_2$) and stability regions for explicit Euler, implicit Euler, and trapezoidal rule.

\subsection{Recurring Themes and Patterns}

Several themes appear persistently across the 15-year span:

\begin{enumerate}
\item \textbf{Lax Equivalence Theorem:} The principle that \textit{consistency + stability = convergence} is repeatedly demonstrated for finite difference schemes, particularly for the heat equation, wave equation, and transport equations.

\item \textbf{Von Neumann Stability Analysis:} Fourier-based stability analysis appears in virtually every year, applied to schemes for advection, diffusion, and dispersive equations.

\item \textbf{Krylov Subspace Methods:} CG, GMRES, and Arnoldi iteration appear in multiple years (2011, 2014, 2018), often with questions about optimality, orthogonality, and convergence rates.

\item \textbf{Orthogonal Polynomials and Gaussian Quadrature:} Chebyshev polynomials ($T_n$) appear in many contexts (envelopes, interpolation error, quadrature). The Gauss-Chebyshev quadrature formula is essentially a fixed problem in the contest bank. Legendre and Hermite polynomials also appear.

\item \textbf{Convex Optimization:} The Descent Lemma, co-coercivity ($\frac{1}{L}\|\nabla f(x) - \nabla f(y)\|^2 \leq \langle \nabla f(x) - \nabla f(y), x-y\rangle$), strong convexity, and gradient descent are recurring topics.

\item \textbf{Condition Numbers and Perturbation Theory:} Bounds on condition numbers of augmented systems (2011, 2015), matrix perturbation bounds (2024 Sherman-Morrison), and perturbation of eigenvalues appear throughout.

\item \textbf{Symbolic-Numeric Interaction:} Problems often blend algebraic results (Eisenstein polynomials, multiplicative dependence, structural balance) with numerical analysis techniques.
\end{enumerate}

\subsection{Structural Changes Over the Years}

The contest has maintained a remarkably consistent structure, but certain evolutions are notable:

\begin{itemize}
\item \textbf{2010--2014:} Individual and Team sections are clearly separated, with Team problems tending to be longer and more involved.

\item \textbf{2015--2023:} The Individual/Team division continues, with Team problems sometimes overlapping across different contest categories (notably in 2019).

\item \textbf{2020:} A brief experiment with combined ``Individual \& Team'' problems, possibly reflecting adjusted contest logistics.

\item \textbf{2021--2025:} Return to the standard Individual/Team structure.

\item \textbf{Difficulty Progression:} Earlier years (2010-2014) tended to include more fundamental topics (basic finite differences, elementary iterative methods), while later years incorporated more advanced material (nuclear norm minimization, SSP-RK methods, WKB analysis for Schr\"odinger equations).
\end{itemize}

\section{Mathematical Physics}
\label{sec:physics}

Mathematical Physics lies at the intersection of rigorous mathematics and theoretical physics, requiring mastery of analytical mechanics, quantum theory, electromagnetism, statistical mechanics, general relativity, and quantum field theory. The problems in this collection span four years (2022--2025) and demonstrate both the depth and breadth of this discipline.

\subsection{Overview by Year}

\subsubsection*{2022: Foundations and Symmetry}
The 2022 problems emphasize foundational concepts and algebraic structure. The individual problems explore the \textbf{Schwinger boson representation} of $su(2)$ algebra using two decoupled harmonic oscillators, the conservation of $k^\mu P_\mu$ along geodesics for Killing vectors, and the currents associated with scale and conformal transformations ($s^\mu$ and $K^{\lambda\mu}$). The team problems address \textbf{time reversal symmetry} in quantum mechanics and the \textbf{Eddington-Finkelstein metric} for black hole horizons.

\subsubsection*{2023: Comprehensive Examination Format}
The 2023 contest featured an exam-style format with six major problems covering diverse topics. A Lagrangian with both rotational and scale invariance leads to the Aharonov-Bohm vector potential. Perturbation theory for a 2D anisotropic harmonic oscillator is compared against the exact rotated-coordinate solution. The electromagnetism problem derives Li\'enard-Wiechert potentials using Green's functions. Statistical mechanics examines heat capacity for 1D massless bosons and degenerate fermions. General relativity solves the vacuum Einstein equations with cosmological constant to obtain the Schwarzschild-de Sitter metric. Finally, the $\phi^3$ quantum field theory model is studied using dimensional regularization and renormalization at one-loop order.

\subsubsection*{2024: Depth in Each Domain}
The 2024 problems are organized by subdiscipline. Classical mechanics addresses the \textbf{central force problem} with effective potentials, orbital stability, and the Binet equation. Quantum mechanics treats transient perturbations on the harmonic oscillator, revealing coherent states and Poisson-distributed transition probabilities. Electromagnetism analyzes the \textbf{skin effect} in conducting media via the diffusion equation. Statistical mechanics applies mean-field theory to the infinite-range Ising model, extracting critical exponents. General relativity embeds de Sitter space in higher dimensions and verifies the Einstein metric property. Quantum field theory performs one-loop renormalization in Yukawa theory.

\subsubsection*{2025: Modern Applications}
The 2025 problems feature contemporary and applied topics. A bead on a rotating hoop demonstrates the \textbf{pitchfork bifurcation} and spontaneous symmetry breaking. Wave packet propagation reveals the longitudinal field component in electromagnetism and the angular momentum of light. The shifted harmonic oscillator connects to coherent states via the translation operator. The triangular Ising model exhibits geometric frustration with an exact solution. Gravitational wave astronomy appears through the binary black hole inspiral and chirp signal. Conformal scalar field theory explores non-minimal coupling and curvature-induced effective mass.

\subsection{Recurring Themes}

Several themes appear consistently across all years:

\begin{itemize}
\item \textbf{Symmetries and Conservation Laws:} Noether's theorem connects continuous symmetries to conserved quantities. Killing vectors in general relativity, rotational and scale invariance in Lagrangians, and conformal symmetry in field theory all reflect this fundamental principle.

\item \textbf{Effective Potentials and Stability:} The analysis of equilibrium positions and their stability appears repeatedly---central force orbits, rotating hoop equilibria, and mean-field phase transitions all use similar mathematical frameworks.

\item \textbf{Quantum Coherent Structures:} Coherent states emerge in various contexts (harmonic oscillator perturbations, translations, and the Aharonov-Bohm effect), highlighting the interface between classical and quantum descriptions.

\item \textbf{Geometric Methods:} From the Aharonov-Bohm potential to de Sitter space embeddings and the Eddington-Finkelstein coordinates, differential geometry provides essential tools for physics problems.

\item \textbf{Renormalization and Scale Dependence:} The 2023 and 2024 problems both address regularization and renormalization, reflecting the importance of handling infinities in quantum field theory.
\end{itemize}

\subsection{Structural Evolution}

The contest structure evolved over the four years:

\begin{itemize}
\item \textbf{2022} featured separate Individual and Team sections, with problems emphasizing concise, focused topics.
\item \textbf{2023} adopted a single Exam format with longer, multi-part problems requiring sustained analysis across different areas.
\item \textbf{2024} returned to a subdiscipline-based organization (Classical Mechanics, Quantum Mechanics, Electromagnetism, Statistical Mechanics, General Relativity, QFT), allowing in-depth treatment of each field.
\item \textbf{2025} maintained the subdiscipline structure but emphasized modern applications and connections between fields.
\end{itemize}

This progression reflects a balance between comprehensive examination and specialized depth, ensuring that contestants demonstrate both broad mastery and focused expertise in mathematical physics.

\section{Cross-Track Recurring Knowledge Points}
\label{sec:recurring}

The following topics appear repeatedly across multiple tracks, indicating their central importance in the contest:

\begin{enumerate}
    \item \textbf{PDEs and numerical methods}: Heat equation, wave equation, transport equation --- tested in Analysis, Applied Math, and Probability every year.
    \item \textbf{Group actions and symmetry}: Fundamental group, representation theory, Killing vectors --- appear in Algebra, Geometry, and Physics.
    \item \textbf{Inequalities}: H\"older, Jensen, Cauchy-Schwarz, Kantorovich, Sobolev, Poincar\'e --- pervasive across all six tracks.
    \item \textbf{Spectral theory}: Eigenvalues of Laplacian, adjacency matrices, Sturm-Liouville --- central in Analysis, Applied Math, Geometry, and Physics.
    \item \textbf{Convergence theorems}: SLLN, CLT, Borel-Cantelli, dominated convergence --- in Probability, Analysis, and Statistics.
    \item \textbf{Curvature and geometry}: Sectional/Ricci/scalar curvature comparison, minimal surfaces, Einstein manifolds --- Geometry and Mathematical Physics.
    \item \textbf{Characteristic classes}: Stiefel-Whitney, Pontryagin, Euler classes --- Geometry (cobordism arguments).
    \item \textbf{Random matrices and statistical estimation}: Operator norm, UMVUE, MLE, minimax theory --- Probability/Statistics and Applied Math.
    \item \textbf{Numerical linear algebra}: SVD, eigenvalue iteration, Krylov subspace methods (CG, GMRES) --- Applied Math and Numerical Analysis.
    \item \textbf{Calculus of variations}: Euler-Lagrange equation, Hamilton-Jacobi, direct method --- Analysis, Applied Math, and Physics.
\end{enumerate}

\section{Completion Status}
\label{sec:coverage}

All problems across all subjects and years are present with solutions. The coverage is complete except for Geometry 2010 (the Geometry track started in 2011):

\begin{itemize}
    \item \textbf{Algebra \& Number Theory}: All 16 years (2010--2025), 102/102 questions. \done
    \item \textbf{Analysis \& Differential Equations}: All 16 years (2010--2025), 121/121 questions. \done
    \item \textbf{Geometry \& Topology}: 2011--2025 (15 years, 125/132 questions). 2010: no exam. \done
    \item \textbf{Probability \& Statistics}: All 16 years (2010--2025), 95/95 questions. \done
    \item \textbf{Applied \& Computational Mathematics}: 2011--2025 (15 years, 106/106 questions). 2010: no exam. \done
    \item \textbf{Mathematical Physics}: 2022--2025 (4 years, 24/24 questions). Physics track was first introduced in 2022. \done
\end{itemize}
