\begin{minted}
  [
  frame=lines,
  framesep=2mm,
  baselinestretch=1,
  fontsize=\footnotesize,
  linenos
  ]
  {coq}

(* Declaration of constants used in the statement of the riddle *)

Context (i : Type).
Context (Rich : i -> Prop).
Context (mother : i -> i).
Context (h : i).

(* Statement of the riddle *)

Theorem rich_mothers :
  (forall x, ~Rich(x) -> Rich(mother(x))) ->
  (forall x,  ~Rich(mother(mother(x))) \/ ~Rich(x)) ->
  False.

(* Proof of the riddle *)

Proof.
  intros H1 H2.
  (** 2 clicks on the conclusion *)

  destruct (H2 h) as [H | H].
  (** DnD of [h] onto [H2], then click on the resulting hypothesis *)

  * pose proof (H1 _ H) as H'.
    (* If one naively uses [apply _ in], then one loses [H] although
       it is needed later! Hence the use of [pose proof]. *)
    (** DnD of [H1] onto [H] *)
    
    destruct (H2 (mother h)) as [H2' | H2'].
    apply H2'. exact H'.
    (** DnD of [H'] onto [H2]. Could also be performed stepwise:  
        - Selection of [mother(h)] in [H']
        - DnD of [H'] onto [H2]
        - Click on the resulting hypothesis
        - DnD of [H2'] onto [H'] *)

    apply H1 in H2'.
    (** DnD of [H1] onto [H2'] *)

    apply H. exact H2'.
    (** DnD of [H] onto [H2'] *)

  * pose proof (H1 _ H) as H'.
    (** DnD of [H1] onto [H] *)

    destruct (H2 (mother h)) as [H2' | H2'].
    2: { apply H2'. exact H'. }
    (** DnD of [H'] onto [H2] *)

    pose proof (H1 _ H2') as H2''.
    (** DnD of [H1] onto [H2'] *)

    destruct (H2 (mother (mother h))) as [H2''' | H2'''].
    apply H2'''. exact H2''.
    (** DnD of [H2''] onto [H2] *)

    apply H1 in H2'''.
    (** DnD of [H1] onto [H2'''] *)

    apply H2'. exact H2'''.
    (** DnD of [H2'] onto [H2'''] *)
Qed.

\end{minted}