\section{2017 Geometry \& Topology}

\subsection{Individual Exam}

\begin{exercise}
Let $M$ be a smooth, compact, oriented $n$-dimensional manifold. Suppose that the Euler characteristic of $M$ is zero. Show that $M$ admits a nowhere vanishing vector field.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Manifold:} A space that locally looks like Euclidean space $\mathbb{R}^n$.

2. \textbf{Vector Field:} An assignment of a tangent vector to each point on the manifold in a smooth manner.

3. \textbf{Euler Characteristic ($\chi(M)$):} A topological invariant (a number) that describes the shape or structure of a topological space regardless of the way it is bent. For instance, the Euler characteristic of a sphere is $2$, and for a torus it is $0$.

4. \textbf{Poincaré-Hopf Theorem:} This theorem connects the topology of a manifold (its Euler characteristic) to the analytical behavior of vector fields on it. It states that for any smooth vector field with isolated zeros on a compact manifold, the sum of the indices of these zeros equals the Euler characteristic of the manifold.

5. \textbf{Index of a Zero:} A number measuring how the vector field "wraps around" a zero point.

\textbf{Intuition:}

This problem is essentially asking us to prove a version of the famous "Hairy Ball Theorem" but in reverse. The Hairy Ball Theorem says you can't comb a hairy sphere flat without creating a cowlick (a zero of the vector field). This is because the Euler characteristic of a sphere is $2$, not $0$.

However, if the "topological debt" (the Euler characteristic) is exactly zero, we are not forced to have any zeros. We can think of zeros as positive and negative charges. If the total sum of these charges is zero, we can mathematically "push" them together so they annihilate each other, leaving a completely smooth, nowhere vanishing field. A physical example is a torus (donut shape), which has $\chi=0$ and can be combed smoothly (imagine water flowing cleanly around the donut ring).

\textbf{Steps:}

\textbf{Step 1: Start with a generic vector field.}
Since $M$ is a smooth compact manifold, we can always construct a smooth vector field on it, say $X$. We can perturb $X$ slightly if necessary so that it only has a finite number of isolated zeros. Let these zero points be $p_1, p_2, \dots, p_k$.

\textbf{Step 2: Apply the Poincaré-Hopf Theorem.}
According to the Poincaré-Hopf Index Theorem, if we sum the indices of all these isolated zeros, we must get exactly the Euler characteristic of the manifold:
\[ \sum_{i=1}^{k} \mathrm{ind}_{p_i}(X) = \chi(M) \]
Since we are given that $\chi(M) = 0$, we know that the sum of the indices of all zeros is exactly zero.

\textbf{Step 3: Handle the $n=1$ case.}
If the manifold is 1-dimensional, the only connected compact orientable manifold is the circle $S^1$. The circle clearly admits a nowhere vanishing vector field (just a constant vector pointing along the curve everywhere, like moving counterclockwise).

\textbf{Step 4: Annihilate the zeros for $n \ge 2$.}
For dimensions $2$ and higher, we can use a "singularity cancellation" technique. Because the sum of the indices is zero, we can pair up zeros with positive indices and zeros with negative indices.

Since the manifold is connected, we can draw a path between any positive index zero and negative index zero. We can locally modify the vector field $X$ along this path to push these two zeros together.

When they meet, their indices add up to $0$, and they annihilate each other, resulting in a local patch with no zeros.

By repeating this pairing and canceling process, we can eventually eliminate all the zeros.

\textbf{Step 5: Conclude.}
Once all zeros are eliminated, the modified vector field is smooth and has no zeros anywhere on $M$. Therefore, $M$ admits a nowhere vanishing vector field.
\end{solution}

\begin{exercise}
Let $S^{2} \xrightarrow{q_{1}} S^{2} \vee S^{2} \xrightarrow{q_{2}} S^{2}$ be the maps that crush out one of the two summands. Let $f : S^{2} \to S^{2} \vee S^{2}$ be a map such that $q_{i} \circ f : S^{2} \to S^{2}$ is a map of degree $d_{i}$. Compute the integral homology groups of $(S^{2} \vee S^{2}) \cup_{f} D^{3}$. Here $D^{3}$ is the unit solid ball with boundary $S^{2}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{Homology Groups ($H_k(X)$):} Algebraic structures (specifically, abelian groups) that count the number of "$k$-dimensional holes" in a topological space. For example, $H_0$ counts connected components, $H_1$ counts 1D loops, and $H_2$ counts 2D cavities.

2. \textbf{Wedge Sum ($S^2 \vee S^2$):} The space formed by taking two 2-dimensional spheres and gluing them together at a single point. It looks like a figure-8 made of balloons.

3. \textbf{Cellular Homology:} A method to compute homology groups by building the space out of basic building blocks called "cells" (points are 0-cells, lines are 1-cells, disks are 2-cells, solid balls are 3-cells, etc.). We then compute boundary maps between these cells.

4. \textbf{Degree of a Map:} An integer that represents how many times a continuous map wraps a sphere around another sphere.

\textbf{Intuition:}

Imagine you have two empty balloons tied together at their knots (this is $S^2 \vee S^2$). Now, you take a solid, filled rubber ball ($D^3$) and you want to glue its outer skin (which is an $S^2$) onto the two balloons.

The map $f$ tells you exactly how to stretch and wrap the skin of the solid ball around the two balloons. The numbers $d_1$ and $d_2$ tell you that the skin wraps around the first balloon $d_1$ times and around the second balloon $d_2$ times.

By gluing the solid ball, we are essentially "filling in" some of the 2D emptiness (cavities) of the balloons. We need to set up the cellular structure to calculate exactly what topological features remain.

\textbf{Steps:}

\textbf{Step 1: Identify the cell structure.}
Let our space be $X = (S^2 \vee S^2) \cup_f D^3$. We can build $X$ using a CW complex structure:
First, start with one point, which is a $0$-cell. Let's call it $e_0$.
Second, we have no lines or loops to add, so there are no $1$-cells.
Third, attach two 2-dimensional disks (the surfaces of the two balloons). These are two $2$-cells, let's call them $e_1$ and $e_2$. They are both attached to the single point $e_0$. At this stage, we have formed the wedge sum $S^2 \vee S^2$.
Finally, attach one 3-dimensional solid ball ($D^3$). This is our single $3$-cell, $e_3$. It is attached via the map $f$.

\textbf{Step 2: Set up the cellular chain complex.}
The chain groups $C_k$ are freely generated by the $k$-cells.
- $C_0 \cong \mathbb{Z}$ (generated by $e_0$)
- $C_1 = 0$ (because there are no 1-cells)
- $C_2 \cong \mathbb{Z} \oplus \mathbb{Z}$ (generated by $e_1$ and $e_2$)
- $C_3 \cong \mathbb{Z}$ (generated by $e_3$)
- $C_k = 0$ for all $k > 3$

\textbf{Step 3: Determine the boundary maps.}
The boundary map $\partial_k: C_k \to C_{k-1}$ tells us how the boundary of the $k$-cells wraps around the $(k-1)$-cells.
- $\partial_1: C_1 \to C_0$ is simply the zero map because $C_1 = 0$.
- $\partial_2: C_2 \to C_1$ is also the zero map because $C_1 = 0$.
- $\partial_3: C_3 \to C_2$ is the crucial one. The boundary of our 3-cell $e_3$ is attached to the two 2-cells. The degrees $d_1$ and $d_2$ precisely provide the coefficients for how it wraps! Therefore, $\partial_3(e_3) = d_1 e_1 + d_2 e_2$.
As a matrix mapping from $\mathbb{Z}$ to $\mathbb{Z}^2$, $\partial_3$ is just the column vector $[d_1, d_2]^T$.

\textbf{Step 4: Calculate the homology groups.}
Homology is defined as $H_k(X) = \ker(\partial_k) / \mathrm{im}(\partial_{k+1})$.

For $H_0(X)$: $\ker(\partial_0) = C_0 \cong \mathbb{Z}$ and $\mathrm{im}(\partial_1) = 0$. So, $H_0(X) = \mathbb{Z} / 0 = \mathbb{Z}$. (This just means the space is connected).

For $H_1(X)$: Since $C_1 = 0$, we simply have $H_1(X) = 0$.

For $H_2(X)$: We need to find $\ker(\partial_2) / \mathrm{im}(\partial_3)$.
Since $\partial_2$ is the zero map, $\ker(\partial_2) = C_2 \cong \mathbb{Z}^2$.
The image $\mathrm{im}(\partial_3)$ is the subgroup generated by the vector $(d_1, d_2)$.
Therefore, $H_2(X) = \mathbb{Z}^2 / \langle(d_1, d_2)\rangle$.
If both $d_1 = 0$ and $d_2 = 0$, then we are modding out by zero, so $H_2(X) = \mathbb{Z}^2$.
If at least one is non-zero, let $g = \gcd(d_1, d_2)$. We can change our basis for $\mathbb{Z}^2$ to make one generator $(d_1/g, d_2/g)$. Then the vector $(d_1, d_2)$ is exactly $g$ times this new generator. This quotient leaves one free $\mathbb{Z}$ part and one finite cyclic part $\mathbb{Z}_g$. So, $H_2(X) \cong \mathbb{Z} \oplus \mathbb{Z}_g$.

For $H_3(X)$: We need to find $\ker(\partial_3) / \mathrm{im}(\partial_4)$. Since $C_4=0$, $\mathrm{im}(\partial_4)=0$, so $H_3(X) = \ker(\partial_3)$.
If $(d_1, d_2) = (0,0)$, then $\partial_3(e_3) = 0$, meaning the whole group $C_3$ is in the kernel. Thus $H_3(X) \cong \mathbb{Z}$.
If at least one is non-zero, then $\partial_3(e_3) \neq 0$. The only element mapping to $0$ is $0$ itself. So $\ker(\partial_3) = 0$, giving $H_3(X) = 0$.

\textbf{Step 5: Final Summary.}
Putting it all together:
$H_0(X) \cong \mathbb{Z}$
$H_1(X) = 0$
If $(d_1, d_2) = (0,0)$, then $H_2(X) \cong \mathbb{Z}^2$ and $H_3(X) \cong \mathbb{Z}$.
If $(d_1, d_2) \neq (0,0)$, then $H_2(X) \cong \mathbb{Z} \oplus \mathbb{Z}_{\gcd(d_1, d_2)}$ and $H_3(X) = 0$.
\end{solution}

\begin{exercise}
Let $X$ and $Y$ be smooth vector fields on a smooth manifold. Prove that the Lie derivative satisfies the identity
\[ L_{X}Y = [X, Y]. \]
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} The Lie derivative $L_X Y$ measures how the field $Y$ changes as it is pushed along the flow of $X$. The Lie bracket $[X, Y]$ measures the non-commutativity of the flows. This identity shows that these two seemingly different concepts—directional derivative of a field and commutator of flows—are exactly the same.

\textbf{Proof:} Let $\phi_t$ be the flow of $X$. For any function $f$:
\[ (L_X Y) f = \lim_{t \to 0} \frac{(\phi_{-t}^* Y) f - Y f}{t} = \lim_{t \to 0} \frac{Y(f \circ \phi_{-t}) \circ \phi_t - Y f}{t} \]
Using the expansion $f \circ \phi_{-t} = f - t X f + O(t^2)$:
\[ (L_X Y) f = \lim_{t \to 0} \frac{Y(f - t X f) \circ \phi_t - Y f}{t} = \lim_{t \to 0} \frac{(Y f - t Y X f) \circ \phi_t - Y f}{t} \]
Since $(Yf) \circ \phi_t = Yf + t XYf$, we get:
\[ (L_X Y) f = \frac{Yf + t XYf - t YXf - Yf}{t} = (XY - YX) f = [X, Y] f. \]
\end{solution}

\begin{exercise}
State and prove the Liouville formula for the geodesic curvature $\kappa_{g}$ along a regular curve on a smooth surface in $\mathbb{R}^{3}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}

1. \textbf{First Fundamental Form ($E, F, G$):} A way to measure distances and angles on a surface. If the surface coordinates are $(u,v)$, the squared length element is $ds^2 = E du^2 + 2F du dv + G dv^2$. For orthogonal coordinates, $F=0$.

2. \textbf{Geodesic Curvature ($\kappa_g$):} How much a curve "turns" within the surface itself. It's the component of the curve's acceleration vector that lies tangent to the surface but perpendicular to the curve's velocity.

3. \textbf{Moving Frame ($\mathbf{e}_1, \mathbf{e}_2$):} A pair of orthonormal basis vectors on the tangent plane at each point.

4. \textbf{Connection Form ($\omega_{12}$):} A 1-form that describes how the basis vectors rotate as you move around the surface.

\textbf{Intuition:}

Imagine you are driving a car on a hilly, uneven surface. The geodesic curvature is how much you have to turn the steering wheel to stay on your path.

If you are just driving straight relative to the local coordinate lines ($u$ and $v$ directions), but those lines themselves are curving because of the shape of the surface, your car will experience a turning force.

Liouville's formula mathematically splits your total steering (geodesic curvature) into two distinct parts:
1. The rate at which the angle of your car $\theta$ changes relative to the local coordinate grid.
2. The rate at which the coordinate grid itself is twisting or turning due to the intrinsic geometry (metric) of the surface.

\textbf{Steps:}

\textbf{Step 1: State the Formula.}
Assume we have an orthogonal coordinate system on the surface, so the first fundamental form is $ds^2 = E du^2 + G dv^2$.
Let our curve be parameterized by arc length $s$. Let $\theta(s)$ be the angle between the tangent vector of the curve and the $u$-coordinate line at any point.
Liouville's formula states:
\[ \kappa_g = \frac{d\theta}{ds} + \frac{1}{2\sqrt{EG}} \left( G_u \frac{dv}{ds} - E_v \frac{du}{ds} \right) \]
where $G_u = \frac{\partial G}{\partial u}$ and $E_v = \frac{\partial E}{\partial v}$.

\textbf{Step 2: Set up the orthonormal frame.}
Because the coordinates are orthogonal, we can define an orthonormal basis for the tangent plane:
$\mathbf{e}_1 = \frac{\mathbf{r}_u}{\sqrt{E}}$ (pointing in the $u$ direction)
$\mathbf{e}_2 = \frac{\mathbf{r}_v}{\sqrt{G}}$ (pointing in the $v$ direction)

The tangent vector to the curve $\mathbf{t}$ forms an angle $\theta$ with $\mathbf{e}_1$:
\[ \mathbf{t} = \cos\theta \, \mathbf{e}_1 + \sin\theta \, \mathbf{e}_2 \]

The normal vector to the curve within the surface is $\mathbf{n}_g$, which is $\mathbf{t}$ rotated by 90 degrees:
\[ \mathbf{n}_g = -\sin\theta \, \mathbf{e}_1 + \cos\theta \, \mathbf{e}_2 \]

\textbf{Step 3: Differentiate the tangent vector.}
Geodesic curvature is defined by $\kappa_g = \langle \nabla_\mathbf{t} \mathbf{t}, \mathbf{n}_g \rangle$, where $\nabla$ is the covariant derivative (taking the derivative and projecting back onto the surface).

Let's differentiate $\mathbf{t}$ with respect to the arc length $s$:
\[ \nabla_\mathbf{t} \mathbf{t} = \nabla_\mathbf{t} (\cos\theta \, \mathbf{e}_1 + \sin\theta \, \mathbf{e}_2) \]
Using the product rule:
\[ \nabla_\mathbf{t} \mathbf{t} = (-\sin\theta \cdot \theta')\mathbf{e}_1 + (\cos\theta \cdot \theta')\mathbf{e}_2 + \cos\theta \nabla_\mathbf{t} \mathbf{e}_1 + \sin\theta \nabla_\mathbf{t} \mathbf{e}_2 \]
\[ \nabla_\mathbf{t} \mathbf{t} = \theta'(-\sin\theta \, \mathbf{e}_1 + \cos\theta \, \mathbf{e}_2) + \cos\theta \nabla_\mathbf{t} \mathbf{e}_1 + \sin\theta \nabla_\mathbf{t} \mathbf{e}_2 \]
Notice that the first term is exactly $\theta' \mathbf{n}_g$.

\textbf{Step 4: Compute the inner product with $\mathbf{n}_g$.}
Now we take the inner product of everything with $\mathbf{n}_g = -\sin\theta \, \mathbf{e}_1 + \cos\theta \, \mathbf{e}_2$:
\[ \kappa_g = \langle \theta' \mathbf{n}_g, \mathbf{n}_g \rangle + \cos\theta \langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{n}_g \rangle + \sin\theta \langle \nabla_\mathbf{t} \mathbf{e}_2, \mathbf{n}_g \rangle \]
Since $\mathbf{n}_g$ is a unit vector, $\langle \mathbf{n}_g, \mathbf{n}_g \rangle = 1$. So the first term is just $\theta'$.
For the other terms, we expand $\mathbf{n}_g$:
\[ \langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{n}_g \rangle = \langle \nabla_\mathbf{t} \mathbf{e}_1, -\sin\theta \, \mathbf{e}_1 + \cos\theta \, \mathbf{e}_2 \rangle = \cos\theta \langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{e}_2 \rangle \]
(Note that $\langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{e}_1 \rangle = 0$ because $\mathbf{e}_1$ is a unit vector).
Similarly,
\[ \langle \nabla_\mathbf{t} \mathbf{e}_2, \mathbf{n}_g \rangle = -\sin\theta \langle \nabla_\mathbf{t} \mathbf{e}_2, \mathbf{e}_1 \rangle = \sin\theta \langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{e}_2 \rangle \]

Combining these into our equation for $\kappa_g$:
\[ \kappa_g = \theta' + (\cos^2\theta + \sin^2\theta) \langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{e}_2 \rangle = \theta' + \langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{e}_2 \rangle \]

The term $\langle \nabla_\mathbf{t} \mathbf{e}_1, \mathbf{e}_2 \rangle$ is defined as the connection form evaluated on the tangent vector, $\omega_{12}(\mathbf{t})$.

\textbf{Step 5: Evaluate the connection form.}
For orthogonal coordinates $ds^2 = E du^2 + G dv^2$, standard differential geometry calculations for the connection form yield:
\[ \omega_{12} = \frac{1}{2\sqrt{EG}} ( -E_v du + G_u dv ) \]
Evaluating this on the tangent vector $\mathbf{t} = \frac{du}{ds}\mathbf{r}_u + \frac{dv}{ds}\mathbf{r}_v$:
\[ \omega_{12}(\mathbf{t}) = \frac{1}{2\sqrt{EG}} \left( G_u \frac{dv}{ds} - E_v \frac{du}{ds} \right) \]

Substituting this back gives the final Liouville formula:
\[ \kappa_g = \frac{d\theta}{ds} + \frac{1}{2\sqrt{EG}} \left( G_u \frac{dv}{ds} - E_v \frac{du}{ds} \right) \]
\end{solution}

\begin{exercise}
On a Riemannian manifold, let $F$ be the set of smooth functions $f$ on $M$ with $|\mathrm{grad}\,f| \leq 1$. For any $x, y$ in the manifold, show that
\[ d(x, y) = \sup\{|f(x) - f(y)| : f \in F\}. \]
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} This is the "dual" definition of distance. Usually, we define distance as the minimum length of a path. Here, we define it as the maximum "height difference" you can achieve with a function that doesn't change too fast (slope $\le 1$). If you can't climb more than 1 meter per meter of horizontal travel, the furthest you can get is your distance away.

\textbf{Proof:}
1. \textbf{$\le$:} For any $f \in F$ and path $\gamma$: $|f(y)-f(x)| = |\int \langle \nabla f, \gamma' \rangle| \le \int |\nabla f| |\gamma'| \le \text{Length}(\gamma)$. Thus $|f(y)-f(x)| \le d(x, y)$.
2. \textbf{$\ge$:} Consider $f_0(z) = d(x, z)$. This function is 1-Lipschitz by triangle inequality, so $|\nabla f_0| \le 1$ almost everywhere. Smooth approximations $f_\epsilon$ satisfy $|\nabla f_\epsilon| \le 1$ and $|f_\epsilon(y)-f_\epsilon(x)| \to d(x, y)$.
\end{solution}

\begin{exercise}
Let $M$ be an $n$-dimensional oriented closed minimal submanifold in an $(n + p)$-dimensional unit sphere $S^{n+p}$. Denote by $K_{M}$ the sectional curvature of $M$. Prove that if $K_{M} > \frac{p-1}{2p-1}$, then $M$ is the great sphere $S^{n}$.
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} This is a curvature "pinching" theorem. It says that if a minimal submanifold is "too flat" (but still positively curved) compared to the sphere it lives in, it must actually be a perfectly flat great sphere. There's no room for "small" minimal wrinkles if the curvature is high.

\textbf{Proof Sketch:} One uses the Simons' identity for the second fundamental form $|h|^2$. The curvature bound $K_M > \frac{p-1}{2p-1}$ ensures that the "curving" term in the Laplacian formula for $\Delta |h|^2$ dominates. Integrating over the closed manifold forces $|h|=0$, meaning the submanifold is totally geodesic.
\end{solution}

\subsection{Team Exam}

\begin{exercise}
Consider the space $X = M_{1} \cup M_{2}$, where $M_{1} \text{ and } M_{2}$ are Möbius bands and $M_{1} \cap M_{2} = \partial M_{1} = \partial M_{2}$. Determine the fundamental group of $X$.
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} Gluing two Möbius bands along their boundaries produces a Klein bottle. A Möbius band "wraps" around its center circle once, but its boundary wraps around twice.

\textbf{Proof:} Each $\pi_1(M_i) \cong \langle a_i \rangle$. The boundary generator $c$ maps to $2 a_i$ in each.
By Seifert-van Kampen: $\pi_1(X) = \langle a_1, a_2 \mid a_1^2 = a_2^2 \rangle$. This is the standard group for the Klein bottle.
\end{solution}

\begin{exercise}
For $n \in \mathbb{Z}$, let $f_{n}$ be a degree $n$ map on $S^{3}$. Compute the integral homology groups of $T_{f_{n}}$.
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} A mapping torus is a "cylinder" $S^3 \times [0, 1]$ with the ends glued after a "twist" by $f_n$. The homology depends on how $f_n$ permutes the cycles of $S^3$.

\textbf{Result:} Using the Wang sequence: $H_0 \cong \mathbb{Z}, H_1 \cong \mathbb{Z}$. $H_2 = 0$. $H_3 \cong \mathbb{Z}_{|n-1|}$. If $n=1$, $H_3 \cong \mathbb{Z}, H_4 \cong \mathbb{Z}$. If $n \neq 1$, $H_4 = 0$.
\end{solution}

\begin{exercise}
Show that geodesic torsion along a curve $C$ on a surface $S$ is given by the specified determinant formula involving the first and second fundamental forms.
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} Geodesic torsion measures how the surface normal "tilts" sideways as you move along a curve. The determinant formula combines the shape of the surface (II) and the direction of the curve ($du, dv$) in a way that extracts this "twist."

\textbf{Proof Sketch:} One calculates $\tau_g = \langle d\mathbf{N}/ds, \mathbf{n}_g \rangle$. By substituting the Weingarten equations and the components of the normals, the algebraic expansion matches the provided determinant.
\end{solution}

\begin{exercise}
On a Riemannian manifold, if $f$ is a smooth function such that $|\mathrm{grad}\,f| = 1$, then the integral curves of $\mathrm{grad}\,f$ are geodesics.
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} If you always walk "straight uphill" at a constant speed, and the slope is uniform, you won't feel any "side-force" that would make you turn. Your path will be a geodesic.

\textbf{Proof:} Let $X = \nabla f$. We want to show $\nabla_X X = 0$.
For any $Y$, $\langle \nabla_Y X, X \rangle = \frac{1}{2} Y \langle X, X \rangle = \frac{1}{2} Y(1) = 0$.
Since $X$ is a gradient, $\text{Hess}(f)$ is symmetric, so $\langle \nabla_X X, Y \rangle = \text{Hess}(f)(X, Y) = \text{Hess}(f)(Y, X) = \langle \nabla_Y X, X \rangle = 0$.
Since this holds for all $Y$, $\nabla_X X = 0$.
\end{solution}
