\section{2020 Geometry \& Topology}

\subsection{Individual \& Team}

\begin{exercise}
\begin{enumerate}
    \item[(a)] Show that $\mathbb{RP}^{2n}$ cannot be the boundary of a compact manifold.
    \item[(b)] Show that $\mathbb{RP}^{3}$ is the boundary of some compact manifold.
\end{enumerate}
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To understand this problem, you need to know about \textit{manifolds}, \textit{boundaries}, and \textit{Stiefel-Whitney classes}. A manifold is a mathematical space that locally looks like Euclidean space. The boundary of a manifold is its edge (like the circle is the boundary of a disk). Stiefel-Whitney classes are algebraic tools used to classify manifolds and their vector bundles.

\vspace{\baselineskip}
\textbf{Intuition:}
This problem explores the concept of \textit{cobordism}. Imagine asking if a given shape can serve as the "skin" or boundary of a higher-dimensional solid. The rule is that a manifold can only be a boundary if certain algebraic invariants, called characteristic numbers, all evaluate to zero. For the even-dimensional projective space $\mathbb{RP}^{2n}$, these numbers do not vanish, meaning it cannot "cap off" a higher-dimensional space. However, $\mathbb{RP}^3$ has zero characteristic numbers, so it can be a boundary.

\vspace{\baselineskip}
\textbf{Steps:}
\textbf{Part (a) - Showing $\mathbb{RP}^{2n}$ is not a boundary:}

Step 1: Use Thom's Theorem. This foundational theorem states that a compact smooth manifold is the boundary of a compact smooth manifold if and only if all of its Stiefel-Whitney numbers are zero.

Step 2: Calculate the total Stiefel-Whitney class for $\mathbb{RP}^{2n}$. The standard formula is $w(\mathbb{RP}^{2n}) = (1+a)^{2n+1}$, where $a$ is the generator of the first cohomology group $H^1(\mathbb{RP}^{2n}; \mathbb{Z}_2)$.

Step 3: Extract the top Stiefel-Whitney class, $w_{2n}$. Using the binomial expansion, the coefficient for $a^{2n}$ is $\binom{2n+1}{2n} = 2n+1$.

Step 4: Reduce modulo 2. In $\mathbb{Z}_2$ arithmetic, since $2n$ is even, $2n+1$ is odd and therefore equals $1$. So, $w_{2n}(\mathbb{RP}^{2n}) = a^{2n} \neq 0$.

Step 5: Evaluate the corresponding Stiefel-Whitney number. The evaluation $\langle w_{2n}, [\mathbb{RP}^{2n}] \rangle = 1$, which is non-zero. Since not all Stiefel-Whitney numbers are zero, $\mathbb{RP}^{2n}$ cannot be a boundary.

\vspace{\baselineskip}
\textbf{Part (b) - Showing $\mathbb{RP}^{3}$ is a boundary:}

Step 1: Calculate the total Stiefel-Whitney class for $\mathbb{RP}^{3}$. Using the same formula, $w(\mathbb{RP}^3) = (1+a)^4 = 1 + 4a + 6a^2 + 4a^3$.

Step 2: Reduce modulo 2. The coefficients $4$, $6$, and $4$ are all even numbers, so they become $0$ in $\mathbb{Z}_2$ coefficients. Thus, $w(\mathbb{RP}^3) = 1$.

Step 3: Conclude with Thom's Theorem. Since all higher Stiefel-Whitney classes ($w_1, w_2, w_3$) are zero, all corresponding Stiefel-Whitney numbers must be zero. Therefore, $\mathbb{RP}^3$ is the boundary of some compact manifold.
\end{solution}

\begin{exercise}
Suppose $M$ is a noncompact, complete $n$-dimensional manifold, and suppose there is an open subset $U \subset M$ and an open set $V \subset \mathbb{R}^{n}$ such that $M \setminus U$ is isomorphic to $\mathbb{R}^{n} \setminus V$. If $\mathrm{Ric} M \geq 0$, show that $M$ is isometric to $\mathbb{R}^{n}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To understand this problem, you need to know about \textit{Riemannian manifolds}, \textit{Ricci curvature}, and the \textit{Bishop-Gromov Volume Comparison Theorem}. Ricci curvature measures how the volume of a small geodesic ball deviates from the volume of a standard ball in flat Euclidean space. The Bishop-Gromov theorem provides a strict rule for how this volume can grow when the curvature is non-negative.

\vspace{\baselineskip}
\textbf{Intuition:}
This is a classic \textit{rigidity} result. Think of Ricci curvature $\geq 0$ as a rule that says "space cannot flare out and grow faster than flat space." If the "end" or far-away regions of the manifold look exactly like flat $\mathbb{R}^n$, it means the volume growth at infinity is as large as possible. Because the space is not allowed to grow faster than flat space anywhere, this maximal growth at infinity "forces" the interior to also be perfectly flat. Any curvature would have slowed down the volume growth.

\vspace{\baselineskip}
\textbf{Steps:}
Step 1: Understand the volume at infinity. We are given that $M \setminus U$ is isometric (identical in geometry) to $\mathbb{R}^n \setminus V$. This means for very large distances $r$, the volume of a ball $B(p, r)$ in $M$ grows exactly like a ball in $\mathbb{R}^n$. Mathematically, $\lim_{r \to \infty} \frac{\text{Vol}(B(p, r))}{r^n} = \omega_n$, where $\omega_n$ is the volume of the unit ball in flat $\mathbb{R}^n$.

Step 2: Apply the Bishop-Gromov Volume Comparison Theorem. For a manifold with non-negative Ricci curvature ($\text{Ric} \geq 0$), this theorem states that the ratio $\frac{\text{Vol}(B(p, r))}{r^n}$ is a non-increasing function of the radius $r$.

Step 3: Analyze the volume ratio for small $r$. For very small radii (as $r \to 0$), any Riemannian manifold looks flat, so the limit of this ratio is also $\omega_n$.

Step 4: Conclude that the ratio is constant. Since the ratio starts at $\omega_n$ for small $r$, is non-increasing, and ends up exactly at $\omega_n$ as $r \to \infty$, the ratio must be perfectly constant. That is, $\frac{\text{Vol}(B(p, r))}{r^n} = \omega_n$ for all $r$.

Step 5: Use the rigidity equality case. The equality case of the Bishop-Gromov theorem states that if a ball has the exact same volume as a Euclidean ball ($\text{Vol}(B(p, r)) = \omega_n r^n$) for all $r$, then the manifold must be perfectly flat and isometric to Euclidean space. Thus, $M$ is isometric to $\mathbb{R}^n$.
\end{solution}

\begin{exercise}
Compute all the homotopy groups of the $n$-torus $T^{n} = S^{1} \times S^{1} \times \cdots \times S^{1}$, $n \geq 2$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To understand this problem, you need to know about \textit{homotopy groups}, \textit{universal covers}, and the \textit{fundamental group}. Homotopy groups measure the different ways spheres of various dimensions can be mapped into a space. A universal cover is a simply connected space that "unrolls" the original space.

\vspace{\baselineskip}
\textbf{Intuition:}
Imagine a 2D torus (the surface of a donut). You can "unroll" it into a flat plane ($\mathbb{R}^2$), which is infinitely large and has no holes. This flat plane is called its universal cover. Because the plane has no holes, you cannot trap a balloon (a 2D sphere or higher) inside it. In topology, covering spaces share all the higher-dimensional hole structures (higher homotopy groups) with their base space. Since the flat plane has no higher-dimensional holes, neither does the torus!

\vspace{\baselineskip}
\textbf{Steps:}
Step 1: Identify the universal cover. The $n$-torus $T^n$ is defined as the product of $n$ circles: $S^1 \times S^1 \times \cdots \times S^1$. The universal cover of a single circle $S^1$ is the real line $\mathbb{R}$. Therefore, the universal cover of the $n$-torus is the product of $n$ real lines, which is $\mathbb{R}^n$.

Step 2: Determine the higher homotopy groups. A well-known theorem states that a space and its universal cover have identical higher homotopy groups ($\pi_k$ for $k \geq 2$). Since $\mathbb{R}^n$ is contractible (it can be continuously shrunk to a single point), all of its homotopy groups are zero. Thus, $\pi_k(T^n) \cong \pi_k(\mathbb{R}^n) = 0$ for all $k \geq 2$.

Step 3: Calculate the fundamental group (the first homotopy group). The fundamental group of a product space is the product of the fundamental groups. Since $\pi_1(S^1) \cong \mathbb{Z}$ (representing the number of times a loop winds around the circle), we have $\pi_1(T^n) \cong \pi_1(S^1)^n \cong \mathbb{Z}^n$.

Step 4: Classify the space. A space whose only non-trivial homotopy group is $\pi_1$ is called an Eilenberg-MacLane space. Therefore, $T^n$ is a $K(\mathbb{Z}^n, 1)$ space.
\end{solution}

\begin{exercise}
Consider the upper half space $\mathbb{H}^{3}$ with metric $g = \frac{dx^{2} + dy^{2} + dz^{2}}{z^{2}}$. Let $P$ be the surface defined by $\{z = x\tan\alpha, z > 0\}$. Compute the mean curvature of $P$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To understand this problem, you need to know about \textit{hyperbolic geometry} (specifically the upper half-space model), \textit{conformal metrics}, and \textit{mean curvature}. The upper half-space model represents hyperbolic space using Euclidean coordinates but with a distorted distance formula. Conformal metrics are metrics that are scaled versions of the flat Euclidean metric.

\vspace{\baselineskip}
\textbf{Intuition:}
In the standard flat world (Euclidean space), a flat plane obviously has zero curvature. However, in the hyperbolic world $\mathbb{H}^3$, distance is heavily distorted near the bottom ($z=0$). Because of this distortion, a tilted plane $z = x \tan \alpha$ actually looks "curved" from the perspective of the hyperbolic geometry, unless it is perfectly vertical. The mean curvature tells us exactly how much this plane bends within the distorted space.

\vspace{\baselineskip}
\textbf{Steps:}
Step 1: Relate the hyperbolic metric to the Euclidean metric. The given hyperbolic metric is $g = \frac{dx^2 + dy^2 + dz^2}{z^2}$. This can be written as $\bar{g} = \phi^2 g_E$, where $g_E$ is the standard Euclidean metric and the scaling factor is $\phi = 1/z$. Here, the dimension $n=3$.

Step 2: Use the formula for conformal change of mean curvature. When the metric is scaled by $\phi^2$, the new mean curvature $\bar{H}$ (defined here as the sum of the principal curvatures) is given by the formula: 
\[ \bar{H} = \frac{1}{\phi} \left( H_E + (n-1) \frac{\partial \ln \phi}{\partial n_E} \right) \]

Step 3: Evaluate the Euclidean terms. For our surface $P$, which is a flat plane in Euclidean coordinates, the Euclidean mean curvature $H_E$ is simply $0$.

Step 4: Compute the normal derivative of the scaling factor. We need the gradient of $\ln \phi$. Since $\ln \phi = \ln (1/z) = -\ln z$, its gradient is $\nabla \ln \phi = (0, 0, -1/z)$. 

Step 5: Find the unit normal vector. The surface is given by the equation $z - x \tan \alpha = 0$. The Euclidean unit normal vector $n_E$ to this plane, pointing upwards, is $n_E = (-\sin \alpha, 0, \cos \alpha)$.

Step 6: Calculate the directional derivative. We take the dot product of the gradient and the normal vector: $\frac{\partial \ln \phi}{\partial n_E} = \nabla \ln \phi \cdot n_E = (0, 0, -1/z) \cdot (-\sin \alpha, 0, \cos \alpha) = -\frac{\cos \alpha}{z}$.

Step 7: Substitute back into the main formula. Using $n=3$, $\phi = 1/z$, and $H_E=0$, we get:
\[ \bar{H} = z \left( 0 + 2 \left( -\frac{\cos \alpha}{z} \right) \right) = -2 \cos \alpha \]
(Note: If your convention defines mean curvature $H$ as the average rather than the sum of principal curvatures, you would divide by $2$, yielding $H = -\cos \alpha$.)
\end{solution}

\begin{exercise}
Suppose $M$ is a compact 2-dimensional Riemannian manifold without boundary, with positive sectional curvature. Show that any two compact closed geodesics on $M$ must intersect with each other.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
To understand this problem, you need to know about \textit{geodesics}, \textit{sectional curvature}, and the \textit{second variation of length}. A geodesic is the shortest path between points on a curved surface (like straight lines in flat space). Sectional curvature describes how the surface bends. The second variation formula tells us whether a geodesic path is truly the shortest or if it can be deformed into an even shorter path.

\vspace{\baselineskip}
\textbf{Intuition:}
Imagine the surface of a globe (which has positive curvature everywhere). The "straightest" possible lines (geodesics) are the great circles, like the equator and the lines of longitude. Notice that it is impossible to draw two great circles that do not cross each other! Positive curvature forces straight lines to eventually bend toward one another. This problem, known as Frankel's Theorem, proves that this intersection property holds true for any compact surface with positive curvature, not just perfect spheres.

\vspace{\baselineskip}
\textbf{Steps:}
Step 1: Set up a proof by contradiction. Assume the opposite of what we want to prove: suppose there are two closed geodesics, $\gamma_1$ and $\gamma_2$, that are completely disjoint (they never intersect).

Step 2: Find the shortest connecting path. Because the manifold and the geodesics are compact, there must be a measurable shortest distance between them. Let $\sigma$ be the shortest geodesic connecting a point on $\gamma_1$ to a point on $\gamma_2$, with a length $L > 0$. 

Step 3: Analyze the intersections. Because $\sigma$ is the absolutely shortest path between the two curves, it must hit both $\gamma_1$ and $\gamma_2$ at right angles (perpendicularly). 

Step 4: Use the second variation of length formula. We want to see what happens if we slightly slide $\sigma$ along the two geodesics. We define a vector field $X$ that points along $\gamma_1$ and is parallel transported along $\sigma$ to $\gamma_2$. The formula for how the length changes (the second derivative $L''(0)$) is:
\[ L''(0) = \int_0^L (|\dot{X}|^2 - R(X, \dot{\sigma}, \dot{\sigma}, X)) ds \]

Step 5: Evaluate the terms in the formula. Because $X$ is parallel transported along $\sigma$, its rate of change $\dot{X}$ is exactly $0$. The term $R(X, \dot{\sigma}, \dot{\sigma}, X)$ represents the curvature. Since the surface has strictly positive sectional curvature ($K > 0$), this term is strictly positive.

Step 6: Find the contradiction. Plugging these into our formula, we get $L''(0) = \int_0^L (0 - (\text{positive number})) ds$, which means $L''(0) < 0$. A negative second derivative means the length is at a maximum, not a minimum! This implies we could slide the path slightly to make it shorter, contradicting our assumption that $\sigma$ was already the shortest path. Therefore, the geodesics must intersect.
\end{solution}

\begin{exercise}
Suppose $\Sigma$ is a smooth compact embedded hypersurface without boundary in $\mathbb{R}^{n}$ for $n \geq 3$. Show that $\Sigma$ is orientable.
\end{exercise}

\begin{solution}
\textbf{Geometric Intuition:} An embedded hypersurface in $\mathbb{R}^n$ always has an "inside" and an "outside". This global separation allows us to pick a consistent normal direction everywhere.

\textbf{Formal Proof:}
By the \textit{Jordan-Brouwer Separation Theorem}, $\mathbb{R}^n \setminus \Sigma$ has exactly two components, one of which (the interior $D$) is bounded.
We can define a global transverse vector field by taking the outward unit normal to $D$. 
A hypersurface is orientable if and only if its normal bundle is trivial. 
Since $\mathbb{R}^n$ is orientable, the tangent bundle $T\mathbb{R}^n$ is trivial. On $\Sigma$, we have the decomposition $T\mathbb{R}^n |_\Sigma = T\Sigma \oplus N\Sigma$.
Since $N\Sigma$ has a non-vanishing section (the outward normal), it is trivial.
Thus $w_1(T\Sigma) = w_1(T\mathbb{R}^n|_\Sigma) - w_1(N\Sigma) = 0 - 0 = 0$.
Hence $\Sigma$ is orientable.
\end{solution}
