\section{2023 Geometry \& Topology}

\subsection{Exam}

\begin{exercise}
\label{geometry:2023:1}
(a) Let $G$ be a Lie group, $\mathfrak { g }$ be its Lie algebra. The Maurer-Cartan form on $G$ is the unique left-invariant $\mathfrak { g }$ -valued 1-form such that $\omega | _ { e } : T _ { e } G  \mathfrak { g }$ is the identity map, where $e$ is the identity in $G$ . Show that the Maurer-Cartan form $\omega$ satisfies the Maurer-Cartan equation $\begin{array} { r } { d \omega + \frac { 1 } { 2 } [ \omega , \omega ] = 0 } \end{array}$ .

(b) Let $G$ be a matrix group $G L ( n , \mathbb { R } )$ , give detailed computation to find the Maurer-Cartan form $\omega$ .

(c) Let $S E ( 3 )$ be the special Euclidean group, i.e. it contains all transformations of $\mathbb { R } ^ { 3 }$ (as 3-dim Euclidean space) of the form ${ \bf x } \mapsto t + A { \bf x }$ , where $\textbf { t } \in \mathbb { R } ^ { 3 }$ and $A \in S O ( 3 )$ . Find an expression of the Maurer Cartan form $\omega$ of $S E ( 3 )$ , also check it satisfies the standard Cartan structure equations in Eulidean space.
\end{exercise}

\begin{exercise}
\label{geometry:2023:2}
Let $M_1$, $M_2$ be closed, connected, oriented 3-manifolds with the first fundamental groups
$$
\pi_ {1} (M _ {1}) = \mathbb {Z} _ {3} \oplus \mathbb {Z} ^ {2}, \quad \pi_ {1} (M _ {2}) = \mathbb {Z} _ {6} \oplus \mathbb {Z} ^ {3},
$$
(a) Find all homology groups $H _ { n } ( M _ { 1 } , \mathbb { Z } )$ and $H _ { n } ( M _ { 2 } , \mathbb { Z } )$ .

(b) Find all homology groups $H _ { n } ( M _ { 1 } \times M _ { 2 } , \mathbb { Q } )$ .

(c) Does there exist a closed connected oriented 3-manifold $M$ with
$$
\pi_ {1} (M) = \mathbb {Z} _ {3} \oplus \mathbb {Z} ^ {2} \quad \mathrm {o r} \quad \pi_ {1} (M) = \mathbb {Z} _ {6} \oplus \mathbb {Z} ^ {3}?
$$
\end{exercise}

\begin{exercise}
\label{geometry:2023:3}
(a) Let $f$ be a diffeomorphism goup of a circle $S ^ { 1 }$ , assume $f$ has no fixed point and it is generated by a smooth vector field, show that $f$ must be conjugate to a rotation.

(b) Show that there is a diffeomorphism $f : S ^ { 1 } \to S ^ { 1 }$ , such that $f$ can not be generated by a smooth vector field but it is arbitrarily closed the identity map $i : S ^ { 1 } \to S ^ { 1 }$ in $C ^ { \infty }$ -topology.
\end{exercise}

\begin{exercise}
\label{geometry:2023:4}
(a) State the Leray-Hirsh theorem.

(b) Let
$$
\begin{array}{l} F l _ {k} (\mathbb {C} ^ {n}) = \{\mathrm {a l l k - f l a g s i n} \mathbb {C} ^ {n} \} \\ = \{(F _ {0}, \dots , F _ {k}) | F _ {i} \mathrm {i s a n} i \mathrm {- d i m s u b s p a c e o f} \mathbb {C} ^ {n}, \mathrm {s . t .} F _ {0} \subset F _ {1} \subset \dots \subset F _ {k} \subset \mathbb {C} ^ {n} \} \\ \end{array}
$$
Let $\Phi : F l _ { k } ( \mathbb { C } ^ { n } ) \to F l _ { k - 1 } ( \mathbb { C } ^ { n } )$ be the projection map sending a $k$ -flag $( F _ { 0 } , \cdots , F _ { k } )$ to a $( k - 1 )$ -flag $( F _ { 0 } , \cdots , F _ { k - 1 } )$ , it is known this is a fiber bundle. What is the fiber of $\Phi$ ?

(c) Compute the Euler Characteristic $\chi ( F l _ { n } ( \mathbb { C } ^ { n } ) )$ .
\end{exercise}

\begin{exercise}
\label{geometry:2023:5}
On the Euclidean space $\mathbb { R } ^ { n }$ , we consider an $n - 1$ form $\alpha$ , which is of class $C ^ { 1 }$ , such that both $\alpha$ and $d \alpha$ are in $L ^ { 1 }$ . Show that $\textstyle \int _ { \mathbb { R } ^ { n } } d \alpha = 0$ .
\end{exercise}

\begin{exercise}
\label{geometry:2023:6}
A complete Riemannian metric $g _ { i j }$ on a smooth manifold $M ^ { n }$ is called a gradient expanding Ricci soliton if there exists a smooth function $f$ on $M ^ { n }$ such that the Ricci tensor Ric of the metric $g$ is given by
$$
R i c + \operatorname {H e s s} f = \lambda g,
$$
for some negative constant $\lambda < 0$ . Show that if $M$ is compact, then a gradient expanding Ricci soliton must be an Einstein metric.
\end{exercise}

\newpage
\subsection{Solutions}

\begin{solution}
\textbf{Prerequisites \& Definitions:}
\begin{itemize}
    \item \textbf{Maurer-Cartan Form $\omega$}: A 1-form on a Lie group $G$ taking values in its Lie algebra $\mathfrak{g}$. It is defined as the left-invariant 1-form which is the identity on $T_e G$\footnote{Left-invariance means $L_g^* \omega = \omega$ for all $g \in G$, where $L_g$ is the left translation $L_g(h) = gh$.}.
    \item \textbf{Lie Algebra $\mathfrak{g}$}: The tangent space at the identity $T_e G$ equipped with a bracket $[\cdot, \cdot]$ that measures the non-commutativity of the group\footnote{For matrix groups, $[X, Y] = XY - YX$.}.
    \item \textbf{Exterior Derivative of Vector-valued Forms}: For $\omega \in \Omega^1(G, \mathfrak{g})$, $d\omega$ is a $\mathfrak{g}$-valued 2-form. For vector fields $X, Y$, $d\omega(X, Y) = X(\omega(Y)) - Y(\omega(X)) - \omega([X, Y])$.
    \item \textbf{The Bracket of 1-forms}: $[\omega, \omega](X, Y) = [\omega(X), \omega(Y)] - [\omega(Y), \omega(X)] = 2[\omega(X), \omega(Y)]$\footnote{This bracket combines the wedge product of the form components with the Lie bracket of the algebra components.}.
\end{itemize}

\textbf{Geometric Intuition:}
The Maurer-Cartan form serves as a "bridge" between the global geometry of the Lie group and the local structure of its Lie algebra. Since the group structure is uniform everywhere (due to left translation), we can identify any tangent space $T_g G$ with $T_e G$ by pushing vectors back using $dL_{g^{-1}}$. The Maurer-Cartan equation $d\omega + \frac{1}{2}[\omega, \omega] = 0$ is essentially a geometric manifestation of the Jacobi identity of the Lie algebra, translated into the language of differential forms. In matrix groups, this is simply the statement that $d(g^{-1}dg) = -g^{-1}dg \wedge g^{-1}dg$.

\textbf{Formal Proof:}

\textbf{(a) Proof of the Maurer-Cartan equation.}
Let $X, Y$ be left-invariant vector fields on $G$. By definition, the value of $\omega$ on a left-invariant vector field is constant:
\sidenote{$\omega(X) = X_e \in \mathfrak{g}$.}
\sidenote{$\omega(Y) = Y_e \in \mathfrak{g}$.}
Now we compute $d\omega$ acting on $X$ and $Y$ using the coordinate-free formula:
\sidenote{$d\omega(X, Y) = X(\omega(Y)) - Y(\omega(X)) - \omega([X, Y])$.}
Since $\omega(X)$ and $\omega(Y)$ are constant $\mathfrak{g}$-valued functions, their directional derivatives $X(\omega(Y))$ and $Y(\omega(X))$ vanish.
\sidenote{$d\omega(X, Y) = -\omega([X, Y])$.}
Because $X$ and $Y$ are left-invariant, their Lie bracket $[X, Y]$ is also left-invariant. Therefore:
\sidenote{$\omega([X, Y]) = [X, Y]_e = [X_e, Y_e] = [\omega(X), \omega(Y)]$.}
Substituting this into the equation for $d\omega$:
\sidenote{$d\omega(X, Y) + [\omega(X), \omega(Y)] = 0$.}
Using the definition of the bracket for 1-forms:
\sidenote{$\frac{1}{2}[\omega, \omega](X, Y) = [\omega(X), \omega(Y)]$.}
Combining these results, we obtain the Maurer-Cartan equation:
\sidenote{$d\omega + \frac{1}{2}[\omega, \omega] = 0$.}

\textbf{(b) Computation for $GL(n, \mathbb{R})$.}
For a matrix group $G \subset GL(n, \mathbb{R})$, let $g$ be the inclusion map (identity matrix coordinate). The differential $dg$ is the identity at each point in the matrix space.
Consider the $\mathfrak{g}$-valued 1-form $\omega = g^{-1}dg$.
First, we check its left-invariance. For $h \in G$, let $L_h(g) = hg$.
\sidenote{$L_h^* \omega = (hg)^{-1} d(hg) = g^{-1} h^{-1} h dg = g^{-1} dg = \omega$.}
Next, evaluate $\omega$ at the identity $e = I$:
\sidenote{$\omega|_I = I^{-1} dI = dI$.}
This is exactly the identity map on $T_I G \cong \mathfrak{gl}(n, \mathbb{R})$.
Thus, the Maurer-Cartan form for $GL(n, \mathbb{R})$ is $\omega = g^{-1}dg$.

\textbf{(c) Maurer-Cartan form of $SE(3)$.}
Elements of $SE(3)$ can be represented as $(3+1) \times (3+1)$ matrices:
\sidenote{$g = \begin{pmatrix} A & t \\ 0 & 1 \end{pmatrix}, \quad A \in SO(3), t \in \mathbb{R}^3$.}
The inverse is given by:
\sidenote{$g^{-1} = \begin{pmatrix} A^T & -A^T t \\ 0 & 1 \end{pmatrix}$.}
The differential is:
\sidenote{$dg = \begin{pmatrix} dA & dt \\ 0 & 0 \end{pmatrix}$.}
The Maurer-Cartan form $\omega = g^{-1}dg$ becomes:
\sidenote{$\omega = \begin{pmatrix} A^T dA & A^T dt \\ 0 & 0 \end{pmatrix}$.}
Let $\pi = A^T dA$ be the $so(3)$-valued "angular" part and $\sigma = A^T dt$ be the $\mathbb{R}^3$-valued "translational" part.
\sidenote{$\omega = \begin{pmatrix} \pi & \sigma \\ 0 & 0 \end{pmatrix}$.}
The Maurer-Cartan equation $d\omega + \omega \wedge \omega = 0$ yields:
\sidenote{$d\begin{pmatrix} \pi & \sigma \\ 0 & 0 \end{pmatrix} + \begin{pmatrix} \pi & \sigma \\ 0 & 0 \end{pmatrix} \wedge \begin{pmatrix} \pi & \sigma \\ 0 & 0 \end{pmatrix} = 0$.}
This gives two block equations:
\sidenote{$d\pi + \pi \wedge \pi = 0$ (Cartan's first structure equation for $SO(3)$).}
\sidenote{$d\sigma + \pi \wedge \sigma = 0$ (Cartan's second structure equation for Euclidean space).}
These match the standard structure equations where $\pi_{ij}$ are the connection forms and $\sigma_i$ are the coframe forms.
\end{solution}

\begin{solution}
\textbf{Prerequisites \& Definitions:}
\begin{itemize}
    \item \textbf{Fundamental Group $\pi_1$ and Homology $H_1$}: For a connected space, $H_1(M, \mathbb{Z})$ is the abelianization of $\pi_1(M)$\footnote{If $\pi_1(M)$ is already abelian, then $H_1(M, \mathbb{Z}) \cong \pi_1(M)$.}.
    \item \textbf{Poincaré Duality}: For a closed, connected, oriented $n$-manifold, $H_k(M, \mathbb{Z}) \cong H^{n-k}(M, \mathbb{Z})$\footnote{Specifically for $n=3$, $H_3 \cong \mathbb{Z}, H_2 \cong H^1$.}.
    \item \textbf{Universal Coefficient Theorem}: Relates cohomology to homology. For a manifold, $H^1(M, \mathbb{Z}) \cong \text{Hom}(H_1(M, \mathbb{Z}), \mathbb{Z})$\footnote{The torsion part of $H_1$ is killed by the Hom functor, leaving only the free part.}.
    \item \textbf{Künneth Formula}: For any field $\mathbb{F}$, $H_n(X \times Y, \mathbb{F}) \cong \bigoplus_{i+j=n} H_i(X, \mathbb{F}) \otimes H_j(Y, \mathbb{F})$.
\end{itemize}

\textbf{Geometric Intuition:}
The homology groups of a 3-manifold are highly constrained by its fundamental group and orientability. The 0th and 3rd groups are fixed for a closed connected oriented manifold. The 1st group is the abelianization of $\pi_1$. The 2nd group, via Poincaré duality and the UCT, captures the "free part" of the 1st group. Intuitively, every "hole" you can wrap a 1D loop around ($H_1$) corresponds to a 2D surface ($H_2$) that can wrap around it.

\textbf{Formal Proof:}

\textbf{(a) Homology groups of $M_1$ and $M_2$.}
For $M_1$, we have $\pi_1(M_1) = \mathbb{Z}_3 \oplus \mathbb{Z}^2$. Since this is abelian:
\sidenote{$H_1(M_1, \mathbb{Z}) = \mathbb{Z}_3 \oplus \mathbb{Z}^2$.}
As $M_1$ is closed, connected, and oriented:
\sidenote{$H_0(M_1, \mathbb{Z}) = \mathbb{Z}, \quad H_3(M_1, \mathbb{Z}) = \mathbb{Z}$.}
By Poincaré duality, $H_2(M_1, \mathbb{Z}) \cong H^1(M_1, \mathbb{Z})$. By the UCT:
\sidenote{$H^1(M_1, \mathbb{Z}) \cong \text{Hom}(\mathbb{Z}_3 \oplus \mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z}^2$.}
Thus, $H_2(M_1, \mathbb{Z}) = \mathbb{Z}^2$.
Similarly for $M_2$ with $\pi_1(M_2) = \mathbb{Z}_6 \oplus \mathbb{Z}^3$:
\sidenote{$H_0(M_2, \mathbb{Z}) = \mathbb{Z}, \quad H_1(M_2, \mathbb{Z}) = \mathbb{Z}_6 \oplus \mathbb{Z}^3, \quad H_2(M_2, \mathbb{Z}) = \mathbb{Z}^3, \quad H_3(M_2, \mathbb{Z}) = \mathbb{Z}$.}

\textbf{(b) Homology of $M_1 \times M_2$ over $\mathbb{Q}$.}
Over $\mathbb{Q}$, torsion vanishes. The Betti numbers for $M_1$ are $(1, 2, 2, 1)$ and for $M_2$ are $(1, 3, 3, 1)$.
Let $P_1(t) = 1 + 2t + 2t^2 + t^3 = (1+t)(1+t+t^2)$.
Let $P_2(t) = 1 + 3t + 3t^2 + t^3 = (1+t)^3$.
\sidenote{$P_{M_1 \times M_2}(t) = P_1(t)P_2(t) = (1+t)^4(1+t+t^2)$.}
\sidenote{$(1+4t+6t^2+4t^3+t^4)(1+t+t^2) = (1+t+t^2) + 4(t+t^2+t^3) + 6(t^2+t^3+t^4) + 4(t^3+t^4+t^5) + (t^4+t^5+t^6)$.}
Summing the terms by power of $t$:
\sidenote{$H_0: 1, \quad H_1: 1+4=5, \quad H_2: 1+4+6=11, \quad H_3: 4+6+4=14$.}
\sidenote{$H_4: 6+4+1=11, \quad H_5: 4+1=5, \quad H_6: 1$.}
So $H_n(M_1 \times M_2, \mathbb{Q})$ has dimensions $1, 5, 11, 14, 11, 5, 1$.

\textbf{(c) Existence of such 3-manifolds.}
We check if $\pi_1 = \mathbb{Z}_3 \oplus \mathbb{Z}^2$ is a valid fundamental group for a closed oriented 3-manifold.
If a closed oriented 3-manifold $M$ has an abelian fundamental group, it must be either finite (like a lens space $\mathbb{Z}_p$), or $\mathbb{Z}$ (like $S^2 \times S^1$), or $\mathbb{Z}^3$ (like $T^3$).
\footnote{This follows from the fact that if $\pi_1$ is abelian, $M$ is either prime or a connected sum. If prime, the Geometrization Theorem limits the possibilities. If it's a connected sum, the fundamental group is a free product, which is only abelian if it has only one non-trivial factor.}
The group $\mathbb{Z}_3 \oplus \mathbb{Z}^2$ is abelian but does not match any of these allowed structures. Thus, no such 3-manifold exists.
\end{solution}

\begin{solution}
\textbf{Prerequisites \& Definitions:}
\begin{itemize}
    \item \textbf{Diffeomorphism}: A smooth bijective map with a smooth inverse.
    \item \textbf{Vector Field Flow}: A vector field $v$ generates a 1-parameter group of diffeomorphisms $\phi_t$. A diffeomorphism $f$ is "generated by a vector field" if $f = \phi_1$ for some $v$\footnote{Such diffeomorphisms are always in the identity component of the diffeomorphism group.}.
    \item \textbf{Fixed Point}: A point $x$ such that $f(x) = x$.
    \item \textbf{Conjugacy}: Two diffeomorphisms $f, g$ are conjugate if there exists $h$ such that $f = h \circ g \circ h^{-1}$.
\end{itemize}

\textbf{Geometric Intuition:}
A vector field on a circle defines a "direction of flow" at every point. If the diffeomorphism $f$ is generated by this flow and has no fixed points, then the flow must be pushing every point in the same direction around the circle without ever stopping. By re-scaling our coordinate system to "even out" the speed of this flow, we can make it look like a constant speed rotation. Part (b) reminds us that even for maps close to the identity, being a flow is a very restrictive "rigid" property — most small perturbations will destroy the ability to represent a map as a continuous flow.

\textbf{Formal Proof:}

\textbf{(a) Conjugacy to rotation.}
Let $f = \phi_1$ be the flow of a smooth vector field $v$ on $S^1$.
Since $f$ has no fixed points, the vector field $v$ must be nowhere zero.
\sidenote{If $v(p) = 0$, then $\phi_t(p) = p$ for all $t$, so $f(p) = p$, contradicting the "no fixed point" assumption.}
Since $v$ is a smooth, non-vanishing vector field on $S^1$, we can define a new coordinate $\theta$ by:
\sidenote{$d\theta = \frac{c}{v(x)} dx$, where $c$ is chosen such that the total integral is $2\pi$.}
In this new coordinate system, the vector field $v$ becomes constant:
\sidenote{$v = c \frac{\partial}{\partial \theta}$.}
The flow of a constant vector field is simply a translation in that coordinate:
\sidenote{$\phi_t(\theta) = \theta + ct$.}
Setting $t=1$, we get $f(\theta) = \theta + c$, which is a rotation.
Thus, $f$ is conjugate to a rotation via the coordinate transformation $\theta$.

\textbf{(b) Existence of non-flow diffeomorphism.}
A diffeomorphism $f$ on $S^1$ is generated by a vector field flow if and only if it is either conjugate to a rotation or it satisfies certain constraints at its fixed points.
According to Kopell's Lemma, if $f$ is a flow and has a fixed point $p$, then the derivatives at the fixed point are constrained.
\sidenote{Specifically, if $f = \phi_1^v$ and $v(p) = 0$, then $f'(p) = e^{v'(p)}$.}
If we construct a diffeomorphism $f$ such that $f(p) = p, f'(p) = 1$, but $f''(p) \neq 0$ and $f$ is not the identity, it cannot be a flow of a \textit{smooth} vector field unless it satisfies infinitely many more conditions.
\sidenote{Consider $f(\theta) = \theta + \epsilon \sin^2(\theta/2)$. For small $\epsilon$, this is close to identity in $C^\infty$ but its fixed point behavior is inconsistent with a smooth flow.}
Such maps are arbitrarily close to the identity in $C^\infty$ topology but fail the "rigid" requirements of being a flow.
\end{solution}

\begin{solution}
\textbf{Prerequisites \& Definitions:}
\begin{itemize}
    \item \textbf{Leray-Hirsch Theorem}: A powerful tool for computing the cohomology of a fiber bundle when the cohomology of the fiber "survives" in the total space.
    \item \textbf{Fiber Bundle $F \to E \xrightarrow{\pi} B$}: A space $E$ that locally looks like $B \times F$.
    \item \textbf{Flag Manifold $Fl_k(\mathbb{C}^n)$}: The space of sequences of nested subspaces $0 \subset F_1 \subset \dots \subset F_k \subset \mathbb{C}^n$ where $\dim F_i = i$.
    \item \textbf{Euler Characteristic $\chi$}: A topological invariant. For a fiber bundle, $\chi(E) = \chi(B) \chi(F)$.
\end{itemize}

\textbf{Geometric Intuition:}
The Leray-Hirsch theorem says that if the cohomology of the fiber can be "pushed up" into the total space such that it still looks like a basis at every point, then the whole space just looks like a product of the base and the fiber from a cohomology perspective. Flag manifolds are naturally built this way: to pick a $k$-flag, you first pick a $(k-1)$-flag, and then you just need to pick one more line inside the remaining $(n-(k-1))$-dimensional space. This "picking a line" is exactly what a projective space does.

\textbf{Formal Proof:}

\textbf{(a) Leray-Hirsch Theorem.}
Let $\pi: E \to B$ be a fiber bundle with fiber $F$. Suppose there exist classes $c_1, \dots, c_r \in H^*(E; R)$ such that for each $b \in B$, the restrictions $i_b^*(c_1), \dots, i_b^*(c_r)$ form a basis for $H^*(F; R)$ as an $R$-module (where $i_b: F \to E$ is the inclusion of the fiber over $b$).
Then $H^*(E; R)$ is a free $H^*(B; R)$-module with basis $\{c_1, \dots, c_r\}$.
\sidenote{$H^*(E; R) \cong H^*(B; R) \otimes_R H^*(F; R)$ as $H^*(B; R)$-modules.}

\textbf{(b) The fiber of $\Phi$.}
The map $\Phi: Fl_k(\mathbb{C}^n) \to Fl_{k-1}(\mathbb{C}^n)$ takes a $k$-flag $(F_1, \dots, F_k)$ and maps it to $(F_1, \dots, F_{k-1})$.
To find the fiber over a fixed $(k-1)$-flag, we need to find all $k$-dimensional subspaces $F_k$ such that:
\sidenote{$F_{k-1} \subset F_k \subset \mathbb{C}^n$.}
This is equivalent to choosing a 1-dimensional subspace in the quotient space $\mathbb{C}^n / F_{k-1}$.
\sidenote{$\dim(\mathbb{C}^n / F_{k-1}) = n - (k-1) = n - k + 1$.}
The space of 1-dimensional subspaces in an $m$-dimensional complex space is $\mathbb{CP}^{m-1}$.
Substituting $m = n-k+1$, we find the fiber is:
\sidenote{Fiber $\cong \mathbb{CP}^{n-k}$.}

\textbf{(c) Euler Characteristic $\chi(Fl_n(\mathbb{C}^n))$.}
We can decompose $Fl_n(\mathbb{C}^n)$ into a sequence of fiber bundles:
\sidenote{$Fl_n \xrightarrow{\mathbb{CP}^0} Fl_{n-1} \xrightarrow{\mathbb{CP}^1} \dots \xrightarrow{\mathbb{CP}^{n-1}} \{pt\}$.}
Using the multiplicative property of the Euler characteristic $\chi(E) = \chi(B)\chi(F)$:
\sidenote{$\chi(Fl_n) = \chi(\mathbb{CP}^{n-1}) \cdot \chi(\mathbb{CP}^{n-2}) \cdot \dots \cdot \chi(\mathbb{CP}^0)$.}
Since $\chi(\mathbb{CP}^k) = k+1$:
\sidenote{$\chi(Fl_n) = n \cdot (n-1) \cdot \dots \cdot 1 = n!$.}
\end{solution}

\begin{solution}
\textbf{Prerequisites \& Definitions:}
\begin{itemize}
    \item \textbf{Stokes' Theorem}: $\int_\Omega d\alpha = \int_{\partial \Omega} \alpha$\footnote{For non-compact manifolds like $\mathbb{R}^n$, we must consider the limit of integrals over compact regions.}.
    \item \textbf{$L^1$ Space}: A function or form is in $L^1$ if its norm is integrable, $\int_{\mathbb{R}^n} |\alpha| < \infty$.
    \item \textbf{Exhaustion by Compact Sets}: We represent $\mathbb{R}^n$ as the limit of balls $B_R$ as $R \to \infty$.
\end{itemize}

\textbf{Geometric Intuition:}
The integral of a total derivative over the whole space should be determined by the "values at the boundary." Since $\mathbb{R}^n$ has no boundary in the traditional sense, we look at the limit as we go to infinity. If the form $\alpha$ is in $L^1$, it must "decay" on average as we move far away. If it decays sufficiently, its contribution on spheres of very large radius must vanish, leading to a zero total integral for $d\alpha$.

\textbf{Formal Proof:}
By definition of the integral over $\mathbb{R}^n$:
\sidenote{$\int_{\mathbb{R}^n} d\alpha = \lim_{R \to \infty} \int_{B_R} d\alpha$.}
Applying Stokes' Theorem to the ball $B_R$ of radius $R$:
\sidenote{$\int_{B_R} d\alpha = \int_{S_R} \alpha$, where $S_R$ is the sphere of radius $R$.}
Since $\alpha$ is in $L^1(\mathbb{R}^n)$, we have:
\sidenote{$\int_0^\infty \left( \int_{S_r} |\alpha| dS \right) dr = \int_{\mathbb{R}^n} |\alpha| dV < \infty$.}
Let $f(r) = \int_{S_r} |\alpha| dS$. Since the integral of $f(r)$ from 0 to $\infty$ converges, it is a property of integrable functions that the limit inferior must be zero:
\sidenote{$\liminf_{r \to \infty} f(r) = 0$.}
This implies there exists a sequence of radii $R_j \to \infty$ such that the integral over the spheres tends to zero:
\sidenote{$\lim_{j \to \infty} \int_{S_{R_j}} \alpha = 0$.}
Since $d\alpha$ is also in $L^1$, the limit $\lim_{R \to \infty} \int_{B_R} d\alpha$ exists. Therefore, this limit must equal the limit along the subsequence $R_j$:
\sidenote{$\int_{\mathbb{R}^n} d\alpha = \lim_{j \to \infty} \int_{S_{R_j}} \alpha = 0$.}
\end{solution}

\begin{solution}
\textbf{Prerequisites \& Definitions:}
\begin{itemize}
    \item \textbf{Ricci Soliton}: A generalization of Einstein metrics. It satisfies $Ric + \text{Hess} f = \lambda g$.
    \item \textbf{Einstein Metric}: A metric where $Ric = \lambda g$ for some constant $\lambda$. This corresponds to $f$ being constant in the soliton equation.
    \item \textbf{Expanding Soliton}: One where the constant $\lambda$ is negative ($\lambda < 0$).
    \item \textbf{Maximum Principle}: On a compact manifold, a function must attain its maximum and minimum. At these points, the gradient is zero and the Hessian has a definite sign.
\end{itemize}

\textbf{Geometric Intuition:}
A Ricci soliton can be thought of as a "steady state" of the Ricci flow. The term $\text{Hess} f$ represents a change in the metric due to a diffeomorphism. On a compact manifold, there is "no room" for the manifold to indefinitely expand in a non-trivial way while maintaining a gradient structure. The geometric constraints of compactness force the potential function $f$ to be "flat" (constant), meaning the expansion is uniform in all directions, which is exactly what an Einstein metric describes.

\textbf{Formal Proof:}
For a gradient Ricci soliton, we have the fundamental identity:
\sidenote{$R + |\nabla f|^2 - 2\lambda f = C$ for some constant $C$.}
Taking the trace of the soliton equation $Ric + \text{Hess} f = \lambda g$:
\sidenote{$R + \Delta f = n \lambda$.}
Let $p \in M$ be a point where $f$ attains its maximum, and $q \in M$ be a point where $f$ attains its minimum.
At the maximum $p$, we have $\nabla f(p) = 0$ and $\Delta f(p) \le 0$.
\sidenote{$R(p) = n\lambda - \Delta f(p) \ge n\lambda$.}
At the minimum $q$, we have $\nabla f(q) = 0$ and $\Delta f(q) \ge 0$.
\sidenote{$R(q) = n\lambda - \Delta f(q) \le n\lambda$.}
Now we use the identity $R + |\nabla f|^2 - 2\lambda f = C$ at these two points:
\sidenote{$R(p) - 2\lambda f(p) = C$.}
\sidenote{$R(q) - 2\lambda f(q) = C$.}
Subtracting these equations:
\sidenote{$R(p) - R(q) = 2\lambda (f(p) - f(q))$.}
From our inequalities, $R(p) \ge n\lambda \ge R(q)$, so the left-hand side is non-negative:
\sidenote{$R(p) - R(q) \ge 0$.}
However, on the right-hand side, since $p$ is the maximum and $q$ is the minimum, $f(p) \ge f(q)$. Since $\lambda < 0$ (expanding soliton):
\sidenote{$2\lambda (f(p) - f(q)) \le 0$.}
The only way for a non-negative number to equal a non-positive number is if both are zero:
\sidenote{$f(p) = f(q)$.}
This implies $f$ is constant throughout the connected manifold $M$.
If $f$ is constant, then $\text{Hess} f = 0$. The soliton equation then reduces to:
\sidenote{$Ric = \lambda g$.}
This is exactly the definition of an Einstein metric.
\end{solution}
