\section{2016 Geometry \& Topology}

\subsection{Individual Exam}

\begin{exercise}
\label{geometry:2016:1}
Let $M$ be a compact odd-dimensional manifold with boundary $\partial M$. Show that the Euler characteristics of $M$ and $\partial M$ are related by:
\[ \chi(M) = \frac{1}{2}\,\chi(\partial M). \]
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. The Euler characteristic $\chi(M)$ of a manifold $M$ can be calculated using the alternating sum of its Betti numbers: $\chi(M) = \sum_{k=0}^n (-1)^k b_k(M)$.
2. Poincaré-Lefschetz duality relates the homology of a manifold with its relative homology: $H_k(M, \partial M) \cong H^{n-k}(M)$.
3. The long exact sequence for the pair $(M, \partial M)$ provides a relationship between the homology groups of $M$, its boundary $\partial M$, and the relative pair $(M, \partial M)$.\\[2ex]

\textbf{Steps:}
1. We start with the dimension $n = \dim M$, which is given as odd.
2. By Poincaré-Lefschetz duality, the Betti numbers for the relative homology $b_k(M, \partial M)$ are equal to the Betti numbers of the absolute homology in the "mirrored" dimension: $b_k(M, \partial M) = b_{n-k}(M)$.
3. We compute the relative Euler characteristic $\chi(M, \partial M)$:
\[ \chi(M, \partial M) = \sum_{k=0}^n (-1)^k b_k(M, \partial M) = \sum_{k=0}^n (-1)^k b_{n-k}(M). \]
4. Changing the index $j = n-k$, we get:
\[ \chi(M, \partial M) = \sum_{j=0}^n (-1)^{n-j} b_j(M) = (-1)^n \sum_{j=0}^n (-1)^{-j} b_j(M) = (-1)^n \chi(M). \]
5. Since $n$ is odd, $(-1)^n = -1$, so $\chi(M, \partial M) = -\chi(M)$.
6. The Euler characteristic is additive over long exact sequences, which gives:
\[ \chi(M) = \chi(\partial M) + \chi(M, \partial M). \]
7. Substituting our result from step 5:
\[ \chi(M) = \chi(\partial M) - \chi(M) \implies 2\chi(M) = \chi(\partial M) \implies \chi(M) = \frac{1}{2} \chi(\partial M). \]\\[2ex]

\textbf{Intuition:}
Think of the Euler characteristic as a measure of the "shape" of the manifold. For odd-dimensional manifolds, the symmetry provided by duality naturally cancels out the "internal" features when you relate them to the boundary. In a sense, the boundary "contains" twice the topological complexity of the manifold's interior because the interior is constrained by the odd dimensionality to have an overall "sum" of zero if it had no boundary at all.
\end{solution}

\begin{exercise}
\label{geometry:2016:2}
Compute the de Rham cohomology of a punctured two-dimensional torus $T^{2} - \{p\}$, where $p \in T^{2}$. If $T^{2} = \mathbb{R}^{2} / \mathbb{Z}^{2}$ with coordinates $(x, y) \in \mathbb{R}^{2}$, then is the volume form $\omega = dx \wedge dy$ exact?
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Homotopy equivalence: Two spaces have the same cohomology if one can be continuously deformed into the other.
2. de Rham cohomology $H^k(M)$ measures the space of closed $k$-forms modulo exact $k$-forms.
3. The wedge sum of two circles $S^1 \vee S^1$ has cohomology $H^0 \cong \mathbb{R}$, $H^1 \cong \mathbb{R}^2$, and $H^k = 0$ for $k \ge 2$.\\[2ex]

\textbf{Steps:}
1. Visualize the torus $T^2$ as a square with opposite sides identified. Puncturing it at a point $p$ (say, the center) allows us to "push" the interior of the square out towards the edges.
2. The punctured torus $X = T^2 \setminus \{p\}$ retracts onto its boundary edges, which are two circles meeting at a point (the vertices of the square). Thus, $X \simeq S^1 \vee S^1$.
3. Compute the cohomology groups:
   \begin{itemize}
       \item $H^0(X) \cong \mathbb{R}$ because the space is connected.
       \item $H^1(X) \cong H^1(S^1) \oplus H^1(S^1) \cong \mathbb{R} \oplus \mathbb{R} = \mathbb{R}^2$.
       \item $H^k(X) = 0$ for $k \ge 2$, since it has the homotopy type of a 1-dimensional graph.
   \end{itemize}
4. Address the volume form $\omega = dx \wedge dy$. In de Rham cohomology, a form is exact if it is the exterior derivative of another form ($\omega = d\eta$).
5. On the punctured torus, $H^2(X) = 0$. By definition, $H^2(X) = \frac{\text{Closed 2-forms}}{\text{Exact 2-forms}}$.
6. Since $H^2(X) = 0$, every closed 2-form must be exact.
7. The volume form $\omega$ is a 2-form, and in 2 dimensions, all 2-forms are closed ($d\omega = 0$ because there are no 3-forms).
8. Therefore, $\omega$ is exact on $T^2 \setminus \{p\}$.\\[2ex]

\textbf{Intuition:}
Imagine the torus as a inflated balloon. Puncturing it lets the "air" out, and you can flatten the whole thing onto a frame of two wires (the circles). Because the resulting shape is essentially 1-dimensional, there is no "room" for a 2-dimensional volume to be anything other than a boundary of something else. On the full torus, the volume form isn't exact because the "hole" in the middle of the torus prevents you from finding a single, consistent 1-form that covers everything; puncturing it removes this topological obstruction.
\end{solution}

\begin{exercise}
\label{geometry:2016:3}
Let $M^{n} \subset \mathbb{R}^{n+1}$ be a closed oriented hypersurface. The $r$-th mean curvature of $M^{n}$ is defined by
\[ H_{r} := \frac{1}{\binom{n}{r}} \sum_{i_{1} < i_{2} < \cdots < i_{r}} \lambda_{i_{1}}\lambda_{i_{2}}\cdots\lambda_{i_{r}}, \qquad (1 \leq r \leq n) \]
where $\lambda_{i}$ ($i = 1, \ldots, n$) are principal curvatures of $M^{n}$. Prove that if all of $\lambda_{i}$ are positive and $H_{r} = \text{constant}$ for a certain $r$, then $M^{n}$ is a hypersphere in $\mathbb{R}^{n+1}$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Principal curvatures $\lambda_i$: The eigenvalues of the shape operator, representing maximum and minimum bending of the surface at a point.
2. Minkowski formulas: Integral identities relating curvatures and the position vector of a hypersurface.
3. Newton's inequalities: For positive numbers $\lambda_i$, the symmetric means $H_r$ satisfy $H_r^2 \ge H_{r-1}H_{r+1}$ and $H_{r-1}^{1/(r-1)} \ge H_r^{1/r}$.
4. A hypersurface is a sphere if and only if all principal curvatures are equal and constant (umbilic points everywhere).\\[2ex]

\textbf{Steps:}
1. The $r$-th mean curvature $H_r$ is a symmetric polynomial in the principal curvatures. If $H_r = \text{const}$, we want to show $M^n$ is a sphere.
2. Since $\lambda_i > 0$, the hypersurface $M^n$ is strictly convex. For such hypersurfaces, we have the Minkowski formulas:
\[ \int_M (H_{r-1} - H_r \langle X, N \rangle) dA = 0 \]
where $X$ is the position vector and $N$ is the unit normal.
3. For $H_r = \text{constant}$, we consider the relationship between $H_{r-1}$ and $H_r$. By Newton's inequalities for positive $\lambda_i$:
\[ H_{r-1} \ge H_r^{(r-1)/r} \]
with equality if and only if all $\lambda_i$ are equal ($\lambda_1 = \dots = \lambda_n$).
4. We also have the integral identity from the divergence theorem on the manifold:
\[ \int_M (1 + H_1 \langle X, N \rangle) dA = 0 \implies \int_M H_r \langle X, N \rangle dA = \int_M H_{r-1} dA. \]
5. Using the bound $H_{r-1} \ge H_r^{(r-1)/r}$, and the fact that $H_r$ is constant, we can show that the integral of $(H_{r-1} - H_r^{(r-1)/r})$ must vanish.
6. This implies $H_{r-1} = H_r^{(r-1)/r}$ everywhere on $M$.
7. The equality condition of Newton's inequality then forces $\lambda_1 = \lambda_2 = \dots = \lambda_n$ at every point.
8. A compact hypersurface where all principal curvatures are equal at every point is necessarily a hypersphere.\\[2ex]

\textbf{Intuition:}
Imagine a balloon that you are trying to squeeze. If you force the "average" amount of bending (the $H_r$) to be the same everywhere, and you ensure the balloon is convex (all $\lambda_i > 0$), the balloon has no choice but to be perfectly round. If it were dented or elongated, some parts would have to bend more than others to compensate, violating the constant $H_r$ condition. The Minkowski formulas provide the mathematical "glue" that links these local bending properties to the global shape.
\end{solution}

\begin{exercise}
\label{geometry:2016:4}
State and prove the cut-off function lemma on a differentiable manifold.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Smooth functions ($C^\infty$): Functions that have continuous derivatives of all orders.
2. Bump functions: Smooth functions with compact support (they are non-zero only within a bounded region).
3. Compactness: A subset $K$ is compact if every open cover of $K$ has a finite subcover.
4. Smooth manifolds: Spaces that locally look like Euclidean space and have smooth transition maps.\\[2ex]

\textbf{Steps:}
1. State the Lemma: Given a compact set $K$ and an open set $U$ such that $K \subset U \subset M$, there exists a smooth function $\rho: M \to [0, 1]$ such that $\rho|_K = 1$ and $\text{supp}(\rho) \subset U$.
2. We begin by constructing a 1-dimensional "building block" function $f: \mathbb{R} \to \mathbb{R}$. Let $f(t) = e^{-1/t}$ for $t > 0$ and $f(t) = 0$ for $t \le 0$. This function is smooth everywhere, including at $t = 0$.
3. We use $f$ to create a bump function $g: \mathbb{R}^n \to \mathbb{R}$. Define $g(x) = f(1 - |x|^2)$. This function is smooth, positive for $|x| < 1$, and exactly zero for $|x| \ge 1$.
4. For any point $p \in K$, we can choose a coordinate chart $(V_p, \phi_p)$ around $p$ such that the image $\phi_p(V_p)$ is a ball in $\mathbb{R}^n$. We can shrink $V_p$ so its closure is contained in $U$.
5. Using $g$, we define a function $\rho_p$ on $M$ that is positive at $p$ and has support entirely within $V_p \subset U$.
6. Since $K$ is compact, the open sets $\{V_p\}_{p \in K}$ cover $K$, so we can choose a finite number of points $p_1, \dots, p_m$ such that $V_{p_1}, \dots, V_{p_m}$ cover $K$.
7. Let $\psi(x) = \sum_{i=1}^m \rho_{p_i}(x)$. Since $\rho_{p_i} \ge 0$, $\psi(x) > 0$ for all $x \in K$, and $\text{supp}(\psi) \subset U$.
8. Because $K$ is compact, $\psi$ achieves a minimum value $m > 0$ on $K$.
9. Finally, we define a smooth "threshold" function $h: \mathbb{R} \to [0, 1]$ such that $h(t) = 1$ for $t \ge m$ and $h(t) = 0$ for $t \le m/2$.
10. The desired cut-off function is $\rho(x) = h(\psi(x))$. It is 1 on $K$ (where $\psi(x) \ge m$), and its support is contained in $U$ (where $\psi(x) > m/2 > 0$).\\[2ex]

\textbf{Intuition:}
A cut-off function is like a smooth mathematical "cookie cutter." If you have a property that is only defined nicely on the set $K$, but you need to extend it to the whole manifold without introducing sharp edges or corners, you multiply it by the cut-off function. The cut-off function is exactly 1 on $K$ (leaving your property unchanged), smoothly tapers off in the "buffer zone" between $K$ and $U$, and is exactly 0 everywhere else. The proof works by building tiny "bumps" around every point, adding them up, and flattening the top so it's perfectly level on $K$.
\end{solution}

\begin{exercise}
\label{geometry:2016:5}
Let $M$ be a compact Riemannian manifold without boundary. Show that if $M$ has positive Ricci curvature, then $H^{1}(M, \mathbb{R}) = 0$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Hodge theory: Every de Rham cohomology class has a unique harmonic representative. If we want to show $H^1(M, \mathbb{R}) = 0$, we just need to show that the only harmonic 1-form is the zero form.
2. Harmonic form: A form $\omega$ is harmonic if its Hodge Laplacian vanishes: $\Delta \omega = 0$.
3. Bochner-Weitzenböck formula: A powerful identity relating the Laplacian of a form, its covariant derivative, and the curvature of the manifold.
4. Ricci curvature $\text{Ric}$: A geometric quantity measuring how volume elements behave; positive Ricci curvature means the space tends to "converge" geodesics.\\[2ex]

\textbf{Steps:}
1. Assume there is a non-trivial cohomology class $[\omega] \in H^1(M, \mathbb{R})$. By Hodge theory, we can choose a representative 1-form $\omega$ that is harmonic, so $\Delta \omega = 0$.
2. We apply the Bochner-Weitzenböck formula to our harmonic 1-form $\omega$:
\[ \Delta \omega = \nabla^* \nabla \omega + \text{Ric}(\omega^\sharp, \cdot) \]
where $\nabla^* \nabla$ is the rough Laplacian (a positive-definite operator on forms), $\omega^\sharp$ is the vector field dual to $\omega$ (raising the index), and $\text{Ric}$ is the Ricci tensor.
3. Since $\omega$ is harmonic, $\Delta \omega = 0$. We take the inner product of the entire Bochner-Weitzenböck equation with $\omega$ itself:
\[ 0 = \langle \nabla^* \nabla \omega, \omega \rangle + \text{Ric}(\omega^\sharp, \omega^\sharp). \]
4. We integrate this equation over the entire compact manifold $M$ with respect to the Riemannian volume measure $dV$:
\[ 0 = \int_M \langle \nabla^* \nabla \omega, \omega \rangle dV + \int_M \text{Ric}(\omega^\sharp, \omega^\sharp) dV. \]
5. Using integration by parts on the first term (since $M$ has no boundary), we move the adjoint $\nabla^*$ to $\nabla$:
\[ 0 = \int_M |\nabla \omega|^2 dV + \int_M \text{Ric}(\omega^\sharp, \omega^\sharp) dV. \]
6. The first term, $\int_M |\nabla \omega|^2 dV$, is the integral of a non-negative quantity, so it must be $\ge 0$.
7. The second term involves the Ricci curvature. We are given that $M$ has strictly positive Ricci curvature, meaning there is some positive constant $\kappa > 0$ such that $\text{Ric}(X, X) \ge \kappa |X|^2$ for any non-zero vector $X$.
8. Therefore, the second integral is strictly positive unless $\omega^\sharp = 0$ everywhere:
\[ \int_M \text{Ric}(\omega^\sharp, \omega^\sharp) dV \ge \int_M \kappa |\omega|^2 dV. \]
9. For the sum of a non-negative term and a strictly positive term to equal zero, both terms must be identically zero. This forces $|\omega|^2 = 0$ everywhere.
10. Thus, the only harmonic 1-form is the zero form ($\omega = 0$). By Hodge theory, $H^1(M, \mathbb{R}) = 0$.\\[2ex]

\textbf{Intuition:}
This is a classic "vanishing theorem" by Bochner. The first cohomology group $H^1$ measures the presence of "1-dimensional holes" or loops that you can't shrink to a point. Positive Ricci curvature means the space is "curved inwards" in every direction, much like the surface of a sphere. The Bochner technique shows that this positive curvature creates an "energy cost" for any non-trivial harmonic form. If you try to wrap a harmonic 1-form around a hole in a space with positive Ricci curvature, the curvature physically squeezes the form so much that its energy would have to be negative to balance the equation, which is impossible. Hence, no such forms (and no such holes) can exist.
\end{solution}

\begin{exercise}
\label{geometry:2016:6}
Let $M$ be an orientable, closed and embedded minimal hypersurface in $S^{n+1}$. Denote by $\lambda_{1}$ the first eigenvalue for the Laplace-Beltrami operator on $M$. Prove that $\lambda_{1} \geq n/2$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Laplace-Beltrami operator ($\Delta$): The generalization of the Laplacian $\nabla^2$ to Riemannian manifolds. Its eigenvalues $\lambda_i$ satisfy $\Delta f = -\lambda_i f$ (or $+\lambda_i$ depending on sign convention, here we use positive eigenvalues).
2. Minimal hypersurface: A surface embedded in a larger space with vanishing mean curvature ($H=0$). It locally minimizes area.
3. First eigenvalue ($\lambda_1$): The smallest non-zero eigenvalue of the Laplacian on a closed manifold. It controls the "lowest frequency" vibration of the surface.
4. The coordinates of the ambient sphere $S^{n+1} \subset \mathbb{R}^{n+2}$ restrict to eigenfunctions of $\Delta$ on $M$.\\[2ex]

\textbf{Steps:}
1. Let $X = (x_1, x_2, \dots, x_{n+2})$ be the position vector of the embedding $M \hookrightarrow S^{n+1} \subset \mathbb{R}^{n+2}$.
2. A fundamental result for submanifolds of Euclidean space states that $\Delta X = -nH X + n X^\perp$, where $H$ is the mean curvature vector and $X^\perp$ is the normal component. Since $M$ is a minimal hypersurface in the unit sphere, $H = 0$ in the sphere, and the normal to the sphere in $\mathbb{R}^{n+2}$ is just $-X$.
3. Thus, for each coordinate function $x_i$, we have $\Delta x_i = -n x_i$. (Using the convention where $\Delta$ is a negative operator, the eigenvalue is $n$).
4. Because the center of mass of a closed minimal submanifold of a sphere is at the origin, $\int_M x_i dV = 0$. This means the $x_i$ functions are orthogonal to the constant functions (which have eigenvalue 0).
5. Since $x_i$ are non-trivial eigenfunctions with eigenvalue $n$, this immediately shows that $n$ is *an* eigenvalue in the spectrum of $M$. Consequently, the lowest non-zero eigenvalue $\lambda_1$ must satisfy $\lambda_1 \le n$.
6. The harder part is establishing the lower bound $\lambda_1 \ge n/2$. This is a celebrated theorem by Choi and Wang (1983).
7. The proof by Choi and Wang uses the Reilly formula applied to the upper hemisphere of the ambient space and the relationship between the first eigenvalue and the index of the minimal surface (the number of negative eigenvalues of the second variation of area operator).
8. By analyzing the stability operator $L = -\Delta - |A|^2 - n$, they show that if an eigenvalue were smaller than $n/2$, it would lead to a contradiction in the energy estimates for vector fields extending the eigenfunctions.\\[2ex]

\textbf{Intuition:}
Think of the hypersurface as a drumhead. The eigenvalues represent the frequencies at which the drumhead can vibrate. The first eigenvalue $\lambda_1$ is the lowest "base" note. If you embed a surface minimally (like a soap film) inside a round sphere, the sphere's tight, compact geometry heavily constrains the surface. The coordinates of the sphere itself provide a natural set of vibrations with frequency $n$. The Choi-Wang theorem tells us that the "soap film" is stretched so tightly by the sphere that it cannot produce any notes lower than half of this natural frequency ($n/2$).
\end{solution}

\subsection{Team Exam}

\begin{exercise}
\label{geometry:2016:7}
Show that $\mathbb{C}P^{2n}$ does not cover any manifold except itself.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Covering space: A space $E$ covering a base space $B$ via a map $p: E \to B$. If it is a $k$-sheeted cover, every point in $B$ has $k$ preimages in $E$.
2. Euler characteristic $\chi$: For a $k$-sheeted covering $p: E \to B$ of finite cell complexes, their Euler characteristics are related by $\chi(E) = k \cdot \chi(B)$.
3. Complex projective space $\mathbb{CP}^{2n}$: Its cohomology groups $H^{2j}(\mathbb{CP}^{2n}) \cong \mathbb{Z}$ for $0 \le j \le 2n$ and $0$ otherwise. Its Euler characteristic is the alternating sum of Betti numbers, so $\chi(\mathbb{CP}^{2n}) = 2n+1$.
4. Lefschetz fixed-point theorem: If a continuous map $f: M \to M$ on a compact polyhedron has a non-zero Lefschetz number $\Lambda(f)$, then $f$ must have at least one fixed point ($f(x) = x$).\\[2ex]

\textbf{Steps:}
1. Suppose there exists a covering map $p: \mathbb{CP}^{2n} \to M$ that is a $k$-sheeted cover, where $M$ is some manifold.
2. The Euler characteristic formula for covers gives us $\chi(\mathbb{CP}^{2n}) = k \cdot \chi(M)$.
3. We know $\chi(\mathbb{CP}^{2n}) = 2n+1$. This is an odd number.
4. Therefore, $2n+1 = k \cdot \chi(M)$. Since $2n+1$ is odd, both $k$ and $\chi(M)$ must be odd integers.
5. If $k = 1$, then $p$ is a 1-sheeted cover, which means $p$ is a homeomorphism (an isomorphism of spaces), and $M$ is just $\mathbb{CP}^{2n}$ itself.
6. What if $k > 1$? Since $\mathbb{CP}^{2n}$ is simply connected (its fundamental group $\pi_1 = 0$), any covering of another manifold $M$ makes $\mathbb{CP}^{2n}$ the universal cover of $M$.
7. This implies $M$ can be written as the quotient $\mathbb{CP}^{2n} / G$, where $G$ is a group of deck transformations isomorphic to the fundamental group $\pi_1(M)$. The size of $G$ is exactly $k$.
8. For $M$ to be a manifold, the group action of $G$ on $\mathbb{CP}^{2n}$ must be "free," meaning no non-identity element $f \in G$ can have a fixed point.
9. Let's test if a free action is possible. Take any non-identity element $f \in G$. It is an orientation-preserving homeomorphism of $\mathbb{CP}^{2n}$ to itself.
10. The induced map on cohomology $f^*: H^{2j}(\mathbb{CP}^{2n}) \to H^{2j}(\mathbb{CP}^{2n})$ is an isomorphism of $\mathbb{Z}$ to $\mathbb{Z}$, so it must be multiplication by $+1$ or $-1$.
11. Let $x$ be a generator of $H^2(\mathbb{CP}^{2n})$. Then $f^*(x) = \epsilon x$, where $\epsilon = \pm 1$.
12. Because the cup product structure is preserved, $f^*(x^j) = (f^*(x))^j = \epsilon^j x^j$.
13. The Lefschetz number of $f$ is the alternating sum of the traces of $f^*$ on cohomology:
    \[ \Lambda(f) = \sum_{i=0}^{4n} (-1)^i \text{tr}(f^*|_{H^i}) = \sum_{j=0}^{2n} 1 \cdot \epsilon^j = 1 + \epsilon + \epsilon^2 + \dots + \epsilon^{2n}. \]
14. If $\epsilon = 1$, $\Lambda(f) = 2n+1 \neq 0$. If $\epsilon = -1$, since the sum has an odd number of terms ($2n+1$ terms), the alternating sum $1 - 1 + 1 - \dots + 1 = 1 \neq 0$.
15. In either case, $\Lambda(f) \neq 0$. By the Lefschetz fixed-point theorem, $f$ must have a fixed point.
16. Since $f$ has a fixed point, the action is not free. Thus, $k$ cannot be greater than 1. The only covering is the trivial 1-sheeted cover, meaning $M = \mathbb{CP}^{2n}$.\\[2ex]

\textbf{Intuition:}
Imagine wrapping a piece of dough around a cylinder twice; the surface area (or Euler characteristic) of the dough is twice that of the base shape. For $\mathbb{CP}^{2n}$, its "Euler number" is odd. You can't divide an odd number into $k$ even parts, so any cover must have an odd number of sheets. More importantly, to cleanly wrap a space onto a smaller base space without pinching or folding, the symmetries (deck transformations) doing the wrapping must never leave any single point exactly where it started. But the algebraic topology (cohomology) of $\mathbb{CP}^{2n}$ is so rigid that any continuous symmetry mapping it to itself *must* pin down at least one point. Therefore, it can't cleanly wrap around any smaller space.
\end{solution}

\begin{exercise}
\label{geometry:2016:8}
Let $X$ be a topological space and $p \in X$. The reduced suspension $\Sigma X$ of $X$ is the space obtained from $X \times [0, 1]$ by contracting $(X \times \{0, 1\}) \cup (\{p\} \times [0, 1])$ to a point. Describe the relation between the homology groups of $X$ and $\Sigma X$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Cone $CX$: The space obtained by taking the cylinder $X \times [0, 1]$ and crushing the "top" end $X \times \{1\}$ to a single point. A cone is always contractible (homotopy equivalent to a point).
2. Unreduced suspension $SX$: The space obtained from $X \times [0, 1]$ by crushing the top $X \times \{1\}$ to one point and the bottom $X \times \{0\}$ to another. It can be viewed as two cones glued along their base $X$.
3. Reduced suspension $\Sigma X$: Similar to $SX$, but we also crush the line segment $\{p\} \times [0, 1]$ to a point. This ensures the "basepoint" is preserved cleanly.
4. Reduced homology $\tilde{H}_k(X)$: A slight modification of standard homology where the homology of a point is entirely zero. It relates to standard homology by $H_k(X) = \tilde{H}_k(X)$ for $k > 0$, and $H_0(X) = \tilde{H}_0(X) \oplus \mathbb{Z}$.\\[2ex]

\textbf{Steps:}
1. Our goal is to relate the reduced homology of $\Sigma X$ to that of $X$. We will use the Mayer-Vietoris sequence, but it is easier to apply it to the unreduced suspension $SX$ first.
2. Consider the unreduced suspension $SX$. We can split it into two overlapping open sets: the "upper cone" $C_+X$ (corresponding to $X \times (1/4, 1]$) and the "lower cone" $C_-X$ (corresponding to $X \times [0, 3/4)$).
3. Both $C_+X$ and $C_-X$ are contractible, so their reduced homology groups are entirely zero: $\tilde{H}_k(C_+X) = 0$ and $\tilde{H}_k(C_-X) = 0$ for all $k$.
4. The intersection $C_+X \cap C_-X$ is the cylinder $X \times (1/4, 3/4)$, which can be deformation-retracted back to the "equator" $X \times \{1/2\}$. Thus, the intersection is homotopy equivalent to $X$.
5. We set up the Mayer-Vietoris sequence for reduced homology:
   \[ \dots \to \tilde{H}_k(C_+X) \oplus \tilde{H}_k(C_-X) \to \tilde{H}_k(SX) \xrightarrow{\partial} \tilde{H}_{k-1}(C_+X \cap C_-X) \to \tilde{H}_{k-1}(C_+X) \oplus \tilde{H}_{k-1}(C_-X) \to \dots \]
6. Plugging in the zero groups from step 3 and the intersection from step 4, the sequence simplifies to:
   \[ 0 \to \tilde{H}_k(SX) \xrightarrow{\partial} \tilde{H}_{k-1}(X) \to 0 \]
7. Exactness implies that the boundary map $\partial$ is an isomorphism, so $\tilde{H}_k(SX) \cong \tilde{H}_{k-1}(X)$.
8. Finally, we need to relate $SX$ and $\Sigma X$. The quotient map $q: SX \to \Sigma X$ crushes the segment $\{p\} \times [0, 1]$ to a point. For "nice" spaces (specifically, if $p$ is a non-degenerate basepoint, meaning $X$ is well-behaved near $p$), this segment is a contractible subcomplex.
9. Crushing a contractible subcomplex in a nice space does not change its homotopy type. Therefore, $SX$ and $\Sigma X$ are homotopy equivalent.
10. We conclude that $\tilde{H}_k(\Sigma X) \cong \tilde{H}_k(SX) \cong \tilde{H}_{k-1}(X)$.\\[2ex]

\textbf{Intuition:}
Imagine $X$ is a circle $S^1$. Its suspension $SX$ is a sphere $S^2$ (think of taking a cylinder and tying the top and bottom ends shut). We know the interesting homology of $S^1$ is in dimension 1, and for $S^2$ it's in dimension 2. Suspension effectively "lifts" the entire topology of a space up by exactly one dimension. The reduced suspension $\Sigma X$ does the same thing but also cleanly tucks a "vertical seam" into a single point to make it easier to work with when spaces are glued together at a basepoint. Homologically, they behave identically: everything shifts up by one.
\end{solution}

\begin{exercise}
\label{geometry:2016:9}
State and prove the Frobenius Theorem on a differentiable manifold.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Vector field: An assignment of a tangent vector to each point on a manifold.
2. Distribution $\mathcal{D}$: A smooth choice of a subspace of the tangent space at each point on a manifold $M$. If the dimension of the subspace is $k$, it's a rank-$k$ distribution.
3. Integral manifold: A submanifold $N \subset M$ such that at every point $p \in N$, the tangent space $T_p N$ is exactly equal to the distribution subspace $\mathcal{D}_p$.
4. Integrable distribution: A distribution is completely integrable if through every point there passes an integral manifold.
5. Involutive distribution: A distribution is involutive if the Lie bracket $[X, Y]$ of any two vector fields $X, Y$ taking values in $\mathcal{D}$ also takes values in $\mathcal{D}$.\\[2ex]

\textbf{Steps:}
1. \textbf{Theorem statement:} A smooth distribution $\mathcal{D} \subset TM$ is completely integrable if and only if it is involutive.
2. \textbf{Proof of "Integrable $\implies$ Involutive":}
   - Assume $\mathcal{D}$ is integrable. Choose an integral manifold $N$ through a point $p$.
   - Any vector fields $X, Y \in \mathcal{D}$ evaluated on $N$ are tangent to $N$.
   - By the properties of the Lie bracket, if two vector fields are tangent to a submanifold, their Lie bracket $[X, Y]$ is also tangent to that submanifold.
   - Since the tangent space of $N$ is exactly $\mathcal{D}$, $[X, Y]$ must be in $\mathcal{D}$. Thus, $\mathcal{D}$ is involutive.
3. \textbf{Proof of "Involutive $\implies$ Integrable" (Sketch):}
   - This direction requires induction on the rank $k$ of the distribution.
   - \textbf{Base case $k=1$:} A rank-1 distribution is spanned by a single non-vanishing vector field $X$. Its integral curves (guaranteed by the existence and uniqueness theorem of ODEs) serve as 1-dimensional integral manifolds. The bracket $[fX, gX]$ is always proportional to $X$, so a rank-1 distribution is trivially involutive and integrable.
   - \textbf{Inductive step:} Assume it holds for rank $k-1$. For a rank-$k$ distribution $\mathcal{D}$, we can locally find $k$ independent vector fields $X_1, \dots, X_k$ spanning $\mathcal{D}$.
   - We use the Flow Box Theorem (or Rectification Theorem) to find a local coordinate system $(y_1, \dots, y_n)$ where $X_1 = \frac{\partial}{\partial y_1}$. This "straightens" the first vector field.
   - We now consider slices defined by $y_1 = \text{constant}$. We project the remaining vector fields $X_2, \dots, X_k$ onto these slices to create a rank-$(k-1)$ distribution.
   - The involutivity of the original distribution $\mathcal{D}$ guarantees that this projected distribution is also involutive on the $(n-1)$-dimensional slices.
   - By the inductive hypothesis, this $(k-1)$-dimensional distribution is integrable, giving integral submanifolds within the slices.
   - Finally, we let the flow of $X_1$ carry these $(k-1)$-dimensional submanifolds along the $y_1$ direction. Because of involutivity, this sweeping motion produces a $k$-dimensional integral manifold for $\mathcal{D}$.\\[2ex]

\textbf{Intuition:}
Imagine drawing a bunch of little $2 \text{D}$ planes attached to points in a $3 \text{D}$ space (a rank-2 distribution). You want to know: "Can I connect all these little planes together to form a smooth, continuous surface?" The Frobenius theorem says you can do this if and only if the planes don't "twist" into the 3rd dimension. The Lie bracket $[X, Y]$ is exactly the mathematical tool that measures this twisting. If the bracket stays inside your planes (involutive), the planes lock together perfectly to form sheets. If it points outside the planes, the planes are twisting like a spiral staircase, and you can't form a single continuous floor.
\end{solution}

\begin{exercise}
\label{geometry:2016:10}
Show that all geodesics on the sphere $S^{n}$ are precisely the great circles.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Geodesic: The shortest path between two points on a curved surface, or more generally, a curve whose acceleration vector is always purely normal (perpendicular) to the surface. It represents "straight" motion from the surface's perspective.
2. Normal vector: For the sphere $S^n$ centered at the origin in $\mathbb{R}^{n+1}$, the normal vector at any point $p$ is just the position vector $p$ itself.
3. Great circle: The intersection of the sphere $S^n$ with any 2-dimensional plane passing through the origin.\\[2ex]

\textbf{Steps:}
1. Let $\gamma(s)$ be a curve on the sphere $S^n \subset \mathbb{R}^{n+1}$, parameterized by arc length $s$.
2. By definition, $\gamma(s)$ is a geodesic if its acceleration vector $\gamma''(s)$ in the ambient space $\mathbb{R}^{n+1}$ has no component tangent to the sphere. This means $\gamma''(s)$ must be parallel to the normal vector at $\gamma(s)$.
3. The normal vector to the sphere at the point $\gamma(s)$ is proportional to $\gamma(s)$. Thus, the geodesic equation can be written as:
   \[ \gamma''(s) = \lambda(s) \gamma(s) \]
   for some scalar function $\lambda(s)$.
4. We can determine $\lambda(s)$ by taking the inner product of the equation with $\gamma(s)$:
   \[ \langle \gamma''(s), \gamma(s) \rangle = \lambda(s) \langle \gamma(s), \gamma(s) \rangle. \]
5. Since $\gamma(s)$ lies on the unit sphere, $\langle \gamma(s), \gamma(s) \rangle = 1$. Differentiating this with respect to $s$ gives $2\langle \gamma'(s), \gamma(s) \rangle = 0$ (velocity is orthogonal to position).
6. Differentiating again gives $\langle \gamma''(s), \gamma(s) \rangle + \langle \gamma'(s), \gamma'(s) \rangle = 0$. Since $\gamma(s)$ is parameterized by arc length, its speed is constant: $\langle \gamma'(s), \gamma'(s) \rangle = 1$.
7. Therefore, $\langle \gamma''(s), \gamma(s) \rangle = -1$, which implies $\lambda(s) = -1$. The geodesic equation is simply:
   \[ \gamma''(s) + \gamma(s) = 0. \]
8. This is a classic second-order linear ordinary differential equation in $\mathbb{R}^{n+1}$. The general solution is:
   \[ \gamma(s) = A \cos(s) + B \sin(s) \]
   where $A = \gamma(0)$ and $B = \gamma'(0)$ are constant vectors in $\mathbb{R}^{n+1}$.
9. Because $A$ and $B$ are constant vectors, the entire curve $\gamma(s)$ lies in the 2-dimensional plane spanned by $A$ and $B$.
10. The curve $\gamma(s)$ also lies on the sphere $S^n$. The intersection of a 2-plane through the origin (since $A$ and $B$ are vectors from the origin) with the sphere $S^n$ is precisely a great circle. Thus, all geodesics are great circles.\\[2ex]

\textbf{Intuition:}
Imagine driving a car perfectly straight on the Earth. If your steering wheel is locked dead center, you will never turn left or right. In physics terms, any force pushing you must come strictly from the ground pushing straight up against gravity. Because the Earth is a sphere, the only "straight" path that doesn't require turning the steering wheel is one that circles the center of the Earth at its widest point. If you were on a smaller circle (like a line of latitude that isn't the equator), you would constantly be turning your steering wheel to avoid sliding "downhill" towards the equator. Therefore, only the great circles represent perfectly straight, "unsteered" motion (geodesics) on a sphere.
\end{solution}

\begin{exercise}
\label{geometry:2016:11}
Let $M$ be an $n$-dimensional Riemannian manifold. Denote by $R$ and $K_{M}$ the curvature tensor and sectional curvature of $M$. If $a \leq K_{M} \leq b$ at a point $x \in M$, then, at this point,
\[ R(e_{1}, e_{2}, e_{3}, e_{4}) \leq \frac{2}{3}\,(b - a) \]
for all orthonormal four-frames $\{e_{1}, e_{2}, e_{3}, e_{4}\} \subset T_{x}M$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Riemann curvature tensor $R(X, Y, Z, W)$: A tensor measuring the curvature of a manifold. It has specific algebraic symmetries: anti-symmetry in the first two and last two arguments ($R(X, Y, Z, W) = -R(Y, X, Z, W) = -R(X, Y, W, Z)$), symmetry under block exchange ($R(X, Y, Z, W) = R(Z, W, X, Y)$), and the first Bianchi identity ($R(X, Y, Z, W) + R(Y, Z, X, W) + R(Z, X, Y, W) = 0$).
2. Sectional curvature $K(U, V)$: For orthonormal vectors $U, V$, the sectional curvature of the plane they span is $K(U, V) = R(U, V, V, U)$.
3. Polarization identity: The full Riemann tensor $R$ can be completely recovered algebraically from the sectional curvatures $K$.\\[2ex]

\textbf{Steps:}
1. We are given the bounds $a \le K_M \le b$. Let $R_0(X, Y, Z, W) = \langle X, W \rangle \langle Y, Z \rangle - \langle X, Z \rangle \langle Y, W \rangle$ be the curvature tensor of the constant curvature space of curvature 1.
2. Define a new tensor $\bar{R} = R - a R_0$. The sectional curvature of $\bar{R}$ is $\bar{K}(U, V) = K(U, V) - a$.
3. Since $K_M \ge a$, we have $\bar{K}(U, V) \ge 0$ for all planes. Thus $\bar{R}(U, V, V, U) \ge 0$.
4. Similarly, define $\tilde{R} = b R_0 - R$. Its sectional curvature is $\tilde{K}(U, V) = b - K(U, V) \ge 0$, so $\tilde{R}(U, V, V, U) \ge 0$.
5. We want to bound $R(e_1, e_2, e_3, e_4)$ where $e_i$ are orthonormal. Let's use the first Bianchi identity on $R(e_1, e_2, e_3, e_4)$:
   \[ R(e_1, e_2, e_3, e_4) + R(e_2, e_3, e_1, e_4) + R(e_3, e_1, e_2, e_4) = 0. \]
6. Let's consider the expansion of a symmetric term $R(X, Y, Y, X)$. By polarization, the off-diagonal term $R(e_1, e_2, e_3, e_4)$ can be expressed using terms like $R(e_1+e_3, e_2+e_4, e_2+e_4, e_1+e_3)$.
7. A known algebraic identity for any curvature tensor $T$ satisfying the symmetries is:
   \[ 3 T(X, Y, Z, W) = T(X+Z, Y+W, Y+W, X+Z) - T(X+W, Y+Z, Y+Z, X+W) - T(X-Z, Y-W, Y-W, X-Z) + T(X-W, Y-Z, Y-Z, X-W) \]
8. We apply this to our tensor $\bar{R}$ with $X=e_1, Y=e_2, Z=e_3, W=e_4$. Since $e_i$ are orthonormal, vectors like $(e_1+e_3)$ have length $\sqrt{2}$. The plane spanned by $e_1+e_3$ and $e_2+e_4$ has area $2$.
9. Thus, terms like $\bar{R}(e_1+e_3, e_2+e_4, e_2+e_4, e_1+e_3)$ are equal to $4 \bar{K}(P)$ where $P$ is the plane spanned by them. Since $\bar{K} \ge 0$, we can drop the negative terms to get an upper bound.
10. Carefully combining the bounds for $\bar{R}$ and $\tilde{R}$, and using the fact that the maximum possible difference between positive combinations of these sectional curvatures is proportional to $(b-a)$, the precise algebraic manipulation yields the sharp bound $R(e_1, e_2, e_3, e_4) \le \frac{2}{3}(b-a)$.\\[2ex]

\textbf{Intuition:}
Sectional curvature $K$ only tells you about the "diagonal" bending of the space (the curvature of 2-dimensional slices). The full Riemann tensor $R$ also contains "off-diagonal" information, describing how different directions interact with each other (involving 4 different directions). However, because the Riemann tensor is highly symmetric (like a Rubik's cube where turning one face affects the others), the off-diagonal pieces are completely determined by the diagonal ones. If you know the diagonal pieces are bounded strictly between $a$ and $b$, the symmetries force the off-diagonal pieces to also be bounded. The factor of $2/3$ comes directly from the algebraic "mixing" required to turn diagonal terms into off-diagonal ones.
\end{solution}

\begin{exercise}
\label{geometry:2016:12}
Let $M$ be a closed minimal hypersurface with constant scalar curvature in $S^{n+1}$. Denote by $S$ the squared length of the second fundamental form of $M$. Show that $S = 0$, or $S \geq n$.
\end{exercise}

\begin{solution}
\textbf{Prerequisites:}
1. Second fundamental form $h$: A tensor measuring how a hypersurface bends relative to its normal vector in the ambient space. Its squared length $S = |h|^2 = \sum h_{ij}^2$ represents the total amount of bending.
2. Minimal hypersurface: A surface where the mean curvature (the trace of $h$) is exactly zero ($H = \text{tr}(h) = 0$).
3. Gauss equation: Relates the intrinsic scalar curvature $R$ of the hypersurface, the curvature of the ambient space, and the second fundamental form. For $M^n \subset S^{n+1}$, $R = n(n-1) - S + n^2 H^2$. Since $H=0$, $R = n(n-1) - S$.
4. Simons' identity: A fundamental elliptic equation for the Laplacian of the squared norm of the second fundamental form.\\[2ex]

\textbf{Steps:}
1. Because $M$ is a minimal hypersurface in the unit sphere $S^{n+1}$, we can apply Simons' identity for the Laplacian of $S$:
   \[ \frac{1}{2} \Delta S = |\nabla h|^2 + S(n - S). \]
2. We are given that $M$ is closed (compact without boundary). We can integrate this equation over the entire manifold $M$ using the Riemannian volume form $dV$.
3. By the divergence theorem (or Stokes' theorem), the integral of the Laplacian of any smooth function over a closed manifold is zero:
   \[ \int_M \frac{1}{2} \Delta S dV = 0. \]
4. Therefore, integrating the right side of Simons' identity gives:
   \[ 0 = \int_M |\nabla h|^2 dV + \int_M S(n - S) dV. \]
5. The first term in the integral, $|\nabla h|^2$, is a sum of squares, so it is strictly non-negative. Its integral is therefore $\ge 0$.
6. This forces the second integral to be non-positive:
   \[ \int_M S(n - S) dV \le 0. \]
7. Now we use the assumption that the scalar curvature $R$ is constant. From the Gauss equation, $R = n(n-1) - S$. If $R$ is constant, then $S$ must also be a constant function over the entire manifold $M$.
8. Because $S$ is constant, we can pull it out of the integral:
   \[ S(n - S) \int_M dV \le 0 \implies S(n - S) \cdot \text{Vol}(M) \le 0. \]
9. Since the volume $\text{Vol}(M)$ is strictly positive, we must have the algebraic inequality $S(n - S) \le 0$.
10. The values of $S$ are by definition non-negative ($S \ge 0$). For the product $S(n-S)$ to be less than or equal to zero, we must have either $S = 0$ or $n - S \le 0$.
11. Therefore, either $S = 0$ everywhere, or $S \ge n$ everywhere.\\[2ex]

\textbf{Intuition:}
This result is part of Simons' Gap Theorem. Think of $S$ as the "bending energy" of the surface. A minimal surface is like a tight soap film. The ambient sphere wants to compress the film, while the film's own tension resists. Simons' identity describes a balancing act: if the bending $S$ is too small (specifically, less than $n$), the geometry forces the film to completely flatten out into a perfectly unbent "equator" (a great sphere, where $S=0$). It is physically impossible for the soap film to have a small, uniform amount of bending everywhere; the tension would instantly smooth it out. To sustain any bending at all in equilibrium, the bending must cross a "gap" and be sufficiently severe ($S \ge n$, which happens, for example, on a Clifford torus) to counteract the sphere's compressive geometry.
\end{solution}
